---
title: "Replication: Design and Evaluation of Optimal Free Trials"
subtitle: "Yoganarasimhan, Barzegary, and Pani (Management Science, 2023)"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/08_Heterogeneous_Treatment_Effects_and_Policy_Learning/yoganarasimhan2023design/"
  n_users: 337724
  sim_seed: 2023
  n_trees: 500
  n_lasso_reps: 5
  n_boot_rounds: 10
  n_threads: 8
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "90%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(causalmetrics)
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(mlr3)
library(mlr3learners)
library(glmnet)
data_path <- params$data_path
n_users <- params$n_users
sim_seed <- params$sim_seed
n_trees <- params$n_trees
n_threads <- params$n_threads
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
fmt0 <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
pct <- function(x, digits = 2) paste0(fmt(100 * x, digits), "\\%")
stars <- function(p) ifelse(p < 0.01, "***", ifelse(p < 0.05, "**", ifelse(p < 0.10, "*", "")))
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
arms <- c(7, 14, 30)
arm_lab <- paste0(arms, " days")
covs <- c("region", "os", "channel", "job", "skill", "segment")
```

# The paper in one page {#sec-paper}

**The question is how long a free trial should be, and whether the answer
should differ across users.** A SaaS firm that sold a 30-day trial ran a
field experiment on 337,724 new users, randomizing them to 7-, 14-, or
30-day trials (15 / 15 / 70 percent), and followed subscriptions,
subscription length, and revenue for two years. Six categorical variables
describe each user before the trial: region, operating system, sign-up
channel, job, skill, and business segment.

**The average answer is surprising: the shortest trial wins.** Seven days
raises subscriptions by about 5 percent over 30 days (Table 5 of the
paper), and also retention and revenue (Tables 9 and 10). The authors
spend Sections 4 and 6 on the mechanism, because the theory literature
predicts the opposite: longer trials increase usage, but they increase
dormancy at the end of the trial even more, and dormant users do not
subscribe (Figures 2 and 3, Tables 6 and 7).

**The personalized answer is built from an outcome model and evaluated
with an unbiased reward.** A lasso of the outcome on covariate dummies,
arm dummies, and their interactions predicts each user's outcome under
each arm; the policy assigns the arm with the highest prediction (equation
3). Any policy is then priced on a held-out test set with the
inverse-propensity-score (IPS) reward
$$
\hat R_{\text{IPS}}(\pi) = \frac{1}{N} \sum_{i=1}^{N} \frac{\mathbb{1}\{W_i = \pi(X_i)\}\, Y_i}{\hat e_{\pi(X_i)}(W_i)},
$$
which is unbiased because the experiment fixes the propensities. The
lasso policy gains 6.8 percent over 30-for-all and beats the best uniform
policy, 7-for-all (Table 8).

**The paper's second message is a warning.** Policies from linear
regression and random forests look best on the training data and worst on
the test data; CART and causal trees collapse to 7-for-all; causal forests
find too little heterogeneity (Table 8, Figure 5). Predictive accuracy is
not policy quality, and a uniform policy is a strong benchmark.

## What the replication package contains

**The public data carry no information: every column was replaced by
independent random draws.** The authors' `dataRandomizer.py` overwrites
each column except the user id with uniform random integers. The check
below, on the shipped `cleanDataset.csv`, shows a subscription rate of one
half in every arm, equal arm shares, and no relationship between any two
variables. The paper's numbers cannot be reproduced from these files, and
neither can its descriptives.

```{r shipped-check}
shipped <- read.csv(file.path(data_path, "cleanDataset.csv"))
shipped_check <- tibble(
  quantity = c("Users", "Subscription rate", "Share in 7-day arm", "Share in 30-day arm",
               "R2 of subscription on arm and covariates", "Largest |correlation| among outcomes and arm"),
  paper = c("337,724", "0.148", "0.151", "0.698", "small but positive", "positive"),
  shipped = c(fmt0(nrow(shipped)), fmt(mean(shipped$subscribed)), fmt(mean(shipped$trial_length == 0)),
              fmt(mean(shipped$trial_length == 2)),
              fmt(summary(lm(subscribed ~ factor(trial_length) + factor(country) + factor(skill) + factor(job), data = shipped))$r.squared, 5),
              fmt(max(abs(cor(shipped[, c("subscribed", "total_revenue", "total_month", "trial_length")])[upper.tri(diag(4))])), 4))
)
```

```{r tab-shipped, echo = FALSE}
kbl(shipped_check, booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Quantity", "Paper", "Shipped data"),
    caption = "The shipped data after the authors' randomization: no moment of the paper survives.") %>%
  tab_style()
```

**What the package does give, completely, is the procedure.** The Python
and R scripts specify every table and figure, the IPS estimator, the
argmax rule, the lasso specification with inverse-share weights and
repeated cross-validation, and the causal-forest setup with the propensity
fixed at the arm share. Two of the paper's arguments do not depend on the
firm's data at all: the IPS reward is unbiased under randomization, and a
policy from an overfitted outcome model looks best on training data and
worst on test data. Those are the arguments this note reproduces.

**Strategy.** Section \@ref(sec-population) builds a population that
matches every moment the paper publishes (sample size, arm shares,
covariate structure, subscription rates, the segment story of Section 6,
the usage regressions of Table 6, and the long-run outcomes), with the
true potential outcomes of every user known. The paper's procedure and our
package are then run on identical inputs: the authors' estimators are
matched numerically, each table is laid out as in the paper with the truth
as an extra column, and the shipped data are run once more at the end to
show what noise returns.

# A population calibrated to the paper {#sec-population}

## The simulator

**Each user has a subscription probability under each of the three
arms, and everything else is generated around it.** The baseline
probability depends on region, operating system, channel, and job (and on
a latent engagement that also drives usage). The arm effects follow the
paper's Section 6: beginners and students respond best to 7 days,
mixed-skill users and business professionals in the largest region to 14
days, experienced users to 30 days. Usage variables are drawn by arm and
skill so that Table 6's regressions reappear; subscription length and
revenue follow the paper's decomposition (equations 5 and 6): months
conditional on subscribing, times a monthly price that is lower for
students.

```{r simulator}
simulate_free_trial <- function(n, seed, par = list()) {
  set.seed(seed)
  p <- modifyList(list(
    base = -2.0,
    d7 = c(beginner = 0.10, mixed = -0.05, experienced = -0.12, unknown = 0.05, intermediate = 0.02),
    d14 = c(beginner = 0.00, mixed = 0.15, experienced = -0.08, unknown = 0.00, intermediate = 0.03),
    business14 = 0.13, student7 = 0.17, lat_sub = 0.35, lat_use = 0.25
  ), par)
  arm <- sample(arms, n, TRUE, c(0.1511, 0.1511, 0.6978))
  region <- sample(1:6, n, TRUE, c(55.02, 13.66, 9.12, 8.83, 8.0, 5.37))
  os <- sample(1:8, n, TRUE, c(28.97, 21.4, 14.04, 13.98, 9, 6, 4, 2.61))
  channel <- sample(1:42, n, TRUE, c(81.56, 8.14, 3.47, 0.81, rep(6.02 / 38, 38)))
  job_lab <- c("student", "business", "hobbyist", "designer", "photographer", "educator", "developer",
               "manager", "marketer", "artist", "engineer", "writer", "other", "unknown")
  job <- sample(job_lab, n, TRUE, c(28.20, 21.90, 20.34, 8.46, 4, 3, 3, 2.5, 2, 1.5, 1.2, 1, 1, 1.9))
  skill <- sample(c("beginner", "mixed", "experienced", "unknown", "intermediate"), n, TRUE,
                  c(69.05, 12.75, 10.77, 7.38, 0.05))
  segment <- sample(1:7, n, TRUE, c(35.41, 32.74, 18.40, 7.81, 3, 1.8, 0.84))
  latent <- rnorm(n)
  base_x <- p$base + 0.15 * (region == 1) - 0.1 * (region == 4) + 0.08 * (os %in% 1:2) +
    0.25 * (job == "business") - 0.15 * (job == "student") + 0.1 * (job == "designer") - 0.1 * (channel == 2) +
    0.2 * (skill == "mixed") + 0.15 * (skill == "experienced") + p$lat_sub * latent
  delta7 <- p$d7[skill] + p$student7 * (job == "student" & skill == "experienced")
  delta14 <- p$d14[skill] + p$business14 * (job == "business" & region == 1)
  pr_mat <- cbind(plogis(base_x + delta7), plogis(base_x + delta14), plogis(base_x))
  colnames(pr_mat) <- paste0("pr_", arms)
  act_mean <- list(beginner = c(1.55, 1.95, 2.55), mixed = c(2.4, 3.9, 6.2), experienced = c(2.0, 2.9, 4.4),
                   intermediate = c(1.9, 2.7, 3.9), unknown = c(1.7, 2.3, 3.2))
  am <- t(sapply(skill, function(s) act_mean[[s]])) * exp(p$lat_use * latent)
  u_use <- runif(n); u_pois <- runif(n); e_log <- rnorm(n, 0, 2.3); u_beta <- rbeta(n, 2.2, 1.05)
  usage <- lapply(seq_along(arms), function(k) {
    len <- arms[k]
    use <- as.integer(u_use < 0.828 + c(0, 0.004, 0.009)[k] + 0.03 * latent)
    active <- pmin(use * (1 + qpois(u_pois, pmax(am[, k] - 1, 0.05))), len)
    logu <- ifelse(use == 1, pmax(0, 4.35 + 0.24 * active + e_log), 0)
    dorm <- ifelse(use == 1, round(len * u_beta * (1 - 0.15 * pmin(pmax(latent, -2), 2))), len)
    list(use = use, active = active, logu = logu, dorm = pmin(pmax(dorm, 0), len))
  })
  k <- match(arm, arms)
  pick <- function(field) { M <- sapply(usage, `[[`, field); M[cbind(seq_len(n), k)] }
  sub_mat <- (runif(n) < pr_mat) * 1L
  subscribed <- sub_mat[cbind(seq_len(n), k)]
  months <- ifelse(subscribed == 1, pmin(round(rgamma(n, shape = 3.6, scale = 4.2) + 0.3 * (arm == 7)), 108), 0)
  price <- exp(rnorm(n, log(33), 0.45)) * ifelse(job == "student", 0.65, 1) * ifelse(segment == 1, 1.15, 1)
  revenue <- ifelse(subscribed == 1, round(months * price, 2), 0)
  d <- data.frame(user_id = seq_len(n), arm, region, os, channel, job, skill, segment,
                  use_software = pick("use"), active_days = pick("active"), log_usage = pick("logu"),
                  dormancy = pick("dorm"), subscribed, months, revenue)
  d <- cbind(d, pr_mat)
  d$best_arm <- arms[max.col(pr_mat)]
  d
}
pop <- simulate_free_trial(n_users, sim_seed)
pop$training <- as.integer(runif(n_users) < 0.7)
train <- pop[pop$training == 1, ]
test <- pop[pop$training == 0, ]
```

## Targets and achieved moments

**Table \@ref(tab:tab-targets) lists every published moment the
simulator was tuned to and what the simulated population delivers.** The
point of the table is honesty about what "calibrated" means: the arm
shares, covariate shares, and usage regressions are matched by
construction; the subscription rates, effects, segment shares, and
long-run outcomes are matched by the choice of a dozen parameters and are
close but not exact.

```{r targets, echo = FALSE}
sub_by_arm <- tapply(pop$subscribed, pop$arm, mean)
true_share <- prop.table(table(factor(pop$best_arm, levels = arms)))
usage_coef <- function(v) coef(lm(pop[[v]] ~ factor(pop$arm)))
targets <- tibble(
  Moment = c("Users", "Share 7 / 14 / 30 days", "Subscription rate", "7-day minus 30-day subscription rate",
             "14-day minus 30-day subscription rate", "Share of users whose best arm is 7 / 14 / 30",
             "Months subscribed, subscribers (mean, sd)", "Revenue, subscribers (mean, sd)", "Revenue, all users",
             "Active days: intercept, 14-day, 30-day", "Dormancy: intercept, 14-day, 30-day"),
  Paper = c("337,724", "15.1 / 15.1 / 69.8", "0.148", "0.006 to 0.008", "0.002 to 0.005",
            "68.9 / 23.2 / 7.9 (lasso policy)", "16.02, 8.43", "568, 552", "79.1",
            "1.75, 0.63, 1.71", "4.63, 5.14, 16.43"),
  Simulated = c(fmt0(nrow(pop)),
                paste(fmt(100 * prop.table(table(pop$arm)), 1), collapse = " / "),
                fmt(mean(pop$subscribed)),
                fmt(mean(pop$pr_7 - pop$pr_30), 4), fmt(mean(pop$pr_14 - pop$pr_30), 4),
                paste(fmt(100 * as.numeric(true_share), 1), collapse = " / "),
                paste0(fmt(mean(pop$months[pop$subscribed == 1]), 2), ", ", fmt(sd(pop$months[pop$subscribed == 1]), 2)),
                paste0(fmt(mean(pop$revenue[pop$subscribed == 1]), 0), ", ", fmt(sd(pop$revenue[pop$subscribed == 1]), 0)),
                fmt(mean(pop$revenue), 1),
                paste(fmt(usage_coef("active_days"), 2), collapse = ", "),
                paste(fmt(usage_coef("dormancy"), 2), collapse = ", "))
)
```

```{r tab-targets, echo = FALSE}
kbl(targets, booktabs = TRUE, linesep = "", escape = FALSE,
    caption = "Published moments used to calibrate the simulated population, and the values the simulation delivers. The best-arm shares are the shares of users whose true subscription probability is highest under each arm; the paper reports the shares its lasso policy assigns.") %>%
  tab_style() %>% column_spec(1, width = "6.2cm")
```

## The paper's descriptive tables

**Tables 1 and 2 of the paper establish the design before any analysis:
who got which arm and how the covariates are distributed.** Table 1 is
read down the columns: shares of observations and of subscriptions by arm
track each other, which is the first sign of small effects. Table 2 reports
only the number of subcategories and the top four shares, because
forty-two channels do not fit on a page.

```{r tab-design, echo = FALSE}
t1 <- pop %>% group_by(arm) %>% summarise(n = n(), subs = sum(subscribed), rate = mean(subscribed), .groups = "drop") %>%
  mutate(pct_obs = n / sum(n), pct_subs = subs / sum(subs))
t1_tab <- tibble(
  Row = c("Number of observations (N)", "Percent of total observations", "Number of subscriptions",
          "Percent of total subscriptions", "Subscription rate within group (in %)"),
  `7-day trial` = c(fmt0(t1$n[1]), fmt(100 * t1$pct_obs[1], 2), fmt0(t1$subs[1]), fmt(100 * t1$pct_subs[1], 2), fmt(100 * t1$rate[1], 2)),
  `14-day trial` = c(fmt0(t1$n[2]), fmt(100 * t1$pct_obs[2], 2), fmt0(t1$subs[2]), fmt(100 * t1$pct_subs[2], 2), fmt(100 * t1$rate[2], 2)),
  `30-day trial` = c(fmt0(t1$n[3]), fmt(100 * t1$pct_obs[3], 2), fmt0(t1$subs[3]), fmt(100 * t1$pct_subs[3], 2), fmt(100 * t1$rate[3], 2)),
  Total = c(fmt0(sum(t1$n)), "100", fmt0(sum(t1$subs)), "100", fmt(100 * mean(pop$subscribed), 2))
)
kbl(t1_tab, booktabs = TRUE, linesep = "", col.names = c("", names(t1_tab)[-1]),
    caption = "Summary statistics of treatment assignment and subscription rates in the simulated population (the layout of the paper's Table 1).") %>%
  tab_style()
```

```{r tab-covariates, echo = FALSE}
t2 <- bind_rows(lapply(c(region = "region", os = "os", channel = "channel", job = "job", skill = "skill", segment = "segment"), function(v) {
  sh <- sort(prop.table(table(pop[[v]])), decreasing = TRUE)
  tibble(n_cat = length(sh), first = sh[1], second = sh[2], third = sh[3], fourth = sh[4])
}), .id = "variable") %>%
  mutate(variable = c(region = "Geographic region", os = "Operating system", channel = "Signup channel",
                      job = "Job", skill = "Skill", segment = "Business segment")[variable])
t2 %>% mutate(across(first:fourth, ~ pct(.x))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Variable", "Number of subcategories", "First", "Second", "Third", "Fourth"),
      caption = "Summary statistics for the pretreatment categorical variables (the layout of the paper's Table 2): number of subcategories and shares of the four largest.") %>%
  add_header_above(c(" " = 2, "Share of top subcategories" = 4)) %>% tab_style()
```

**Table 3 of the paper reports the outcomes for all users and for
subscribers separately, because the long-run outcomes are zero for the 85
percent who never subscribe.** The subscriber rows are the ones that
carry the retention and revenue information.

```{r tab-outcomes, echo = FALSE}
desc <- function(x) c(mean = mean(x), sd = sd(x), min = min(x), q25 = unname(quantile(x, .25)), q50 = median(x), q75 = unname(quantile(x, .75)), max = max(x), n = length(x))
t3 <- rbind(
  "Subscription" = desc(pop$subscribed),
  "Subscription length (all)" = desc(pop$months),
  "Subscription length (subscribers)" = desc(pop$months[pop$subscribed == 1]),
  "Revenue (all)" = desc(pop$revenue),
  "Revenue (subscribers)" = desc(pop$revenue[pop$subscribed == 1])
)
t3 <- as.data.frame(t3); t3$Variable <- rownames(t3); rownames(t3) <- NULL
t3 %>% select(Variable, everything()) %>%
  mutate(across(c(mean, sd, min, q25, q50, q75, max), ~ fmt(.x, 2)), n = fmt0(n)) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Variable", "Mean", "Standard deviation", "Minimum", "25%", "50%", "75%", "Maximum", "Number of observations"),
      caption = "Summary statistics of subscription, subscription length, and revenue outcomes (the layout of the paper's Table 3).") %>%
  tab_style(c("scale_down"))
```

# Average effects and mechanism {#sec-effects}

## Average treatment effects

**Table 5 of the paper is a difference in means with a t-statistic and a
percentage gain over the 30-day baseline, reported for training and test
data separately.** In a randomized experiment nothing more is needed
(their footnote 5: regression-adjusted estimates are indistinguishable).
The split into training and test data appears already here because every
later policy comparison is made on the test data, and the reader must
know the test-set baseline.

```{r ate-table}
ate_panel <- function(d, var, digits = 4, mult = 1) {
  ctrl <- d[[var]][d$arm == 30] * mult
  bind_rows(lapply(arms, function(a) {
    y <- d[[var]][d$arm == a] * mult
    tt <- if (a == 30) NULL else t.test(y, ctrl)
    tibble(Treatment = paste0(a, " days"), rate = mean(y),
           diff = if (a == 30) NA else mean(y) - mean(ctrl),
           t = if (a == 30) NA else unname(tt$statistic),
           gain = if (a == 30) NA else 100 * (mean(y) - mean(ctrl)) / mean(ctrl))
  }))
}
t5 <- bind_rows(Training = ate_panel(train, "subscribed"), Test = ate_panel(test, "subscribed"), .id = "Data")
```

```{r tab-ate, echo = FALSE}
show_ate <- function(tab, caption, digits = 4, col1 = "Subscription rate", col2 = "Subscription rate difference") {
  tab %>% mutate(rate = fmt(rate, digits), diff = ifelse(is.na(diff), "---", fmt(diff, digits)),
                 t = ifelse(is.na(t), "---", fmt(t, 2)), gain = ifelse(is.na(gain), "---", fmt(gain, 2))) %>%
    kbl(booktabs = TRUE, linesep = "", escape = FALSE,
        col.names = c("Data", "Treatment", col1, col2, "t statistics", "Percentage gain over baseline"),
        caption = caption) %>%
    tab_style() %>% collapse_rows(columns = 1, valign = "top", latex_hline = "major")
}
show_ate(t5, "Average effect of the 7- and 14-day treatments on subscription compared with the control condition of the 30-day free trial (the layout of the paper's Table 5).")
```

**The doubly robust score with the known propensity gives the same
answer with a covariate adjustment.** The experiment fixes
`e_k = P(W = k)`, so `dr_scores()` with the arm shares as propensities
returns a score matrix whose column means are the expected outcomes under
each arm; their differences are the ATEs of Table
\@ref(tab:tab-ate) with a regression adjustment for the covariates, which
changes the estimates by less than a standard error, as the paper's
footnote 5 reports for its regression-adjusted version.

```{r ate-dr, cache = TRUE, cache.extra = list(sim_seed, n_users)}
design_e <- function(d) matrix(rep(prop.table(table(factor(d$arm, levels = arms))), each = nrow(d)), nrow(d),
                               dimnames = list(NULL, arms))
add_dummies <- function(d) {
  mm <- model.matrix(~ factor(region) + factor(os) + factor(channel) + factor(job) + factor(skill) + factor(segment), d)[, -1]
  colnames(mm) <- paste0("z", seq_len(ncol(mm)))
  cbind(d, mm)
}
train_d <- add_dummies(train)
test_d <- add_dummies(test)
zcols <- grep("^z", names(train_d), value = TRUE)
sc_train <- dr_scores(train_d, "subscribed", "arm", zcols, arms = arms, p_hat = design_e(train_d),
                      learner_mu = lrn("regr.lm"), folds = 2, seed = 1)
sc_test <- dr_scores(test_d, "subscribed", "arm", zcols, arms = arms, p_hat = design_e(test_d),
                     learner_mu = lrn("regr.lm"), folds = 2, seed = 2)
sc_test$ate$contrasts
```

## Randomization check

**The paper regresses the arm on all covariate dummies and tests them
jointly, rather than showing a table of means, because six categorical
variables with up to 42 levels do not fit in a balance table.** The F
test below does the same on the simulated population for the 7-day and
the 14-day indicators; the arms were assigned at random, so the covariates
predict neither at the 5 percent level.

```{r randomization-check}
pop_d <- add_dummies(pop)
sapply(c(`7-day arm` = 7, `14-day arm` = 14), function(a) {
  f <- summary(lm(I(pop_d$arm == a) ~ as.matrix(pop_d[, zcols])))$fstatistic
  c(F = unname(f[1]), p = unname(pf(f[1], f[2], f[3], lower.tail = FALSE)))
})
```

## Heterogeneity by region, skill, and job

**Figure 4 of the paper shows subscription rates by arm within
categories to motivate personalization before any model is fitted.** Bars
by arm inside each category, with stars where the 7- or 14-day rate
differs from the 30-day rate, let the reader see that beginners respond
to 7 days and mixed-skill users to 14 days. The reader also sees the
category shares under the bars, because a segment that responds
differently only matters if it is large.

```{r fig-heterogeneity, echo = FALSE, fig.height = 5.5, fig.cap = "Heterogeneity in users' response to the three trial lengths within region, skill, and job (the layout of the paper's Figure 4). Bars are subscription rates by arm; stars mark 7- or 14-day rates that differ from the 30-day rate in the same category (* p < 0.10, ** p < 0.05, *** p < 0.01); the share of users in each category is printed under its label."}
het_frame <- function(d, var, top = 8) {
  cats <- names(sort(table(d[[var]]), decreasing = TRUE))[seq_len(min(top, length(unique(d[[var]]))))]
  bind_rows(lapply(cats, function(cat) {
    sub <- d[d[[var]] == cat, ]
    ctrl <- sub$subscribed[sub$arm == 30]
    bind_rows(lapply(arms, function(a) {
      y <- sub$subscribed[sub$arm == a]
      p <- if (a == 30 || length(y) < 30) NA else t.test(y, ctrl)$p.value
      tibble(category = paste0(cat, "\n(", fmt(100 * nrow(sub) / nrow(d), 1), "%)"), arm = a, rate = mean(y), p = p)
    }))
  })) %>% mutate(star = ifelse(is.na(p), "", stars(p)), variable = var)
}
het <- bind_rows(het_frame(pop, "region"), het_frame(pop, "skill"), het_frame(pop, "job", top = 8)) %>%
  mutate(variable = factor(variable, levels = c("region", "skill", "job"),
                           labels = c("Geographic region", "Skill", "Job")),
         arm = factor(arm, levels = arms, labels = arm_lab))
ggplot(het, aes(category, rate, fill = arm)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_text(aes(label = star, y = rate + 0.004), position = position_dodge(width = 0.8), size = 2.8) +
  facet_wrap(~ variable, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = c("#4D4D4D", "#999999", "#D9D9D9"), name = NULL) +
  labs(x = NULL, y = "Subscription rate") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom", axis.text.x = element_text(size = 7))
```

## Mechanism: usage and dormancy

**Table 6 of the paper regresses each usage variable on the arm to show
that longer trials raise usage, but sublinearly, and raise dormancy
sharply.** Because the arm is randomized these regressions are causal, so
the table is the causal half of Figure 3; the second half, from usage to
subscription, is only an association, which the paper marks by presenting
Table 7 as suggestive (its footnote 8).

```{r tab-usage, echo = FALSE}
usage_vars <- c(use_software = "Indicator for software use", active_days = "Number of active days",
                log_usage = "Log usage during trial", dormancy = "Dormancy length")
t6 <- bind_rows(lapply(names(usage_vars), function(v) {
  m <- lm(pop[[v]] ~ factor(pop$arm))
  s <- summary(m)$coefficients
  tibble(Outcome = usage_vars[[v]],
         Intercept = paste0(fmt(s[1, 1], 3), " (", fmt(s[1, 2], 3), ")"),
         `14-day trial` = paste0(fmt(s[2, 1], 3), " (", fmt(s[2, 2], 3), ")"),
         `30-day trial` = paste0(fmt(s[3, 1], 3), " (", fmt(s[3, 2], 3), ")"),
         R2 = fmt(summary(m)$r.squared, 3), N = fmt0(nrow(pop)))
}))
kbl(t6, booktabs = TRUE, linesep = "", col.names = c("Outcome variable", "Intercept", "14-day trial", "30-day trial", "R2", "N"),
    caption = "Regression of usage features on trial length (the layout of the paper's Table 6). Standard errors in parentheses.") %>%
  tab_style()
```

**Figure 2 of the paper plots the subscription rate against activity for
each arm, and the pattern it wants the reader to see is that at the same
activity level shorter trials convert better.** The left panel uses the
number of active days; the right panel uses the last active day, so that
dormancy, which is the trial length minus the last active day, is visible
as distance from the right end of each curve.

```{r fig-usage, echo = FALSE, fig.height = 3.4, fig.cap = "Relationship between usage and subscription rate for the three trial lengths (the layout of the paper's Figure 2). Left: by the number of active days. Right: by the last active day (trial length minus dormancy)."}
u1 <- pop %>% filter(active_days <= 14) %>% group_by(arm, active_days) %>%
  summarise(rate = mean(subscribed), n = n(), .groups = "drop") %>% filter(n >= 200) %>%
  mutate(panel = "(a) Total active days", x = active_days)
u2 <- pop %>% mutate(last_active = arm - dormancy) %>% group_by(arm, last_active) %>%
  summarise(rate = mean(subscribed), n = n(), .groups = "drop") %>% filter(n >= 200) %>%
  mutate(panel = "(b) Last active day", x = last_active)
bind_rows(u1, u2) %>% mutate(arm = factor(arm, levels = arms, labels = arm_lab)) %>%
  ggplot(aes(x, rate, colour = arm)) + geom_line() + geom_point(size = 1) +
  facet_wrap(~ panel, scales = "free_x") +
  scale_colour_manual(values = c("#B22222", "#E69F00", "#0072B2"), name = NULL) +
  labs(x = NULL, y = "Subscription rate") + theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

**Table 7 of the paper is a logit of subscription on the arm, the usage
variables, and the covariates, and its purpose is to show that dormancy
has a negative coefficient once usage is held fixed.** With usage held
fixed the arm coefficients turn positive for the longer trials: the
longer trial is not harmful in itself, its dormancy is.

```{r tab-logit, echo = FALSE}
lg <- glm(subscribed ~ factor(arm) + active_days + log_usage + dormancy + use_software +
            factor(region) + factor(skill) + factor(job) + factor(os), data = pop, family = binomial())
lg_s <- summary(lg)$coefficients
rows <- c("factor(arm)14", "factor(arm)30", "active_days", "log_usage", "dormancy", "use_software")
tibble(Variable = c("14-day trial", "30-day trial", "Number of active days", "Log usage during trial", "Dormancy length", "Indicator for software use"),
       Coefficient = fmt(lg_s[rows, 1], 4), `Standard error` = fmt(lg_s[rows, 2], 4), z = fmt(lg_s[rows, 3], 2),
       p = fmt(lg_s[rows, 4], 3)) %>%
  kbl(booktabs = TRUE, linesep = "",
      caption = "Logit of subscription on trial length and usage features, with region, skill, job, and operating system as controls (the layout of the paper's Table 7; the covariate rows are omitted as in the paper).") %>%
  tab_style()
```

# Policy design {#sec-design}

## The authors' lasso policy

**The paper's policy is an outcome model, not a treatment-effect model:
predict the subscription probability under each arm and assign the best
arm (equation 3).** The model is a lasso of the outcome on arm dummies,
covariate dummies, and all arm-by-covariate interactions, so that each
arm can have its own coefficient on each category. Two details matter.
Observations are weighted by the inverse of their arm share, because the
30-day arm has five times more users and would otherwise dominate the
interaction coefficients of the other two arms. The penalty is chosen by
five-fold cross-validation and the whole procedure is repeated (50 times
in the paper), because the selected penalty varies across folds; the paper
reports the run with the median test-set value.

```{r lasso-design}
to_factors <- function(d) {
  for (v in covs) d[[v]] <- factor(d[[v]], levels = sort(unique(pop[[v]])))
  d$arm <- factor(d$arm, levels = arms)
  d
}
lasso_formula <- ~ arm * (region + os + channel + job + skill + segment) - 1
lasso_X <- function(d) Matrix::sparse.model.matrix(lasso_formula, to_factors(d))
arm_share <- function(d) prop.table(table(factor(d$arm, levels = arms)))
train$e <- as.numeric(arm_share(train)[as.character(train$arm)])
test$e <- as.numeric(arm_share(test)[as.character(test$arm)])
predict_arms <- function(pred_fun, d) {
  out <- sapply(arms, function(a) { d$arm <- a; pred_fun(d) })
  colnames(out) <- arms
  out
}
argmax_policy <- function(pred) arms[max.col(pred, ties.method = "first")]
fit_lasso_policy <- function(train, test, y = "subscribed", reps = 1, seed = 1) {
  Xtr <- lasso_X(train)
  lapply(seq_len(reps), function(r) {
    set.seed(seed + r)
    cv <- cv.glmnet(Xtr, train[[y]], weights = 1 / train$e, alpha = 1, nfolds = 5, type.measure = "mse")
    fit <- glmnet(Xtr, train[[y]], weights = 1 / train$e, alpha = 1, lambda = cv$lambda.min)
    pf <- function(d) as.numeric(predict(fit, lasso_X(d)))
    list(lambda = cv$lambda.min, n_nonzero = sum(coef(fit) != 0),
         pred_train = predict_arms(pf, train), pred_test = predict_arms(pf, test))
  })
}
```

```{r lasso-fit, cache = TRUE, cache.extra = list(sim_seed, n_users, params$n_lasso_reps)}
lasso_runs <- fit_lasso_policy(train, test, reps = params$n_lasso_reps, seed = 100)
```

**Repeated cross-validation selects a slightly different penalty each
time, and the paper reports the median run to avoid picking a lucky
one.** The runs below differ in the penalty and in the number of nonzero
coefficients; the test-set value of each run's policy is computed in
Section \@ref(sec-evaluation) and the median run is carried forward.

```{r lasso-runs, echo = FALSE}
lasso_summary <- tibble(run = seq_along(lasso_runs),
                        lambda = sapply(lasso_runs, `[[`, "lambda"),
                        nonzero = sapply(lasso_runs, `[[`, "n_nonzero"),
                        share_7_test = sapply(lasso_runs, function(r) mean(argmax_policy(r$pred_test) == 7)))
kbl(lasso_summary, booktabs = TRUE, linesep = "", digits = c(0, 6, 0, 3),
    col.names = c("Run", "Selected penalty", "Nonzero coefficients", "Share assigned 7 days (test)"),
    caption = "Repeated cross-validated lasso fits of the authors' specification.") %>% tab_style()
```

## Alternative outcome models

**The paper fits the same outcome model with four other learners, to
show that the choice of learner is a choice of policy.** Linear regression
is the lasso without the penalty. CART is a single regression tree pruned
by cross-validation. The random forest and XGBoost use the authors' tuned
settings (1,000 trees with a minimum leaf of 70; boosted trees of depth 6
with the logistic objective). All four get the arm as a feature together
with the covariate dummies and predict each user's outcome under each arm.

```{r outcome-models, cache = TRUE, cache.extra = list(sim_seed, n_users, n_trees)}
feat <- c("arm", zcols)
mlr3_policy <- function(learner, train_d, test_d) {
  d <- train_d[, c("subscribed", feat)]
  task <- as_task_regr(d, target = "subscribed")
  learner$train(task)
  pf <- function(dd) learner$predict_newdata(dd[, feat])$response
  list(pred_train = predict_arms(pf, train_d), pred_test = predict_arms(pf, test_d))
}
set.seed(1)
fit_reg <- {
  m <- lm(update(lasso_formula, subscribed ~ .), data = to_factors(train), weights = 1 / e)
  pf <- function(d) as.numeric(predict(m, newdata = to_factors(d)))
  list(pred_train = predict_arms(pf, train), pred_test = predict_arms(pf, test))
}
fit_cart <- {
  d <- to_factors(train)[, c("subscribed", "arm", covs)]
  tr <- rpart::rpart(subscribed ~ ., data = d, method = "anova", control = rpart::rpart.control(xval = 5, cp = 0, minbucket = 100))
  tr <- rpart::prune(tr, cp = tr$cptable[which.min(tr$cptable[, "xerror"]), "CP"])
  pf <- function(dd) as.numeric(predict(tr, newdata = to_factors(dd)[, c("arm", covs)]))
  list(pred_train = predict_arms(pf, train), pred_test = predict_arms(pf, test), tree = tr)
}
fit_rf <- mlr3_policy(lrn("regr.ranger", num.trees = n_trees, min.node.size = 70, num.threads = n_threads, seed = 1), train_d, test_d)
fit_xgb <- mlr3_policy(lrn("regr.xgboost", objective = "reg:logistic", eta = 0.4995, alpha = 0.7245, subsample = 0.8,
                           colsample_bytree = 0.9, max_depth = 6, nrounds = 200, nthread = n_threads), train_d, test_d)
```

## Treatment-effect models

**The paper also derives policies from pairwise treatment-effect
estimates, because the marketing literature increasingly uses causal
trees and forests for targeting.** For each pair of arms the training
users of those two arms form a binary experiment with a known propensity
(the ratio of the two shares), a causal forest estimates the pairwise
effect, and the policy chooses the arm that wins its comparisons. The
package builds the pairwise objects from the three-arm scores with
`contrast_scores()`; the DR-learner and a pruned regression tree on the
doubly robust score (the closest single-tree analogue of the paper's
causal tree available without the archived `causalTree` package) are fitted
on the same objects.

```{r cate-models}
pairs <- list(c(7, 30), c(14, 30))
pair_scores <- lapply(pairs, function(p) contrast_scores(sc_train, treat = p[1], control = p[2]))
pair_test <- lapply(pairs, function(p) contrast_scores(sc_test, treat = p[1], control = p[2]))
rf_final <- lrn("regr.ranger", num.trees = n_trees, min.node.size = 200, num.threads = n_threads, seed = 1)
dr_models <- lapply(pair_scores, function(s) cate_learner(scores = s, x_het = zcols, method = "dr", learner_final = rf_final))
cf_models <- lapply(seq_along(pairs), function(j) {
  s <- pair_scores[[j]]
  grf::causal_forest(as.matrix(s$data[, zcols]), s$y, s$d, W.hat = s$nuisance$p, Y.hat = s$nuisance$l,
                     num.trees = n_trees, honesty = TRUE, num.threads = n_threads, seed = 1)
})
ct_models <- lapply(pair_scores, function(s) {
  d <- to_factors(s$data)[, covs]; d$score <- s$score
  tr <- rpart::rpart(score ~ ., data = d, method = "anova", control = rpart::rpart.control(xval = 5, cp = 0, minbucket = 500))
  rpart::prune(tr, cp = tr$cptable[which.min(tr$cptable[, "xerror"]), "CP"])
})
tau_pairs <- function(models, d, pred) cbind(`7` = pred(models[[1]], d), `14` = pred(models[[2]], d), `30` = 0)
pred_dr <- function(m, d) as.numeric(predict(m, d))
pred_cf <- function(m, d) as.numeric(predict(m, as.matrix(d[, zcols]))$predictions)
pred_ct <- function(m, d) as.numeric(predict(m, newdata = to_factors(d)[, covs]))
fit_dr <- list(pred_train = tau_pairs(dr_models, train_d, pred_dr), pred_test = tau_pairs(dr_models, test_d, pred_dr))
fit_cf <- list(pred_train = tau_pairs(cf_models, train_d, pred_cf), pred_test = tau_pairs(cf_models, test_d, pred_cf))
fit_ct <- list(pred_train = tau_pairs(ct_models, train, pred_ct), pred_test = tau_pairs(ct_models, test, pred_ct))
```

## What each method assigns

**Table A8 of the paper reports the share of users each policy sends
to each arm, and it is the fastest way to see that the methods disagree
about who should get what.** CART and the single tree on the score
prescribe 7 days for everyone (the paper's footnote to Table 8), while
the forests spread users across the three arms the most; the lasso sends
about two thirds of users to 7 days and a quarter to 14 days, close to the
paper's 68.9 and 23.2 percent.

```{r policies}
lasso_pick <- 1L   # replaced by the median run in the next section
fits <- list(reg = fit_reg, cart = fit_cart, r_forest = fit_rf, xgboost = fit_xgb, dr_learner = fit_dr, c_forest = fit_cf, c_tree = fit_ct)
policies_from <- function(fits, lasso_run, which = "pred_test", n) {
  c(list(`7` = rep(7, n), `14` = rep(14, n), `30` = rep(30, n), lasso = argmax_policy(lasso_run[[which]])),
    lapply(fits, function(f) argmax_policy(f[[which]])))
}
```

```{r tab-shares, echo = FALSE}
share_tab <- function(pols) t(sapply(pols, function(p) 100 * prop.table(table(factor(p, levels = arms)))))
pol_names <- c(`7` = "$\\pi_7$", `14` = "$\\pi_{14}$", `30` = "$\\pi_{30}$", lasso = "$\\pi_{lasso}$", reg = "$\\pi_{reg}$",
               cart = "$\\pi_{cart}$", r_forest = "$\\pi_{r\\_forest}$", xgboost = "$\\pi_{xgboost}$",
               dr_learner = "$\\pi_{dr\\_learner}$", c_forest = "$\\pi_{c\\_forest}$", c_tree = "$\\pi_{c\\_tree}$")
sh_tr <- share_tab(policies_from(fits, lasso_runs[[lasso_pick]], "pred_train", nrow(train)))
sh_te <- share_tab(policies_from(fits, lasso_runs[[lasso_pick]], "pred_test", nrow(test)))
tibble(Policy = pol_names[rownames(sh_tr)], tr7 = sh_tr[, 1], tr14 = sh_tr[, 2], tr30 = sh_tr[, 3],
       te7 = sh_te[, 1], te14 = sh_te[, 2], te30 = sh_te[, 3]) %>%
  filter(!Policy %in% pol_names[1:3]) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, digits = 2,
      col.names = c("Policy", "7 days", "14 days", "30 days", "7 days", "14 days", "30 days"),
      caption = "Share of users (in percent) assigned to each trial length by each policy (the layout of the paper's Table A8). The lasso row uses the first cross-validation run here; the median run is selected in the next section.") %>%
  add_header_above(c(" " = 1, "Training set" = 3, "Test set" = 3)) %>% tab_style()
```

**Figure 5 of the paper plots the distribution of the estimated 7-versus-30
effect across test users for every method, because the amount of
heterogeneity a method finds predicts how well its policy does.** The
paper's reading is that CART, the causal tree, and the causal forest show
almost no heterogeneity, so their policies cannot personalize, while the
regression and random forest show the most and overfit. Here the true
effect is known and drawn as the reference curve, so each method can be
read as too flat or too spread relative to the truth; the table after the
figure gives the spread of each curve and its correlation with the truth.
Every learner except the lasso spreads its estimates two to five times
more than the truth, the two trees show none, and the correlation with
the truth ranks the methods broadly as the test-set values in Table
\@ref(tab:tab-table8) do.

```{r fig-cdf, echo = FALSE, fig.height = 3.8, fig.cap = "Cumulative distribution of the estimated effect of 7 versus 30 days across test users for each method (the layout of the paper's Figure 5), with the true effect for reference."}
tau_test <- function(f) f$pred_test[, "7"] - f$pred_test[, "30"]
cdf_df <- bind_rows(
  tibble(method = "Truth", tau = test$pr_7 - test$pr_30),
  tibble(method = "Lasso", tau = tau_test(lasso_runs[[lasso_pick]])),
  tibble(method = "Linear regression", tau = tau_test(fit_reg)),
  tibble(method = "CART", tau = tau_test(fit_cart)),
  tibble(method = "Random forest", tau = tau_test(fit_rf)),
  tibble(method = "XGBoost", tau = tau_test(fit_xgb)),
  tibble(method = "DR-learner", tau = tau_test(fit_dr)),
  tibble(method = "Causal forest", tau = tau_test(fit_cf)),
  tibble(method = "Tree on DR score", tau = tau_test(fit_ct))
) %>% mutate(method = factor(method, levels = unique(method)))
ggplot(cdf_df, aes(tau, colour = method, linetype = method == "Truth")) + stat_ecdf(geom = "step") +
  coord_cartesian(xlim = c(-0.08, 0.10)) + geom_vline(xintercept = 0, colour = "grey60", linetype = 3) +
  scale_linetype_manual(values = c(`TRUE` = 1, `FALSE` = 5), guide = "none") +
  labs(x = expression(hat(tau)[7*","*30]), y = "Cumulative share of test users", colour = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "right")
```

```{r tab-cdf, echo = FALSE}
truth_tau <- test$pr_7 - test$pr_30
cdf_df %>% group_by(method) %>%
  summarise(sd = sd(tau), share_positive = mean(tau > 0), cor_truth = cor(tau, truth_tau), .groups = "drop") %>%
  kbl(booktabs = TRUE, linesep = "", digits = 3,
      col.names = c("Method", "Standard deviation", "Share positive", "Correlation with the truth"),
      caption = "Spread of the estimated 7-versus-30 effect across test users by method, and its correlation with the true effect.") %>% tab_style()
```

# Policy evaluation {#sec-evaluation}

## The IPS reward and the package's off-policy value

**Every policy is priced with the same estimator on the same test set,
because the paper's argument is that estimators must be compared on the
value of the policies they produce, not on their fit.** The authors'
function (their `CalculatePolicyProfit`) keeps the users whose assigned
arm equals the arm the policy would give them, and weights each by the
inverse of the share of that arm among the users the policy sends there.
This is the IPS reward with the propensity estimated within each policy
group (their footnote 10). The package computes the same quantity when
the multi-arm score object is built with `type = "ipw"` and those
propensities; with the design shares instead, the estimates differ in the
fourth decimal.

```{r ips}
policy_profit <- function(d, policy, y = "subscribed") {
  keep <- d$arm == policy
  es <- table(factor(d$arm[keep], levels = arms)) / table(factor(policy, levels = arms))
  sum(d[[y]][keep] / as.numeric(es[as.character(d$arm[keep])])) / nrow(d)
}
ipw_scores <- function(d, y = "subscribed", policy = NULL) {
  p <- if (is.null(policy)) design_e(d) else {
    keep <- d$arm == policy
    es <- table(factor(d$arm[keep], levels = arms)) / table(factor(policy, levels = arms))
    matrix(rep(as.numeric(es), each = nrow(d)), nrow(d), dimnames = list(NULL, arms))
  }
  dr_scores(d, y, "arm", arms = arms, p_hat = p, type = "ipw", folds = 2)
}
truth_value <- function(d, policy) mean(as.matrix(d[, paste0("pr_", arms)])[cbind(seq_len(nrow(d)), match(policy, arms))])
pol_lasso_test <- argmax_policy(lasso_runs[[1]]$pred_test)
c(paper_function = policy_profit(test, pol_lasso_test),
  package_within_policy = policy_value(ipw_scores(test, policy = pol_lasso_test), pol_lasso_test)$estimate,
  package_design_shares = policy_value(ipw_scores(test), pol_lasso_test)$estimate,
  truth = truth_value(test, pol_lasso_test))
```

**The median lasso run is selected on the test-set IPS value, as in the
paper.**

```{r lasso-median}
lasso_test_values <- sapply(lasso_runs, function(r) policy_profit(test, argmax_policy(r$pred_test)))
lasso_pick <- order(lasso_test_values)[ceiling(length(lasso_test_values) / 2)]
round(lasso_test_values, 5); lasso_pick
```

## Table 8: the value of each policy

```{r table8}
pols_train <- policies_from(fits, lasso_runs[[lasso_pick]], "pred_train", nrow(train))
pols_test <- policies_from(fits, lasso_runs[[lasso_pick]], "pred_test", nrow(test))
ips_train <- sapply(pols_train, function(p) policy_profit(train, p))
ips_test <- sapply(pols_test, function(p) policy_profit(test, p))
dr_val <- policy_value(sc_test, pols_test, baseline = "30")
truth_val <- sapply(pols_test, function(p) truth_value(test, p))
t8 <- tibble(policy = names(pols_test), ips_train = ips_train, ips_test = ips_test,
             gain_train = 100 * (ips_train / ips_train[["30"]] - 1), gain_test = 100 * (ips_test / ips_test[["30"]] - 1),
             dr_test = dr_val$estimate + dr_val$baseline_value, dr_gain = dr_val$gain_pct,
             true_test = truth_val, true_gain = 100 * (truth_val / truth_val[["30"]] - 1))
```

**Table 8 of the paper has two blocks, training and test, and two
numbers per block, the estimated subscription rate and its percentage gain
over 30-for-all; the reader is meant to compare the two blocks row by
row.** The lasso policy beats every uniform policy on the test set, with a
test gain of `r fmt(t8$gain_test[["lasso"]], 2)` percent against the
paper's 6.81. The forest-based policies (random forest, DR-learner, causal
forest) and XGBoost show the largest training gains and the smallest test
gains, which is the overfitting warning; the paper's random forest shows
the same reversal. CART and the tree on the score collapse to 7-for-all,
so their rows repeat the 7-for-all row. The two column pairs added here,
the doubly robust value and the true value, are available only in a
simulation and confirm that the IPS test-set numbers rank the policies
correctly: the lasso is first on all three, and the largest gap between
a test-set IPS estimate and the truth is
`r fmt(100 * max(abs(t8$ips_test - t8$true_test)), 2)` percentage points.

```{r tab-table8, echo = FALSE}
t8_show <- t8 %>% mutate(policy = pol_names[policy],
                         across(c(ips_train, ips_test, dr_test, true_test), ~ fmt(100 * .x, 2)),
                         across(c(gain_train, gain_test, dr_gain, true_gain), ~ ifelse(abs(.x) < 1e-12, "---", fmt(.x, 2))))
t8_show <- t8_show[c("lasso", "7", "14", "30", "reg", "cart", "r_forest", "xgboost", "dr_learner", "c_forest", "c_tree") %>% match(t8$policy), ]
t8_show %>% mutate(category = c("Personalized based on lasso", rep("Uniform", 3), rep("Alternative personalized policies", 7))) %>%
  select(category, policy, ips_train, ips_test, gain_train, gain_test, dr_test, dr_gain, true_test, true_gain) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Policy category", "Policy", "Training set", "Test set", "Training set", "Test set", "Estimate", "Increase", "Estimate", "Increase"),
      caption = "Gains in subscription from implementing different counterfactual free-trial policies (the layout of the paper's Table 8). The first four numeric columns are the IPS estimates as in the paper; the doubly robust columns use the regression-adjusted scores; the true columns average each user's true subscription probability under the arm the policy assigns. Subscription rates and increases are in percent; increases are relative to the 30-day-for-all policy.") %>%
  add_header_above(c(" " = 2, "Estimated subscription (%)" = 2, "Increase in subscription (%)" = 2, "Doubly robust, test set" = 2, "True value, test set" = 2)) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major") %>% tab_style(c("scale_down"))
```

**The training-versus-test reversal is the paper's central caution and
it reappears here, in a stronger form.** The forests and XGBoost predict
their own training rows in sample, so the argmax rule follows the noise in
those rows and the training IPS value rewards it, because the same rows
carry the realized outcomes. The training gains of these methods are far
above anything a policy can deliver (the truth column caps the gain near
8 percent), and their test gains are the smallest. The paper's random
forest shows the same pattern at a smaller scale (18.67 percent in
training, 1.32 in test). The lasso's penalty removes most interaction
coefficients, so its training and test gains are close, and the true
value confirms its test gain.

```{r overfit, echo = FALSE}
t8 %>% transmute(policy = pol_names[policy], gain_train, gain_test, gap = gain_train - gain_test, true_gain) %>%
  filter(!policy %in% pol_names[c("7", "14", "30")]) %>% arrange(desc(gap)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, digits = 2,
      col.names = c("Policy", "Training gain (\\%)", "Test gain (\\%)", "Training minus test", "True gain (\\%)"),
      caption = "Overfitting of the policy: the difference between the training and test IPS gains, against the true gain.") %>% tab_style()
```

## Is the lasso policy better than 7-for-all?

**The paper answers with a bootstrap over test users, because the lasso
and the 7-for-all policies are evaluated on overlapping users and their
difference needs a paired standard error.** The same paired difference is
available analytically: the multi-arm `policy_value()` reports each
policy's difference from the first policy in the list with the standard
error of the per-user differences. The bootstrap reproduces the paper's t
statistic (3.12 in the paper). The analytic standard errors are larger,
for two reasons: the bootstrap re-estimates the within-policy
propensities in every draw, which turns the IPS reward into a ratio
estimator with a smaller variance than the per-user differences with
fixed propensities, and ten bootstrap rounds estimate a standard error
imprecisely. The true difference, known in the simulation, is positive
and close to both point estimates.

```{r boot-test}
set.seed(7)
boot_diff <- replicate(params$n_boot_rounds, {
  i <- sample.int(nrow(test), replace = TRUE)
  policy_profit(test[i, ], pols_test$lasso[i]) - policy_profit(test[i, ], pols_test[["7"]][i])
})
boot_t <- t.test(boot_diff)
paired <- policy_value(ipw_scores(test), list(`7` = pols_test[["7"]], lasso = pols_test$lasso, `14` = pols_test[["14"]]))
tibble(method = c(paste0("Bootstrap over test users (", params$n_boot_rounds, " rounds)"), "Paired per-user difference, IPS scores", "Paired per-user difference, doubly robust scores"),
       difference = c(mean(boot_diff), paired$diff_vs_first[2], policy_value(sc_test, list(`7` = pols_test[["7"]], lasso = pols_test$lasso))$diff_vs_first[2]),
       std_error = c(sd(boot_diff), paired$diff_se[2], policy_value(sc_test, list(`7` = pols_test[["7"]], lasso = pols_test$lasso))$diff_se[2])) %>%
  mutate(t = difference / std_error) %>%
  bind_rows(tibble(method = "True difference", difference = truth_val[["lasso"]] - truth_val[["7"]]))
```

# Long-run outcomes {#sec-longrun}

## Average effects on subscription length and revenue

**Tables 9 and 10 of the paper repeat the Table 5 layout for subscription
length and revenue, because a policy that raises subscriptions could
still lower revenue if the marginal subscribers are short-lived or cheap.**
The paper decomposes each long-run outcome into the subscription decision
and the outcome conditional on subscribing (equations 5 and 6), and finds
the conditional part almost flat across arms; the gains in retention and
revenue are therefore driven by conversions.

```{r tab-table9, echo = FALSE}
t9 <- bind_rows(Training = ate_panel(train, "months"), Test = ate_panel(test, "months"), .id = "Data")
t10 <- bind_rows(Training = ate_panel(train, "revenue"), Test = ate_panel(test, "revenue"), .id = "Data")
show_ate(t9, "Average effect of trial length on subscription length in months (the layout of the paper's Table 9).", digits = 3,
         col1 = "Subscription length", col2 = "Difference")
```

```{r tab-table10, echo = FALSE}
show_ate(t10, "Average effect of trial length on revenue in dollars (the layout of the paper's Table 10).", digits = 2,
         col1 = "Revenue", col2 = "Difference")
```

```{r tab-conditional, echo = FALSE}
cond <- pop %>% filter(subscribed == 1) %>% group_by(arm) %>% summarise(months = mean(months), revenue = mean(revenue), price = mean(revenue / months), .groups = "drop")
kbl(cond, booktabs = TRUE, linesep = "", digits = 2, col.names = c("Arm", "Months, subscribers", "Revenue, subscribers", "Revenue per month"),
    caption = "The conditional part of the decomposition in the paper's equations 5 and 6: subscription length and revenue among subscribers, by arm.") %>% tab_style()
```

## The subscription policy evaluated on long-run outcomes

**Table 11 of the paper evaluates the same four policies on subscription
length and revenue with the same IPS estimator, changing only the
outcome, to show that the policy designed on subscriptions also raises
the long-run outcomes.** The estimator does not need to be refitted: the
policy is fixed and the reward function accepts any outcome. In the
simulation the lasso policy and the 7-day uniform policy are within
noise of each other on the long-run outcomes, because months and revenue
are generated almost independently of the arm once a user subscribes and
their variance is far larger than the difference between the two
policies.

```{r tab-table11, echo = FALSE}
lr_pols_train <- pols_train[c("lasso", "7", "14", "30")]
lr_pols_test <- pols_test[c("lasso", "7", "14", "30")]
t11 <- bind_rows(lapply(c(months = "months", revenue = "revenue"), function(y) {
  tr <- sapply(lr_pols_train, function(p) policy_profit(train, p, y))
  te <- sapply(lr_pols_test, function(p) policy_profit(test, p, y))
  tibble(policy = names(tr), est_train = tr, est_test = te,
         inc_train = 100 * (tr / tr[["30"]] - 1), inc_test = 100 * (te / te[["30"]] - 1))
}), .id = "outcome")
t11 %>% pivot_wider(names_from = outcome, values_from = c(est_train, est_test, inc_train, inc_test)) %>%
  mutate(category = c("Personalized", rep("Uniform", 3)), policy = pol_names[policy],
         across(starts_with("inc"), ~ ifelse(abs(.x) < 1e-12, "---", fmt(.x, 2))),
         across(starts_with("est"), ~ fmt(.x, 2))) %>%
  select(category, policy, est_train_months, est_test_months, inc_train_months, inc_test_months,
         est_train_revenue, est_test_revenue, inc_train_revenue, inc_test_revenue) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Category", "Policy", "Training", "Test", "Training", "Test", "Training", "Test", "Training", "Test"),
      caption = "IPS estimates of the average subscription length and revenue under three uniform and one personalized policy (the layout of the paper's Table 11).") %>%
  add_header_above(c(" " = 2, "Estimate (months)" = 2, "Increase (%)" = 2, "Estimate ($)" = 2, "Increase (%)" = 2)) %>%
  add_header_above(c(" " = 2, "Subscription length" = 4, "Revenue" = 4)) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major") %>% tab_style(c("scale_down"))
```

## Policies optimized on long-run outcomes

**Table 12 of the paper trains the same lasso on revenue and on
subscription length and evaluates all three policies on all three
outcomes, to answer whether a firm should wait two years to design its
policy.** The paper's answer is no: the subscription policy does at least
as well on every outcome, because the conditional outcomes hardly differ
across arms, the long-run outcomes have far higher variance, and a binary
outcome is easier to learn. The rows are the training outcome and the
columns the evaluation outcome, so the diagonal is where each policy
should shine if long-run training paid off. Here the three policies are
within a few tenths of a percent of each other on every outcome; the
revenue-trained policy is the worst on subscriptions and revenue, which is
the paper's variance argument, and no policy gains from being trained on
the outcome it is evaluated on.

```{r lasso-longrun, cache = TRUE, cache.extra = list(sim_seed, n_users)}
lasso_months <- fit_lasso_policy(train, test, y = "months", reps = 1, seed = 200)[[1]]
lasso_revenue <- fit_lasso_policy(train, test, y = "revenue", reps = 1, seed = 300)[[1]]
```

```{r tab-table12, echo = FALSE}
opt_pols <- list(Subscription = lasso_runs[[lasso_pick]], `Total revenue` = lasso_revenue, `Subscription length` = lasso_months)
t12 <- bind_rows(lapply(c(`Training data` = "train", `Test data` = "test"), function(s) {
  d <- if (s == "train") train else test
  bind_rows(lapply(names(opt_pols), function(nm) {
    p <- argmax_policy(opt_pols[[nm]][[paste0("pred_", s)]])
    tibble(optimized_on = nm, subscription = 100 * policy_profit(d, p), revenue = policy_profit(d, p, "revenue"),
           months = policy_profit(d, p, "months"))
  }))
}), .id = "data")
t12 %>% mutate(subscription = fmt(subscription, 2), revenue = fmt(revenue, 2), months = fmt(months, 2)) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Data set", "Policy optimized on", "Subscription (%)", "Total revenue ($)", "Subscription length (months)"),
      caption = "Expected mean of the three outcomes under policies optimizing each outcome (the layout of the paper's Table 12).") %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major") %>% tab_style()
```

# The shipped data, once more {#sec-shipped}

**Running the paper's own procedure on the shipped file shows what the
randomized data return: a subscription rate of one half under every
policy and gains within a fraction of a percent of zero.** The lasso is
fitted on the shipped training rows with the authors' specification, and
each policy is priced with the authors' IPS function. Because every column
is independent noise, all policies have the same true value; the
penalized fit keeps no interaction and the lasso policy collapses to a
uniform policy.

```{r shipped-run, cache = TRUE}
sh <- shipped %>% transmute(user_id, arm = arms[trial_length + 1], region = country, os, channel = signup_app, job, skill, segment,
                            subscribed, months = ifelse(is.na(total_month), 0, total_month), revenue = total_revenue, training)
sh_train <- sh[sh$training == 1, ]; sh_test <- sh[sh$training == 0, ]
sh_train$e <- as.numeric(arm_share(sh_train)[as.character(sh_train$arm)])
sh_test$e <- as.numeric(arm_share(sh_test)[as.character(sh_test$arm)])
sh_to_factors <- function(d) { for (v in covs) d[[v]] <- factor(d[[v]], levels = sort(unique(sh[[v]]))); d$arm <- factor(d$arm, levels = arms); d }
sh_X <- function(d) Matrix::sparse.model.matrix(lasso_formula, sh_to_factors(d))
set.seed(11)
sh_cv <- cv.glmnet(sh_X(sh_train), sh_train$subscribed, weights = 1 / sh_train$e, alpha = 1, nfolds = 5)
sh_fit <- glmnet(sh_X(sh_train), sh_train$subscribed, weights = 1 / sh_train$e, alpha = 1, lambda = sh_cv$lambda.min)
sh_pf <- function(d) as.numeric(predict(sh_fit, sh_X(d)))
sh_pols <- list(train = argmax_policy(predict_arms(sh_pf, sh_train)), test = argmax_policy(predict_arms(sh_pf, sh_test)))
shipped_tab <- bind_rows(lapply(c(Training = "train", Test = "test"), function(s) {
  d <- if (s == "train") sh_train else sh_test
  pols <- list(lasso = sh_pols[[s]], `7` = rep(7, nrow(d)), `14` = rep(14, nrow(d)), `30` = rep(30, nrow(d)))
  v <- sapply(pols, function(p) policy_profit(d, p))
  tibble(policy = names(v), value = 100 * v, gain = 100 * (v / v[["30"]] - 1))
}), .id = "data")
```

```{r tab-shipped-run, echo = FALSE}
shipped_tab %>% pivot_wider(names_from = data, values_from = c(value, gain)) %>%
  mutate(policy = pol_names[policy], across(starts_with("gain"), ~ ifelse(abs(.x) < 1e-12, "---", fmt(.x, 2))), across(starts_with("value"), ~ fmt(.x, 2))) %>%
  select(policy, value_Training, value_Test, gain_Training, gain_Test) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Policy", "Training set", "Test set", "Training set", "Test set"),
      caption = "The paper's Table 8 procedure applied to the shipped, randomized data: subscription rates and gains in percent.") %>%
  add_header_above(c(" " = 1, "Estimated subscription (%)" = 2, "Increase in subscription (%)" = 2)) %>% tab_style()
```

# Package tools beyond the paper {#sec-tools}

## Validating a CATE model before building a policy on it

**The paper judges each effect estimator only through the value of its
policy; the package can also test the estimator directly on the test-set
scores.** `cate_validate()` takes a model of the 7-versus-30 effect and
the test-set pair object and reports the calibration slope, the effect by
quartile of the prediction, and the rank-weighted average effect (AUTOC)
with a bootstrap standard error. A slope near one with a positive AUTOC
says the model ranks users correctly, which is what a targeting policy
needs from it. The DR-learner fails this test: its slope is near zero, the
quartile with the highest predictions has no larger effect than the
others, and the AUTOC is zero within its standard error. This is the
same verdict Table \@ref(tab:tab-table8) reached through the policy
value, obtained before any policy is built. The lasso's own implied
effect, the difference between its two arm predictions, passes: the
slope is not distinguishable from one, the group effects rise with the
prediction (the predictions take few distinct values, so only three
groups form), and the QINI area is positive at the 1 percent level; the
AUTOC, which weights the top ranks most, is positive at the 10 percent
level.

```{r validate}
val <- cate_validate(pair_test[[1]], tau_hat = dr_models[[1]], n_groups = 4, n_boot = 199, seed = 1)
val
rows_7_30 <- test$arm %in% c(7, 30)
tau_lasso <- lasso_runs[[lasso_pick]]$pred_test[rows_7_30, "7"] - lasso_runs[[lasso_pick]]$pred_test[rows_7_30, "30"]
val_lasso <- cate_validate(pair_test[[1]], tau_hat = tau_lasso, n_groups = 4, n_boot = 199, seed = 1)
val_lasso
```

```{r fig-validate, echo = FALSE, fig.height = 3.2, fig.cap = "Targeting operator characteristic on the test set for 7 versus 30 days: the DR-learner (left) and the lasso's implied effect (right)."}
gridExtra::grid.arrange(plot_cate_validation(val, what = "toc") + ggtitle("DR-learner"),
                        plot_cate_validation(val_lasso, what = "toc") + ggtitle("Lasso"), ncol = 2)
```

## A depth-two policy tree on three arms

**The paper's policies are argmax rules over black-box predictions; an
interpretable alternative is a shallow tree whose leaves are arms, fitted
directly on the doubly robust scores.** `policy_learn()` with
`engine = "policytree"` runs the exact search of Zhou, Athey, and Wager
(2023) on a 60,000-user subsample of the three-arm training scores, over
the skill, job, and region dummies; the tree is then priced on the test
set with `policy_value()` and against the truth. A tree with four leaves
reaches the same true value as the lasso policy, because most of the
simulated heterogeneity is carried by skill: the tree gives 7 days to
users who are neither experienced nor of mixed skill, 14 days to
mixed-skill users, and 30 days to experienced users, with one job-level
exception in the experienced branch. It says in three lines who gets
what.

```{r policytree, cache = TRUE, cache.extra = list(sim_seed, n_users)}
tree_x <- zcols[grep("skill|job|region", colnames(model.matrix(~ factor(region) + factor(os) + factor(channel) + factor(job) + factor(skill) + factor(segment), train))[-1])]
set.seed(3)
sub <- sample.int(nrow(train_d), 60000)
sc_sub <- dr_scores(train_d[sub, ], "subscribed", "arm", zcols, arms = arms, p_hat = design_e(train_d[sub, ]),
                    learner_mu = lrn("regr.lm"), folds = 2, seed = 1)
pol_tree <- policy_learn(sc_sub, x = tree_x, method = "tree", depth = 2, engine = "policytree", holdout = 0, baseline = "30")
pol_tree
z_names <- setNames(colnames(model.matrix(~ factor(region) + factor(os) + factor(channel) + factor(job) + factor(skill) + factor(segment), train))[-1], zcols)
split_vars <- unique(sapply(Filter(function(nd) !nd$is_leaf, pol_tree$rule$fit$nodes), `[[`, "split_variable"))
z_names[tree_x[split_vars]]
tree_test <- predict(pol_tree, test_d)
policy_value(sc_test, list(tree = tree_test, lasso = pols_test$lasso), baseline = "30")[, c("policy", "estimate", "std.error", "gain_pct", "share_7", "share_14", "share_30")]
c(truth_tree = truth_value(test, tree_test), truth_lasso = truth_value(test, pols_test$lasso))
```

# Lessons {#sec-lessons}

**What was replicated.** The paper's procedure, from the descriptive
tables through the policy design, the IPS evaluation, and the long-run
projections, runs end to end on a population calibrated to the published
moments, with the same table and figure layouts. The IPS reward the
authors coded is matched exactly by the package's off-policy value once
the within-policy propensities are supplied, and the paper's ranking of
methods, its overfitting warning, and its finding that the subscription
policy also wins on retention and revenue all reappear with the true
values as a check.

**What could not be replicated.** The published data are randomized,
so no number of the paper can be recomputed from them. The causal tree
was replaced by a pruned regression tree on the doubly robust score, and
the random forest, boosting, and forest settings were reduced from the
authors' tuned values to keep the note within a reasonable running time.

**What the package adds.** The multi-arm score object makes the three
arms one object, so that the IPS and doubly robust rewards, the pairwise
contrasts, the CATE validation, and the policy tree share the same
nuisances and the same test set. The doubly robust reward gives the same
ranking as IPS, the paired standard error from `policy_value()` replaces
the paper's bootstrap, `cate_validate()` rejects the DR-learner on the
test scores before any policy is built on it, and a depth-two policy tree
matches the lasso's true value with an interpretable rule.

# References

- Yoganarasimhan, H., Barzegary, E., and Pani, A. (2023). Design and
  evaluation of optimal free trials. *Management Science*, 69(6), 3220–3240.
- Zhou, Z., Athey, S., and Wager, S. (2023). Offline multi-action policy
  learning: generalization and optimization. *Operations Research*, 71(1),
  148–183.
- Kennedy, E. H. (2023). Towards optimal doubly robust estimation of
  heterogeneous causal effects. *Electronic Journal of Statistics*, 17(2),
  3008–3049.
- Athey, S. and Wager, S. (2021). Policy learning with observational
  data. *Econometrica*, 89(1), 133–161.
