---
title: "Replication: Carbon Taxes and CO2 Emissions"
subtitle: "Synthetic control, illustrated with Andersson (American Economic Journal: Economic Policy, 2019)"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/09_Synthetic_Control/andersson2019carbon/"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "90%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(causalmetrics)
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(haven)
library(sandwich)
library(lmtest)
library(ivreg)
data_path <- params$data_path
data_md5 <- tools::md5sum(file.path(data_path, c("carbontax_data.dta", "regression_data.Rds", "disentangling_data.dta", "descr_Sweden.Rds")))
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
pct <- function(x, digits = 1) paste0(formatC(100 * x, digits = digits, format = "f"), " percent")
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
rep_theme <- function() theme_minimal(base_size = 11) + theme(legend.position = "bottom", panel.grid.minor = element_blank())
tax_line <- geom_vline(xintercept = 1989.5, linetype = "dotted")
```

# The paper in one page {#sec-paper}

**The question is whether a carbon tax reduces emissions, answered with
one country and thirty years of history.** Sweden extended its value-added
tax to transport fuel in 1990 and introduced a carbon tax on it in 1991,
on top of an existing energy tax. Andersson (2019) asks what happened to
per-capita CO2 emissions from transport, the sector the tax covers. The
answer has to come from a comparison, and the natural comparison, the
average of other OECD countries, does not trend like Sweden before 1990.

**The design is a synthetic control with a donor pool chosen by hand and
weights chosen by the data.** From 25 OECD countries the paper keeps 14
donors: it drops the countries that introduced carbon taxes on transport
(Finland, Norway, the Netherlands) or made large fuel-tax changes
(Germany, Italy, the United Kingdom), two countries with fuel tourism
(Austria, Luxembourg), and two whose development is unlike Sweden's
(Turkey, Ireland). Synthetic Sweden is the convex combination of the 14
donors that best reproduces Sweden's 1980 to 1989 averages of GDP per
capita, motor vehicles, gasoline consumption, and urban population, and
its emissions in 1970, 1980, and 1989. The weights on the predictors are
chosen jointly with the country weights to minimize the pre-treatment
mean squared prediction error of emissions over 1960 to 1989.

**The headline is a reduction of almost 11 percent in an average
post-treatment year.** Emissions in Sweden are 0.29 metric tons per capita
below synthetic Sweden on average over 1990 to 2005, and 0.35 tons, or
12.5 percent, below it in 2005. The paper then splits the effect between
the VAT and the carbon tax with a time-series regression of gasoline
consumption on the tax-exclusive price and the carbon tax, finds that
consumers respond about three times more strongly to the carbon tax than
to an equivalent price change, and attributes 6.3 percent of the 10.9
percent to the carbon tax alone.

**This notebook reproduces every table and figure from the replication
data, adds the checks of Ferman and Pinto (2021), and reads each object
as the paper does.** Section \@ref(sec-data) describes the data. Section
\@ref(sec-descriptive) reproduces the descriptive figures and the
pass-through regression. Section \@ref(sec-synthetic) builds synthetic
Sweden (Tables 1 and 2, Figures 4 and 5). Section \@ref(sec-placebo) runs
the placebo tests (Figures 6 to 9). Section \@ref(sec-confounder) checks
the recession (Figures 11 and 12). Section \@ref(sec-disentangle)
reproduces Table 3 and Figures 13 and 14. Section \@ref(sec-fp) adds the
imperfect-fit diagnostics. All estimation uses `synth_control()` and its
companions from `causalmetrics`; the regressions use `lm()`, `ivreg()`,
and Newey-West standard errors from `sandwich`.

# Data {#sec-data}

**The panel is 15 countries over 1960 to 2005 with the outcome and five
predictors.** The file `carbontax_data.dta` comes from the replication
package as redistributed in the RTutor problem set of the paper. The
outcome is CO2 emissions from transport in metric tons per capita (World
Bank). The predictors are GDP per capita (PPP, 2005 dollars), motor
vehicles per 1,000 people, gasoline consumption per capita (kilograms of
oil equivalent), urban population (percent), and population density.
Poland's GDP is missing before 1970, which matters only for the 1970
in-time placebo, where the paper's code drops Poland.

```{r data}
carbon <- as.data.frame(read_dta(file.path(data_path, "carbontax_data.dta")))
carbon$treated <- as.integer(carbon$country == "Sweden" & carbon$year >= 1990)
donors <- sort(setdiff(unique(carbon$country), "Sweden"))
xs <- c("GDP_per_capita", "vehicles_capita", "gas_cons_capita", "urban_pop")
x_labels <- c(GDP_per_capita = "GDP per capita", vehicles_capita = "Motor vehicles (per 1,000 people)",
              gas_cons_capita = "Gasoline consumption per capita", urban_pop = "Urban population",
              `CO2_transport_capita(1989)` = "CO2 from transport per capita 1989",
              `CO2_transport_capita(1980)` = "CO2 from transport per capita 1980",
              `CO2_transport_capita(1970)` = "CO2 from transport per capita 1970")
c(countries = length(unique(carbon$country)), years = length(unique(carbon$year)),
  pre_years = sum(unique(carbon$year) < 1990), post_years = sum(unique(carbon$year) >= 1990))
donors
```

**Three further files serve the descriptive figures and the second
analysis.** `descr_Sweden.Rds` holds the Swedish price and tax components
and the fuel consumption series (Figures 1 and 2 of the paper).
`regression_data.Rds` holds the annual Swedish series for 1970 to 2015
used in the pass-through regression and in Table 3: nominal and real
prices, the energy and carbon taxes with and without VAT, the crude oil
price, log gasoline consumption per capita, GDP per capita, urbanization,
and unemployment. `disentangling_data.dta` holds the simulated emission
paths under the three tax scenarios of Figure 13 and the synthetic
control gap for Figure 14.

```{r data-other}
descr <- readRDS(file.path(data_path, "descr_Sweden.Rds"))
reg <- readRDS(file.path(data_path, "regression_data.Rds"))
disent <- as.data.frame(read_dta(file.path(data_path, "disentangling_data.dta")))
c(descr = nrow(descr), regression = nrow(reg), regression_complete = sum(!is.na(reg$log_gas_cons)), disentangling = nrow(disent))
```

# Descriptive evidence and tax incidence {#sec-descriptive}

**Figure 1 of the paper shows that the tax reform changed the price of
fuel, which is the mechanism the whole design relies on.** The energy tax
existed before 1990; the VAT and the carbon tax were added in 1990 and
1991, and the carbon tax rose sharply after 2000 while the energy tax was
reduced. The figure is there so the reader sees a large, dated change in
the treatment variable.

```{r fig-taxes, fig.height = 3.4, fig.cap = "Real energy tax, carbon tax, and total tax on gasoline in Sweden, SEK per liter (paper's Figure 1, panel A)."}
descr %>% select(year, en_tax, CO2_tax, total_tax) %>%
  pivot_longer(-year, names_to = "component", values_to = "value") %>%
  mutate(component = recode(component, en_tax = "Energy tax", CO2_tax = "Carbon tax", total_tax = "Total tax")) %>%
  ggplot(aes(x = year, y = value, colour = component, linetype = component)) +
  geom_line() + tax_line +
  labs(x = NULL, y = "Real tax, SEK per liter", colour = NULL, linetype = NULL) + rep_theme()
```

**Figure 3 of the paper is the reason for the method: Sweden and the
donor average do not share a trend.** Before 1980 the two series grow at
a similar pace; between 1980 and 1989 Sweden's emissions grow twice as
fast. A difference-in-differences estimate would attribute that trend
difference to the tax. The figure is placed before any estimation so that
the reader accepts the need for a weighted comparison.

```{r fig3, fig.height = 3.4, fig.cap = "Per-capita CO2 emissions from transport, Sweden versus the unweighted average of the 14 donor countries (paper's Figure 3)."}
donor_avg <- carbon %>% filter(country != "Sweden") %>% group_by(year) %>%
  summarise(y = mean(CO2_transport_capita), .groups = "drop") %>% mutate(series = "OECD sample")
sweden <- carbon %>% filter(country == "Sweden") %>% transmute(year, y = CO2_transport_capita, series = "Sweden")
plot_paths <- function(d, ylab = "Metric tons per capita (CO2 from transport)") {
  ggplot(d, aes(x = year, y = y, colour = series, linetype = series)) +
    geom_line() + tax_line +
    scale_colour_manual(values = c("black", "grey45", "#B22222")[seq_along(unique(d$series))]) +
    scale_y_continuous(limits = c(0, 3)) +
    labs(x = NULL, y = ylab, colour = NULL, linetype = NULL) + rep_theme()
}
plot_paths(bind_rows(sweden, donor_avg))
```

**The pass-through regression (paper's equation 2) establishes that tax
changes reach consumers, so that tax elasticities can be read as demand
elasticities.** The nominal retail price is regressed in first
differences on the crude oil price and the sum of the energy and carbon
taxes, 1970 to 2015. The paper reports a coefficient of 1.15 on taxes
with a 95 percent interval of 0.90 to 1.39.

```{r passthrough}
pass <- reg %>% arrange(year) %>%
  mutate(delta_p = p_nom - lag(p_nom), delta_oil = oil_p - lag(oil_p), delta_tax = en_CO2_tax - lag(en_CO2_tax))
pass_fit <- lm(delta_p ~ delta_oil + delta_tax, data = pass)
cbind(estimate = coef(pass_fit), confint(pass_fit))[c("delta_oil", "delta_tax"), ]
```

The point estimate on taxes is the paper's 1.15. The interval above uses
ordinary least-squares standard errors and is wider than the paper's; the
paper does not state which standard errors it used for this regression,
and the replication package's code for it is not part of the RTutor data.

# Synthetic Sweden {#sec-synthetic}

**The synthetic control is built exactly as in the paper: four
covariates averaged over 1980 to 1989, three emission lags, and predictor
weights chosen to minimize the pre-treatment MSPE over 1960 to 1989.**
The `Synth` package solves the same nested problem with a different
optimizer; the pre-treatment MSPE is flat near its minimum, so the
predictor weights $V$ differ across optimizers while the country weights
agree to about 0.04. The country weights are what the paper interprets.

```{r synth-main, cache = TRUE, cache.extra = data_md5}
set.seed(1)
sc_sweden <- synth_control(carbon, id = "country", time = "year", y = "CO2_transport_capita", d = "treated",
                           x = xs, lags = c(1989, 1980, 1970), pre_window = 1980:1989, v = "mspe",
                           v_control = list(maxit = 2000, starts = 5))
sc_sweden
```

**Table 1 reports the predictor means, and its purpose is to show that
synthetic Sweden resembles Sweden on the drivers of emissions while the
donor average does not.** The paper's last column is a
population-weighted average of the donors; the replication data do not
include population, so the column here is unweighted, as in the RTutor
version. GDP per capita, vehicles, urban population, and the three
emission lags match closely. Gasoline consumption does not, and the paper
attributes that to the small weight $V$ assigns to it.

```{r tab1}
bal <- sc_sweden$balance
tab1 <- data.frame(Variables = unname(x_labels[bal$predictor]), Sweden = bal$treated,
                   `Synth. Sweden` = bal$synthetic, `OECD sample` = bal$donor_mean, `V matrix` = bal$v, check.names = FALSE)
kable(tab1, digits = c(0, 3, 3, 3, 3), booktabs = TRUE, row.names = FALSE,
      caption = "CO2 emissions from transport, predictor means before the tax reform (paper's Table 1). The OECD column is an unweighted donor average.") %>% tab_style()
```

**Table 2 reports the country weights, and the paper reads them as a
plausibility check: Denmark, Belgium, and New Zealand, which resemble
Sweden in institutions, GDP, and geography, make up three quarters of
synthetic Sweden.** The paper's weights are Denmark 0.384, Belgium 0.195,
New Zealand 0.177, Greece 0.090, the United States 0.088, and Switzerland
0.061, with the rest at or near zero.

```{r tab2}
w <- sc_sweden$weights
published <- c(Australia = 0.001, Belgium = 0.195, Canada = 0, Denmark = 0.384, France = 0, Greece = 0.090,
               Iceland = 0.001, Japan = 0, `New Zealand` = 0.177, Poland = 0.001, Portugal = 0, Spain = 0,
               Switzerland = 0.061, `United States` = 0.088)
tab2 <- data.frame(Country = w$country, Weight = round(w$weight, 3), Published = published[w$country])
kable(cbind(tab2[1:7, ], tab2[8:14, ]), booktabs = TRUE, row.names = FALSE,
      caption = "Country weights in synthetic Sweden (paper's Table 2), replication against published values.") %>% tab_style()
c(max_abs_difference = max(abs(tab2$Weight - tab2$Published)), pre_rmspe = sc_sweden$pre_rmspe,
  concentration = unname(sc_sweden$concentration[["l2"]]))
```

**Figure 4 is the identification argument in one picture: synthetic
Sweden tracks Sweden for thirty years and departs from it after 1990.**
The paper reports an average absolute pre-treatment difference of about
0.03 tons. The post-treatment distance is the estimate.

```{r fig4, fig.height = 3.4, fig.cap = "Per-capita CO2 emissions from transport, Sweden versus synthetic Sweden (paper's Figure 4)."}
paths <- sc_sweden$effects %>% transmute(year = time, Sweden = treated, `Synthetic Sweden` = synthetic) %>%
  pivot_longer(-year, names_to = "series", values_to = "y")
plot_paths(paths)
```

**Figure 5 plots the gap, and the paper reads three numbers off it: the
2005 reduction, the average reduction, and the reduction in 2000, the
last year before the energy tax was lowered.** The published values are
12.5 percent (0.35 tons) in 2005, 10.9 percent (0.29 tons) on average
over 1990 to 2005, and 11.5 percent (0.31 tons) in 2000. Percentages are
gaps relative to the synthetic level.

```{r fig5, fig.height = 3.4, fig.cap = "Gap in per-capita CO2 emissions from transport between Sweden and synthetic Sweden (paper's Figure 5)."}
eff <- sc_sweden$effects
ggplot(eff, aes(x = time, y = gap)) + geom_line() + geom_hline(yintercept = 0, linetype = "dotted") + tax_line +
  scale_y_continuous(limits = c(-0.4, 0.4)) +
  labs(x = NULL, y = "Gap in metric tons per capita") + rep_theme()
headline <- eff %>% filter(post) %>%
  summarise(average_gap = mean(gap), average_pct = mean(gap / synthetic),
            gap_2005 = gap[time == 2005], pct_2005 = gap[time == 2005] / synthetic[time == 2005],
            gap_2000 = gap[time == 2000], pct_2000 = gap[time == 2000] / synthetic[time == 2000],
            pre_abs_diff = mean(abs(eff$gap[!eff$post])))
round(as.data.frame(headline), 3)
```

The replication gives an average reduction of
`r fmt(headline$average_gap, 2)` tons (`r pct(-headline$average_pct)`),
a 2005 reduction of `r fmt(headline$gap_2005, 2)` tons
(`r pct(-headline$pct_2005)`), and a 2000 reduction of
`r fmt(headline$gap_2000, 2)` tons (`r pct(-headline$pct_2000)`),
against the paper's 0.29 (10.9 percent), 0.35 (12.5 percent), and 0.31
(11.5 percent).

# Placebo tests {#sec-placebo}

**The in-time placebos (Figure 6) move the tax to 1980 and to 1970 and
ask the method to find an effect where there was none.** Each placebo
uses only the data before 1990, shorter predictor windows, and emission
lags shifted back, following the paper's code: for 1980, covariates over
1970 to 1979 and lags in 1979, 1970, and 1965; for 1970, covariates over
1960 to 1969, lags in 1969, 1965, and 1960, and Poland dropped because
its GDP is missing before 1970.

```{r fig6, cache = TRUE, cache.extra = data_md5, fig.height = 3.2, fig.cap = "Placebo in-time tests: the tax moved to 1980 (left) and to 1970 (right), estimated on data before 1990 (paper's Figure 6)."}
set.seed(1)
pt_1980 <- synth_placebo(sc_sweden, type = "time", placebo_time = 1980, lags = c(1979, 1970, 1965), pre_window = 1970:1979)
sc_no_poland <- synth_control(carbon[carbon$country != "Poland", ], id = "country", time = "year",
                              y = "CO2_transport_capita", d = "treated", x = xs, lags = c(1989, 1980, 1970),
                              pre_window = 1980:1989, v = "mspe", v_control = list(maxit = 2000, starts = 5))
pt_1970 <- synth_placebo(sc_no_poland, type = "time", placebo_time = 1970, lags = c(1969, 1965, 1960), pre_window = 1960:1969)
placebo_paths <- bind_rows(
  pt_1980$effects %>% transmute(panel = "Panel A. 1980", year = time, Sweden = treated, `Synthetic Sweden` = synthetic),
  pt_1970$effects %>% transmute(panel = "Panel B. 1970", year = time, Sweden = treated, `Synthetic Sweden` = synthetic)
) %>% pivot_longer(c(Sweden, `Synthetic Sweden`), names_to = "series", values_to = "y")
ggplot(placebo_paths, aes(x = year, y = y, colour = series, linetype = series)) +
  geom_line() + facet_wrap(~ panel) +
  geom_vline(data = data.frame(panel = c("Panel A. 1980", "Panel B. 1970"), x = c(1979.5, 1969.5)),
             aes(xintercept = x), linetype = "dotted") +
  scale_colour_manual(values = c("black", "grey45")) + scale_y_continuous(limits = c(0, 3)) +
  labs(x = NULL, y = "Metric tons per capita", colour = NULL, linetype = NULL) + rep_theme()
c(placebo_1980_gap = pt_1980$estimate, placebo_1970_gap = pt_1970$estimate, real_gap = sc_sweden$estimate)
```

**The in-space placebos (Figure 7) reassign the tax to each donor; the
paper first shows all 14 gaps and then drops the donors whose
pre-treatment MSPE exceeds 20 times Sweden's, which leaves nine.** In the
restricted set Sweden's post-1990 gap is the largest, so the
permutation p-value is 1/10. Each placebo is fitted with the same
predictors and with the other donors as its pool.

```{r fig7, cache = TRUE, cache.extra = data_md5, fig.height = 3.4, fig.cap = "Permutation test: Sweden's gap (black) against the placebo gaps of all 14 donors (left) and of the nine donors with pre-treatment MSPE within 20 times Sweden's (right) (paper's Figure 7)."}
pl <- synth_placebo(sc_sweden, mspe_limit = 20)
gap_long <- function(keep, panel) {
  g <- pl$gaps[, keep, drop = FALSE]
  data.frame(panel = panel, year = rep(pl$time, ncol(g)), gap = as.vector(g),
             id = rep(colnames(g), each = length(pl$time)),
             series = rep(ifelse(colnames(g) == "Sweden", "Sweden", "Control countries"), each = length(pl$time)))
}
d7 <- bind_rows(gap_long(rep(TRUE, ncol(pl$gaps)), "Panel A. All 14 donors"), gap_long(pl$kept, "Panel B. MSPE within 20 times Sweden's"))
ggplot() +
  geom_line(data = filter(d7, series != "Sweden"), aes(x = year, y = gap, group = id), colour = "grey70") +
  geom_line(data = filter(d7, series == "Sweden"), aes(x = year, y = gap), colour = "black", linewidth = 0.8) +
  facet_wrap(~ panel) + geom_hline(yintercept = 0, linetype = "dotted") + tax_line +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(x = NULL, y = "Gap in metric tons per capita") + rep_theme()
c(kept_after_cut = sum(pl$kept) - 1L, sweden_rank_after_cut = pl$mspe$rank[1L], p_value_after_cut = pl$p_value)
```

**The ratio test (Figure 8) avoids the arbitrary cut by ranking every
country on its post-to-pre MSPE ratio; Sweden's ratio is the largest of
all 15, so the p-value is 1/15.** A large ratio needs both a good
pre-treatment fit and a large post-treatment gap, which is why the paper
prefers it with few donors.

```{r fig8, fig.height = 3.6, fig.cap = "Ratio test: post-treatment MSPE over pre-treatment MSPE for Sweden and the 14 donors (paper's Figure 8)."}
pl_all <- synth_placebo(sc_sweden)
plot_synth(pl_all, type = "ratio")
c(sweden_rank = pl_all$mspe$rank[1L], p_value = pl_all$p_value, sweden_ratio = pl_all$mspe$ratio[1L])
```

**The leave-one-out test (Figure 9) refits synthetic Sweden without each
donor that has weight above 0.001; the paper reads the range of
reductions, 8.8 percent without Switzerland to 13.0 percent without
Denmark, as the sensitivity to any single donor.** All refits keep the
estimate well away from zero and above the paper's DiD estimate of 8.1
percent.

```{r fig9, cache = TRUE, cache.extra = data_md5, fig.height = 3.4, fig.cap = "Leave-one-out distribution of synthetic Sweden (paper's Figure 9)."}
loo <- synth_loo(sc_sweden, min_weight = 0.001)
plot_synth(loo)
loo_tab <- loo$table %>% mutate(pct = sapply(seq_along(loo$fits), function(i) {
  e <- loo$fits[[i]]$effects; mean(e$gap[e$post] / e$synthetic[e$post]) }))
loo_tab %>% mutate(estimate = round(estimate, 3), pct = round(100 * pct, 1), pre_rmspe = round(pre_rmspe, 4)) %>%
  select(dropped, weight, estimate, pct, pre_rmspe)
```

**The full-sample test of the paper (Figure 10) re-estimates with all 24
OECD countries; the replication data contain only the 14 donors, so that
figure cannot be reproduced.** The paper reports that only the United
Kingdom would receive a sizeable weight (0.128) and that the estimate is
nearly unchanged.

# The recession as a possible confounder {#sec-confounder}

**Figures 11 and 12 answer the objection that the 1991 to 1993 recession,
not the tax, cut emissions.** The check reuses the emission weights on GDP
per capita. If synthetic Sweden reproduces Sweden's GDP through the
recession, the recession is common to both and cannot explain the
emissions gap. The paper then compares the two recessions: the GDP gap
fell by about 2,300 dollars in both, while the emissions gap rose by 0.086
tons in 1975 to 1978 and fell by 0.291 tons in 1989 to 1993. GDP per
capita is missing for Poland before 1970, so the check runs on 1970 to
2005; Poland's weight is zero.

```{r fig11, fig.height = 3.4, fig.cap = "GDP per capita, Sweden versus synthetic Sweden with the emission weights; shaded areas are the 1976 to 1978 and 1991 to 1993 recessions (paper's Figure 11)."}
gdp <- synth_control(carbon[carbon$year >= 1970, ], id = "country", time = "year", y = "GDP_per_capita", d = "treated",
                     weights = sc_sweden)
recessions <- data.frame(start = c(1976, 1991), end = c(1978, 1993))
gdp_paths <- gdp$effects %>% transmute(year = time, Sweden = treated, `Synthetic Sweden` = synthetic) %>%
  pivot_longer(-year, names_to = "series", values_to = "y")
ggplot() +
  geom_rect(data = recessions, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), fill = "grey85") +
  geom_line(data = gdp_paths, aes(x = year, y = y, colour = series, linetype = series)) + tax_line +
  scale_colour_manual(values = c("black", "grey45")) +
  scale_y_continuous(limits = c(0, 35000), labels = scales::comma) +
  labs(x = NULL, y = "GDP per capita (PPP, 2005 USD)", colour = NULL, linetype = NULL) + rep_theme()
```

```{r fig12, fig.height = 3.6, fig.cap = "Gaps between Sweden and synthetic Sweden in CO2 emissions (left axis) and GDP per capita (right axis) (paper's Figure 12)."}
gaps <- merge(data.frame(year = eff$time, emissions = eff$gap), data.frame(year = gdp$effects$time, gdp = gdp$effects$gap), by = "year")
ggplot() +
  geom_rect(data = recessions, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), fill = "grey85") +
  geom_hline(yintercept = 0, linetype = "dotted") + tax_line +
  geom_line(data = gaps, aes(x = year, y = emissions, colour = "CO2 emissions (left axis)")) +
  geom_line(data = gaps, aes(x = year, y = gdp / 5000, colour = "GDP per capita (right axis)")) +
  scale_y_continuous(limits = c(-0.4, 0.4), sec.axis = sec_axis(~ . * 5000, name = "Gap in GDP per capita (PPP, 2005 USD)", labels = scales::comma)) +
  scale_colour_manual(values = c("black", "#B22222")) +
  labs(x = NULL, y = "Gap in metric tons per capita", colour = NULL) + rep_theme()
change <- function(v, from, to) v[gaps$year == to] - v[gaps$year == from]
data.frame(recession = c("1975 to 1978", "1989 to 1993"),
           gdp_gap_change = c(change(gaps$gdp, 1975, 1978), change(gaps$gdp, 1989, 1993)),
           emissions_gap_change = c(change(gaps$emissions, 1975, 1978), change(gaps$emissions, 1989, 1993)),
           gdp_gap_post_average = mean(gaps$gdp[gaps$year >= 1990]) / mean(gdp$effects$synthetic[gdp$effects$post]))
```

# Disentangling the carbon tax and the VAT {#sec-disentangle}

**Table 3 estimates the gasoline demand response separately to the
carbon-tax-exclusive price and to the carbon tax, on Swedish annual data
for 1970 to 2011.** The model is log gasoline consumption per capita on
the real price excluding the carbon tax (with VAT), the real carbon tax
(with VAT), a dummy for 1991 onward, a trend, and, across columns, GDP per
capita, urbanization, and unemployment. Columns 5 and 6 instrument the
price with the real energy tax and with the crude oil price. Standard
errors are Newey-West with 16 lags. The table exists to show that the
carbon tax coefficient is about three times the price coefficient, and
that the difference survives controls and instruments.

```{r tab3}
reg42 <- reg %>% filter(year >= 1970, year <= 2011, !is.na(log_gas_cons))
f1 <- log_gas_cons ~ p_real_vat + real_CO2_tax_vat + d_CO2_tax + t
f2 <- update(f1, . ~ . + gdp_cap)
f3 <- update(f2, . ~ . + urban_pop)
f4 <- update(f3, . ~ . + unempl)
ols <- list(lm(f1, reg42), lm(f2, reg42), lm(f3, reg42), lm(f4, reg42))
iv_en <- ivreg(log_gas_cons ~ p_real_vat + real_CO2_tax_vat + d_CO2_tax + t + gdp_cap + urban_pop + unempl |
                 real_CO2_tax_vat + d_CO2_tax + t + gdp_cap + urban_pop + unempl + real_en_tax_vat, data = reg42)
iv_oil <- ivreg(log_gas_cons ~ p_real_vat + real_CO2_tax_vat + d_CO2_tax + t + gdp_cap + urban_pop + unempl |
                  real_CO2_tax_vat + d_CO2_tax + t + gdp_cap + urban_pop + unempl + oil_p_real, data = reg42)
models <- c(ols, list(iv_en, iv_oil))
nw <- lapply(models, function(m) NeweyWest(m, lag = 16, prewhite = FALSE))
weak_f <- sapply(list(iv_en, iv_oil), function(m) summary(m, diagnostics = TRUE)$diagnostics["Weak instruments", "statistic"])
equal_p <- sapply(seq_along(models), function(i) {
  b <- coef(models[[i]]); V <- nw[[i]]
  d <- b[["p_real_vat"]] - b[["real_CO2_tax_vat"]]
  v <- V["p_real_vat", "p_real_vat"] + V["real_CO2_tax_vat", "real_CO2_tax_vat"] - 2 * V["p_real_vat", "real_CO2_tax_vat"]
  2 * pnorm(-abs(d / sqrt(v)))
})
rows <- data.frame(term = c("p-value: beta1 = beta2", "Instrument F-statistic"),
                   rbind(formatC(equal_p, digits = 3, format = "f"), c(rep("", 4), formatC(weak_f, digits = 2, format = "f"))))
names(rows) <- c("term", paste0("(", 1:6, ")"))
modelsummary::modelsummary(
  setNames(models, c("OLS (1)", "OLS (2)", "OLS (3)", "OLS (4)", "IV EnTax (5)", "IV OilPrice (6)")),
  vcov = nw, stars = FALSE, fmt = 4,
  coef_map = c(p_real_vat = "Gas price with VAT", real_CO2_tax_vat = "Carbon tax with VAT", d_CO2_tax = "Dummy carbon tax",
               t = "Trend", gdp_cap = "GDP per capita", urban_pop = "Urban population", unempl = "Unemployment rate",
               `(Intercept)` = "Constant"),
  gof_map = list(list(raw = "nobs", clean = "Observations", fmt = 0), list(raw = "r.squared", clean = "R2", fmt = 2)),
  add_rows = rows, output = "kableExtra",
  title = "Estimation results from gasoline consumption regressions (paper's Table 3). Newey-West standard errors with 16 lags in parentheses.") %>%
  tab_style()
```

The coefficients and Newey-West standard errors match the paper's Table
3 to the third decimal in every column. The instrument F-statistics
differ: `ivreg`'s diagnostic is the classical first-stage F, while the
paper reports 3.57 and 310.9, presumably with heteroskedasticity and
autocorrelation robust standard errors; the conclusion is the same, the
energy tax is a weak instrument and the oil price is not.

**The elasticities follow by multiplying the semi-elasticities of column
4 by the 1990 real price of 8.48 SEK: the paper reports a price
elasticity of -0.51 and a carbon tax elasticity of -1.57.** The ratio of
about three is the paper's explanation for why a simulation using the
price elasticity alone understates the effect of the tax.

```{r elasticities}
b4 <- coef(ols[[4]])
c(price_semi = b4[["p_real_vat"]], tax_semi = b4[["real_CO2_tax_vat"]],
  price_elasticity = 8.48 * b4[["p_real_vat"]], tax_elasticity = 8.48 * b4[["real_CO2_tax_vat"]],
  ratio = b4[["real_CO2_tax_vat"]] / b4[["p_real_vat"]])
```

**Figure 13 uses the column 4 semi-elasticities to simulate emissions
under three scenarios and reads the carbon tax's share of the reduction
off the distances between the lines.** The three series are supplied in
the replication data: no carbon tax and no VAT, VAT without the carbon
tax, and both. Up to 2000 the two taxes contribute about equally; after
2000 the carbon tax increases dominate. The paper then applies the carbon
tax's share of the simulated reduction in each year to the synthetic
control gap of that year, and attributes 6.3 percent, or 0.17 tons per
capita, of the average post-treatment reduction to the carbon tax, 0.15
tons (5.5 percent) in 2000 and 0.27 tons (9.4 percent, three quarters of
the total) in 2005.

```{r fig13, fig.height = 3.4, fig.cap = "Simulated per-capita CO2 emissions from transport under three tax scenarios, 1970 to 2005 (paper's Figure 13)."}
d13 <- disent %>% filter(year >= 1970, year <= 2005) %>%
  select(year, `Carbon tax and VAT` = CarbonTaxandVAT, `No carbon tax, with VAT` = NoCarbonTaxWithVAT,
         `No carbon tax, no VAT` = NoCarbonTaxNoVAT) %>%
  pivot_longer(-year, names_to = "scenario", values_to = "y")
ggplot(d13, aes(x = year, y = y, colour = scenario, linetype = scenario)) + geom_line() + tax_line +
  scale_colour_manual(values = c("black", "grey45", "#B22222")) +
  labs(x = NULL, y = "Metric tons per capita (CO2 from transport)", colour = NULL, linetype = NULL) + rep_theme()
carbon_share <- disent %>% filter(year >= 1990, year <= 2005) %>%
  transmute(year, share = (CarbonTaxandVAT - NoCarbonTaxWithVAT) / (CarbonTaxandVAT - NoCarbonTaxNoVAT)) %>%
  left_join(eff %>% transmute(year = time, gap, synthetic), by = "year") %>%
  mutate(carbon_tax_gap = share * gap, carbon_tax_pct = carbon_tax_gap / synthetic)
carbon_only <- carbon_share %>%
  summarise(average_gap = mean(carbon_tax_gap), average_pct = mean(carbon_tax_pct),
            gap_2000 = carbon_tax_gap[year == 2000], pct_2000 = carbon_tax_pct[year == 2000],
            gap_2005 = carbon_tax_gap[year == 2005], pct_2005 = carbon_tax_pct[year == 2005], share_2005 = share[year == 2005])
round(as.data.frame(carbon_only), 3)
```

**Figure 14 puts the simulation next to the synthetic control gap; the
two agree through 2000 and diverge afterwards, because the simulation
holds the energy tax fixed while the real tax reform lowered it as the
carbon tax rose.** The paper reads the 2005 divergence, -0.757 against
-0.355, as what the carbon tax increases would have done had the energy
tax not been cut.

```{r fig14, fig.height = 3.4, fig.cap = "Gap in per-capita CO2 emissions from transport: synthetic control versus simulation (paper's Figure 14)."}
d14 <- disent %>% transmute(year, `Synthetic control result` = CO2_reductions_synth, `Simulation result` = CO2_reductions_simulation) %>%
  pivot_longer(-year, names_to = "series", values_to = "y")
ggplot() +
  geom_rect(data = data.frame(start = 2000, end = 2005), aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), fill = "grey85") +
  geom_line(data = d14, aes(x = year, y = y, colour = series, linetype = series)) +
  geom_hline(yintercept = 0, linetype = "dotted") + tax_line +
  scale_colour_manual(values = c("black", "#B22222")) +
  labs(x = NULL, y = "Gap in metric tons per capita", colour = NULL, linetype = NULL) + rep_theme()
disent %>% filter(year %in% c(2000, 2005)) %>% select(year, CO2_reductions_synth, CO2_reductions_simulation)
```

# Checks the paper did not run {#sec-fp}

**Ferman and Pinto (2021) showed that a synthetic control with an
imperfect pre-treatment fit is biased when the weights miss the treated
unit's level or when assignment is correlated with the factors, and
recommended four checks; Sweden passes them.** The demeaned synthetic
control matches trends only and gives almost the same estimate, so the
level is matched. The fit remains close after subtracting the period mean
of all countries. The weights are spread over four effective donors. And
the specification test does not find the demeaned synthetic control and
DiD to disagree.

```{r fp-checks, cache = TRUE, cache.extra = data_md5}
sc_all_lags <- synth_control(carbon, "country", "year", "CO2_transport_capita", "treated")
sc_demeaned <- synth_control(carbon, "country", "year", "CO2_transport_capita", "treated", demean = TRUE)
did <- sdid_weights(carbon, "country", "year", "CO2_transport_capita", "treated", estimator = "did")
spec <- synth_spec_test(sc_demeaned)
conf <- synth_conformal(sc_all_lags, per_period = TRUE)
data.frame(estimator = c("Andersson's specification", "All 30 lags", "Demeaned, all lags", "DiD"),
           estimate = c(sc_sweden$estimate, sc_all_lags$estimate, sc_demeaned$estimate, did$estimate),
           pre_rmspe = c(sc_sweden$pre_rmspe, sc_all_lags$pre_rmspe, sc_demeaned$pre_rmspe, NA),
           concentration = c(sc_sweden$concentration[["l2"]], sc_all_lags$concentration[["l2"]], sc_demeaned$concentration[["l2"]], 1 / 14))
c(demeaned_intercept = sc_demeaned$intercept, spec_test_p = spec$p_value, conformal_joint_p = conf$p_value,
  per_period_rejections_at_5pct = sum(conf$per_period$p_value <= 0.05))
```

```{r fig-demeaned, fig.height = 3.2, fig.cap = "Sweden and synthetic Sweden after subtracting each year's mean over all 15 countries: the fit the weights achieved beyond the common trend (Ferman and Pinto's Figure 2.C for this application)."}
plot_synth(sc_sweden, type = "demeaned_path")
```

# Conclusion {#sec-conclusion}

**The replication reproduces the paper's estimates and every robustness
check within the tolerance one expects from a different optimizer.** The
country weights agree with Table 2 to within
`r fmt(max(abs(tab2$Weight - tab2$Published)), 2)`, the average
post-treatment reduction is `r pct(-headline$average_pct)` against the
paper's 10.9 percent, the in-time placebos find nothing, the ratio test
puts Sweden first among 15, the leave-one-out range brackets the estimate,
and the recession comparison reproduces the paper's numbers to three
decimals. Table 3 and the disentangling follow from the supplied series.

**What the package adds is the imperfect-fit reading.** The demeaned
estimate, the detrended fit, the weight concentration, and the
specification test are the checks a referee would now ask for, and they
support the paper's design.

# Session information {.unnumbered}

```{r session}
sessionInfo()
```
