---
title: "Causal Inference in Economics and Marketing"
subtitle: "Section 07: Difference-in-Differences and Its Variants"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, fig.align = "center", fig.pos = "H",
  out.width = "95%", collapse = TRUE, comment = "#>"
)
# Code and output in footnotesize.
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
# This lecture lives in inst/lectures/07_did/ of the package source. When
# knitted from the source tree it loads the development version of the
# package; when the package is installed it uses the installed helpers.
helper_rel <- file.path("vignette_helpers", "07_Difference_in_Differences_and_Panel", "did_lecture_helpers.R")
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
  helper_path <- file.path(pkg_root, "inst", helper_rel)
} else {
  library(causalmetrics)
  helper_path <- system.file(helper_rel, package = "causalmetrics")
}
library(fixest)
library(dplyr)
library(ggplot2)
library(kableExtra)
source(helper_path)
n_sims_staggered <- 40
n_sims_covariates <- 200
fmt <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
run_honest <- requireNamespace("HonestDiD", quietly = TRUE)
```

# The problem {#sec-problem}

## Same question, a new source of variation

Sections 03 and 04 identified treatment effects by comparing treated and
untreated units that look alike. Selection on unobservables broke that
strategy. Difference-in-differences (DiD) uses a second dimension: time.
It compares the *change* in outcomes of treated units to the *change* among
untreated units. Level differences between the groups cancel. Only the
difference in trends must be assumed away.

That assumption is **parallel trends**: absent treatment, the treated
group's average outcome would have moved like the untreated group's. It
allows selection on anything fixed over time. It allows common shocks. It
does not allow selection on future trends.

DiD is the right tool when three things hold.

- A group is exposed to a policy at a known date and another group is not.
- Both groups are observed before and after that date.
- The reason for exposure is not the group's outcome trajectory.

The data can be a panel (the same units over time) or repeated
cross-sections (fresh draws from the same populations each period).

## One experiment, the classic regression breaks three ways

In practice DiD is run as a two-way fixed effects (TWFE) regression,
$$
Y_{it} = \alpha_i + \lambda_t + \beta\, D_{it} + \varepsilon_{it},
$$
with $D_{it} = 1$ once unit $i$ is treated. Section \@ref(sec-framework)
shows that in the two-group, two-period case $\beta$ is exactly the DiD of
means. Applied work then extends the regression in three ways: adoption
dates differ across units (staggered adoption), leads and lags of treatment
are added to trace dynamics, and covariates are added linearly. Each
extension can fail, and the failures are large.

**Design.** The staggered experiment follows Sant'Anna's lectures. There
are 1000 units in 40 states. States adopt in one of four cohorts, 1986,
1992, 1998, or 2004. The panel runs from 1980 to 2003, so the 2004 cohort is
untreated within the sample and serves as the comparison group. The effect
grows with exposure and differs across cohorts:
$\tau_{it} = \mu_g (e + 1)$ at event time $e = t - g$, with
$\mu = (3, 2, 1)$ for the three treated cohorts. Untreated outcomes have
unit effects, a common trend, and cohort-specific levels.

```{r staggered-experiment, cache = TRUE, cache.extra = tools::md5sum(helper_path)}
results <- run_staggered_experiment(n_sims = n_sims_staggered, n_units = 1000, seed = 1)
tab <- summarise_staggered_experiment(results)
pick <- function(estimator, e, column) tab[[column]][tab$estimator == estimator & tab$event_time == e]
```

**Estimators.** Five estimators of the event-study parameters, the average
effect $e$ periods after adoption for $e = 0, \dots, 5$.

| Estimator | What it does |
|---|---|
| Static TWFE | one coefficient on $D_{it}$, reported at every $e$ |
| Dynamic TWFE, binned | leads and lags of adoption, endpoints binned at $-5$ and $6+$ |
| Sun-Abraham | cohort-by-event-time coefficients, last cohort as comparison, aggregated |
| Callaway-Sant'Anna, not yet treated | `att_gt()` cells with not-yet-treated comparisons, aggregated by `aggregate_att()` |
| Imputation (BJS) | `did_imputation()`: unit and time effects fitted on untreated cells, effects imputed |

```{r tab-staggered, echo = FALSE}
tab %>% filter(event_time %in% c(0, 2, 4)) %>%
  mutate(across(c(truth, mean, bias, sd), ~ round(.x, 2))) %>%
  arrange(event_time, estimator) %>% select(event_time, estimator, truth, mean, bias, sd) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Event time", "Estimator", "Truth", "Mean estimate", "Bias", "SD"),
      caption = "Five estimators of the event-study parameters in the staggered design, 40 simulations.") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")
```

```{r fig-staggered, echo = FALSE, fig.height = 4.8, fig.cap = "Mean estimate and a two-standard-deviation range across simulations. The two TWFE regressions miss the truth by a wide margin; the three estimators built from clean two-by-two comparisons track it."}
plot_staggered_experiment(results)
```

The second failure needs only two periods. The design is the Kang-Schafer
style example of Sant'Anna's lecture 5: four covariates drive both selection
into treatment and the trend of untreated outcomes, and the true effect is
zero.

```{r covariate-experiment, cache = TRUE, cache.extra = tools::md5sum(helper_path)}
cov_results <- run_covariate_experiment(n_sims = n_sims_covariates, n = 1000, seed = 1)
cov_tab <- summarise_covariate_experiment(cov_results, truth = 0)
pick_cov <- function(estimator, column) cov_tab[[column]][cov_tab$estimator == estimator]
```

```{r tab-covariates, echo = FALSE}
kbl(cov_tab, booktabs = TRUE, digits = 2, linesep = "",
    col.names = c("Estimator", "Bias", "SD", "RMSE", "Mean SE", "95\\% coverage"), escape = FALSE,
    caption = "Two-period design with covariate-specific trends; true effect 0; 200 simulations.") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9)
```

## Reading the results

- **Static TWFE** reports one number, `r fmt(pick("Static TWFE", 0, "mean"))`,
  for every event time. It overstates the immediate effect by
  `r fmt(pick("Static TWFE", 0, "bias"))` and understates the effect four
  periods out by `r fmt(-pick("Static TWFE", 4, "bias"))`. Its spread across
  simulations is ten times that of the good estimators, because the
  weights it puts on cohorts and periods move with the realized cohort
  sizes. The regression averages many two-by-two comparisons, and some of
  them use already-treated units, whose effects are still growing, as
  controls (Section \@ref(sec-staggered-problems)).
- **Dynamic TWFE** is biased at every event time, with bias
  `r fmt(pick("Dynamic TWFE, binned", 0, "bias"))` at $e = 0$ and
  `r fmt(pick("Dynamic TWFE, binned", 4, "bias"))` at $e = 4$, and its
  pre-treatment coefficients are not zero either. Each coefficient is
  contaminated by effects from other event times and cohorts.
- **Sun-Abraham, Callaway-Sant'Anna, and imputation** are centred on the
  truth. They differ only in which comparisons they allow and which
  pre-periods they use (Section \@ref(sec-staggered-solutions)).
- **TWFE with covariates** in the two-period design has bias
  `r fmt(pick_cov("TWFE + X", "bias"))` and coverage
  `r fmt(pick_cov("TWFE + X", "coverage"))`. Interacting the covariates with
  the post indicator helps only if the trend is linear in $X$. Outcome
  regression, inverse probability weighting, and the doubly robust estimator
  all have coverage near 0.95 (Section \@ref(sec-covariates)).

None of these failures come from heterogeneity itself. They come from
letting least squares choose the comparisons. The cure is the same
throughout the lecture: write down the target, the assumption, and the
comparison, and build the estimator from them.

## Roadmap

Section \@ref(sec-framework) sets up potential outcomes, the assumptions,
and the identification argument. Section \@ref(sec-covariates) handles
covariates. Section \@ref(sec-dynamics) studies dynamics with two groups.
Sections \@ref(sec-staggered-problems) and \@ref(sec-staggered-solutions)
diagnose and fix staggered adoption. Section \@ref(sec-assess) asks how to
assess parallel trends, Section \@ref(sec-inference) how to do inference.
Section \@ref(sec-beyond) covers continuous doses, treatments that switch
off, triple differences, and nonlinear models. Section \@ref(sec-sdid)
covers synthetic DiD. Sections \@ref(sec-practice) and \@ref(sec-extensions)
close with replications, a checklist, and pointers.

# Just enough framework {#sec-framework}

## Potential outcomes indexed by adoption time

Units $i = 1, \dots, n$ are observed in periods $t = 1, \dots, T$. Let
$G_i \in \{2, \dots, T\} \cup \{\infty\}$ be the period in which unit $i$
is first treated, with $G_i = \infty$ for units never treated in the sample.
Treatment is absorbing: once treated, always treated (treatments that switch
off are in Section \@ref(sec-beyond)). The potential outcome $Y_{it}(g)$ is
the outcome in period $t$ if the unit were first treated in $g$. Observed
outcomes are $Y_{it} = \sum_g \mathbf{1}\{G_i = g\}\, Y_{it}(g)$.

The building block of everything that follows is the **group-time average
treatment effect**
$$
ATT(g, t) = \mathbb{E}\bigl[Y_{it}(g) - Y_{it}(\infty) \mid G_i = g\bigr],
\qquad t \ge g .
$$
It is the effect at calendar time $t$ of having started treatment at $g$,
for the cohort that did start at $g$. The canonical two-by-two is the case
$T = 2$, $G \in \{2, \infty\}$, and $ATT = ATT(2, 2)$. Summaries across
$(g, t)$, by exposure length, cohort, or calendar time, come in Section
\@ref(sec-staggered-solutions).

The target is an effect on the treated. Parallel trends will let us
impute what treated units would have done without treatment; it says
nothing about what untreated units would have done with it. The ATE needs
a second assumption in the other direction, which is rarely credible.

## Three assumptions

> **Assumption 1 (no interference).** $Y_{it} = \sum_g \mathbf{1}\{G_i = g\} Y_{it}(g)$.
> One unit's treatment does not change another unit's outcome.

> **Assumption 2 (no anticipation).** $Y_{it}(g) = Y_{it}(\infty)$ for all
> $t < g$. Effects are zero before treatment starts.

> **Assumption 3 (parallel trends).** For the two-by-two,
> $$
> \mathbb{E}[Y_{2}(\infty) - Y_{1}(\infty) \mid G = 2] = \mathbb{E}[Y_{2}(\infty) - Y_{1}(\infty) \mid G = \infty].
> $$
> With staggered adoption, for every treated cohort $g$ and every $t \ge g$,
> either the *never-treated* version
> $$
> \mathbb{E}[Y_{t}(\infty) - Y_{t-1}(\infty) \mid G = g] = \mathbb{E}[Y_{t}(\infty) - Y_{t-1}(\infty) \mid G = \infty],
> $$
> or the *not-yet-treated* version, which replaces the right-hand side by
> the units still untreated at $s \ge t$: $\mathbb{E}[\,\cdot \mid D_{s} = 0, G \ne g]$.
> The *conditional* version adds "given $X$" to both sides.

Three remarks.

- The two-by-two assumption restricts only the untreated potential
  outcome, and only its change. Levels may differ between groups by any
  amount; that is what "selection on fixed unobservables" means.
- The never-treated and not-yet-treated versions are different
  assumptions. The second uses more comparison units, so it is more
  precise, but it requires later cohorts to be valid controls for earlier
  ones. Both restrict post-treatment periods only. TWFE and the imputation
  estimator implicitly assume parallel trends in *all* periods, including
  before treatment (Section \@ref(sec-staggered-solutions)).
- With covariates, an overlap condition is needed: $P(G = g \mid X) < 1$,
  so every treated unit has comparable untreated units.

## Identification of the 2x2 ATT: skeleton {#sec-identification}

Compare treated and untreated units after treatment:
$$
\mathbb{E}[Y_2 \mid G = 2] - \mathbb{E}[Y_2 \mid G = \infty]
= \underbrace{ATT}_{\text{target}} + \underbrace{\mathbb{E}[Y_2(\infty) \mid G = 2] - \mathbb{E}[Y_2(\infty) \mid G = \infty]}_{\text{selection bias}} .
$$
The bias term contains the counterfactual $\mathbb{E}[Y_2(\infty) \mid G = 2]$.
Three steps replace it by observables.

1. **Parallel trends** rewrites the counterfactual as the treated group's
   pre-period level plus the untreated group's change:
   $\mathbb{E}[Y_2(\infty) \mid G = 2] = \mathbb{E}[Y_1(\infty) \mid G = 2] + \mathbb{E}[Y_2(\infty) - Y_1(\infty) \mid G = \infty]$.
2. **No anticipation** turns $Y_1(\infty)$ into $Y_1(2)$ for the treated,
   which is the observed $Y_1$.
3. **No interference** turns every remaining potential outcome into an
   observed one.

Hence
$$
ATT = \bigl(\mathbb{E}[Y_2 \mid G = 2] - \mathbb{E}[Y_1 \mid G = 2]\bigr)
- \bigl(\mathbb{E}[Y_2 \mid G = \infty] - \mathbb{E}[Y_1 \mid G = \infty]\bigr) .
$$
The right-hand side is a difference of two before-after differences, hence
the name. Appendix \@ref(app-identification) writes the steps out.

**Estimation.** Replace expectations by sample means. In a panel, average
the change $\Delta Y_i = Y_{i2} - Y_{i1}$ within each group and take the
difference. The same number is the coefficient on the interaction in the
regression $Y_{it} = \alpha + \gamma\, \mathbf{1}\{G_i = 2\} + \lambda\, \mathbf{1}\{t = 2\} + \beta\, \mathbf{1}\{G_i = 2\}\mathbf{1}\{t = 2\} + \varepsilon_{it}$,
the coefficient on the treated dummy in the regression of $\Delta Y$ on it,
and the coefficient on $D_{it}$ in a TWFE regression with unit and time
effects. All four agree to the last digit.

```{r two-by-two-equivalence}
two <- sim_did_panel(n_units = 500, n_periods = 2, groups = 2, x_trend = 0, seed = 2)
two$treat <- as.integer(two$g == 2)
two$post <- as.integer(two$time == 2)
wide <- with(two, data.frame(id = id[time == 1], treat = treat[time == 1],
                             dy = y[time == 2] - y[time == 1]))
by_hand <- mean(wide$dy[wide$treat == 1]) - mean(wide$dy[wide$treat == 0])
c(by_hand = by_hand,
  interaction = coef(feols(y ~ treat * post, data = two))[["treat:post"]],
  first_difference = coef(feols(dy ~ treat, data = wide))[["treat"]],
  twfe = coef(feols(y ~ treat:post | id + time, data = two))[["treat:post"]],
  att_gt = att_gt(two, id = "id", time = "time", group = "g", y = "y", n_boot = 0)$att_gt$att)
```

The regression is a device for computing means. It does not change the
target or the assumption. Once the design has more than two groups or two
periods, that convenience disappears: the regression starts choosing
comparisons on its own (Section \@ref(sec-staggered-problems)).

## From two periods to long differences {#sec-long-differences}

With several post-treatment periods, apply the argument period by period.
Parallel trends from $t - 1$ to $t$ for every $t \ge g$ chains into
$$
\mathbb{E}[Y_t(\infty) \mid G = g] = \mathbb{E}[Y_{g-1} \mid G = g] + \mathbb{E}[Y_t - Y_{g-1} \mid G = \infty],
$$
so
$$
ATT(g, t) = \mathbb{E}[Y_t - Y_{g-1} \mid G = g] - \mathbb{E}[Y_t - Y_{g-1} \mid G = \infty] .
$$
The comparison is a **long difference** from the last pre-treatment period
$g - 1$ to $t$. Intermediate periods drop out; they are used only through
the assumption. Before treatment, the same assumptions imply that the
short difference from $t - 1$ to $t$ has zero DiD, which is the placebo
check of Section \@ref(sec-assess).

```{r long-difference}
one <- sim_did_panel(n_units = 400, n_periods = 6, groups = 4, never_share = 0.5, seed = 3)
long_diff <- one %>%
  group_by(id, g) %>%
  summarise(ld = y[time == 6] - y[time == 3], .groups = "drop") %>%
  summarise(att = mean(ld[g == 4]) - mean(ld[g == 0])) %>% pull(att)
cell <- att_gt(one, id = "id", time = "time", group = "g", y = "y", n_boot = 0)$att_gt
c(long_difference_by_hand = long_diff, att_gt_4_6 = cell$att[cell$group == 4 & cell$time == 6])
```

## What parallel trends really assumes {#sec-functional-form}

Parallel trends is a statement about a specific transformation of the
outcome. If it holds for $Y$ it generally fails for $\log Y$, and the two
DiDs answer different questions with different assumptions. Roth and
Sant'Anna (2023) show that parallel trends holds for every monotone
transformation only when it holds for the whole distribution of untreated
outcomes, and that this happens only when the population is a mixture of
two types: units for which treatment is as good as random, and units
whose untreated outcome distribution does not change over time.

A two-line example. Treated units have higher outcome levels. Nothing
happens over time except a common additive shift, so parallel trends holds
in levels and the true effect is zero.

```{r functional-form}
set.seed(4)
n <- 4000
treat <- rep(c(0, 1), each = n / 2)
y1 <- exp(rnorm(n, mean = ifelse(treat == 1, 2, 1), sd = 0.5))
y2 <- y1 + 3
did <- function(v1, v2) mean((v2 - v1)[treat == 1]) - mean((v2 - v1)[treat == 0])
c(levels = did(y1, y2), logs = did(log(y1), log(y2)))
```

The DiD in logs is far from zero. A common shift of 3 units is a smaller
proportional change for the group with higher levels, so the log trends are
not parallel. The choice of scale is part of the identifying assumption and
should be argued from the economics of the outcome, not from fit. Section
\@ref(sec-assess) shows a test that can reject the joint hypothesis that
parallel trends holds on every scale.

## Inference from influence functions {#sec-influence}

Write the panel two-by-two estimator as
$\hat\theta = \bar{\Delta Y}_{G = 2} - \bar{\Delta Y}_{G = \infty}$. A
first-order expansion gives
$$
\sqrt{n}\,(\hat\theta - \theta) = \frac{1}{\sqrt n}\sum_{i=1}^n \xi_i + o_p(1),
\qquad
\xi_i = \frac{\mathbf{1}\{G_i = 2\}}{P(G = 2)}\bigl(\Delta Y_i - \mathbb{E}[\Delta Y \mid G = 2]\bigr)
- \frac{\mathbf{1}\{G_i = \infty\}}{P(G = \infty)}\bigl(\Delta Y_i - \mathbb{E}[\Delta Y \mid G = \infty]\bigr).
$$
$\xi_i$ is the **influence function**: unit $i$'s contribution to the
estimation error. It has mean zero, so the central limit theorem gives
$\sqrt{n}(\hat\theta - \theta) \to N(0, \mathbb{E}[\xi^2])$, and the
standard error is $\sqrt{\sum_i \hat\xi_i^2} / n$. Three things follow.

- **Clustering** at a level above the unit sums $\xi_i$ within clusters
  before squaring. The unit is the smallest admissible level, because the
  two periods of a unit are one draw.
- **Many parameters** share one machinery. Every $ATT(g, t)$ has an
  influence function, and any weighted average of cells has the
  corresponding weighted average of influence functions. Covariances come
  free.
- **The multiplier bootstrap** draws i.i.d. weights $V_i$ with mean zero
  and variance one and forms $\hat\theta^* = \hat\theta + n^{-1}\sum_i V_i \hat\xi_i$.
  No re-estimation is needed. The distribution of
  $\max_k |\hat\theta_k^* - \hat\theta_k| / \hat{se}_k$ across draws gives
  the critical value of a **simultaneous confidence band** over all cells.

`att_gt()` returns the influence functions and uses them for all of this.
On the two-period data, the influence-function standard error equals the
heteroskedasticity-robust standard error of the first-difference
regression up to the finite-sample factor, and the bootstrap agrees with
both.

```{r influence-se}
fit0 <- att_gt(two, id = "id", time = "time", group = "g", y = "y", n_boot = 0)
fitb <- att_gt(two, id = "id", time = "time", group = "g", y = "y", n_boot = 999, seed = 1)
xi <- fit0$inffunc[, 1]
c(influence_function = sqrt(sum(xi^2)) / nrow(wide),
  att_gt_analytic = fit0$att_gt$std.error,
  fixest_hc1 = se(feols(dy ~ treat, data = wide, vcov = "hetero"))[["treat"]],
  multiplier_bootstrap = fitb$att_gt$std.error)
```

Appendix \@ref(app-influence) derives $\xi_i$ for the panel and the
repeated cross-section cases and states the bootstrap result.

# Covariates: three faces of conditional parallel trends {#sec-covariates}

## Why covariates

Parallel trends is often more plausible within groups of similar units than
overall. Employment in northern and southern states, earnings of young and
old workers, sales of large and small retailers all trend differently.
If treated units are richer in the fast-trending type, the unconditional
comparison confounds the effect with composition. The fix is to assume
parallel trends **conditional on** pre-treatment covariates $X$:
$$
\mathbb{E}[Y_2(\infty) - Y_1(\infty) \mid G = 2, X] = \mathbb{E}[Y_2(\infty) - Y_1(\infty) \mid G = \infty, X] .
$$
Two rules. Covariates must be measured before treatment, or be unaffected
by it; controlling for a post-treatment variable blocks part of the effect.
And overlap is needed: $P(G = 2 \mid X) < 1$, so every treated unit has
untreated units with similar $X$.

A classic case is the **Ashenfelter dip**: people enter training after a bad
year, so their pre-period earnings are transitorily low and mean-revert
afterwards. Conditioning on a longer earnings history, or moving the base
period back, is the remedy; a two-period DiD with the dip year as the base
is not.

## The tempting regression and why it fails {#sec-twfe-covariates}

The natural move is to add $X$ to the regression,
$$
Y_{it} = \alpha + \gamma\, \mathbf{1}\{G_i = 2\} + \lambda\, \mathbf{1}\{t = 2\} + \beta\, \mathbf{1}\{G_i = 2\}\mathbf{1}\{t = 2\} + X_i'\delta + \varepsilon_{it}.
$$
Read off its four conditional means. The untreated change is
$\lambda$ for every value of $X$: the regression assumes there are no
covariate-specific trends, which is exactly what the covariates were
supposed to allow. The treated change is $\lambda + \beta$ for every $X$:
the effect is assumed homogeneous across covariate values. When either
restriction fails, $\beta$ is not the ATT. The Section 1 table showed a
bias of `r fmt(pick_cov("TWFE + X", "bias"))` with zero coverage.

Interacting $X$ with the post indicator allows covariate-specific trends
that are linear in $X$. That worked in Section 1 because the design was
linear. It is a modelling assumption, not a fix.

```{r covariates-one-draw}
kd <- simulate_covariate_dgp(n = 1000, seed = 5)
kd$treat <- as.integer(kd$g == 2); kd$post <- as.integer(kd$time == 2)
c(twfe_plus_x = coef(feols(y ~ treat:post + x1 + x2 + x3 + x4 | id + time, data = kd))[["treat:post"]],
  twfe_post_x = coef(feols(y ~ treat:post + post:(x1 + x2 + x3 + x4) | id + time, data = kd))[["treat:post"]],
  dr = att_gt(kd, id = "id", time = "time", group = "g", y = "y",
              x = c("x1", "x2", "x3", "x4"), n_boot = 0)$att_gt$att)
```

## Outcome regression, IPW, doubly robust {#sec-three-faces}

Separate identification from estimation. Under conditional parallel
trends, with $D = \mathbf{1}\{G = 2\}$, $\Delta Y = Y_2 - Y_1$,
$p(X) = P(D = 1 \mid X)$ and $m_\Delta(X) = \mathbb{E}[\Delta Y \mid D = 0, X]$,
the ATT has three equivalent expressions.

| Face | Estimand | What must be modelled |
|---|---|---|
| Outcome regression (Heckman, Ichimura, Todd 1997) | $\mathbb{E}[\Delta Y \mid D = 1] - \mathbb{E}[m_\Delta(X) \mid D = 1]$ | the untreated change given $X$ |
| Inverse probability weighting (Abadie 2005) | $\mathbb{E}[w_1 \Delta Y] - \mathbb{E}[w_0 \Delta Y]$, $w_1 = \tfrac{D}{\mathbb{E}[D]}$, $w_0 = \tfrac{p(X)(1 - D)/(1 - p(X))}{\mathbb{E}[p(X)(1 - D)/(1 - p(X))]}$ | the propensity score |
| Doubly robust (Sant'Anna and Zhao 2020) | $\mathbb{E}\bigl[(w_1 - w_0)\,(\Delta Y - m_\Delta(X))\bigr]$ | either one correctly |

The doubly robust estimand is the IPW estimand applied to the residual
$\Delta Y - m_\Delta(X)$. If the outcome model is right, the residual has
conditional mean zero among the untreated, so any weights work. If the
propensity score is right, the weights balance $X$, so any outcome model
works. When both are right it reaches the semiparametric efficiency bound.
`att_gt(method = )` implements the three with a logit and a linear
regression as working models, reproducing the `DRDID` package.

```{r three-faces}
sapply(c("reg", "ipw", "dr"), function(m)
  att_gt(kd, id = "id", time = "time", group = "g", y = "y",
         x = c("x1", "x2", "x3", "x4"), method = m, n_boot = 0)$att_gt[, c("att", "std.error")])
```

The guarantee is about *specification*, not about *fit*. When the true
drivers are nonlinear transformations of $X$ and the analyst enters $X$
linearly in both models, nothing is guaranteed.

```{r covariate-misspecified, cache = TRUE, cache.extra = tools::md5sum(helper_path)}
mis_results <- run_covariate_experiment(n_sims = 100, n = 1000, seed = 2, misspecify = TRUE)
mis_tab <- summarise_covariate_experiment(mis_results, truth = 0)
pick_mis <- function(estimator, column) mis_tab[[column]][mis_tab$estimator == estimator]
```

```{r tab-misspecified, echo = FALSE}
kbl(mis_tab[mis_tab$estimator != "TWFE + post:X", ], booktabs = TRUE, digits = 2, linesep = "",
    col.names = c("Estimator", "Bias", "SD", "RMSE", "Mean SE", "95\\% coverage"), escape = FALSE,
    caption = "Both working models misspecified (the true drivers are nonlinear in X); true effect 0; 100 simulations.") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9)
```

In this design the linear outcome model happens to extrapolate well, so
outcome regression is nearly unbiased (bias
`r fmt(pick_mis("Outcome regression", "bias"))`). The misspecified logit
produces extreme weights, IPW is biased
(`r fmt(pick_mis("IPW", "bias"))`), and the doubly robust estimator, which
applies those weights to the residuals, is the most biased of the three
(`r fmt(pick_mis("Doubly robust", "bias"))`, coverage
`r fmt(pick_mis("Doubly robust", "coverage"))`). Double robustness insures
against one wrong model. With two wrong models it can be worse than either
alone. The way out is flexible estimation of both nuisances, which is the
next subsection.

## DR-DiD is the ATT score of Section 04 on the differenced outcome {#sec-dr-dml}

Compare the doubly robust estimand with the ATT score of Section 04. With
$\Delta Y$ in place of $Y$ and $m_\Delta$ in place of $\mu_0$, they are the
same object: the augmented inverse probability weighting score for the
average effect on the treated. The only difference is the normalization
of the control weights, by their own mean here and by $\mathbb{E}[D]$ in
Section 04, which is immaterial in large samples. Everything Section 04
established therefore carries over: the score is Neyman orthogonal, its
remainder is the product of the two nuisance errors, and machine learners
with cross-fitting may estimate the nuisances.

```{r dr-equals-dml}
wide_kd <- with(kd, data.frame(id = id[time == 1], d = treat[time == 1],
                               dy = y[time == 2] - y[time == 1],
                               x1 = x1[time == 1], x2 = x2[time == 1], x3 = x3[time == 1], x4 = x4[time == 1]))
p_hat <- fitted(glm(d ~ x1 + x2 + x3 + x4, data = wide_kd, family = binomial))
m_hat <- predict(lm(dy ~ x1 + x2 + x3 + x4, data = wide_kd[wide_kd$d == 0, ]), newdata = wide_kd)
c(att_gt_dr = att_gt(kd, id = "id", time = "time", group = "g", y = "y",
                     x = c("x1", "x2", "x3", "x4"), n_boot = 0)$att_gt$att,
  est_dml_att = est_dml(wide_kd, y = "dy", d = "d", model = "irm", estimand = "ATT",
                        p_hat = p_hat, mu0_hat = m_hat, mu1_hat = m_hat)$estimate)
```

## Machine-learning nuisances and cross-fitting {#sec-ml-nuisances}

Supplying `mlr3` learners to `att_gt()` switches each cell to cross-fitted
nuisances: the propensity score and the untreated change are predicted
out of fold, and the doubly robust score is evaluated on the held-out
units. This is the recipe of Section 04 inside each $(g, t)$ cell. It
needs learners that converge fast enough, roughly $n^{-1/4}$ for the
product of the two errors, and it pays off when the functional forms are
unknown, as in the misspecified design.

```{r ml-nuisances, eval = requireNamespace("mlr3", quietly = TRUE) && requireNamespace("mlr3learners", quietly = TRUE) && requireNamespace("ranger", quietly = TRUE)}
library(mlr3); library(mlr3learners)
mis <- simulate_covariate_dgp(n = 2000, seed = 6, misspecify = TRUE)
rbind(
  parametric_dr = att_gt(mis, id = "id", time = "time", group = "g", y = "y",
                         x = c("x1", "x2", "x3", "x4"), n_boot = 0)$att_gt[, c("att", "std.error")],
  forest_dr = att_gt(mis, id = "id", time = "time", group = "g", y = "y", x = c("x1", "x2", "x3", "x4"),
                     learner_p = lrn("classif.ranger", predict_type = "prob", num.trees = 300),
                     learner_or = lrn("regr.ranger", num.trees = 300),
                     folds = 5, seed = 1, n_boot = 0)$att_gt[, c("att", "std.error")]
)
```

## Repeated cross-sections {#sec-rcs}

With repeated cross-sections the same units are not observed twice, so
$\Delta Y$ does not exist. The estimators use four conditional means, one
per group and period, and need one more assumption: the composition of
each group does not change between periods (stationarity). Two doubly
robust versions exist. The first models only the untreated outcomes in
each period; the second also models the treated outcomes, which makes it
efficient (Sant'Anna and Zhao 2020). `att_gt(sampling = "rcs")` uses the
second. For a given sample size, repeated cross-sections are always less
precise than a panel, and more so when the pre and post samples are
unbalanced.

```{r rcs}
set.seed(7)
rcs <- kd[sample(nrow(kd), 1500), ]
rcs$row <- seq_len(nrow(rcs))
rbind(
  panel = att_gt(kd, id = "id", time = "time", group = "g", y = "y",
                 x = c("x1", "x2", "x3", "x4"), n_boot = 0)$att_gt[, c("att", "std.error", "n_treated", "n_control")],
  repeated_cross_section = att_gt(rcs, id = "row", time = "time", group = "g", y = "y",
                                  x = c("x1", "x2", "x3", "x4"), sampling = "rcs", n_boot = 0)$att_gt[, c("att", "std.error", "n_treated", "n_control")]
)
```

# Dynamics with two groups {#sec-dynamics}

## ATT(g,t) as an event study {#sec-event-study}

One treated cohort $g$, one comparison group, many periods. Section
\@ref(sec-long-differences) identified $ATT(g, t)$ for $t \ge g$ by the
long difference from $g - 1$. Indexing by event time $e = t - g$ turns the
cells into an **event study**: the effect $e$ periods after adoption.
Cells with $e < 0$ are placebos. Under no anticipation and parallel trends
they equal zero, and they are what a pre-trends plot shows.

Two conventions exist for the pre-treatment cells.

- **Varying base period** (default): the cell at $e < 0$ compares $t - 1$
  to $t$, a short difference. It shows period-by-period deviations from
  parallel trends.
- **Universal base period**: every cell is compared to $g - 1$. Post
  cells are unchanged. Pre cells show the cumulative deviation between
  $t$ and $g - 1$, so the plot is anchored at $e = -1$ and reads like a
  standard event-study figure. Use it for sensitivity analysis (Section
  \@ref(sec-assess)).

```{r es-two-groups}
esd <- sim_did_panel(n_units = 500, n_periods = 10, groups = 6, never_share = 0.5, mu = 1, seed = 8)
es_var <- att_gt(esd, id = "id", time = "time", group = "g", y = "y", control_group = "never",
                 base_period = "varying", n_boot = 199, seed = 1)
es_uni <- att_gt(esd, id = "id", time = "time", group = "g", y = "y", control_group = "never",
                 base_period = "universal", n_boot = 199, seed = 1)
plot_event_study(event_study_frame("Varying base period" = es_var, "Universal base period" = es_uni))
```

The post-treatment cells coincide. The pre-treatment cells differ only in
what they cumulate; both are centred at zero here because parallel trends
holds in the design.

## Anticipation {#sec-anticipation}

If units respond before the official start, because the policy was
announced or phased in, no anticipation fails for the periods just before
$g$. The base period $g - 1$ is then already "treated", and every post
effect is measured against a contaminated baseline. The remedy is
**limited anticipation**: assume no effects before $g - k$ for a known $k$,
strengthen parallel trends to hold from $g - k - 1$ on, and use $g - k - 1$
as the base period. `att_gt(anticipation = k)` does exactly that.

```{r anticipation}
ant <- sim_did_panel(n_units = 600, n_periods = 8, groups = 5, never_share = 0.5, mu = 2,
                     anticipation = 1, seed = 9)
truth_ant <- attr(ant, "att_gt")
cmp <- lapply(0:1, function(k) {
  f <- att_gt(ant, id = "id", time = "time", group = "g", y = "y", control_group = "never",
              anticipation = k, n_boot = 0)$att_gt
  data.frame(anticipation = k, event_time = f$event_time, att = round(f$att, 2))
})
tab_ant <- merge(cmp[[1]], cmp[[2]], by = "event_time", suffixes = c("_k0", "_k1"))
tab_ant$truth <- ifelse(tab_ant$event_time >= 0, truth_ant$att[match(tab_ant$event_time + 5, truth_ant$time)],
                        ifelse(tab_ant$event_time == -1, 0.5 * truth_ant$att[1], 0))
tab_ant[tab_ant$event_time >= -2 & tab_ant$event_time <= 2, c("event_time", "att_k0", "att_k1", "truth")]
```

In the design, half of the first-period effect appears one period early.
With `anticipation = 0` every post-treatment effect is understated by that
amount, and the cell at $e = -1$ looks like a pre-trend. With
`anticipation = 1` the base moves to $g - 2$, the post effects are right,
and the cell at $e = -1$ correctly measures the anticipation effect. The
lesson is to date treatment at the announcement, or to set $k$ from
institutional knowledge, not from the plot.

## TWFE event-study regressions {#sec-twfe-es}

The regression version adds leads and lags of adoption,
$$
Y_{it} = \alpha_i + \lambda_t + \sum_{k \ne -1} \gamma_k \mathbf{1}\{t - G_i = k\} + \varepsilon_{it},
$$
with the period before adoption omitted. With one cohort, a never-treated
comparison group, and **all** leads and lags included, the model is
saturated and $\hat\gamma_k$ equals the universal-base $ATT(g, g + k)$ of
`att_gt()` exactly. Binning the endpoints, for example one dummy for
$k \le -4$ and one for $k \ge 3$, is an extra restriction: it assumes the
effect is constant inside the bin, and the other coefficients change.

```{r twfe-es}
esd$rel <- ifelse(esd$g > 0, esd$time - esd$g, -1000)
tw_all <- feols(y ~ i(rel, ref = c(-1, -1000)) | id + time, data = esd, cluster = ~id)
esd$rel_bin <- ifelse(esd$g == 0, -1000, pmin(pmax(esd$rel, -3), 2))
tw_bin <- feols(y ~ i(rel_bin, ref = c(-1, -1000)) | id + time, data = esd, cluster = ~id)
fr <- event_study_frame("att_gt" = es_uni, "TWFE_all" = tw_all, "TWFE_binned" = tw_bin)
fr %>% filter(event_time %in% c(-3, -2, 0, 1, 2)) %>%
  select(method, event_time, estimate) %>%
  tidyr::pivot_wider(names_from = method, values_from = estimate) %>%
  as.data.frame()
```

The saturated regression and `att_gt()` agree to the last digit. The
binned regression does not, and the difference is not a bug but a
restriction. Report all event times and let the reader choose what to
look at.

## Simultaneous confidence bands {#sec-bands}

An event study is many parameters. Pointwise 95 percent intervals cover
each cell 95 percent of the time, but the probability that *all* cells are
covered at once is much lower, and the eye reads the curve as a whole. A
**simultaneous band** fixes the joint coverage. `att_gt()` computes it
from the multiplier bootstrap: draw the joint distribution of the
studentized cells, take the maximum absolute deviation across cells, and
use its 95th percentile as the critical value. The band is wider than the
pointwise interval by that factor.

```{r bands}
c(pointwise = qnorm(0.975), simultaneous = es_uni$crit_val)
es_uni$pretest
```

The pre-test is a joint Wald test that all pre-treatment cells are zero,
computed from the covariance of their influence functions. It answers a
narrower question than it seems to; Section \@ref(sec-assess) returns to
it.

# Staggered adoption: what goes wrong {#sec-staggered-problems}

## Static TWFE as a weighted average {#sec-bacon}

With several adoption dates, the static regression
$Y_{it} = \alpha_i + \lambda_t + \beta D_{it} + \varepsilon_{it}$ still
computes a difference-in-differences, but a weighted average of every
two-by-two comparison the data allow (Goodman-Bacon 2021):
$$
\hat\beta = \sum_{k} s_{kU}\, \hat\beta_{kU} + \sum_{k < l} \bigl( s_{kl}^{k}\, \hat\beta_{kl}^{k} + s_{kl}^{l}\, \hat\beta_{kl}^{l} \bigr).
$$
The three types are: cohort $k$ against never-treated units; the earlier
cohort $k$ against the later cohort $l$ before $l$ adopts, where $l$ is a
clean control; and the later cohort $l$ against the earlier cohort $k$
*after* $k$ has adopted. In the third type the control group is already
treated. If its effect is still changing, its trend is not the untreated
trend, and the comparison is biased. The weights are positive and sum to
one, but they depend on group sizes and on how central each adoption date
is in the panel, not on anything a researcher would choose.

```{r bacon}
sd1 <- simulate_staggered_dgp(n_units = 1000, seed = 11)
bd <- bacon_decomp(sd1, id = "id", time = "year", y = "y", d = "treated")
bd$by_type
c(twfe = bd$twfe, weighted_sum = bd$check)
```

The later-versus-earlier comparisons carry a substantial weight and a
negative estimate, although every true effect is positive. That is the
mechanism behind the Section 1 result.

## Negative weights {#sec-negative-weights}

de Chaisemartin and D'Haultfoeuille (2020) give a second reading. Under
parallel trends, the same coefficient is a weighted sum of the treatment
effects of the treated unit-periods,
$\beta = \sum_{(i,t):\, D_{it} = 1} w_{it}\, \tau_{it}$, with weights that
sum to one but can be negative. The weight of a cell is the residual of
$D_{it}$ on unit and period effects, divided by the sum of those residuals.
Cells treated for a long time in periods when most units are treated get
negative weights, because their treatment status is *below* what the two
fixed effects predict.

```{r negative-weights}
tw <- twfe_weights(sd1, id = "id", time = "year", d = "treated", y = "y")
c(share_negative = tw$share_negative, sum_negative = tw$sum_negative)
tw$by_cell %>% group_by(g) %>%
  summarise(first_year = min(time), last_year = max(time), min_cell_weight = min(weight),
            n_negative = sum(weight < 0), cohort_weight = sum(weight), .groups = "drop") %>%
  as.data.frame()
```

A negative weight on a positive effect pulls the estimate down. When
effects grow with exposure, the negative weights fall on the largest
effects, and the coefficient can even change sign.

## Dynamic TWFE and contamination {#sec-contamination}

Adding leads and lags does not help. Sun and Abraham (2021) show that
with heterogeneous effects across cohorts, the coefficient on lag $k$ is a
weighted average of $ATT(g, g + k)$ across cohorts, with weights that need
not be convex, *plus* contamination terms from every other lag and lead.
Two consequences matter in practice. Pre-treatment coefficients can be
nonzero even when no anticipation and parallel trends hold, so a
pre-trends plot from this regression can reject a valid design. And
binning the endpoints, which imposes constant effects far from adoption,
moves the contamination into the window of interest.

```{r contamination}
sd1$rel <- ifelse(sd1$g > 0, sd1$year - sd1$g, -1000)
tw_dyn <- feols(y ~ i(rel, ref = c(-1, -1000)) | id + year, data = sd1, cluster = ~state)
cs1 <- att_gt(sd1, id = "id", time = "year", group = "g", y = "y", control_group = "notyet", n_boot = 0)
cs1_dyn <- aggregate_att(cs1, type = "dynamic", min_e = -5, max_e = 5, n_boot = 0)
truth_sd1 <- attr(sd1, "truth")
plot_event_study(event_study_frame("TWFE, all leads and lags" = tw_dyn, "Callaway-Sant'Anna" = cs1_dyn)) +
  geom_point(data = truth_sd1, aes(x = event_time, y = att), inherit.aes = FALSE, shape = 4, size = 3) +
  coord_cartesian(xlim = c(-5, 5)) +
  labs(subtitle = "Crosses: true event-study parameters; no anticipation in the design")
```

```{r contamination-table}
event_study_frame("TWFE" = tw_dyn, "CS" = cs1_dyn) %>%
  filter(event_time %in% c(-4, -2, 0, 2, 4)) %>%
  select(method, event_time, estimate) %>%
  tidyr::pivot_wider(names_from = method, values_from = estimate) %>%
  left_join(truth_sd1, by = "event_time") %>% as.data.frame()
```

The regression shows a pre-trend that does not exist and misstates the
post-treatment effects. The cell-based estimator shows neither.

## What is not the problem {#sec-not-the-problem}

None of this is caused by heterogeneity as such. If effects were the same
for every cohort and constant over time, the forbidden comparisons would
be harmless and static TWFE would be fine.

```{r homogeneous}
sd0 <- simulate_staggered_dgp(n_units = 1000, mu = c(2, 2, 2, 2), seed = 12)
sd0$y <- sd0$y - sd0$tau + 2 * sd0$treated   # constant effect of 2
c(twfe = coef(feols(y ~ treated | id + year, data = sd0))[["treated"]],
  cs = aggregate_att(att_gt(sd0, id = "id", time = "year", group = "g", y = "y", n_boot = 0),
                     type = "simple", n_boot = 0)$overall$estimate)
```

The problem is that least squares chooses the comparisons and the weights.
Heterogeneous dynamics are the normal case in economics and marketing,
so the estimator must be built to survive them. That is the next section.

# Staggered adoption: reliable estimators {#sec-staggered-solutions}

## Identification, aggregation, inference as three steps {#sec-three-steps}

Every reliable estimator in this section follows the same order, which is
the reverse of "run the regression and interpret the coefficient".

1. **Identify** each $ATT(g, t)$ from a stated parallel trends assumption
   and a stated comparison group (Section \@ref(sec-long-differences)).
2. **Aggregate** the cells with weights chosen for the question, by
   exposure length, cohort, or calendar time.
3. **Infer** from the influence functions of the cells, so that the
   aggregate inherits valid standard errors and simultaneous bands.

Least squares performs all three steps at once with weights nobody chose.
Separating them is the whole content of Callaway and Sant'Anna (2021).

## The building block ATT(g,t) {#sec-att-gt}

`att_gt()` estimates every cell. The comparison group is a choice with
consequences.

- **Never treated**: one fixed comparison group. Parallel trends is
  assumed between each cohort and the never-treated units only.
- **Not yet treated**: every unit still untreated at the later of $t$ and
  the base period. More units, hence more precision, at the price of
  assuming parallel trends among all cohorts as well.

```{r att-gt-both}
cs_never <- att_gt(sd1, id = "id", time = "year", group = "g", y = "y", control_group = "never",
                   n_boot = 199, seed = 1)
cs_notyet <- att_gt(sd1, id = "id", time = "year", group = "g", y = "y", control_group = "notyet",
                    n_boot = 199, seed = 1)
tidy(cs_never) %>% filter(group == 1992, event_time %in% 0:3) %>%
  mutate(se_notyet = tidy(cs_notyet)$std.error[match(term, tidy(cs_notyet)$term)]) %>%
  select(group, time, event_time, estimate, se_never = std.error, se_notyet)
```

```{r fig-att-gt, fig.height = 4.5, fig.cap = "Group-time effects with not-yet-treated comparisons; error bars are pointwise intervals and ribbons are the simultaneous band."}
tidy(cs_notyet) %>%
  ggplot(aes(x = time, y = estimate, colour = factor(post))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = band.low, ymax = band.high, group = 1), fill = "grey70", alpha = 0.3, colour = NA) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.3) +
  facet_wrap(~ group, labeller = label_both) +
  scale_colour_manual(values = c("0" = "grey40", "1" = "#0072B2"), labels = c("pre-treatment", "post-treatment")) +
  labs(x = "Year", y = "ATT(g, t)", colour = NULL) + did_theme()
```

## Aggregations {#sec-aggregations}

With cohort shares $\pi_g = P(G = g \mid G \ne \infty)$, the four
summaries of `aggregate_att()` are

| Type | Parameter | Question |
|---|---|---|
| dynamic | $\theta_D(e) = \sum_g \pi_{g \mid e}\, ATT(g, g + e)$ | effect $e$ periods after adoption |
| group | $\theta_S(g) = \frac{1}{T - g + 1}\sum_{t \ge g} ATT(g, t)$ | average effect experienced by cohort $g$ |
| calendar | $\theta_C(t) = \sum_{g \le t} \pi_{g \mid t}\, ATT(g, t)$ | average effect in period $t$ among the treated |
| simple | $\sum_{g \le t} \pi_{g}\, ATT(g, t) / \sum_{g \le t} \pi_g$ | one number, overweighting early cohorts |

The dynamic aggregate at $e$ averages only the cohorts observed $e$
periods after adoption. Its composition therefore changes with $e$, and
the curve can bend for that reason alone. `balance_e = e_0` keeps only
cohorts observed for at least $e_0$ periods, so every point averages the
same cohorts.

```{r aggregations}
agg_dyn <- aggregate_att(cs_notyet, type = "dynamic", min_e = -4, max_e = 5)
agg_grp <- aggregate_att(cs_notyet, type = "group")
agg_cal <- aggregate_att(cs_notyet, type = "calendar")
agg_simple <- aggregate_att(cs_notyet, type = "simple")
rbind(dynamic = agg_dyn$overall, group = agg_grp$overall, calendar = agg_cal$overall, simple = agg_simple$overall)
agg_grp$by[, c("group", "estimate", "std.error")]
```

The four overall numbers answer four questions. The group aggregate
weights cohorts by size; the simple aggregate weights *cells* by cohort
size, so cohorts treated early, with many post periods, dominate. Report
the one that matches the question, and say which it is.

## Sun and Abraham {#sec-sun-abraham}

Sun and Abraham (2021) fix the dynamic regression rather than abandon it:
interact the event-time dummies with cohort dummies, use the never-treated
(or last-treated) cohort as the only comparison, estimate the saturated
regression, and aggregate the cohort-specific coefficients with cohort
shares. Without covariates this is numerically the Callaway-Sant'Anna
estimator with the never-treated comparison and the universal base
period. `fixest::sunab()` does the interaction and the aggregation.

```{r sunab}
sa1 <- feols(y ~ sunab(g, year) | id + year, data = sd1, cluster = ~id)
cs_uni <- att_gt(sd1, id = "id", time = "year", group = "g", y = "y", control_group = "never",
                 base_period = "universal", n_boot = 0)
event_study_frame("Sun-Abraham" = sa1, "CS never, universal" = aggregate_att(cs_uni, type = "dynamic", n_boot = 0)) %>%
  filter(event_time %in% c(-3, 0, 2, 4)) %>% select(method, event_time, estimate) %>%
  tidyr::pivot_wider(names_from = method, values_from = estimate) %>% as.data.frame()
```

## Imputation estimators {#sec-imputation}

Borusyak, Jaravel, and Spiess (2024) and Gardner (2022) take the model
seriously in the untreated state only. Fit $Y_{it} = \alpha_i + \lambda_t + \varepsilon_{it}$
on untreated observations, never-treated units and not-yet-treated
periods, impute $\hat Y_{it}(0)$ for every treated observation, and
average $Y_{it} - \hat Y_{it}(0)$ by event time or cohort. Wooldridge
(2021) obtains the same estimates from one regression with cohort-by-period
interactions.

The estimator uses every pre-treatment period as the baseline, not only
$g - 1$. That is efficient when parallel trends holds in all periods and
the errors are homoskedastic and serially uncorrelated, and it is the
reason its standard errors are often smaller. The price is a stronger
assumption, parallel trends in all periods, and a separate pre-trends
check, run on the untreated sample with lead dummies.

```{r imputation}
imp1 <- did_imputation(sd1, id = "id", time = "year", group = "g", y = "y", horizons = 0:5, pre_window = 3)
imp1
```

## Stacked DiD and local-projections DiD {#sec-stacked}

Two regression-based alternatives are common in applied work.

**Stacked DiD** (Cengiz et al. 2019) builds one clean data set per cohort:
the cohort, the never-treated units, and the units not yet treated within
an event window, then stacks the data sets and runs a TWFE event study
with data-set-specific unit and period effects. Every comparison is clean
by construction. The implicit weights across cohorts are those of least
squares on the stacked sample, roughly proportional to cohort size.

```{r stacked}
stack <- bind_rows(lapply(c(1986, 1992, 1998), function(g0) {
  sd1 %>% filter(g == g0 | g == 0 | g > g0 + 5, year >= g0 - 5, year <= g0 + 5) %>%
    mutate(stack_id = g0, rel = ifelse(g == g0, year - g0, -1000))
}))
stacked <- feols(y ~ i(rel, ref = c(-1, -1000)) | id^stack_id + year^stack_id, data = stack, cluster = ~id)
```

**Local-projections DiD** (Dube, Girardi, Jorda, and Taylor 2023) runs,
for each horizon $h$, a regression of $Y_{i,t+h} - Y_{i,t-1}$ on an
indicator for adopting at $t$, using only clean controls (units not yet
treated by $t + h$). It is a sequence of long-difference two-by-twos, one
per horizon, and delivers the same logic as the cell-based estimator in
regression form.

## Comparing estimators {#sec-compare}

```{r fig-compare, fig.height = 4.5, fig.cap = "Four heterogeneity-robust estimators on one draw of the staggered design, with the truth as crosses."}
imp1_frame <- event_study_frame("Imputation" = imp1) %>% filter(event_time >= 0)
cmp_frame <- event_study_frame("Callaway-Sant'Anna" = agg_dyn, "Sun-Abraham" = sa1,
                               "Stacked" = stacked, "Imputation" = imp1_frame) %>%
  filter(event_time >= -4, event_time <= 5)
plot_event_study(cmp_frame) +
  geom_point(data = truth_sd1, aes(x = event_time, y = att), inherit.aes = FALSE, shape = 4, size = 3)
```

```{r compare-table}
cmp_frame %>% filter(event_time %in% c(-2, 0, 3, 5)) %>% select(method, event_time, estimate) %>%
  tidyr::pivot_wider(names_from = method, values_from = estimate) %>%
  left_join(truth_sd1, by = "event_time") %>% as.data.frame()
```

| Estimator | Parallel trends used | Comparison group | Covariates | Precision |
|---|---|---|---|---|
| Callaway-Sant'Anna | post-treatment periods only | never or not yet treated | regression, IPW, doubly robust, per cell | baseline |
| Sun-Abraham | post-treatment periods only | never or last treated | linear in the regression | equals CS without covariates |
| Imputation (BJS, Gardner, Wooldridge) | all periods | all untreated observations | linear in the first stage | higher under homoskedastic, uncorrelated errors |
| Stacked | all periods within the window | never and not yet treated in the window | linear in the regression | depends on the window |

The estimators agree when their assumptions hold. They disagree when
parallel trends holds only after treatment, when pre-periods are
contaminated by anticipation, or when covariates matter, and then the
choice is about assumptions, not about software.

## Covariates in staggered designs {#sec-staggered-covariates}

Covariates enter `att_gt()` cell by cell, with a propensity score and an
outcome regression fitted on each cell's sample, so the conditional
parallel trends assumption of Section \@ref(sec-covariates) is used at the
cell level. The regression-based estimators can include covariates only
linearly, and a time-invariant covariate must be interacted with time to
matter at all.

```{r staggered-covariates}
sdx <- sim_did_panel(n_units = 800, n_periods = 9, groups = c(4, 7), never_share = 0.4,
                     mu = c(2, 1), x_trend = 2, x_select = 2, seed = 13)
sdx$x1t <- sdx$x1 * sdx$time
cs_x <- aggregate_att(att_gt(sdx, id = "id", time = "time", group = "g", y = "y", x = "x1",
                             control_group = "notyet", n_boot = 0), type = "dynamic", n_boot = 0)
truth_x <- attr(sdx, "att_gt") %>% mutate(event_time = time - group) %>%
  inner_join(cs_x$weights, by = c("event_time", "group")) %>%
  group_by(event_time) %>% summarise(truth = sum(weight * att), .groups = "drop")
cs_nox <- aggregate_att(att_gt(sdx, id = "id", time = "time", group = "g", y = "y",
                               control_group = "notyet", n_boot = 0), type = "dynamic", n_boot = 0)
sa_x <- feols(y ~ sunab(g, time) + x1t | id + time, data = sdx, cluster = ~id)
twfe_x <- feols(y ~ treated + x1t | id + time, data = sdx, cluster = ~id)
c(static_twfe_plus_x1t = coef(twfe_x)[["treated"]])
event_study_frame("CS with x1" = cs_x, "CS without x1" = cs_nox, "Sun-Abraham + x1*t" = sa_x) %>%
  filter(event_time %in% 0:2) %>% select(method, event_time, estimate) %>%
  tidyr::pivot_wider(names_from = method, values_from = estimate) %>%
  left_join(truth_x, by = "event_time") %>% as.data.frame()
```

The static regression with the covariate trend is far from every true
effect. The cell estimator without the covariate is biased upward, because
early adopters have higher `x1` and hence steeper untreated trends; with
the covariate it is centred. Sun-Abraham with `x1 * t` also works here,
but only because the covariate trend is linear in the design.

# Assessing parallel trends {#sec-assess}

## Pre-trends tests and what they cannot tell {#sec-pretrends}

A pre-trends plot tests parallel trends *before* treatment. The
identifying assumption is about trends *after* treatment. The test is
informative only through an extra step: if trends were parallel before,
they are probably parallel after. That step is economics, not statistics,
and it fails when treatment is timed in response to a change in the
outcome, which is the usual reason policies are adopted.

Even on its own terms the test is weak. Roth (2022) shows that
pre-trends tests have low power against violations large enough to
matter. `pretrend_power()` makes this concrete: it takes the event-study
estimates and their covariance, posits a linear violation of parallel
trends with a given slope, and reports the probability that the standard
pre-test (any pre-period coefficient significant at 5 percent) rejects.

```{r pretrend-power}
pp <- pretrend_power(agg_dyn, n_sim = 3000, seed = 1)
pp$detectable
```

A violation that the test detects only half the time would bias the
effect at $e = 4$ by five times that slope. In this design the effects are
large relative to such slopes; in many applications they are not.

## Conditioning on passing the pre-test {#sec-pretest-bias}

Worse, the estimates that survive the test are not a random sample of
the estimates. When a real violation exists, the draws that fail to reject
are those in which the noise happened to offset the pre-trend, and the
same noise pushes the post-treatment estimates away from the truth. The
bias conditional on passing exceeds the unconditional bias.

```{r pretest-bias}
pp50 <- pretrend_power(agg_dyn, slope = pp$detectable$slope[1], n_sim = 3000, seed = 1)
cbind(power = round(pp50$slopes[[1]]$power, 2), pp50$slopes[[1]]$bias)
```

The practical conclusion is not to abandon pre-trends plots, but to stop
treating a passed test as a license. Report the plot, report the power of
the test, and quantify how conclusions change under plausible violations.

## Sensitivity analysis {#sec-honestdid}

Rambachan and Roth (2023) replace "parallel trends holds" by a bound on
how far it can fail, indexed by the pre-treatment evidence. Two families
of bounds are common.

- **Relative magnitudes** $\Delta^{RM}(\bar M)$: the post-treatment
  violation in any period is at most $\bar M$ times the largest
  pre-treatment violation. $\bar M = 1$ says trends after treatment
  deviate no more than they did before.
- **Smoothness** $\Delta^{SD}(M)$: the violation evolves smoothly; the
  change in its slope between consecutive periods is at most $M$.
  $M = 0$ allows a linear pre-trend to continue.

For each bound the method returns a confidence set for the effect that
is valid for every violation in the set. The **breakdown value** is the
bound at which zero enters the set. `honestdid_inputs()` extracts the
coefficients and their covariance from a dynamic aggregate estimated with
the universal base period, so the reference period is exactly zero.

```{r honestdid, eval = run_honest}
csu_h <- att_gt(sd1, id = "id", time = "year", group = "g", y = "y", control_group = "notyet",
                base_period = "universal", n_boot = 0)
dyn_h <- aggregate_att(csu_h, type = "dynamic", min_e = -4, max_e = 3, n_boot = 0)
hi <- honestdid_inputs(dyn_h)
c(numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods)
l1 <- HonestDiD::basisVector(1, hi$numPostPeriods)
orig <- HonestDiD::constructOriginalCS(betahat = hi$betahat, sigma = hi$sigma, numPrePeriods = hi$numPrePeriods,
                                       numPostPeriods = hi$numPostPeriods, l_vec = l1)
rm <- HonestDiD::createSensitivityResults_relativeMagnitudes(
  betahat = hi$betahat, sigma = hi$sigma, numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods,
  Mbarvec = c(0.5, 1, 2), l_vec = l1, grid.lb = orig$lb - 2, grid.ub = orig$ub + 2)
sm <- HonestDiD::createSensitivityResults(
  betahat = hi$betahat, sigma = hi$sigma, numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods,
  Mvec = c(0, 0.05, 0.1), l_vec = l1)
rbind(as.data.frame(orig)[, c("lb", "ub", "method")], as.data.frame(rm)[, c("lb", "ub", "method")],
      as.data.frame(sm)[, c("lb", "ub", "method")])
```

The grid for the relative-magnitudes set is passed explicitly, because
the default grid is built for coefficients on the scale of the package's
examples and can miss the estimate entirely.

The rows show the effect at $e = 0$ under the original assumption, under
relative-magnitude violations up to half, once, and twice the largest
pre-treatment deviation, and under smoothness bounds of 0, 0.05, and 0.1
per period. The estimated pre-trends are small here, so even $\bar M = 2$
leaves the conclusion intact. When they are not small, this table is the
honest statement of what the design can deliver.

## Functional-form test {#sec-ff-test}

Section \@ref(sec-functional-form) showed that parallel trends on one
scale does not imply it on another. Roth and Sant'Anna (2023) give a
test of the stronger claim that it holds on every scale, which is the
same as parallel trends of the whole distribution. Under that claim the
counterfactual distribution of treated outcomes after treatment is
$F_{1,\text{pre}} + F_{0,\text{post}} - F_{0,\text{pre}}$, which must be a
distribution function: its implied density must be non-negative
everywhere. A negative implied mass anywhere rejects distributional
parallel trends, and with it the hope that the scale does not matter.

```{r ff-test}
breaks <- quantile(c(y1, y2), probs = seq(0, 1, 0.1))
mass <- function(v) as.numeric(table(cut(v, breaks, include.lowest = TRUE))) / length(v)
implied <- mass(y1[treat == 1]) + mass(y2[treat == 0]) - mass(y1[treat == 0])
round(implied, 3)
```

The negative entries say that no monotone transformation of this outcome
satisfies parallel trends, so the level and log analyses cannot both be
right, which is what Section \@ref(sec-functional-form) found.

## Placebos and design checks {#sec-placebos}

- **Placebo outcomes**: an outcome the policy cannot affect should show no
  effect; if it does, the comparison group is trending differently.
- **Placebo timing**: pretend treatment started earlier and estimate on
  the pre-period only; the cells of `att_gt()` with `post = 0` are this
  check, done period by period.
- **Anticipation**: date treatment at the announcement or use
  `anticipation = k`; a jump at $e = -1$ is a symptom (Section
  \@ref(sec-anticipation)).
- **Composition**: in repeated cross-sections, test that the distribution
  of covariates within groups does not change over time.
- **Estimator agreement**: Callaway-Sant'Anna with never and with not-yet
  comparisons, Sun-Abraham, and imputation should agree when parallel
  trends holds in all periods; a gap between them is evidence about the
  pre-period assumption, not a choice of software.

# Inference {#sec-inference}

## What level to cluster {#sec-cluster-level}

Two views justify clustering. In the **sampling** view, units are drawn
from a population in clusters, and standard errors must reflect the
cluster draw: at least the unit, because a unit's periods are one draw,
and higher when units are sampled by state or firm. In the **design**
view, treatment is assigned at some level, and the uncertainty is about
which clusters were assigned; standard errors should then be clustered
at the assignment level. Both views lead to the same practical rule:
cluster at the level of treatment assignment, and never below the unit.

```{r cluster-level}
agg_unit <- aggregate_att(cs_notyet, type = "simple", n_boot = 199, seed = 1)
cs_state <- att_gt(sd1, id = "id", time = "year", group = "g", y = "y", control_group = "notyet",
                   cluster = "state", n_boot = 199, seed = 1)
agg_state <- aggregate_att(cs_state, type = "simple")
rbind(cluster_unit = agg_unit$overall, cluster_state = agg_state$overall)
```

Treatment is assigned by state, so the state-clustered standard error is
the relevant one; it is several times larger than the unit-clustered one.

## Few clusters {#sec-few-clusters}

Cluster-robust standard errors rely on a central limit theorem across
clusters. With few clusters, and especially few *treated* clusters, the
approximation fails, standard errors are biased downward, and tests
over-reject. The wild cluster bootstrap repairs moderate cases but not
the case of one or two treated clusters. Randomization inference asks a
different question: if treatment timing had been assigned at random
across clusters, how unusual is the observed statistic? Under random
timing this test is exact for the sharp null of no effect for any unit,
and with a studentized statistic it is also asymptotically valid for the
weak null of a zero average effect (Roth and Sant'Anna 2023).

```{r few-clusters}
few <- sim_did_panel(n_units = 300, n_periods = 8, groups = 5, never_share = 0.75, mu = 1, seed = 14)
few$state <- (few$id - 1) %/% 15 + 1
few$g <- ave(few$g, few$state, FUN = function(v) v[1])
few$treated <- as.integer(few$g > 0 & few$time >= few$g)
c(treated_states = length(unique(few$state[few$g > 0])), states = length(unique(few$state)))
asy <- feols(y ~ treated | id + time, data = few, cluster = ~state)
perm <- did_permutation_test(few, id = "id", time = "time", group = "g", y = "y",
                             cluster = "state", n_perm = 499, seed = 1)
c(asymptotic_p = fixest::pvalue(asy)[["treated"]], permutation_p = perm$p.value)
```

## Design-based inference and random timing {#sec-random-timing}

When treatment timing really is random, as in a randomized rollout,
parallel trends is implied but the DiD estimator is not efficient. Roth
and Sant'Anna (2023) show that the efficient estimator in a large class
subtracts from the simple post-treatment comparison a regression
adjustment on pre-treatment differences, in the spirit of covariate
adjustment in experiments, and that standard-error reductions of a
factor of two are common. They also provide the studentized permutation
test used above. `staggered_efficient()` implements the estimator
(reproducing the `staggered` package): `beta = 1` is the
Callaway-Sant'Anna comparison, `beta = 0` the difference in means, and
the default is the plug-in efficient `beta*`. The gain is largest when
pre-treatment outcomes predict post-treatment outcomes weakly, so that
subtracting the pre-period difference adds noise rather than removing it.

```{r random-timing}
set.seed(19)
n_rt <- 600; T_rt <- 8
g_rt <- sample(c(3, 5, 7, Inf), n_rt, replace = TRUE)   # random timing
y_rt <- matrix(rnorm(n_rt * T_rt), n_rt) + outer(rep(1, n_rt), seq_len(T_rt) / 4)   # little serial correlation
for (i in seq_len(n_rt)) if (is.finite(g_rt[i])) y_rt[i, seq_len(T_rt) >= g_rt[i]] <- y_rt[i, seq_len(T_rt) >= g_rt[i]] + 1
rt <- data.frame(id = rep(seq_len(n_rt), each = T_rt), time = rep(seq_len(T_rt), n_rt),
                 g = rep(g_rt, each = T_rt), y = as.vector(t(y_rt)))
rbind(
  efficient = staggered_efficient(rt, id = "id", time = "time", group = "g", y = "y", estimand = "simple",
                                  n_perm = 199, seed = 1)$estimates[, c("estimate", "std.error", "beta", "fisher_p.value")],
  did_beta_1 = staggered_efficient(rt, id = "id", time = "time", group = "g", y = "y", estimand = "simple",
                                   beta = 1, n_perm = 199, seed = 1)$estimates[, c("estimate", "std.error", "beta", "fisher_p.value")]
)
```

With independent noise across periods the efficient `beta*` is near zero,
and the DiD standard error is about 1.4 times larger, the square root of
two that differencing costs. Random timing is a stronger assumption than
parallel trends; under it the gain in precision is free, and without it
the DiD comparison remains the safe choice.

# Beyond binary absorbing treatments {#sec-beyond}

## Continuous and multi-valued doses {#sec-dose}

Many treatments come in doses: a minimum wage increase of 5 or 15
percent, an ad budget, a tax rate. With a dose $D \ge 0$ and untreated
units at $D = 0$, two questions arise, and Callaway, Goodman-Bacon, and
Sant'Anna (2024) show they need different assumptions.

- **Level effects** $ATT(d \mid d) = \mathbb{E}[Y_2(d) - Y_2(0) \mid D = d]$:
  the effect of dose $d$ for units that received $d$. Parallel trends
  between dose-$d$ units and untreated units identifies it by the usual
  DiD, $\mathbb{E}[\Delta Y \mid D = d] - \mathbb{E}[\Delta Y \mid D = 0]$.
- **Causal responses** $ACRT(d \mid d) = \partial \mathbb{E}[Y_2(l) \mid D = d] / \partial l$
  at $l = d$: the effect of a marginal increase in the dose. Comparing
  $ATT(d \mid d)$ across doses does not identify it, because units that
  chose different doses may have different responses; the comparison
  mixes the causal response with selection into dose. Identification
  needs **strong parallel trends**: every dose group would have followed
  the same average path at every dose, which rules out selection on
  gains.

The two-way fixed effects coefficient on a continuous dose is
$\mathrm{Cov}(\Delta Y, D) / \mathrm{Var}(D)$. It has two readings. It is a
weighted sum of level effects with weights that can be negative, and it
is a weighted sum of causal-response increments with non-negative weights
that integrate to one. The second reading is the appealing one and is
the one that needs strong parallel trends. `att_dose()` reports both
weight sets so the reader sees which assumption is being invoked.

```{r dose}
set.seed(15)
nd <- 1500
dose <- ifelse(runif(nd) < 0.3, 0, runif(nd, 0.5, 3))
dd <- data.frame(id = rep(seq_len(nd), each = 2), time = rep(1:2, nd), dose = rep(dose, each = 2))
dd$y <- rep(rnorm(nd), each = 2) + 0.5 * dd$time +
  (dd$time == 2) * (2 * dd$dose - 0.3 * dd$dose^2) + rnorm(2 * nd, sd = 0.5)
ad <- att_dose(dd, id = "id", time = "time", y = "y", dose = "dose", dose_type = "spline",
               n_bins = 4, n_boot = 199, seed = 1)
ad$by_dose[, c("bin", "dose", "att", "std.error")]
ad$twfe_weights
```

```{r fig-dose, fig.height = 3.6, fig.cap = "Causal responses by dose bin (points) and the spline estimate with a uniform band; the true response is 2 - 0.6 d."}
plot_att_dose(ad, "acr")
```

In the design the response curve is $2d - 0.3 d^2$, so the true causal
response is $2 - 0.6 d$ and falls with the dose. The level weights put
negative mass on low doses; the slope weights are positive and sum to one
over the dose gaps. A researcher who reports the TWFE coefficient as "the
effect of a unit of dose" is using the second reading and therefore
strong parallel trends, whether or not they say so.

## Treatments that turn on and off {#sec-on-off}

When treatment can be switched off and on, a unit's history is a
sequence $(d_1, \dots, d_T)$, and there are $2^{T-1}$ possible groups
instead of $T$. The building block becomes $ATT(\mathbf{d}, t)$, the effect
of sequence $\mathbf{d}$ versus never being treated. Identification works as
in Section \@ref(sec-long-differences) once groups are defined by the
first exposure date and parallel trends is assumed across all sequences.
Mechanically, `att_gt()` with the first exposure period as the group
variable computes the right cells. The difficulty is interpretation: the
cell for a first-exposure cohort at time $t$ averages units that stayed
treated, switched off, and switched back on. de Chaisemartin and
D'Haultfoeuille (2024) propose normalizing the event-study aggregate by
the average number of treated periods, which reads as an effect "per unit
of treatment" only under no carryover. With rich switching, the honest
conclusion is often that DiD is not the right tool, and that sequential
methods for time-varying treatments (marginal structural models, panel
matching) fit better.

## Triple differences {#sec-ddd}

A triple difference adds a third dimension: within each group and period,
some units are eligible for the policy and others are not (Gruber 1994).
The estimand subtracts the DiD among ineligible units from the DiD among
eligible units. It identifies the ATT under a weaker assumption than
either DiD alone: the *difference* in untreated trends between eligible
and ineligible units must be the same in the treated and comparison
groups. Without covariates the estimate is the coefficient on the triple
interaction, or equivalently the difference of two `att_gt()` cells.
With covariates the shortcut fails: Ortiz-Villavicencio and Sant'Anna
(2025) show that neither the triple-interaction regression with controls
nor the difference of two conditional DiDs recovers the ATT, and they
give a doubly robust estimator built directly from the triple-difference
identification. The same holds for staggered adoption.

```{r ddd}
set.seed(16)
n3 <- 2000
d3 <- data.frame(id = rep(seq_len(n3), each = 2), time = rep(1:2, n3),
                 treat = rep(rbinom(n3, 1, 0.5), each = 2), elig = rep(rbinom(n3, 1, 0.5), each = 2))
d3$post <- as.integer(d3$time == 2)
d3$y <- rnorm(2 * n3) + 0.4 * d3$post * d3$treat + 0.3 * d3$post * d3$elig +
  1.5 * d3$post * d3$treat * d3$elig  # effect 1.5 on eligible units in the treated group
ddd <- feols(y ~ treat:post:elig + treat:post + post:elig | id + time, data = d3, cluster = ~id)
did_two <- function(e) coef(feols(y ~ treat:post | id + time, data = subset(d3, elig == e)))[["treat:post"]]
c(triple_interaction = coef(ddd)[["treat:post:elig"]], difference_of_dids = did_two(1) - did_two(0))
```

## Nonlinear and distributional DiD {#sec-nonlinear}

Section \@ref(sec-functional-form) showed that the scale is part of the
assumption. Three situations call for models beyond the linear DiD.

- **Counts and positive outcomes with zeros.** Logs are undefined at
  zero and the log DiD has no clean interpretation. Wooldridge (2023)
  proposes a Poisson quasi-likelihood with unit and period effects, whose
  identifying assumption is parallel trends in *ratios*: the proportional
  change in untreated means is the same across groups. The coefficient
  gives the proportional ATT as $\exp(\beta) - 1$, and `fixest::fepois()`
  estimates it with the same syntax as the linear model. This is the
  workhorse of the gravity literature (Section \@ref(sec-practice)).
- **Distributional effects.** Changes-in-changes (Athey and Imbens 2006)
  assumes that the untreated outcome distribution changes over time
  through one monotone map common to both groups, estimates the map from
  the comparison group's pre and post quantiles, and applies it to the
  treated group's pre-period distribution. It identifies the entire
  counterfactual distribution, so quantile effects follow, and it is
  invariant to monotone transformations by construction. Callaway and Li
  (2019) give quantile treatment effects on the treated under a
  copula-type assumption on the joint distribution of changes.

```{r poisson}
set.seed(17)
np <- 3000
pd <- data.frame(id = rep(seq_len(np), each = 2), time = rep(1:2, np), treat = rep(rbinom(np, 1, 0.5), each = 2))
pd$post <- as.integer(pd$time == 2)
mu <- exp(rep(rnorm(np, sd = 0.5), each = 2) + 0.2 * pd$post + 0.3 * pd$post * pd$treat)
pd$y <- rpois(2 * np, mu)
c(proportional_att_poisson = exp(coef(fepois(y ~ treat:post | id + time, data = pd))[["treat:post"]]) - 1,
  truth = exp(0.3) - 1,
  level_did_over_baseline = coef(feols(y ~ treat:post | id + time, data = pd))[["treat:post"]] /
    mean(pd$y[pd$treat == 1 & pd$post == 0]))
```

# Synthetic difference-in-differences {#sec-sdid}

## When the comparison group is not credible {#sec-sdid-problem}

DiD averages all untreated units into one comparison group. With a
handful of aggregate units, say one treated state and forty others, that
average has no reason to trend like the treated state. Synthetic control
(Abadie, Diamond, and Hainmueller 2010) replaces the average by a
weighted combination of untreated units chosen to reproduce the treated
unit's pre-treatment path. Synthetic DiD (Arkhangelsky et al. 2021)
combines the two ideas: it keeps the unit and period effects of DiD, so
levels need not match, and it adds unit weights that align the untreated
*trends* with the treated trend and period weights that make the
pre-treatment periods predictive of the post-treatment ones. The estimator
solves
$$
(\hat\tau, \hat\alpha, \hat\beta) = \arg\min \sum_i \sum_t \bigl(Y_{it} - \alpha_i - \beta_t - \tau D_{it}\bigr)^2 \hat\omega_i \hat\lambda_t ,
$$
a weighted two-way fixed effects regression. Plain DiD is the case of
uniform weights; synthetic control is the case of fitted unit weights,
no period weights, and no unit effects.

## Unit and time weights {#sec-sdid-weights}

The unit weights solve a regularized constrained regression of the
treated pre-treatment average on the untreated units' pre-treatment
paths, with an intercept, so only trends must match. The period weights
solve the analogous problem in the other direction, predicting untreated
post-treatment averages from their pre-treatment periods. The
regularization $\zeta$ is set from the noise in first differences, and
the weights are sparse in practice. `sdid_weights()` returns them, with a
cell-weight column for the regression step in `fixest`.

The design below has a latent factor: units load differently on a common
time trend, and treated units load more. Plain DiD compares treated units
to an average that trends differently; synthetic DiD reweights the
untreated units to match.

```{r sdid-sim}
set.seed(18)
n0 <- 30; n1 <- 5; t0 <- 10; t1 <- 5
nn <- n0 + n1; TT <- t0 + t1
f <- cumsum(rnorm(TT)) + seq_len(TT) / 2
load <- c(runif(n0, 0, 2), runif(n1, 1.5, 2.5))
Y <- outer(rnorm(nn), rep(1, TT)) + outer(load, f) + matrix(rnorm(nn * TT, sd = 0.3), nn)
Y[(n0 + 1):nn, (t0 + 1):TT] <- Y[(n0 + 1):nn, (t0 + 1):TT] + 2
blk <- data.frame(id = rep(seq_len(nn), each = TT), time = rep(seq_len(TT), nn), y = as.vector(t(Y)),
                  d = as.integer(rep(seq_len(nn) > n0, each = TT) & rep(seq_len(TT) > t0, nn)))
ests <- sapply(c("did", "sc", "difp", "sdid"), function(m)
  sdid_weights(blk, id = "id", time = "time", y = "y", d = "d", estimator = m)$estimate)
round(ests, 3)
sw <- sdid_weights(blk, id = "id", time = "time", y = "y", d = "d")
fit_w <- feols(y ~ d | id + time, data = merge(blk, sw$cell_weights, by = c("id", "time")), weights = ~weight)
c(closed_form = sw$estimate, weighted_fixest = coef(fit_w)[["d"]])
```

The true effect is 2. DiD is far off because the untreated average has a
flatter trend than the treated units. Synthetic control reduces the bias
but does not remove it, because without unit effects its weights must
match levels as well as trends. Synthetic DiD matches trends only and
recovers the effect, and its closed form equals the weighted regression.

## Inference {#sec-sdid-inference}

Three resampling schemes are available. The **placebo** scheme draws
untreated units, treats them as if treated, and re-estimates; it is the
choice when treated units are few, because it uses the untreated units'
variability. The **bootstrap** resamples units with replacement and needs
several treated units. The **jackknife** leaves one unit out. All three
re-solve the weights on each draw, so the standard error includes the
uncertainty from weight estimation.

```{r sdid-se}
sapply(c("placebo", "bootstrap", "jackknife"), function(m)
  sdid_se(sw, method = m, n_reps = 100, seed = 1)$std.error)
```

## Diagnostics {#sec-sdid-diagnostics}

```{r fig-sdid, fig.height = 3.8, fig.cap = "Treated average and the weighted untreated average; bars show the period weights."}
plot_sdid(sw)
```

Read three things from the fit. The pre-treatment paths should coincide
after the unit effects are removed; a gap there is a failure of the
weights, not evidence of an effect. The unit weights should not
concentrate on one or two donors, which makes the estimate fragile to
those units. And the period weights show which pre-treatment periods are
doing the work; weights on distant periods with different dynamics are a
warning. With staggered adoption, run the estimator cohort by cohort and
average the results with cohort weights. The companion synthetic control
note in this series covers the Abadie method, its placebo tests, and the
`synthdid` package in depth.

# Practice {#sec-practice}

## Replications and anchors {#sec-replications}

The replication notebooks for this section live in the notes folder and
are planned for the package's replication directory for Section 07; they
are not yet part of the package. Each anchors one part of the lecture.

| Paper | What it anchors |
|---|---|
| de Chaisemartin and D'Haultfoeuille (2020), newspapers and turnout; union wages | negative weights and the switchers-versus-stayers estimator (Section \@ref(sec-staggered-problems)) |
| Freyaldenhoven, Hansen, and Shapiro (2019); Roth (2022) | pre-trends, power, and the pre-test bias (Section \@ref(sec-assess)) |
| Roth and Sant'Anna (2023), functional form; Cengiz et al. (2019) minimum wage bins | scale dependence and the distributional test (Sections \@ref(sec-functional-form), \@ref(sec-ff-test)) |
| Elenev et al. (2024), staggered health policy with spillovers | staggered estimators and the interference assumption (Sections \@ref(sec-staggered-solutions), \@ref(sec-extensions)) |
| Nagengast and Yotov (2025), trade agreements in gravity models | nonlinear DiD with Poisson fixed effects (Section \@ref(sec-nonlinear)) |
| Arkhangelsky et al. (2021), California Proposition 99; Li (2024), forward DiD | synthetic DiD (Section \@ref(sec-sdid)) |
| Planned: Callaway and Sant'Anna (2021) minimum wage; Sant'Anna and Zhao (2020); Goodman-Bacon (2021); Rambachan and Roth (2023) | `att_gt()` validation, the three faces of covariates, the decomposition, sensitivity analysis |

## A DiD checklist {#sec-checklist}

Adapted from Sant'Anna's course.

1. Plot the treatment rollout: who is treated when, and who never.
2. Count units per cohort and per comparison group; small cohorts have
   noisy cells and small comparison groups have overlap problems.
3. Plot raw average outcomes by cohort over calendar time before any
   estimation.
4. State the comparison group and the parallel trends assumption in
   words: who decides treatment, what do they know, what selection is
   allowed. Date treatment at the announcement if units can anticipate.
5. Estimate the event study with `att_gt()` and no covariates, universal
   base period, uniform bands. Read the pre-treatment cells, then the
   post-treatment ones.
6. If unconditional parallel trends is implausible, add pre-treatment
   covariates with the doubly robust cells, and check overlap. Do not add
   them linearly to a TWFE regression.
7. Compare never-treated and not-yet-treated comparisons, Sun-Abraham,
   and imputation. Disagreement is information about the pre-period
   assumption.
8. Report the power of the pre-test (`pretrend_power()`) and a sensitivity
   analysis (HonestDiD) rather than a passed test.
9. Cluster at the level of treatment assignment; with few treated
   clusters, use `did_permutation_test()`.
10. If parallel trends is not credible with any comparison group,
    consider synthetic DiD, or a different design.

# Extensions {#sec-extensions}

- **Efficient DiD under parallel trends in all periods.** Chen,
  Sant'Anna, and Xie (2024) show that when the assumption holds before
  and after treatment, the model is overidentified, and derive the
  estimator that weights the pre-treatment periods optimally given the
  serial correlation of outcomes. It nests Callaway-Sant'Anna and the
  imputation estimator as special cases and dominates both.
- **Heterogeneous effects.** The doubly robust cell score is a pseudo
  outcome with conditional mean equal to the conditional ATT given
  covariates. Regressing it on a subset of covariates gives the best
  linear approximation of $CATT(g, t, Z)$ (Callaway, Chen, and Sant'Anna
  2024), and a meta-learner built on it estimates it flexibly (the DiD
  analogue of Section 08's DR-learner).
- **Selection and parallel trends.** Ghanem, Sant'Anna, and Wuthrich
  (2022) derive the selection mechanisms under which parallel trends
  holds, and Marx, Tamer, and Tang (2024) show that forward-looking
  adoption decisions generally violate it. Both give the vocabulary for
  step 4 of the checklist.
- **Compositional changes.** In repeated cross-sections the stationarity
  assumption can fail; Sant'Anna and Xu (2023) give estimators that allow
  the covariate distribution to change over time and a test for it.
- **Spillovers.** When treated units affect their neighbours, the
  no-interference assumption fails and the comparison group is
  contaminated. Elenev et al. (2024) and Butts (2023) estimate direct and
  spillover effects by defining exposure through distance to treated
  units.
- **Interactive fixed effects and matrix completion.** Parallel trends is
  the one-factor case of a model with unit loadings on common factors.
  Interactive fixed effects (Bai 2009; Xu 2017) and matrix completion
  (Athey et al. 2021) impute untreated outcomes under a low-rank factor
  structure; synthetic DiD is a weighted special case. Liu, Wang, and Xu
  (2024) unify these as counterfactual estimators with placebo and
  equivalence tests.
- **Dynamic treatment regimes.** With treatments chosen sequentially in
  response to past outcomes, the target becomes the effect of a
  treatment path, identified by the g-formula and estimated by sequential
  regression or dynamic DML rather than DiD (Section 08).

# Appendix {#sec-appendix}

## Influence functions for the two-by-two {#app-influence}

**Panel.** With $D_i = \mathbf{1}\{G_i = 2\}$, $\pi = P(D = 1)$, and
$\Delta Y_i = Y_{i2} - Y_{i1}$, the estimator is
$$
\hat\theta = \frac{\mathbb{E}_n[D\,\Delta Y]}{\mathbb{E}_n[D]} - \frac{\mathbb{E}_n[(1 - D)\,\Delta Y]}{\mathbb{E}_n[1 - D]} .
$$
Each ratio is a smooth function of two sample means. A first-order
expansion of $a_n / b_n$ around $(a, b)$ gives
$a_n / b_n - a / b = (a_n - a)/b - a (b_n - b)/b^2 + o_p(n^{-1/2})$.
Applied to the first ratio with $a = \mathbb{E}[D \Delta Y]$ and $b = \pi$,
$$
\frac{\mathbb{E}_n[D\,\Delta Y]}{\mathbb{E}_n[D]} - \mathbb{E}[\Delta Y \mid D = 1]
= \mathbb{E}_n\Bigl[\frac{D}{\pi}\bigl(\Delta Y - \mathbb{E}[\Delta Y \mid D = 1]\bigr)\Bigr] + o_p(n^{-1/2}),
$$
and likewise for the second ratio. Subtracting,
$\hat\theta - \theta = \mathbb{E}_n[\xi_i] + o_p(n^{-1/2})$ with
$$
\xi_i = \frac{D_i}{\pi}\bigl(\Delta Y_i - \mathbb{E}[\Delta Y \mid D = 1]\bigr) - \frac{1 - D_i}{1 - \pi}\bigl(\Delta Y_i - \mathbb{E}[\Delta Y \mid D = 0]\bigr).
$$
$\xi_i$ has mean zero and variance $\mathbb{E}[\xi^2] = \mathrm{Var}(\Delta Y \mid D = 1)/\pi + \mathrm{Var}(\Delta Y \mid D = 0)/(1 - \pi)$,
so $\sqrt n(\hat\theta - \theta) \to N(0, \mathbb{E}[\xi^2])$ by the
central limit theorem, and the plug-in $\hat\xi_i$ gives the standard
error $\sqrt{\sum_i \hat\xi_i^2}/n$.

**Repeated cross-sections.** With $T_i \in \{1, 2\}$ the period of
observation and four cells $(t, d)$, the estimator is the DiD of four
cell means, and the same expansion gives
$$
\xi_i = \xi_{i}^{(2,1)} - \xi_{i}^{(1,1)} - \xi_{i}^{(2,0)} + \xi_{i}^{(1,0)},
\qquad
\xi_{i}^{(t,d)} = \frac{\mathbf{1}\{T_i = t, D_i = d\}}{P(T = t, D = d)}\bigl(Y_i - \mathbb{E}[Y \mid T = t, D = d]\bigr).
$$

**Clusters.** If observations are sampled in clusters, replace
$\hat\xi_i$ by cluster sums $\hat\xi_c = \sum_{i \in c}\hat\xi_i$; the
variance is $\sum_c \hat\xi_c^2 / n^2$.

**Covariates.** With parametric first steps the influence function of
the regression, IPW, and doubly robust estimators includes correction
terms for the estimated logit and regression coefficients (Sant'Anna
and Zhao 2020, Appendix). `att_gt()` reports these exact influence
functions. With cross-fitted learners it reports the plain orthogonal
score, whose first-order corrections vanish by the argument of the next
subsection.

## Orthogonality and remainder of the DR-DiD score {#app-orthogonality}

Write the panel doubly robust score with $\pi = \mathbb{E}[D]$ and
nuisances $\eta = (p, m)$:
$$
\psi(W; \theta, \eta) = \frac{1}{\pi}\Bigl(D - \frac{p(X)(1 - D)}{1 - p(X)}\Bigr)\bigl(\Delta Y - m(X)\bigr) - \frac{D}{\pi}\,\theta .
$$
Two facts: $\mathbb{E}[D \mid X] = p_0(X)$, so
$\mathbb{E}[p_0(X)(1 - D)/(1 - p_0(X)) \mid X] = p_0(X)$; and
$\mathbb{E}[\Delta Y - m_0(X) \mid D = 0, X] = 0$.

*Direction in $m$.* Replacing $m_0$ by $m_0 + t\Delta_m$ changes the
moment by $-t\,\mathbb{E}[(D - p_0(1 - D)/(1 - p_0))\Delta_m(X)]/\pi$.
Conditioning on $X$, the bracket has mean $p_0 - p_0 = 0$. The derivative
is zero.

*Direction in $p$.* Only the weight $p(1 - D)/(1 - p)$ depends on $p$,
with derivative $(1 - D)\Delta_p(X)/(1 - p_0(X))^2$. It multiplies
$\Delta Y - m_0(X)$ on untreated units, whose conditional mean is zero.
The derivative is zero. The score is Neyman orthogonal.

*Exact remainder.* With fixed $\hat p, \hat m$, write
$\Delta_p = \hat p - p_0$ and $\Delta_m = \hat m - m_0$. Conditioning on
$X$ and using the two facts, every term involving $\Delta Y - m_0(X)$ on
untreated units vanishes, and what remains is a product of the two
nuisance errors:
$$
\mathbb{E}[\psi(W; \theta_0, \hat\eta)] = \frac{1}{\pi}\,\mathbb{E}\Bigl[\frac{\hat p(X) - p_0(X)}{1 - \hat p(X)}\,\bigl(m_0(X) - \hat m(X)\bigr)\Bigr].
$$
It is zero when either nuisance is correct (double robustness), and it
is $o(n^{-1/2})$ when the product of the $L^2$ errors is, which is the
rate condition of Section 04. The Hajek normalization used in
`att_gt()` divides the control term by its own mean instead of $\pi$;
the difference is $O_p(n^{-1/2})$ times the remainder and does not affect
the argument.

## The Bacon decomposition {#app-bacon}

> **Theorem (Goodman-Bacon 2021).** In a balanced panel with $K$ timing
> groups and possibly a never-treated group $U$, the static TWFE
> coefficient satisfies
> $\hat\beta = \sum_{k} s_{kU}\hat\beta_{kU} + \sum_{k < l}\bigl(s_{kl}^{k}\hat\beta_{kl}^{k} + s_{kl}^{l}\hat\beta_{kl}^{l}\bigr)$,
> where each $\hat\beta$ is a two-by-two DiD estimate and the weights are
> non-negative and sum to one. With $n_k$ the population share of group
> $k$, $\bar D_k$ the share of periods it spends treated, and $\hat V^D$
> the variance of the doubly demeaned treatment indicator,
> $s_{kU} = (n_k + n_U)^2\, n_{kU}(1 - n_{kU})\,\bar D_k(1 - \bar D_k)/\hat V^D$
> with $n_{kU} = n_k/(n_k + n_U)$, and the two timing weights are given in
> the paper's Theorem 1.

*Sketch.* By the Frisch-Waugh-Lovell theorem, $\hat\beta = \mathrm{Cov}(\tilde D, Y)/\mathrm{Var}(\tilde D)$
where $\tilde D$ is $D$ with unit and period means removed. Restricting
the sample to any pair of groups and applying the same formula gives that
pair's two-by-two estimate. The full-sample covariance is a sum over
pairs of the within-pair covariances, because $\tilde D$ is piecewise
constant across groups and periods; dividing each by $\mathrm{Var}(\tilde D)$
yields the weights, which are each pair's share of the treatment
variance. `bacon_decomp()` computes the pieces on group-by-period means,
and its `check` component verifies the identity numerically.

## Identification of ATT(g,t) {#app-identification}

**Never-treated comparison.** Fix $g$ and $t \ge g$. Parallel trends for
each period $s = g, \dots, t$,
$\mathbb{E}[Y_s(\infty) - Y_{s-1}(\infty) \mid G = g] = \mathbb{E}[Y_s(\infty) - Y_{s-1}(\infty) \mid G = \infty]$,
summed over $s$ gives
$\mathbb{E}[Y_t(\infty) - Y_{g-1}(\infty) \mid G = g] = \mathbb{E}[Y_t(\infty) - Y_{g-1}(\infty) \mid G = \infty]$.
No anticipation replaces $Y_{g-1}(\infty)$ by the observed $Y_{g-1}$ for
group $g$, and no interference replaces the remaining potential outcomes
by observed ones. Hence
$ATT(g, t) = \mathbb{E}[Y_t - Y_{g-1} \mid G = g] - \mathbb{E}[Y_t - Y_{g-1} \mid G = \infty]$.

**Not-yet-treated comparison.** Replace $G = \infty$ by the event
$D_t = 0, G \ne g$ in every step; the assumption must then hold for that
changing set, which is the not-yet-treated version of Assumption 3.

**Covariates.** With conditional parallel trends and overlap, the same
argument conditional on $X$ identifies $ATT(g, t \mid X)$, and averaging
over the distribution of $X$ in group $g$ gives the outcome regression,
IPW, and doubly robust estimands of Section \@ref(sec-three-faces) with
$\Delta Y$ replaced by $Y_t - Y_{g-1}$ and $D$ by $\mathbf{1}\{G = g\}$
within the cell sample (Callaway and Sant'Anna 2021, Theorem 1).

## Multiplier bootstrap and simultaneous bands {#app-bootstrap}

Let $\hat\theta = (\hat\theta_1, \dots, \hat\theta_K)$ be the cell
estimates with $\sqrt n(\hat\theta - \theta) = n^{-1/2}\sum_i \xi_i + o_p(1)$
jointly, and let $V_1, \dots, V_n$ be i.i.d. draws, independent of the
data, with mean zero and variance one (Mammen's two-point distribution or
Rademacher). Define $\hat\theta^* = \hat\theta + n^{-1}\sum_i V_i \hat\xi_i$.
By the conditional multiplier central limit theorem,
$\sqrt n(\hat\theta^* - \hat\theta)$ converges conditionally on the data
to the same Gaussian limit as $\sqrt n(\hat\theta - \theta)$
(Callaway and Sant'Anna 2021, Theorem 3). Therefore:

- the bootstrap standard error, taken as the interquartile range of
  $\hat\theta_k^* - \hat\theta_k$ divided by that of a standard normal, is
  consistent and robust to outlying draws;
- with $\hat c_{1-\alpha}$ the $(1 - \alpha)$ quantile across draws of
  $\max_k |\hat\theta_k^* - \hat\theta_k| / \hat{se}_k$, the band
  $\hat\theta_k \pm \hat c_{1-\alpha}\,\hat{se}_k$ covers all $K$ cells
  jointly with asymptotic probability $1 - \alpha$.

With clusters, sum $\hat\xi_i$ within clusters and draw one $V$ per
cluster. Aggregations inherit the result because their influence
functions are linear combinations of the cell influence functions plus
the estimation error of the aggregation weights, which is itself a
mean-zero average.

## References

- Abadie, Alberto. 2005. "Semiparametric Difference-in-Differences Estimators." *Review of Economic Studies*.
- Abadie, Alberto, Alexis Diamond, and Jens Hainmueller. 2010. "Synthetic Control Methods for Comparative Case Studies." *Journal of the American Statistical Association*.
- Arkhangelsky, Dmitry, Susan Athey, David A. Hirshberg, Guido W. Imbens, and Stefan Wager. 2021. "Synthetic Difference-in-Differences." *American Economic Review*.
- Athey, Susan, and Guido W. Imbens. 2006. "Identification and Inference in Nonlinear Difference-in-Differences Models." *Econometrica*.
- Athey, Susan, and Guido W. Imbens. 2022. "Design-Based Analysis in Difference-in-Differences Settings with Staggered Adoption." *Journal of Econometrics*.
- Athey, Susan, Mohsen Bayati, Nikolay Doudchenko, Guido Imbens, and Khashayar Khosravi. 2021. "Matrix Completion Methods for Causal Panel Data Models." *Journal of the American Statistical Association*.
- Borusyak, Kirill, Xavier Jaravel, and Jann Spiess. 2024. "Revisiting Event-Study Designs: Robust and Efficient Estimation." *Review of Economic Studies*.
- Butts, Kyle. 2023. "Difference-in-Differences Estimation with Spatial Spillovers." Working paper.
- Callaway, Brantly, Andrew Goodman-Bacon, and Pedro H. C. Sant'Anna. 2024. "Difference-in-Differences with a Continuous Treatment." Working paper, arXiv:2107.02637.
- Callaway, Brantly, and Tong Li. 2019. "Quantile Treatment Effects in Difference in Differences Models with Panel Data." *Quantitative Economics*.
- Callaway, Brantly, and Pedro H. C. Sant'Anna. 2021. "Difference-in-Differences with Multiple Time Periods." *Journal of Econometrics*.
- Cengiz, Doruk, Arindrajit Dube, Attila Lindner, and Ben Zipperer. 2019. "The Effect of Minimum Wages on Low-Wage Jobs." *Quarterly Journal of Economics*.
- Chang, Neng-Chieh. 2020. "Double/Debiased Machine Learning for Difference-in-Differences Models." *The Econometrics Journal*.
- Chen, Xiaohong, Pedro H. C. Sant'Anna, and Haitian Xie. 2024. "Efficient Difference-in-Differences and Event Study Estimators." Working paper.
- Chernozhukov, Victor, Christian Hansen, Nathan Kallus, Martin Spindler, and Vasilis Syrgkanis. 2026. *Applied Causal Inference Powered by ML and AI*, chapter 16. Online, version 0.1.2.
- de Chaisemartin, Clement, and Xavier D'Haultfoeuille. 2020. "Two-Way Fixed Effects Estimators with Heterogeneous Treatment Effects." *American Economic Review*.
- de Chaisemartin, Clement, and Xavier D'Haultfoeuille. 2024. "Difference-in-Differences Estimators of Intertemporal Treatment Effects." *Review of Economics and Statistics*.
- Dube, Arindrajit, Daniele Girardi, Oscar Jorda, and Alan M. Taylor. 2023. "A Local Projections Approach to Difference-in-Differences Event Studies." NBER Working Paper 31184.
- Freyaldenhoven, Simon, Christian Hansen, and Jesse M. Shapiro. 2019. "Pre-Event Trends in the Panel Event-Study Design." *American Economic Review*.
- Gardner, John. 2022. "Two-Stage Differences in Differences." Working paper.
- Ghanem, Dalia, Pedro H. C. Sant'Anna, and Kaspar Wuthrich. 2022. "Selection and Parallel Trends." Working paper, arXiv:2203.09001.
- Goodman-Bacon, Andrew. 2021. "Difference-in-Differences with Variation in Treatment Timing." *Journal of Econometrics*.
- Gruber, Jonathan. 1994. "The Incidence of Mandated Maternity Benefits." *American Economic Review*.
- Heckman, James J., Hidehiko Ichimura, and Petra E. Todd. 1997. "Matching as an Econometric Evaluation Estimator." *Review of Economic Studies*.
- Liu, Licheng, Ye Wang, and Yiqing Xu. 2024. "A Practical Guide to Counterfactual Estimators for Causal Inference with Time-Series Cross-Sectional Data." *American Journal of Political Science*.
- Marx, Philip, Elie Tamer, and Xun Tang. 2024. "Parallel Trends and Dynamic Choices." *Journal of Political Economy: Microeconomics*.
- Ortiz-Villavicencio, Marcelo, and Pedro H. C. Sant'Anna. 2025. "Better Understanding Triple Differences Estimators." Working paper.
- Rambachan, Ashesh, and Jonathan Roth. 2023. "A More Credible Approach to Parallel Trends." *Review of Economic Studies*.
- Roth, Jonathan. 2022. "Pretest with Caution: Event-Study Estimates after Testing for Parallel Trends." *American Economic Review: Insights*.
- Roth, Jonathan, and Pedro H. C. Sant'Anna. 2023. "When Is Parallel Trends Sensitive to Functional Form?" *Econometrica*.
- Roth, Jonathan, and Pedro H. C. Sant'Anna. 2023. "Efficient Estimation for Staggered Rollout Designs." *Journal of Political Economy: Microeconomics*.
- Roth, Jonathan, Pedro H. C. Sant'Anna, Alyssa Bilinski, and John Poe. 2023. "What's Trending in Difference-in-Differences? A Synthesis of the Recent Econometrics Literature." *Journal of Econometrics*.
- Sant'Anna, Pedro H. C. 2025. *Causal Inference Using Difference-in-Differences*. Lecture slides, Emory University.
- Sant'Anna, Pedro H. C., and Qi Xu. 2023. "Difference-in-Differences with Compositional Changes." Working paper, arXiv:2304.13925.
- Sant'Anna, Pedro H. C., and Jun Zhao. 2020. "Doubly Robust Difference-in-Differences Estimators." *Journal of Econometrics*.
- Sun, Liyang, and Sarah Abraham. 2021. "Estimating Dynamic Treatment Effects in Event Studies with Heterogeneous Treatment Effects." *Journal of Econometrics*.
- Wooldridge, Jeffrey M. 2021. "Two-Way Fixed Effects, the Two-Way Mundlak Regression, and Difference-in-Differences Estimators." Working paper.
- Wooldridge, Jeffrey M. 2023. "Simple Approaches to Nonlinear Difference-in-Differences with Panel Data." *The Econometrics Journal*.
