---
title: "Causal Inference in Economics and Marketing"
subtitle: "Section 08: Heterogeneous Treatment Effects and Policy Learning"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, fig.align = "center", fig.pos = "H",
  out.width = "95%", collapse = TRUE, comment = "#>"
)
# Code and output in footnotesize.
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
# This lecture lives in inst/lectures/08_hte/ of the package source. When
# knitted from the source tree it loads the development version of the
# package; when the package is installed it uses the installed helpers.
helper_rel <- file.path("vignette_helpers", "08_hte", "hte_lecture_helpers.R")
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
  helper_path <- file.path(pkg_root, "inst", helper_rel)
} else {
  library(causalmetrics)
  helper_path <- system.file(helper_rel, package = "causalmetrics")
}
library(dplyr)
library(ggplot2)
library(kableExtra)
library(mlr3)
library(mlr3learners)
source(helper_path)
helper_md5 <- tools::md5sum(helper_path)
n_sims_learners <- 60
n_sims_policy <- 40
fmt <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
run_grf <- requireNamespace("grf", quietly = TRUE)
run_policytree <- requireNamespace("policytree", quietly = TRUE)
x5 <- paste0("x", 1:5)
```

# From "does it work" to "for whom" and "whom to treat" {#sec-problem}

## Three questions

Sections 02 to 07 answered one question: what is the average effect of a
treatment? The average treatment effect (ATE), the ATT, and the ATT(g,t) of
the last section are all averages. They inform a uniform decision: roll the
policy out to everyone, or to no one.

Two further questions sit behind most applied work.

- **For whom does it work?** The conditional average treatment effect
  (CATE), $\tau(x) = E[Y(1) - Y(0) \mid X = x]$, describes how the effect
  varies with observed characteristics.
- **Whom should we treat?** A policy $\pi(x) \in \{0, 1\}$ assigns treatment
  from characteristics. Its value is the average gain it produces.

A positive ATE can hide a subgroup that is harmed. A null ATE can hide a
subgroup that responds strongly. Marketing lives on this: a discount that is
worthless on average may pay for itself among price-sensitive customers.
Development economics lives on it too: a cash transfer may be worth most to
households that are not the poorest.

## Why the CATE is a harder object

Every target so far was a number. The CATE is a function. That changes what
we can hope for.

- Estimating a function is a prediction problem. Prediction rates are slower
  than $n^{-1/2}$ and confidence intervals do not come for free.
- Even with both potential outcomes observed, estimating $\tau(x)$ would be
  as hard as estimating the regression of $Y(1) - Y(0)$ on $X$.
- Genovese and Wasserman (2008) show that adaptive confidence bands for a
  curve do not exist in general. Bands must target a simpler surrogate, or
  restrict the dimension of $X$.

The response is to relax the goal, and the lecture is organized by how much
we relax it (Chernozhukov et al. 2026, ch. 14).

| Goal | What we get | Section |
|---|---|---|
| A low-dimensional summary: best linear predictor or group effects | Root-$n$ inference, uniform bands | \@ref(sec-summaries) |
| Confidence intervals at a point with forests | Valid in low dimensions, brittle otherwise | \@ref(sec-forests) |
| A flexible CATE model with small mean squared error | No intervals; out-of-sample scores and tests | \@ref(sec-learners), \@ref(sec-selection) |
| Tests that a model has found heterogeneity | Calibration, targeting curves | \@ref(sec-validation) |
| A good policy, without caring about the CATE's magnitude | Regret bounds, honest values | \@ref(sec-policy) |

## Identification

Conditional exogeneity, $D \perp (Y(0), Y(1)) \mid Z$, identifies the CATE
for any $X \subseteq Z$:
$$
\tau(X) = E\big[\, g(1, Z) - g(0, Z) \mid X \,\big], \qquad g(d, z) = E[Y \mid D = d, Z = z].
$$
Two remarks matter later.

- $X$ may be a small subset of the controls $Z$. Confounding is handled by
  $Z$; heterogeneity is described by $X$. The high-dimensional CATE
  $\delta(Z) = E[Y(1) - Y(0) \mid Z]$ and its coarsening $\tau(X) = E[\delta(Z) \mid X]$
  are different objects.
- On $Z$, the effect among the treated equals the effect among the
  controls: $E[Y(1) - Y(0) \mid Z, D = 1] = E[Y(1) - Y(0) \mid Z, D = 0] = \delta(Z)$.
  This fails for a subset $X$, because the distribution of $Z$ given $X$
  differs across arms. The X-learner of Section \@ref(sec-x-learner) is built
  on the first fact.

## The first CATE model: an interaction regression {#sec-interaction}

The simplest CATE model is the regression with interactions,
$Y = \beta_0 + \beta_1 D + \beta_2' X + D\, X'\beta_3 + e$, which gives
$\hat\tau(x) = \hat\beta_1 + x'\hat\beta_3$ (Facure 2023, ch. 18). We use a
simulated design throughout. Think of $D$ as a marketing nudge, $Y$ as
spending, and $x_1, \dots, x_5$ as customer characteristics. The true CATE
is $1 + x_1 + 0.5 (x_2^2 - 1)$; assignment depends on $x_1$ and $x_2$; the
baseline is nonlinear in $x_2$.

```{r running-data}
dat <- sim_hte(n = 4000, dgp = "smooth", seed = 1)
train <- dat[1:2000, ]
score_set <- dat[2001:3000, ]
test <- dat[3001:4000, ]
fit_int <- lm(y ~ d * (x1 + x2 + x3 + x4 + x5), data = train)
tau_int <- coef(fit_int)["d"] + as.matrix(test[, x5]) %*% coef(fit_int)[paste0("d:", x5)]
c(rmse = sqrt(mean((tau_int - test$tau_true)^2)), cor = cor(tau_int, test$tau_true))
```

The regression catches the linear part of the heterogeneity in $x_1$ and
misses the curvature in $x_2$. It also has no protection against
misspecification of the baseline: the terms in $X$ must absorb confounding
and describe heterogeneity at the same time. Figure \@ref(fig:interaction)
shows both facts.

```{r interaction, echo = FALSE, fig.cap = "Interaction-regression CATE against the truth on the test sample. Left: along x1 the linear model tracks the truth. Right: along x2 it cannot bend."}
df <- tibble(x1 = test$x1, x2 = test$x2, truth = test$tau_true, model = as.numeric(tau_int))
p1 <- ggplot(df, aes(x1)) + geom_point(aes(y = truth), colour = "grey60", size = 0.6) +
  geom_point(aes(y = model), colour = hte_colors[["DR-learner"]], size = 0.6) +
  labs(y = "CATE", subtitle = "Grey: truth; blue: interaction regression") + hte_theme()
p2 <- ggplot(df, aes(x2)) + geom_point(aes(y = truth), colour = "grey60", size = 0.6) +
  geom_point(aes(y = model), colour = hte_colors[["DR-learner"]], size = 0.6) +
  labs(y = NULL, subtitle = " ") + hte_theme()
patchwork::wrap_plots(p1, p2)
```

The rest of the lecture separates the two jobs. Confounding is removed
once, by cross-fitted nuisances. Heterogeneity is then a regression of a
single constructed variable on $X$.

## Roadmap

Section \@ref(sec-score) builds that variable, the doubly robust
pseudo-outcome, and shows why its conditional mean is the CATE. Section
\@ref(sec-learners) turns any regression method into a CATE model and
compares the meta-learners in three designs. Section \@ref(sec-summaries)
gives valid inference on summaries and on forests. Sections
\@ref(sec-selection) and \@ref(sec-validation) select and validate a model
out of sample. Section \@ref(sec-policy) learns and evaluates policies.
Section \@ref(sec-traps) collects traps, Section \@ref(sec-practice) the
replications and a checklist. The appendix holds the formal statements.

# One signal, many uses: the doubly robust pseudo-outcome {#sec-score}

## Definition and the key property

Let $p(z) = P(D = 1 \mid Z = z)$ and $g(d, z) = E[Y \mid D = d, Z = z]$,
with $\eta = (p, g)$. The doubly robust pseudo-outcome is
$$
Y(\eta) = g(1, Z) - g(0, Z) + H(p)\,\big(Y - g(D, Z)\big), \qquad
H(p) = \frac{D}{p(Z)} - \frac{1 - D}{1 - p(Z)}.
$$
It is the summand of the AIPW estimator of Section 04. Its mean is the ATE.
Its conditional mean is the CATE:
$$
E[Y(\eta_0) \mid X] = \tau(X).
$$
*Skeleton.* Condition on $Z$ first. Given $Z$, $E[H(p_0)(Y - g_0(D, Z)) \mid Z] = 0$
because $E[D(Y - g_0(1,Z)) \mid Z] = p_0(Z) \cdot 0$ and likewise for the
control term. What remains is $g_0(1, Z) - g_0(0, Z) = \delta(Z)$. Then
average over $Z$ given $X$.

The property that makes everything else work is **conditional Neyman
orthogonality**: the derivative of $E[Y(\eta) \mid X]$ with respect to
$\eta$ at $\eta_0$ is zero. Errors in $\hat p$ and $\hat g$ enter the
conditional mean only through their product,
$$
E[Y(\eta_0) - Y(\hat\eta) \mid X] = E\big[(H(p_0) - H(\hat p))(g_0(D, Z) - \hat g(D, Z)) \mid X\big],
$$
so the pseudo-outcome is as good as the true one whenever the product of the
two nuisance errors is small. Appendix \@ref(app-orthogonality) proves the
identity in three lines.

## Cross-fitting the nuisances

As in Section 04, the nuisances are cross-fitted: $\hat p$ and $\hat g$ are
trained on folds that exclude the observation they predict. The final
regression of $Y(\hat\eta)$ on $X$ may then use all rows (Chernozhukov et
al. 2026, Remark 15.1.1). Overfitting in the first stage cannot leak into
the second, because the label of row $i$ never saw row $i$.

## Code: building the scores

`dr_scores()` builds the vector once. Every later function consumes the
returned object.

```{r scores}
sc_train <- dr_scores(train, y = "y", d = "d", x = x5, seed = 1)
sc_train
```

With linear nuisances the outcome model is misspecified (the baseline is
quadratic in $x_2$) while the logistic propensity model is correctly
specified. Random forests fit both nuisances flexibly. In a simulation we
can also plug in the true nuisances, which no analyst has. The regression
of the score on the true CATE, whose slope should be one, summarizes how
close each version gets.

```{r scores-three}
rf_first <- hte_forest(min_node = 20, trees = 300)
rf_final <- hte_forest(min_node = 100, trees = 300)
sc_train_rf <- dr_scores(train, "y", "d", x5, learner_mu = rf_first, seed = 1)
sc_train_oracle <- dr_scores(train, "y", "d", p_hat = "p_true", mu0_hat = "mu0_true", mu1_hat = "mu1_true")
slope_tab <- sapply(list(linear = sc_train, forest = sc_train_rf, true_nuisances = sc_train_oracle), function(s) {
  b <- cate_blp(s, ~ tau_true, uniform = FALSE)$coefficients
  c(slope_on_true_cate = b$estimate[2], hc1_se = b$std.error[2], sd_score = sd(s$score))
})
round(slope_tab, 3)
```

The slope moves toward one and the score's standard deviation falls as
the nuisances improve. The gap that remains with estimated nuisances is
the product term above: small on average, but concentrated where the
outcome error and the propensity error are both large, which here is the
tail of $x_2$. Figure \@ref(fig:score-mean) checks the key property with
the forest nuisances, which the rest of the lecture uses: a running mean of
the score along $x_1$ follows the truth.

```{r score-mean, echo = FALSE, fig.cap = "The conditional mean of the pseudo-outcome is the CATE. Left: the score against the true CATE, with the OLS line (slope reported). Right: a binned mean of the score along x1 (points) against the true CATE averaged in the same bins (line)."}
slope <- slope_tab["slope_on_true_cate", "forest"]
df <- tibble(score = sc_train_rf$score, truth = train$tau_true, x1 = train$x1)
p1 <- ggplot(df, aes(truth, score)) + geom_point(alpha = 0.25, size = 0.6) +
  geom_smooth(method = "lm", se = FALSE, colour = hte_colors[["DR-learner"]]) +
  coord_cartesian(ylim = c(-15, 20)) +
  labs(x = "True CATE", y = "Pseudo-outcome", subtitle = paste0("OLS slope = ", fmt(slope))) + hte_theme()
bins <- df %>% mutate(bin = cut(x1, breaks = quantile(x1, 0:20 / 20), include.lowest = TRUE)) %>%
  group_by(bin) %>% summarise(x1 = mean(x1), score = mean(score), truth = mean(truth), .groups = "drop")
p2 <- ggplot(bins, aes(x1)) + geom_line(aes(y = truth), colour = "black") +
  geom_point(aes(y = score), colour = hte_colors[["DR-learner"]], size = 2) +
  labs(y = "Mean in bin", subtitle = "Binned score (points) and true CATE (line)") + hte_theme()
patchwork::wrap_plots(p1, p2)
```

The score is noisy: an individual value divides a residual by a propensity.
That noise is the price of unbiasedness at the individual level, and it is
why the rest of the lecture never looks at one score alone. It averages
them, regresses them, or classifies their sign.

## Two alternatives and why they lose

`dr_scores(type = "ipw")` returns $Y H(\hat p)$, the inverse-propensity
signal, whose conditional mean is also the CATE. `type = "reg"` returns
$\hat g(1, Z) - \hat g(0, Z)$, the plug-in signal. The first is unbiased only
if the propensity is right and has much larger variance; the second is
unbiased only if the outcome model is right and has no variance protection.
The doubly robust signal combines them.

```{r score-types}
sc_ipw <- dr_scores(train, "y", "d", x5, type = "ipw", seed = 1)
sc_reg <- dr_scores(train, "y", "d", x5, type = "reg", learner_mu = rf_first, seed = 1)
type_tab <- sapply(list(dr = sc_train_rf, ipw = sc_ipw, reg = sc_reg), function(s) {
  b <- cate_blp(s, ~ tau_true, uniform = FALSE)$coefficients
  c(slope_on_true_cate = b$estimate[2], hc1_se = b$std.error[2], sd_score = sd(s$score))
})
round(type_tab, 3)
```

The plug-in signal has the smallest variance and a slope below one: a
regularized outcome model shrinks the heterogeneity it cannot represent.
The inverse-propensity signal needs only the propensity, which is right
here, but its standard deviation is
`r fmt(type_tab["sd_score", "ipw"] / type_tab["sd_score", "dr"], 1)`
times that of the doubly robust signal. The doubly robust signal keeps
the slope of the inverse-propensity signal at a fraction of its variance.

## Connection to Section 04

The mean of the score is the AIPW estimate of the ATE, with the same
standard error as `est_aipw()`:

```{r score-ate}
tidy(sc_train_rf)
```

Everything below is a regression of the score on something: on a
dictionary of covariate functions (Section \@ref(sec-summaries)), on a
machine learner (Section \@ref(sec-learners)), on a candidate model's own
predictions (Section \@ref(sec-validation)), or on the sign of a policy
(Section \@ref(sec-policy)).

# Meta-learners: turning a regression method into a CATE model {#sec-learners}

## The idea and a notation

A meta-learner decomposes CATE estimation into regression problems and hands
each to an off-the-shelf method (Künzel et al. 2019). Write
$O_H(\{X_i, L_i, W_i\})$ for a regression oracle: it takes covariates,
labels, and optional weights and returns an estimate of the weighted
best predictor $\arg\min_{h \in H} E[W (L - h(X))^2]$ (Chernozhukov et al.
2026, Section 15.1). Every meta-learner below ends with one call to this
oracle on the heterogeneity variables $X$. The learners differ in the label
$L_i$ they construct, and in how they construct it. All first stages are
cross-fitted; the final call uses all rows.

## S- and T-learners {#sec-s-t}

The **S-learner** fits one model of $Y$ on $(D, Z)$ and sets
$L_i = \hat g(1, Z_i) - \hat g(0, Z_i)$. The **T-learner** fits one model
per arm, $\hat g_1$ on the treated and $\hat g_0$ on the controls, and sets
$L_i = \hat g_1(Z_i) - \hat g_0(Z_i)$. Both converge to the best
approximation of the CATE in the final oracle's class if the outcome models
are consistent. Both fail in characteristic ways.

- The S-learner regularizes the treatment like any other covariate. When
  the effect is small relative to the baseline variation, a penalized or
  tree-based learner shrinks it toward zero, or drops $D$ entirely.
- The T-learner cannot share information across arms. When one arm is
  small, its model is heavily regularized while the other is flexible, and
  the difference of two mismatched fits produces spurious heterogeneity.

The second failure is Figure \@ref(fig:t-failure), on the first design of
Chernozhukov et al. (2026, Example 15.1.1): 5 percent treated, a constant
effect of 0.5, a baseline with a jump on $[0.6, 0.8]$, and noise of 0.05.

```{r t-failure, echo = FALSE, fig.cap = "Design 1 (5 percent treated, constant effect, jump in the baseline), one sample of n = 500. Left: the cross-fitted outcome models of the T-learner; the control model finds the jump, the treated model, fitted on about 25 points, finds it only noisily. Right: the resulting CATE estimates of the T- and X-learners against the truth."}
dgp1 <- sim_hte(500, dgp = "simple_cate", seed = 4)
fl1 <- fit_all_learners(dgp1, "x1", newdata = dgp1, learner = hte_forest(min_node = 5, trees = 200),
                        learner_final = hte_forest(min_node = 50, trees = 200))
df1 <- tibble(x1 = dgp1$x1, d = dgp1$d, y = dgp1$y,
              mu0 = fl1$scores$nuisance$mu0, mu1 = fl1$scores$nuisance$mu1) %>% arrange(x1)
p1 <- ggplot(df1, aes(x1)) +
  geom_point(aes(y = y, colour = factor(d, labels = c("control", "treated"))), size = 0.7, alpha = 0.6) +
  geom_line(aes(y = mu0), colour = "black", linewidth = 0.8) +
  geom_line(aes(y = mu1), colour = hte_colors[["T-learner"]], linewidth = 0.8) +
  scale_colour_manual(values = c(control = "grey60", treated = hte_colors[["T-learner"]]), name = NULL) +
  labs(y = "Outcome", subtitle = "Outcome models: control (black), treated (orange)") + hte_theme()
p2 <- plot_learner_fits(dgp1, fl1$preds, c("T-learner", "X-learner")) + labs(subtitle = "CATE: black = truth")
patchwork::wrap_plots(p1, p2)
```

## X-learner {#sec-x-learner}

The X-learner (Künzel et al. 2019) uses the fact that the effect is the
same among treated and controls once we condition on $Z$. Each arm can
therefore estimate the CATE with only *one* imputed counterfactual:
$$
\delta^T(Z) = E[Y - g_0(Z) \mid Z, D = 1], \qquad
\delta^C(Z) = E[g_1(Z) - Y \mid Z, D = 0].
$$
The treated version needs only the control model $g_0$, which is estimated
on the large arm; the control version needs only $g_1$. The two estimates
are blended with the propensity, $\hat\delta(Z) = (1 - \hat p)\hat\delta^T + \hat p\,\hat\delta^C$,
so that where treatment is rare, the version that uses the well-estimated
control model gets the weight. The label is $L_i = \hat\delta(Z_i)$. The
right panel of Figure \@ref(fig:t-failure) shows the repair.

Two refinements come from the covariate-shift view: $\hat\delta^T$ is
trained on the treated but evaluated everywhere, so training it with weights
$(1 - \hat p)^2 / \hat p$ (and $\hat\delta^C$ with $\hat p^2 / (1 - \hat p)$)
targets the population where it will be used (`adapt = TRUE`, the
domain-adapted X-learner). And the X-learner's outcome models can feed the
DR-learner below in place of the T-learner's.

## DR-learner {#sec-dr-learner}

The **DR-learner** regresses the pseudo-outcome of Section \@ref(sec-score)
on $X$: $L_i = Y_i(\hat\eta)$ (Kennedy 2023; Foster and Syrgkanis 2023).
Its error bound is
$$
E\big[(\hat\tau - \tau^*)^2\big] \lesssim r_n^2 + \mathrm{Err}(\hat g)^2\, \mathrm{Err}(H(\hat p))^2,
$$
where $r_n^2$ is the rate the oracle would achieve with the true label and
$\tau^*$ is the best approximation in the oracle's class. Only the product
of the nuisance errors enters: this is double robustness for a function
(Appendix \@ref(app-learner-bounds) sketches the argument). The price is
variance. The label divides a residual by $\hat p$ or $1 - \hat p$, and when
overlap is weak in parts of the covariate space the label takes extreme
values there.

## R-learner {#sec-r-learner}

The **R-learner** (Nie and Wager 2021) starts from Robinson's
decomposition. If the CATE depends on $X$ only, then
$$
Y - E[Y \mid Z] = \tau(X)\,(D - E[D \mid Z]) + \epsilon, \qquad E[\epsilon \mid D, Z] = 0,
$$
so $\tau$ minimizes $E[(\tilde Y - \tau(X)\tilde D)^2]$ with $\tilde Y = Y - l(Z)$,
$\tilde D = D - p(Z)$. Rewriting the loss as
$E[\tilde D^2 (\tilde Y / \tilde D - \tau(X))^2]$ makes it a weighted
regression: label $L_i = \tilde Y_i / \tilde D_i$, weight
$W_i = \tilde D_i^2$. The final oracle must accept weights. The error bound
is
$$
E\big[(\hat\tau - \tau_0)^2\big] \lesssim r_n^2 + \mathrm{Err}(\hat p)^4 + \mathrm{Err}(\hat p)^2 \mathrm{Err}(\hat l)^2,
$$
so the propensity matters more than in the DR-learner, but nothing is
divided by it.

What does the R-learner estimate when the CATE depends on more than $X$,
or when the oracle's class is misspecified? Not the best approximation of
$\tau_0$, but an overlap-weighted one:
$$
E\big[(\tilde Y - \tau(X)\tilde D)^2\big] = E\big[(\delta(Z) - \tau(X))^2 \mathrm{Var}(D \mid Z)\big] + E[\epsilon^2].
$$
*Skeleton.* Substitute $\tilde Y = \delta(Z)\tilde D + \epsilon$, expand, use
$E[\epsilon \mid D, Z] = 0$ to kill the cross term, and
$E[\tilde D^2 \mid Z] = \mathrm{Var}(D \mid Z)$. Regions where treatment is
nearly deterministic get no weight. That is a bias relative to the DR-learner's
target, and a large variance reduction.

## Guidelines and an experiment {#sec-guidelines}

Table \@ref(tab:guidelines) condenses the discussion.

```{r guidelines, echo = FALSE}
tibble(
  Learner = c("S", "T", "X", "DR", "R"),
  `Needs` = c("outcome model", "outcome models", "outcome models (+ propensity to blend)",
              "outcome or propensity model", "propensity and outcome model"),
  `Converges to` = c("best approximation of the CATE", "best approximation", "best approximation",
                     "best approximation", "overlap-weighted approximation"),
  `Weak when` = c("effect small relative to baseline (shrinks it)", "one arm is small",
                  "CATE as complex as the baseline", "overlap weak (extreme labels)",
                  "propensity poorly estimated; ignores low-overlap regions")
) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Meta-learners at a glance.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  column_spec(2:4, width = "4cm")
```

Two experiments make the guidelines concrete. The first uses the three
designs of Chernozhukov et al. (2026, Example 15.1.1) with random-forest
oracles: design 1 as above; design 2 with a simple baseline and the jump
in the CATE, 5 percent treated; design 3 like design 2 with 95 percent
treated. Learners are fitted on 70 percent of a sample, the two ensembles
of Section \@ref(sec-ensembles) are stacked on the remaining 30 percent,
and the root mean squared error is computed against the true CATE on a
fresh sample.

```{r learner-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5, n_sims_learners)}
mc <- run_metalearner_experiment(n_sims = n_sims_learners, n_grid = c(500, 2000), seed = 1)
mc_tab <- summarise_metalearner_experiment(mc)
```

```{r tab-learner-mc, echo = FALSE}
mc_tab %>%
  mutate(cell = paste0(fmt(mean_rmse, 3), " (", fmt(sd_rmse, 3), ")")) %>%
  select(dgp, n, learner, cell) %>%
  tidyr::pivot_wider(names_from = learner, values_from = cell) %>%
  mutate(dgp = sub(":.*", "", dgp)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Design", "n", "S", "T", "X", "DA X", "DR", "R", "Q-agg.", "Best"),
      caption = paste0("Root mean squared error of the CATE (mean and standard deviation over ",
                       n_sims_learners, " simulations) with random-forest oracles. Design 1: constant effect, ",
                       "jump in the baseline, 5 percent treated. Design 2: jump in the effect, simple baseline, ",
                       "5 percent treated. Design 3: as design 2 with 95 percent treated.")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")
```

```{r learner-mc-numbers, echo = FALSE}
best_of <- function(design, size) {
  t <- mc_tab %>% filter(grepl(design, dgp), n == size, !learner %in% c("Q-aggregation", "Best single"))
  as.character(t$learner[which.min(t$mean_rmse)])
}
```

Three patterns stand out.

- Design 1 rewards outcome modelling with borrowed strength: the X-learner
  is best (`r best_of("DGP 1", 500)` at $n = 500$), because the control
  model is easy and the effect is constant. The learners that divide by
  the propensity pay for it: with about 17 treated units in the training
  part, the DR-learner's label multiplies their residuals by 20.
- Design 3 rewards the T-learner: with 95 percent treated the treated model
  is the good one and the effect is complex, so borrowing the control model
  hurts (`r best_of("DGP 3", 500)` is best at $n = 500$).
- Growing $n$ from 500 to 2000 lowers every learner's error but does not
  change which learner is best within a design. The learners that divide by
  the propensity remain behind with forest final stages even at
  $n = 2000$: their labels are noisy, and a forest fitted to noisy labels
  needs large leaves and many observations. The ensembles track the best
  base learner in every design without being told which one it is.

The book obtains the same qualitative picture with gradient-boosted
oracles, with the DR- and R-learners closer to the front in designs 1 and
2. The lesson is that the ranking depends on the oracle, the sample size,
and the design: there is no single best meta-learner, and Section
\@ref(sec-selection) replaces the guesswork with a data-driven choice.

The second experiment isolates double robustness. It uses the running
"smooth" design with *linear* oracles. The outcome regressions are
misspecified (the baseline is quadratic in $x_2$) while the logistic
propensity model is exactly right. The final stage is linear in $x_1$ and
$x_2^2$, which contains the true CATE.

```{r misspec-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5)}
mis <- run_misspecification_experiment(n_sims = 200, n = 2000, seed = 2)
mis_tab <- summarise_misspecification_experiment(mis)
```

```{r tab-misspec, echo = FALSE}
mis_tab %>%
  mutate(across(c(bias_rms, sd_rms, rmse), ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Learner", "Bias", "Standard deviation", "RMSE"),
      caption = "Misspecified linear outcome models, correct logistic propensity; n = 2000, 200 simulations. Bias is the root mean square over test points of the mean error across simulations; the standard deviation is the root mean square of the simulation standard deviation; RMSE combines both.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

The S-, T-, and X-learners inherit the outcome model's bias and cannot
remove it with more data; their errors are almost entirely bias. The DR-
and R-learners have almost no bias, because the correct propensity model
protects them as the error bounds promise, and their errors are variance,
which shrinks with $n$. This is the reason
to prefer them whenever the outcome model is the shaky part of the
analysis, which in observational data it usually is.

## Multi-arm and other extensions

With $K$ arms the S- and T-learners extend directly (one model with the arm
as a feature, or $K$ models), and pairwise X- and R-learners can be built
for each contrast. The R-learner then depends on the reference arm, and
Zhou, Zhang, and Tu (2023) propose a reference-free version that estimates
$E[Y^{(k)} - Y \mid X]$ for every arm under a simplex constraint. Neural
networks with a shared representation and a balancing penalty (Shalit,
Johansson, and Sontag 2017) are another route to outcome modelling under
covariate shift. None of these change the logic of this section: a
pseudo-label, then a regression.

## Interpreting a black-box CATE {#sec-interpret}

A forest-based $\hat\tau$ is not readable. Two devices restore
interpretation without new assumptions.

- **Distillation**: fit a shallow regression tree to $\hat\tau(X_i)$ on
  $X_i$ and read its splits.
- **Group differences**: compare covariate means between the top and
  bottom groups of $\hat\tau$ (Section \@ref(sec-validation) reports them).

```{r distill}
dr_fit <- cate_learner(scores = sc_train_rf, x_het = x5, method = "dr", learner_final = rf_final)
tau_test <- predict(dr_fit, test)
distill <- rpart::rpart(tau_hat ~ ., data = data.frame(tau_hat = tau_test, test[, x5]),
                        control = rpart::rpart.control(maxdepth = 2, minbucket = 50, cp = 0.01))
distill
```

The distillation is fitted to predictions on the test sample, never to
in-sample fitted values of a forest, which overfit their own noisy labels.
Its root splits on `r as.character(distill$frame$var[1])`, the variable
that drives the effect linearly in the design; the splits below use
`r paste(setdiff(unique(as.character(distill$frame$var)), c("<leaf>", as.character(distill$frame$var[1]))), collapse = " and ")`.

# Inference on summaries of the CATE {#sec-summaries}

## Best linear predictor {#sec-blp}

The most useful relaxation replaces $\tau(x)$ by its best linear
approximation in a dictionary $p(x)$ of $d \ll n$ functions:
$$
\beta_0 = \arg\min_\beta E\big[(\tau(X) - p(X)'\beta)^2\big]
        = E[p(X)p(X)']^{-1} E[p(X)\,\tau(X)]
        = E[p(X)p(X)']^{-1} E[p(X)\,Y(\eta_0)],
$$
where the last equality uses $E[Y(\eta_0) \mid X] = \tau(X)$. So $\beta_0$
is the population regression coefficient of the pseudo-outcome on $p(X)$,
and its sample analogue is ordinary least squares of $Y_i(\hat\eta)$ on
$p(X_i)$ (Semenova and Chernozhukov 2021; Chernozhukov et al. 2026, Section
14.2).

*Skeleton of why OLS inference is valid.* Write $\hat\beta - \beta_0 =
\hat Q^{-1} \mathbb{E}_n[p(X)(Y(\hat\eta) - p(X)'\beta_0)]$ with $\hat Q =
\mathbb{E}_n[p(X)p(X)']$. Replace $Y(\hat\eta)$ by $Y(\eta_0)$: the
difference has conditional mean equal to the product of the nuisance
errors (Section \@ref(sec-score)), which is $o(n^{-1/2})$ under the rate
condition of Section 04, and a conditional variance that vanishes with
consistency of the nuisances. What remains is OLS with the oracle
outcome, whose HC sandwich is the right variance. The dictionary can be
group indicators, polynomials, or splines. When it is group indicators the
coefficients are group average treatment effects (GATEs).

`cate_blp()` runs the regression, reports HC1 standard errors, and adds a
simultaneous band from the multiplier bootstrap of Section 07 applied to the
influence functions $\hat Q^{-1} p(X_i) \hat e_i$.

```{r blp}
blp <- cate_blp(sc_train_rf, ~ x1 + I(x2^2), seed = 1)
blp
blp_oracle <- cate_blp(sc_train_oracle, ~ x1 + I(x2^2), uniform = FALSE)
blp_oracle$coefficients %>% select(term, estimate, std.error) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

The true CATE is $0.5 + x_1 + 0.5 x_2^2$, so the three coefficients target
$0.5$, $1$, and $0.5$. The coefficient on $x_1$ is on target. The
coefficient on $x_2^2$ is `r fmt(blp$coefficients$estimate[3])` with the
forest nuisances and `r fmt(blp_oracle$coefficients$estimate[3])` with the
true nuisances: the difference is the product-of-errors term of Section
\@ref(sec-score), which is largest in the tail of $x_2$, where the forest's
outcome error and the propensity's leverage are both large. Better
nuisances, not a different second stage, close that gap. The Wald test
rejects a constant effect. The simultaneous critical value is above 1.96
because it covers all three coefficients at once.

Figure \@ref(fig:blp-curve) evaluates the curve along $x_1$ with pointwise
and simultaneous bands. The simultaneous band covers the whole curve with
probability 0.95; the pointwise band covers one pre-specified point.

```{r blp-curve, fig.cap = "Best linear predictor of the CATE along x1 (x2 fixed at 0) with pointwise (dark) and simultaneous (light) 95 percent bands. The dashed line is the true CATE."}
grid <- data.frame(x1 = seq(-2.5, 2.5, by = 0.1), x2 = 0)
plot_cate_blp(blp, grid, seed = 1) +
  geom_line(data = grid, aes(x1, 0.5 + x1), linetype = "dashed", colour = "black")
```

## Group average treatment effects {#sec-gate}

With group indicators as the dictionary, `cate_gate()` returns the doubly
robust GATEs with pointwise and simultaneous intervals and a joint test of
equality.

```{r gate}
q_x1 <- cut(train$x1, quantile(train$x1, 0:4 / 4), include.lowest = TRUE, labels = paste0("Q", 1:4))
gate <- cate_gate(sc_train_rf, q_x1, seed = 1)
gate
```

```{r gate-plot, echo = FALSE, fig.height = 3.4, fig.cap = "Group average treatment effects by quartile of x1. Thick bars: pointwise 95 percent intervals; thin bars: simultaneous intervals valid for all four groups at once."}
plot_cate_gate(gate)
```

The choice between pointwise and simultaneous intervals is a choice of
question. A pre-registered interest in the bottom quartile alone uses the
pointwise interval. A claim that "effects differ across quartiles" needs
the joint test or the simultaneous intervals.

## A subset of covariates is enough {#sec-subset}

The dictionary used $x_1$ and $x_2$ while confounding was handled by all
five covariates inside the scores. Nothing required $X = Z$. That is the
practical advantage of the pseudo-outcome route: heterogeneity can be
described in the two or three variables a decision maker can act on, while
the nuisances use everything.

## Continuous treatments {#sec-continuous-blp}

With a continuous treatment the natural target is the derivative
$E[\partial_t E[Y \mid T = t, X] \mid X]$ (Facure 2023, ch. 18). Under the
partially linear interaction model
$Y = (p(X)'\beta)\, T + f(Z) + e$, partialling out gives
$\tilde Y = (p(X)'\beta)\, \tilde T + e$ with $\tilde Y = Y - E[Y \mid Z]$,
$\tilde T = T - E[T \mid Z]$. So $\beta$ is OLS of $\tilde Y$ on
$p(X)\tilde T$, which is the R-learner loss of Section \@ref(sec-r-learner)
with a linear CATE. The residuals come from `est_dml(model = "plr")`; the
regression on $p(X)\tilde T$ is one `lm()` call with HC standard errors.
The heterogeneity is then a slope that varies with $X$, exactly the
"sensitivity" that price experiments estimate.

## Forests with confidence intervals {#sec-forests}

Inference at a point $x$ is harder than inference on a summary. Random
forests adapted for inference (Wager and Athey 2018; Athey, Tibshirani, and
Wager 2019) deliver it in low dimensions. The ingredients are:

- **honesty**: one subsample chooses the splits, another fills the leaves;
- **subsampling** without replacement, with subsample size $s \ll n$;
- **balanced splits** and a positive probability that each variable is
  tried at each split;
- fully grown trees with small leaves.

Under these conditions predictions are asymptotically normal, and a
variance estimate (from the variation across subsamples) gives confidence
intervals. Two constructions are in use.

- A **doubly robust forest** is an honest regression forest on the
  pseudo-outcome $Y(\hat\eta)$. Its validity in high dimensions of $Z$
  needs only the product of nuisance errors to be small conditionally on $X$
  (Oprescu, Syrgkanis, and Wu 2019).
- A **causal forest** solves the R-learner moment locally:
  $E[(\tilde Y - \tau(x)\tilde D)\tilde D \mid X = x] = 0$, using forest
  weights $K(x, X_i)$ that measure how often $X_i$ shares a leaf with $x$.
  This is the generalized random forest of `grf`, and it is valid when the
  CATE depends on the forest's variables only.

The theory is exact only under the tuning it assumes; cross-validated
hyperparameters break it. In practice the intervals are a useful signal of
where the CATE is uncertain, not a guarantee. The comparison below uses the
same data for `grf::causal_forest()` and for the DR-learner with a forest
final stage.

```{r forests, eval = run_grf}
library(grf)
cf <- causal_forest(as.matrix(train[, x5]), train$y, train$d, num.trees = 1000, seed = 1)
pred_cf <- predict(cf, as.matrix(test[, x5]), estimate.variance = TRUE)
dr_forest <- cate_learner(scores = sc_train_rf, x_het = x5, method = "dr",
                          learner_final = lrn("regr.ranger", num.trees = 500, min.node.size = 50))
c(rmse_causal_forest = sqrt(mean((pred_cf$predictions - test$tau_true)^2)),
  rmse_dr_forest = sqrt(mean((predict(dr_forest, test) - test$tau_true)^2)),
  coverage_causal_forest = mean(abs(pred_cf$predictions - test$tau_true) <= 1.96 * sqrt(pred_cf$variance.estimates)))
```

```{r forests-plot, echo = FALSE, eval = run_grf, fig.cap = "Causal forest (with pointwise 95 percent intervals) and DR-learner forest predictions along x1 on the test sample, other covariates as observed. The black line is the true CATE averaged in bins of x1."}
df <- tibble(x1 = test$x1, truth = test$tau_true, cf = pred_cf$predictions,
             cf_se = sqrt(pred_cf$variance.estimates), drf = predict(dr_forest, test)) %>%
  mutate(bin = cut(x1, breaks = quantile(x1, 0:25 / 25), include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarise(x1 = mean(x1), truth = mean(truth), cf = mean(cf), cf_se = mean(cf_se), drf = mean(drf), .groups = "drop")
ggplot(df, aes(x1)) +
  geom_ribbon(aes(ymin = cf - 1.96 * cf_se, ymax = cf + 1.96 * cf_se), fill = hte_colors[["R-learner"]], alpha = 0.2) +
  geom_line(aes(y = truth), colour = "black", linewidth = 1) +
  geom_line(aes(y = cf, colour = "Causal forest (grf)"), linewidth = 0.8) +
  geom_line(aes(y = drf, colour = "DR-learner, forest final stage"), linewidth = 0.8) +
  scale_colour_manual(values = c("Causal forest (grf)" = hte_colors[["R-learner"]],
                                 "DR-learner, forest final stage" = hte_colors[["DR-learner"]]), name = NULL) +
  labs(y = "CATE (binned means)") + hte_theme()
```

Both forests find the slope in $x_1$; the doubly robust one is noisier
because it regresses a noisy label. The causal forest's pointwise intervals
cover the true CATE for
`r round(100 * mean(abs(pred_cf$predictions - test$tau_true) <= 1.96 * sqrt(pred_cf$variance.estimates)))`
percent of the test units, below the nominal 95 percent: the default
tuning is not the tuning the theory assumes, and the intervals are a signal
of uncertainty rather than a guarantee. Where the two forests disagree, the
disagreement itself is informative: the R-learner moment weights regions by
$\mathrm{Var}(D \mid Z)$, the DR-learner does not (Section
\@ref(sec-r-learner)).

# Choosing a CATE model out of sample {#sec-selection}

## Why prediction metrics do not work

A CATE model cannot be scored by the usual out-of-sample mean squared error:
the target $\tau(X_i)$ is never observed, and neither is $Y_i(1) - Y_i(0)$.
Comparing models by how well they predict $Y$ rewards the baseline, not the
effect (Facure 2023, "Why prediction metrics are dangerous for causal
models"). The pseudo-outcome solves this: it is an observed variable whose
conditional mean is the target.

## The doubly robust loss and its differences {#sec-dr-loss}

For a candidate $\tau_m$ define, on a scoring sample,
$$
\hat L(\tau_m) = \mathbb{E}_n\big[(Y(\hat\eta) - \tau_m(X))^2\big].
$$
The loss itself contains the irreducible variance of the pseudo-outcome. Its
differences do not: for two candidates,
$$
L(\tau_i; \eta_0) - L(\tau_j; \eta_0) = E[(\tau_i(X) - \tau_0(X))^2] - E[(\tau_j(X) - \tau_0(X))^2].
$$
*Skeleton.* Expand the squares; the cross term is
$-2E[Y(\eta_0)(\tau_i - \tau_j)] = -2E[\tau_0(\tau_i - \tau_j)]$ by the key
property, and $E[Y(\eta_0)^2]$ cancels in the difference. So ranking by
$\hat L$ ranks by mean squared error against the true CATE. The difference
is Neyman orthogonal in $\eta$, and Theorem 15.2.1 of Chernozhukov et al.
(2026) gives $\sqrt n(\hat L_i - \hat L_j - \delta^*_{ij}) \to N(0, V)$ with
$V$ the variance of the per-observation loss difference (Appendix
\@ref(app-dr-loss)). Confidence intervals for the comparison follow.

The interpretable version compares each model with the constant-effect
model: $S(\tau) = 1 - \hat L(\tau) / \hat L(\hat\theta)$, in $(-\infty, 1]$,
is the fraction of the constant model's loss the candidate removes.

## Code: scoring the meta-learners {#sec-score-code}

We fit the five meta-learners on the training sample with forests, then
score them on the scoring sample. The scoring sample gets its own
cross-fitted pseudo-outcomes.

```{r score-fit}
fit_one <- function(m, ...) cate_learner(scores = sc_train_rf, x = x5, x_het = c("x1", "x2"), method = m,
                                         learner = rf_first, learner_final = rf_final, ...)
models <- list(s_learner = fit_one("s"), t_learner = fit_one("t"), x_learner = fit_one("x"),
               dr_learner = fit_one("dr"), r_learner = fit_one("r"))
sc_score <- dr_scores(score_set, "y", "d", x5, learner_mu = rf_first, seed = 2)
cs <- do.call(cate_score, c(list(sc_score), models))
cs
```

```{r score-truth, echo = FALSE}
truth_rmse <- sapply(models, function(m) sqrt(mean((predict(m, score_set) - score_set$tau_true)^2)))
```

The doubly robust losses order the models
`r paste(cs$table$model[order(cs$table$loss)], collapse = ", ")`; the true
root mean squared errors on the same rows, which only a simulation can
compute, order them
`r paste(names(sort(truth_rmse)), collapse = ", ")`. The confidence intervals
say which differences are real: a candidate whose interval for `diff`
includes zero has not been shown to beat a constant effect.

## Ensembles {#sec-ensembles}

Selecting the single best model has a $\sqrt{\log(M)/n}$ excess risk. Stacking
the candidates by minimizing the doubly robust loss over convex weights
achieves $M/n$; Q-aggregation, which adds the penalty $\sum_m w_m \hat L(\tau_m)$
to the stacked loss, achieves $\log(M)/n$, the best possible dependence on
both $M$ and $n$ (Lecué and Rigollet 2014; Lan and Syrgkanis 2023). The
intercept is fixed at the ATE and the base predictions are centered, so the
ensemble never moves the average effect.

```{r ensemble}
ens <- do.call(cate_ensemble, c(list(sc_score, method = "q"), models))
round(ens$ensemble$weights, 3)
ens_rules <- sapply(c("q", "convex", "best", "ols"), function(r) {
  e <- do.call(cate_ensemble, c(list(sc_score, method = r), models))
  sqrt(mean((predict(e, test) - test$tau_true)^2))
})
best_single_rmse <- min(sapply(models, function(m) sqrt(mean((predict(m, test) - test$tau_true)^2))))
round(c(ens_rules, truth_best_single = best_single_rmse), 3)
```

The best base learner on the test sample has root mean squared error
`r fmt(best_single_rmse, 3)`;
the convex ensemble reaches `r fmt(ens_rules["convex"], 3)`, Q-aggregation
`r fmt(ens_rules["q"], 3)`, and the single best-scored model
`r fmt(ens_rules["best"], 3)`. No rule needs to know which base learner is
best, and the convex rule comes closest. The scoring sample has only
`r nrow(score_set)` rows, so its ranking of the learners is imperfect
(Section \@ref(sec-score-code) showed the doubly robust losses and the true
errors ordering the learners differently), and the ensembles inherit that
noise. Unconstrained least squares can extrapolate outside the convex hull
and is less stable.

## What the pipeline cannot give

The ensemble comes without confidence intervals for its predictions and can
change with the random split, the seeds of the forests, or a few removed
observations. Report sensitivity to these choices. For statements with
error control, hold out a third sample and use the tests of the next
section.

# Validating the chosen model {#sec-validation}

## Three samples

Train on one sample, score and stack on a second, test on a third. The test
sample sees the final model only as a fixed function $\tau^*$. Everything
in this section is computed on the test sample's own cross-fitted
pseudo-outcomes.

```{r test-scores}
sc_test <- dr_scores(test, "y", "d", x5, learner_mu = rf_first, seed = 3)
val <- cate_validate(sc_test, ens, n_boot = 999, seed = 1)
```

## Heterogeneity test from the best linear predictor {#sec-het-test}

Regress $Y(\hat\eta)$ on $(1, \tau^*(X) - \bar\tau^*)$. The slope converges
to
$$
\beta_1 = \frac{\mathrm{Cov}(\tau_0(X), \tau^*(X))}{\mathrm{Var}(\tau^*(X))}
        = \frac{\mathrm{Cov}(Y(1) - Y(0), \tau^*(X))}{\mathrm{Var}(\tau^*(X))}.
$$
A significant slope means the model is correlated with the individual
effect, which also proves that heterogeneity in $X$ exists. A slope of one
means the model's scale is right. The intercept is the ATE because the
regressor is centered.

```{r het-test}
val$blp
```

Here the slope is `r fmt(val$blp$estimate[2])` with standard error
`r fmt(val$blp$std.error[2])`: the test does not reject a zero slope, while
the grouped diagnostics below detect heterogeneity clearly. The slope is a
regression on a heavy-tailed outcome and a few extreme predictions carry
much of its variance; the grouped statistics average within quantile
groups and are far more stable. Report both.

## Calibration {#sec-calibration}

A calibrated CATE model satisfies $E[Y(1) - Y(0) \mid \tau^*(X) = t] = t$.
The mean squared error decomposes as calibration error plus distortion,
$E[(\tau^* - \tau_0)^2] = \mathrm{CAL}_2(\tau^*) + E[\mathrm{Var}(\tau_0(X) \mid \tau^*(X))]$
(the proof is the variance decomposition with $E[\tau_0 \mid \tau^*]$ as the
conditional mean), so any consistent model is eventually calibrated, but a
model can be calibrated without being accurate. The binned check compares,
in each quantile group of $\tau^*$, the doubly robust GATE with the mean
prediction.

```{r calibration}
val$calibration$table
c(cal1 = val$calibration$cal1, cal1_normalized = val$calibration$cal1_normalized,
  p_equal = val$calibration$test_equal$p.value)
```

```{r calibration-plot, echo = FALSE, fig.height = 3.6, fig.cap = "Calibration of the ensemble on the test sample: doubly robust group average effect (with 95 percent intervals) against the mean prediction in each quartile of predicted CATE. Points on the diagonal are calibrated."}
plot_cate_validation(val, "calibration")
```

## Targeting curves {#sec-toc}

Suppose we can treat only a share $q$ of the population and rank units by
$\tau^*$. The **TOC** curve (Chernozhukov et al. 2026, Section 15.3) is the
average effect among the top $q$ minus the ATE,
$$
\mathrm{TOC}(q) = E\big[Y(1) - Y(0) \mid \tau^*(X) \ge \mu(\tau^*, q)\big] - \mathrm{ATE}
              = \mathrm{Cov}\Big(Y(1) - Y(0),\ \frac{\mathbb{1}\{\tau^*(X) \ge \mu(\tau^*, q)\}}{q}\Big),
$$
and the **QINI** curve scales it by the share treated,
$\mathrm{QINI}(q) = q\, \mathrm{TOC}(q) = \mathrm{Cov}(Y(1) - Y(0), \mathbb{1}\{\tau^*(X) \ge \mu(\tau^*, q)\})$.
The covariance forms (Appendix \@ref(app-uplift)) show that both are
identified by replacing $Y(1) - Y(0)$ with $Y(\eta_0)$, so both are
averages of the pseudo-outcome with fixed weights. Their influence
functions give pointwise standard errors; the multiplier bootstrap gives a
one-sided simultaneous band. If the band lies above zero at any $q$, the
model has found heterogeneity with statistical significance. The areas
under the curves, AUTOC and AUQC, integrate over all ranks and are the
rank-weighted average treatment effects of Yadlowsky et al. (2021).

```{r toc}
val$areas
val$heterogeneity_statistic
```

```{r toc-plot, echo = FALSE, fig.cap = "TOC curve of the ensemble on the test sample with pointwise (dotted) and one-sided simultaneous (shaded) 95 percent lower bounds. Treating the top 20 percent by predicted CATE yields an average effect about TOC(0.2) above the ATE."}
plot_cate_validation(val, "toc")
```

The TOC curve reads: treating the top 5 percent by predicted CATE yields
an average effect `r fmt(val$curves$toc[val$curves$q == 0.05])` above the
ATE of `r fmt(val$ate)`; the gain declines to zero as $q$ reaches one. The
QINI curve reads as total gain over treating a random share $q$; it peaks
at the share where marginal units stop being better than average.

## Group differences {#sec-group-diff}

When the top and bottom groups are separated statistically, describing them
is the last diagnostic: covariate means in the two groups, or a shallow
classification tree that separates them.

```{r group-diff}
val$group_diff %>% mutate(across(where(is.numeric), ~ round(.x, 3)))
```

In the design, $x_1$ drives the effect linearly and $x_2$ quadratically.
The table shows the top quartile has a much higher mean of $x_1$, while the
mean of $x_2$ barely differs: a quadratic effect changes the spread of
$x_2$, not its mean, and a comparison of means cannot see it. $x_3$ to
$x_5$ do not differ, as they should not.

## Ties

A model that predicts the same value for many units (a tree, a coarse
rule) leaves the top-$q$ set ill defined. The package treats units above
the threshold deterministically and units at the threshold with the
probability that fills the share $q$ (Chernozhukov et al. 2026, Remark
15.3.1); the areas use the average weight within each tied block, which is
the expectation of that rule.

# Policy learning {#sec-policy}

## Policies and their value {#sec-policy-value}

A policy $\pi: X \to \{0, 1\}$ says whom to treat. Its gain over treating
no one, net of a per-unit cost $c$, is
$$
V(\pi) = E\big[\pi(X)\,(Y(1) - Y(0) - c)\big] = E\big[\pi(X)\,(Y(\eta_0) - c)\big],
$$
the second equality by the key property of Section \@ref(sec-score), since
$\pi(X)$ is a function of $X$. So the value of a *fixed* policy is an
average of the pseudo-outcome, with a standard error from its influence
function. The gain over treating everyone is $E[(\pi(X) - 1)(Y(\eta_0) - c)]$.
Under a budget that allows a share $q$ to be treated, the best rule treats
the top $q$ by $\tau(X)$ and its value is $q \cdot (\mathrm{TOC}(q) + \mathrm{ATE})$
in the notation of Section \@ref(sec-toc).

```{r policy-value}
policy_value(sc_test, list(sign_of_ensemble = predict(ens, test) > 0,
                           treat_all = rep(1, nrow(test)),
                           oracle = test$tau_true > 0), cost = 0.5)
```

With a cost of 0.5 per treated unit, treating everyone is clearly worse
than the rule that treats units whose predicted effect exceeds the cost.
The rule that knows the true CATE and the estimated rule are
indistinguishable within sampling error: the difference between them is
smaller than two of its standard errors.

## Optimal policies are classifiers {#sec-classification}

Without constraints the optimal policy is $\pi^*(x) = \mathbb{1}\{\tau(x) > c\}$:
only the *sign* of $\tau(x) - c$ matters. Learning a policy is therefore
closer to classification than to regression. Mistakes cost more where
$|\tau(x) - c|$ is large, so it is cost-sensitive classification. Formally,
with $\Gamma = Y(\eta_0) - c$,
$$
\arg\max_\pi V(\pi) = \arg\max_\pi E\big[(2\pi(X) - 1)\,\Gamma\big]
                    = \arg\max_\pi E\big[\mathbb{1}\{2\pi(X) - 1 = \mathrm{sign}(\Gamma)\}\,|\Gamma|\big].
$$
*Skeleton.* $(2\pi - 1)\Gamma = (2\pi - 1)\mathrm{sign}(\Gamma)|\Gamma|$, and
$(2\pi - 1)\mathrm{sign}(\Gamma)$ equals $+1$ when the policy matches the
sign and $-1$ otherwise; adding the constant $E|\Gamma|$ and halving gives
the indicator form. So any weighted classifier, trained with labels
$\mathrm{sign}(\hat\Gamma_i)$ and weights $|\hat\Gamma_i|$, maximizes the
empirical value.

## Empirical welfare maximization in a restricted class {#sec-ewm}

Plugging a CATE model into the sign rule is one option. Empirical welfare
maximization (Kitagawa and Tetenov 2018; Athey and Wager 2021) is another:
choose the rule in a class $\Pi$ that maximizes the sample value
$\hat V(\pi) = \mathbb{E}_n[\pi(X)\hat\Gamma]$. Restricting $\Pi$ has two
purposes. A depth-two tree or a linear rule can be explained, audited, and
implemented, which a forest cannot. And the regret of the learned rule,
$$
R(\hat\pi) = \max_{\pi \in \Pi} V(\pi) - V(\hat\pi)
           \lesssim \sqrt{\frac{V^*\, \mathrm{VC}(\Pi)}{n}}
           + \|H(\hat p) - H(p_0)\|\,\|\hat g - g_0\|,
$$
is controlled by the complexity of $\Pi$ (its VC dimension), not by the
complexity of the CATE (Athey and Wager 2021, Theorem 1; Appendix
\@ref(app-regret) sketches the reduction). The nuisance errors enter as a
product, as everywhere else in this lecture. The plug-in rule inherits the
CATE model's estimation error instead, which is larger when the CATE is
harder to learn than its sign.

## Code: learning a tree {#sec-policy-code}

The `"policy"` design has a CATE that is piecewise constant on a
depth-two tree in $x_1$, $x_2$, $x_3$ (plus a small smooth term in $x_4$),
confounded assignment, and a nonlinear baseline. `policy_learn()` searches
depth-one and depth-two trees exactly, holding out half of the rows to
report an honest value.

```{r policy-tree}
pol_dat <- sim_hte(4000, dgp = "policy", seed = 2)
sc_pol <- dr_scores(pol_dat, "y", "d", x5, learner_mu = rf_first, seed = 1)
pol <- policy_learn(sc_pol, method = "tree", depth = 2, seed = 1)
pol
attr(pol_dat, "optimal_tree")
```

```{r policy-tree-plot, echo = FALSE, fig.height = 3, fig.cap = "The depth-two policy tree learned by empirical welfare maximization on the policy design. It recovers the structure of the optimal rule."}
plot_policy_tree(pol)
```

The estimation-sample value is optimistic: it is the maximum over many
candidate trees. The held-out value is not. `engine = "policytree"`
delegates the search to the `policytree` package (Sverdrup et al. 2020),
which handles any depth with the same objective:

```{r policytree, eval = run_policytree}
pol_pt <- policy_learn(sc_pol, method = "tree", depth = 2, engine = "policytree", seed = 1)
c(agreement = mean(predict(pol_pt, pol_dat) == predict(pol, pol_dat)))
```

## Regret of trees versus plug-in rules {#sec-regret-mc}

Figure \@ref(fig:policy-mc) compares four rules on the policy design as $n$
grows: depth-one and depth-two trees from empirical welfare maximization,
and the signs of T- and DR-learner CATE models. Regret is measured with the
true CATE on a large test sample.

```{r policy-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5, n_sims_policy), fig.height = 3.8, fig.cap = "Regret (value of the optimal rule minus value of the learned rule, in outcome units) on the policy design, mean over simulations with a two-standard-error band. The depth-two tree has the right class and learns fastest; the depth-one tree has the wrong class and its regret does not vanish; plug-in sign rules need to learn the magnitude before the sign."}
pmc <- run_policy_experiment(n_sims = n_sims_policy, n_grid = c(500, 1000, 2000, 4000), seed = 3)
pmc_tab <- pmc %>% group_by(n, policy) %>%
  summarise(regret = mean(regret), se = sd(regret) / sqrt(n()), .groups = "drop")
ggplot(pmc_tab, aes(n, regret, colour = policy)) +
  geom_ribbon(aes(ymin = regret - 2 * se, ymax = regret + 2 * se, fill = policy), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) + geom_point() +
  scale_x_log10(breaks = c(500, 1000, 2000, 4000)) +
  scale_colour_manual(values = c("Tree, depth 1" = "#999999", "Tree, depth 2" = hte_colors[["DR-learner"]],
                                 "Sign of T-learner CATE" = hte_colors[["T-learner"]],
                                 "Sign of DR-learner CATE" = hte_colors[["R-learner"]]), name = NULL) +
  scale_fill_manual(values = c("Tree, depth 1" = "#999999", "Tree, depth 2" = hte_colors[["DR-learner"]],
                               "Sign of T-learner CATE" = hte_colors[["T-learner"]],
                               "Sign of DR-learner CATE" = hte_colors[["R-learner"]]), guide = "none") +
  labs(x = "n (log scale)", y = "Regret") + hte_theme() + guides(colour = guide_legend(nrow = 2))
```

Two lessons. The class must contain a good rule: the depth-one tree
cannot express the optimal policy and its regret stays flat. Within the
right class, learning the sign directly is easier than learning the
magnitude: at every $n$ the depth-two tree beats the plug-in rules built
from forests.

## Budgets and other classes {#sec-policy-classes}

- **Budget**: `policy_learn(method = "budget", budget = q, tau_hat = model)`
  treats the top share $q$ ranked by a CATE model, the constrained optimum
  when the ranking is right. Its value is the QINI curve at $q$ plus $q$
  times the ATE.
- **Linear rules** $\mathbb{1}\{x'b > 0\}$: Kitagawa and Tetenov solve the
  exact problem by integer programming; the package fits the weighted
  logistic surrogate, which is fast and usually close.
- **Any classifier**: `method = "classifier"` passes labels and weights to
  an `mlr3` classifier; deeper trees or forests give flexible but opaque
  rules.

```{r policy-classes}
tau_pol <- cate_learner(scores = sc_pol, x_het = x5, method = "dr", learner_final = rf_final)
rules <- list(
  tree2 = pol,
  budget30 = policy_learn(sc_pol, method = "budget", budget = 0.3, tau_hat = tau_pol, seed = 1),
  linear = policy_learn(sc_pol, method = "linear", seed = 1)
)
do.call(rbind, lapply(names(rules), function(nm) cbind(rule = nm, rules[[nm]]$value[2, ])))
```

## Targeting impact versus deprivation {#sec-frontier}

Haushofer, Niehaus, Paramo, Miguel, and Walker (2025) ask whom a cash
transfer should reach: the poorest, or those who gain most. With a
concave utility $u$ and predicted untreated outcome $\hat y_0$ (deprivation)
and effect $\hat\tau$ (impact), a planner treats units in order of
$u(\hat y_0 + \hat\tau) - u(\hat y_0)$. Linear utility ranks on impact;
strong curvature ranks on deprivation. `policy_frontier()` traces the
overlap of the chosen set with impact-only and deprivation-only targeting
as curvature grows, and evaluates each rule with the doubly robust value.

```{r frontier}
fr <- policy_frontier(sc_test, ens, y0_hat = sc_test$nuisance$mu0, budget = 0.3,
                      utility = "cara", curvature = c(0, 0.5, 1, 2, 5))
fr
```

```{r frontier-plot, echo = FALSE, fig.height = 3.6, fig.cap = "Units selected by a planner with CARA curvature 1 and a 30 percent budget, in the plane of predicted untreated outcome and predicted effect. The planner trades impact for deprivation along the frontier."}
plot_policy_frontier(fr, curvature = 1)
```

In the running design impact and deprivation are only weakly related, so
the planner's choice moves smoothly from one corner to the other; the
value of the rule falls as curvature rises because the value counts effects,
not utility. The paper's finding is that in Kenya the two criteria pick
different households, so deprivation-only targeting leaves impact on the
table.

## Cautions {#sec-policy-cautions}

- **Variance and overlap.** A candidate policy that treats where the data
  rarely treated has a noisy value; regret bounds carry the worst overlap
  ratio. Variance penalization, distributionally robust objectives, and
  pessimistic (lower-confidence-bound) objectives address this, at the cost
  of leaving the classification reformulation.
- **Adaptive data.** When assignment probabilities were updated during
  collection (bandits), the pseudo-outcome must use the realized
  probabilities and stabilizing weights (Zhan, Ren, Athey, and Zhou 2024).
- **Dynamic regimes.** Sequential treatments need backward induction over
  stages; Section 07's L19 material and Q-learning cover it.
- **Honesty.** Report the value on data the search never saw. The
  in-sample value of the best of many rules is biased upward.

# Traps {#sec-traps}

## Binary outcomes and the curvature trap {#sec-binary-trap}

With a binary outcome the effect on the probability scale depends on where
a unit sits on the response curve. Facure (2023, ch. 23) gives the cleanest
example. A randomized nudge, income that moves only the baseline, and age
that moves only the latent effect:
$$
Y^* = -4.5 + 0.001\,\mathrm{income} + D\,(1 + 0.01\,\mathrm{age}) + \varepsilon, \qquad Y = \mathbb{1}\{Y^* > t\}.
$$
On the latent scale, income has no effect heterogeneity. On the
probability scale it does, through the slope of the normal CDF, and it can
dominate age.

```{r binary-trap, echo = FALSE, fig.cap = "Cumulative effect curves (Facure 2023): units ordered by a covariate, difference in means between treated and controls among the top share. Left: latent outcome, where only age matters. Right: the binary conversion with a 12 percent base rate, where income (which moves only the baseline) sorts units by treatment effect more sharply than age."}
bin <- sim_hte(50000, dgp = "binary_outcome", threshold = 2, seed = 1)
curves <- bind_rows(
  cumulative_effect_curve(bin, "d", "latent", "age") %>% mutate(outcome = "Latent outcome", ordered_by = "age"),
  cumulative_effect_curve(bin, "d", "latent", "income") %>% mutate(outcome = "Latent outcome", ordered_by = "income"),
  cumulative_effect_curve(bin, "d", "y", "age") %>% mutate(outcome = "Conversion (12 percent)", ordered_by = "age"),
  cumulative_effect_curve(bin, "d", "y", "income") %>% mutate(outcome = "Conversion (12 percent)", ordered_by = "income")
)
ggplot(curves, aes(share, effect, colour = ordered_by)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ factor(outcome, levels = c("Latent outcome", "Conversion (12 percent)")), scales = "free_y") +
  scale_colour_manual(values = c(age = hte_colors[["DR-learner"]], income = hte_colors[["T-learner"]]), name = "Ordered by") +
  labs(x = "Share of units (ordered from high to low)", y = "Effect among the top share") + hte_theme()
```

The trap has a use. When conversion is rare, units with a high baseline
are on the steep part of the curve and respond most; when conversion is
common, units with a low baseline do. A CATE model on the probability
scale is right about the effect on conversion, which is what a business
cares about, but the ranking it produces is driven by the baseline, not by
the latent sensitivity. Decide which scale the decision lives on before
modelling heterogeneity.

## Extrapolation of parametric summaries {#sec-extrapolation}

A best linear predictor in a polynomial of one covariate is fitted where
the data are dense and extrapolated where they are not. Heavy tails in
the covariate produce wild curves in the tails, while forests regularize
toward the nearest data (Chernozhukov et al. 2026, Section 15.5). Plot the
covariate's distribution next to the curve, and report the curve only over
the central range.

## Specification search {#sec-search}

CATE estimates are noisy and rarely the paper's main object, so they are
prone to being reported when interesting and dropped when not (Childers
2024). Pre-register the heterogeneity dimensions or the analysis pipeline;
use held-out data for every claim; treat data-driven subgroups as
hypotheses to be tested in the next experiment.

## Heterogeneity that is not the CATE {#sec-distributions}

Two other questions hide under "heterogeneous effects".

- **Effects on distributions.** Quantile treatment effects
  $Q_{Y(1)}(u) - Q_{Y(0)}(u)$ and effects on the CDF are identified by the
  same adjustment formulas as the mean, applied to $\mathbb{1}\{Y \le y\}$
  (distribution regression) or to the check loss (quantile regression;
  Chernozhukov, Fernández-Val, and Melly 2013). Parallel trends is the one
  design where this fails, because the assumption is about means (Athey
  and Imbens 2006).
- **Distributions of effects.** The distribution of $Y(1) - Y(0)$ is *not*
  identified without further assumptions, because the two potential
  outcomes are never observed together. Only bounds (Makarov bounds from
  the Fréchet-Hoeffding inequalities) or functional-form assumptions
  deliver it.

The CATE is neither: it is the conditional mean of the effect, the one
feature of the effect distribution that adjustment identifies at every $x$.

```{r qte}
qte <- sapply(c(0.1, 0.25, 0.5, 0.75, 0.9), function(u) {
  q1 <- quantreg::rq(y ~ 1, tau = u, data = subset(bin, d == 1))
  q0 <- quantreg::rq(y ~ 1, tau = u, data = subset(bin, d == 0))
  coef(q1) - coef(q0)
})
round(setNames(qte, paste0("QTE at ", c(0.1, 0.25, 0.5, 0.75, 0.9))), 3)
```

Quantile effects of the nudge on the binary outcome are zero at every
quantile except where the treated distribution crosses the threshold; on
the latent outcome they would all be about 1.4. The two scales tell
different stories, again.

# Practice {#sec-practice}

## Replications and anchors {#sec-replications}

Four papers anchor this section; their replication notes live next to
the lecture in the notes repository.

- **Athey and Wager (2021), policy learning with observational data.**
  Depth-limited policy trees from doubly robust scores in the California
  GAIN experiments; the simulation evidence (their Figure 2 and Appendix
  Table III) is reproducible from the saved outputs, and the tree search
  is `policy_learn()` or `policytree`.
- **Haushofer, Niehaus, Paramo, Miguel, and Walker (2025), targeting impact
  versus deprivation.** Causal forests predict untreated consumption and
  the transfer effect; a planner with CARA or CRRA utility trades them off.
  `policy_frontier()` reproduces the exercise from any pair of predictions.
- **Yoganarasimhan, Barzegary, and Pani (2023), optimal free trials.**
  Seven trial lengths as arms, lasso, boosting, causal trees and forests
  compared by out-of-sample scoring; a multi-arm case that reduces to
  pairwise contrasts and the scoring of Section \@ref(sec-selection).
- **Zhan, Ren, Athey, and Zhou (2024), policy learning with adaptively
  collected data.** Doubly robust scores with adaptive weights when the
  assignment probabilities changed during the experiment.

## A checklist for heterogeneity and targeting {#sec-checklist}

1. State the question: description (for whom), or decision (whom to
   treat). The second needs the value and a policy class, not the CATE's
   magnitude.
2. Split the data: train, score, test. Report which sample produced each
   number.
3. Build the pseudo-outcome with cross-fitted nuisances. Report the
   nuisance learners, their fit, and the propensity distribution.
4. For inference on summaries, use the best linear predictor or GATEs with
   simultaneous bands. Say which covariates define heterogeneity and which
   handle confounding.
5. For flexible models, fit several meta-learners and forests, score them
   on the scoring sample with confidence intervals, and stack.
6. Validate the chosen model on the test sample: heterogeneity slope,
   calibration, TOC and QINI with bands, AUTOC.
7. For policies, learn in a class you can implement; report the held-out
   value with a standard error, against treating everyone and no one.
8. Check stability across seeds and splits; state the scale (latent,
   probability, money) the decision lives on.

# Appendix {#sec-appendix}

## Conditional orthogonality of the pseudo-outcome {#app-orthogonality}

Let $\eta = (p, g)$ and write $\mu_d = g(d, \cdot)$. Conditioning on $Z$,
$$
E[Y(\eta) \mid Z] = \mu_1 - \mu_0 + \frac{p_0}{p}\,(\mu_{1,0} - \mu_1) - \frac{1 - p_0}{1 - p}\,(\mu_{0,0} - \mu_0),
$$
where $\mu_{d,0} = g_0(d, Z)$. Collect terms:
$$
E[Y(\eta_0) - Y(\eta) \mid Z]
 = \Big(1 - \frac{p_0}{p}\Big)(\mu_{1,0} - \mu_1) - \Big(1 - \frac{1 - p_0}{1 - p}\Big)(\mu_{0,0} - \mu_0)
 = (H(p_0) - H(p))\,(g_0(D, Z) - g(D, Z))
$$
in conditional expectation, since $E[D \mid Z] = p_0$. Each factor vanishes
at $\eta_0$, so the Gateaux derivative in any direction is zero, and the
bias is the product of the two errors. Averaging over $Z$ given $X$ gives
the statement of Section \@ref(sec-score).

## Error bounds of the DR- and R-learners {#app-learner-bounds}

*DR-learner.* Let $\hat\tau$ be the oracle's output on $\{X_i, Y_i(\hat\eta)\}$
and $\bar\tau$ its output on the true labels $\{X_i, Y_i(\eta_0)\}$. For
square-loss oracles that are stable (a Lipschitz map from labels to
predictions, as least squares in a fixed class or a regularized learner
under standard conditions), $\|\hat\tau - \bar\tau\|_{L_2(X)} \lesssim
\|E[Y(\hat\eta) - Y(\eta_0) \mid X]\|_{L_2(X)} + \text{(noise term)}$. The
first term equals $\|E[(H(p_0) - H(\hat p))(g_0 - \hat g) \mid X]\|$ by
Appendix \@ref(app-orthogonality), bounded by the product of the nuisance
$L_4$ errors by Cauchy-Schwarz. The noise term is second order because
the label perturbation has conditional mean given by the product and
conditional variance that vanishes with consistency. Then
$\|\hat\tau - \tau^*\| \le \|\bar\tau - \tau^*\| + \|\hat\tau - \bar\tau\|$
gives $r_n + \mathrm{Err}(\hat g)\mathrm{Err}(H(\hat p))$; squaring gives
the bound in Section \@ref(sec-dr-learner). Foster and Syrgkanis (2023)
give the general theorem (orthogonal statistical learning), Kennedy (2023)
sharp rates for specific oracles.

*R-learner.* The population loss is $L(\tau; \eta) = E[(Y - l(Z) - \tau(X)(D - p(Z)))^2]$.
Its derivative in $\tau$ at $(\tau_0, \eta_0)$ is orthogonal to $\eta$ in
the sense that $\partial_\eta \partial_\tau L(\tau_0; \eta_0) = 0$, which is
the partialling-out orthogonality of Section 04 applied conditionally. The
second-order remainder contains $E[(\hat p - p_0)^2 (\tau - \tau_0)^2]$-type
terms and $E[(\hat p - p_0)(\hat l - l_0)(\tau - \tau_0)]$; bounding them
by Cauchy-Schwarz gives the $\mathrm{Err}(\hat p)^4$ and
$\mathrm{Err}(\hat p)^2\mathrm{Err}(\hat l)^2$ terms of Section
\@ref(sec-r-learner) (Nie and Wager 2021, Theorem 1; Foster and Syrgkanis
2023).

## Loss differences are asymptotically normal {#app-dr-loss}

Let $\Delta_i = (Y_i(\eta) - \tau_i)^2 - (Y_i(\eta) - \tau_j)^2 = (\tau_i - \tau_j)(\tau_i + \tau_j - 2Y(\eta))$
evaluated at $X_i$. Then $\hat\delta_{ij} = \mathbb{E}_n[\Delta_i]$ is a
sample mean whose population value at $\eta_0$ is
$E[(\tau_i - \tau_0)^2] - E[(\tau_j - \tau_0)^2]$ (expand and use
$E[Y(\eta_0) \mid X] = \tau_0$). The dependence on $\eta$ enters only
through $E[Y(\eta) \mid X]$, so the functional is Neyman orthogonal and its
nuisance error is the product bound of Appendix \@ref(app-orthogonality)
weighted by $\tau_i - \tau_j$. Under
$\sqrt n\,\|H(\hat p) - H(p_0)\|_{L_2}\|\hat g - g_0\|_{L_2} \to 0$ and
consistency, $\sqrt n(\hat\delta_{ij} - \delta^*_{ij}) \to N(0, V)$ with
$V = \mathrm{Var}(\Delta)$, estimated by the sample variance of
$\hat\Delta_i$ (Chernozhukov et al. 2026, Theorem 15.2.1). The variance is
bounded below by $4E[(\tau_i - \tau_j)^2 \mathrm{Var}(Y(\eta_0) \mid X)]$
(their Appendix 15.A), which is why models must be separated by more than
the estimation error for the comparison to be informative.

## TOC and QINI as covariances {#app-uplift}

Let $A = Y(1) - Y(0)$ and $B_q = \mathbb{1}\{\tau^*(X) \ge \mu(\tau^*, q)\} / P(\tau^*(X) \ge \mu(\tau^*, q))$,
so $E[B_q] = 1$. Then $\mathrm{TOC}(q) = E[A \mid B_q > 0] - E[A] = E[AB_q] - E[A]E[B_q] = \mathrm{Cov}(A, B_q)$.
Likewise $\mathrm{QINI}(q) = (E[A \mid B_q > 0] - E[A])\,P(B_q > 0) = \mathrm{Cov}(A, \mathbb{1}\{B_q > 0\})$.
Since $B_q$ is a function of $X$ and $E[A \mid X] = E[Y(\eta_0) \mid X]$,
both covariances are unchanged when $A$ is replaced by $Y(\eta_0)$. The
influence function of $\widehat{\mathrm{TOC}}(q)$ is
$\psi_q = (Y(\eta_0) - \theta_0)(\mathbb{1}\{\cdot\}/\pi_0(q) - 1) - \mathrm{TOC}(q)$
with $\theta_0$ the ATE and $\pi_0(q)$ the treated share; the estimated
ATE and share are nuisances whose first-order effect vanishes because the
functional is orthogonal to both (the covariance form is invariant to the
centering). Joint normality over a grid of $q$ and the simultaneous band
follow from the high-dimensional central limit theorem and the multiplier
bootstrap of Section 07 (Chernozhukov et al. 2026, Theorem 15.3.1).
Integrating over all $q$ with $u = k/n$ gives the rank weights of Section
\@ref(sec-toc): the unit ranked $r$ contributes $\sum_{k \ge r} 1/k - 1$ to
AUTOC.

## Regret of empirical welfare maximization {#app-regret}

Let $\hat\pi = \arg\max_{\pi \in \Pi} \hat V(\pi)$ with
$\hat V(\pi) = \mathbb{E}_n[\pi(X)\hat\Gamma]$ and $\pi^* = \arg\max_{\pi \in \Pi} V(\pi)$.
Then
$$
V(\pi^*) - V(\hat\pi) = [V(\pi^*) - \hat V(\pi^*)] + [\hat V(\pi^*) - \hat V(\hat\pi)] + [\hat V(\hat\pi) - V(\hat\pi)]
                     \le 2 \sup_{\pi \in \Pi} |\hat V(\pi) - V(\pi)|,
$$
because the middle bracket is at most zero by the definition of
$\hat\pi$. The supremum splits into the oracle part, with the true
$\Gamma$, and the nuisance part. The oracle part is the supremum of an
empirical process indexed by $\Pi$ and is $O_p(\sqrt{V^*\,\mathrm{VC}(\Pi)/n})$
by a uniform law of large numbers with the variance proxy $V^*$. The
nuisance part is $\sup_\pi |\mathbb{E}_n[\pi(X)(\hat\Gamma - \Gamma)]|$,
whose leading term is the product of nuisance errors by the orthogonality
of the pseudo-outcome, with cross-fitting removing the dependence between
$\hat\eta$ and the sample (Athey and Wager 2021, Theorem 1 and Lemma 2).

## References

- Athey, Susan, and Guido W. Imbens. 2006. "Identification and Inference in Nonlinear Difference-in-Differences Models." *Econometrica*.
- Athey, Susan, Julie Tibshirani, and Stefan Wager. 2019. "Generalized Random Forests." *Annals of Statistics*.
- Athey, Susan, and Stefan Wager. 2021. "Policy Learning with Observational Data." *Econometrica*.
- Chernozhukov, Victor, Iván Fernández-Val, and Blaise Melly. 2013. "Inference on Counterfactual Distributions." *Econometrica*.
- Chernozhukov, Victor, Christian Hansen, Nathan Kallus, Martin Spindler, and Vasilis Syrgkanis. 2026. *Applied Causal Inference Powered by ML and AI*, chapters 14 and 15.
- Childers, David. 2024. "Outcome and Effect Heterogeneity." Lecture notes.
- Facure, Matheus. 2023. *Causal Inference for the Brave and True*, chapters 18, 21, and 23.
- Foster, Dylan J., and Vasilis Syrgkanis. 2023. "Orthogonal Statistical Learning." *Annals of Statistics*.
- Genovese, Christopher, and Larry Wasserman. 2008. "Adaptive Confidence Bands." *Annals of Statistics*.
- Haushofer, Johannes, Paul Niehaus, Carlos Paramo, Edward Miguel, and Michael Walker. 2025. "Targeting Impact versus Deprivation." *American Economic Review*.
- Kennedy, Edward H. 2023. "Towards Optimal Doubly Robust Estimation of Heterogeneous Causal Effects." *Electronic Journal of Statistics*.
- Kitagawa, Toru, and Aleksey Tetenov. 2018. "Who Should Be Treated? Empirical Welfare Maximization Methods for Treatment Choice." *Econometrica*.
- Künzel, Sören R., Jasjeet S. Sekhon, Peter J. Bickel, and Bin Yu. 2019. "Metalearners for Estimating Heterogeneous Treatment Effects Using Machine Learning." *Proceedings of the National Academy of Sciences*.
- Lan, Hui, and Vasilis Syrgkanis. 2023. "Causal Q-Aggregation for CATE Model Selection." arXiv:2310.16945.
- Lecué, Guillaume, and Philippe Rigollet. 2014. "Optimal Learning with Q-Aggregation." *Annals of Statistics*.
- Nie, Xinkun, and Stefan Wager. 2021. "Quasi-Oracle Estimation of Heterogeneous Treatment Effects." *Biometrika*.
- Oprescu, Miruna, Vasilis Syrgkanis, and Zhiwei Steven Wu. 2019. "Orthogonal Random Forest for Causal Inference." *ICML*.
- Semenova, Vira, and Victor Chernozhukov. 2021. "Debiased Machine Learning of Conditional Average Treatment Effects and Other Causal Functions." *The Econometrics Journal*.
- Shalit, Uri, Fredrik D. Johansson, and David Sontag. 2017. "Estimating Individual Treatment Effect: Generalization Bounds and Algorithms." *ICML*.
- Sverdrup, Erik, Ayush Kanodia, Zhengyuan Zhou, Susan Athey, and Stefan Wager. 2020. "policytree: Policy Learning via Doubly Robust Empirical Welfare Maximization over Trees." *Journal of Open Source Software*.
- Wager, Stefan, and Susan Athey. 2018. "Estimation and Inference of Heterogeneous Treatment Effects Using Random Forests." *Journal of the American Statistical Association*.
- Yadlowsky, Steve, Scott Fleming, Nigam Shah, Emma Brunskill, and Stefan Wager. 2021. "Evaluating Treatment Prioritization Rules via Rank-Weighted Average Treatment Effects." arXiv:2111.07966.
- Yoganarasimhan, Hema, Ebrahim Barzegary, and Abhishek Pani. 2023. "Design and Evaluation of Optimal Free Trials." *Management Science*.
- Zhan, Ruohan, Zhimei Ren, Susan Athey, and Zhengyuan Zhou. 2024. "Policy Learning with Adaptively Collected Data." *Management Science*.
- Zhou, Junyi, Ying Zhang, and Wanzhu Tu. 2023. "A Reference-Free R-Learner for Treatment Recommendation." *Statistical Methods in Medical Research*.
