---
title: "Replication: First-Generation Elite"
subtitle: "The Role of School Social Networks (Cattan, Salvanes, and Tominey, AER 2025): an audit of the IV mediation design and a calibrated substitute for the registry data"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/10_mediation/cattan2025/"
  n_boot: 199
  n_sims: 100
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "90%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(causalmetrics)
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(fixest)
data_path <- params$data_path
n_boot <- params$n_boot
n_sims <- params$n_sims
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
fmt0 <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
paren <- function(x, digits = 3) paste0("(", fmt(x, digits), ")")
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
dofiles <- list.files(file.path(data_path, "DoFiles_Sept2025"), pattern = "\\.do$")
```

# The paper in one page {#sec-paper}

**The question is whether going to high school with the children of the
elite makes low-income students more likely to enter elite degrees, and
the answer is yes, through a channel that runs against a channel through
grades.** Norwegian registry data follow every student who finished
middle school between 2002 and 2010 into high school, university, and
the labour market. Exposure to elite peers is the share of a student's
high-school cohort whose parents hold an elite degree, standardized.
Within-school variation across cohorts identifies the effect, with
school, cohort, and program fixed effects. For low-SES students a one
standard deviation rise in exposure raises enrollment in an elite degree
by 1.1 percentage points (Table 2), about a fifth of the baseline rate.

**The mechanism section is an instrumental-variables mediation
analysis.** Elite peers lower grades (competition, or teacher grading),
and grades govern admission, so the total effect hides a negative
channel. To separate the channels the authors instrument the high-school
grade point average with a lottery: in the third year of high school
students are randomly assigned to a written mathematics exam, which is
graded blind and raises the GPA of those assigned. Table 4 reports the
first stage (GPA on peers and on the lottery), the second stage
(enrollment on peers and instrumented GPA), and the decomposition: the
direct effect is the second-stage peer coefficient $\gamma_1$, the
indirect effect is the first-stage peer coefficient times the
instrumented GPA effect, $\delta_1 \gamma_2$, and the total effect is
their sum. For low-SES students the direct effect (0.026) is larger than
the total (0.011) because the indirect effect through grades is negative
($-0.015$).

**Why this paper anchors the endogenous-mediator part of Section 10.**
The mediator is a choice-laden outcome, GPA, and the decomposition needs
an exogenous shifter of it. The lottery is an encouragement design for
the mediator. The extra assumption, that the effect of GPA on enrollment
is the same for the lottery's compliers as for the students whose grades
the peers move (Huber 2019), is what the paper's Appendix Figure A13
checks by re-estimating $\gamma_2$ across subgroups.

**What the replication package holds and does not hold.** The registry
data cannot be released; the package contains the complete do-files, the
auxiliary inputs (crosswalks, CPI series, subject codes, occupation
lists), and the README. There is no synthetic dataset. This note audits
the code path that produces Table 4, then builds a design calibrated to
the published Table 4 and runs `mediate_iv()` on it to reproduce the
table's layout, the homogeneity check, and the weak-instrument column.

# Audit of the replication package {#sec-audit}

**The package is a code map: `r length(dofiles)` do-files, no analysis
data.** `Master_Replication.do` sets the paths to a restricted
`replication_data` folder and calls the others in order.

```{r audit-table, echo = FALSE}
tibble(
  `Do-file` = c("Replication\\_ElitePeers.do, Analysis\\_Replication.do", "Analysis\\_IV\\_Replication.do", "FigureA13\\_Replication.do",
                "simulate\\_replication.do", "Analysis\\_Longrun\\_Replication.do, RankRank\\_Replication.do", "TeacherData\\_Replication.do, MeasurementError\\_Replication.do, Placebo\\_Replication.do"),
  Produces = c("Tables 1-3, Figure 1: exposure measure, total effects by SES", "Table 4 (IV mediation), Table A11 (lottery balance)",
               "Figure A13: gamma2 by subgroup (homogeneity check)", "Section VII: reallocation of low-SES pupils across schools and predicted mobility",
               "Tables 5-6: earnings and rank-rank mobility", "Appendix: teacher composition, measurement error, placebo cohorts"),
  `Missing input` = c("elitetemp.dta (analysis file from registries)", "tab\\_vitnemalfag\\_vgskole (exam registry), elitetemp.dta",
                      "elitetemp.dta", "elitetemp.dta, rankrank\\_Replication.dta", "earnings registries", "teacher and census registries")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "The replication package by output and by missing input. Every analysis file is built from Norwegian registries that the package cannot include.") %>%
  tab_style() %>% column_spec(1, width = "5.2cm") %>% column_spec(2, width = "5.5cm") %>% column_spec(3, width = "4cm")
```

**The Table 4 specification is explicit in `Analysis_IV_Replication.do`.**
The instrument `writtenmaths` is built from the exam registry as an
indicator for a written mathematics exam in the third year of a
three-year high school. Then, for the low-SES sample:

```
xtivreg2 elitedegree mean_pexecutive_schoolstd $controlsminus spec alternative_educ
    (gpa_tot_sd = writtenmaths) [pweight = meansize2] if parentedu == 1, fe first
```

with `xtset skole_orgnr1` (school fixed effects), cohort dummies and the
parental-education, gender, origin, middle-school GPA, and parental
occupation controls in `$controlsminus`, high-school program dummies
(`spec`, `alternative_educ`), and weights equal to the school size. The
same call on the high-SES sample gives columns 3 and 4, and `xtreg ...,
fe` without GPA gives the OLS columns. Three observations from the audit:

- The OLS columns of Table 4 are the total-effect regressions without
  GPA (the Table 2 specification on the IV sample), not OLS with GPA as a
  regressor. Panel C's "total effect" is that coefficient.
- The IV call carries `[pweight = meansize2]`, which makes `xtivreg2`
  report heteroskedasticity-robust standard errors; the table note says
  "clustered at the school level". The cluster option appears in the
  Table A11 regressions of the same file but not in the Table 4 line.
- Panel C reports no standard errors for the direct, indirect, and total
  effects. `mediate_iv()` bootstraps the whole system, which supplies
  them.

**`simulate_replication.do` is not a data generator.** It is the Section
VII exercise: it reallocates low-SES students across schools cohort by
cohort to raise their exposure, re-estimates the enrollment and earnings
regressions by SES with `parmest`, and predicts outcomes under the new
allocation. It reads `elitetemp.dta` on its first line. There is
therefore no simulated dataset in the package to run, and the design
below is this note's substitute.

# A design calibrated to Table 4 {#sec-design}

**The design keeps the structure of the paper's identification and the
magnitudes of its Table 4.** Schools with sizes drawn around the paper's
mean (58,304 pupils in 500 schools), nine cohorts, three programs,
school-cohort exposure to elite peers standardized across school-cohorts,
an unobserved ability that raises both GPA and enrollment (so that OLS
on GPA is biased), a binary lottery for the written exam with the
published first stage, and school-cohort shocks so that clustering by
school matters. The outcome is kept linear so that the coefficients are
the parameters of the design.

```{r design-def}
sim_cattan <- function(n_schools = 500, mean_size = 117, delta1 = -0.039, delta2 = 0.094,
                       gamma1 = 0.026, gamma2 = 0.384, kappa = 0.15, sd_sc = 0.25, sd_gpa = 1.3, sd_y = 0.4,
                       seed = 1) {
  set.seed(seed)
  size <- pmax(10, round(rgamma(n_schools, shape = 4, scale = mean_size / 4)))
  n <- sum(size)
  school <- rep(seq_len(n_schools), size)
  cohort <- sample(2002:2010, n, replace = TRUE)
  program <- sample(1:3, n, replace = TRUE)
  school_fe <- rnorm(n_schools, 0, 0.3)[school]
  cohort_fe <- (cohort - 2006) * 0.01
  key <- paste(school, cohort)
  sc_ids <- unique(key)
  peer_raw <- rnorm(length(sc_ids), 0.10, 0.06)[match(key, sc_ids)]
  peer <- (peer_raw - mean(peer_raw)) / sd(peer_raw)
  u_sc_gpa <- rnorm(length(sc_ids), 0, sd_sc)[match(key, sc_ids)]
  u_sc_y <- rnorm(length(sc_ids), 0, sd_sc / 4)[match(key, sc_ids)]
  female <- rbinom(n, 1, 0.5)
  ability <- rnorm(n)
  z <- rbinom(n, 1, 0.5)
  gpa <- school_fe + cohort_fe + delta1 * peer + delta2 * z + 0.1 * female + 0.5 * ability + u_sc_gpa + rnorm(n, 0, sd_gpa)
  y <- 0.15 + 0.5 * school_fe + cohort_fe + gamma1 * peer + gamma2 * gpa + kappa * ability + 0.02 * female + u_sc_y + rnorm(n, 0, sd_y)
  structure(data.frame(y = y, gpa = gpa, peer = peer, z = z, female = female, school = school, cohort = cohort,
                       program = program, size = size[school]),
            truth = list(direct = gamma1, indirect = delta1 * gamma2, total = gamma1 + delta1 * gamma2,
                         first_stage = delta1, mediator_effect = gamma2, lottery = delta2))
}
low <- sim_cattan(seed = 1)
high <- sim_cattan(n_schools = 409, mean_size = 49, delta1 = -0.010, delta2 = 0.043, gamma1 = 0.049, gamma2 = 0.384, seed = 2)
c(low_n = nrow(low), low_schools = length(unique(low$school)), high_n = nrow(high), high_schools = length(unique(high$school)))
```

**The high-SES design differs in three published ways.** Fewer pupils
(19,968 in 409 schools), a smaller first-stage effect of the lottery
(0.043 against 0.094, with an F of 5.6 against 76.9), and a larger direct
effect (0.049). The published $\gamma_2$ for high-SES students is 1.771
with a standard error of 0.714, which a weak instrument produces from the
same 0.384 as easily as from anything else; the design keeps 0.384 so
that the weak-instrument column can be read against a known truth.

# Table 4 in the published layout {#sec-table4}

**One call per SES group gives every entry of the table.** `mediate_iv()`
fits the first stage, the instrumented second stage, and the reduced
forms with `fixest`, with the school, cohort, and program fixed effects,
the school-size weights, and clustering by school, and bootstraps the
system by school for the decomposition's standard errors.

```{r table4-fit, cache = TRUE, cache.extra = list(n_boot)}
fit_low <- mediate_iv(low, "y", "peer", "gpa", z = "z", x = "female", fe = c("school", "cohort", "program"),
                      cluster = "school", weights = "size", n_boot = n_boot, seed = 1)
fit_high <- mediate_iv(high, "y", "peer", "gpa", z = "z", x = "female", fe = c("school", "cohort", "program"),
                       cluster = "school", weights = "size", n_boot = n_boot, seed = 1)
```

```{r table4-layout, echo = FALSE}
col_iv <- function(f, dat) {
  e <- f$effects; get <- function(t) e$estimate[e$term == t]; se <- function(t) e$std.error[e$term == t]
  fs <- coeftable(f$models$first_stage)
  iv <- coeftable(f$models$iv)
  c(fmt(fs["peer", 1]), paren(fs["peer", 2]), fmt(fs["z", 1]), paren(fs["z", 2]), fmt(f$first_stage_F, 2),
    fmt(get("direct")), paren(se("direct")), fmt(get("mediator_effect")), paren(se("mediator_effect")),
    paste0(fmt(get("direct")), " ", paren(se("direct"))), paste0(fmt(get("indirect")), " ", paren(se("indirect"))),
    paste0(fmt(get("total_z")), " ", paren(se("total_z"))),
    fmt0(f$n), fmt0(length(unique(dat$school))))
}
col_ols <- function(f, dat) {
  e <- f$effects
  ols <- coeftable(f$models$reduced_form)
  c("", "", "", "", "", fmt(ols["peer", 1]), paren(ols["peer", 2]), "", "", "", "", "", fmt0(f$n), fmt0(length(unique(dat$school))))
}
rows4 <- c("Proportion of parents with elite degree (std)", "", "Student took written math exam (IV)", "", "F-statistic",
           "Proportion of parents with elite degree (std)", "", "High school GPA", "",
           "Direct effect", "Indirect effect", "Total effect", "Number of pupils", "Number of schools")
pub4 <- tibble(` ` = rows4,
               `Low SES OLS` = c("", "", "", "", "", "0.011", "(0.004)", "", "", "", "", "", "58,304", "500"),
               `Low SES IV` = c("-0.039", "(0.007)", "0.094", "(0.011)", "76.86", "0.026", "(0.004)", "0.384", "(0.053)", "0.026", "-0.015", "0.011", "58,304", "500"),
               `High SES OLS` = c("", "", "", "", "", "0.032", "(0.009)", "", "", "", "", "", "19,968", "409"),
               `High SES IV` = c("-0.010", "(0.010)", "0.043", "(0.018)", "5.57", "0.049", "(0.019)", "1.771", "(0.714)", "0.049", "-0.018", "0.031", "19,968", "409"))
ours4 <- tibble(` ` = rows4, `Low SES OLS` = col_ols(fit_low, low), `Low SES IV` = col_iv(fit_low, low),
                `High SES OLS` = col_ols(fit_high, high), `High SES IV` = col_iv(fit_high, high))
ours4 %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcccc",
      caption = "Table 4 layout from the calibrated design: Panel A first stage (GPA on peers and the lottery, with the first-stage F), Panel B second stage (enrollment on peers and instrumented GPA; OLS columns are the total-effect regressions without GPA), Panel C the decomposition with school-bootstrap standard errors that the paper does not report. School, cohort, and program fixed effects; weights by school size; standard errors clustered by school.") %>%
  pack_rows("Panel A. First-stage outcome: high school GPA", 1, 5) %>%
  pack_rows("Panel B. Second-stage outcome: enrollment to elite degree", 6, 9) %>%
  pack_rows("Panel C. Decomposition", 10, 12) %>%
  tab_style("scale_down")
pub4 %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcccc", caption = "Table 4 as published (Cattan, Salvanes, and Tominey 2025), for comparison.") %>%
  pack_rows("Panel A. First-stage outcome: high school GPA", 1, 5) %>%
  pack_rows("Panel B. Second-stage outcome: enrollment to elite degree", 6, 9) %>%
  pack_rows("Panel C. Decomposition", 10, 12) %>%
  tab_style("scale_down")
```

```{r table4-numbers, echo = FALSE}
gl <- function(t) fit_low$effects$estimate[fit_low$effects$term == t]
sl <- function(t) fit_low$effects$std.error[fit_low$effects$term == t]
gh <- function(t) fit_high$effects$estimate[fit_high$effects$term == t]
sh <- function(t) fit_high$effects$std.error[fit_high$effects$term == t]
tr_low <- attr(low, "truth")
```

**The low-SES column reproduces the published structure with the
published magnitudes, and the decomposition now carries standard
errors.** The direct effect is `r fmt(gl("direct"))` (`r fmt(sl("direct"))`)
for a design value of `r fmt(tr_low$direct)`, the indirect effect
`r fmt(gl("indirect"))` (`r fmt(sl("indirect"))`) for `r fmt(tr_low$indirect)`,
and the total `r fmt(gl("total_z"))` for `r fmt(tr_low$total)`. The
first-stage F is `r fmt(fit_low$first_stage_F, 1)` against the paper's
76.9; both are the conventional F, which `xtivreg2` prints under `first`
and `fixest` reports as `ivf`, and the gap is a matter of how much
residual variance the design puts in GPA rather than a difference in
definition. The identity total = direct + indirect holds
exactly for the reduced form that includes the lottery, which is the
`total_z` row of the object; the `total` row without the lottery differs
by sampling error only.

**The high-SES column shows what a first-stage F of
`r fmt(fit_high$first_stage_F, 1)` does to the decomposition.** The
instrumented GPA effect is `r fmt(gh("mediator_effect"))` with a standard
error of `r fmt(sh("mediator_effect"))`, and the indirect effect's
interval covers zero and values several times the design's
`r fmt(attr(high, "truth")$indirect)`. The paper's 1.771 (0.714) is the
same phenomenon on the real data. The Anderson-Rubin set from Section
05's `iv_ar_confidence_set()` shows what a weak-instrument-robust
interval for $\gamma_2$ looks like in this column. The helper takes
controls but not fixed effects, so it is run on the high-SES design
without the school, cohort, and program effects; its point estimate
therefore differs from the table's, and the object of interest is the
shape of the set, which is unbounded.

```{r ar-set}
ar_high <- iv_ar_confidence_set(high, "y", "gpa", z = "z", x = c("peer", "female"), cluster = "school", weights = "size")
ar_high
```

# Figure A13: the homogeneity check {#sec-a13}

**The decomposition assumes that the lottery compliers and the students
whose grades the peers move share the same $\gamma_2$.** The paper checks
this by re-estimating the second stage in subgroups (Appendix Figure
A13). `mediate_iv(homogeneity_by = )` runs the system within each group.
In the design $\gamma_2$ is homogeneous by construction, so the figure
shows what the check looks like when the assumption holds: estimates
scattered around one value with intervals that widen with the group's
first stage.

```{r a13-fit, cache = TRUE, cache.extra = list(n_boot)}
low$cohort_group <- ifelse(low$cohort <= 2004, "2002-04", ifelse(low$cohort <= 2007, "2005-07", "2008-10"))
low$sex <- ifelse(low$female == 1, "female", "male")
low$program_f <- paste0("program ", low$program)
hom <- bind_rows(lapply(c("sex", "cohort_group", "program_f"), function(g) {
  h <- mediate_iv(low, "y", "peer", "gpa", z = "z", x = "female", fe = c("school", "cohort", "program"),
                  cluster = "school", weights = "size", homogeneity_by = g, n_boot = 0)$homogeneity
  # standard errors of the subgroup mediator effects from the subgroup IV fits
  se <- sapply(h$group, function(lev) {
    sub <- low[low[[g]] == lev, ]
    coeftable(feols(y ~ peer + female | school + cohort + program | gpa ~ z, data = sub, weights = ~size, cluster = ~school))["fit_gpa", 2]
  })
  mutate(h, dimension = g, se = se)
}))
```

```{r a13-plot, echo = FALSE, fig.height = 3.4, fig.cap = "Figure A13 analogue: the instrumented effect of GPA on enrollment by subgroup in the calibrated low-SES design, with 95 percent intervals clustered by school. The dashed line is the design's gamma2 = 0.384."}
hom %>% mutate(group = factor(group, levels = unique(group))) %>%
  ggplot(aes(x = group, y = mediator_effect, colour = dimension)) +
  geom_hline(yintercept = 0.384, linetype = "dashed", colour = "grey40") +
  geom_pointrange(aes(ymin = mediator_effect - 1.96 * se, ymax = mediator_effect + 1.96 * se)) +
  scale_colour_manual(values = c("#0072B2", "#D55E00", "#009E73"), name = NULL) +
  labs(x = NULL, y = "Effect of GPA on enrollment (IV)") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r a13-table, echo = FALSE}
hom %>% transmute(Dimension = dimension, Group = group, N = fmt0(n), `First stage (peers)` = fmt(first_stage), `GPA effect` = fmt(mediator_effect), `(se)` = paren(se), `Direct effect` = fmt(direct)) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "The homogeneity check as a table: subgroup first stages, instrumented GPA effects, and direct effects.") %>%
  tab_style()
```

**What a failure would look like.** If $\gamma_2$ differed between, say,
programs, the subgroup estimates would separate by more than their
intervals, and the product $\delta_1 \gamma_2$ would be a weighted
average whose weights depend on which students the lottery moves rather
than on which students the peers move. The paper's Figure A13 shows
estimates that are stable across gender, cohort, and program, and reads
that as support for the assumption. It is a necessary check, not a
sufficient one: homogeneity across observed groups does not imply
homogeneity across the unobserved compliance types.

# Sampling properties of the decomposition {#sec-mc}

**Repeated draws of the low-SES design show that the system bootstrap
is calibrated and that the weak-instrument column is not.** For each of
`r n_sims` draws the direct and indirect effects are estimated with
`mediate_iv()` without the bootstrap; the spread across draws is
compared with the bootstrap standard error from the first draw.

```{r mc, cache = TRUE, cache.extra = list(n_sims)}
mc <- bind_rows(lapply(seq_len(n_sims), function(s) {
  dl <- sim_cattan(seed = 100 + s)
  dh <- sim_cattan(n_schools = 409, mean_size = 49, delta1 = -0.010, delta2 = 0.043, gamma1 = 0.049, gamma2 = 0.384, seed = 500 + s)
  fl <- mediate_iv(dl, "y", "peer", "gpa", z = "z", x = "female", fe = c("school", "cohort", "program"), cluster = "school", weights = "size", n_boot = 0)$effects
  fh <- mediate_iv(dh, "y", "peer", "gpa", z = "z", x = "female", fe = c("school", "cohort", "program"), cluster = "school", weights = "size", n_boot = 0)$effects
  tibble(sim = s, sample = rep(c("low SES", "high SES"), each = 3), term = rep(c("direct", "indirect", "mediator_effect"), 2),
         estimate = c(fl$estimate[match(c("direct", "indirect", "mediator_effect"), fl$term)], fh$estimate[match(c("direct", "indirect", "mediator_effect"), fh$term)]))
}))
```

```{r mc-table, echo = FALSE}
truth_tab <- tibble(sample = rep(c("low SES", "high SES"), each = 3), term = rep(c("direct", "indirect", "mediator_effect"), 2),
                    truth = c(0.026, -0.039 * 0.384, 0.384, 0.049, -0.010 * 0.384, 0.384))
boot_se <- tibble(sample = rep(c("low SES", "high SES"), each = 3), term = rep(c("direct", "indirect", "mediator_effect"), 2),
                  boot_se = c(sapply(c("direct", "indirect", "mediator_effect"), sl), sapply(c("direct", "indirect", "mediator_effect"), sh)))
mc %>% group_by(sample, term) %>% summarise(mean = mean(estimate), sd = sd(estimate), median = median(estimate), .groups = "drop") %>%
  left_join(truth_tab, by = c("sample", "term")) %>% left_join(boot_se, by = c("sample", "term")) %>%
  mutate(sample = factor(sample, levels = c("low SES", "high SES"))) %>% arrange(sample, term) %>%
  transmute(Sample = sample, Effect = term, Truth = fmt(truth), `Mean across draws` = fmt(mean), `Median` = fmt(median), `SD across draws` = fmt(sd), `Bootstrap SE (draw 1)` = fmt(boot_se)) %>%
  kbl(booktabs = TRUE, linesep = "", caption = paste0("Sampling distribution of the decomposition over ", n_sims, " draws of each design, against the design values and the school-bootstrap standard errors of the first draw.")) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "top") %>%
  tab_style()
```

```{r mc-numbers, echo = FALSE}
mc_sum <- mc %>% group_by(sample, term) %>% summarise(mean = mean(estimate), sd = sd(estimate), median = median(estimate), .groups = "drop")
g_mc <- function(s, t, col) mc_sum[[col]][mc_sum$sample == s & mc_sum$term == t]
mc_sum_med <- function(s, t) mc_sum$median[mc_sum$sample == s & mc_sum$term == t]
```

**In the low-SES design the estimator is centred and the bootstrap
matches the sampling spread; in the high-SES design neither holds.** The
mean of the low-SES direct effect across draws is
`r fmt(g_mc("low SES", "direct", "mean"))` with a spread of
`r fmt(g_mc("low SES", "direct", "sd"))`, against a bootstrap standard
error of `r fmt(sl("direct"))`. The high-SES instrumented GPA effect is
roughly centred (mean `r fmt(g_mc("high SES", "mediator_effect", "mean"))`,
median `r fmt(mc_sum_med("high SES", "mediator_effect"))`, design value
0.384) but its spread is `r fmt(g_mc("high SES", "mediator_effect", "sd"))`,
ten times the low-SES one. The school bootstrap is not a remedy there:
the first draw's bootstrap standard errors (`r fmt(sh("direct"))` for the
direct effect, `r fmt(sh("mediator_effect"))` for the GPA effect) are many
times even that spread, because replicates in which the first stage lands
near zero produce arbitrarily large ratios. A weak first stage has to be
reported as such, which is what the paper's F of 5.6 does; its high-SES
decomposition (direct 0.049, indirect $-0.018$) should be read with that
column's first stage in mind.

# What the replication says {#sec-reading}

**Nothing in the paper's Table 4 can be recomputed without the
registries, and everything in its design can be.** The do-files fix the
specification exactly, and the calibrated design reproduces the layout,
the magnitudes, and the two features the paper relies on: a strong
lottery for low-SES students and a weak one for high-SES students.

**`mediate_iv()` is the paper's three regressions plus the standard
errors the paper leaves out.** The direct effect is the second-stage
peer coefficient, the indirect effect is the product, and the total
effect is the reduced form; the school bootstrap of the system gives the
decomposition's standard errors, and the homogeneity argument
re-estimates the system by subgroup.

**Two audit findings.** The Table 4 IV line carries probability weights
without a cluster option, so its reported standard errors are robust
rather than school-clustered as the note states; and the "simulated
data" one might hope for in `simulate_replication.do` is the Section VII
reallocation exercise, which needs the same registry file as everything
else.

## Mapping to the lecture's checklist

1. Estimands: total effect of peer exposure, direct effect holding GPA,
   indirect effect through GPA; identified by the lottery plus
   homogeneity.
2. Graph: peers to GPA to enrollment, peers to enrollment directly,
   ability confounding GPA and enrollment, lottery to GPA only.
3. Total effect and the effect on the mediator first: Table 2 and Panel
   A.
4. Natural effects need an instrumented mediator here: `mediate_iv()`.
5. Sensitivity: to the homogeneity assumption, via subgroups (Figure
   A13); to the exclusion of the lottery, argued from the blind grading.
6. Instrument: the written-exam lottery; weak in the high-SES sample.
7. No treatment-induced confounder is modelled.
8. No regression accounting.

# References

- Cattan, Sarah, Kjell G. Salvanes, and Emma Tominey. 2025. "First-Generation Elite: The Role of School Social Networks." *American Economic Review* 115(12), 4369-4403. Replication package: openICPSR.
- Dippel, Christian, Andreas Ferrara, and Stephan Heblich. 2020. "Causal Mediation Analysis in Instrumental-Variables Regressions." *Stata Journal*.
- Huber, Martin. 2019. "A Review of Causal Mediation Analysis for Assessing Direct and Indirect Treatment Effects." IZA Discussion Paper 12660.
