---
title: "Replication: LinkedIn(to) Job Opportunities"
subtitle: "Experimental Evidence from Job Readiness Training (Wheeler, Garlick, Johnson, Shaw, and Gargano, AEJ: Applied 2022)"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/10_mediation/wheeler2022/"
  n_boot_wild: 1000
  n_boot: 499
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "90%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(causalmetrics)
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(fixest)
library(mlr3)
library(mlr3learners)
data_path <- params$data_path
n_boot_wild <- params$n_boot_wild
n_boot <- params$n_boot
data_md5 <- tools::md5sum(file.path(data_path, "panel_clean_final.dta"))
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
fmt0 <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
paren <- function(x, digits = 3) paste0("(", fmt(x, digits), ")")
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
```

# The paper in one page {#sec-paper}

**The question is whether teaching work seekers to use LinkedIn raises
their employment, and the answer is a 7 percentage point gain that
persists for a year.** Harambee, a South African job-readiness program,
ran 30 training cohorts in 2016. Cohorts were paired into 14
stratification blocks by city and date, and one cohort per block was
randomized to receive a LinkedIn training module inside the program.
Treated candidates were 31 percentage points more likely to hold a
LinkedIn account and 7 percentage points more likely to be employed at
the end of the program, with effects of the same size 6 and 12 months
later.

**The identification is the randomization; the mechanism is the harder
question and the paper treats it as such.** The experiment identifies the
effect of the training on employment and the effect on every LinkedIn
measure (Table 2). It does not identify how much of the employment gain
runs through LinkedIn use, because nobody randomized LinkedIn use. The
paper assembles four pieces of evidence in Section IV: the nonexperimental
association between account ownership and employment (Table 4), effect
heterogeneity by baseline account (Table 5), effects on candidate
mechanisms (Table 6), and, in Appendix C, two "share explained"
statistics under stated assumptions. It says that none of this is
prespecified and that the assumptions are strong.

**Why this paper anchors Section 10.** It is a clean experiment whose
authors reach for mediation only with explicit caveats. The Appendix C
shares are the linear mediation model of the lecture, estimated as a
system with delta-method standard errors, and the mechanism tables are
the "effects on mediators" step that every mediation analysis should
report first. The public data reproduce all of it.

**What the public replication package holds.** `panel_clean_final.dta`
(1,638 candidates, one row each, with the three survey waves stacked as
variables), the Stata scripts for the paper and the appendices, the
auxiliary programs (wild bootstrap, extensive/intensive decomposition),
and the `.tex` and `.csv` artifacts of every table. The LinkedIn usage
components behind the intensive-margin indices were redacted by LinkedIn
before release, so four columns of Table 6 and the component detail of
Table 2 cannot be rebuilt. Every reproduced number below is printed next
to the published one.

# Data and specifications {#sec-data}

```{r load}
wh <- as.data.frame(haven::zap_labels(haven::read_dta(file.path(data_path, "panel_clean_final.dta"))))
wh$strata <- factor(wh$strata)
wh$cohort <- factor(wh$studref)
c(candidates = nrow(wh), treated = sum(wh$treated), cohorts = nlevels(wh$cohort), blocks = nlevels(wh$strata))
```

**Every treatment effect in the paper is one regression.** The outcome is
regressed on the treatment indicator with stratification-block fixed
effects, and standard errors are clustered by cohort (30 clusters). In
Stata this is `reghdfe y treated, absorb(strata) cluster(studref)`; here it
is one `fixest::feols` call. The helper below returns the coefficient, the
standard error, the control-group mean, and the sample sizes the paper's
tables report. `fixest`'s default small-sample correction matches
`reghdfe`'s for this design (block fixed effects are not nested within
cohorts), which the tables confirm to three decimals.

```{r itt-helper}
itt <- function(y, data = wh, cond = NULL) {
  d <- data[!is.na(data[[y]]), ]
  fit <- feols(as.formula(paste(y, "~ treated | strata")), data = d, cluster = ~studref)
  ct <- coeftable(fit)
  out <- tibble(outcome = y, estimate = ct["treated", "Estimate"], std.error = ct["treated", "Std. Error"],
                control_mean = mean(d[[y]][d$treated == 0]), n = nobs(fit), n_cohorts = length(unique(d$studref)),
                adj_r2 = unname(r2(fit, "ar2")))
  if (!is.null(cond)) out$control_mean_cond <- mean(d[[y]][d$treated == 0 & d[[cond]] == 1], na.rm = TRUE)
  out
}
```

**Covariates enter only in Table 4 and in our additions, with the paper's
missing-value convention.** The do-file replaces missing education dummies
by zero and, for every other covariate with missing values, adds a
missing indicator and a zero-filled copy. We build the same list.

```{r covariates}
covs <- c("age", "gender_composite", "prev_employ", "cog_num", "cog_comm", "cog_cft",
          "edn_dummy3", "edn_dummy4", "region1", "region2")
covs_miss <- character(0)
for (v in covs) {
  if (v %in% c("edn_dummy3", "edn_dummy4")) {
    wh[[v]][is.na(wh[[v]])] <- 0
    covs_miss <- c(covs_miss, v)
  } else if (anyNA(wh[[v]])) {
    wh[[paste0(v, "_miss")]] <- as.integer(is.na(wh[[v]]))
    wh[[paste0(v, "_nomiss")]] <- ifelse(is.na(wh[[v]]), 0, wh[[v]])
    covs_miss <- c(covs_miss, paste0(v, "_miss"), paste0(v, "_nomiss"))
  } else {
    covs_miss <- c(covs_miss, v)
  }
}
covs_miss
```

# Table 1: sample characteristics and balance {#sec-table1}

**Table 1 establishes that randomization produced comparable arms, and the
paper's balance test is a regression.** For each characteristic the
do-file regresses the variable on the two arm indicators without a
constant, clustered by cohort, and tests their equality; that is the
clustered t-test of the treatment coefficient. The standardized
difference divides the arm gap by the pooled standard deviation. Two
characteristics are cohort-level (size and completion rate), and their
tests run on 30 cohort observations.

```{r table1}
table1_row <- function(v, label, binary = FALSE, cohort_level = FALSE) {
  if (cohort_level) {
    d <- wh %>% group_by(studref) %>% summarise(x = first(.data[[v]]), treated = first(treated), .groups = "drop")
    fit <- lm(x ~ treated, data = d)
    p <- coeftable(feols(x ~ treated, data = d, vcov = "hetero"))["treated", "Pr(>|t|)"]
  } else {
    d <- wh[!is.na(wh[[v]]), ]; d$x <- d[[v]]
    p <- coeftable(feols(x ~ treated, data = d, cluster = ~studref))["treated", "Pr(>|t|)"]
  }
  q <- quantile(d$x, c(0.1, 0.9))
  tibble(Variable = label, N = nrow(d), Mean = mean(d$x), `Std. dev.` = sd(d$x),
         `10th` = if (binary) NA else q[[1]], `90th` = if (binary) NA else q[[2]],
         `p-value` = p, `Std. diff.` = (mean(d$x[d$treated == 1]) - mean(d$x[d$treated == 0])) / sd(d$x))
}
t1 <- bind_rows(
  table1_row("age", "Age"), table1_row("cog_num", "Numeracy score"), table1_row("cog_comm", "Communications score"),
  table1_row("cog_cft", "Cognitive score"), table1_row("gender_composite", "Gender", binary = TRUE),
  table1_row("edn_over12", "High school education", binary = TRUE),
  table1_row("edn_postsecond", "Post-secondary education", binary = TRUE),
  table1_row("edn_uni", "University education", binary = TRUE), table1_row("prev_employ", "Previously employed", binary = TRUE),
  table1_row("cohort_size", "Size of cohort", cohort_level = TRUE),
  table1_row("bridge_p_complete", "Program completion rate", cohort_level = TRUE))
published1 <- read.csv(file.path(data_path, "output", "summstats.csv"), skip = 1, header = TRUE, check.names = FALSE)
t1$`Published p` <- published1[["p-value"]][match(t1$Variable, trimws(published1$Variable))]
t1$`Published std. diff.` <- published1[["Std Diff"]][match(t1$Variable, trimws(published1$Variable))]
t1 %>% mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  kbl(booktabs = TRUE, linesep = "", digits = 2, caption = "Table 1 of the paper: sample characteristics, balance p-values, and standardized differences, with the published p-values and standardized differences in the last two columns.") %>%
  tab_style("scale_down")
```

**The reproduced p-values and standardized differences match the
published ones, including the one communication-score imbalance at
p = 0.05.** The paper reads that imbalance as chance among eleven tests
and returns to it in Appendix B7 with heterogeneity by communication
skill. Balance on baseline employment (previously employed, 38 percent)
matters most for what follows, because previous employment is the
strongest predictor of both LinkedIn use and later employment (Table 4).

# Table 2: effects on LinkedIn account opening and usage {#sec-table2}

**Table 2 is the first stage of any mechanism story: the training moved
LinkedIn use, by a lot.** Three outcomes: any account between baseline
and 12 months, an account opened during training, and the usage index
(first principal component of ten usage measures, standardized in the
control group). The usage index rows also split the effect into an
extensive margin (more accounts, valued at the control-group usage of
account holders) and an intensive margin (more use per account). The
decomposition needs only the account regression, the index regression,
and the control mean of the index among account holders, so the shares
reproduce from the public data even though the ten components do not.

```{r table2}
t2 <- bind_rows(itt("li_account"), itt("account_in_bridge"), itt("summary_li_pca", cond = "li_account"))
ext <- t2$estimate[1] * t2$control_mean_cond[3]
shares <- c(extensive = ext / t2$estimate[3], intensive = (t2$estimate[3] - ext) / t2$estimate[3])
pub2 <- list(est = c(0.314, 0.395, 0.935), se = c(0.049, 0.048, 0.145), shares = c(0.319, 0.682),
             mean = c(0.484, 0.091, 0.000), cond = 0.950, n = c(1638, 1579, 1300), r2 = c(0.140, 0.260, 0.187))
tibble(
  ` ` = c("Treated cohort", "", "Share of treatment effect due to more accounts", "Share due to using accounts more",
          "Control group mean", "Control mean | account", "Number of respondents", "Number of cohorts", "Adjusted R2"),
  `(1) LinkedIn account` = c(fmt(t2$estimate[1]), paren(t2$std.error[1]), "", "", fmt(t2$control_mean[1]), "", fmt0(t2$n[1]), t2$n_cohorts[1], fmt(t2$adj_r2[1])),
  `(2) Account during training` = c(fmt(t2$estimate[2]), paren(t2$std.error[2]), "", "", fmt(t2$control_mean[2]), "", fmt0(t2$n[2]), t2$n_cohorts[2], fmt(t2$adj_r2[2])),
  `(3) LI usage index` = c(fmt(t2$estimate[3]), paren(t2$std.error[3]), fmt(shares[1]), fmt(shares[2]), fmt(abs(t2$control_mean[3])), fmt(t2$control_mean_cond[3]), fmt0(t2$n[3]), t2$n_cohorts[3], fmt(t2$adj_r2[3])),
  `Published (1)` = c(fmt(pub2$est[1]), paren(pub2$se[1]), "", "", fmt(pub2$mean[1]), "", fmt0(pub2$n[1]), 30, fmt(pub2$r2[1])),
  `Published (3)` = c(fmt(pub2$est[3]), paren(pub2$se[3]), fmt(pub2$shares[1]), fmt(pub2$shares[2]), fmt(pub2$mean[3]), fmt(pub2$cond), fmt0(pub2$n[3]), 30, fmt(pub2$r2[3]))
) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lccccc",
      caption = "Table 2 of the paper: treatment effects on LinkedIn account opening and usage, block fixed effects, standard errors clustered by cohort; columns 1 and 3 of the published table for comparison.") %>%
  tab_style("scale_down")
```

**All three coefficients, their standard errors, the control means, and
the two shares reproduce.** The usage index effect of `r fmt(t2$estimate[3])`
standard deviations is the number the Appendix C shares divide by. About
a third of it is more accounts and two thirds is more use per account,
which the paper reads as candidates learning to use the platform rather
than being pushed to open a dormant profile.

# Figure 1: employment effects, with wild bootstrap p-values {#sec-figure1}

**Figure 1 is the paper's headline: employment rises by 7 percentage
points at the end of the program and stays up at 6 and 12 months.** The
figure plots the control mean and the treated mean (control mean plus the
fixed-effect ITT) at 0, 6, and 12 months with 95 percent intervals. With
30 clusters the paper adds wild cluster bootstrap p-values (Cameron,
Gelbach, and Miller 2008), computed by the `wildboot` program in the
replication package: Rademacher weights drawn once per cohort, 1,000
replications, the null imposed.

**The wild bootstrap is defined here, in the note, not in the package, in
the form the authors' program uses.** Their `wildboot` program fits the
unrestricted model with block fixed effects, forms the fitted values with
the treatment coefficient set to zero and keeps the unrestricted
residuals, then rebuilds the outcome in each replication by flipping the
sign of each cohort's residuals with probability one half and refits. The
p-value is the share of bootstrap t-statistics at least as large in
absolute value as the observed clustered t. The textbook version
(Cameron, Gelbach, and Miller 2008; Roodman et al. 2019) takes both the
fitted values and the residuals from the restricted regression; the
function offers both.

**One detail of the shipped program matters for the published numbers.**
Inside `wildboot` the observed t-statistic uses the clustered standard
error passed as an argument, but the bootstrap regressions reference a
local macro named `myvce` that the program never defines, so each
replication's t-statistic is computed with conventional standard errors.
The `boot_vcov` argument reproduces that behaviour with `"iid"`; the
clustered replications are `"cluster"`.

```{r wild-boot-def}
wild_cluster_boot_t <- function(data, y, d = "treated", fe = "strata", cluster = "studref",
                                n_boot = 1000, seed = 1, residuals = c("authors", "restricted"),
                                boot_vcov = c("cluster", "iid")) {
  residuals <- match.arg(residuals)
  boot_vcov <- match.arg(boot_vcov)
  dat <- data[complete.cases(data[, c(y, d, fe, cluster)]), ]
  cl_f <- as.formula(paste0("~", cluster))
  fit <- feols(as.formula(paste(y, "~", d, "|", fe)), data = dat, cluster = cl_f)
  b <- coef(fit)[[d]]
  t_obs <- coeftable(fit)[d, "t value"]
  if (residuals == "authors") {
    yhat <- fitted(fit) - b * dat[[d]]
    ehat <- resid(fit)
  } else {
    fit0 <- feols(as.formula(paste(y, "~ 1 |", fe)), data = dat)
    yhat <- fitted(fit0)
    ehat <- resid(fit0)
  }
  cl <- factor(dat[[cluster]])
  G <- nlevels(cl)
  vc <- if (boot_vcov == "cluster") cl_f else "iid"
  set.seed(seed)
  t_star <- vapply(seq_len(n_boot), function(r) {
    s <- sample(c(-1, 1), G, replace = TRUE)[as.integer(cl)]
    dat$.ystar <- yhat + s * ehat
    fb <- feols(as.formula(paste(".ystar ~", d, "|", fe)), data = dat, vcov = vc)
    coeftable(fb)[d, "t value"]
  }, numeric(1))
  list(estimate = b, std.error = coeftable(fit)[d, "Std. Error"], t = t_obs,
       p_wild = mean(abs(t_star) >= abs(t_obs)), p_cluster = coeftable(fit)[d, "Pr(>|t|)"],
       n_boot = n_boot, n_clusters = G)
}
```

```{r figure1-fit, cache = TRUE, cache.extra = list(data_md5, n_boot_wild)}
emp <- c("employed1_all", "employed2_all", "employed3_all")
f1 <- bind_rows(lapply(seq_along(emp), function(k) {
  y <- emp[k]
  d <- wh[!is.na(wh[[y]]), ]
  b0 <- coef(feols(as.formula(paste(y, "~ treated")), data = d, cluster = ~studref))[["(Intercept)"]]
  r <- itt(y)
  wa <- wild_cluster_boot_t(wh, y, n_boot = n_boot_wild, seed = 1, residuals = "authors", boot_vcov = "cluster")
  wi <- wild_cluster_boot_t(wh, y, n_boot = n_boot_wild, seed = 1, residuals = "authors", boot_vcov = "iid")
  wr <- wild_cluster_boot_t(wh, y, n_boot = n_boot_wild, seed = 1, residuals = "restricted", boot_vcov = "cluster")
  tibble(month = c(0, 6, 12)[k], outcome = y, control = b0, estimate = r$estimate, std.error = r$std.error,
         treated = b0 + r$estimate, p_cluster = wa$p_cluster, p_wild_authors = wa$p_wild, p_wild_iid = wi$p_wild, p_wild_restricted = wr$p_wild)
}))
```

```{r figure1, echo = FALSE, fig.height = 3.6, fig.cap = "Figure 1 of the paper: employment status by survey wave. Red circles are control-group means; blue squares are the control mean plus the block-fixed-effect treatment effect, with 95 percent intervals from standard errors clustered by cohort."}
ggplot(f1, aes(x = month)) +
  geom_errorbar(aes(ymin = treated - 1.96 * std.error, ymax = treated + 1.96 * std.error), width = 0.6, colour = "#0072B2") +
  geom_point(aes(y = control), colour = "#D55E00", shape = 16, size = 3) +
  geom_point(aes(y = treated), colour = "#0072B2", shape = 15, size = 3) +
  scale_x_continuous(breaks = c(0, 6, 12), labels = c("End of program", "6 months", "12 months")) +
  labs(x = NULL, y = "Employment rate") +
  theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
```

```{r figure1-table, echo = FALSE}
f1 %>%
  transmute(Wave = c("End of program", "6 months", "12 months"), Estimate = fmt(estimate), `Std. error` = paren(std.error),
            `Cluster p` = fmt(p_cluster), `Wild p, clustered` = fmt(p_wild_authors), `Wild p, restricted` = fmt(p_wild_restricted),
            `Wild p, as shipped` = fmt(p_wild_iid),
            Published = c("0.070 (0.021) [<0.001]", "0.081 (0.039) [0.138]", "0.069 (0.024) [<0.001]")) %>%
  kbl(booktabs = TRUE, linesep = "", caption = paste0("Employment effects behind Figure 1 with cluster-robust and wild cluster bootstrap p-values (", fmt0(n_boot_wild), " Rademacher replications, 30 cohorts). 'Clustered' uses the authors' residual construction with clustered t-statistics in the replications; 'restricted' uses restricted residuals; 'as shipped' reproduces the program's conventional bootstrap t-statistics. The published column quotes the figure note: estimate (standard error) [wild bootstrap p-value].")) %>%
  tab_style("scale_down")
```

**The point estimates and standard errors reproduce, and the published
p-values are reproduced only by the program as shipped.** With clustered
t-statistics in the replications, the wild bootstrap p-values are
`r fmt(f1$p_wild_authors[1])`, `r fmt(f1$p_wild_authors[2])`, and
`r fmt(f1$p_wild_authors[3])`; with the conventional standard errors the
program actually uses they are `r fmt(f1$p_wild_iid[1])`,
`r fmt(f1$p_wild_iid[2])`, and `r fmt(f1$p_wild_iid[3])`, which are the
published "<0.001", 0.138, and "<0.001" up to Monte Carlo error. The
mixed statistic is not a valid bootstrap-t (the observed and bootstrap
t-statistics are on different scales), and in this design it understates
the p-values at the end of the program and at 12 months. The conclusion
does not change: the end-of-program and 12-month effects are significant
by every version, and the 6-month effect is borderline by every version.
The restricted-residual version gives `r fmt(f1$p_wild_restricted[1])`,
`r fmt(f1$p_wild_restricted[2])`, and `r fmt(f1$p_wild_restricted[3])`.

# Table 3: effects on job type {#sec-table3}

**Table 3 asks whether the new jobs are different jobs.** Hours worked
(winsorized), currently employed in the first job, more than one
employer, permanent contract, and promotion, at 6 and 12 months. The
outcomes are coded zero for the non-employed, so the conditional control
means (among the employed) show whether the effect is more jobs or
different jobs. The paper concludes it is more jobs: hours rise because
employment rises, not because the employed work longer (Appendix B3
decomposes this).

```{r table3}
jobtype <- c("hourswork_wins", "employ_1tcurrent", "numemployers_bin", "permemploy", "promoted")
labels3 <- c("Hours worked", "Employed in first job", "More than one employer", "Permanent contract", "Promoted")
t3a <- bind_rows(lapply(paste0("tr_", jobtype), itt, cond = "employed2_all"))
t3b <- bind_rows(lapply(paste0("tr_w2_", jobtype), itt, cond = "employed3_all"))
panel <- function(t, published) {
  tibble(` ` = c("Treated cohort", "", "Control group mean", "Control mean | employment", "Number of respondents", "Adjusted R2", "Published estimate", "Published SE"),
         !!!setNames(lapply(seq_len(nrow(t)), function(j) c(fmt(t$estimate[j]), paren(t$std.error[j]), fmt(t$control_mean[j]), fmt(t$control_mean_cond[j]), fmt0(t$n[j]), fmt(t$adj_r2[j]), fmt(published$est[j]), paren(published$se[j]))), labels3))
}
pub3a <- list(est = c(4.200, 0.107, 0.001, 0.026, 0.007), se = c(1.700, 0.040, 0.021, 0.026, 0.010))
pub3b <- list(est = c(2.879, 0.126, -0.044, 0.034, -0.023), se = c(1.029, 0.026, 0.025, 0.025, 0.021))
bind_rows(panel(t3a, pub3a), panel(t3b, pub3b)) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Table 3 of the paper: treatment effects on job type six months (Panel A) and twelve months (Panel B) after the program, with the published estimates in the last two rows of each panel.") %>%
  pack_rows("Panel A: six months after program completion", 1, 8) %>%
  pack_rows("Panel B: twelve months after program completion", 9, 16) %>%
  tab_style("scale_down")
```

**Every coefficient and standard error in both panels reproduces.** Hours
rise by about four per week at six months; the only job-quality outcome
that moves is being in the first job at twelve months, which is retention
rather than quality.

# Table 4: the nonexperimental association {#sec-table4}

**Table 4 is the first mediation-style piece of evidence, and the paper
labels it nonexperimental.** Among candidates, does having a LinkedIn
account at baseline predict previous employment, and, in the control
group, employment at the end of the program? The regressions are plain
OLS with heteroskedasticity-robust standard errors; columns 2 and 4 add
the covariates. In the language of Section 10, the column 4 coefficient
is the delta of the Appendix C shares for a different mediator (the
baseline account rather than the account by 12 months): a
mediator-outcome association in the untreated, whose causal reading needs
the mediator to be as good as random given the covariates.

```{r table4}
covs_noemp <- setdiff(covs_miss, c("prev_employ_miss", "prev_employ_nomiss"))
ctrl <- wh[wh$treated == 0, ]
fits4 <- list(
  feols(prev_employ ~ s1_linkedin, data = wh, vcov = "hetero"),
  feols(as.formula(paste("prev_employ ~ s1_linkedin +", paste(covs_noemp, collapse = " + "))), data = wh, vcov = "hetero"),
  feols(employed1_all ~ s1_linkedin, data = ctrl, vcov = "hetero"),
  feols(as.formula(paste("employed1_all ~ s1_linkedin +", paste(covs_miss, collapse = " + "))), data = ctrl, vcov = "hetero"))
t4 <- sapply(fits4, function(f) c(coeftable(f)["s1_linkedin", 1:2], n = nobs(f), r2 = unname(r2(f, "ar2"))))
tibble(` ` = c("LI account at baseline", "", "Covariates included?", "Number of respondents", "Adjusted R2", "Published"),
       `(1) Previously employed` = c(fmt(t4[1, 1]), paren(t4[2, 1]), "", fmt0(t4[3, 1]), fmt(t4[4, 1]), "0.130 (0.035)"),
       `(2)` = c(fmt(t4[1, 2]), paren(t4[2, 2]), "Y", fmt0(t4[3, 2]), fmt(t4[4, 2]), "0.090 (0.035)"),
       `(3) Employed at end of program` = c(fmt(t4[1, 3]), paren(t4[2, 3]), "", fmt0(t4[3, 3]), fmt(t4[4, 3]), "0.068 (0.044)"),
       `(4)` = c(fmt(t4[1, 4]), paren(t4[2, 4]), "Y", fmt0(t4[3, 4]), fmt(t4[4, 4]), "0.035 (0.049)")) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Table 4 of the paper: nonexperimental relationship between employment and a LinkedIn account at baseline (columns 3-4 use control-group candidates only), heteroskedasticity-robust standard errors.") %>%
  tab_style("scale_down")
```

**The association is positive and shrinks by a third to a half once
covariates enter, which is the paper's own warning about selection into
LinkedIn use.** The end-of-program association in the control group
(column 4, `r fmt(t4[1, 4])`) is not significant. The lecture's
sensitivity analysis in Section \@ref(sec-beyond) puts a number on how
much residual selection would erase the mediated share.

# Table 5: heterogeneity by baseline account {#sec-table5}

**Table 5 asks whether the training helped those who already had an
account, and finds the effect concentrated among those who did not.** The
specification interacts the treatment with the baseline account indicator
and interacts the block fixed effects with it too, so each group has its
own randomization strata. The published table reports the effect for
candidates without an account (the main coefficient) and the difference
for those with one. `reghdfe` drops singleton fixed-effect cells, which
`fixest` does with `fixef.rm = "singleton"`; one cell is a singleton in
each wave, hence the sample sizes.

```{r table5}
wh$fe_times_li <- interaction(wh$s1_linkedin, wh$strata, drop = TRUE)
t5 <- bind_rows(lapply(emp, function(y) {
  d <- wh[!is.na(wh[[y]]) & !is.na(wh$s1_linkedin), ]
  fit <- feols(as.formula(paste(y, "~ treated + treated:s1_linkedin | fe_times_li")), data = d, cluster = ~studref, fixef.rm = "singleton")
  ct <- coeftable(fit)
  tibble(outcome = y, b_treat = ct["treated", 1], se_treat = ct["treated", 2], b_int = ct["treated:s1_linkedin", 1], se_int = ct["treated:s1_linkedin", 2],
         mean_acc = mean(d[[y]][d$treated == 0 & d$s1_linkedin == 1]), mean_noacc = mean(d[[y]][d$treated == 0 & d$s1_linkedin == 0]),
         n = nobs(fit), adj_r2 = unname(r2(fit, "ar2")))
}))
pub5 <- list(b = c(0.083, 0.099, 0.087), sb = c(0.023, 0.041, 0.026), i = c(-0.031, -0.106, -0.067), si = c(0.030, 0.067, 0.051))
tibble(` ` = c("Treated cohort", "", "... x baseline LinkedIn account", "", "Control group mean | account", "Control group mean | no account", "Number of respondents", "Adjusted R2", "Published: treated", "Published: interaction"),
       !!!setNames(lapply(1:3, function(j) c(fmt(t5$b_treat[j]), paren(t5$se_treat[j]), fmt(t5$b_int[j]), paren(t5$se_int[j]), fmt(t5$mean_acc[j]), fmt(t5$mean_noacc[j]), fmt0(t5$n[j]), fmt(t5$adj_r2[j]),
                                               paste(fmt(pub5$b[j]), paren(pub5$sb[j])), paste(fmt(pub5$i[j]), paren(pub5$si[j])))), c("(1) End of program", "(2) 6 months", "(3) 12 months"))) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Table 5 of the paper: employment effects by prior LinkedIn account, block fixed effects interacted with the baseline-account indicator, standard errors clustered by cohort.") %>%
  tab_style()
```

**The interaction is negative in every wave and never significant, which
the paper reads as the training working through the account margin.** In
the language of Section 08, Table 5 is a group average treatment effect
by one baseline covariate. The package estimates the same two group
effects from a doubly robust score with the blocks and covariates as
controls, without a linear interaction model.

```{r table5-gate}
d5 <- wh[!is.na(wh$employed1_all) & !is.na(wh$s1_linkedin), ]
d5$account_baseline <- factor(ifelse(d5$s1_linkedin == 1, "account", "no account"))
sc5 <- dr_scores(d5, "employed1_all", "treated", x = c("strata", covs_miss), seed = 1)
g5 <- cate_gate(sc5, groups = "account_baseline", n_boot = 499, seed = 1)
g5$table
```

```{r table5-gate-numbers, echo = FALSE}
g5t <- g5$table
gate_acc <- g5t$estimate[g5t$group == "account"]
gate_noacc <- g5t$estimate[g5t$group == "no account"]
gate_se_acc <- g5t$std.error[g5t$group == "account"]
reg_acc <- t5$b_treat[1] + t5$b_int[1]
```

**Both approaches put the effect among candidates without an account and
neither can resolve the effect among those with one.** The score gives
`r fmt(gate_noacc)` for the `r fmt0(g5t$n[g5t$group == "no account"])`
candidates without a baseline account, close to the regression's
`r fmt(t5$b_treat[1])`. For the `r fmt0(g5t$n[g5t$group == "account"])`
with an account it gives `r fmt(gate_acc)` against the regression's
implied `r fmt(reg_acc)`, a gap well inside the group's standard error of
`r fmt(gate_se_acc)`. That group is a sixth of the sample, so neither
estimate says much about it; the regression borrows strength across
groups through its functional form and the score does not.

# Table 6: effects on potential mechanisms {#sec-table6}

**Table 6 is the "effects on the mediators" step, and the paper runs it for
ten candidate mechanisms.** Supply-side information (jobs and profiles
viewed), placement in the targeted sector, placement through a partner
firm, demand-side information (profile completeness and views), number of
connections, job applications on the platform, self-placement, program
completion, an engagement index, and an aspirations index. Each is an ITT
with block fixed effects. Four of the ten (the LinkedIn usage indices and
platform applications) were built from data LinkedIn redacted; the six
others reproduce.

```{r table6}
mech <- c("targetsector", "harambee_partner_original", "selfplaced", "s2_completed", "summary_engage", "summary_aspire2")
labels6 <- c("(2) Employed in target sector", "(3) Employed at partner firm", "(7) Self-placed", "(8) Completed program", "(9) Engagement index", "(10) Aspirations index")
t6 <- bind_rows(lapply(mech, itt))
pub6 <- list(est = c(0.155, -0.024, -0.013, -0.023, 0.151, 0.065), se = c(0.034, 0.059, 0.009, 0.030, 0.122, 0.054),
             mean = c(0.539, 0.342, 0.056, 0.882, 0.000, 0.000), n = c(1626, 1626, 1626, 1612, 1246, 1231))
tibble(` ` = c("Treated cohort", "", "Control group mean", "Number of respondents", "Number of cohorts", "Adjusted R2", "Published estimate", "Published SE", "Published N"),
       !!!setNames(lapply(seq_along(mech), function(j) c(fmt(t6$estimate[j]), paren(t6$std.error[j]), fmt(t6$control_mean[j]), fmt0(t6$n[j]), t6$n_cohorts[j], fmt(t6$adj_r2[j]), fmt(pub6$est[j]), paren(pub6$se[j]), fmt0(pub6$n[j]))), labels6)) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Table 6 of the paper, the six columns that the public data support: treatment effects on potential mechanisms with block fixed effects and cohort-clustered standard errors. Columns 1, 4, 5, and 6 of the published table (supply-side information 0.316 (0.073), demand-side information 0.842 (0.144), connections 0.559 (0.093), job applications 0.009 (0.004)) use redacted LinkedIn data.") %>%
  tab_style("scale_down")
```

**The six reproduced columns match the paper, and the pattern is the
paper's argument for the mechanism.** Treated candidates are 15
percentage points more likely to be placed in the targeted sector
(finance and call centres), with no change in partner-firm placement,
self-placement, completion, engagement, or aspirations. Together with the
redacted columns (large effects on information and connections, none on
on-platform applications), the paper concludes that LinkedIn worked as an
information and signalling tool inside the program's own placement
channel, not as a job board.

# Appendix Table C1: the mediation shares {#sec-tableC1}

**Table C1 converts the two first stages into a share of the employment
effect explained by LinkedIn use, under two explicit assumptions.** For a
mediator $M$ (the account by 12 months, or the usage index) the do-file
estimates four regressions as a system with `suest` and cohort clustering:
employment on treatment ($\beta$), $M$ on treatment ($\gamma$),
employment on treatment and $M$ ($\tilde\beta$, $\alpha$), and employment
on $M$ in the control group ($\delta$). The shares are
$S_1 = \alpha\gamma/\beta$ and $S_2 = \delta\gamma/\beta$ with delta-method
standard errors from the system covariance. $S_1$ assumes that the
mediator-outcome relation is unconfounded given treatment and block; $S_2$
replaces the mediator coefficient by its control-group value, which the
paper motivates as "treatment increases LinkedIn use but does not change
the relationship between LinkedIn use and employment".

**`mediate_reg()` is this system.** With `method = "delta"` it fits the
same four regressions, computes the joint covariance from the stacked
influence functions clustered by cohort (what `suest` does, with its
$G/(G-1)$ cluster correction and no degrees-of-freedom factor), and
reports the shares with their standard errors in `shares`. Block fixed
effects enter as a factor covariate. Panel A's standard errors below are
read from the same stacked covariance, so that they match `suest` rather
than `reghdfe` (which adds the $(n-1)/(n-k)$ factor and gives 0.021
instead of 0.020 for $\beta$).

```{r tableC1}
c1_fit <- function(m) {
  d <- wh[complete.cases(wh[, c("employed1_all", "treated", m)]), ]
  mediate_reg(d, "employed1_all", "treated", m, x = "strata", method = "delta", cluster = "studref")
}
fit_acc <- c1_fit("li_account")
fit_idx <- c1_fit("summary_li_pca")
c1_col <- function(f) {
  m <- f$spec$m
  c(fmt(coef(f$models$total)[["treated"]]), paren(sqrt(vcov_of(f, "total"))),
    fmt(coef(f$models$mediator[[1]])[["treated"]]), paren(sqrt(vcov_of(f, "mediator"))),
    fmt(coef(f$models$outcome)[["treated"]]), paren(sqrt(vcov_of(f, "outcome_d"))),
    fmt(coef(f$models$outcome)[[m]]), paren(sqrt(vcov_of(f, "outcome_m"))),
    fmt(coef(f$models$control)[[m]]), paren(sqrt(vcov_of(f, "control"))),
    fmt(f$shares$estimate[2]), paren(f$shares$std.error[2]), fmt(f$shares$estimate[3]), paren(f$shares$std.error[3]), fmt0(f$n))
}
vcov_of <- function(f, which) {
  # suest's cluster-robust covariance: G/(G-1) only, no (n-1)/(n-k) factor
  rows <- complete.cases(wh[, c("employed1_all", "treated", f$spec$m)])
  cl <- wh$studref[rows]
  n <- sum(rows)
  m <- f$spec$m
  se2 <- function(model, term, subset = seq_len(n)) {
    infl <- causalmetrics:::.cm_infl(model, n, subset)
    causalmetrics:::.cm_stacked_vcov(infl, cl)[term, term]
  }
  switch(which,
    total = se2(f$models$total, "treated"),
    mediator = se2(f$models$mediator[[1]], "treated"),
    outcome_d = se2(f$models$outcome, "treated"),
    outcome_m = se2(f$models$outcome, m),
    control = se2(f$models$control, m, which(wh$treated[rows] == 0)))
}
rows_c1 <- c("Treatment effect on employment (beta)", "", "Treatment effect on LinkedIn use (gamma)", "",
             "Treatment effect on employment | LinkedIn use (beta tilde)", "", "Employment-LinkedIn association | treatment (alpha)", "",
             "Employment-LinkedIn association in control group (delta)", "", "S1 = alpha gamma / beta", "", "S2 = delta gamma / beta", "", "Sample size")
tibble(` ` = rows_c1, `(1) LinkedIn account` = c1_col(fit_acc),
       `Published (1)` = c("0.070", "(0.020)", "0.321", "(0.049)", "0.019", "(0.026)", "0.158", "(0.027)", "0.146", "(0.026)", "0.729", "(0.306)", "0.672", "(0.266)", "1626"),
       `(2) LI usage index` = c1_col(fit_idx),
       `Published (2)` = c("0.083", "(0.020)", "0.954", "(0.145)", "0.028", "(0.025)", "0.058", "(0.014)", "0.059", "(0.017)", "0.668", "(0.253)", "0.678", "(0.259)", "1288")) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Appendix Table C1 of the paper: relationship between treatment, initial employment, and LinkedIn use. Panel A reports the system's parameters, Panel B the shares of the treatment effect explained by LinkedIn use with delta-method standard errors; block fixed effects in every regression, standard errors clustered by cohort.") %>%
  pack_rows("Panel A: parameter estimates", 1, 10) %>%
  pack_rows("Panel B: share of treatment effect explained by LinkedIn use", 11, 14) %>%
  tab_style("scale_down")
```

**Every entry, including the delta-method standard errors of the shares,
reproduces to three decimals.** The shares say that two thirds to three
quarters of the employment effect would run through LinkedIn use if the
mediator were unconfounded; their standard errors say that the interval
covers everything from zero to more than the whole effect. The paper's
phrase is "suggestive evidence".

# Beyond the paper: natural effects, machine learning, sensitivity, and accounting {#sec-beyond}

**The four additions below keep the paper's data and specifications and
ask what the Section 10 tools add.** Each is labelled as ours.

## Natural effects with an interaction

**The shares are the linear model's proportion mediated; allowing the
account's effect to differ by treatment changes little.** With
`interaction = TRUE` the outcome model adds treatment times mediator, and
`mediate_reg()` reports both decompositions by simulation.

```{r beyond-interaction}
d_acc <- wh[complete.cases(wh[, c("employed1_all", "treated", "li_account")]), ]
fit_int <- mediate_reg(d_acc, "employed1_all", "treated", "li_account", x = "strata", interaction = TRUE,
                       cluster = "studref", n_sim = 1000, seed = 1)
fit_int$effects
fit_int$shares
```

```{r beyond-interaction-numbers, echo = FALSE}
int_coef <- coef(fit_int$models$outcome)[["treated:li_account"]]
```

**The interaction coefficient is `r fmt(int_coef)` and not significant:
the account's association with employment is similar in both arms.** The
two decompositions differ by the mediated interaction,
`r fmt(fit_int$effects$estimate[fit_int$effects$term == "nde_total"] - fit_int$effects$estimate[fit_int$effects$term == "nde"])`,
and the proportion mediated is `r fmt(fit_int$shares$estimate[1], 2)` with
a simulation standard error of `r fmt(fit_int$shares$std.error[1], 2)`:
a ratio whose denominator is a 3-standard-error total effect has a wide
interval whichever model computes it.

## Double machine learning with the covariates

**The linear system conditions on blocks only; the influence-function
estimator adds the ten covariates with flexible learners.** The estimand
is unchanged (sequential ignorability given block and covariates), so
what changes is the credibility of the mediator model and the outcome
model. Random forests for the outcome and logistic regressions for the two
propensities; five folds.

```{r beyond-dml, cache = TRUE, cache.extra = list(data_md5)}
rf <- lrn("regr.ranger", num.trees = 500, min.node.size = 20)
fit_dml <- mediate_dml(d_acc, "employed1_all", "treated", "li_account", x = c("strata", covs_miss),
                       learner_y = rf, folds = 5, n_rep = 3, seed = 1)
fit_dml$effects
```

```{r beyond-dml-numbers, echo = FALSE}
nie_dml <- fit_dml$effects$estimate[fit_dml$effects$term == "nie"]
nie_lin <- fit_acc$effects$estimate[fit_acc$effects$term == "nie"]
```

**The indirect effect with forests and covariates is `r fmt(nie_dml)`
against `r fmt(nie_lin)` from the linear system.** The covariates change
the split little, which is what balance (Table 1) and the small
covariate adjustment in Table 4 lead one to expect. The standard errors
are of the same order; the influence-function estimator does not buy
precision here, it buys robustness to the functional form.

## Sensitivity to an unmeasured mediator-outcome confounder

**How much residual correlation between account opening and employment
would erase the share.** `mediate_sensitivity()` traces the indirect
effect against the correlation between the errors of the mediator and
outcome regressions and reports where it crosses zero.

```{r beyond-sensitivity, fig.height = 3.4, fig.cap = "The indirect effect of the training through the LinkedIn account as a function of the correlation between the account and employment errors. Sequential ignorability is rho = 0; the effect vanishes at the value reported as rho zero."}
sens <- mediate_sensitivity(fit_acc, rho = seq(-0.5, 0.5, by = 0.02))
c(rho_zero = sens$rho_zero, rho_tilde = sens$rho_tilde)
plot_mediate_sensitivity(sens)
```

**An error correlation of `r fmt(sens$rho_zero, 2)` is enough.** For scale,
the covariates of Table 4 move the account coefficient from 0.130 to
0.090 among all candidates, a modest amount of observed selection; an
unobserved trait of similar strength would take the share to zero. This
is the quantitative form of the paper's caution.

## Gelbach accounting of the mechanism variables

**Adding the six public mechanism variables to the employment regression
and attributing the change to each one.** Gelbach's (2016) decomposition
assigns the fall in the treatment coefficient to each added covariate
without depending on the order of addition.

```{r beyond-gelbach, fig.height = 3.2, fig.cap = "Gelbach decomposition of the end-of-program employment effect on the LinkedIn account and the five public mechanism variables, block fixed effects, cluster bootstrap intervals."}
mech_g <- c("li_account", "targetsector", "harambee_partner_original", "selfplaced", "s2_completed", "summary_engage")
d_g <- wh[complete.cases(wh[, c("employed1_all", "treated", mech_g)]), ]
g_wh <- gelbach_decomp(d_g, "employed1_all", "treated", x_add = mech_g, fe = "strata", cluster = "studref",
                       n_boot = n_boot, seed = 1)
g_wh$coefficients
g_wh$contributions[, c("covariate", "Gamma", "beta_full", "contribution", "std.error", "share")]
plot_gelbach(g_wh)
```

**The decomposition is an accounting of where the jobs came from, not a
mechanism.** Placement in the targeted sector absorbs almost all of the
coefficient because it is nearly the same event as being employed at the
end of the program in this population. The LinkedIn account's own
contribution is small once placement variables are in the regression,
which is the lecture's point about post-treatment covariates that sit
downstream of the mediator.

## The first step by subgroup

**Where did the training move the mediator?** The account ITT by baseline
account and by education, from the doubly robust score.

```{r beyond-gate}
d_m <- wh[complete.cases(wh[, c("li_account", "treated", "s1_linkedin")]), ]
d_m$account_baseline <- factor(ifelse(d_m$s1_linkedin == 1, "account", "no account"))
d_m$education <- factor(ifelse(d_m$edn_dummy4 == 1, "university", ifelse(d_m$edn_dummy3 == 1, "post-secondary", "secondary")))
sc_m <- dr_scores(d_m, "li_account", "treated", x = c("strata", covs_miss), seed = 1)
cate_gate(sc_m, groups = "account_baseline", n_boot = 499, seed = 1)$table
cate_gate(sc_m, groups = "education", n_boot = 499, seed = 1)$table
```

# What the replication says {#sec-reading}

**The public data reproduce the paper.** Tables 1 to 6, Figure 1, and
Appendix Table C1 match to three decimals wherever the inputs are public;
the four redacted Table 6 columns are the only gap. The one finding about
the code is that the wild bootstrap program computes its replication
t-statistics with conventional standard errors because of an undefined
local macro; the clustered version gives larger p-values for the two
significant waves without changing the conclusion.

**The mediation content is the linear system of Section 10, run as a
regression system with a delta method.** `mediate_reg()` reproduces the
shares and their standard errors because it computes the same stacked
covariance. Its other outputs, the potential-outcome means and both
decompositions, are free once the system is fitted.

**The additions change no conclusion and sharpen the caveat.** Forests and
covariates leave the indirect effect where the linear system put it; an
error correlation of about `r fmt(sens$rho_zero, 2)` would erase it; the
Gelbach accounting shows that placement variables absorb the coefficient
for a reason that has nothing to do with LinkedIn. The paper's own
reading, "suggestive evidence" that the training worked through LinkedIn
use, is the right one.

## Mapping to the lecture's checklist

1. Estimand: total effect (RCT, Figure 1), effects on mediators (Tables 2
   and 6), and a proportion mediated under sequential ignorability
   (Table C1). Stated as such in the paper.
2. Graph: treatment randomized; the account is a post-treatment choice
   whose confounders with employment (motivation, prior experience) are
   partly observed (Table 4).
3. Total and mediator effects first: done, with cluster-robust and wild
   bootstrap inference.
4. Natural effects with and without interaction, and with machine
   learning: this note's Section \@ref(sec-beyond).
5. Sensitivity: rho at zero of `r fmt(sens$rho_zero, 2)`.
6. No instrument for the mediator and no front-door structure.
7. No treatment-induced confounder is modelled; program completion is a
   candidate and moves little (Table 6).
8. Regression accounting: the Gelbach decomposition, labelled as
   accounting.

# References

- Cameron, A. Colin, Jonah B. Gelbach, and Douglas L. Miller. 2008. "Bootstrap-Based Improvements for Inference with Clustered Errors." *Review of Economics and Statistics*.
- Gelbach, Jonah B. 2016. "When Do Covariates Matter? And Which Ones, and How Much?" *Journal of Labor Economics*.
- Imai, Kosuke, Luke Keele, and Teppei Yamamoto. 2010. "Identification, Inference and Sensitivity Analysis for Causal Mediation Effects." *Statistical Science*.
- Roodman, David, Morten Orregaard Nielsen, James G. MacKinnon, and Matthew D. Webb. 2019. "Fast and Wild: Bootstrap Inference in Stata Using boottest." *Stata Journal*.
- Wheeler, Laurel, Robert Garlick, Eric Johnson, Patrick Shaw, and Marissa Gargano. 2022. "LinkedIn(to) Job Opportunities: Experimental Evidence from Job Readiness Training." *American Economic Journal: Applied Economics* 14(2), 101-125. Replication package: openICPSR.
