---
title: "Using est_dml(): learners, external nuisances, and repeated cross-fitting"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using est_dml(): learners, external nuisances, and repeated cross-fitting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)

library(causalmetrics)

run_mlr3 <- requireNamespace("mlr3", quietly = TRUE) &&
  requireNamespace("mlr3learners", quietly = TRUE)
run_ranger <- run_mlr3 && requireNamespace("ranger", quietly = TRUE)
if (run_mlr3) {
  library(mlr3)
  library(mlr3learners)
}

# ---- locate Python and the bundled predictor script ----------------------
py_script <- system.file("python", "nuisance_predictor.py", package = "causalmetrics")

.python_has_deps <- function(bin) {
  if (!nzchar(bin) || !file.exists(bin)) return(FALSE)
  status <- suppressWarnings(system2(
    bin,
    args   = c("-c", shQuote("import pandas, numpy, sklearn")),
    stdout = FALSE, stderr = FALSE
  ))
  identical(as.integer(status), 0L)
}
python_candidates <- unique(c(
  Sys.getenv("CAUSALMETRICS_PYTHON", unset = ""),
  unname(Sys.which("python3")),
  unname(Sys.which("python")),
  "/opt/anaconda3/bin/python"
))
python_candidates <- python_candidates[nzchar(python_candidates)]
python_ok  <- vapply(python_candidates, .python_has_deps, logical(1))
python_bin <- if (any(python_ok)) python_candidates[python_ok][[1]] else ""
run_python <- nzchar(python_bin) && nzchar(py_script)

options(modelsummary_get = "broom")
```

```{r status-note, echo=FALSE, results="asis"}
cat(
  "> **Status:** mlr3 learners ", if (run_mlr3) "available" else "not installed",
  "; ranger ", if (run_ranger) "available" else "not installed",
  "; Python with pandas/numpy/scikit-learn ",
  if (run_python) paste0("detected at `", python_bin, "`") else "not found (those chunks are shown but not run)",
  ".\n", sep = ""
)
```

## What `est_dml()` does

`est_dml()` implements the double/debiased machine learning (DML) recipe:
a Neyman-orthogonal score, cross-fitted nuisance predictions, and
influence-function inference. It owns only the causal layer. The nuisance
functions can be produced by `mlr3` learners inside the function or by
anything else that can write a column of out-of-fold predictions.

Two models are available.

| `model` | Target | Nuisances | Score |
|---|---|---|---|
| `"plr"` | coefficient $\theta$ in $Y = \theta D + g(X) + \varepsilon$ | $\ell(X) = E[Y \mid X]$, $m(X) = E[D \mid X]$ | partialling out: $(\tilde Y - \theta \tilde D)\tilde D$ |
| `"irm"` | ATE or ATT of a binary $D$ | $p(X) = P(D = 1 \mid X)$, $\mu_0(X)$, $\mu_1(X)$ | doubly robust (AIPW) |

The interactive model is `est_aipw()` with the choice of fold solution and
repeated cross-fitting added; with the same nuisances and folds the two functions
return identical numbers. The partially linear model is the generalized
Frisch-Waugh-Lovell estimator of Robinson (1988): residualize $Y$ and $D$
on $X$ with any learner, then regress the residuals on each other.

## A partially linear example

The outcome depends on the covariates nonlinearly, and the treatment is a
continuous dose confounded by the same covariates. The true effect is 1.5.

```{r simulate-plr}
set.seed(1)
n  <- 1500
x1 <- rnorm(n)
x2 <- rnorm(n)
d  <- 0.5 * x1 + sin(x2) + rnorm(n)
y  <- 1.5 * d + x1^2 + exp(x2 / 2) + rnorm(n)
dat <- data.frame(y = y, d = d, x1 = x1, x2 = x2)
```

A linear regression with the covariates as controls gets the functional
form of the confounding wrong:

```{r naive-ols}
coef(summary(lm(y ~ d + x1 + x2, data = dat)))["d", ]
```

`est_dml()` with its default learners is the same linear partialling-out,
but cross-fitted. The interesting version uses random forests for both
nuisances:

```{r dml-plr, eval=run_ranger}
fit_lin <- est_dml(dat, y = "y", d = "d", x = c("x1", "x2"), seed = 1)

set.seed(1)   # ranger draws its own randomness from the session RNG
fit_rf <- est_dml(
  dat, y = "y", d = "d", x = c("x1", "x2"),
  learner_l = lrn("regr.ranger", num.trees = 300),
  learner_m = lrn("regr.ranger", num.trees = 300),
  folds = 5, seed = 1
)
fit_rf
```

The print method reports the cross-fitted root mean squared error of each
nuisance. Those numbers are the first diagnostic to read: DML can only
remove confounding that the learners actually capture.

```{r dml-plr-diag, eval=run_ranger}
fit_lin$diagnostics$nuisance
fit_rf$diagnostics$nuisance
fit_rf$diagnostics$identification
```

The identification block reports the mean squared treatment residual
$E_n[\tilde D^2]$, the denominator of the estimator. When it is small
relative to $\mathrm{var}(D)$, the covariates predict the treatment almost
perfectly and the coefficient is weakly identified. Here about two thirds
of the treatment variance survives partialling out.

```{r resid-plot, eval=run_ranger, fig.cap="Residualized outcome against residualized treatment. The slope of the fitted line is the DML estimate."}
res <- fit_rf$residuals
plot_task(res, ggplot2::aes(x = d_tilde, y = y_tilde)) +
  ggplot2::geom_point(alpha = 0.3, size = 1) +
  ggplot2::geom_abline(intercept = 0, slope = fit_rf$estimate, linewidth = 0.9) +
  ggplot2::labs(x = "D residual: D - m_hat(X)", y = "Y residual: Y - l_hat(X)")
```

## Why cross-fitting matters

The same forests fitted on the full sample and evaluated on the same
observations absorb part of the noise into the nuisance predictions. The
treatment residuals shrink and the estimate drifts.

```{r no-crossfit, eval=run_ranger}
set.seed(1)
fit_insample <- est_dml(
  dat, y = "y", d = "d", x = c("x1", "x2"),
  learner_l = lrn("regr.ranger", num.trees = 300),
  learner_m = lrn("regr.ranger", num.trees = 300),
  cross_fit = FALSE
)
data.frame(
  fit = c("cross-fitted", "full sample"),
  estimate = c(fit_rf$estimate, fit_insample$estimate),
  std.error = c(fit_rf$std.error, fit_insample$std.error),
  mean_sq_D_resid = c(fit_rf$diagnostics$identification$mean_sq_treatment_residual,
                      fit_insample$diagnostics$identification$mean_sq_treatment_residual)
)
```

`cross_fit = FALSE` exists to make this demonstration possible. It is not a
setting to use for results.

## Pooled or fold-averaged solutions, and repeated cross-fitting

`solve = "pooled"` (the default, DML2 in the paper) stacks the scores from
all folds and solves once; `solve = "fold_average"` (DML1) solves fold by
fold and averages the fold estimates. The fold-level solutions are always
stored, and their spread is a useful diagnostic of how much the estimate
depends on the particular partition.

```{r folds, eval=run_ranger}
fit_rf$fold_estimates
```

Repeating the whole procedure on several random partitions and taking the
median removes the dependence on any single split. The reported variance
adds the spread of the repetition estimates around their median, as in
Chernozhukov et al. (2018, Section 3.4).

```{r reps, eval=run_ranger}
set.seed(2)
fit_rep <- est_dml(
  dat, y = "y", d = "d", x = c("x1", "x2"),
  learner_l = lrn("regr.ranger", num.trees = 300),
  learner_m = lrn("regr.ranger", num.trees = 300),
  folds = 5, n_rep = 5, seed = 1
)
fit_rep$repetitions
c(estimate = fit_rep$estimate, std.error = fit_rep$std.error)
```

## The interactive model: ATE and ATT

With a binary treatment and heterogeneous effects, `model = "irm"` uses the
doubly robust score. The ATT score needs only the propensity score and the
untreated outcome regression.

```{r simulate-irm}
set.seed(3)
p   <- plogis(-0.2 + 0.8 * x1 - 0.4 * (x2 > 0))
db  <- rbinom(n, 1, p)
tau <- 1 + 0.5 * x1
yb  <- 1 + x1 + x2^2 + db * tau + rnorm(n)
dat_irm <- data.frame(y = yb, d = db, x1 = x1, x2 = x2)
c(ATE = mean(tau), ATT = mean(tau[db == 1]))
```

```{r dml-irm, eval=run_ranger}
learners_irm <- list(
  learner_p   = lrn("classif.ranger", predict_type = "prob", num.trees = 300),
  learner_mu0 = lrn("regr.ranger", num.trees = 300),
  learner_mu1 = lrn("regr.ranger", num.trees = 300)
)
set.seed(4)
fit_ate <- do.call(est_dml, c(list(dat_irm, y = "y", d = "d", x = c("x1", "x2"),
                                   model = "irm", estimand = "ATE", seed = 1), learners_irm))
set.seed(4)
fit_att <- do.call(est_dml, c(list(dat_irm, y = "y", d = "d", x = c("x1", "x2"),
                                   model = "irm", estimand = "ATT", seed = 1), learners_irm))
fit_ate
fit_att
```

Propensity clipping (`p_clip`) and trimming (`trim`) work as in
`est_aipw()`, and the same overlap, weight, and effective-sample-size
diagnostics are returned under `diagnostics$propensity` and
`diagnostics$weights`.

## External nuisances: the contract

Whatever produces the nuisance predictions must return the rows of the input
data in the same order, with these columns added.

| `model` | Required columns | Meaning |
|---|---|---|
| `"plr"` | `l_hat`, `m_hat` | out-of-fold $\hat\ell(X_i)$ and $\hat m(X_i)$ |
| `"irm"` | `p_hat`, `mu0_hat`, `mu1_hat` | out-of-fold propensity and arm-specific outcome regressions (`mu1_hat` not needed for the ATT) |
| both | `fold_id` (optional) | the fold each row was held out in |

Two rules are not visible in the column list. Predictions must be
out-of-fold: a row's nuisance must come from a model that did not see that
row. And `est_dml()` cannot check this; with supplied nuisances it trusts
the caller. Passing `fold_id` is still worthwhile, because the per-fold
solutions are then reported as a diagnostic.

The bundled Python script implements the contract for both models.

```{r call-python, eval=run_python, purl=FALSE}
tmp_in  <- tempfile(fileext = ".csv")
tmp_out <- tempfile(fileext = ".csv")
data.table::fwrite(dat, tmp_in)

status <- system2(
  command = python_bin,
  args = c(
    shQuote(py_script),
    "--model",  "plr",
    "--input",  shQuote(tmp_in),
    "--output", shQuote(tmp_out),
    "--y",      "y",
    "--d",      "d",
    "--x",      "x1,x2",
    "--folds",  "5",
    "--seed",   "1"
  )
)
if (!identical(as.integer(status), 0L)) {
  stop("Python predictor failed with exit status ", status, ".")
}
augmented <- data.table::fread(tmp_out)
head(augmented)
```

```{r call-python-fallback, eval=!run_python, echo=FALSE}
message("Skipping the Python chunks: no interpreter with pandas, numpy, and scikit-learn was found.")
```

Sanity checks before estimating: no missing predictions, folds of similar
size, and predictions on the scale of the targets.

```{r sanity, eval=run_python, purl=FALSE}
augmented[, .(
  n         = .N,
  na_l_hat  = sum(is.na(l_hat)),
  na_m_hat  = sum(is.na(m_hat)),
  rmse_l    = sqrt(mean((y - l_hat)^2)),
  rmse_m    = sqrt(mean((d - m_hat)^2))
)]
augmented[, .N, by = fold_id][order(fold_id)]
```

```{r run-external, eval=run_python, purl=FALSE}
fit_py <- est_dml(
  augmented, y = "y", d = "d",
  l_hat = "l_hat", m_hat = "m_hat", fold_id = "fold_id"
)
fit_py
fit_py$fold_estimates
```

Because the Python folds travel with the data, the R learners can be
cross-fitted on exactly the same partition, which makes the two producers
directly comparable:

```{r same-folds, eval=run_python && run_ranger, purl=FALSE}
set.seed(5)
fit_same_folds <- est_dml(
  augmented, y = "y", d = "d", x = c("x1", "x2"),
  learner_l = lrn("regr.ranger", num.trees = 300),
  learner_m = lrn("regr.ranger", num.trees = 300),
  fold_id = "fold_id"
)
data.frame(
  producer  = c("Python (sklearn HGB)", "R (ranger, same folds)", "Truth"),
  estimate  = c(fit_py$estimate, fit_same_folds$estimate, 1.5),
  std.error = c(fit_py$std.error, fit_same_folds$std.error, NA_real_)
)
```

The same script with `--model irm` writes `p_hat`, `mu0_hat`, `mu1_hat`, and
`fold_id` for `est_dml(model = "irm")` or `est_aipw()`; see
`vignette("usage_aipw_external_nuisances")`.

## Reporting with modelsummary

`tidy()` and `glance()` methods are registered for both estimators, so the
objects go straight into `modelsummary()`:

```{r modelsummary, eval=run_ranger}
modelsummary::modelsummary(
  list("PLR, linear" = fit_lin, "PLR, forests" = fit_rf,
       "IRM ATE" = fit_ate, "IRM ATT" = fit_att),
  gof_map = c("nobs", "method", "estimand"),
  output = "markdown"
)
```

## Caveats

- **DML is estimation, not identification.** Conditional ignorability,
  a pre-treatment adjustment set, and overlap are still assumptions to
  argue for. Orthogonality and cross-fitting only protect the estimate from
  the errors of the nuisance learners.
- **What the partially linear coefficient means.** When effects are
  heterogeneous, $\theta$ is a weighted average of conditional effects with
  overlap weights $m(X)(1 - m(X))$, not the ATE. Use `model = "irm"` for
  the ATE or ATT of a binary treatment.
- **Weak residual variation.** Read `diagnostics$identification`. If the
  covariates almost determine the treatment, no learner choice will make
  the coefficient precise.
- **Learner quality and tuning.** Compare the cross-fitted RMSE across
  learners and pick the best one for each nuisance; tune inside the
  training folds, never on the full sample.
- **Repetitions.** With flexible learners on moderate samples, `n_rep = 5`
  or more makes the reported estimate independent of one lucky or unlucky
  split.
- **Supplied nuisances must be out-of-fold.** In-sample predictions
  reintroduce exactly the bias that cross-fitting removes, and the function
  cannot detect it.

## Related material

- The Section 04 lecture vignette derives the scores and the orthogonality
  and cross-fitting arguments used here.
- `vignette("usage_aipw_external_nuisances")` covers the interactive model
  contract in more detail.
- Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C.,
  Newey, W., and Robins, J. (2018). Double/debiased machine learning for
  treatment and structural parameters. *The Econometrics Journal*, 21(1),
  C1-C68.
