---
title: "Using causalmetrics for mediation and mechanisms"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using causalmetrics for mediation and mechanisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
                      message = FALSE, warning = FALSE)
library(causalmetrics)
library(mlr3)
library(mlr3learners)
set.seed(1)
run_ranger <- requireNamespace("ranger", quietly = TRUE)
```

## The question and the objects

The functions in this part of the package answer one question: how does a
treatment $D$ produce its effect on an outcome $Y$? A mediator $M$ is a
variable on the path from $D$ to $Y$. Writing $M(d)$ for the mediator under
treatment $d$ and $Y(d, m)$ for the outcome under treatment $d$ and
mediator level $m$, the estimands are

- the total effect $E[Y(1, M(1)) - Y(0, M(0))]$,
- the natural direct effect (NDE) $E[Y(1, M(0)) - Y(0, M(0))]$: change the
  treatment, hold the mediator at its untreated value,
- the natural indirect effect (NIE) $E[Y(1, M(1)) - Y(1, M(0))]$: hold the
  treatment, let the mediator move,
- the controlled direct effect (CDE) $E[Y(1, m) - Y(0, m)]$ at a fixed
  mediator level $m$.

Total = NDE + NIE by construction. Every estimator below returns a data
frame `effects` with these terms, and `tidy()` returns it. Identification
needs more than an experiment on $D$: sequential ignorability (Imai, Keele,
and Yamamoto 2010) also requires the mediator to be as good as randomly
assigned given $D$ and covariates $X$. The package makes the assumption
explicit rather than hiding it, and offers a sensitivity analysis, a
controlled-direct-effect estimator that survives some violations, an
instrumental-variables version, and a front-door version for a different
identifying structure.

`sim_mediation()` generates data from nine designs whose true effects are
stored in the `truth` attribute. Every example below compares an estimate
with that truth.

```{r designs}
dat <- sim_mediation(3000, dgp = "linear", seed = 1)
head(dat, 3)
str(attr(dat, "truth")[c("total", "nde", "nie", "cde")])
```

## Regression-based mediation: `mediate_reg()`

### The linear system

With a linear mediator model $M = \alpha_0 + \alpha_1 D + X'\alpha + e_M$
and a linear outcome model $Y = \beta_0 + \beta_1 D + \beta_2 M + X'\beta +
e_Y$, the NDE is $\beta_1$ and the NIE is $\alpha_1 \beta_2$ (Baron and
Kenny 1986). `mediate_reg()` fits the two models by `lm()` and computes the
effects by g-computation, which reduces to these products here. Three
inference methods are available: the quasi-Bayesian simulation of Imai et
al. (the default), a nonparametric bootstrap, and Sobel's delta method for
the linear single-mediator case.

```{r reg-linear}
fit_lin <- mediate_reg(dat, "y", "d", "m", x = c("x1", "x2"), method = "delta")
fit_lin
tidy(fit_lin)
```

The `shares` element reports the proportion mediated and the two shares
used by Wheeler et al. (2022): `S1` multiplies the mediator coefficient
from the full outcome model by the treatment effect on the mediator, `S2`
uses the mediator-outcome association in the control group instead.

```{r reg-shares}
fit_lin$shares
```

The four potential-outcome means behind the effects are in `potential`, and
`plot_mediation()` draws the effects with their intervals.

```{r reg-plot, fig.height=3.2}
fit_lin$potential
plot_mediation(fit_lin)
```

### Treatment-mediator interaction

When the effect of the mediator depends on the treatment, the product of
coefficients is wrong and the two decompositions differ: the pure NDE plus
the total NIE, or the total NDE plus the pure NIE (VanderWeele 2015). With
`interaction = TRUE` both are returned, as `nde` and `nie` (pure direct,
total indirect) and `nde_total` and `nie_pure`.

```{r reg-interaction}
dat_int <- sim_mediation(3000, dgp = "interaction", seed = 2)
fit_int <- mediate_reg(dat_int, "y", "d", "m", x = c("x1", "x2"), interaction = TRUE, seed = 1)
fit_int
unlist(attr(dat_int, "truth")[c("nde", "nie", "nde_total", "nie_pure", "cde")])
```

Ignoring the interaction misses part of the indirect effect:

```{r reg-interaction-wrong}
mediate_reg(dat_int, "y", "d", "m", x = c("x1", "x2"), method = "delta")$effects[1:3, ]
```

### Binary outcomes and mediators

`outcome = "logit"` and `mediator = "logit"` switch the models. Effects are
then on the probability scale, computed by g-computation over the sample.
The simulation method draws coefficients from their sampling distribution
and recomputes the effects for each draw.

```{r reg-binary}
dat_bin <- sim_mediation(3000, dgp = "binary", seed = 3)
fit_bin <- mediate_reg(dat_bin, "y", "d", "m", x = c("x1", "x2"),
                       outcome = "logit", mediator = "logit", n_sim = 500, seed = 1)
fit_bin$effects[1:3, ]
unlist(attr(dat_bin, "truth")[c("total", "nde", "nie")])
```

### Several parallel mediators

`m` may name several mediators that do not cause each other. The output
adds one `nie_<mediator>` term per mediator; they sum to the joint `nie`.

```{r reg-parallel}
dat_par <- sim_mediation(3000, dgp = "parallel", seed = 4)
fit_par <- mediate_reg(dat_par, "y", "d", c("m1", "m2"), x = c("x1", "x2"), n_sim = 500, seed = 1)
fit_par$effects
unlist(attr(dat_par, "truth")[c("nde", "nie", "nie_m1", "nie_m2")])
```

### Clustered data and the bootstrap

`cluster` names a cluster column; the coefficient covariance and the
bootstrap are clustered on it.

```{r reg-cluster}
dat$school <- sample(1:60, nrow(dat), replace = TRUE)
fit_cl <- mediate_reg(dat, "y", "d", "m", x = c("x1", "x2"), method = "bootstrap",
                      cluster = "school", n_boot = 99, seed = 1)
fit_cl$effects[1:3, ]
```

## Sensitivity to an unmeasured mediator-outcome confounder

An experiment randomizes $D$ but not $M$. If an unmeasured variable affects
both the mediator and the outcome, the errors of the two regressions are
correlated, and the estimated NIE is biased. `mediate_sensitivity()`
computes the NIE that a linear model would imply for each value of the
error correlation $\rho$ (Imai, Keele, and Yamamoto 2010, Theorem 3) and
reports the $\rho$ at which the NIE crosses zero. The `correlated_errors`
design has $\rho = 0.5$; the estimate at $\rho = 0$ is off, the estimate at
the true $\rho$ is right.

```{r sensitivity, fig.height=3.5}
dat_rho <- sim_mediation(4000, dgp = "correlated_errors", rho = 0.5, seed = 5)
fit_rho <- mediate_reg(dat_rho, "y", "d", "m", x = c("x1", "x2"), method = "delta")
sens <- mediate_sensitivity(fit_rho, rho = seq(-0.8, 0.8, by = 0.1))
sens
sens$curve[sens$curve$rho %in% c(0, 0.5), ]
attr(dat_rho, "truth")$nie
plot_mediate_sensitivity(sens)
```

## Double machine learning mediation: `mediate_dml()`

`mediate_reg()` needs correct parametric models. `mediate_dml()` uses the
efficient influence function for $E[Y(d, M(d'))]$ under sequential
ignorability (Tchetgen Tchetgen and Shpitser 2012; Farbmacher et al. 2022),
whose nuisances are the outcome regression $\mu_d(M, X)$, two propensity
scores $p(D = 1 \mid X)$ and $p(D = 1 \mid M, X)$, and a nested regression
of $\mu_d(M, X)$ on $X$ among units with $D = d'$. All are cross-fitted
with `mlr3` learners, and the mediator can be binary, continuous, or a
vector. The defaults are linear and logistic regressions, which reproduce
the linear case.

```{r dml-linear}
fit_dml <- mediate_dml(dat, "y", "d", "m", x = c("x1", "x2"), seed = 1)
fit_dml
```

In the `nonlinear` design the mediator and outcome models have curvature
and an interaction. Linear learners are biased; random forests are not.

```{r dml-nonlinear, eval=run_ranger}
dat_nl <- sim_mediation(3000, dgp = "nonlinear", seed = 6)
rf <- lrn("regr.ranger", num.trees = 200, min.node.size = 10)
rfc <- lrn("classif.ranger", num.trees = 200, min.node.size = 10, predict_type = "prob")
fit_nl_lin <- mediate_dml(dat_nl, "y", "d", "m", x = c("x1", "x2"), seed = 1)
fit_nl_rf <- mediate_dml(dat_nl, "y", "d", "m", x = c("x1", "x2"),
                         learner_y = rf, learner_d = rfc, seed = 1)
cbind(truth = unlist(attr(dat_nl, "truth")[c("total", "nde", "nie")]),
      linear = fit_nl_lin$effects$estimate[1:3], forest = fit_nl_rf$effects$estimate[1:3])
```

The `diagnostics` element reports the propensity overlap and the share of
units trimmed; `psi` holds the influence-function values used for the
standard errors, one column per potential outcome.

```{r dml-diagnostics}
fit_dml$diagnostics[c("propensity_x", "propensity_mx")]
head(fit_dml$psi, 3)
```

With a binary mediator and `m_ref`, the controlled direct effect at that
mediator level is added.

```{r dml-cde}
fit_dml_bin <- mediate_dml(dat_bin, "y", "d", "m", x = c("x1", "x2"), m_ref = 0, seed = 1)
fit_dml_bin$effects
attr(dat_bin, "truth")$cde
```

## Controlled direct effects with treatment-induced confounders: `mediate_cde()`

Natural effects are not identified when a variable affected by the
treatment also confounds the mediator and the outcome. The controlled
direct effect still is, by sequential g-estimation (Acharya, Blackwell, and
Sen 2016): estimate the mediator's effect in a regression that includes the
intermediate confounder, subtract it from the outcome, and regress the
result on the treatment and pre-treatment covariates only. The one-step
regression that conditions on the intermediate confounder is biased,
because that confounder is itself an outcome of the treatment.

```{r cde}
dat_z <- sim_mediation(3000, dgp = "post_treatment_confounder", seed = 7)
fit_cde <- mediate_cde(dat_z, "y", "d", "m", x_pre = c("x1", "x2"), x_post = "z", n_boot = 99, seed = 1)
fit_cde
attr(dat_z, "truth")$cde
```

## Mechanism accounting in regressions: `gelbach_decomp()`

Applied papers often add candidate mediators to a regression and read the
drop in the treatment coefficient as the part "explained" by them. Gelbach
(2016) shows how to attribute that drop to each added covariate in a way
that does not depend on the order of addition: the contribution of
covariate $k$ is the product of its coefficient in the full regression and
the coefficient on $D$ in the regression of covariate $k$ on $D$ and the
base covariates. The contributions sum exactly to the change.

```{r gelbach, fig.height=3}
g <- gelbach_decomp(dat_par, "y", "d", x_base = c("x1", "x2"), x_add = c("m1", "m2"),
                    groups = list(both = c("m1", "m2")), n_boot = 99, seed = 1)
g
plot_gelbach(g)
```

The decomposition is a description of the regressions. In this linear
design without interaction it coincides with the per-mediator indirect
effects; with post-treatment covariates and unmeasured confounding it does
not; the lecture notes discuss when each reading is
warranted. Fixed effects are passed through `fe` and use `fixest`.

## An instrumented mediator: `mediate_iv()`

When the mediator is endogenous but an instrument $Z$ shifts it, the
system of Cattan, Salvanes, and Tominey (2025) estimates the direct effect
as the coefficient on $D$ in the second stage where $M$ is instrumented by
$Z$, and the indirect effect as the product of the first-stage effect of
$D$ on $M$ and the instrumented mediator effect. The `iv_mediator` design
has an unobserved variable that raises both the mediator and the outcome,
so the OLS mediator coefficient is too large and the OLS direct effect is
wrong.

```{r iv}
dat_iv <- sim_mediation(3000, dgp = "iv_mediator", seed = 8)
fit_iv <- mediate_iv(dat_iv, "y", "d", "m", z = "z", x = c("x1", "x2"), n_boot = 99, seed = 1)
fit_iv
unlist(attr(dat_iv, "truth")[c("total", "direct", "indirect", "first_stage", "mediator_effect")])
```

The decomposition assumes that the mediator effect is the same for the
instrument's compliers and for the population. `homogeneity_by` re-runs the
system within subgroups and reports the mediator effect in each, a check
in the spirit of the authors' Figure A13.

```{r iv-homogeneity}
dat_iv$group <- ifelse(dat_iv$x1 > 0, "high x1", "low x1")
mediate_iv(dat_iv, "y", "d", "m", z = "z", x = c("x1", "x2"), homogeneity_by = "group",
           n_boot = 49, seed = 1)$homogeneity
```

Fixed effects (`fe`) and clustered bootstraps (`cluster`) are supported;
the `fixest` models are returned in `models`.

## Front-door adjustment: `front_door()`

The front-door criterion (Pearl 1995) identifies the effect of $D$ on $Y$
when their confounding is unobserved, provided a mediator carries the whole
effect, is unconfounded given $D$, and its effect on $Y$ is unconfounded
given $D$. The `front_door` design has exactly that structure: the
regression of $Y$ on $D$ (the "backdoor" estimate) is biased, the front-door
estimate is not. `method = "regression"` plugs two regressions into the
formula; `method = "aipw"` uses the influence function of Fulcher et al.
(2020) with cross-fitted learners (binary mediator).

```{r front-door}
dat_fd <- sim_mediation(4000, dgp = "front_door", seed = 9)
fd_reg <- front_door(dat_fd, "y", "d", "m", x = c("x1", "x2"), n_boot = 99, seed = 1)
fd_reg
fd_aipw <- front_door(dat_fd, "y", "d", "m", x = c("x1", "x2"), method = "aipw", seed = 1)
fd_aipw$effects
attr(dat_fd, "truth")$total
```

## Weights

`mediate_reg()`, `mediate_iv()`, and `gelbach_decomp()` accept a `weights`
column: every regression is weighted, the robust and clustered
covariances use the weights, the g-computation averages over units with
the weights, and the bootstrap carries them. Survey or design weights
(Cattan et al. weight by school size) go here.

```{r weights}
dat$w <- runif(nrow(dat), 0.5, 2)
fit_w <- mediate_reg(dat, "y", "d", "m", x = c("x1", "x2"), weights = "w", method = "delta")
rbind(unweighted = fit_lin$effects$estimate[1:3], weighted = fit_w$effects$estimate[1:3])
dat_iv$w <- runif(nrow(dat_iv), 0.5, 2)
mediate_iv(dat_iv, "y", "d", "m", z = "z", x = c("x1", "x2"), weights = "w", n_boot = 49, seed = 1)$effects[1:4, ]
dat_par$w <- runif(nrow(dat_par), 0.5, 2)
gelbach_decomp(dat_par, "y", "d", x_base = c("x1", "x2"), x_add = c("m1", "m2"), weights = "w", n_boot = 0)$contributions
```

## Which function when

| Situation | Function |
|---|---|
| Mediator as good as random given $D$ and $X$, parametric models acceptable | `mediate_reg()` |
| Same assumption, flexible nuisances and root-n inference | `mediate_dml()` |
| Unmeasured mediator-outcome confounder, how large would it need to be | `mediate_sensitivity()` |
| A treatment-induced confounder of the mediator and the outcome; only the CDE is wanted | `mediate_cde()` |
| "How much of the coefficient do these covariates absorb", by covariate or group | `gelbach_decomp()` |
| Endogenous mediator with its own instrument | `mediate_iv()` |
| Unobserved treatment-outcome confounding, but a mediator that satisfies the front-door criterion | `front_door()` |

## References

Acharya, A., Blackwell, M., and Sen, M. (2016). Explaining causal findings
without bias. *American Political Science Review*, 110(3), 512-529.

Cattan, S., Salvanes, K. G., and Tominey, E. (2025). First-generation
elite: the role of school social networks. *American Economic Review*,
115(12), 4369-4403.

Farbmacher, H., Huber, M., Laffers, L., Langen, H., and Spindler, M.
(2022). Causal mediation analysis with double machine learning. *The
Econometrics Journal*, 25(2), 277-300.

Fulcher, I. R., Shpitser, I., Marealle, S., and Tchetgen Tchetgen, E. J.
(2020). Robust inference on population indirect causal effects. *JRSS-B*,
82(1), 199-214.

Gelbach, J. B. (2016). When do covariates matter? And which ones, and how
much? *Journal of Labor Economics*, 34(2), 509-543.

Imai, K., Keele, L., and Yamamoto, T. (2010). Identification, inference and
sensitivity analysis for causal mediation effects. *Statistical Science*,
25(1), 51-71.

Pearl, J. (1995). Causal diagrams for empirical research. *Biometrika*,
82(4), 669-688.

Tchetgen Tchetgen, E. J. and Shpitser, I. (2012). Semiparametric theory for
causal mediation analysis. *Annals of Statistics*, 40(3), 1816-1845.

VanderWeele, T. J. (2015). *Explanation in Causal Inference*. Oxford
University Press.

Wheeler, L., Garlick, R., Johnson, E., Shaw, P., and Gargano, M. (2022).
LinkedIn(to) job opportunities. *AEJ: Applied Economics*, 14(2), 101-125.
