---
title: "Using causalmetrics for synthetic control: weights, placebos, conformal inference, and staggered adoption"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using causalmetrics for synthetic control: weights, placebos, conformal inference, and staggered adoption}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
                      message = FALSE, warning = FALSE)
library(causalmetrics)
set.seed(1)
```

## What the package adds

Synthetic control is not a regression, so nothing in `fixest` or `estimatr`
produces it. `causalmetrics` implements the estimator and the inference
tools that the applied literature uses with it.

| Task | Function |
|---|---|
| Donor weights from outcome lags and covariates, with equal, regression-based, or MSPE-optimized predictor weights; demeaned synthetic control; relaxed constraints; ridge penalty; ridge augmentation; one fit per treated unit under staggered adoption | `synth_control()` |
| In-space placebos (MSPE-ratio p-values) and in-time placebos (backdating) | `synth_placebo()` |
| Conformal p-values and confidence sets (Chernozhukov, Wuthrich, and Zhu 2021) | `synth_conformal()` |
| Specification test contrasting the demeaned synthetic control with DiD (Ferman and Pinto 2021) | `synth_spec_test()` |
| Leave-one-donor-out refits | `synth_loo()` |
| Paths, gaps, detrended fit, weights, balance, placebo bands | `plot_synth()` |
| A linear factor model to simulate from | `sim_synth_panel()` |

The synthetic difference-in-differences weights of `sdid_weights()` (lecture
07) are the same family: `synth_control()` with all outcome lags equals
`sdid_weights(estimator = "sc")`, and with `demean = TRUE` it equals the
`"difp"` estimator averaged over the post-treatment periods.

## A panel from a linear factor model

`sim_synth_panel()` draws untreated outcomes `c_j + delta_t + lambda_t' mu_j + e_jt`.
The treated unit here loads more than the donors on the first factor, and
its fixed effect sits above theirs, so neither the donor average nor a
level match is a good counterfactual.

```{r data}
dat <- sim_synth_panel(n_donors = 20, t_pre = 20, t_post = 5, sd_e = 0.7,
                       select_loading = 1.2, select_level = 1.5, effect = 2, seed = 3)
head(dat)
```

## Fitting the synthetic control

The default uses every pre-treatment outcome as a predictor, equal
predictor weights, and weights on the simplex.

```{r fit}
fit <- synth_control(dat, id = "id", time = "time", y = "y", d = "d")
fit
plot_synth(fit, type = "path")
```

The object reports the pre-treatment fit (`pre_rmspe`), the weight
concentration `||w||^2` of Ferman and Pinto (2021), and a balance table.
Two variants matter in practice. The demeaned synthetic control matches
trends rather than levels. The ridge-augmented estimator adds a regression
correction for the remaining imbalance.

```{r variants}
fit_dm <- synth_control(dat, id = "id", time = "time", y = "y", d = "d", demean = TRUE)
fit_aug <- synth_control(dat, id = "id", time = "time", y = "y", d = "d", demean = TRUE, augment = "ridge")
did <- sdid_weights(dat, id = "id", time = "time", y = "y", d = "d", estimator = "did")
c(did = did$estimate, sc = fit$estimate, demeaned = fit_dm$estimate, augmented = fit_aug$estimate)
plot_synth(fit_dm, type = "demeaned_path")
```

Covariates enter as pre-treatment averages, outcome lags can be chosen, and
`v = "mspe"` runs the nested optimization of the predictor weights.

```{r covariates}
dat$x1 <- 0.5 * dat$y0 + rnorm(nrow(dat))
fit_x <- synth_control(dat, id = "id", time = "time", y = "y", d = "d", x = "x1",
                       lags = c(10, 15, 20), pre_window = 11:20, v = "mspe")
fit_x$balance
```

## Inference

In-space placebos reassign the treatment to each donor; the p-value is the
rank of the treated unit's post/pre MSPE ratio.

```{r placebo}
pl <- synth_placebo(fit_dm)
pl
plot_synth(pl)
plot_synth(pl, type = "ratio")
```

An in-time placebo moves the treatment date into the pre-treatment period
and should find nothing.

```{r placebo-time}
synth_placebo(fit_dm, type = "time", placebo_time = 12)$estimate
```

Conformal inference tests a sharp null on the post-treatment effects and
inverts the test into a confidence set.

```{r conformal}
ci <- synth_conformal(fit_dm)
ci
```

The specification test of Ferman and Pinto (2021) contrasts the demeaned
synthetic control with DiD; a low p-value says both are biased.

```{r spec}
st <- synth_spec_test(fit_dm)
st
plot_synth(st)
```

Leave-one-donor-out shows how much the estimate depends on any single donor.

```{r loo}
synth_loo(fit_dm)
```

## Several treated units and staggered adoption

With `treated_units = "separate"`, each treated unit gets its own synthetic
control from the units still untreated over its window, and the gaps are
averaged in event time.

```{r staggered}
stag <- sim_synth_panel(n_donors = 20, n_treated = 4, t_pre = 15, t_post = 8, effect = 2,
                        adoption = c(14, 16, 18, 20), seed = 4)
fit_sep <- synth_control(stag, id = "id", time = "time", y = "y", d = "d",
                         treated_units = "separate", horizon = 4, demean = TRUE)
fit_sep
fit_sep$unit_table
plot_synth(fit_sep)
```

`tidy()` and `glance()` work on every fit for `modelsummary` tables.

```{r tidy}
tidy(fit_dm)
glance(fit_dm)
```
