---
title: "Using est_aipw() with external (Python) nuisance predictions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using est_aipw() with external Python nuisance predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)

library(causalmetrics)

# ---- locate Python and the bundled predictor script ----------------------
# python_bin <- Sys.which("python3")
python_bin <- "/opt/anaconda3/bin/python"
if (!nzchar(python_bin)) python_bin <- Sys.which("python")

py_script <- system.file(
  "vignette_helpers", "03_Selection_on_Observables", "nuisance_predictor.py",
  package = "causalmetrics"
)

# Decide whether to run the Python chunks. We require:
#   (a) a python interpreter on PATH,
#   (b) the bundled script to exist,
#   (c) the script's runtime deps (pandas, numpy, sklearn) to import.
.python_ok <- function() {
  if (!nzchar(python_bin) || !nzchar(py_script)) return(FALSE)
  status <- suppressWarnings(system2(
    python_bin,
    args   = c("-c", shQuote("import pandas, numpy, sklearn")),
    stdout = FALSE, stderr = FALSE
  ))
  identical(as.integer(status), 0L)
}
run_python <- .python_ok()
```

```{r status-note, echo=FALSE, results="asis"}
if (run_python) {
  cat(
    "> **Status:** Python detected at `", python_bin,
    "`; the external-predictor chunks will run for real.\n",
    sep = ""
  )
} else {
  cat(
    "> **Status:** No working Python found (need Python 3 with `pandas`, ",
    "`numpy`, and `scikit-learn`). The external-predictor chunks below are ",
    "shown for reference but not executed. The R-only cross-check still runs.\n",
    sep = ""
  )
}
```

## What this vignette does

`causalmetrics::est_aipw()` separates *what the estimator does* (compute the
AIPW score, its standard error, and overlap diagnostics) from *how the
nuisance functions* — propensity $e(X)$ and outcome regressions $\mu_0(X),
\mu_1(X)$ — are produced. As long as something hands the estimator a data
frame with out-of-fold predictions in the expected columns, the estimator
does not care whether those predictions came from base R, `mlr3`, `tidymodels`,
PyTorch, or a corporate ML platform.

This vignette walks through the canonical external-predictor workflow using
the Python script bundled with the package. The same pattern applies to any
producer that respects the contract laid out below.

When you might want this:

- the most appropriate learner for your problem lives in Python (recent
  `sklearn` estimators, `xgboost` with custom objectives, a fine-tuned
  PyTorch model);
- your team already has a Python ML pipeline you want to reuse instead of
  rewriting in R;
- the nuisance fitting is heavy enough to deserve a separate, possibly
  remote, compute environment;
- you want reproducibility from a CI pipeline that already produces a CSV.

## The contract

The R-side estimator consumes a single data frame. Whatever produces the
nuisance predictions must emit rows in the same order as the input data with
at least the following columns:

| Column      | Type        | Required | Meaning                                                                 |
|-------------|-------------|----------|-------------------------------------------------------------------------|
| `y`         | numeric     | yes      | observed outcome                                                        |
| `d`         | $\{0, 1\}$  | yes      | observed binary treatment                                               |
| `p_hat`     | $(0, 1)$    | yes      | out-of-fold propensity: $\hat e(X_i) = \hat{\mathbb{P}}(D_i = 1 \mid X_i)$ |
| `mu0_hat`   | numeric     | yes      | out-of-fold control-arm outcome regression $\hat\mu_0(X_i)$             |
| `mu1_hat`   | numeric     | yes      | out-of-fold treated-arm outcome regression $\hat\mu_1(X_i)$             |
| `fold_id`   | integer     | optional | which fold $i$ was held out in; recorded in diagnostics if supplied     |

Two requirements that are *not* visible in the column list but matter:

1. **Out-of-fold.** Predictions for row $i$ must come from a model that did
   not see row $i$ during training. Reusing in-fold predictions reintroduces
   the overfitting bias that AIPW + cross-fitting is designed to avoid.
   When `cross_fit = FALSE`, `est_aipw()` trusts the caller on this point —
   it has no way to check.
2. **Arm-specific outcome fits.** `mu0_hat` should come from a model trained
   only on controls in the training folds; `mu1_hat` should come from a
   model trained only on treated units in the training folds. (This is the
   T-learner convention. S-learner-style joint models also work but require
   slightly different handling, beyond the scope of this vignette.)

Things the predictor is *not* responsible for: propensity clipping (handled
by `est_aipw(p_clip = ...)`), identification, choice of estimand, or
inference.

## A simulated example

We use a small linear-treatment, mildly nonlinear-confounding DGP so we can
verify the answer is close to the true ATE of 2:

```{r simulate-data}
set.seed(1)
n   <- 1000
x1  <- rnorm(n)
x2  <- rbinom(n, 1, 0.5)
p   <- plogis(-0.2 + 0.6 * x1 - 0.4 * x2)
d   <- rbinom(n, 1, p)
mu0 <- 1 + x1 + x2
tau <- 2
y   <- mu0 + tau * d + rnorm(n)

dat <- data.frame(y = y, d = d, x1 = x1, x2 = x2)
str(dat)
```

## Calling the bundled Python predictor

The script lives under `inst/vignette_helpers/03_Selection_on_Observables/`
inside the installed package and is discoverable via `system.file()`. The
calling pattern is: write a CSV, invoke the script via `system2()`, read
the augmented CSV back in.

```{r locate-script}
py_script   # path to nuisance_predictor.py
python_bin  # path to python interpreter
```

```{r call-python, eval=run_python}
tmp_in  <- tempfile(fileext = ".csv")
tmp_out <- tempfile(fileext = ".csv")
on.exit(unlink(c(tmp_in, tmp_out)), add = TRUE)

data.table::fwrite(dat, tmp_in)

status <- system2(
  command = python_bin,
  args = c(
    shQuote(py_script),
    "--input",  shQuote(tmp_in),
    "--output", shQuote(tmp_out),
    "--y",      "y",
    "--d",      "d",
    "--x",      "x1,x2",
    "--folds",  "5",
    "--seed",   "1"
  )
)
if (!identical(as.integer(status), 0L)) {
  stop("Python predictor failed with exit status ", status, ".")
}

augmented <- data.table::fread(tmp_out)
head(augmented)
```

```{r call-python-fallback, eval=!run_python, echo=FALSE}
message(
  "Skipping Python call. The next 'sanity checks' and 'run AIPW with ",
  "Python nuisances' sections will also be skipped. The R-only ",
  "cross-check at the end demonstrates the same workflow without Python."
)
```

## Sanity-checking what Python returned

Before running the estimator, look at the columns Python produced. Three
quick checks catch the majority of broken-pipeline failures:

```{r sanity-1, eval=run_python}
# 1. No missing nuisance predictions, no malformed propensity scores.
augmented[, .(
  n_rows      = .N,
  na_p_hat    = sum(is.na(p_hat)),
  na_mu0      = sum(is.na(mu0_hat)),
  na_mu1      = sum(is.na(mu1_hat)),
  p_min       = round(min(p_hat), 3),
  p_max       = round(max(p_hat), 3)
)]
```

```{r sanity-2, eval=run_python}
# 2. Fold sizes look balanced and stratified by treatment.
augmented[, .(n = .N, n_treated = sum(d), n_control = sum(1 - d)),
          by = fold_id][order(fold_id)]
```

```{r sanity-3, eval=run_python, fig.height=3, fig.cap="Propensity-score overlap on the Python-produced p_hat. Mass in both groups across the support means overlap is reasonable."}
ggplot(augmented, aes(x = p_hat, fill = factor(d), color = factor(d))) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 25, alpha = 0.55, position = "identity") +
  scale_fill_manual(values = c("0" = "#F5C4B3", "1" = "#B5D4F4"),
                    labels = c("Control", "Treated"), name = NULL) +
  scale_color_manual(values = c("0" = "#993C1D", "1" = "#185FA5"),
                     labels = c("Control", "Treated"), name = NULL) +
  labs(x = "p_hat (out-of-fold propensity)", y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

If any of these fail — NAs in nuisance columns, propensities pinned at 0/1,
fold sizes wildly unequal — go back to the predictor before running AIPW.
The estimator can't tell good nuisances from bad ones.

## Running `est_aipw()` with the Python nuisances

Since Python already did the cross-fitting, set `cross_fit = FALSE` and pass
`fold_id` so the diagnostics record which fold each unit was held out on.

```{r run-aipw, eval=run_python}
fit_py <- est_aipw(
  data      = augmented,
  y         = "y",
  d         = "d",
  p_hat     = "p_hat",
  mu0_hat   = "mu0_hat",
  mu1_hat   = "mu1_hat",
  fold_id   = "fold_id",
  cross_fit = FALSE,
  p_clip    = c(0.01, 0.99)
)

fit_py
```

The returned object carries the point estimate, an influence-function SE,
overlap diagnostics, and (if supplied) a per-fold breakdown:

```{r aipw-diag, eval=run_python}
str(fit_py$diagnostics, max.level = 2)
```

Use the per-fold summary to spot a fold that drives the estimate — a sign
the nuisance model is unstable on that subsample.

## An R-only cross-check on the same data

The pattern above is not Python-specific. Below we produce nuisance
predictions in base R using logistic regression for the propensity and
ordinary least squares for the outcome regressions, with manual stratified
$K$-fold cross-fitting. This both demonstrates the contract is genuinely
language-agnostic and gives a reference point for the Python estimate.

```{r r-only-crossfit}
k <- 5
set.seed(1)

# Stratified folds by treatment so each fold has both arms.
fold_id <- integer(nrow(dat))
fold_id[dat$d == 1] <- sample(rep(seq_len(k), length.out = sum(dat$d == 1)))
fold_id[dat$d == 0] <- sample(rep(seq_len(k), length.out = sum(dat$d == 0)))

p_hat   <- numeric(nrow(dat))
mu0_hat <- numeric(nrow(dat))
mu1_hat <- numeric(nrow(dat))

for (j in seq_len(k)) {
  train <- dat[fold_id != j, ]
  test  <- dat[fold_id == j, ]

  ps_fit  <- glm(d ~ x1 + x2, data = train, family = binomial())
  mu0_fit <- lm (y ~ x1 + x2, data = subset(train, d == 0))
  mu1_fit <- lm (y ~ x1 + x2, data = subset(train, d == 1))

  p_hat  [fold_id == j] <- predict(ps_fit,  newdata = test, type = "response")
  mu0_hat[fold_id == j] <- predict(mu0_fit, newdata = test)
  mu1_hat[fold_id == j] <- predict(mu1_fit, newdata = test)
}

dat_r <- cbind(dat,
               p_hat = p_hat, mu0_hat = mu0_hat,
               mu1_hat = mu1_hat, fold_id = fold_id)

fit_r <- est_aipw(
  data      = dat_r,
  y         = "y",
  d         = "d",
  p_hat     = "p_hat",
  mu0_hat   = "mu0_hat",
  mu1_hat   = "mu1_hat",
  fold_id   = "fold_id",
  cross_fit = FALSE,
  p_clip    = c(0.01, 0.99)
)
fit_r
```

```{r compare-estimates, eval=run_python}
# Side-by-side comparison.
data.frame(
  source   = c("Python (sklearn)", "R (glm + lm)", "Truth"),
  estimate = c(fit_py$estimate, fit_r$estimate, tau),
  se       = c(fit_py$se,       fit_r$se,       NA_real_)
)
```

Both producers respect the same contract, so both estimates target the same
estimand. Differences between them reflect the choice of learner, not the
estimator: the sklearn pipeline uses `HistGradientBoostingRegressor` for the
outcome surface, which captures any mild nonlinearity better than OLS at
the cost of a bit of variance; the R baseline uses linear models, which are
correctly specified for this DGP and therefore unbiased. On a real dataset
the two will rarely agree to three decimals — that is a feature, not a bug.

## When to use this pattern

- You want to plug an existing Python ML stack into the estimator without
  rewriting it in R.
- The right learner for your data only lives in the Python ecosystem
  (recent `sklearn` releases, PyTorch-based deep models, `lightgbm` with
  custom configurations).
- Nuisance training is heavy and lives on a separate machine, GPU node, or
  CI job that already emits a CSV.
- You want the predictor versioned and tested *separately* from the
  estimator. With this contract you can swap in a new predictor, rerun the
  estimator, and compare estimates without touching the R-side code.

## Caveats and extensions

**File I/O has overhead.** A `tempfile()` round-trip is fine for one-shot
analysis; for repeated calls inside an optimization loop or a Shiny app
use `reticulate` to call the predictor in-process. The contract is
identical — only the transport changes.

**The Python script is a reference, not a mandate.** Anything that emits a
CSV with the contract's columns plugs in. Swap `HistGradientBoostingRegressor`
for `xgboost`, a fine-tuned PyTorch model, or a calibrated random forest;
the R side does not change. The bundled script's source is available at
`r ifelse(nzchar(py_script), py_script, "(installed-package path)")`.

**Cross-fitting is the caller's responsibility when `cross_fit = FALSE`.**
If you produce predictions outside R, you also produce the folds. The
`fold_id` column exists so `est_aipw()` can report per-fold diagnostics, not
because it does any sample-splitting itself in this mode.

**Sample-size strata for binary covariates.** If a covariate level appears
in only one treatment arm in a training fold, the arm-specific outcome
regression will refuse to fit (or fit and refuse to predict on unseen
levels). Either drop the level, pool it, or use a learner that handles
unseen categories — the bundled script raises a clear error rather than
producing silent garbage.

**Same workflow for other languages.** Julia, Spark, an internal REST API —
any producer that emits a CSV with the right columns plugs in unchanged.
The contract is the integration point.

## Related material

- The method itself, with derivations and visualizations, lives in
  `vignettes/lectures/03_selection_on_observables/`.
- The cross-fitting workflow with R-native ML learners (`mlr3`, `grf`,
  `glmnet`) is covered in `vignettes/lectures/04_double_ml/`.
- The bundled Python script's source is at
  `inst/vignette_helpers/03_Selection_on_Observables/nuisance_predictor.py`.
