---
title: "Using causalmetrics for heterogeneous treatment effects and policy learning"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using causalmetrics for heterogeneous treatment effects and policy learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
                      message = FALSE, warning = FALSE)
library(causalmetrics)
library(mlr3)
library(mlr3learners)
set.seed(1)
run_ranger <- requireNamespace("ranger", quietly = TRUE)
run_grf <- requireNamespace("grf", quietly = TRUE)
run_policytree <- requireNamespace("policytree", quietly = TRUE)
```

## One object, many uses

Every function in this part of the package consumes the same input: the
cross-fitted doubly robust pseudo-outcome

$$
Y_i(\hat\eta) = \hat\mu_1(Z_i) - \hat\mu_0(Z_i)
  + \Big(\frac{D_i}{\hat p(Z_i)} - \frac{1 - D_i}{1 - \hat p(Z_i)}\Big)\,(Y_i - \hat\mu_{D_i}(Z_i)),
$$

whose conditional mean given any function of the covariates is the
conditional average treatment effect (CATE). `dr_scores()` builds it once.
Then:

| Task | Function |
|---|---|
| CATE models by meta-learning (S, T, X, DR, R) on any `mlr3` learner | `cate_learner()` |
| Best linear predictor of the CATE, group average effects, simultaneous bands | `cate_blp()`, `cate_gate()` |
| Out-of-sample model comparison with confidence intervals; stacking | `cate_score()`, `cate_ensemble()` |
| Validation: heterogeneity test, calibration, TOC and QINI curves, AUTOC | `cate_validate()` |
| Value of a treatment rule; learning rules (trees, linear, classifiers, budgets) | `policy_value()`, `policy_learn()` |
| Targeting impact versus deprivation for a planner with curvature | `policy_frontier()` |
| Multi-arm scores, off-policy evaluation, and trees over arms | `dr_scores()` with several arms, `contrast_scores()`, `policy_value()`, `policy_learn()` |
| Simulated designs with a known CATE | `sim_hte()` |

Learners come from `mlr3`, causal forests from `grf`, exact policy trees of
any depth from `policytree`. The package owns the causal layer: scores,
inference, scoring, validation, and welfare.

## Data with a known CATE

The `"smooth"` design has five standard normal covariates, confounded
assignment, a nonlinear baseline, and a CATE that depends on `x1` and `x2`
only. We split it into a training sample, a scoring sample, and a test
sample.

```{r data}
dat <- sim_hte(n = 4500, dgp = "smooth", seed = 1)
x <- paste0("x", 1:5)
train <- dat[1:2000, ]
score_set <- dat[2001:3250, ]
test <- dat[3251:4500, ]
head(dat[, c("y", "d", "x1", "x2", "tau_true", "p_true")], 3)
```

## Step 1: the scores

With default learners (logistic regression for the propensity score, linear
regression for the outcomes) and five folds:

```{r scores}
sc_train <- dr_scores(train, y = "y", d = "d", x = x, seed = 1)
sc_train
tidy(sc_train)
```

The score is noisy at the individual level (its 1st and 99th percentiles are
far apart), but its conditional mean is the CATE, which is all the later
steps need.

Flexible learners replace the defaults. Any `mlr3` learner works; here we
use random forests for the outcome regressions.

```{r scores-rf, eval = run_ranger}
rf <- lrn("regr.ranger", num.trees = 300, min.node.size = 20)
sc_train_rf <- dr_scores(train, "y", "d", x, learner_mu = rf, seed = 1)
sc_train_rf$diagnostics$nuisance
```

## Step 2: CATE models

`cate_learner()` reuses the scores. Each method produces a cross-fitted
pseudo-label and then fits `learner_final` to it on `x_het`, using all rows.
We restrict heterogeneity to `x1` and `x2`.

```{r learners, eval = run_ranger}
rf_final <- lrn("regr.ranger", num.trees = 300, min.node.size = 50)
methods <- c(dr = "dr", r = "r", x = "x", t = "t", s = "s")
fits <- lapply(methods, function(m) {
  cate_learner(scores = sc_train_rf, x_het = c("x1", "x2"), method = m,
               learner = rf, learner_final = rf_final)
})
fits$dr
sapply(fits, function(f) sqrt(mean((predict(f, test) - test$tau_true)^2)))
```

The last line is the root mean squared error against the true CATE on the
test sample, something only a simulation can show. The next steps replace
it with statistics that are available in real data.

A parametric final stage is one line: pass `learner_final = lrn("regr.lm")`
and the DR-learner becomes the best linear predictor of the next section.

## Step 3: inference on summaries

The best linear predictor of the CATE in a dictionary of covariate functions
is ordinary least squares of the score on that dictionary, with
heteroskedasticity-robust standard errors and a simultaneous band from the
multiplier bootstrap.

```{r blp}
blp <- cate_blp(sc_train, ~ x1 + I(x2^2), seed = 1)
blp
grid <- data.frame(x1 = seq(-2, 2, by = 0.25), x2 = 0)
plot_cate_blp(blp, grid, seed = 1)
```

The true CATE is `1 + x1 + 0.5 (x2^2 - 1)`, so the coefficients on `x1` and
`x2^2` should be near 1 and 0.5; the intercept picks up the constant.
Group average effects use the same machinery with group indicators:

```{r gate}
q_x1 <- cut(train$x1, quantile(train$x1, 0:4 / 4), include.lowest = TRUE,
            labels = paste0("Q", 1:4))
gate <- cate_gate(sc_train, q_x1, seed = 1)
gate
plot_cate_gate(gate)
```

## Step 4: causal forests next to the meta-learners

`grf::causal_forest()` estimates the CATE with a forest built on the
R-learner moment and returns confidence intervals. Its predictions and its
own doubly robust scores drop into the same evaluation functions. The
pseudo-outcome `grf` uses equals `dr_scores()` with `mu0 = Y.hat - W.hat tau`,
`mu1 = Y.hat + (1 - W.hat) tau`.

```{r grf, eval = run_grf}
library(grf)
cf <- causal_forest(as.matrix(train[, x]), train$y, train$d, num.trees = 500, seed = 1)
tau_cf <- predict(cf)$predictions
sc_grf <- dr_scores(train, "y", "d", p_hat = cf$W.hat,
                    mu0_hat = cf$Y.hat - cf$W.hat * tau_cf,
                    mu1_hat = cf$Y.hat + (1 - cf$W.hat) * tau_cf, p_clip = c(0, 1))
all.equal(sc_grf$score, as.numeric(get_scores(cf)))
```

## Step 5: choose a model out of sample

On the scoring sample, `cate_score()` computes the doubly robust loss of each
candidate and its difference from the constant-effect model with a confidence
interval. Differences of losses estimate differences of mean squared errors
against the true CATE, which is why they can be compared with confidence.

```{r score, eval = run_ranger}
sc_score <- dr_scores(score_set, "y", "d", x, learner_mu = rf, seed = 2)
cs <- cate_score(sc_score, dr_learner = fits$dr, r_learner = fits$r,
                 x_learner = fits$x, t_learner = fits$t, s_learner = fits$s)
cs
```

Avoid model names that partially match `scores`, such as `s` or `sc`; R
would assign them to that argument.

`cate_ensemble()` stacks the candidates. Q-aggregation keeps weights on the
simplex and penalizes each model by its own loss.

```{r ensemble, eval = run_ranger}
ens <- cate_ensemble(sc_score, dr_learner = fits$dr, r_learner = fits$r,
                     x_learner = fits$x, t_learner = fits$t, s_learner = fits$s,
                     method = "q")
ens
sqrt(mean((predict(ens, test) - test$tau_true)^2))
```

## Step 6: validate on the test sample

`cate_validate()` takes a fixed model and the test-sample scores. It reports
the heterogeneity regression (slope ideally 1), calibration by quartile of the
prediction, the TOC and QINI curves with one-sided simultaneous bands, and the
areas under them.

```{r validate, eval = run_ranger}
sc_test <- dr_scores(test, "y", "d", x, learner_mu = rf, seed = 3)
val <- cate_validate(sc_test, ens, n_boot = 499, seed = 1)
val
plot_cate_validation(val, "calibration")
plot_cate_validation(val, "toc")
```

`AUTOC` and `AUQC` integrate the curves over all ranks and match
`grf::rank_average_treatment_effect()` when the same scores are used.

## Step 7: from CATE to policy

A policy is a rule `X -> {0, 1}`. Its gain over treating no one is the mean
of `pi(X) (Y(eta) - cost)`, an average of the score, so it comes with a
standard error.

```{r policy-value, eval = run_ranger}
policy_value(sc_test, list(
  ensemble_sign = predict(ens, test) > 0,
  treat_all = rep(1, nrow(test)),
  oracle = test$tau_true > 0
))
```

`policy_learn()` maximizes the empirical welfare over a restricted class.
The `"policy"` design has a CATE that is piecewise constant on a depth-two
tree, so the exact search should recover it.

```{r policy-learn}
pol_dat <- sim_hte(4000, dgp = "policy", seed = 2)
sc_pol <- dr_scores(pol_dat, "y", "d", x, seed = 1)
pol <- policy_learn(sc_pol, method = "tree", depth = 2, seed = 1)
pol
attr(pol_dat, "optimal_tree")
plot_policy_tree(pol)
```

Half of the rows were held out before the search, so the second row of the
value table is an honest estimate. `engine = "policytree"` delegates the
search to that package and allows any depth:

```{r policytree, eval = run_policytree}
pol_pt <- policy_learn(sc_pol, method = "tree", depth = 2, engine = "policytree", seed = 1)
pol_pt$value
mean(predict(pol_pt, pol_dat) == predict(pol, pol_dat))
```

Other classes: linear rules (`method = "linear"`), any weighted `mlr3`
classifier (`method = "classifier"`), and the budgeted rule that treats the
top share of a CATE model (`method = "budget"`, with `tau_hat`). The fitted
object behind a rule is in `$model`, so the usual inspection tools apply:
for a gradient-boosting classifier, the variable importance of the policy.

```{r classifier-model, eval = requireNamespace("xgboost", quietly = TRUE)}
xgb <- lrn("classif.xgboost", predict_type = "prob", nrounds = 100, max_depth = 3)
pol_xgb <- policy_learn(sc_pol, method = "classifier", learner = xgb, seed = 1)
pol_xgb$value
head(xgboost::xgb.importance(model = pol_xgb$model$model), 5)
```

Unit weights enter both functions through `weights`: survey weights, or the
time-decaying weights `h_t` that policy learning with adaptively collected
data requires (Zhan, Ren, Athey, and Zhou 2024). The objective becomes
`sum(w_i Gamma_i pi_i) / sum(w_i)` and the standard error is the adaptively
weighted one, `sqrt(sum(w_i^2 (c_i - value)^2)) / sum(w_i)`.

```{r policy-weights}
h <- 1 / seq_len(nrow(pol_dat))^0.5
pol_w <- policy_learn(sc_pol, method = "tree", depth = 2, weights = h, seed = 1)
pol_w$value
policy_value(sc_pol, list(weighted = pol_w, unweighted = pol), weights = h)
```

Bootstrap percentile intervals for policy values, the construction used in
much of the targeting literature, are one argument away. The resampled
estimates are kept in the `"boot"` attribute so any other level can be read
off, for instance a 75 percent interval next to the 95 percent one.

```{r policy-boot}
pv <- policy_value(sc_pol, list(tree = pol, treat_all = rep(1, nrow(pol_dat))),
                   n_boot = 500, seed = 1)
pv[, c("policy", "estimate", "std.error", "conf.low", "conf.high", "boot.low", "boot.high")]
apply(attr(pv, "boot"), 2, quantile, probs = c(0.125, 0.875))
```

## Step 8: targeting impact versus deprivation

`policy_frontier()` implements the planner of Haushofer et al. (2025): with
utility curvature `a`, a unit is worth treating in proportion to
`u(y0 + tau) - u(y0)`. Linear utility targets impact; strong curvature
targets the most deprived.

```{r frontier, eval = run_ranger}
fr <- policy_frontier(sc_test, ens, y0_hat = sc_test$nuisance$mu0, budget = 0.3,
                      utility = "cara", curvature = c(0, 0.5, 2, 10))
fr
plot_policy_frontier(fr, curvature = 2)
```

## Step 9: more than two arms

Many experiments compare several treatments. With a treatment column that
has more than two values, `dr_scores()` returns the multi-arm score matrix
`Gamma[i, k] = mu_k(x_i) + 1{W_i = k}(Y_i - mu_k(x_i)) / e_k(x_i)`, one
column per arm: its column means are the expected outcomes under each arm,
and `Gamma[i, pi(x_i)]` averaged over units is the expected outcome under
any arm-valued policy `pi`. That is the object of multi-action policy
learning (Athey and Wager 2021; the `policytree` package) and of off-policy
evaluation: with `type = "ipw"` the score is `1{W_i = k} Y_i / e_k(x_i)`
and `policy_value()` returns the inverse-propensity-score reward.

```{r multi-arm}
set.seed(3)
n <- 3000
arm <- sample(c("7", "14", "30"), n, replace = TRUE, prob = c(0.15, 0.15, 0.7))
skill <- sample(c("beginner", "mixed", "expert"), n, replace = TRUE, prob = c(0.6, 0.25, 0.15))
x1 <- rnorm(n)
mu <- cbind(`7` = 0.14 + 0.04 * (skill == "beginner"), `14` = 0.14 + 0.04 * (skill == "mixed"),
            `30` = 0.14 + 0.04 * (skill == "expert")) + 0.02 * x1
y <- rbinom(n, 1, mu[cbind(seq_len(n), match(arm, colnames(mu)))])
trial <- data.frame(y, arm, x1, beginner = as.integer(skill == "beginner"), mixed = as.integer(skill == "mixed"))
# a randomized design: the propensities are the assignment shares
e <- matrix(rep(c(0.15, 0.15, 0.7), each = n), n, dimnames = list(NULL, c("7", "14", "30")))
sc3 <- dr_scores(trial, "y", "arm", c("x1", "beginner", "mixed"), arms = c("7", "14", "30"),
                 p_hat = e, seed = 1)
sc3
```

A policy is a vector of arm labels. `policy_value()` prices it against a
uniform policy that gives one arm to everyone:

```{r multi-arm-value}
rule <- ifelse(trial$beginner == 1, "7", ifelse(trial$mixed == 1, "14", "30"))
policy_value(sc3, list(by_skill = rule, all_7 = rep("7", n)), baseline = "30")
```

`policy_learn()` searches trees whose leaves are arms, exactly for depth
one and two and through `policytree` for any depth, and reports the
held-out expected outcome and gain:

```{r multi-arm-tree}
pol3 <- policy_learn(sc3, x = c("x1", "beginner", "mixed"), depth = 2, baseline = "30", seed = 1)
pol3
```

Every binary tool applies to one pair of arms through `contrast_scores()`,
which builds the binary score object of a pair from the stored nuisances
(rows of the two arms, pairwise propensity `e_t / (e_t + e_c)`):

```{r multi-arm-contrast}
sc_7_30 <- contrast_scores(sc3, treat = "7", control = "30")
# the pair object keeps only the rows of the two arms, so build groups from its data
cate_gate(sc_7_30, ifelse(sc_7_30$data$beginner == 1, "beginner", "other"), n_boot = 199)
```

## External nuisances

Any engine can supply the out-of-fold predictions, exactly as for
`est_dml()`. The bundled Python script writes `p_hat`, `mu0_hat`, `mu1_hat`
columns; pass them by name and no learner is fitted.

```{r external, eval = FALSE}
sc_py <- dr_scores(train_with_predictions, "y", "d", x,
                   p_hat = "p_hat", mu0_hat = "mu0_hat", mu1_hat = "mu1_hat",
                   fold_id = "fold_id")
```

## References

- Athey, S. and Wager, S. (2021). Policy learning with observational data. *Econometrica*, 89(1), 133-161.
- Chernozhukov, V., Hansen, C., Kallus, N., Spindler, M., and Syrgkanis, V. (2026). *Applied Causal Inference Powered by ML and AI*, chapters 14-15.
- Haushofer, J., Niehaus, P., Paramo, C., Miguel, E., and Walker, M. (2025). Targeting impact versus deprivation. *American Economic Review*.
- Kennedy, E. H. (2023). Towards optimal doubly robust estimation of heterogeneous causal effects. *Electronic Journal of Statistics*, 17(2).
- Künzel, S. R., Sekhon, J. S., Bickel, P. J., and Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. *PNAS*, 116(10).
- Nie, X. and Wager, S. (2021). Quasi-oracle estimation of heterogeneous treatment effects. *Biometrika*, 108(2).
- Semenova, V. and Chernozhukov, V. (2021). Debiased machine learning of conditional average treatment effects and other causal functions. *The Econometrics Journal*, 24(2).
- Zhan, R., Ren, Z., Athey, S., and Zhou, Z. (2024). Policy learning with adaptively collected data. *Management Science*, 70(8).
- Yadlowsky, S., Fleming, S., Shah, N., Brunskill, E., and Wager, S. (2021). Evaluating treatment prioritization rules via rank-weighted average treatment effects. arXiv:2111.07966.
