---
title: "Using causalmetrics with the rdrobust family for regression discontinuity designs"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using causalmetrics with the rdrobust family for regression discontinuity designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
                      message = FALSE, warning = FALSE)
library(causalmetrics)
set.seed(1)
run_rd <- requireNamespace("rdrobust", quietly = TRUE)
run_honest <- requireNamespace("RDHonest", quietly = TRUE)
run_density <- requireNamespace("rddensity", quietly = TRUE)
run_locrand <- requireNamespace("rdlocrand", quietly = TRUE)
run_multi <- requireNamespace("rdmulti", quietly = TRUE)
run_hte <- requireNamespace("rdhte", quietly = TRUE)
run_mlr3 <- requireNamespace("mlr3", quietly = TRUE) && requireNamespace("mlr3learners", quietly = TRUE) &&
  requireNamespace("ranger", quietly = TRUE)
knitr::opts_chunk$set(eval = run_rd)
```

## What the package adds to the rdrobust family

The packages of Cattaneo, Titiunik, and coauthors estimate regression
discontinuity (RD) designs: `rdrobust` (local polynomial estimation with
robust bias-corrected inference and bandwidth selection), `rddensity`
(manipulation tests), `RDHonest` (honest inference), `rdlocrand` (local
randomization), `rdmulti` (multiple cutoffs and scores), `rdhte`
(heterogeneity at the cutoff). `causalmetrics` calls them as they are and
adds what they do not do:

| Task | Function |
|---|---|
| Flexible covariate adjustment with cross-fitted learners before `rdrobust` | `rd_adjust()` |
| Effects away from the cutoff under conditional independence | `rd_extrapolate()` |
| Weak-identification-robust intervals for fuzzy RD | `rd_weak_iv()` |
| Balance with a joint test, placebo cutoffs, bandwidth sensitivity, donut holes, as one object | `rd_checks()`, `plot_rd_checks()` |
| RD plots as modifiable ggplot objects | `rd_plot()`, `rd_bins()` |
| One tidy table across packages | `rd_frame()`, `tidy()` for `RDHonest` and `rdhte` |
| Regression kink reporting and the four kink graphs | `rd_kink()`, `plot_rd_kink()` |
| Simulated designs with known effects | `sim_rd()` |

All examples use simulated data from `sim_rd()`, whose effect at the
cutoff is known.

## Sharp RD with rdrobust, and the plot

```{r sharp}
dat <- sim_rd(3000, "lee", seed = 1)
attr(dat, "tau_true")
fit <- rdrobust::rdrobust(dat$y, dat$x, c = 0)
summary(fit)
rd_plot(dat, "y", "x")
```

`rd_plot()` returns a ggplot object. The default bins are the IMSE-optimal
quantile-spaced bins of `rdrobust::rdplot()`; the fitted lines are global
fourth-order polynomials on each side. Everything can be changed with
`+`:

```{r plot-modify}
rd_plot(dat, "y", "x", fit = "local", ci = TRUE, n_bins = 25) +
  ggplot2::labs(title = "Local linear fit inside the MSE-optimal bandwidth",
                x = "Running variable", y = "Outcome") +
  ggplot2::theme_bw()
head(rd_bins(dat, "y", "x", n_bins = 10))
```

## The diagnostic battery

`rd_checks()` runs the main estimate, covariate balance with a joint test,
placebo cutoffs on each side, bandwidth sensitivity, and donut-hole
estimates. Each piece is also available on its own.

```{r checks}
datc <- sim_rd(3000, "covariates", seed = 2)
ck <- rd_checks(datc, "y", "x", covariates = paste0("z", 1:4))
ck
plot_rd_checks(ck, "sensitivity")
plot_rd_checks(ck, "balance")
```

The joint balance test stacks the influence functions of the local linear
jumps of all covariates at one bandwidth, which `rdrobust` does not
provide.

## Flexible covariate adjustment

`rdrobust(covs = )` adjusts linearly. `rd_adjust()` subtracts a
cross-fitted machine-learning prediction of the outcome from the
covariates before the local polynomial step. Any function of the
covariates leaves the RD estimand unchanged, so no rate condition is
needed and the only effect is a smaller variance.

```{r adjust, eval = run_rd && run_mlr3}
library(mlr3)
library(mlr3learners)
adj_lin <- rd_adjust(datc, "y", "x", covariates = paste0("z", 1:4), seed = 1)
adj_rf <- rd_adjust(datc, "y", "x", covariates = paste0("z", 1:4),
                    learner = lrn("regr.ranger", num.trees = 300, min.node.size = 20), pooled = FALSE, seed = 1)
rd_frame(raw = adj_lin$fits$raw, linear = adj_lin$fits$adjusted, forest = adj_rf$fits$adjusted, methods = "robust")
```

The design has a nonlinear covariate effect, so the forest removes more
outcome variance than the linear model and the standard error falls
further. The adjusted outcome is in `adj_rf$data$y_adj` for any further
use.

## Honest inference and local randomization

```{r honest, eval = run_rd && run_honest}
hn <- RDHonest::RDHonest(y ~ x, data = dat, cutoff = 0)
tidy(hn)
```

```{r locrand, eval = run_rd && run_locrand}
lr <- rdlocrand::rdrandinf(dat$y, dat$x, wl = -0.05, wr = 0.05, quietly = TRUE, ci = 0.05)
```

```{r frame, eval = run_rd && run_honest && run_locrand}
rd_frame(rdrobust = fit, honest = hn, local_random = lr, methods = "robust")
```

## Manipulation

```{r density, eval = run_rd && run_density}
man <- sim_rd(4000, "manipulated", seed = 3)
summary(rddensity::rddensity(man$x, c = 0))
rd_plot(man, "y", "x")
plot_rd_checks(rd_donut(man, "y", "x"))
```

## Fuzzy RD and weak first stages

```{r fuzzy}
fz <- sim_rd(3000, "fuzzy", compliance = 0.7, seed = 4)
rd_weak_iv(fz, "y", "d", "x")
```

The table shows the local two-stage least squares estimate with its Wald
interval, the Anderson-Rubin set, and the `rdrobust` fuzzy rows at the
same bandwidth. With a weak first stage the Anderson-Rubin set widens or
becomes unbounded while the Wald interval does not, which is the warning
sign.

## Effects away from the cutoff

```{r extrapolate}
cia <- sim_rd(4000, "cia", seed = 5)
ex <- rd_extrapolate(cia, "y", "x", covariates = "z", window = 0.5)
ex
plot_rd_extrapolate(ex, truth = attr(cia, "tau_true"))
```

The conditional independence test does not reject on either side, and the
average effects above and below the cutoff recover the true effect, which
in this design is constant.

## Multiple cutoffs and heterogeneity

```{r multi, eval = run_rd && run_multi}
mc_dat <- sim_rd(4000, "multi_cutoff", seed = 6)
mc <- rdmulti::rdmc(mc_dat$y, mc_dat$x, C = mc_dat$cutoff, verbose = FALSE, plot = FALSE)
rd_frame(multi = mc)
```

```{r hte, eval = run_rd && run_hte}
datc$group <- as.integer(datc$z3 > 0)
ht <- rdhte::rdhte(datc$y, datc$x, c = 0, covs.hte = datc$group)
tidy(ht)
```

## Regression kink

```{r kink}
kd <- sim_rd(4000, "kink", seed = 7)
kk <- rd_kink(kd, "y", "x", d = "d", elasticity = TRUE)
kk
```

```{r kink-plot, fig.height = 6, eval = run_rd && requireNamespace("patchwork", quietly = TRUE)}
plot_rd_kink(kk)
```

## References

- Angrist, J. D. and Rokkanen, M. (2015). Wanna get away? Regression discontinuity estimation of exam school effects away from the cutoff. *Journal of the American Statistical Association*, 110(512).
- Calonico, S., Cattaneo, M. D., and Titiunik, R. (2014). Robust nonparametric confidence intervals for regression-discontinuity designs. *Econometrica*, 82(6).
- Calonico, S., Cattaneo, M. D., and Titiunik, R. (2015). Optimal data-driven regression discontinuity plots. *Journal of the American Statistical Association*, 110(512).
- Card, D., Lee, D. S., Pei, Z., and Weber, A. (2015). Inference on causal effects in a generalized regression kink design. *Econometrica*, 83(6).
- Feir, D., Lemieux, T., and Marmer, V. (2016). Weak identification in fuzzy regression discontinuity designs. *Journal of Business and Economic Statistics*, 34(2).
- Noack, C., Olma, T., and Rothe, C. (2024). Flexible covariate adjustments in regression discontinuity designs. arXiv:2107.07942.
