---
title: "Using causalmetrics for difference-in-differences: from two-by-two to staggered, continuous, and synthetic designs"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using causalmetrics for difference-in-differences: from two-by-two to staggered, continuous, and synthetic designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
                      message = FALSE, warning = FALSE)
library(causalmetrics)
library(fixest)
set.seed(1)
run_honest <- requireNamespace("HonestDiD", quietly = TRUE)
run_mlr3 <- requireNamespace("mlr3", quietly = TRUE) && requireNamespace("mlr3learners", quietly = TRUE) &&
  requireNamespace("ranger", quietly = TRUE)
```

## What the package adds to fixest

`fixest` already estimates every regression a difference-in-differences (DiD)
analysis needs: the two-by-two, two-way fixed effects (TWFE), event-study
dummies with `i()`, Sun and Abraham with `sunab()`, and clustered standard
errors. `causalmetrics` adds the pieces that are not regressions:

| Task | Function |
|---|---|
| Group-time effects `ATT(g, t)` with never- or not-yet-treated comparisons, covariates by regression, IPW, or doubly robust scores, influence functions, uniform bands | `att_gt()` |
| Event-study, cohort, calendar, and overall aggregates with valid inference | `aggregate_att()` |
| Imputation estimator (Borusyak, Jaravel, Spiess; Gardner) | `did_imputation()` |
| Why static TWFE fails: Goodman-Bacon decomposition and negative weights | `bacon_decomp()`, `twfe_weights()` |
| One frame and one plot for all estimators | `event_study_frame()`, `plot_event_study()` |
| Pre-test power and conditional bias (Roth 2022); HonestDiD inputs | `pretrend_power()`, `honestdid_inputs()` |
| Randomization inference with few clusters | `did_permutation_test()` |
| Efficient estimator under random treatment timing (Roth-Sant'Anna) | `staggered_efficient()` |
| Continuous or multi-valued doses | `att_dose()` |
| Synthetic DiD weights for a weighted fixest regression | `sdid_weights()`, `sdid_se()`, `plot_sdid()` |
| A staggered-adoption simulator | `sim_did_panel()` |

All examples use simulated data, so the truth is known.

```{r data}
dat <- sim_did_panel(n_units = 400, n_periods = 10, groups = c(4, 7), never_share = 0.4,
                     mu = c(3, 1), x_trend = 0.3, seed = 1)
head(dat)
truth <- attr(dat, "att_gt")
head(truth)
```

Units in cohort 4 have effects `3 * (e + 1)` at event time `e`, cohort 7 has
`1 * (e + 1)`, and untreated outcomes trend with `x1`, so parallel trends
holds only conditional on `x1`.

## The two-by-two in fixest and in att_gt

Take cohort 4 against the never treated in periods 3 and 4. The regression
coefficient on the interaction and `att_gt()` on that single cell coincide.

```{r two-by-two}
two <- subset(dat, (g == 4 | g == 0) & time %in% c(3, 4))
two$post <- as.integer(two$time == 4)
two$treat <- as.integer(two$g == 4)
c(fixest = coef(feols(y ~ treat:post | id + time, data = two, cluster = ~id))[["treat:post"]],
  att_gt = att_gt(two, id = "id", time = "time", group = "g", y = "y", n_boot = 0)$att_gt$att)
```

## Covariates: regression, IPW, doubly robust

Adding `x1` linearly to the TWFE regression does not recover the effect when
trends depend on `x1`. The three conditional estimators do. On panel data the
doubly robust cell equals the ATT score of `est_dml()` on the differenced
outcome.

```{r covariates}
truth_44 <- truth$att[truth$group == 4 & truth$time == 4]
twfe_x <- feols(y ~ treat:post + x1 | id + time, data = two, cluster = ~id)
fits <- sapply(c("reg", "ipw", "dr"), function(m) {
  att_gt(two, id = "id", time = "time", group = "g", y = "y", x = "x1", method = m, n_boot = 0)$att_gt$att
})
round(c(truth = truth_44, twfe_plus_x = coef(twfe_x)[["treat:post"]], fits), 3)
```

## Two groups, many periods: event studies

With one cohort, `att_gt()` returns the event-study cells directly. The TWFE
event-study regression with all leads and lags gives the same numbers when
the never treated are the only comparison group.

```{r two-group-es}
one <- subset(dat, g %in% c(0, 7))
es <- att_gt(one, id = "id", time = "time", group = "g", y = "y", control_group = "never",
             base_period = "universal", n_boot = 199, seed = 1)
one$rel <- ifelse(one$g > 0, one$time - one$g, -1000)
tw <- feols(y ~ i(rel, ref = c(-1, -1000)) | id + time, data = one, cluster = ~id)
frame <- event_study_frame("att_gt (universal base)" = es, "fixest i()" = tw)
plot_event_study(frame)
```

## Staggered adoption: why static TWFE fails

```{r staggered-twfe}
static <- feols(y ~ treated | id + time, data = dat, cluster = ~id)
coef(static)[["treated"]]
bacon <- bacon_decomp(dat, id = "id", time = "time", y = "y", d = "treated")
bacon$by_type
w <- twfe_weights(dat, id = "id", time = "time", d = "treated", y = "y")
c(share_negative = w$share_negative, sum_negative = w$sum_negative)
```

The "later vs earlier treated" comparisons use already-treated units, whose
outcomes are still rising, as controls. Their weight is not small.

## Staggered adoption: Callaway-Sant'Anna, Sun-Abraham, imputation

```{r staggered-cs}
cs <- att_gt(dat, id = "id", time = "time", group = "g", y = "y", x = "x1",
             control_group = "notyet", n_boot = 199, seed = 1)
cs
dyn <- aggregate_att(cs, type = "dynamic", min_e = -4, max_e = 4)
dyn
aggregate_att(cs, type = "group")$by
```

A time-invariant covariate is collinear with the unit effects, so the
regression-based estimators need it interacted with time; `att_gt()` uses it
in the cells directly.

```{r staggered-compare}
dat$x1t <- dat$x1 * dat$time
sa <- feols(y ~ sunab(g, time) + x1t | id + time, data = dat, cluster = ~id)
imp <- did_imputation(dat, id = "id", time = "time", group = "g", y = "y", x = "x1t", pre_window = 4)
dat$rel <- ifelse(dat$g > 0, dat$time - dat$g, -1000)
tw_all <- feols(y ~ i(rel, ref = c(-1, -1000)) + x1t | id + time, data = dat, cluster = ~id)
frame <- event_study_frame("Callaway-Sant'Anna" = dyn, "Sun-Abraham" = sa,
                           "Imputation" = imp, "TWFE event study" = tw_all)
plot_event_study(subset(frame, event_time >= -4 & event_time <= 4))
```

The Callaway-Sant'Anna, Sun-Abraham, and imputation curves track the true
dynamic effects; the TWFE event study does not, because its leads and lags
are contaminated by the other cohort.

## Assessing parallel trends

`pretrend_power()` asks what linear violation the pre-test would detect and
what bias survives when the test passes.

```{r pretrends}
pp <- pretrend_power(dyn, slope = 0.1, n_sim = 2000, seed = 1)
pp
```

With HonestDiD installed, `honestdid_inputs()` supplies the coefficients and
covariance for the sensitivity analysis.

```{r honest, eval = run_honest}
csu <- att_gt(dat, id = "id", time = "time", group = "g", y = "y", x = "x1",
              control_group = "notyet", base_period = "universal", n_boot = 0)
dyn_u <- aggregate_att(csu, type = "dynamic", min_e = -3, max_e = 2, n_boot = 0)
hi <- honestdid_inputs(dyn_u)
rm <- HonestDiD::createSensitivityResults_relativeMagnitudes(
  betahat = hi$betahat, sigma = hi$sigma,
  numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods,
  Mbarvec = c(0.5, 1, 1.5), l_vec = HonestDiD::basisVector(1, hi$numPostPeriods)
)
rm
```

## Few treated clusters: randomization inference

```{r permutation}
small <- sim_did_panel(n_units = 40, n_periods = 8, groups = 5, never_share = 0.6, mu = 2, seed = 2)
small$state <- small$id
did_permutation_test(small, id = "id", time = "time", group = "g", y = "y",
                     cluster = "state", n_perm = 299, seed = 1)
```

## Random treatment timing

When timing is randomly assigned, `staggered_efficient()` estimates the
Roth-Sant'Anna efficient estimator; `beta = 1` recovers the
Callaway-Sant'Anna weighting for comparison.

```{r staggered-efficient}
rt <- sim_did_panel(n_units = 400, n_periods = 8, groups = c(3, 5, 7), never_share = 0.3, x_select = 0, seed = 5)
rbind(efficient = staggered_efficient(rt, id = "id", time = "time", group = "g", y = "y",
                                      estimand = "eventstudy", event_time = 0:2)$estimates[, 1:4],
      did = staggered_efficient(rt, id = "id", time = "time", group = "g", y = "y",
                                estimand = "eventstudy", event_time = 0:2, beta = 1)$estimates[, 1:4])
```

## Continuous treatment

```{r dose}
n <- 800
dose <- ifelse(runif(n) < 0.3, 0, runif(n, 0.5, 3))
dd <- data.frame(id = rep(1:n, each = 2), time = rep(1:2, n), dose = rep(dose, each = 2))
dd$y <- rep(rnorm(n), each = 2) + 0.5 * dd$time + (dd$time == 2) * (2 * dd$dose - 0.3 * dd$dose^2) + rnorm(2 * n, sd = 0.5)
ad <- att_dose(dd, id = "id", time = "time", y = "y", dose = "dose", dose_type = "spline", n_bins = 4, n_boot = 99, seed = 1)
ad
plot_att_dose(ad, "acr")
```

The level effects need only parallel trends against the untreated; the
causal responses need strong parallel trends. The TWFE weights show that the
same regression coefficient can be read as a non-convex average of level
effects or a convex average of slopes.

## Synthetic difference-in-differences

`sdid_weights()` solves for the unit and time weights; the estimate is the
coefficient of a weighted fixest regression.

```{r sdid}
block <- sim_did_panel(n_units = 40, n_periods = 12, groups = 9, never_share = 0.85, mu = 2, seed = 3)
sw <- sdid_weights(block, id = "id", time = "time", y = "y", d = "treated")
sw
fit_sdid <- feols(y ~ treated | id + time, data = merge(block, sw$cell_weights, by = c("id", "time")), weights = ~weight)
c(closed_form = sw$estimate, fixest = coef(fit_sdid)[["treated"]])
sdid_se(sw, method = "placebo", n_reps = 100, seed = 1)[c("estimate", "std.error")]
plot_sdid(sw)
```

## Repeated cross-sections and machine-learning nuisances

```{r rcs}
rcs <- dat[sample(nrow(dat), 2000), ]
rcs$row <- seq_len(nrow(rcs))
att_gt(rcs, id = "row", time = "time", group = "g", y = "y", x = "x1", sampling = "rcs",
       control_group = "never", n_boot = 0)$att_gt[1:4, c("group", "time", "att", "std.error")]
```

```{r learners, eval = run_mlr3}
library(mlr3); library(mlr3learners)
ml <- att_gt(subset(dat, g %in% c(0, 4) & time <= 5), id = "id", time = "time", group = "g", y = "y",
             x = c("x1", "x2"), learner_p = lrn("classif.ranger", predict_type = "prob", num.trees = 200),
             learner_or = lrn("regr.ranger", num.trees = 200), folds = 3, seed = 1, n_boot = 0)
ml$att_gt[, c("group", "time", "att", "std.error")]
```
