---
title: "Using causalmetrics for instrumental variables: 2SLS, control functions, orthogonal scores, weak instruments, compliers, and marginal treatment effects"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using causalmetrics for instrumental variables: 2SLS, control functions, orthogonal scores, weak instruments, compliers, and marginal treatment effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
                      message = FALSE, warning = FALSE)
library(causalmetrics)
library(estimatr)
library(fixest)
library(mlr3)
library(mlr3learners)
lgr::get_logger("mlr3")$set_threshold("warn")
set.seed(1)
run_mlogit <- requireNamespace("mlogit", quietly = TRUE) && requireNamespace("dfidx", quietly = TRUE)
run_ranger <- requireNamespace("ranger", quietly = TRUE)
```

## What the package adds to iv_robust and feols

Two-stage least squares (2SLS), its first stage and reduced form, robust and
clustered standard errors, and fixed effects are the job of
`estimatr::iv_robust()` and `fixest::feols()`. `causalmetrics` adds what
those do not give in one place:

| Task | Function |
|---|---|
| Effective first-stage F (Montiel Olea-Pflueger) next to the usual ones | `iv_first_stage()` |
| Weak-instrument-robust confidence sets (Anderson-Rubin) | `iv_ar_confidence_set()`, `plot_ar_set()` |
| Plausibly exogenous instruments (Conley-Hansen-Rossi) | `iv_plausibly_exogenous()` |
| Control functions: first-stage or generalized residuals, two-step bootstrap, average partial effects | `cf_residuals()`, `cf_bootstrap()`, `cf_ape()` |
| Partially linear IV with machine-learned controls; the LATE with doubly robust scores | `est_dml(model = "pliv")`, `est_dml(model = "iivm")` |
| Complier shares and characteristics (Abadie's kappa) | `complier_profile()` |
| Complier CATE and policies for compliers | `late_scores()`, `late_blp()` |
| What 2SLS with several instruments estimates | `iv_late_weights()` |
| Leave-out leniency instruments | `leniency_instrument()` |
| Marginal treatment effects and policy-relevant averages | `mte_curve()`, `plot_mte()` |
| Shift-share diagnostics: Rotemberg weights, shock-level regression | `ssiv_rotemberg()`, `ssiv_shock_level()` |
| Bounds on the ATE with a binary instrument | `iv_ate_bounds()` |
| Sensitivity to unobserved confounding when there is no instrument | `dml_sensitivity()`, `plot_dml_sensitivity()` |
| Simulated designs with known truth | `sim_iv()` |

All examples use simulated data.

## The standard toolkit, once

The `"linear"` design has a treatment `d` driven by an instrument `z`, an
unobserved confounder, and nonlinear functions of five covariates; the true
effect is 1. The instrument is excluded only conditional on `x1^2` and
`sin(2 x2)`, so these sections include those terms as controls; the DML
section below shows what happens when the functional form is not known.

```{r tsls}
dat <- sim_iv(2000, dgp = "linear", seed = 1)
x <- paste0("x", 1:5)
# the instrument is excluded conditional on x1^2 and sin(2 x2); include them
dat$x1sq <- dat$x1^2
dat$s2x2 <- sin(2 * dat$x2)
xc <- c("x1sq", "s2x2", "x3", "x4", "x5")
ols <- lm_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5, data = dat)
first <- lm_robust(d ~ z + x1sq + s2x2 + x3 + x4 + x5, data = dat)
rf <- lm_robust(y ~ z + x1sq + s2x2 + x3 + x4 + x5, data = dat)
tsls <- iv_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5 | z + x1sq + s2x2 + x3 + x4 + x5, data = dat)
c(ols = coef(ols)[["d"]], first_stage = coef(first)[["z"]], reduced_form = coef(rf)[["z"]],
  wald = coef(rf)[["z"]] / coef(first)[["z"]], tsls = coef(tsls)[["d"]])
```

With `fixest`, the same regression and its first-stage statistics:

```{r fixest}
fe <- feols(y ~ x1sq + s2x2 + x3 + x4 + x5 | d ~ z, data = dat, vcov = "hetero")
coef(fe)["fit_d"]
fitstat(fe, type = c("ivf", "ivwald"))
```

`iv_first_stage()` adds the effective F, which is the statistic to compare
with weak-instrument thresholds when errors are heteroskedastic or
clustered:

```{r first-stage}
iv_first_stage(dat, d = "d", z = "z", x = xc)
```

## Weak instruments

The `"weak"` design sets the first-stage coefficient so that the
concentration parameter is small. The Wald interval is then unreliable; the
Anderson-Rubin set is not.

```{r weak}
weak <- sim_iv(500, dgp = "weak", concentration = 2, seed = 2)
ar <- iv_ar_confidence_set(weak, "y", "d", "z", x = paste0("x", 1:3))
ar
plot_ar_set(ar)
```

The set can be unbounded or the whole line, which is the honest answer when
the data cannot rule out a zero first stage. With a strong instrument it is
close to the Wald interval:

```{r ar-strong}
iv_ar_confidence_set(dat, "y", "d", "z", x = xc)$intervals
```

If the exclusion restriction is only approximately true, the union of
intervals over a range of direct effects of the instrument shows how much
that costs:

```{r plausexog}
iv_plausibly_exogenous(dat, "y", "d", "z", x = xc, gamma_grid = seq(-0.2, 0.2, by = 0.1))
```

## Control functions

Adding the first-stage residual to the second stage reproduces 2SLS in the
linear model. The coefficient on the residual is the endogeneity parameter
and its t test is the Wu-Hausman test.

```{r cf-linear}
cf <- cf_residuals(dat, d = "d", z = "z", x = xc)
cf_ols <- lm_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5 + v_hat, data = cf)
c(control_function = coef(cf_ols)[["d"]], tsls = coef(tsls)[["d"]])
tidy(cf_ols)[tidy(cf_ols)$term == "v_hat", c("estimate", "std.error", "p.value")]
```

The point of the control function is the nonlinear second stage. In the
`"probit_cf"` design the outcome is binary and the endogenous regressor is
continuous. Plugging the fitted first stage into a probit (the "forbidden
regression") is wrong; the control function is right, and the object to
report is the average partial effect computed from the average structural
function. Standard errors come from bootstrapping both stages.

```{r cf-probit}
dp <- sim_iv(2000, dgp = "probit_cf", rho = 0.6, seed = 3)
first_stage <- function(df) cf_residuals(df, d = "d", z = "z", x = "x1")
second_stage <- function(df) glm(y ~ d + x1 + v_hat, data = df, family = binomial("probit"))
aug <- first_stage(dp)
fit_cf <- second_stage(aug)
naive <- glm(y ~ d + x1, data = dp, family = binomial("probit"))
c(truth = attr(dp, "ape"), control_function = as.numeric(cf_ape(fit_cf, aug, d = "d")),
  naive_probit = as.numeric(cf_ape(naive, dp, d = "d")))
cf_bootstrap(dp, first_stage, second_stage, n_boot = 199, seed = 1,
             ape = function(fit, df) cf_ape(fit, df, d = "d"))
```

For a binary endogenous regressor the linear residual is not the right
control; `family = "probit"` returns the generalized residual of a probit
first stage instead:

```{r cf-binary}
dp$d_binary <- as.integer(dp$d > 0)
head(cf_residuals(dp, d = "d_binary", z = "z", x = "x1", family = "probit")$v_hat)
```

### Price endogeneity in a discrete-choice demand model (Petrin and Train)

Petrin and Train (2010) apply the control function to multinomial logit
demand: regress price on cost shifters, put the residual in the utility.
We simulate `T` markets with `J` products, an unobserved product-market
quality shock that raises both price and demand, and a cost shifter `w`
that moves price only. The true price coefficient is `-1`.

```{r pt-data, eval = run_mlogit}
library(mlogit)
library(dfidx)
n_markets <- 150; J <- 3; n_per_market <- 40
mk <- expand.grid(market = seq_len(n_markets), alt = seq_len(J))
mk$xi <- rnorm(nrow(mk), sd = 0.6)                      # unobserved quality shock
mk$w <- rnorm(nrow(mk))                                  # cost shifter
mk$price <- 2 + 0.6 * mk$w + 0.8 * mk$xi + rnorm(nrow(mk), sd = 0.3)
mk$beta_j <- c(0, 0.3, -0.2)[mk$alt]
# consumers
cons <- expand.grid(market = seq_len(n_markets), id = seq_len(n_per_market))
cons$cid <- seq_len(nrow(cons))
long <- merge(cons, mk, by = "market")
long <- long[order(long$cid, long$alt), ]
long$util <- long$beta_j - 1 * long$price + long$xi + -log(-log(runif(nrow(long))))
long$choice <- ave(long$util, long$cid, FUN = function(u) as.integer(u == max(u)))
```

The control function is estimated at the product-market level (the level
at which price is set) and merged back to the consumer-alternative rows.

```{r pt-cf, eval = run_mlogit}
mk_cf <- cf_residuals(transform(mk, alt2 = as.integer(alt == 2), alt3 = as.integer(alt == 3)),
                      d = "price", z = "w", x = c("alt2", "alt3"))
long <- merge(long, mk_cf[, c("market", "alt", "v_hat")], by = c("market", "alt"))
long <- long[order(long$cid, long$alt), ]
dd <- dfidx(long, idx = c("cid", "alt"))
naive_logit <- mlogit(choice ~ price | 1, data = dd)
cf_logit <- mlogit(choice ~ price + v_hat | 1, data = dd)
rbind(naive = coef(naive_logit)[c("price")], control_function = coef(cf_logit)[c("price")])
coef(cf_logit)["v_hat"]
```

The naive logit attributes part of the demand shock to price and
understates price sensitivity; the control function recovers the true
coefficient of `-1`. Standard errors for the two-step estimator come from
`cf_bootstrap()` with the market as the bootstrap cluster:

```{r pt-boot, eval = run_mlogit}
pt_first <- function(df) {
  mk_b <- unique(df[, c("market", "alt", "price", "w")])
  mk_b <- transform(mk_b, alt2 = as.integer(alt == 2), alt3 = as.integer(alt == 3))
  mk_b <- cf_residuals(mk_b, d = "price", z = "w", x = c("alt2", "alt3"))
  df$v_hat <- NULL
  out <- merge(df, mk_b[, c("market", "alt", "v_hat")], by = c("market", "alt"))
  out[order(out$cid, out$alt), ]
}
pt_second <- function(df) {
  df$cid2 <- as.integer(factor(paste(df$cid, df$.boot_draw)))
  coef(mlogit(choice ~ price + v_hat | 1, data = dfidx(df, idx = c("cid2", "alt"))))
}
# cluster bootstrap by market; consumers within a resampled market keep their rows
boot_pt <- function(n_boot, seed) {
  set.seed(seed)
  markets <- unique(long$market)
  draws <- replicate(n_boot, {
    pick <- sample(markets, length(markets), replace = TRUE)
    df <- do.call(rbind, lapply(seq_along(pick), function(k) {
      d <- long[long$market == pick[k], ]; d$market <- k; d$.boot_draw <- k; d
    }))
    pt_second(pt_first(df))[c("price", "v_hat")]
  })
  apply(draws, 1, sd)
}
boot_pt(30, seed = 1)
```

## Orthogonal scores: machine-learned controls

`est_dml(model = "pliv")` residualizes the outcome, the treatment, and the
instrument on the covariates with cross-fitted learners and runs IV on the
residuals. Suppose we did not know that `x1^2` and `sin(2 x2)` are the right
controls and passed the raw covariates: linear learners leave bias that
forests remove.

```{r pliv, eval = run_ranger}
rf <- lrn("regr.ranger", num.trees = 300, min.node.size = 20)
pliv_lin <- est_dml(dat, "y", "d", x, z = "z", model = "pliv", seed = 1, weak_iv = TRUE)
pliv_rf <- est_dml(dat, "y", "d", x, z = "z", model = "pliv",
                   learner_l = rf, learner_m = rf, learner_z = rf, seed = 1, weak_iv = TRUE)
pliv_rf
rbind(tidy(pliv_lin), tidy(pliv_rf))
```

`weak_iv = TRUE` adds the Anderson-Rubin set computed from the cross-fitted
score, which remains valid however weak the residualized instrument is.

## The LATE with doubly robust scores

With a binary instrument and a binary treatment, `model = "iivm"` estimates
the local average treatment effect as the ratio of two AIPW-type scores. The
`"late"` design has always-takers, never-takers, and compliers whose effect
is `2 + x1`.

```{r iivm}
dl <- sim_iv(3000, dgp = "late", seed = 4)
late <- est_dml(dl, "y", "d", c("x1", "x2"), z = "z", model = "iivm", seed = 1)
late
c(true_late = attr(dl, "late"), true_ate = attr(dl, "ate"))
```

Who are the compliers? `complier_profile()` reports the compliance-group
shares and covariate means with Abadie's kappa weights:

```{r compliers}
complier_profile(dl, d = "d", z = "z", covariates = c("x1", "x2"), x = "x1", n_boot = 99, seed = 1)
```

The two score components are pseudo-outcomes, so the heterogeneity tools
apply to compliers. `late_blp()` projects both on a dictionary and forms
the ratio, the complier CATE:

```{r late-blp}
ls <- late_scores(late)
late_blp(ls, ~ x1, newdata = data.frame(x1 = c(-1, 0, 1)))[, c("x1", "estimate", "std.error", "conf.low", "conf.high")]
```

## Several instruments

Two-stage least squares with several instruments is a weighted average of
the instrument-specific Wald estimates. The weights can be negative:

```{r late-weights}
dat$z2 <- dat$z + rnorm(nrow(dat))
dat$z3 <- rnorm(nrow(dat))
iv_late_weights(dat, "y", "d", c("z", "z2", "z3"), x = xc)
```

## Judge leniency

Examiner designs use the leave-out mean decision of the assigned examiner
as the instrument:

```{r judge}
dj <- sim_iv(6000, dgp = "judge", n_judges = 40, seed = 5)
dj <- leniency_instrument(dj, examiner = "judge", d = "d", x = "x1")
iv_first_stage(dj, d = "d", z = "leniency", x = "x1", cluster = "judge")
coef(iv_robust(y ~ d + x1 | leniency + x1, data = dj, clusters = judge))[["d"]]
```

## Marginal treatment effects

With a continuous instrument, `mte_curve()` estimates the propensity score,
fits the outcome as a function of the propensity, and differentiates. The
weights turn the curve into the ATE, ATT, ATU, a LATE for a shift in the
propensity, and a policy-relevant effect.

```{r mte}
dm <- sim_iv(4000, dgp = "mte", seed = 6)
m <- mte_curve(dm, "y", "d", "z", x = "x1", degree = 3, late_range = c(0.3, 0.7),
               policy = function(p) pmin(p + 0.1, 0.99), n_boot = 49, seed = 1)
m
c(true_ate = attr(dm, "ate"), true_att = attr(dm, "att"))
plot_mte(m)
plot_mte(m, "weights")
```

## Shift-share instruments

For a Bartik instrument, `ssiv_rotemberg()` reports which shocks carry the
estimate and `ssiv_shock_level()` the equivalent shock-level regression with
shock-level standard errors:

```{r ssiv}
ds <- sim_iv(600, dgp = "shift_share", n_industries = 15, seed = 7)
shares <- paste0("s", 1:15)
ssiv_rotemberg(ds, "y", "d", shares = shares, shocks = attr(ds, "shocks"), x = "x1")
ssiv_shock_level(ds, "y", "d", shares = shares, shocks = attr(ds, "shocks"), x = "x1")
```

## Bounds and sensitivity

Without the LATE assumptions, the instrument still bounds the ATE of a
binary outcome:

```{r bounds}
dl$y_binary <- as.integer(dl$y > 1)
iv_ate_bounds(dl, "y_binary", "d", "z", n_boot = 99, seed = 1)
```

And without any instrument, `dml_sensitivity()` bounds the omitted-confounder
bias of a partially linear DML estimate in terms of partial R-squareds:

```{r sensitivity}
fit_plr <- est_dml(dat, "y", "d", xc, seed = 1)
s <- dml_sensitivity(fit_plr, r2_y = 0.05, r2_d = 0.03)
s
plot_dml_sensitivity(s, benchmarks = data.frame(r2_d = 0.03, r2_y = 0.05, label = "benchmark"))
```

## References

- Abadie, A. (2003). Semiparametric instrumental variable estimation of treatment response models. *Journal of Econometrics*, 113(2).
- Andrews, I., Stock, J. H., and Sun, L. (2019). Weak instruments in instrumental variables regression. *Annual Review of Economics*, 11.
- Borusyak, K., Hull, P., and Jaravel, X. (2022). Quasi-experimental shift-share research designs. *Review of Economic Studies*, 89(1).
- Chernozhukov, V., Hansen, C., Kallus, N., Spindler, M., and Syrgkanis, V. (2026). *Applied Causal Inference Powered by ML and AI*, chapters 12-13.
- Conley, T. G., Hansen, C. B., and Rossi, P. E. (2012). Plausibly exogenous. *Review of Economics and Statistics*, 94(1).
- Goldsmith-Pinkham, P., Sorkin, I., and Swift, H. (2020). Bartik instruments: what, when, why, and how. *American Economic Review*, 110(8).
- Heckman, J. J. and Vytlacil, E. (2005). Structural equations, treatment effects, and econometric policy evaluation. *Econometrica*, 73(3).
- Mogstad, M., Torgovitsky, A., and Walters, C. R. (2021). The causal interpretation of two-stage least squares with multiple instrumental variables. *American Economic Review*, 111(11).
- Montiel Olea, J. L. and Pflueger, C. (2013). A robust test for weak instruments. *Journal of Business & Economic Statistics*, 31(3).
- Petrin, A. and Train, K. (2010). A control function approach to endogeneity in consumer choice models. *Journal of Marketing Research*, 47(1).
- Wooldridge, J. M. (2015). Control function methods in applied econometrics. *Journal of Human Resources*, 50(2).
