---
title: "Replication: Does Premium Version Adoption in mHealth Improve User Engagement and Health-Related Outcomes?"
subtitle: "Jiang, Uetake, and Yang (Marketing Science 2026): the Callaway-Sant'Anna workflow on a freemium app, with causalmetrics"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/07_did/jiang2026does/"
  n_boot: 999
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "95%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(causalmetrics)
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(kableExtra)
library(fixest)
library(MatchIt)
data_path <- params$data_path
n_boot <- params$n_boot
data_md5 <- tools::md5sum(file.path(data_path, "data", c("simulated_data.csv", "simulated_data_robust_accuracy.csv")))
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
fmt0 <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
paren <- function(x, digits = 3) paste0("(", fmt(x, digits), ")")
est_se <- function(e, s, digits = 3) paste0(fmt(e, digits), " ", paren(s, digits))
ci_of <- function(e, s, digits = 3) paste0("[", fmt(e - qnorm(0.975) * s, digits), ", ", fmt(e + qnorm(0.975) * s, digits), "]")
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
run_dcdh <- requireNamespace("DIDmultiplegtDYN", quietly = TRUE) && requireNamespace("polars", quietly = TRUE)
run_honest <- requireNamespace("HonestDiD", quietly = TRUE)
outcomes6 <- c("track_exercise_avg", "track_food_avg", "within_track_avg",
               "exercisecalories_avg", "lower_initial_avg", "weight_loss_norm_avg")
labels6 <- c("Exercise tracking", "Food tracking", "Within budget",
             "Exercise calories", "Positive weight loss", "Weight loss amount")
names(labels6) <- outcomes6
eng4 <- outcomes6[1:4]
```

# The paper in one page {#sec-paper}

**The question is whether paying for the premium tier of a freemium
fitness app changes what users do and what they achieve.** Jiang, Uetake,
and Yang (2026) observe 11,873 users of a food- and exercise-tracking app
for 15 weeks after registration. Some pay $39.99 for a year of premium
features in weeks 1 to 7; most never do. The outcomes are four engagement
measures (days tracking food, days tracking exercise, days tracking and
staying within the calorie budget, exercise calories logged) and two
health measures (any weight loss relative to the starting weight, pounds
lost).

**The answer is an immediate lift that fades and does not reach the
scale.** Premium adopters track more in the adoption week, the lift
attenuates within about five weeks, and weight loss shows at most a
modest, delayed, and imprecise improvement. A mechanism section splits
adopters by their exposure to the free version before upgrading: the lift
is concentrated among low-exposure users, which fits hedonic decline
(novelty wears off) rather than sunk costs (everyone paid the same fee).

**The methodological point is why this paper anchors Section 07.**
Adoption is staggered and chosen, and later adopters upgrade while their
engagement is already rising. A pooled two-way fixed effects regression
therefore overstates both the size and the persistence of the effects.
The paper's preferred estimator is Callaway and Sant'Anna (2021) with the
doubly robust score, not-yet-treated comparisons, and covariates in both
nuisances, followed by the Rambachan and Roth (2023) sensitivity analysis
and a matched-sample design for time-varying selection. That is exactly
the workflow of the lecture, and every step below runs on the package's
functions: `att_gt()`, `aggregate_att()`, `event_study_frame()`,
`plot_event_study()`, `honestdid_inputs()`, plus `bacon_decomp()`,
`twfe_weights()`, `pretrend_power()`, `did_imputation()`, and
`sdid_weights()` in the extensions.

**What the replication package allows.** The proprietary data are under a
nondisclosure agreement. The authors ship two simulated datasets that
preserve the panel structure and all variables, the complete analysis
code (15 R scripts and one Stata script), and a README that says the
scripts are illustrational: numbers from the simulated data differ from
the paper. The standard for matching in this note is therefore the
authors' scripts run on the simulated data; published numbers are printed
for orientation only and labelled as such. Section \@ref(sec-verify)
verifies the match standard directly.

# Data and setting {#sec-data}

**One row per user-week, with adoption timing, outcomes, and the
covariates the design conditions on.** `weeks_join` counts weeks since
registration (1 to 15). `week_upgrade` is the adoption week; values above
100 mean the user never upgrades in the window. The `first_7_*` columns
are week-1 covariates: engagement, weight, and three national Google
search indexes for app- and weight-loss-related terms.

```{r load}
Data <- read.csv(file.path(data_path, "data", "simulated_data.csv"))
Data$norm_distance <- Data$initial_distance / Data$initialweight
users <- Data %>% arrange(userid, weeks_join) %>% group_by(userid) %>% slice(1) %>% ungroup()
c(users = nrow(users), user_weeks = nrow(Data), weeks = max(Data$weeks_join),
  adopters_w1_7 = sum(users$week_upgrade <= 7), never = sum(users$week_upgrade > 100))
```

**The estimation sample drops week-1 adopters so every treated user has a
pre-period.** Adopters in weeks 2 to 7 plus never-adopters form the DiD
sample used from Table 3 onward. The group variable `g` codes
never-adopters as 0, the package's convention.

```{r sample}
dat <- Data %>% filter((week_upgrade >= 2 & week_upgrade <= 7) | week_upgrade > 100) %>%
  mutate(g = ifelse(week_upgrade > 100, 0L, week_upgrade),
         rel = ifelse(g == 0, -1000L, weeks_join - week_upgrade),
         rel_b = ifelse(rel <= -4, -1000L, pmin(rel, 13L)))
xv <- c("initialweight", "heightinches", "age", "first_day", "ini_goalweight", "ini_goalbmi",
        "ini_bmi", "gender", "norm_distance",
        "first_7_track_food_avg", "first_7_foodcalories_avg", "first_7_track_exercise_avg",
        "first_7_within_track_avg", "first_7_exercisecalories_avg", "first_7_lower_initial_avg",
        "first_7_weight_loss_norm_avg", "first_7_app_weight_loss_search_avg",
        "first_7_app_premium_search_avg", "first_7_weight_loss_premium_search_avg")
table(users$week_upgrade[users$week_upgrade <= 7])
```

## Table 1: summary statistics {#sec-table1}

**Table 1 describes the full sample of 11,873 users, per-user averages
over the 15 weeks.** The published sample is 29 percent male, 42 years
old, 204 pounds at the start, aiming to lose 19 percent of the initial
weight; food is tracked on 58 percent of days. The simulated data are
drawn to preserve these properties.

```{r table1}
sumdat <- Data %>% filter(week_upgrade <= 7 | week_upgrade > 100) %>%
  arrange(userid, weeks_join) %>% group_by(userid) %>%
  mutate(across(c(track_food_avg, track_exercise_avg, within_track_avg,
                  exercisecalories_avg, foodcalories_avg, budgetcalories_avg), mean,
                .names = "{.col}_mean")) %>% slice(1) %>% ungroup()
t1_vars <- c(gender = "Male", age = "Age", initialweight = "Initial weight (lb)",
             ini_goalweight = "Target weight (lb)", initial_distance = "Initial distance to goal weight (lb)",
             norm_distance = "Proportional distance to goal weight",
             budgetcalories_avg = "Suggested budget of calories (kcal)",
             foodcalories_avg_mean = "Food calories logged (kcal)",
             exercisecalories_avg_mean = "Exercise calories logged (kcal)",
             track_food_avg_mean = "Avg. proportion of days tracked food",
             track_exercise_avg_mean = "Avg. proportion of days tracked exercise",
             within_track_avg_mean = "Avg. proportion of days tracked and within budget",
             active_days = "Number of active days")
t1 <- tibble(Variable = unname(t1_vars),
             Mean = sapply(names(t1_vars), function(v) mean(sumdat[[v]], na.rm = TRUE)),
             `Std. dev.` = sapply(names(t1_vars), function(v) sd(sumdat[[v]], na.rm = TRUE)))
t1 %>% mutate(across(where(is.numeric), ~ fmt(.x))) %>%
  bind_rows(tibble(Variable = "N", Mean = fmt0(nrow(sumdat)), `Std. dev.` = "")) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lrr",
      caption = "Table 1 of the paper: summary statistics for the full sample (simulated data; per-user averages over the 15-week window).") %>%
  tab_style()
```

## Figure 1: when users upgrade {#sec-figure1}

**Adoption is front-loaded and staggered, which is the whole
identification problem.** Most upgrades happen in week 1; the histogram
motivates both the staggered design and the decision to drop week-1
adopters, who have no pre-period.

```{r figure1, echo = FALSE, fig.height = 3.2, fig.cap = "Figure 1 of the paper: number of premium upgrades in each week since registration (simulated data)."}
users %>% filter(week_upgrade <= 7) %>% count(week_upgrade) %>%
  ggplot(aes(x = factor(paste0("Week ", week_upgrade), levels = paste0("Week ", 1:7)), y = n)) +
  geom_col(fill = "#c9c58a", width = 0.95) +
  labs(x = "Week Upgrade", y = "Frequency") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

## Table 2: who upgrades {#sec-table2}

**Table 2 profiles adopters with a linear probability model and shows
why selection is the threat.** The published estimates say adopters are
older, aim to lose a larger share of their weight, log more exercise
calories in week 1, and are less likely to have tracked food consistently
in week 1; the app-name premium search index predicts upgrading. The
regression uses classical standard errors, as the authors' `lm` call
does.

```{r table2}
f1 <- ever_upgrade ~ initialweight + heightinches + age + first_day + ini_goalweight +
  ini_goalbmi + ini_bmi + norm_distance + gender + first_7_track_food_avg +
  first_7_foodcalories_avg + first_7_track_exercise_avg + first_7_within_track_avg +
  first_7_exercisecalories_avg + first_7_lower_initial_avg + first_7_weight_loss_norm_avg
f2 <- update(f1, . ~ . + first_7_app_premium_search_avg + first_7_app_weight_loss_search_avg +
               first_7_weight_loss_premium_search_avg)
users_r <- users %>% filter((week_upgrade >= 2 & week_upgrade <= 7) | week_upgrade > 100)
m1 <- feols(f1, data = users, vcov = "iid")
m2 <- feols(f2, data = users_r, vcov = "iid")
cm <- c(initialweight = "Initial weight", heightinches = "Height", age = "Age",
        first_day = "Start date", ini_goalweight = "Initial goal weight",
        ini_goalbmi = "Initial goal BMI", ini_bmi = "Initial BMI",
        norm_distance = "Prop. distance to goal weight", gender = "Male",
        first_7_track_food_avg = "Prop. days tracked food (wk 1)",
        first_7_foodcalories_avg = "Food calories (wk 1)",
        first_7_track_exercise_avg = "Prop. days tracked exercise (wk 1)",
        first_7_within_track_avg = "Prop. days within budget (wk 1)",
        first_7_exercisecalories_avg = "Exercise calories (wk 1)",
        first_7_lower_initial_avg = "Prop. days below initial weight (wk 1)",
        first_7_weight_loss_norm_avg = "Avg. weight change (wk 1)",
        first_7_app_premium_search_avg = "App premium searches (wk 1)",
        first_7_app_weight_loss_search_avg = "App weight loss searches (wk 1)",
        first_7_weight_loss_premium_search_avg = "Weight loss premium searches (wk 1)")
modelsummary::modelsummary(list("Full sample" = m1, "Restricted sample" = m2),
  output = "kableExtra", coef_map = cm, stars = c(`*` = 0.10, `**` = 0.05, `***` = 0.01),
  gof_map = list(list(raw = "nobs", clean = "Observations", fmt = 0),
                 list(raw = "r.squared", clean = "R2", fmt = 3)),
  fmt = 3, title = "Table 2 of the paper: linear probability model of upgrading within seven weeks (classical standard errors, as in the authors' script). The restricted sample is the DiD sample: adopters in weeks 2-7 and never-adopters.") %>%
  tab_style("scale_down")
```

# The staggered DiD design {#sec-design}

**The estimand is the group-time average treatment effect and everything
else is an aggregation of it.** For adoption cohort $g$ and week $t$,
$ATT(g,t) = E[Y_t(1) - Y_t(0) \mid G = g]$. Callaway and Sant'Anna
(2021) estimate each cell by a doubly robust comparison of the outcome
change from $g-1$ to $t$ between cohort $g$ and users not yet treated at
$t$, with the covariates entering a propensity score and an outcome
regression. Under limited anticipation and conditional parallel trends,
either nuisance being correct suffices. Dynamic aggregation over event
time $e = t - g$ gives the event studies of Figures 2 and 3; the simple
aggregation gives the overall ATT of Table 3; the multiplier bootstrap
gives simultaneous bands.

**The specification, exactly as in the authors' scripts.**

- Sample: adopters in weeks 2-7 plus never-adopters (`r fmt0(length(unique(dat$userid)))` users).
- Comparison group: not yet treated (never-adopters plus later cohorts).
- Method: doubly robust; logistic propensity, linear outcome regression.
- Covariates: demographics, start date, initial goals and BMI,
  proportional distance to goal, week-1 engagement and weight, week-1
  search indexes (19 variables).
- Benchmark: TWFE event study with user and week fixed effects, dummies
  for event times $-3$ to $12$ and $13{+}$, reference: never-adopters and
  event times $\le -4$; pooled TWFE with a single post indicator.
  Standard errors clustered by user everywhere.

## Verifying the match standard {#sec-verify}

**Our `att_gt()` reproduces the `did` package on this data to every
digit, and the TWFE helper reproduces the authors' dummy regression
exactly.** The chunk runs both packages on food tracking and compares;
it also compares `fixest::i()` with the authors' hand-built dummies.

```{r verify, cache = TRUE, cache.extra = data_md5}
a_cm <- att_gt(dat, id = "userid", time = "weeks_join", group = "g", y = "track_food_avg",
               x = xv, method = "dr", control_group = "notyet", n_boot = 0)
s_cm <- aggregate_att(a_cm, type = "simple", n_boot = 0)
b_did <- suppressWarnings(did::att_gt(yname = "track_food_avg", tname = "weeks_join",
  idname = "userid", gname = "week_upgrade", xformla = reformulate(xv), data = dat,
  control_group = "notyettreated", est_method = "dr"))
s_did <- did::aggte(b_did, type = "simple")
tw_i <- feols(track_food_avg ~ i(rel_b, ref = -1000) | userid + weeks_join, data = dat, cluster = ~userid)
da <- dat
for (k in 0:12) da[[paste0("u_", k)]] <- as.integer(da$rel == k)
da$u_13a <- as.integer(da$rel >= 13 & da$g != 0)
for (k in 1:3) da[[paste0("u_p", k)]] <- as.integer(da$rel == -k)
tw_a <- feols(reformulate(c(paste0("u_p", 3:1), paste0("u_", 0:12), "u_13a"),
                          response = "track_food_avg"), data = da,
              fixef = c("userid", "weeks_join"), cluster = ~userid)
c(att_gt_vs_did = abs(s_cm$overall$estimate - s_did$overall.att),
  twfe_i_vs_dummies = max(abs(sort(coef(tw_i)) - sort(coef(tw_a)))))
```

**Both differences are zero to machine precision.** From here on, every
Callaway-Sant'Anna object is the package's, and every published number is
quoted for orientation.

```{r helpers, include = FALSE}
# TWFE event study in the authors' specification; returns a fixest fit.
tw_es <- function(y, data) {
  feols(as.formula(paste0(y, " ~ i(rel_b, ref = -1000) | userid + weeks_join")),
        data = data, cluster = ~userid)
}
# Event-study plot in the paper's style: pre in salmon, post in blue.
paper_es <- function(frame, title, band = FALSE, ylab = "Estimates") {
  frame$period <- factor(ifelse(frame$event_time < 0, "Pre", "Post"), levels = c("Pre", "Post"))
  lo <- if (band && all(!is.na(frame$band.low))) frame$band.low else frame$conf.low
  hi <- if (band && all(!is.na(frame$band.high))) frame$band.high else frame$conf.high
  ggplot(frame, aes(x = event_time, y = estimate, colour = period)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_point(size = 1.4) +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.25, linewidth = 0.4) +
    scale_colour_manual(values = c(Pre = "salmon", Post = "blue"), name = NULL) +
    scale_x_continuous(breaks = seq(min(frame$event_time), max(frame$event_time), by = 2)) +
    labs(title = title, x = "Time Relative to Treatment", y = ylab) +
    theme_minimal(base_size = 9) +
    theme(plot.title = element_text(size = 8.5, face = "bold"), legend.position = "right",
          panel.grid.minor = element_blank())
}
frame_tw <- function(fit) event_study_frame(TWFE = fit)
frame_cs <- function(dyn) event_study_frame(CS = dyn)
```

# Results: event studies and the overall ATT {#sec-results}

```{r cs-main, cache = TRUE, cache.extra = list(data_md5, n_boot)}
cs_main <- lapply(setNames(outcomes6, outcomes6), function(y)
  att_gt(dat, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
         method = "dr", control_group = "notyet", n_boot = n_boot, seed = 678))
dyn_main <- lapply(cs_main, aggregate_att, type = "dynamic", min_e = -3, seed = 678)
simple_main <- lapply(cs_main, aggregate_att, type = "simple", seed = 678)
tw_main <- lapply(setNames(outcomes6, outcomes6), function(y) tw_es(y, dat))
```

## Figures 2 and 3: TWFE against Callaway-Sant'Anna {#sec-fig23}

**The two columns of each figure are the paper's argument in one
picture.** The left column is the TWFE event study: a sharp rise at
adoption that appears to persist for three months. The right column is
the Callaway-Sant'Anna event study with simultaneous bands: the same
sharp rise, but it attenuates within weeks. The published reading is that
TWFE overstates persistence because later adopters upgrade on rising
trajectories and contaminate the comparisons.

```{r figure2, echo = FALSE, fig.height = 11.5, fig.cap = "Figure 2 of the paper: impact of premium adoption on the four engagement outcomes. Left column: TWFE event studies (pointwise 95 percent intervals, clustered by user; reference: never-adopters and event times at or below -4). Right column: Callaway-Sant'Anna doubly robust event studies with not-yet-treated comparisons and simultaneous confidence bands."}
plots2 <- list()
for (y in eng4) {
  plots2[[length(plots2) + 1]] <- paper_es(frame_tw(tw_main[[y]]), paste0("TWFE: ", labels6[y]))
  plots2[[length(plots2) + 1]] <- paper_es(frame_cs(dyn_main[[y]]), paste0("Staggered DiD: ", labels6[y]), band = TRUE)
}
wrap_plots(plots2, ncol = 2)
```

```{r figure3, echo = FALSE, fig.height = 6, fig.cap = "Figure 3 of the paper: impact of premium adoption on the two health outcomes, same layout as Figure 2."}
plots3 <- list()
for (y in outcomes6[5:6]) {
  plots3[[length(plots3) + 1]] <- paper_es(frame_tw(tw_main[[y]]), paste0("TWFE: ", labels6[y]))
  plots3[[length(plots3) + 1]] <- paper_es(frame_cs(dyn_main[[y]]), paste0("Staggered DiD: ", labels6[y]), band = TRUE)
}
wrap_plots(plots3, ncol = 2)
```

```{r fig23-numbers, echo = FALSE}
e0 <- sapply(dyn_main, function(d) d$by$estimate[d$by$event_time == 0])
e5 <- sapply(dyn_main, function(d) d$by$estimate[d$by$event_time == 5])
e10 <- sapply(dyn_main, function(d) d$by$estimate[d$by$event_time == 10])
```

**On the simulated data the levels and the TWFE-CS ordering reproduce;
the decay is clear for the exercise outcomes and slower for the food
outcomes.** Exercise tracking falls from `r fmt(e0["track_exercise_avg"])`
in the adoption week to `r fmt(e5["track_exercise_avg"])` five weeks
later, and exercise calories from `r fmt(e0["exercisecalories_avg"], 1)`
to `r fmt(e5["exercisecalories_avg"], 1)`. Food tracking starts at
`r fmt(e0["track_food_avg"])`, is still `r fmt(e5["track_food_avg"])` at
week five, and declines to `r fmt(e10["track_food_avg"])` by week ten;
budget adherence behaves like food tracking. The published estimates
attenuate faster (food tracking insignificant after week five), so this
simulated draw damps the decay without changing its direction. The
health outcomes show wide bands and no effect covered by the band at any
horizon.

## Table 3: overall ATT, pooled TWFE against Callaway-Sant'Anna {#sec-table3}

**The overall ATT makes the overstatement quantitative.** The pooled
TWFE coefficient (one post indicator, user and week fixed effects) is
compared with the simple aggregation of the group-time effects.

```{r table3}
tw_pool <- lapply(setNames(outcomes6, outcomes6), function(y)
  feols(as.formula(paste0(y, " ~ upgraded | userid + weeks_join")), data = dat, cluster = ~userid))
row_tw <- function(y) {
  ct <- coeftable(tw_pool[[y]])["upgraded", ]
  tibble(Variable = labels6[y], ATT = fmt(ct[1]), `Std. error` = fmt(ct[2]), `95\\% CI` = ci_of(ct[1], ct[2]))
}
row_cs <- function(y) {
  o <- simple_main[[y]]$overall
  tibble(Variable = labels6[y], ATT = fmt(o$estimate), `Std. error` = fmt(o$std.error), `95\\% CI` = ci_of(o$estimate, o$std.error))
}
pub3 <- tibble(Variable = labels6,
  `Published TWFE` = c("0.124", "0.163", "0.124", "47.657", "0.036", "1.584"),
  `Published CS` = c("0.074", "0.060", "0.048", "32.183", "0.026", "0.487"))
bind_rows(bind_rows(lapply(outcomes6, row_tw)), bind_rows(lapply(outcomes6, row_cs))) %>%
  left_join(pub3, by = "Variable") %>%
  mutate(`Published ATT` = c(pub3$`Published TWFE`, pub3$`Published CS`)) %>%
  select(Variable, ATT, `Std. error`, `95\\% CI`, `Published ATT`) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, align = "lcccc",
      caption = "Table 3 of the paper: overall ATT estimates from the pooled TWFE regression and from the simple aggregation of the Callaway-Sant'Anna group-time effects, standard errors clustered by user. The published column shows the proprietary-data estimates for orientation.") %>%
  pack_rows("Panel A: pooled TWFE estimates", 1, 6) %>%
  pack_rows("Panel B: Callaway and Sant'Anna (2021) estimates", 7, 12) %>%
  tab_style()
```

```{r table3-numbers, echo = FALSE}
gap <- sapply(outcomes6, function(y) coef(tw_pool[[y]])[["upgraded"]] / simple_main[[y]]$overall$estimate)
gap5 <- gap[setdiff(outcomes6, "lower_initial_avg")]
```

**TWFE is larger for the four engagement outcomes and the weight-loss
amount, here by factors of `r fmt(min(gap5), 1)` to
`r fmt(max(gap5), 1)`.** Positive weight loss is small and imprecise in
both panels, and its two estimates are indistinguishable. In the paper
the ratios run from about 1.5 (exercise calories) to 3.3 (weight loss
amount). Extension \@ref(sec-ext-bacon) decomposes where the gap comes
from.

# Sensitivity and robustness {#sec-robust}

## Figure 4: overlap {#sec-figure4}

**The overlap assumption is checked by the propensity-score densities by
cohort.** The score is a pooled logistic regression of the premium
indicator on the design covariates plus first-week weight, as in the
authors' script; nondegenerate, overlapping densities support the
weighting.

```{r figure4, fig.height = 3.8, fig.cap = "Figure 4 of the paper: density of the predicted propensity score across adoption cohorts (pooled logistic regression on the design covariates plus first-week weight)."}
ps_x <- c(xv, "first_7_weight_avg")
ps_fit <- glm(reformulate(ps_x, response = "upgraded"), data = dat, family = binomial())
dat$pscore <- predict(ps_fit, newdata = dat, type = "response")
dat$cohort_lab <- ifelse(dat$g == 0, "Never", paste0("Week ", dat$g))
ggplot(dat, aes(x = pscore, colour = cohort_lab)) +
  geom_density() +
  scale_colour_manual(values = c("Week 2" = "orange", "Week 3" = "blue", "Week 4" = "brown",
                                 "Week 5" = "green", "Week 6" = "pink", "Week 7" = "darkgreen",
                                 "Never" = "purple"), name = "Upgrade Week") +
  coord_cartesian(xlim = c(0, 0.1), ylim = c(0, 60)) +
  labs(x = "Predicted Propensity Score", y = "Density") +
  theme_classic(base_size = 11) + theme(legend.position = "top")
```

## Figures 5 and 6: pre-adoption trajectories {#sec-fig56}

**The raw trajectories are the selection story.** Mean outcomes by week
since registration, one line per adoption cohort (including week-1
adopters) plus never-adopters, with each cohort's adoption week marked.
In the paper, later adopters ride rising engagement into their upgrade
and never-adopters drift down; weight-loss trajectories rise for
everyone, steepest for late adopters.

```{r figure56, echo = FALSE, fig.height = 8.5, fig.cap = "Figures 5 and 6 of the paper: pre-adoption trajectories. Mean outcome by week since registration for each adoption cohort and for never-adopters; dashed vertical lines mark adoption weeks."}
traj <- Data %>%
  mutate(cohort_lab = case_when(week_upgrade <= 7 ~ paste0("Week ", week_upgrade),
                                week_upgrade > 100 ~ "Never Adopters")) %>%
  filter(!is.na(cohort_lab))
cols_traj <- c("Week 1" = "red", "Week 2" = "orange", "Week 3" = "blue", "Week 4" = "brown",
               "Week 5" = "green", "Week 6" = "pink", "Week 7" = "darkgreen",
               "Never Adopters" = "purple")
traj_plot <- function(y) {
  traj %>% group_by(weeks_join, cohort_lab) %>%
    summarise(m = mean(.data[[y]], na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = weeks_join, y = m, colour = cohort_lab)) +
    geom_line(linewidth = 0.5) +
    geom_vline(xintercept = 1:7, linetype = "dashed", colour = cols_traj[1:7], linewidth = 0.3) +
    scale_colour_manual(values = cols_traj, name = NULL) +
    scale_x_continuous(breaks = seq(1, 15, 2)) +
    labs(x = "Weeks Since Joined", y = labels6[y]) +
    theme_minimal(base_size = 8.5) + theme(panel.grid.minor = element_blank())
}
wrap_plots(lapply(outcomes6, traj_plot), ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
```

## Figure 7: Rambachan-Roth sensitivity {#sec-figure7}

**The sensitivity analysis asks how large a violation of conditional
parallel trends the significant effects survive.** The relative-magnitudes
family bounds the post-treatment violation by $\bar M$ times the largest
pre-treatment violation; the conditional least-favorable hybrid interval
is computed for the week-of-adoption effect over a grid of $\bar M$.
`att_gt()` is re-run with the universal base period, so every pre-period
coefficient is a genuine pre-trend estimate, and `honestdid_inputs()`
hands the coefficients and their clustered covariance to HonestDiD. The
published thresholds are $\bar M \le 1.6$ for exercise tracking,
$\le 1.1$ for food tracking, and $\le 1.0$ for the other two engagement
outcomes.

```{r cs-universal, cache = TRUE, cache.extra = data_md5}
cs_uni <- lapply(setNames(eng4, eng4), function(y)
  att_gt(dat, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
         method = "dr", control_group = "notyet", base_period = "universal", n_boot = 0))
dyn_uni <- lapply(cs_uni, aggregate_att, type = "dynamic", min_e = -4, max_e = 3, n_boot = 0)
```

```{r honest, cache = TRUE, cache.extra = data_md5, eval = run_honest}
honest_one <- function(dyn, mbars = seq(0.5, 1.8, by = 0.1)) {
  hi <- honestdid_inputs(dyn)
  l1 <- HonestDiD::basisVector(1, hi$numPostPeriods)
  orig <- HonestDiD::constructOriginalCS(betahat = hi$betahat, sigma = hi$sigma,
    numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods, l_vec = l1)
  rob <- HonestDiD::createSensitivityResults_relativeMagnitudes(betahat = hi$betahat,
    sigma = hi$sigma, numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods,
    Mbarvec = mbars, l_vec = l1, gridPoints = 100)
  list(orig = as.data.frame(orig), rob = as.data.frame(rob))
}
honest_res <- lapply(dyn_uni, honest_one)
```

```{r figure7, echo = FALSE, eval = run_honest, fig.height = 7, fig.cap = "Figure 7 of the paper: conditional least-favorable hybrid confidence intervals for the week-of-adoption effect under relative-magnitudes violations of size Mbar times the largest pre-treatment violation. The leftmost red interval is the original confidence interval under conditional parallel trends."}
honest_plot <- function(hr, title) {
  rob <- hr$rob; orig <- hr$orig
  df <- bind_rows(tibble(M = min(rob$Mbar) - 0.15, lb = orig$lb, ub = orig$ub, which = "Original"),
                  tibble(M = rob$Mbar, lb = rob$lb, ub = rob$ub, which = "C-LF"))
  ggplot(df, aes(x = M, colour = which)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = lb, ymax = ub), width = 0.04, linewidth = 0.5) +
    scale_colour_manual(values = c("C-LF" = "#0072B2", "Original" = "#D55E00"), name = NULL) +
    labs(title = title, x = "Mbar", y = "Effect at event time 0") +
    theme_minimal(base_size = 9) + theme(plot.title = element_text(size = 9, face = "bold"))
}
wrap_plots(Map(honest_plot, honest_res, labels6[eng4]), ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
```

```{r figure7-numbers, echo = FALSE, eval = run_honest}
last_sig <- sapply(honest_res, function(hr) {
  sig <- hr$rob$lb > 0
  if (all(sig)) Inf else if (any(sig)) max(hr$rob$Mbar[sig]) else 0
})
last_txt <- ifelse(is.infinite(last_sig), "beyond the plotted 1.8", fmt(last_sig, 1))
```

**On the simulated data the week-of-adoption effects stay significant up
to `r paste(paste0(labels6[eng4], ": ", last_txt), collapse = "; ")`.**
The reading is the paper's: the effects survive violations comparable to
the worst pre-period deviation, and extreme violations erase them.
Extension \@ref(sec-ext-honest) extends the exercise to more horizons and
the smoothness family, on a finer grid.

## Figure 8: matched sample for time-varying selection {#sec-figure8}

**The matched design conditions on recent trajectories, not only week-1
levels.** Adopters in weeks 5 and 6 are matched one-to-ten to
never-adopters on demographics, goals, week-1 engagement, and weeks 3-4
engagement, weight, and search intensities, with a generalized additive
model propensity (the authors' seed 123). A TWFE event study on the
matched panel then adjusts for what the matching balanced.

```{r figure8-match, cache = TRUE, cache.extra = data_md5}
recent <- Data %>% arrange(userid, weeks_join) %>% group_by(userid) %>%
  mutate(across(c(track_food_avg, foodcalories_avg, track_exercise_avg, within_track_avg,
                  exercisecalories_avg, lower_initial_avg, weight_loss_norm_avg, weight_avg,
                  app_premium_search_avg, app_weight_loss_search_avg, weight_loss_premium_search_avg),
                list(w3 = ~nth(.x, 3), w4 = ~nth(.x, 4), w5 = ~nth(.x, 5)),
                .names = "{.col}_{.fn}")) %>% ungroup()
sub8 <- recent %>% group_by(userid) %>%
  filter(week_upgrade %in% c(5, 6) | ever_upgrade == 0) %>% slice(1) %>% ungroup()
f8 <- reformulate(c("initialweight", "heightinches", "age", "first_day", "ini_goalweight",
  "ini_goalbmi", "ini_bmi", "gender", "norm_distance",
  "first_7_track_food_avg", "first_7_foodcalories_avg", "first_7_track_exercise_avg",
  "first_7_within_track_avg", "first_7_exercisecalories_avg",
  paste0(c("track_food_avg", "foodcalories_avg", "track_exercise_avg", "within_track_avg",
           "exercisecalories_avg", "lower_initial_avg", "weight_loss_norm_avg", "weight_avg",
           "app_weight_loss_search_avg", "app_premium_search_avg", "weight_loss_premium_search_avg"),
         rep(c("_w3", "_w4"), each = 11))), response = "ever_upgrade")
set.seed(123)
m8 <- matchit(f8, data = sub8, method = "nearest", distance = "gam", ratio = 10, replacement = FALSE)
ids8 <- match.data(m8)$userid
dat8 <- Data %>% filter(userid %in% ids8) %>%
  mutate(g = ifelse(week_upgrade > 100, 0L, week_upgrade),
         rel = ifelse(g == 0, -1000L, weeks_join - week_upgrade),
         rel_b = ifelse(rel <= -4, -1000L, pmin(rel, 13L)))
c(adopters = sum(sub8$ever_upgrade), matched_users = length(ids8))
```

```{r figure8, echo = FALSE, fig.height = 8.5, fig.cap = "Figure 8 of the paper: TWFE event studies on the matched sample (adopters in weeks 5-6 matched 1:10 to never-adopters on demographics, goals, week-1 engagement, and weeks 3-4 engagement, weight, and search intensities). Pointwise 95 percent intervals, clustered by user."}
tw8 <- lapply(setNames(outcomes6, outcomes6), function(y) tw_es(y, dat8))
wrap_plots(lapply(outcomes6, function(y)
  paper_es(frame_tw(tw8[[y]]), paste0("Matched: ", labels6[y]))), ncol = 2)
```

**The immediate lift and the attenuation survive the matching, and the
pre-periods flatten.** That is the published conclusion: the response is
not only a motivational surge that jointly causes upgrading and
engagement. Extension \@ref(sec-ext-matchcs) re-estimates this design
with the main estimator instead of TWFE.

# Mechanisms: exposure to the free version {#sec-mechanism}

## Tables 4 and 5: the exposure split {#sec-table45}

**Hedonic decline and sunk costs both predict the immediate lift; only
hedonic decline predicts that prior exposure dampens it.** Adopters are
split at the median days of tracking before the upgrade. Table 4 shows
the two groups are otherwise comparable. Table 5 estimates the
Callaway-Sant'Anna effects separately for each group against the
never-adopters, at the adoption week and one week after.

```{r table4}
med_track <- median(users$total_track_before[users$week_upgrade >= 2 & users$week_upgrade <= 7])
never_ids <- users$userid[users$week_upgrade > 100]
dat_exp <- dat %>% mutate(distance_goal = weight_avg - goalweight)
adopt_at <- dat_exp %>% filter(weeks_join == week_upgrade)
low_at <- adopt_at %>% filter(total_track_before < med_track)
high_at <- adopt_at %>% filter(total_track_before >= med_track)
t4_vars <- c(total_track_before = "Number of prior days tracked food/exercise", gender = "Male",
             age = "Age", initialweight = "Initial weight (lb)", heightinches = "Height (inch)",
             ini_goalweight = "Initial goal weight (lb)", ini_goalbmi = "Initial goal BMI",
             weight_avg = "Current weight (lb)", distance_goal = "Current distance to goal (lb)")
t4 <- bind_rows(lapply(names(t4_vars), function(v) {
  tt <- t.test(low_at[[v]], high_at[[v]])
  tibble(Variable = t4_vars[v], `Low exposure` = fmt(mean(low_at[[v]], na.rm = TRUE)),
         `High exposure` = fmt(mean(high_at[[v]], na.rm = TRUE)),
         `t-statistic` = fmt(tt$statistic), `p-value` = fmt(tt$p.value))
}))
t4 %>% bind_rows(tibble(Variable = "Number of observations", `Low exposure` = fmt0(nrow(low_at)),
                        `High exposure` = fmt0(nrow(high_at)), `t-statistic` = "", `p-value` = "")) %>%
  kbl(booktabs = TRUE, linesep = "",
      caption = "Table 4 of the paper: comparison of means between low- and high-exposure adopters (median split on days of food or exercise tracking before the upgrade), with two-sample t-tests.") %>%
  tab_style()
```

```{r table5, cache = TRUE, cache.extra = data_md5}
dat_low <- dat %>% filter((total_track_before < med_track & g > 0) | userid %in% never_ids)
dat_high <- dat %>% filter((total_track_before >= med_track & g > 0) | userid %in% never_ids)
t5_run <- function(data) lapply(setNames(eng4, eng4), function(y) {
  a <- att_gt(data, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
              method = "dr", control_group = "notyet", n_boot = 499, seed = 1234)
  aggregate_att(a, type = "dynamic", min_e = -3, na.rm = TRUE, seed = 1234)$by
})
t5_low <- t5_run(dat_low)
t5_high <- t5_run(dat_high)
```

```{r table5-table, echo = FALSE}
t5_row <- function(by, lab) {
  w0 <- by[by$event_time == 0, ]; w1 <- by[by$event_time == 1, ]
  tibble(Outcome = lab, `Estimate (wk 0)` = fmt(w0$estimate), `SE (wk 0)` = fmt(w0$std.error),
         `Estimate (wk 1)` = fmt(w1$estimate), `SE (wk 1)` = fmt(w1$std.error))
}
bind_rows(lapply(eng4, function(y) bind_rows(
  t5_row(t5_low[[y]], paste0(labels6[y], " (low exposure)")),
  t5_row(t5_high[[y]], paste0(labels6[y], " (high exposure)"))))) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcccc",
      col.names = c("Outcome", "Estimate", "Std. error", "Estimate", "Std. error"),
      caption = "Table 5 of the paper: heterogeneous effects by pre-adoption exposure. Callaway-Sant'Anna dynamic effects at the adoption week and one week after, estimated separately for low- and high-exposure adopters against never-adopters.") %>%
  add_header_above(c(" " = 1, "Immediate (week 0)" = 2, "Short term (week 1)" = 2)) %>%
  tab_style()
```

```{r table5-numbers, echo = FALSE}
g50 <- function(l, y, e) l[[y]]$estimate[l[[y]]$event_time == e]
```

**Low-exposure adopters respond more, as the paper reports.** Food
tracking one week after upgrading rises by `r fmt(g50(t5_low, "track_food_avg", 1))`
for low-exposure adopters and `r fmt(g50(t5_high, "track_food_avg", 1))`
for high-exposure adopters (published: 0.167 against 0.041, a wider gap
than this simulated draw produces). The same ordering holds for every
engagement outcome, most sharply for exercise calories. The sunk cost is
the same $39.99 for everyone, so the asymmetry points to hedonic
decline.

## Figures 9 and 10: matched samples by exposure {#sec-fig910}

**The sharpest version matches each exposure group to its own controls.**
Week-5 adopters are split at their own median, each group is matched to
never-adopters on recent (weeks 3-4) engagement and search (the authors'
probit, ratio 2, seed 123 for the low group; GAM, ratio 1, seed 234 for
the high group), and TWFE event studies run per group.

```{r fig910-match, cache = TRUE, cache.extra = data_md5}
med5 <- median(users$total_track_before[users$week_upgrade == 5])
f910 <- reformulate(c("initialweight", "heightinches", "age", "first_day", "ini_goalweight",
  "ini_goalbmi", "ini_bmi", "gender",
  "track_food_avg_w3", "track_exercise_avg_w3",
  paste0(c("track_food_avg", "foodcalories_avg", "track_exercise_avg", "within_track_avg",
           "exercisecalories_avg", "lower_initial_avg", "weight_loss_norm_avg",
           "app_weight_loss_search_avg", "app_premium_search_avg", "weight_loss_premium_search_avg"),
         "_w4")), response = "ever_upgrade")
match_ids <- function(cut_low, seed, dist, ratio, link = NULL) {
  sub <- recent %>% group_by(userid) %>%
    filter(week_upgrade == 5 | ever_upgrade == 0) %>% slice(1) %>% ungroup() %>%
    filter(if (cut_low) total_track_before < med5 else total_track_before >= med5)
  set.seed(seed)
  args <- list(f910, data = sub, method = "nearest", distance = dist, ratio = ratio, replacement = FALSE)
  if (!is.null(link)) args$link <- link
  match.data(do.call(matchit, args), data = sub)$userid
}
ids_low5 <- match_ids(TRUE, 123, "glm", 2, link = "probit")
ids_high5 <- match_ids(FALSE, 234, "gam", 1)
panel_of <- function(ids) Data %>% filter(userid %in% ids) %>%
  mutate(g = ifelse(week_upgrade > 100, 0L, week_upgrade),
         rel = ifelse(g == 0, -1000L, weeks_join - week_upgrade),
         rel_b = ifelse(rel <= -4, -1000L, pmin(rel, 13L)))
pan_low5 <- panel_of(ids_low5); pan_high5 <- panel_of(ids_high5)
c(low = length(ids_low5), high = length(ids_high5))
```

```{r figure910, echo = FALSE, fig.height = 11.5, fig.cap = "Figures 9 and 10 of the paper: TWFE event studies for week-5 adopters split at the median of pre-adoption exposure, each group matched to never-adopters on demographics, goals, and weeks 3-4 engagement, weight-loss, and search variables. Left column: low exposure; right column: high exposure."}
p910 <- list()
for (y in eng4) {
  p910[[length(p910) + 1]] <- paper_es(frame_tw(tw_es(y, pan_low5)), paste0("Low exposure: ", labels6[y]))
  p910[[length(p910) + 1]] <- paper_es(frame_tw(tw_es(y, pan_high5)), paste0("High exposure: ", labels6[y]))
}
wrap_plots(p910, ncol = 2)
```

**Only the low-exposure group shows the clear lift, as published.** The
high-exposure panels sit near zero throughout. The same design for
week-6 adopters (published as Online Figures D1 and D2) is reproduced in
Section \@ref(sec-appendix-d).

# Appendix replications {#sec-appendix}

## Appendix A: alternative sample {#sec-appendix-a}

**The published check re-runs everything on weeks 2-6 adopters and a
3 percent random sample of never-adopters.** The point is that the
baseline pattern is not driven by the week-7 cohort (19 users here) or by
the size of the control pool.

```{r appendix-a, cache = TRUE, cache.extra = data_md5}
set.seed(123)
never_3pct <- sample(never_ids, round(0.03 * length(never_ids)))
dat_a <- dat %>% filter((g >= 2 & g <= 6) | userid %in% never_3pct)
cs_a <- lapply(setNames(outcomes6, outcomes6), function(y)
  att_gt(dat_a, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
         method = "dr", control_group = "notyet", n_boot = 499, seed = 678))
dyn_a <- lapply(cs_a, aggregate_att, type = "dynamic", min_e = -3, seed = 678)
```

```{r appendix-a-fig, echo = FALSE, fig.height = 8.5, fig.cap = "Appendix Figures A1-A2 of the paper (Callaway-Sant'Anna panels): event studies on the alternative sample of weeks 2-6 adopters and a 3 percent random sample of never-adopters."}
wrap_plots(lapply(outcomes6, function(y)
  paper_es(frame_cs(dyn_a[[y]]), paste0("Alt. sample: ", labels6[y]), band = TRUE)), ncol = 2)
```

## Appendix B1: measurement-thresholded outcomes {#sec-appendix-b}

**The second simulated file applies physiologically credible thresholds
to food and weight logs.** The Callaway-Sant'Anna overall ATTs on that
file sit next to the baseline Table 3 panel.

```{r appendix-b, cache = TRUE, cache.extra = data_md5}
DataB <- read.csv(file.path(data_path, "data", "simulated_data_robust_accuracy.csv"))
DataB$norm_distance <- DataB$initial_distance / DataB$initialweight
dat_b <- DataB %>% filter((week_upgrade >= 2 & week_upgrade <= 7) | week_upgrade > 100) %>%
  mutate(g = ifelse(week_upgrade > 100, 0L, week_upgrade))
simple_b <- lapply(setNames(outcomes6, outcomes6), function(y) {
  a <- att_gt(dat_b, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
              method = "dr", control_group = "notyet", n_boot = 499, seed = 678)
  aggregate_att(a, type = "simple", seed = 678)$overall
})
```

```{r appendix-b-table, echo = FALSE}
bind_rows(lapply(outcomes6, function(y) tibble(Variable = labels6[y],
  `Baseline ATT` = est_se(simple_main[[y]]$overall$estimate, simple_main[[y]]$overall$std.error),
  `Thresholded ATT` = est_se(simple_b[[y]]$estimate, simple_b[[y]]$std.error)))) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcc",
      caption = "Appendix Table B1 of the paper: Callaway-Sant'Anna overall ATTs on the accuracy-thresholded data next to the baseline estimates (standard errors in parentheses).") %>%
  tab_style()
```

## Appendix C1: de Chaisemartin and d'Haultfoeuille {#sec-appendix-c1}

**The first alternative estimator replaces the Callaway-Sant'Anna
aggregation with the de Chaisemartin-d'Haultfoeuille dynamic estimator.**
The authors run `did_multiplegt_dyn` with 13 effects and 2 placebos,
clustered by user; the table mirrors their Panel B (the average total
effect per switcher).

```{r appendix-c1, cache = TRUE, cache.extra = data_md5, eval = run_dcdh}
suppressPackageStartupMessages(library(polars))
dcdh <- lapply(setNames(outcomes6, outcomes6), function(y) {
  pdf(NULL)
  m <- DIDmultiplegtDYN::did_multiplegt_dyn(as.data.frame(dat), y, "userid", "weeks_join",
                                            "upgraded", effects = 13, placebo = 2,
                                            cluster = "userid", graph_off = TRUE)
  dev.off()
  m$results$ATE
})
```

```{r appendix-c1-table, echo = FALSE, eval = run_dcdh}
bind_rows(lapply(outcomes6, function(y) {
  r <- dcdh[[y]]
  tibble(Variable = labels6[y], ATT = fmt(r[1, "Estimate"]), `Std. error` = fmt(r[1, "SE"]),
         `95\\% CI` = paste0("[", fmt(r[1, "LB CI"]), ", ", fmt(r[1, "UB CI"]), "]"),
         `CS ATT` = fmt(simple_main[[y]]$overall$estimate))
})) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, align = "lcccc",
      caption = "Appendix Table C1 of the paper: de Chaisemartin-d'Haultfoeuille average effect per switcher (13 effects, 2 placebos, clustered by user), with the Callaway-Sant'Anna simple ATT for comparison.") %>%
  tab_style()
```

**The two heterogeneity-robust estimators agree on the engagement
outcomes, as the paper reports for the proprietary data.** The
weight-loss amount differs more, with both estimates inside each other's
wide intervals. Both estimators sit well below the pooled TWFE of
Table 3 for engagement.

## Appendix C2: synthetic DiD {#sec-appendix-c2}

**The second alternative is synthetic DiD, which the authors run in Stata
cohort by cohort.** `sdid_weights()` is the package's implementation; it
requires a block design, so it is applied per adoption cohort and the
cohort estimates are combined with weights proportional to the number of
treated cells, which is how the Stata command aggregates. Cohort 2 has a
single pre-period and is excluded (the regularization needs two); no
Stata output ships with the package, so these numbers stand without a
comparison, and standard errors are omitted because each placebo
replication refits the weights over 11,091 control units.

```{r appendix-c2, cache = TRUE, cache.extra = data_md5}
sdid_cohort <- function(y, gg) {
  dg <- dat %>% filter(g == 0 | g == gg)
  w <- sdid_weights(dg, id = "userid", time = "weeks_join", y = y, d = "upgraded")
  c(estimate = w$estimate, n1 = w$N1, post = 15 - gg + 1)
}
sdid_tab <- lapply(setNames(outcomes6, outcomes6), function(y) {
  cells <- sapply(3:7, function(gg) sdid_cohort(y, gg))
  wts <- cells["n1", ] * cells["post", ]
  list(by_cohort = cells["estimate", ], overall = sum(cells["estimate", ] * wts) / sum(wts))
})
```

```{r appendix-c2-table, echo = FALSE}
bind_rows(lapply(outcomes6, function(y) {
  s <- sdid_tab[[y]]
  tibble(Variable = labels6[y],
         !!!setNames(lapply(fmt(s$by_cohort), identity), paste0("g = ", 3:7)),
         `SDID overall` = fmt(s$overall), `CS ATT` = fmt(simple_main[[y]]$overall$estimate))
})) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lccccccc",
      caption = "Appendix Table C2 of the paper, in R: synthetic DiD estimates by adoption cohort (cohort 2 excluded: one pre-period) and their treated-cell-weighted combination, with the Callaway-Sant'Anna ATT for comparison.") %>%
  tab_style("scale_down")
```

**Synthetic DiD lands near Callaway-Sant'Anna for the exercise and
health outcomes and above it for food tracking and budget adherence.**
It uses no covariates and builds its counterfactual by reweighting
never-adopters to match each cohort's pre-period path, so it differs
most where the week-1 covariates matter for selection. Later cohorts,
which upgrade on rising trajectories, also show the largest cohort
estimates, echoing Figure 5.

## Appendix D: the TWFE coefficient tables and the week-6 exposure design {#sec-appendix-d}

**Online Tables D1 and D2 print the coefficients behind the left panels
of Figures 2 and 3.** They come straight from the same fits.

```{r appendix-d-table, echo = FALSE}
dtab <- lapply(outcomes6, function(y) {
  fr <- frame_tw(tw_main[[y]])
  setNames(est_se(fr$estimate, fr$std.error), fr$event_time)
})
et <- sort(unique(as.numeric(names(dtab[[1]]))))
tibble(`Event time` = ifelse(et == 13, "13+", as.character(et)),
       !!!setNames(lapply(dtab, function(v) v[as.character(et)]), labels6)) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcccccc",
      caption = "Online Tables D1-D2 of the paper: TWFE event-study coefficients for all six outcomes (standard errors in parentheses, clustered by user). Reference: never-adopters and event times at or below -4.") %>%
  tab_style("scale_down")
```

**Online Figures D1 and D2 repeat the exposure-matched design for week-6
adopters** (GAM matching, ratios 4 and 3, seed 123). Food and exercise
tracking are shown; the pattern matches the week-5 figures.

```{r appendix-d-fig, cache = TRUE, cache.extra = data_md5, echo = FALSE, fig.height = 6, fig.cap = "Online Figures D1-D2 of the paper: the exposure-matched TWFE design for week-6 adopters, food and exercise tracking. Left: low exposure; right: high exposure."}
med6 <- median(users$total_track_before[users$week_upgrade == 6])
f_d <- reformulate(c("initialweight", "heightinches", "age", "first_day", "ini_goalweight",
  "ini_goalbmi", "ini_bmi", "gender", "track_food_avg_w4", "foodcalories_avg_w4",
  paste0(c("track_food_avg", "foodcalories_avg", "track_exercise_avg", "within_track_avg",
           "exercisecalories_avg", "lower_initial_avg", "weight_loss_norm_avg",
           "app_weight_loss_search_avg", "app_premium_search_avg", "weight_loss_premium_search_avg"),
         "_w5")), response = "ever_upgrade")
match6 <- function(cut_low, ratio) {
  sub <- recent %>% group_by(userid) %>%
    filter(week_upgrade == 6 | ever_upgrade == 0) %>% slice(1) %>% ungroup() %>%
    filter(if (cut_low) total_track_before < med6 else total_track_before >= med6)
  set.seed(123)
  M <- matchit(f_d, data = sub, method = "nearest", distance = "gam",
               ratio = ratio, replacement = FALSE)
  match.data(M, data = sub)$userid
}
pan_low6 <- panel_of(match6(TRUE, 4)); pan_high6 <- panel_of(match6(FALSE, 3))
pd <- list()
for (y in c("track_food_avg", "track_exercise_avg")) {
  pd[[length(pd) + 1]] <- paper_es(frame_tw(tw_es(y, pan_low6)), paste0("Week 6, low exposure: ", labels6[y]))
  pd[[length(pd) + 1]] <- paper_es(frame_tw(tw_es(y, pan_high6)), paste0("Week 6, high exposure: ", labels6[y]))
}
wrap_plots(pd, ncol = 2)
```

# Extensions {#sec-extensions}

**Eight extensions, each answering a question the paper raises but does
not settle, all on the same sample and labelled as ours.**

## Where the TWFE-CS gap comes from {#sec-ext-bacon}

**The Goodman-Bacon decomposition writes the pooled TWFE as a weighted
average of all two-by-two comparisons, and the forbidden ones carry the
overstatement.** `bacon_decomp()` gives every comparison's weight and
estimate; `twfe_weights()` gives the de Chaisemartin-d'Haultfoeuille
cell weights and their negative share.

```{r ext-bacon, cache = TRUE, cache.extra = data_md5}
bacon_food <- bacon_decomp(dat, id = "userid", time = "weeks_join", y = "track_food_avg", d = "upgraded")
bacon_food$by_type
tw_w <- twfe_weights(dat, id = "userid", time = "weeks_join", d = "upgraded", y = "track_food_avg")
c(share_negative = tw_w$share_negative, sum_negative = tw_w$sum_negative)
```

```{r ext-bacon-fig, echo = FALSE, fig.height = 3.6, fig.cap = "Extension 1: Goodman-Bacon decomposition of the pooled TWFE for food tracking. Each point is one two-by-two comparison; the horizontal line is the TWFE coefficient, the dashed line the Callaway-Sant'Anna ATT."}
ggplot(bacon_food$decomposition, aes(x = weight, y = estimate, colour = type)) +
  geom_hline(yintercept = bacon_food$twfe, colour = "grey30") +
  geom_hline(yintercept = simple_main[["track_food_avg"]]$overall$estimate, linetype = "dashed", colour = "grey30") +
  geom_point(size = 2) +
  scale_colour_manual(values = c("#0072B2", "#D55E00", "#009E73", "#CC79A7"), name = NULL) +
  labs(x = "Weight", y = "2x2 DiD estimate") +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

```{r ext-bacon-numbers, echo = FALSE}
bt <- bacon_food$by_type
w_never <- bt$weight[grepl("never", bt$type)]
```

**The decomposition acquits the usual suspects and convicts the
aggregation.** Negative weights are absent
(`r fmt(100 * tw_w$share_negative, 1)` percent of cells), and the
comparisons that use already-treated units as controls carry about
`r fmt(100 * (1 - w_never), 1)` percent of the weight. Instead,
`r fmt(100 * w_never, 1)` percent of the TWFE coefficient is the static
treated-versus-never comparison, which averages the whole post-adoption
window without covariates. The Callaway-Sant'Anna ATT differs because it
adjusts for the week-1 covariates and weights the group-time cells by
cohort size over a defined window, not because forbidden comparisons
dominate here.

## Which cohorts carry the effect {#sec-ext-groups}

**Group and calendar aggregations, and a balanced event study, unpack
the composition of Figures 2 and 3.** Later cohorts (the ones with rising
pre-adoption engagement in Figure 5) contribute only short horizons; the
balanced aggregation holds the cohort mix fixed across event times.

```{r ext-groups, cache = TRUE, cache.extra = data_md5}
grp <- lapply(cs_main[eng4], aggregate_att, type = "group", seed = 678)
bal <- lapply(cs_main[c("track_food_avg", "track_exercise_avg")], aggregate_att,
              type = "dynamic", min_e = -3, balance_e = 8, seed = 678)
```

```{r ext-groups-table, echo = FALSE}
bind_rows(lapply(eng4, function(y) {
  b <- grp[[y]]$by
  tibble(Outcome = labels6[y], !!!setNames(lapply(seq_len(nrow(b)), function(i) est_se(b$estimate[i], b$std.error[i])),
                                           paste0("g = ", b$group)))
})) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcccccc",
      caption = "Extension 2: cohort-specific ATTs (average post-treatment effect per adoption cohort), standard errors in parentheses.") %>%
  tab_style("scale_down")
```

```{r ext-balance-fig, echo = FALSE, fig.height = 3.6, fig.cap = "Extension 2: unbalanced against balanced (eight post periods per cohort) dynamic aggregation. The unbalanced curve changes cohort composition as the horizon grows; the balanced curve holds it fixed."}
wrap_plots(lapply(c("track_food_avg", "track_exercise_avg"), function(y) {
  fr <- event_study_frame(Unbalanced = dyn_main[[y]], Balanced = bal[[y]])
  plot_event_study(fr) + labs(title = labels6[y], y = "Estimates") +
    theme_minimal(base_size = 9) + theme(plot.title = element_text(size = 9, face = "bold"), legend.position = "bottom")
}), ncol = 2, guides = "collect") & theme(legend.position = "bottom")
```

**The attenuation is not a composition artifact.** The balanced curve
attenuates like the unbalanced one over the common window, so the decay
in Figure 2 is within cohorts, not a changing mix of them.

## How much power the pre-trend evidence has {#sec-ext-power}

**Three pre-period estimates cannot rule out much, and Roth's (2022)
calculation says how much.** For each outcome, `pretrend_power()` finds
the linear violation that the pre-test would detect with 50 and 80
percent power and the bias such an undetected trend would leave in the
week-0 effect.

```{r ext-power, cache = TRUE, cache.extra = data_md5}
pow <- lapply(dyn_main[c("track_food_avg", "track_exercise_avg")], function(d) {
  p0 <- pretrend_power(d, seed = 1)
  s50 <- p0$detectable$slope[p0$detectable$target_power == 0.5]
  p1 <- pretrend_power(d, slope = s50, seed = 1)
  list(detectable = p0$detectable, bias0 = p1$slopes[[1]]$bias)
})
```

```{r ext-power-table, echo = FALSE}
bind_rows(lapply(c("track_food_avg", "track_exercise_avg"), function(y) {
  p <- pow[[y]]
  b0 <- p$bias0[p$bias0$event_time == 0, ]
  tibble(Outcome = labels6[y],
         `Slope detected, 50\\% power` = fmt(p$detectable$slope[1], 4),
         `Slope detected, 80\\% power` = fmt(p$detectable$slope[2], 4),
         `Week-0 effect` = fmt(e0[y]),
         `Bias at 50\\%-power slope` = fmt(b0$unconditional_bias, 4),
         `Conditional bias (test passed)` = fmt(b0$conditional_bias, 4))
})) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      caption = "Extension 3: power of the pre-trend evidence (Roth 2022). The detectable slope is the linear violation per week the three pre-period coefficients would flag with the stated power; the last columns give the week-0 bias such a trend would leave, unconditionally and conditional on the pre-test passing.") %>%
  tab_style("scale_down")
```

**A violation detected only half the time would move the week-0 effect
by a small fraction of its size.** That is why Figure 7's conclusions
survive moderate $\bar M$: the effects are an order of magnitude larger
than the trends the pre-periods could hide.

## Breakdown values by outcome and horizon {#sec-ext-honest}

**Figure 7 covers event time 0; the honest question extends across
horizons and to the smoothness family.** For each engagement outcome the
table reports the relative-magnitudes breakdown value (the smallest
$\bar M$ at which zero enters the interval) at event time 0 and for the
average of event times 0 to 3, and the smoothness breakdown $M$ at event
time 0.

```{r ext-honest, cache = TRUE, cache.extra = data_md5, eval = run_honest}
breakdown_rm <- function(dyn, l_vec) {
  hi <- honestdid_inputs(dyn)
  rob <- HonestDiD::createSensitivityResults_relativeMagnitudes(betahat = hi$betahat, sigma = hi$sigma,
    numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods,
    Mbarvec = seq(0.2, 3, by = 0.1), l_vec = l_vec, gridPoints = 100)
  cov0 <- rob$lb <= 0 & rob$ub >= 0
  if (any(cov0)) min(rob$Mbar[cov0]) else Inf
}
breakdown_sm <- function(dyn, l_vec, m_max) {
  hi <- honestdid_inputs(dyn)
  rob <- HonestDiD::createSensitivityResults(betahat = hi$betahat, sigma = hi$sigma,
    numPrePeriods = hi$numPrePeriods, numPostPeriods = hi$numPostPeriods,
    Mvec = seq(0, m_max, length.out = 11), l_vec = l_vec)
  cov0 <- rob$lb <= 0 & rob$ub >= 0
  if (any(cov0)) min(rob$M[cov0]) else Inf
}
honest_ext <- lapply(setNames(eng4, eng4), function(y) {
  d <- dyn_uni[[y]]
  np <- honestdid_inputs(d)$numPostPeriods
  l0 <- HonestDiD::basisVector(1, np)
  lavg <- rep(1 / np, np)
  m_max <- if (y == "exercisecalories_avg") 25 else if (y == "track_exercise_avg") 0.15 else 0.06
  c(rm0 = breakdown_rm(d, l0), rm_avg = breakdown_rm(d, lavg), sm0 = breakdown_sm(d, l0, m_max))
})
```

```{r ext-honest-table, echo = FALSE, eval = run_honest}
bind_rows(lapply(eng4, function(y) tibble(Outcome = labels6[y],
  `RM breakdown, e = 0` = fmt(honest_ext[[y]]["rm0"], 1),
  `RM breakdown, avg e = 0-3` = fmt(honest_ext[[y]]["rm_avg"], 1),
  `Smoothness breakdown, e = 0` = fmt(honest_ext[[y]]["sm0"], 3)))) %>%
  kbl(booktabs = TRUE, linesep = "",
      caption = "Extension 4: breakdown values. RM columns: the smallest Mbar (violations up to Mbar times the largest pre-period deviation) at which zero enters the confidence set. Smoothness column: the smallest per-week slope change M with the same property.") %>%
  tab_style()
```

**Averaging over the first month is less robust than the impact effect,
because later horizons let violations accumulate.** A smoothness
breakdown of zero for exercise calories says a continued linear
pre-trend alone removes its significance, matching its weaker
relative-magnitudes robustness. This sharpens the paper's reading: the
credible finding is the short-lived lift, not its persistence.

## The matched sample with the main estimator {#sec-ext-matchcs}

**Figure 8 switches estimators along with samples; holding the estimator
fixed isolates what the matching does.** `att_gt()` runs on the matched
panel of Section \@ref(sec-figure8), and the frame overlays it on the
matched-sample TWFE.

```{r ext-matchcs, cache = TRUE, cache.extra = data_md5}
cs_m <- lapply(setNames(c("track_food_avg", "track_exercise_avg"), c("track_food_avg", "track_exercise_avg")),
  function(y) {
    a <- att_gt(dat8, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
                method = "dr", control_group = "notyet", n_boot = 499, seed = 678)
    aggregate_att(a, type = "dynamic", min_e = -3, na.rm = TRUE, seed = 678)
  })
```

```{r ext-matchcs-fig, echo = FALSE, fig.height = 3.8, fig.cap = "Extension 5: the matched sample of Figure 8 estimated with Callaway-Sant'Anna next to the matched-sample TWFE."}
wrap_plots(lapply(c("track_food_avg", "track_exercise_avg"), function(y) {
  fr <- event_study_frame(`Matched TWFE` = tw8[[y]], `Matched CS` = cs_m[[y]])
  fr <- fr[fr$event_time <= 9, ]
  plot_event_study(fr) + labs(title = labels6[y], y = "Estimates") +
    theme_minimal(base_size = 9) + theme(plot.title = element_text(size = 9, face = "bold"))
}), ncol = 2, guides = "collect") & theme(legend.position = "bottom")
```

**The two agree on the matched sample, so the robustness claim does not
depend on switching estimators.** The lift and the attenuation are in
the design, not the method.

## Anticipation {#sec-ext-anticipation}

**Figure 5's rising pre-adoption activity could be anticipation rather
than selection.** `att_gt(anticipation = 1)` treats the week before
adoption as already affected and moves the base period one week back.

```{r ext-anticipation, cache = TRUE, cache.extra = data_md5}
ant <- lapply(setNames(c("track_food_avg", "track_exercise_avg"), c("track_food_avg", "track_exercise_avg")),
  function(y) {
    a <- att_gt(dat, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
                method = "dr", control_group = "notyet", anticipation = 1, n_boot = 499, seed = 678)
    aggregate_att(a, type = "dynamic", min_e = -3, na.rm = TRUE, seed = 678)$by
  })
```

```{r ext-anticipation-table, echo = FALSE}
bind_rows(lapply(c("track_food_avg", "track_exercise_avg"), function(y) {
  b0 <- dyn_main[[y]]$by; b1 <- ant[[y]]
  bind_rows(lapply(c(-1, 0, 1, 2), function(e) tibble(Outcome = labels6[y], `Event time` = e,
    `Baseline` = est_se(b0$estimate[b0$event_time == e], b0$std.error[b0$event_time == e]),
    `Anticipation = 1` = if (any(b1$event_time == e)) est_se(b1$estimate[b1$event_time == e], b1$std.error[b1$event_time == e]) else "")))
})) %>%
  kbl(booktabs = TRUE, linesep = "", align = "lccc",
      caption = "Extension 6: allowing one week of anticipation. The base period moves from g-1 to g-2; a large change in the week-0 effect would signal that the g-1 outcome is already treated.") %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "top") %>%
  tab_style()
```

**Exercise tracking is stable; the food-tracking impact effect grows
when the decision week counts as treated.** With one week of
anticipation the food effect at adoption rises from
`r fmt(dyn_main[["track_food_avg"]]$by$estimate[dyn_main[["track_food_avg"]]$by$event_time == 0])`
to `r fmt(ant[["track_food_avg"]]$estimate[ant[["track_food_avg"]]$event_time == 0])`,
which says part of the food response in this draw begins in the week the
upgrade happens rather than strictly after it. The baseline week
$-1$ coefficient is small and insignificant, so the lift-then-decay
reading survives either dating; the exercise is a reminder that the
anticipation window is a modelling choice worth reporting.

## Exposure as a gradient {#sec-ext-tercile}

**The median split of Table 5 is the coarsest version of the hedonic
decline test; terciles sharpen it.** Adopters are split into thirds by
pre-adoption tracking days, and the week-0 and week-1 effects are
estimated per tercile.

```{r ext-tercile, cache = TRUE, cache.extra = data_md5}
adopters <- users %>% filter(week_upgrade >= 2, week_upgrade <= 7)
terc <- quantile(adopters$total_track_before, c(1/3, 2/3))
terc_of <- function(v) cut(v, c(-Inf, terc, Inf), labels = c("Low", "Middle", "High"))
terc_run <- function(lev) {
  ids <- adopters$userid[terc_of(adopters$total_track_before) == lev]
  dd <- dat %>% filter(userid %in% c(ids, never_ids))
  lapply(setNames(c("track_food_avg", "track_exercise_avg"), c("track_food_avg", "track_exercise_avg")),
    function(y) {
      a <- att_gt(dd, id = "userid", time = "weeks_join", group = "g", y = y, x = xv,
                  method = "dr", control_group = "notyet", n_boot = 499, seed = 1234)
      aggregate_att(a, type = "dynamic", min_e = -3, na.rm = TRUE, seed = 1234)$by
    })
}
terc_res <- lapply(setNames(c("Low", "Middle", "High"), c("Low", "Middle", "High")), terc_run)
```

```{r ext-tercile-fig, echo = FALSE, fig.height = 3.6, fig.cap = "Extension 7: week-0 and week-1 effects by tercile of pre-adoption exposure, with 95 percent intervals. Hedonic decline predicts the monotone decline in the lift; sunk costs predict a flat profile."}
terc_df <- bind_rows(lapply(names(terc_res), function(lev) bind_rows(lapply(c("track_food_avg", "track_exercise_avg"), function(y) {
  b <- terc_res[[lev]][[y]]
  bind_rows(lapply(c(0, 1), function(e) tibble(Tercile = lev, Outcome = labels6[y], event = paste("Week", e),
    estimate = b$estimate[b$event_time == e], se = b$std.error[b$event_time == e])))
}))))
terc_df$Tercile <- factor(terc_df$Tercile, levels = c("Low", "Middle", "High"))
ggplot(terc_df, aes(x = Tercile, y = estimate, colour = event)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
                  position = position_dodge(width = 0.4)) +
  facet_wrap(~ Outcome) +
  scale_colour_manual(values = c("#0072B2", "#D55E00"), name = NULL) +
  labs(x = "Pre-adoption exposure tercile", y = "Effect") +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
```

**The high-exposure tercile sits well below the other two; low and
middle are indistinguishable.** The profile is a step down at high
exposure rather than a straight line, which still contradicts the flat
profile sunk costs predict: users who had already exhausted the shared
features respond least. The novelty margin appears to bind only once
exposure is high.

## An estimator horse race {#sec-ext-race}

**The lecture's comparison of heterogeneity-robust estimators, run on
this application.** Callaway-Sant'Anna (covariates, not-yet-treated),
Sun-Abraham (interaction-weighted, `fixest::sunab()`), the imputation
estimator (`did_imputation()`), and the binned TWFE, all for food
tracking.

```{r ext-race, cache = TRUE, cache.extra = data_md5}
dat$g_sa <- ifelse(dat$g == 0, 10000L, dat$g)
sa <- feols(track_food_avg ~ sunab(g_sa, weeks_join) | userid + weeks_join, data = dat, cluster = ~userid)
imp <- did_imputation(dat, id = "userid", time = "weeks_join", group = "g", y = "track_food_avg",
                      pre_window = 3)
race <- event_study_frame(`Callaway-Sant'Anna` = dyn_main[["track_food_avg"]],
                          `Sun-Abraham` = sa, Imputation = imp,
                          `TWFE (binned)` = tw_main[["track_food_avg"]])
```

```{r ext-race-fig, echo = FALSE, fig.height = 4.2, fig.cap = "Extension 8: four estimators of the food-tracking event study. The binned TWFE sits far above the three heterogeneity-robust estimators at every horizon."}
plot_event_study(race %>% filter(event_time >= -3, event_time <= 13)) +
  labs(y = "Effect on food tracking") + theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

```{r ext-race-table, echo = FALSE}
race %>% filter(event_time %in% c(0, 1, 5, 10)) %>%
  mutate(cell = est_se(estimate, std.error)) %>%
  select(method, event_time, cell) %>%
  pivot_wider(names_from = event_time, values_from = cell, names_prefix = "e = ") %>%
  kbl(booktabs = TRUE, linesep = "", align = "lcccc", col.names = c("Estimator", "e = 0", "e = 1", "e = 5", "e = 10"),
      caption = "Extension 8: the food-tracking event study at four horizons across estimators (standard errors in parentheses).") %>%
  tab_style()
```

**Callaway-Sant'Anna and Sun-Abraham agree to the third decimal, the
imputation estimator sits slightly above them, and the binned TWFE is
far above all three at every horizon.** The TWFE gap is the reference
group: its coefficients compare against never-adopters and distant
pre-periods jointly, so cohort-level differences load onto the event
dummies. The imputation estimator fits the untreated model on all
pre-periods, which buys precision at the price of pre-period parallel
trends; its closeness to Callaway-Sant'Anna says that assumption is not
doing damage here. Synthetic DiD, the remaining estimator, is in Section
\@ref(sec-appendix-c2).

# What the replication says {#sec-reading}

**The workflow reproduces end to end on the package.** On the simulated
data our `att_gt()` matches the `did` package to machine precision, the
TWFE helper matches the authors' dummy regressions exactly, and every
table and figure of the paper and its appendix is rebuilt in the
published layout with `att_gt()`, `aggregate_att()`,
`event_study_frame()`, `plot_event_study()`, and `honestdid_inputs()`,
plus MatchIt as the matching engine and HonestDiD, DIDmultiplegtDYN, and
`sdid_weights()` for the sensitivity and alternative estimators.

**The paper's lessons survive the added cuts, with two honest
qualifications from the simulated draw.** The TWFE-versus-CS gap is the
static, covariate-free treated-versus-never comparison, not negative
weights or forbidden controls; the attenuation is within cohorts, not
composition, though this draw damps the food-tracking decay; the
exposure effect is a step down at high exposure rather than a monotone
gradient; the matched sample gives the same answer under the main
estimator; one week of anticipation leaves exercise unchanged and grows
the food impact effect, a dating choice worth reporting; and Sun-Abraham
and the imputation estimator agree with Callaway-Sant'Anna where their
assumptions overlap.

**What to remember for one's own staggered designs.** Drop cohorts
without pre-periods; state the comparison group; put covariates in both
nuisances; aggregate deliberately (dynamic, group, balanced); report
simultaneous bands; quantify how much the pre-periods can actually rule
out before calling trends parallel; and treat TWFE as a diagnostic, not
an estimate.

# References

- Baker, A., Callaway, B., Cunningham, S., Goodman-Bacon, A., and Sant'Anna, P. H. C. 2026. "Difference-in-Differences Designs: A Practitioner's Guide." Working paper.
- Callaway, B., and Sant'Anna, P. H. C. 2021. "Difference-in-Differences with Multiple Time Periods." *Journal of Econometrics*.
- de Chaisemartin, C., and d'Haultfoeuille, X. 2020. "Two-Way Fixed Effects Estimators with Heterogeneous Treatment Effects." *American Economic Review*.
- Goodman-Bacon, A. 2021. "Difference-in-Differences with Variation in Treatment Timing." *Journal of Econometrics*.
- Jiang, Y., Uetake, K., and Yang, N. 2026. "Does Premium Version Adoption in mHealth Improve User Engagement and Health-Related Outcomes?" *Marketing Science*, Articles in Advance. Replication package: INFORMS.
- Rambachan, A., and Roth, J. 2023. "A More Credible Approach to Parallel Trends." *Review of Economic Studies*.
- Roth, J. 2022. "Pretest with Caution: Event-Study Estimates after Testing for Parallel Trends." *AER: Insights*.
- Sun, L., and Abraham, S. 2021. "Estimating Dynamic Treatment Effects in Event Studies with Heterogeneous Treatment Effects." *Journal of Econometrics*.
