---
title: "Causal Inference in Economics and Marketing"
subtitle: "Section 05: Instrumental Variables and Unobserved Confounding"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, fig.align = "center", fig.pos = "H",
  out.width = "95%", collapse = TRUE, comment = "#>"
)
# Code and output in footnotesize.
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
# This lecture lives in inst/lectures/05_iv/ of the package source. When
# knitted from the source tree it loads the development version of the
# package; when the package is installed it uses the installed helpers.
helper_rel <- file.path("vignette_helpers", "05_iv", "iv_lecture_helpers.R")
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
  helper_path <- file.path(pkg_root, "inst", helper_rel)
} else {
  library(causalmetrics)
  helper_path <- system.file(helper_rel, package = "causalmetrics")
}
library(estimatr)
library(fixest)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(mlr3)
library(mlr3learners)
source(helper_path)
helper_md5 <- tools::md5sum(helper_path)
sim_md5 <- if (file.exists(file.path(pkg_root, "R", "sim-iv.R"))) tools::md5sum(file.path(pkg_root, "R", "sim-iv.R")) else "installed"
n_sims_pliv <- 100
n_sims_late <- 100
n_sims_weak <- 400
n_sims_cf <- 200
fmt <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
run_mlogit <- requireNamespace("mlogit", quietly = TRUE) && requireNamespace("dfidx", quietly = TRUE)
x5 <- paste0("x", 1:5)
```

# When conditioning fails {#sec-problem}

## Two graphs we cannot tell apart

Sections 03 and 04 assumed that, given observed controls, treatment is as
good as random. Every estimator there, from regression to double machine
learning, was a way to condition well. None of them helps when a variable
that drives both the treatment and the outcome is not observed.

The classic example is schooling and earnings. Ability raises schooling and
raises earnings. Suppose we observe schooling $D$, earnings $Y$, and
background $X$, but not ability $A$. Two structural models are consistent
with any joint distribution of $(Y, D, X)$: one in which $D$ causes $Y$, and
one in which $A$ causes both and $D$ has no effect. No amount of data
separates them (Chernozhukov et al. 2026, ch. 12). Conditioning on $X$ is
not enough because $A$ is not in $X$.

The same problem has other names: demand estimation (a demand shock moves
both price and quantity), program evaluation with self-selection (those who
expect to gain enroll), and any marketing setting where the firm targets
the customers it expects to respond.

## What can be said without an instrument {#sec-sensitivity}

In the partially linear model
$$
Y = \theta D + \delta A + f_Y(X) + \epsilon_Y, \qquad D = \gamma A + f_D(X) + \epsilon_D,
$$
partialling out $X$ and regressing $\tilde Y$ on $\tilde D$ gives
$\theta + \phi$ with $\phi = \delta\gamma / E[\tilde D^2]$. The bias is
bounded in terms of two partial $R^2$: how much of the residual variation in
$\tilde D$ the confounder explains ($R^2_{D \sim A}$), and how much of the
residual variation in $\tilde Y$ it explains after $\tilde D$
($R^2_{Y \sim A \mid D}$):
$$
\phi^2 = \frac{R^2_{Y\sim A\mid D}\, R^2_{D\sim A}}{1 - R^2_{D\sim A}}
        \cdot \frac{E[(\tilde Y - \beta \tilde D)^2]}{E[\tilde D^2]}
$$
(Chernozhukov et al. 2026, Theorem 12.2.1; Cinelli and Hazlett 2020;
appendix \@ref(app-ovb)). The second factor is estimated from the DML
residuals. The first is the analyst's belief about the confounder, often
benchmarked against the observed controls.

`dml_sensitivity()` applies the bound to a `cm_dml` fit. The running
"linear" design of this lecture has an unobserved $A$ with $\delta = 1$ and
$\gamma = 0.8$, so the partialling-out estimate is biased upward.

```{r sensitivity}
dat <- sim_iv(n = 2000, dgp = "linear", seed = 1)
rf <- iv_forest()
fit_plr <- est_dml(dat, "y", "d", x5, learner_l = rf, learner_m = rf, seed = 1)
sens <- dml_sensitivity(fit_plr, r2_y = 0.10, r2_d = 0.10)
sens
```

The estimate is `r fmt(fit_plr$estimate)` for a true effect of 1. A
confounder that explains 10 percent of the residual variation in both
variables would move it by `r fmt(sens$bias)`, so the interval of estimates
consistent with that scenario is
[`r fmt(sens$bounds[1])`, `r fmt(sens$bounds[2])`]. The robustness value
says a confounder would need partial $R^2$ of about
`r fmt(sens$robustness_value[1])` on both sides to bring the estimate to
zero. In this design the true confounder is that strong. Figure
\@ref(fig:sensitivity-plot) shows the contours.

```{r sensitivity-plot, echo = FALSE, fig.height = 3.8, fig.cap = "Contours of the omitted-confounder bias bound as a function of the two partial R-squareds. The red contour marks confounders strong enough to bring the estimate to zero."}
plot_dml_sensitivity(sens, benchmarks = data.frame(r2_d = 0.10, r2_y = 0.10, label = "benchmark"))
```

Sensitivity analysis tells us how wrong we could be. It does not tell us the
answer. For that we need a second source of variation.

## Instruments and Wright's path diagram {#sec-wright}

An instrument $Z$ is a variable that moves the treatment but affects the
outcome only through the treatment. Philip Wright (1928) drew the first
causal path diagram to make the point: if $Z \to D \to Y$ and $Z$ has no
other path to $Y$, then the effect of $Z$ on $Y$ equals the effect of $Z$ on
$D$ times the effect of $D$ on $Y$. Two regressions we can run give the
product and one factor; the ratio is the effect we want.

Examples that appear in this lecture: an offer to participate (a nudge) as
an instrument for participation; a supply shifter (weather where the
product is grown) for price; distance to college for schooling; random
assignment of judges for incarceration; national industry shocks
apportioned by local industry shares for local exposure to trade.

The price is a new assumption. Conditional exogeneity said "no unobserved
confounder of $D$". The exclusion restriction says "no unobserved path from
$Z$ to $Y$ except through $D$". Whether that is more or less credible
depends entirely on where $Z$ comes from. When $Z$ is randomized, it is
credible by design.

## Roadmap

Section \@ref(sec-linear) sets up instrumental variables in the linear
model and the reporting conventions. Section \@ref(sec-late) asks what IV
estimates when effects are heterogeneous. Section \@ref(sec-cf) develops
the control-function approach, which equals two-stage least squares in the
linear model and extends to nonlinear ones. Section \@ref(sec-orthogonal)
gives the orthogonal scores that let machine learning handle the controls.
Section \@ref(sec-weak) treats weak instruments. Section \@ref(sec-many)
covers several instruments, judges, marginal treatment effects, and
constructed instruments. Section \@ref(sec-platforms) turns to
experiments and platforms, Section \@ref(sec-traps) to traps, Section
\@ref(sec-practice) to the replications. The appendix holds the proofs.

# Instrumental variables in the linear model {#sec-linear}

## Relevance, exclusion, and the Wald ratio

Take the structural equation $Y = \alpha + \theta D + e$ with
$\mathrm{Cov}(D, e) \neq 0$. An instrument $Z$ satisfies

- **relevance**: $\mathrm{Cov}(Z, D) \neq 0$, and
- **exogeneity with exclusion**: $\mathrm{Cov}(Z, e) = 0$.

Then $\mathrm{Cov}(Z, Y) = \theta\,\mathrm{Cov}(Z, D)$, so
$$
\theta = \frac{\mathrm{Cov}(Z, Y)}{\mathrm{Cov}(Z, D)}
       = \frac{E[Y \mid Z = 1] - E[Y \mid Z = 0]}{E[D \mid Z = 1] - E[D \mid Z = 0]}
$$
for a binary instrument: the Wald ratio, reduced form over first stage. In
an encouragement design the numerator is the intention-to-treat effect and
the denominator is the compliance rate; the ratio rescales the ITT to the
effect of the treatment itself.

## Two-stage least squares {#sec-2sls}

With controls $X$ and one or more instruments, two-stage least squares
(2SLS) regresses $D$ on $(Z, X)$, forms $\hat D$, and regresses $Y$ on
$(\hat D, X)$. Equivalently, it solves the moment condition
$E[(Y - \alpha - \theta D - X'\beta)\,(Z, X)] = 0$. Three facts to keep in
mind.

- The second-stage OLS standard errors are wrong, because they use
  residuals computed with $\hat D$ rather than $D$. Software reports the
  correct ones; hand-rolled two stages do not.
- With more instruments than endogenous regressors the model is
  overidentified and 2SLS combines the instruments with weights from the
  first stage (Section \@ref(sec-many-weights)).
- "Endogenous" is a statement about a structural error, not about
  correlation in the data. It is meaningful only once the structural
  equation is written down.

The running design has one instrument, five covariates, and a true effect of
1. The instrument is excluded conditional on $h(x) = 0.5 x_1^2 + 0.5\sin(2x_2)$,
which also enters the outcome; in this section we include the two terms as
controls, as an analyst who knew the functional form would. Section
\@ref(sec-orthogonal) drops that knowledge. Every regression an applied
paper reports appears in one table.

```{r tsls}
dat$x1sq <- dat$x1^2
dat$s2x2 <- sin(2 * dat$x2)
xc <- c("x1sq", "s2x2", "x3", "x4", "x5")
ols <- lm_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5, data = dat)
first <- lm_robust(d ~ z + x1sq + s2x2 + x3 + x4 + x5, data = dat)
reduced <- lm_robust(y ~ z + x1sq + s2x2 + x3 + x4 + x5, data = dat)
tsls <- iv_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5 | z + x1sq + s2x2 + x3 + x4 + x5, data = dat)
```

```{r tab-tsls, echo = FALSE}
tibble(
  Regression = c("OLS: y on d, X", "First stage: d on z, X", "Reduced form: y on z, X", "2SLS: y on d, X (z instrument)"),
  Coefficient = c("d", "z", "z", "d"),
  Estimate = c(coef(ols)[["d"]], coef(first)[["z"]], coef(reduced)[["z"]], coef(tsls)[["d"]]),
  SE = c(ols$std.error[["d"]], first$std.error[["z"]], reduced$std.error[["z"]], tsls$std.error[["d"]])
) %>%
  mutate(across(c(Estimate, SE), ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "The four regressions to report. True effect 1. The Wald ratio of the two middle rows equals the 2SLS row.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

OLS is biased upward by the confounder, as Section \@ref(sec-sensitivity)
anticipated. The 2SLS estimate is close to 1 with a standard error about
`r fmt(tsls$std.error[["d"]] / ols$std.error[["d"]], 1)` times the OLS one:
the price of using only the variation the instrument induces.

## Standard errors {#sec-se}

Report heteroskedasticity-robust standard errors by default and cluster at
the level at which the instrument varies (a policy that varies by state,
weather that varies by market-day, a judge assigned to a court-year cell).
`iv_robust()` and `feols()` do both. With `fixest`:

```{r fixest}
fe <- feols(y ~ x1sq + s2x2 + x3 + x4 + x5 | d ~ z, data = dat, vcov = "hetero")
coeftable(fe)["fit_d", ]
fitstat(fe, type = c("ivf", "ivwald"))
```

## Overidentification and endogeneity tests {#sec-tests}

With more instruments than endogenous regressors, the Sargan-Hansen test
asks whether the instruments give compatible estimates. A rejection means
that at least one instrument violates exclusion, *or* that the instruments
identify different effects for different people (Section
\@ref(sec-late)). A non-rejection with instruments that share a bias is
uninformative. Treat it as a test of the homogeneous-effect functional form
rather than as a license.

The Wu-Hausman test compares OLS and 2SLS. It tests whether endogeneity
matters, not whether the instrument is valid, and has low power when the
instrument is weak (2SLS then drifts toward OLS). Section \@ref(sec-cf)
shows it is the t test on the control function.

## Reporting {#sec-reporting}

A first stage, a reduced form, and the IV estimate, always together, with
the same controls and the same sample. The reader needs the first stage to
judge relevance and the reduced form to see the raw effect of the
instrument; a paper whose abstract does not say what the instrument is and
why it is excluded has usually not thought hard about either.

Wayne's earlier note on Oreopoulos (2006) is the template: a change in the
minimum school-leaving age in Britain (1947) and Northern Ireland (1957)
instruments years of schooling in a regression of log earnings, with a
quartic in birth cohort as controls. Its Table 1 reports the first stage
(the jump in years of schooling at the reform cohort), the reduced form
(the jump in earnings), and the IV ratio side by side, for each country and
pooled with a country fixed effect. That table structure is what to imitate.

# Heterogeneous effects: what IV estimates {#sec-late}

## Potential outcomes indexed by the instrument

Let $D(z)$ be the treatment a unit would take under instrument value $z$
and $Y(d)$ the outcome under treatment $d$. With a binary instrument and a
binary treatment, each unit is one of four types (Imbens and Angrist 1994):

| $D(0)$ | $D(1)$ | Type |
|---|---|---|
| 0 | 1 | complier |
| 1 | 1 | always-taker |
| 0 | 0 | never-taker |
| 1 | 0 | defier |

The LATE assumptions are: (i) independence, $Z \perp (Y(0), Y(1), D(0), D(1))$
(conditionally on $X$ if needed); (ii) exclusion, $Y$ depends on $Z$ only
through $D$; (iii) relevance, $P(D(1) = 1) \neq P(D(0) = 1)$; (iv)
monotonicity, no defiers. Monotonicity is natural when the instrument is an
offer of something people want, and much less natural when the instrument is
an assigned decision maker, since two judges need not order every defendant
the same way (Section \@ref(sec-judges)).

## The LATE theorem {#sec-late-theorem}

Under (i)-(iv),
$$
\frac{E[Y \mid Z = 1] - E[Y \mid Z = 0]}{E[D \mid Z = 1] - E[D \mid Z = 0]}
= E\big[Y(1) - Y(0) \mid D(1) > D(0)\big],
$$
the average effect among compliers. *Skeleton.* By independence and
exclusion the numerator is $E[Y(D(1)) - Y(D(0))] = E[(Y(1) - Y(0))(D(1) - D(0))]$.
The difference $D(1) - D(0)$ is $1$ for compliers, $0$ for always- and
never-takers, and $-1$ for defiers; without defiers the numerator is the
complier effect times the complier share. The denominator is the complier
share. Divide (appendix \@ref(app-late)).

Three consequences.

- The IV estimand is not the ATE. It is the effect for the units the
  instrument moves. Different instruments move different people and
  estimate different LATEs; overidentification tests can reject for that
  reason alone.
- The complier share is estimable (the first stage), and so are the
  compliers' characteristics (Section \@ref(sec-compliers)), even though no
  individual complier can be named.
- With a continuous or multi-valued instrument, IV is a weighted average of
  effects for the units moved at each margin (Sections \@ref(sec-many-weights)
  and \@ref(sec-mte)).

## IV as a compliance-weighted average {#sec-weighted}

With heterogeneous effects $\beta_i$ and individual compliance
$\pi_i = P(D_i = 1 \mid Z_i = 1) - P(D_i = 1 \mid Z_i = 0)$, the Wald ratio
converges to $E[\pi_i \beta_i] / E[\pi_i]$: units the instrument moves more
get more weight. It equals the ATE if effects are homogeneous or if
$\beta_i$ is uncorrelated with $\pi_i$. When treatment is only available
through the instrument ($\pi_i = p_{1i}$), the units most likely to take it
up when offered carry the weight; if those are the units with the largest
gains, IV exceeds the ATE. When treatment is compulsory under $Z = 1$ and
voluntary otherwise ($\pi_i = 1 - p_{0i}$), the sign reverses.

The running "late" design makes this concrete: compliers have effect
$2 + x_1$, always-takers $3 + x_1$, never-takers $0.5 + x_1$. The population
ATE and the LATE differ.

```{r late-design}
dl <- sim_iv(n = 4000, dgp = "late", seed = 2)
c(late = attr(dl, "late"), ate = attr(dl, "ate"), complier_share = attr(dl, "complier_share"))
wald <- iv_robust(y ~ d | z, data = dl)
c(wald = coef(wald)[["d"]], se = wald$std.error[["d"]])
```

The Wald estimate is the LATE, not the ATE, and nothing in the data would
reveal the always-takers' larger effect.

## Who the compliers are {#sec-compliers}

Abadie (2003) shows that the weights
$$
\kappa_i = 1 - \frac{D_i (1 - Z_i)}{1 - \pi(X_i)} - \frac{(1 - D_i) Z_i}{\pi(X_i)},
\qquad \pi(X) = P(Z = 1 \mid X),
$$
average to the complier share and that $E[\kappa\, g(X)] / E[\kappa]$ is the
complier mean of any function of $X$. The intuition: units with $D = 1,
Z = 0$ are always-takers and units with $D = 0, Z = 1$ are never-takers;
subtracting them (with inverse-propensity weights that undo the assignment
of $Z$) leaves compliers. The same logic gives the always-taker and
never-taker means. `complier_profile()` reports the table.

```{r compliers}
cp <- complier_profile(dl, d = "d", z = "z", covariates = c("x1", "x2"), x = "x1", n_boot = 199, seed = 1)
cp
```

```{r compliers-truth, echo = FALSE}
truth_cp <- c(mean(dl$x1[dl$type == "complier"]), mean(dl$x2[dl$type == "complier"]))
```

The complier means of $x_1$ and $x_2$ are `r fmt(cp$table$compliers[1])`
and `r fmt(cp$table$compliers[2])`; the true values in the simulation are
`r fmt(truth_cp[1])` and `r fmt(truth_cp[2])`. This is the table to report
whenever a LATE is the headline: it tells the reader for whom the effect was
estimated, which is the first step in arguing whether it extrapolates.

## Conditional LATE with covariates {#sec-conditional-late}

When independence holds only conditionally on $X$, the conditional Wald
ratio identifies the conditional LATE at each $x$, and the unconditional
LATE is the ratio of averages (Frölich 2007; Chernozhukov et al. 2026,
Theorem 12.4.1):
$$
\theta = \frac{E\{E[Y \mid Z = 1, X] - E[Y \mid Z = 0, X]\}}{E\{E[D \mid Z = 1, X] - E[D \mid Z = 0, X]\}},
$$
the ATE of $Z$ on $Y$ over the ATE of $Z$ on $D$. Two-stage least squares
with linear controls is *not* this estimand: it is the linear projection
version, which coincides only when both the reduced form and the first stage
are linear and additively separable in $X$ and the compliance does not vary
with $X$. Section \@ref(sec-late-score) estimates the ratio of averages with
doubly robust scores for each ATE and shows the difference in a Monte Carlo.

## Testing the assumptions and bounding the ATE {#sec-tests-bounds}

Independence and monotonicity together imply inequalities on the observed
distribution (Kitagawa 2015; Mourifié and Wan 2017): for any set $A$,
$P(Y \in A, D = 1 \mid Z = 1) \ge P(Y \in A, D = 1 \mid Z = 0)$ and
$P(Y \in A, D = 0 \mid Z = 0) \ge P(Y \in A, D = 0 \mid Z = 1)$. A
violation refutes the model; the absence of a violation does not confirm
it. Report the test when the instrument is not randomized.

The LATE assumptions point-identify only the complier effect. For a binary
outcome, independence alone bounds the ATE for everyone: the instrument
gives two sets of no-assumption bounds on $E[Y(1)]$ and $E[Y(0)]$, one per
instrument value, and their intersection is valid (Manski 1990; Balke and
Pearl 1997 give the sharp version). `iv_ate_bounds()` computes the
intersection bounds.

```{r bounds}
dl$y_binary <- as.integer(dl$y > 1)
iv_ate_bounds(dl, "y_binary", "d", "z", n_boot = 199, seed = 1)
```

The bounds without the instrument have width one; the instrument narrows
them without any monotonicity assumption. When the policy question is
"treat everyone", this is the honest answer, and the LATE is a point
inside it.

# Control functions {#sec-cf}

## The idea

Two-stage least squares removes endogeneity by projection: it keeps only
the part of $D$ that the instrument predicts. A control function removes it
by conditioning: it adds to the outcome equation the part of $D$ the
instrument does *not* predict. Write the first stage as
$D = g(Z, X) + V$. If the dependence between $D$ and the structural error
$U$ runs entirely through $V$,
$$
E[U \mid D, Z, X] = E[U \mid V],
$$
then conditioning on $V$ makes $D$ exogenous. $V$ is not observed, but its
estimate, the first-stage residual $\hat V$, is. Put $\hat V$ in the second
stage as an extra regressor. That regressor is the control function
(Heckman and Robb 1985; Wooldridge 2015).

The condition is stronger than what 2SLS needs. 2SLS needs
$E[U Z] = 0$; the control function needs $V$ independent of $(Z, X)$, or at
least $E[U \mid V, Z, X] = E[U \mid V]$ with a known (usually linear) form.
In exchange it identifies the structural function and works in models where
"plug in the fitted value" has no meaning.

## Linear model: the same number as 2SLS {#sec-cf-linear}

In the linear model, OLS of $Y$ on $(D, X, \hat V)$ gives exactly the 2SLS
coefficient on $D$, just- or over-identified. *Skeleton.* $D = \hat D + \hat V$
where $\hat D$ is the first-stage fitted value. Including $\hat V$ as a
regressor is the same as including $\hat D$ and $\hat V$ (a reparameterization).
By Frisch-Waugh-Lovell, the coefficient on $D$ in that regression is the
coefficient on $\hat D$ after partialling out $\hat V$ and $X$; $\hat D$ is
orthogonal to $\hat V$ by construction, so this is the regression of $Y$ on
$\hat D$ and $X$, which is 2SLS (appendix \@ref(app-cf)).

```{r cf-linear}
cf <- cf_residuals(dat, d = "d", z = "z", x = xc)
cf_ols <- lm_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5 + v_hat, data = cf)
c(control_function = coef(cf_ols)[["d"]], tsls = coef(tsls)[["d"]])
tidy(cf_ols)[tidy(cf_ols)$term == "v_hat", c("estimate", "std.error", "statistic", "p.value")]
```

The coefficient on $\hat V$ is the endogeneity parameter, and its t test is
the Wu-Hausman test of Section \@ref(sec-tests). The standard error on
$D$ printed by the second-stage regression is *not* the 2SLS standard error:
$\hat V$ is a generated regressor, and the naive formula ignores its
sampling variation. It is correct under the null of no endogeneity, which
is why the t test on $\hat V$ is fine, and wrong otherwise. Use the 2SLS
standard errors, or bootstrap both stages (Section \@ref(sec-cf-inference)).

## Nonlinear outcomes {#sec-cf-nonlinear}

Now let the outcome be binary, $Y = \mathbb{1}\{\alpha + \theta D + X'\beta + U > 0\}$,
with $U$ correlated with $D$. Two temptations and one fix.

- **Naive probit** of $Y$ on $(D, X)$ ignores the endogeneity.
- **The forbidden regression** runs the probit on $(\hat D, X)$. It is wrong
  because $E[\Phi(\alpha + \theta D + \cdots) \mid Z, X] \neq \Phi(\alpha + \theta \hat D + \cdots)$:
  the nonlinearity does not pass through the expectation.
- **The control function** (Rivers and Vuong 1988) runs the probit on
  $(D, X, \hat V)$. Under joint normality of $(U, V)$, $U = \rho V + \varepsilon$
  with $\varepsilon$ independent of everything, so conditioning on $V$
  restores a probit with a rescaled index.

The coefficients of the control-function probit are scaled by
$\sigma_{U \mid V}$, so they are not comparable to the naive ones. The object
to report is the average partial effect (APE), computed from the average
structural function $\mathrm{ASF}(d, x) = E_V[\Phi(\alpha + \theta d + x'\beta + \rho V)]$,
averaging over the empirical distribution of $\hat V$ (Blundell and Powell
2004). `cf_ape()` does the averaging; Poisson, tobit, and fractional
responses use the same recipe.

```{r cf-probit-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5, sim_md5, n_sims_cf)}
cf_mc <- run_cf_experiment(n_sims = n_sims_cf, n = 2000, rho = 0.6, seed = 3)
cf_tab <- cf_mc %>% group_by(estimator) %>%
  summarise(bias = mean(ape - truth), sd = sd(ape), rmse = sqrt(mean((ape - truth)^2)), .groups = "drop") %>%
  mutate(estimator = factor(estimator, levels = c("Naive probit", "Forbidden regression", "Control function"))) %>%
  arrange(estimator)
```

```{r tab-cf, echo = FALSE}
cf_tab %>% mutate(across(c(bias, sd, rmse), ~ fmt(.x, 4))) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Estimator", "Bias", "SD", "RMSE"),
      caption = paste0("Average partial effect of an endogenous continuous regressor in a probit model, ",
                       n_sims_cf, " simulations, n = 2000, error correlation 0.6. The truth is the average derivative of the structural probability.")) %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

```{r cf-probit-numbers, echo = FALSE}
pick_cf <- function(e, col) cf_tab[[col]][cf_tab$estimator == e]
```

The naive probit's APE is biased by `r fmt(pick_cf("Naive probit", "bias"), 3)`
on a truth of about 0.15, half again too large. The forbidden regression
has bias `r fmt(pick_cf("Forbidden regression", "bias"), 3)` here, which
is small because the design is close to linear in the index; it is not a
consistent estimator and can be far off when the nonlinearity matters. The
control function has bias `r fmt(pick_cf("Control function", "bias"), 3)`.

## Binary endogenous regressor {#sec-cf-binary}

When $D$ is binary, a linear first stage is not a structural equation and
its residual is not the right control: $D - \hat p$ is a function of the
same threshold-crossing error for two different reasons. Three options.

- A **bivariate probit** models $(D, Y)$ jointly with normal errors. It
  identifies the ATE, but the normality does all the work.
- A **generalized residual** from a probit first stage,
  $D \lambda(z'\hat\gamma) - (1 - D)\lambda(-z'\hat\gamma)$ with $\lambda$
  the inverse Mills ratio, is the control function under the
  threshold-crossing model (Wooldridge 2015, Section 5);
  `cf_residuals(family = "probit")` returns it.
- **Give up the ATE** and report the LATE (Section \@ref(sec-late)) or
  the marginal treatment effect curve (Section \@ref(sec-mte)), which is
  the control-function method with the propensity score as the control and
  no functional-form assumption on how $U$ depends on it.

```{r cf-generalized}
dp <- sim_iv(3000, dgp = "probit_cf", rho = 0.6, seed = 4)
dp$d_binary <- as.integer(dp$d > 0)
gr <- cf_residuals(dp, d = "d_binary", z = "z", x = "x1", family = "probit")
summary(gr$v_hat)
```

## Heterogeneous effects {#sec-cf-heterogeneous}

With a random coefficient, $Y = \alpha + \beta_i D + X'\gamma + U$, the
control function needs an extra term. If $E[\beta_i \mid V] = \bar\beta + \psi V$,
then $E[Y \mid D, X, V] = \alpha + \bar\beta D + \psi D V + X'\gamma + \rho V$:
include $\hat V$ and $D \hat V$ (Garen 1984; the correlated random
coefficient model). The coefficient on $D$ is the average effect. Heckman's
(1979) selection correction is the same idea for a binary selection
equation: the inverse Mills ratio is the control function. And the marginal
treatment effect of Section \@ref(sec-mte) is what remains when the
dependence of $U$ on the control is left unrestricted.

```{r cf-garen}
cf_g <- lm_robust(y ~ d + x1sq + s2x2 + x3 + x4 + x5 + v_hat + d:v_hat, data = cf)
tidy(cf_g)[tidy(cf_g)$term %in% c("d", "v_hat", "d:v_hat"), c("term", "estimate", "std.error")]
```

In the running design the effect is homogeneous, so the interaction is
zero within sampling error and the coefficient on $D$ does not move.

## Dynamics and demand {#sec-cf-demand}

Two applications from the replications show the reach of the method.

- **Accumulated exposure.** Hsieh et al. (2026) estimate the effect of
  television advertising on purchases with an ad stock
  $AS_{it} = \sum_l \lambda^l A_{i,t-l}$. Their instrument is realized minus
  expected exposure (Section \@ref(sec-exposure)); the first-stage residuals
  are accumulated into a control-function stock
  $CFS_{it} = \sum_l \lambda_{CF}^l \hat\varepsilon_{i,t-l}$ that enters the
  purchase model. The endogenous variable accumulates, so the control does
  too.
- **Price in discrete choice.** Petrin and Train (2010) put the residual
  of a regression of price on cost shifters into the utility of a
  multinomial logit. The vignette `usage_iv` simulates markets with an
  unobserved quality shock that raises both price and demand and shows the
  naive logit halving the price coefficient while the control function
  recovers it. This is the workhorse for price endogeneity in marketing
  when instruments exist at the product-market level but the demand model
  is nonlinear.

## Inference for two-step estimators {#sec-cf-inference}

Every control-function estimator has a generated regressor. Second-stage
standard errors that treat $\hat V$ as data understate uncertainty unless
the endogeneity parameter is zero. Two fixes: the Murphy-Topel correction,
which needs the two scores, or a bootstrap of both stages, which needs
nothing. `cf_bootstrap()` re-runs a user-supplied first and second stage on
pairs (or clusters) of observations and reports standard errors and
percentile intervals for coefficients and APEs.

```{r cf-bootstrap}
first_stage <- function(df) cf_residuals(df, d = "d", z = "z", x = "x1")
second_stage <- function(df) glm(y ~ d + x1 + v_hat, data = df, family = binomial("probit"))
cf_bootstrap(dp, first_stage, second_stage, n_boot = 199, seed = 1,
             ape = function(fit, df) cf_ape(fit, df, d = "d"))
c(true_ape = attr(dp, "ape"))
```

Cluster the bootstrap at the level at which the instrument varies. When the
first stage is estimated at a coarser level than the second (prices at the
product-market level, choices at the consumer level), resample at the
coarser level, as the vignette's demand example does.

# Orthogonal scores for IV {#sec-orthogonal}

## Partially linear IV {#sec-pliv}

Return to the partially linear model with an instrument,
$$
Y = \theta D + f_Y(X) + U, \qquad E[U Z \mid X] = 0,
$$
where the controls $X$ enter through an unknown function. Partial out $X$
from everything: $\tilde Y = Y - E[Y \mid X]$, $\tilde D = D - E[D \mid X]$,
$\tilde Z = Z - E[Z \mid X]$. Then $E[(\tilde Y - \theta \tilde D)\tilde Z] = 0$
and
$$
\theta = \frac{E[\tilde Y \tilde Z]}{E[\tilde D \tilde Z]},
$$
Wright's ratio on residualized variables (Chernozhukov et al. 2026,
Theorem 12.3.1). The moment $\psi(W; \theta, \eta) = (Y - \ell(X) - \theta(D - r(X)))(Z - m(X))$
with nuisances $\eta = (\ell, r, m)$ is Neyman orthogonal: its derivative
with respect to each nuisance vanishes at the truth (appendix
\@ref(app-orthogonal-iv)). So the recipe of Section 04 applies: cross-fit
the three regressions with any learner, form the residuals, and run IV on
them. Standard errors are those of IV on the residuals. The score needs all
three nuisances to converge at $n^{-1/4}$; there is no pair of nuisances
that protects against the failure of the third, so this is orthogonal but
not doubly robust.

What does $\theta$ mean when the outcome model is nonlinear? Chernozhukov
et al. (2026, Theorems 12.3.2-12.3.3) show the ratio identifies the average
marginal effect $E[\partial_d Y(d)]$ if either the effect of $D$ or the
effect of $Z$ on $D$ is homogeneous in the unobservable. When both vary
with the unobservable, the ratio is a weighted average of effects, as in
Section \@ref(sec-weighted), and the LATE machinery of Section
\@ref(sec-late-score) is the right description.

## Code and a Monte Carlo {#sec-pliv-code}

In the running design the instrument is excluded only conditional on
$h(x) = 0.5 x_1^2 + 0.5 \sin(2 x_2)$, which also enters the outcome. Linear
controls leave part of $h$ in the error, and the residualized instrument is
correlated with it.

```{r pliv-fit}
pliv_lin <- est_dml(dat, "y", "d", x5, z = "z", model = "pliv", seed = 1)
pliv_rf <- est_dml(dat, "y", "d", x5, z = "z", model = "pliv",
                   learner_l = rf, learner_m = rf, learner_z = rf, seed = 1, weak_iv = TRUE)
pliv_rf
rbind(linear = tidy(pliv_lin)[, c("estimate", "std.error")], forests = tidy(pliv_rf)[, c("estimate", "std.error")])
```

```{r pliv-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5, sim_md5, n_sims_pliv)}
pliv_mc <- run_pliv_experiment(n_sims = n_sims_pliv, n = 2000, seed = 2)
pliv_tab <- summarise_experiment(pliv_mc, truth = 1) %>%
  mutate(estimator = factor(estimator, levels = c("OLS", "2SLS, linear controls", "DML-IV, forests"))) %>%
  arrange(estimator)
```

```{r tab-pliv, echo = FALSE}
pliv_tab %>% mutate(across(c(bias, sd, rmse, mean_se, coverage), ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Estimator", "Bias", "SD", "RMSE", "Mean SE", "95\\% coverage"),
      caption = paste0("Partially linear IV design with a nonlinear confounder path; true effect 1; n = 2000; ",
                       n_sims_pliv, " simulations.")) %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

```{r fig-pliv, echo = FALSE, fig.height = 3.2, fig.cap = "Sampling distributions of the three estimators in the partially linear IV design. OLS is biased by the unobserved confounder; 2SLS with linear controls is biased because the instrument is excluded only conditional on nonlinear functions of the covariates; DML-IV with forests removes most of that bias."}
plot_sampling(pliv_mc, truth = 1)
```

```{r pliv-numbers, echo = FALSE}
pick_pliv <- function(e, col) pliv_tab[[col]][pliv_tab$estimator == e]
```

OLS carries the confounder bias (`r fmt(pick_pliv("OLS", "bias"))`). 2SLS
with linear controls removes it and adds a different one
(`r fmt(pick_pliv("2SLS, linear controls", "bias"))`): the instrument is
not excluded given linear $X$. DML-IV with forests has bias
`r fmt(pick_pliv("DML-IV, forests", "bias"))` and coverage
`r fmt(pick_pliv("DML-IV, forests", "coverage"))`; what remains is the
forests' error in learning $h$, which shrinks with $n$ as Section 04
discussed. The instrument is an assumption about conditioning as much as an
assumption about exclusion, and the conditioning is where machine learning
enters.

## The LATE score: apply the AIPW construction twice {#sec-late-score}

With a binary $Z$ and binary $D$, the conditional LATE of Section
\@ref(sec-conditional-late) is a ratio of two average treatment effects of
$Z$: on $Y$ and on $D$. Each is identified by conditioning on $X$ and has
the doubly robust score of Section 04 with the instrument propensity
$p(X) = P(Z = 1 \mid X)$ in the weights (Chernozhukov et al. 2026, Section
13.2):
$$
\psi(W; \theta, \eta) = \underbrace{\mu_1(X) - \mu_0(X) + H(p)(Y - \mu_Z(X))}_{\text{effect of } Z \text{ on } Y}
 - \theta\,\underbrace{\big[m_1(X) - m_0(X) + H(p)(D - m_Z(X))\big]}_{\text{effect of } Z \text{ on } D},
$$
with $\mu_z(X) = E[Y \mid Z = z, X]$, $m_z(X) = E[D \mid Z = z, X]$, and
$H(p) = Z/p - (1 - Z)/(1 - p)$. The score is linear in $\theta$, so
$\hat\theta$ is the ratio of the two sample means and the influence function
is the score divided by the mean of the second bracket, the complier share.
Each bracket is doubly robust in the sense of Section 04, and the ratio is
Neyman orthogonal (appendix \@ref(app-orthogonal-iv)). This is the doubly
robust IV estimator that Section 04's extension list promised.

```{r iivm}
late_dml <- est_dml(dl, "y", "d", c("x1", "x2"), z = "z", model = "iivm",
                    learner_mu0 = rf, learner_mu1 = rf, seed = 1, weak_iv = TRUE)
late_dml
c(true_late = attr(dl, "late"))
```

```{r late-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5, sim_md5, n_sims_late)}
late_mc <- run_late_experiment(n_sims = n_sims_late, n = 2000, seed = 4)
late_tab <- late_mc %>% group_by(estimator) %>%
  summarise(bias = mean(estimate - truth), sd = sd(estimate), rmse = sqrt(mean((estimate - truth)^2)),
            mean_se = mean(std.error), coverage = mean(abs(estimate - truth) <= 1.96 * std.error), .groups = "drop")
```

```{r tab-late, echo = FALSE}
late_tab %>% mutate(across(c(bias, sd, rmse, mean_se, coverage), ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Estimator", "Bias", "SD", "RMSE", "Mean SE", "95\\% coverage"),
      caption = paste0("LATE design with compliance and effects that vary with the covariates; n = 2000; ",
                       n_sims_late, " simulations. Bias is relative to each sample's population LATE.")) %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

```{r late-numbers, echo = FALSE}
pick_late <- function(e, col) late_tab[[col]][late_tab$estimator == e]
```

In this design the compliance probability and the complier effect both
vary with $x_1$, and the untreated outcome is quadratic in $x_2$. Both
estimators are close to the population LATE on average (biases
`r fmt(pick_late("2SLS, linear controls", "bias"), 3)` and
`r fmt(pick_late("DML-LATE", "bias"), 3)`). The doubly robust score is more
precise, with root mean squared error `r fmt(pick_late("DML-LATE", "rmse"))`
against `r fmt(pick_late("2SLS, linear controls", "rmse"))`, because the
forests absorb the quadratic term that linear controls leave in the
residual. Its standard error also matches its sampling variation (coverage
`r fmt(pick_late("DML-LATE", "coverage"))`), while the 2SLS interval
over-covers (`r fmt(pick_late("2SLS, linear controls", "coverage"))`). The
two estimands coincide here because the compliance varies linearly with
$x_1$; when it does not, 2SLS targets a linear projection and the score
targets the ratio of averages, and only the second is the LATE.

## Observed heterogeneity: the complier CATE {#sec-late-blp}

The two brackets of the LATE score are pseudo-outcomes in the sense of
Section 08: conditional on $X$, their means are the conditional effects of
$Z$ on $Y$ and on $D$. Projecting each on a dictionary $p(X)$ and taking
the ratio gives the complier CATE $E[Y(1) - Y(0) \mid X, \text{complier}]$
when it is linear in the dictionary (Chernozhukov et al. 2026, eq. 12.3.1;
Abadie 2003). `late_scores()` extracts the two scores and `late_blp()`
forms the ratio with delta-method standard errors.

```{r late-blp}
ls <- late_scores(late_dml)
late_blp(ls, ~ x1, newdata = data.frame(x1 = c(-1, 0, 1)))[, c("x1", "estimate", "std.error", "conf.low", "conf.high")]
```

The complier effect in the design is $2 + x_1$. The same scores feed
`cate_learner()` for a flexible complier CATE and `policy_learn()` for
rules whose value is measured among compliers; `grf::instrumental_forest()`
is the forest version.

## Proxy controls as a technical instrument {#sec-proxy}

Sometimes the confounder is unobserved but two noisy proxies of it are:
two test scores for ability, education and family size for household
income. Griliches and Chamberlain's insight is that one proxy can be put in
the outcome equation in place of the confounder, and the other proxy used
as an instrument for it, because the measurement errors are independent.
In the partially linear form this is `est_dml(model = "pliv")` with the
outcome proxy as the endogenous regressor and the treatment proxy as the
instrument, and the treatment of interest as an exogenous control
(Chernozhukov et al. 2026, Section 12.5). The proximal-inference literature
extends the idea to nonlinear models.

## Control function or orthogonal score? {#sec-cf-vs-score}

Both start from the same first stage. The control function conditions on
$\hat V$ inside a parametric second stage and identifies the structural
function under an independence assumption on $V$. The orthogonal score
projects on the residualized instrument and identifies an average effect
under an exclusion restriction, letting machine learning absorb $X$. Use
the control function when the second stage must be nonlinear and structural
(a probit, a demand system, a stock of exposure); use the orthogonal score
when the target is an average effect and the controls are many or their
functional form unknown. In the linear model with linear controls, they are
the same regression.

# Weak instruments {#sec-weak}

## Why a small first stage breaks the approximation {#sec-weak-problem}

With one instrument and controls partialled out, $\hat\theta - \theta = \mathbb{E}_n[\tilde Z U] / \mathbb{E}_n[\tilde Z \tilde D]$.
The usual approximation treats the denominator as fixed at its limit. When
$E[\tilde Z \tilde D]$ is small relative to the sampling variation of its
estimate, the denominator can be near zero, and the ratio is heavy-tailed,
biased toward OLS (the numerator and denominator share the endogenous
variation), and far from normal. The strength of the first stage is
summarized by the concentration parameter $\mu^2 = n\,\pi^2 \mathrm{Var}(\tilde Z) / \mathrm{Var}(V)$,
of which the first-stage F is the sample analogue. Figure
\@ref(fig:fig-weak) shows the sampling distribution of 2SLS for three values.

```{r weak-mc, echo = FALSE, cache = TRUE, cache.extra = list(helper_md5, sim_md5, n_sims_weak)}
weak_mc <- run_weak_iv_experiment(n_sims = n_sims_weak, n = 500, concentrations = c(2, 10, 50), seed = 5)
weak_tab <- weak_mc %>% group_by(concentration) %>%
  summarise(median_F = median(first_stage_F), median_2sls = median(tsls), mean_ols = mean(ols),
            sd_2sls = sd(tsls), wald_coverage = mean(wald_covers), ar_coverage = mean(ar_covers),
            ar_bounded = mean(ar_bounded), .groups = "drop")
```

```{r fig-weak, echo = FALSE, fig.height = 3.2, fig.cap = "Sampling distribution of the 2SLS estimate (true effect 1, n = 500) for concentration parameters 2, 10, and 50. The dotted line is the mean OLS estimate. Estimates beyond three units from the truth are trimmed for display."}
weak_plot <- weak_mc %>% mutate(concentration = factor(paste0("concentration = ", concentration),
                                                       levels = paste0("concentration = ", c(2, 10, 50)))) %>%
  filter(abs(tsls - 1) < 3)
ols_lines <- weak_mc %>% group_by(concentration) %>% summarise(ols = mean(ols), .groups = "drop") %>%
  mutate(concentration = factor(paste0("concentration = ", concentration), levels = levels(weak_plot$concentration)))
ggplot(weak_plot, aes(x = tsls)) +
  geom_histogram(bins = 50, fill = iv_colors[["2SLS"]], alpha = 0.8, colour = "white", linewidth = 0.1) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(data = ols_lines, aes(xintercept = ols), linetype = "dotted", colour = iv_colors[["OLS"]]) +
  facet_wrap(~ concentration, scales = "free_y") +
  labs(x = "2SLS estimate", y = "Simulations") + iv_theme()
```

```{r tab-weak, echo = FALSE}
weak_tab %>% mutate(across(-concentration, ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Concentration", "Median F", "Median 2SLS", "Mean OLS", "SD 2SLS", "Wald coverage", "AR coverage", "AR set bounded"),
      caption = paste0("Weak-instrument Monte Carlo, ", n_sims_weak, " simulations per design, n = 500, true effect 1, nominal 95 percent coverage.")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

```{r weak-numbers, echo = FALSE}
weak_tail <- weak_mc %>% group_by(concentration) %>% summarise(tail = mean(abs(tsls - 1) > 1), .groups = "drop")
```

With a concentration parameter of 2 the median first-stage F is
`r fmt(weak_tab$median_F[1], 1)`. The 2SLS distribution is wide and heavy
tailed: `r fmt(100 * weak_tail$tail[1], 0)` percent of the samples land
more than one unit from the truth, against `r fmt(100 * weak_tail$tail[3], 0)`
percent at a concentration parameter of 50, and its mean is pulled toward
the OLS value of about `r fmt(weak_tab$mean_ols[1], 1)`. The Wald interval
covers the truth in `r fmt(100 * weak_tab$wald_coverage[1], 0)` percent of
samples instead of 95. At a concentration parameter of 50 the
approximation is fine. The F of 10 in between is the rule of thumb of
Staiger and Stock (1997): with F around 10 the bias of 2SLS relative to
OLS is about 10 percent. Lee, McCrary, Moreira, and Porter (2022) show that
the rule is not enough for inference: across all values of the
endogeneity correlation, a first-stage F must exceed about 104.7 for the
usual 1.96 critical value to deliver 95 percent coverage, and they tabulate
the inflated critical values to use below that.

## Diagnostics and their limits {#sec-weak-diagnostics}

Report the first-stage F, the robust F, and the effective F of Montiel
Olea and Pflueger (2013), which is the statistic to compare with
weak-instrument thresholds when errors are heteroskedastic or clustered.
`iv_first_stage()` gives all three. Then be careful with what the numbers
license. A pretest ("proceed with the Wald interval if F > 10") distorts
the inference that follows: conditional on passing, the estimator's
distribution is not the unconditional one (Andrews, Stock, and Sun 2019).
The diagnostics are for transparency. The decision should be to use
inference that does not need them.

```{r first-stage}
iv_first_stage(dat, d = "d", z = "z", x = x5)
```

## Inference that does not need a strong first stage {#sec-ar}

The Anderson-Rubin (1949) test never inverts the first stage. For a
candidate value $\theta_0$, form $Y - \theta_0 D$ and test whether the
instrument predicts it: under $H_0: \theta = \theta_0$ the coefficient of
$\tilde Z$ in a regression of $\tilde Y - \theta_0 \tilde D$ on $\tilde Z$
is zero, and the robust Wald statistic is $\chi^2$ with as many degrees of
freedom as instruments, however weak the instrument. Collecting all
$\theta_0$ that are not rejected gives a confidence set. With one instrument
the set is the solution of a quadratic inequality: bounded when the
instrument is strong, unbounded or the whole real line when it is not. That
is the honest answer. The reduced-form test ($\theta = 0$ if and only if
the reduced-form coefficient is zero) is the special case $\theta_0 = 0$.

```{r ar}
weak <- sim_iv(500, dgp = "weak", concentration = 3, seed = 6)
ar <- iv_ar_confidence_set(weak, "y", "d", "z", x = paste0("x", 1:3))
ar
```

```{r ar-plot, echo = FALSE, fig.height = 3.4, fig.cap = "The Anderson-Rubin statistic as a function of the candidate coefficient in a weak-instrument sample. The confidence set is where the statistic lies below the critical value; with a weak instrument it is wider than the Wald interval, asymmetric around the 2SLS estimate, and can be unbounded."}
plot_ar_set(ar)
```

In the Monte Carlo, the Anderson-Rubin set covers the truth in
`r fmt(100 * weak_tab$ar_coverage[1], 0)`, `r fmt(100 * weak_tab$ar_coverage[2], 0)`,
and `r fmt(100 * weak_tab$ar_coverage[3], 0)` percent of samples at the three
concentration values, at the cost of being unbounded in
`r fmt(100 * (1 - weak_tab$ar_bounded[1]), 0)` percent of the weakest
samples. Other tests (the conditional likelihood ratio, the Lagrange
multiplier) improve power when the model is overidentified; with one
instrument the Anderson-Rubin test is optimal (Andrews, Stock, and Sun
2019).

## Weak identification with machine-learned controls {#sec-weak-dml}

The Anderson-Rubin statistic is Neyman's $C(\alpha)$ statistic for the
orthogonal moment: $C(\theta_0) = n\, \bar\psi(\theta_0)^2 / \widehat{\mathrm{Var}}(\psi(\theta_0))$
with $\psi$ evaluated at cross-fitted nuisances. Because the moment is
orthogonal, the nuisance estimation error has no first-order effect on
$C(\theta_0)$ either, so the test and the set inherit their validity with
machine-learned residuals (Chernozhukov et al. 2026, Theorem 13.3.1). The
same construction works for any orthogonal score, including the LATE score,
which is what `weak_iv = TRUE` computes in `est_dml()`.

```{r weak-dml}
pliv_rf$weak_iv$set
c(wald_lower = pliv_rf$conf.low, wald_upper = pliv_rf$conf.high)
```

With a strong residualized instrument the two intervals agree. The
Chernozhukov et al. (2026) example of institutions and growth, with settler
mortality as the instrument and a first-stage t of about 2 after forests,
is the case where they do not: the robust set's lower end is higher than
the Wald interval's, which changes the economic conclusion.

## Exclusion violations are amplified by weak first stages {#sec-plausibly}

If the instrument has a small direct effect $\gamma$ on the outcome, the
IV estimand is $\theta + \gamma / \pi$: the violation is divided by the
first-stage coefficient. Bound, Jaeger, and Baker (1995) made the point
for quarter of birth as an instrument for schooling, where a first stage of
0.1 years multiplies any direct effect by ten. Conley, Hansen, and Rossi
(2012) turn it into an interval: allow $\gamma$ in a range, re-estimate
with $Y - \gamma Z$ as the outcome, and take the union of the confidence
intervals.

```{r plausibly}
pe <- iv_plausibly_exogenous(dat, "y", "d", "z", x5, gamma_grid = seq(-0.2, 0.2, by = 0.1))
pe
attr(pe, "union")
```

A direct effect of the instrument of 0.2 in either direction, which is a
fifth of the first-stage coefficient here, moves the estimate by about
`r fmt(abs(pe$estimate[1] - pe$estimate[3]) / 2, 2)`. With a first stage
one tenth as large the same violation would move it ten times as much.

# Many instruments, many effects, constructed instruments {#sec-many}

## Several instruments: what 2SLS averages {#sec-many-weights}

With several instruments the 2SLS coefficient is a weighted average of the
just-identified estimates that use one instrument at a time,
$$
\hat\theta_{2SLS} = \sum_k \omega_k \hat\theta_k, \qquad
\hat\theta_k = \frac{\mathrm{Cov}(Z_k, Y)}{\mathrm{Cov}(Z_k, D)}, \qquad
\omega_k = \frac{\hat\pi_k \mathrm{Cov}(Z_k, D)}{\sum_j \hat\pi_j \mathrm{Cov}(Z_j, D)},
$$
with $\hat\pi$ the first-stage coefficients (appendix \@ref(app-weights)).
The weights sum to one but need not be positive. With homogeneous effects
every $\hat\theta_k$ estimates the same number and the weights are
irrelevant. With heterogeneous effects each instrument has its own
compliers and its own LATE, and Mogstad, Torgovitsky, and Walters (2021)
show that the Imbens-Angrist monotonicity condition applied to the vector
of instruments (every unit responds to every instrument in the same
direction) is much stronger than it looks. Under the weaker *partial
monotonicity* (each instrument is monotone on its own), the 2SLS weights
on the instrument-specific LATEs can be negative, and the sign of
$\hat\theta_{2SLS}$ can differ from the sign of every LATE. Their test
checks the sign patterns of the first stages and reduced forms that
positive weights require. `iv_late_weights()` reports the decomposition.

```{r late-weights}
dat$z2 <- dat$z + rnorm(nrow(dat))
dat$z3 <- rnorm(nrow(dat)) + 0.1 * dat$z
lw <- iv_late_weights(dat, "y", "d", c("z", "z2", "z3"), x = xc)
lw
```

Here the effect is homogeneous, so all three instrument-specific estimates
are near 1 and the weights only redistribute precision. The instrument with
the small first stage gets a small weight; when a weight is negative and
the effects differ, the 2SLS number is not an average of anything a policy
would deliver. The remedy is to report the instrument-specific estimates
and the weights, or to move to the marginal treatment effect framework of
Section \@ref(sec-mte), which makes the weighting explicit.

## Judge leniency and examiner designs {#sec-judges}

The most common instrument in recent applied work is the assigned decision
maker: judges for incarceration, examiners for disability insurance,
reviewers for patents, loan officers for credit. Cases are assigned at
random within a court and year; judges differ in leniency; leniency moves
the treatment. The instrument is the leave-one-out leniency of the assigned
judge,
$$
Z_i = \frac{1}{n_{j(i)} - 1} \sum_{k \neq i:\, j(k) = j(i)} D_k,
$$
computed within the randomization cell and after residualizing $D$ on
controls so that leniency is compared among judges who see the same case
mix. Leaving out the own case avoids the mechanical correlation between $Z_i$
and $D_i$ that would bias 2SLS toward OLS. Standard errors cluster at the
judge level.

```{r judges}
dj <- sim_iv(6000, dgp = "judge", n_judges = 40, seed = 7)
dj <- leniency_instrument(dj, examiner = "judge", d = "d", x = "x1")
iv_first_stage(dj, d = "d", z = "leniency", x = "x1", cluster = "judge")
jfit <- iv_robust(y ~ d + x1 | leniency + x1, data = dj, clusters = judge)
c(estimate = coef(jfit)[["d"]], se = jfit$std.error[["d"]])
```

The design's monotonicity condition is the delicate part. Monotonicity
across judges requires that a stricter judge incarcerates every defendant a
more lenient judge would; judges who are strict on drug cases and lenient
on theft violate it. What 2SLS identifies under the weaker *average
monotonicity* condition (Frandsen, Lefgren, and Leslie 2023) is a
positively weighted average of individual effects only if the
judge-specific first stages and effects are positively correlated; the same
sign checks as in Section \@ref(sec-many-weights) apply, and the exclusion
restriction (judges affect outcomes only through the decision) is itself
testable with the leniency distribution.

## Continuous instruments and the marginal treatment effect {#sec-mte}

The LATE model with a binary instrument is a threshold model in disguise:
monotonicity is equivalent to $D = \mathbb{1}\{p(Z, X) \ge V\}$ with
$V \sim U(0, 1)$ independent of $(Z, X)$, where $p$ is the propensity of
treatment and $V$ the unit's resistance to it (Vytlacil 2002). With a
continuous instrument the propensity takes many values, and comparing units
with propensities $p$ and $p + h$ identifies the effect for units with
$V \in (p, p + h]$. Taking $h \to 0$ gives the marginal treatment effect
$$
\mathrm{MTE}(u, x) = E[Y(1) - Y(0) \mid X = x, V = u] = \frac{\partial E[Y \mid X = x, p(Z, X) = u]}{\partial u},
$$
the local instrumental variables estimand of Heckman and Vytlacil (2005).
This is the control-function approach of Section \@ref(sec-cf) with the
propensity as the control and no functional-form assumption about how the
unobservable enters: $E[Y \mid X, p] = X'\beta_0 + p X'(\beta_1 - \beta_0) + K(p)$
and $K'(u)$ is the part of the MTE that varies with resistance.

Every standard estimand is a weighted integral of the curve (appendix
\@ref(app-mte)): the ATE weights all $u$ equally; the ATT weights
$u$ by $P(p > u) / E[p]$ (units with low resistance are over-represented
among the treated); the ATU by $P(p \le u) / E[1 - p]$; the LATE of an
instrument shift from $p_0$ to $p_1$ uniformly on $(p_0, p_1]$; and a policy
that changes the propensity from $p$ to $p'$ has the policy-relevant
treatment effect with weights $(F_p(u) - F_{p'}(u)) / (E[p'] - E[p])$. The
curve can be integrated only over the support of the estimated propensity;
outside it the estimates are extrapolations of the polynomial.

```{r mte}
dm <- sim_iv(5000, dgp = "mte", seed = 8)
m <- mte_curve(dm, "y", "d", "z", x = "x1", degree = 3, late_range = c(0.3, 0.7),
               policy = function(p) pmin(p + 0.1, 0.99), n_boot = 99, seed = 1)
m
c(true_ate = attr(dm, "ate"), true_att = attr(dm, "att"))
```

```{r mte-plot, echo = FALSE, fig.height = 3.6, fig.cap = "Estimated marginal treatment effect curve (polynomial of degree three in the propensity) with bootstrap bands, and the true curve of the design (dashed). The effect falls with resistance: units easy to move into treatment gain most, which is why the ATT exceeds the ATE."}
truth_mte <- attr(dm, "mte")
plot_mte(m) + geom_line(data = data.frame(u = m$curve$u, mte = truth_mte(m$curve$u)), aes(u, mte), linetype = "dashed")
```

```{r mte-weights, echo = FALSE, fig.height = 3.2, fig.cap = "The weights that turn the MTE curve into the ATE, ATT, ATU, the LATE of a propensity shift from 0.3 to 0.7, and the policy-relevant effect of raising every propensity by 0.1."}
plot_mte(m, "weights")
```

The MTE framework answers the question the LATE cannot: what would a
policy that moves a different margin do? It costs a continuous instrument
with wide support and a parametric or smoothing choice for $K$. Mogstad,
Torgovitsky, and Walters (2021) use instrument-specific MTE curves to show
that two instruments for college attendance move different people; the
replication reproduces that exercise.

## Many candidate instruments {#sec-many-candidates}

When many variables could serve as instruments, two problems appear. Using
all of them in 2SLS adds bias: each instrument adds a little of the
endogenous variation back into $\hat D$, and with $K$ instruments the bias
toward OLS grows with $K / \mu^2$ (Bekker 1994). Limited-information
maximum likelihood and jackknife IV (which predicts $D_i$ from a first stage
that excludes unit $i$) reduce it. Choosing among instruments by looking at
the outcome invalidates inference.

Belloni, Chen, Chernozhukov, and Hansen (2012) solve both problems with the
lasso: select instruments by their first-stage predictive power alone,
which is a prediction problem and does not use $Y$, then run 2SLS with the
selected set (or the lasso first-stage prediction as the instrument). Under
approximate sparsity the resulting estimator is as good as knowing the
relevant instruments. In `est_dml(model = "pliv")` the same idea is the
choice of `learner_m` for $E[Z \mid X]$ and of the instrument set; the `hdm`
package implements `rlassoIV()` directly. Gilchrist and Sands (2016) use
this to choose among dozens of weather variables as instruments for opening
weekend movie attendance, then estimate how much later attendance responds
to the exogenous opening shock: social spillovers identified from rain.

## A constructed instrument: shift-share {#sec-ssiv}

Many instruments are built rather than found. The shift-share (Bartik)
instrument combines local exposure shares with aggregate shocks:
$$
B_i = \sum_k s_{ik}\, g_k,
$$
local industry employment shares times national industry growth (Bartik
1991), or, in Autor, Dorn, and Hanson (2013), commuting-zone industry
shares times the growth of Chinese imports in other rich countries. Two
identification stories justify it and they have different implications.

- **Exogenous shares** (Goldsmith-Pinkham, Sorkin, and Swift 2020): the
  shares are as good as randomly assigned given controls, and the shocks
  only provide the weights. Then $B_i$ is a weighted combination of $K$
  instruments (the shares), and the estimate is a weighted average of the
  share-specific just-identified estimates with Rotemberg weights
  $\alpha_k = g_k \mathrm{Cov}(s_k, D^\perp) / \sum_j g_j \mathrm{Cov}(s_j, D^\perp)$.
  The weights say which industries drive the estimate; a few large weights
  mean the identifying assumption is about those industries.
- **Exogenous shocks** (Borusyak, Hull, and Jaravel 2022): the shocks are
  as good as randomly assigned across industries, and the shares can be
  endogenous. Then the estimate equals a shock-level regression of the
  exposure-weighted outcome on the exposure-weighted treatment instrumented
  by the shock, and inference should be at the shock level, with as many
  effective observations as there are independent shocks.

`ssiv_rotemberg()` and `ssiv_shock_level()` compute the two diagnostics.

```{r ssiv}
ds <- sim_iv(600, dgp = "shift_share", n_industries = 15, seed = 9)
sh <- paste0("s", 1:15)
rot <- ssiv_rotemberg(ds, "y", "d", shares = sh, shocks = attr(ds, "shocks"), x = "x1")
rot
ssiv_shock_level(ds, "y", "d", shares = sh, shocks = attr(ds, "shocks"), x = "x1")
```

The two point estimates coincide by construction; the standard errors do
not, because the second counts `r rot$n` regions as
`r length(attr(ds, "shocks"))` independent shocks. Which one is right
depends on which story the paper tells. The replication of Autor, Dorn,
and Hanson (2013) works through both.

# Instruments in experiments and platforms {#sec-platforms}

## Encouragement designs and A/B tests as instruments {#sec-encouragement}

A randomized offer with imperfect take-up is the cleanest instrument there
is: independence holds by design, exclusion holds if the offer itself does
not change behavior, and monotonicity holds if no one is put off by being
offered. The intention-to-treat effect is the reduced form; the LATE is the
effect among those the offer moved. Digital platforms run these designs
constantly: a recommendation shown at random is an instrument for the
action recommended, a nudge is an instrument for the behavior nudged. The
tools of Sections \@ref(sec-late) and \@ref(sec-late-score) apply without
change, and the complier profile answers the platform's own question, which
is who reacts to the nudge.

## Realized minus expected exposure {#sec-exposure}

Hsieh et al. (2026) estimate the effect of television advertising on
purchases from single-source data. Exposure is endogenous: the brand
targets shows whose viewers are likely buyers. Their instrument uses the
timing of ads within a show. Given that a household watched a portion of a
targeted show, whether it saw the ad depends on whether the ad aired during
that portion, and the expected exposure is the integral of the ad-timing
density over the watched segment. The instrument is realized minus expected
exposure, which is mean zero conditional on targeting and viewing, and
shifts actual exposure one for one. The second stage is a purchase model
with an ad stock, estimated with the control-function stock of Section
\@ref(sec-cf-demand). The replication draft simulates the pipeline
(exposure, purchase, first stage, second stage) with the public code.

## When instruments are not exogenous shocks {#sec-not-shocks}

Garcia, Grewal, and coauthors (2022) estimate hotel demand from managerial
responses to automated price recommendations. The natural instruments
(prices at other hotels) load on correlated demand shocks; the paper argues
they are not excluded and instead conditions on the recommendation, which
absorbs the manager's information, using post-double-selection to handle
the high-dimensional interactions of recommendations with calendar
variables. The lesson for this lecture is that an instrument is a claim
about a mechanism, and when the mechanism is not credible the better design
is the one that conditions on the information the decision maker used.

# Traps {#sec-traps}

## Exclusion failures dressed as natural experiments

Rain is random, but farmers, moviegoers, and voters all respond to rain
through more than one channel. Quarter of birth is random, but it is
correlated with parental characteristics and with health. The question is
never whether $Z$ is random; it is whether $Z$ has *no* other path to $Y$.
Weak instruments make small failures large (Section \@ref(sec-plausibly)).

## LATE is not the ATE

Report the compliers, and say whose effect the number describes. When the
policy question is universal ("treat everyone"), the LATE is one point
inside the bounds of Section \@ref(sec-tests-bounds), and the MTE curve
tells whether the marginal units gain more or less than the average.

## Forbidden regressions

Fitted values from a nonlinear first stage plugged into 2SLS, or fitted
values from a linear first stage plugged into a nonlinear second stage, are
both wrong. The first can be repaired by using the nonlinear fitted value as
an *instrument* in a linear 2SLS; the second by the control function
(Section \@ref(sec-cf-nonlinear)).

## IV larger than OLS

Systematic reviews find that IV estimates exceed OLS far more often than
the usual endogeneity stories predict. Measurement error in $D$ explains
some (OLS is attenuated); weak instruments and small exclusion failures
explain more, because both push IV away from zero in the direction of the
reduced-form noise. Young (2022) shows that IV inference is also unusually
sensitive to a few high-leverage observations. Report the first stage, the
reduced form, a weak-IV-robust interval, and the sensitivity of the estimate
to dropping influential units.

## Bad controls, again

Controls measured after the instrument is realized, or affected by the
treatment, break the exclusion restriction in the same way they broke
conditional exogeneity in Section 03. Judge fixed effects in a leniency
design are fine; the judge's sentence length is not.

# Practice {#sec-practice}

## Replications and anchors {#sec-replications}

- **Mogstad, Torgovitsky, and Walters (2021).** Two instruments for college
  attendance, the 2SLS weights, instrument-specific MTE curves, and the
  test of positive weights; `iv_late_weights()` and `mte_curve()`.
- **Hsieh et al. (2026).** Realized-minus-expected exposure as an
  instrument for advertising, and the control-function stock;
  `cf_residuals()` with user-written accumulation.
- **Garcia et al. (2022).** Why external instruments fail for hotel demand
  and how conditioning on the recommendation with post-double-selection
  replaces them; `est_dml()` and `hdm`.
- **Gilchrist and Sands (2016).** Lasso-chosen weather instruments for
  opening-weekend attendance and social spillovers; `hdm::rlassoIV()`,
  `iv_first_stage()`, `iv_ar_confidence_set()`.
- **Autor, Dorn, and Hanson (2013).** The China shock as a shift-share
  instrument; `ssiv_rotemberg()`, `ssiv_shock_level()`.
- **Oreopoulos (2006).** The school-leaving age reform as an instrument for
  schooling, from Wayne's earlier note; the reporting template of Section
  \@ref(sec-reporting).

## An IV checklist {#sec-checklist}

1. Write the structural equation and say what the error contains. Name
   the instrument in the abstract and give the mechanism for exclusion.
2. Report the first stage, the reduced form, and the IV estimate with the
   same controls and sample; cluster where the instrument varies.
3. Report first-stage strength (F, robust F, effective F) and a
   weak-instrument-robust interval regardless of the F.
4. Say whose effect it is: complier share and characteristics; instrument-
   specific estimates and 2SLS weights when there are several instruments.
5. With a nonlinear outcome, use a control function and report average
   partial effects with bootstrap standard errors.
6. With many controls or unknown functional forms, use the orthogonal IV
   scores with cross-fitting and report the nuisance fit.
7. For a constructed instrument, state which identification story you rely
   on and report the diagnostics that story implies.
8. Report a sensitivity analysis to exclusion violations, and, if the paper
   also reports a no-instrument estimate, its sensitivity to unobserved
   confounding.

# Appendix {#sec-appendix}

## The LATE theorem and the kappa identities {#app-late}

By consistency $D = Z D(1) + (1 - Z) D(0)$ and $Y = D Y(1) + (1 - D) Y(0)$.
By independence, $E[Y \mid Z = 1] = E[Y(D(1))]$ and $E[Y \mid Z = 0] = E[Y(D(0))]$,
so the reduced form is $E[Y(D(1)) - Y(D(0))] = E[(D(1) - D(0))(Y(1) - Y(0))]$
after substituting $Y(d) = Y(0) + d(Y(1) - Y(0))$. Partition by type:
$D(1) - D(0)$ equals 1 for compliers, 0 for always- and never-takers, and
$-1$ for defiers, so the reduced form equals
$P(\text{complier})\,E[Y(1) - Y(0) \mid \text{complier}] - P(\text{defier})\,E[Y(1) - Y(0) \mid \text{defier}]$.
The first stage is $E[D(1) - D(0)] = P(\text{complier}) - P(\text{defier})$.
Monotonicity removes the defier terms and the ratio is the complier effect.

For the kappa weights, note that under monotonicity a unit with $D = 1,
Z = 0$ is an always-taker and a unit with $D = 0, Z = 1$ a never-taker, and
that by independence
$E[g(X) D(1 - Z) / (1 - \pi(X))] = E[g(X) D(0)] = E[g(X) \mathbb{1}\{\text{always-taker}\}]$
and $E[g(X)(1 - D)Z / \pi(X)] = E[g(X)(1 - D(1))] = E[g(X)\mathbb{1}\{\text{never-taker}\}]$.
Hence $E[\kappa g(X)] = E[g(X)] - E[g(X)\mathbb{1}\{\text{AT}\}] - E[g(X)\mathbb{1}\{\text{NT}\}] = E[g(X)\mathbb{1}\{\text{complier}\}]$,
and dividing by $E[\kappa] = P(\text{complier})$ gives the complier mean
(Abadie 2003, Theorem 3.1).

## The omitted-confounder bias bound {#app-ovb}

With $\tilde A$ normalized to unit variance, $\gamma = E[\tilde A \tilde D]$
and $\delta = E[\bar A \bar Y] / E[\bar A^2]$ where bars denote residuals
after projecting on $\tilde D$. The bias is $\phi = \delta\gamma / E[\tilde D^2]$.
Writing $R^2_{D \sim A} = \gamma^2 / E[\tilde D^2]$ and
$R^2_{Y \sim A \mid D} = \delta^2 E[\bar A^2] / E[\bar Y^2]$ with
$E[\bar A^2] = 1 - R^2_{D \sim A}$ and $E[\bar Y^2] = E[(\tilde Y - \beta\tilde D)^2]$,
substitute and simplify to obtain the formula of Section
\@ref(sec-sensitivity) (Chernozhukov et al. 2026, Appendix 12.A). The
robustness value solves $\phi(r, r) = |\hat\theta|$, a quadratic in $r$.

## Control functions: equivalence with 2SLS and the average structural function {#app-cf}

*Equivalence.* Let $\hat D = P_{[Z, X]} D$ and $\hat V = D - \hat D$. The
regression of $Y$ on $(D, X, \hat V)$ spans the same column space as the
regression on $(\hat D, X, \hat V)$, so the coefficient on $D$ equals the
coefficient on $\hat D$ in the latter. By Frisch-Waugh-Lovell that
coefficient is obtained by regressing $Y$ on the residual of $\hat D$ after
projecting on $(X, \hat V)$. Since $\hat D$ is orthogonal to $\hat V$ (a
first-stage fitted value is orthogonal to its residual) and $X$ is in the
first-stage regressors, the residual of $\hat D$ after $(X, \hat V)$ equals
the residual of $\hat D$ after $X$ alone, and the coefficient is that of
the regression of $Y$ on $\hat D$ and $X$: two-stage least squares. The
argument does not use the number of instruments.

*Average structural function.* If $U = \rho V + \varepsilon$ with
$\varepsilon \perp (D, Z, X, V)$ and $\varepsilon \sim N(0, \sigma^2)$, then
$P(Y = 1 \mid D, X, V) = \Phi((\alpha + \theta D + X'\beta + \rho V)/\sigma)$,
a probit in $(D, X, V)$ with coefficients scaled by $1/\sigma$. The
structural probability at $(d, x)$ is
$\mathrm{ASF}(d, x) = E_V[\Phi((\alpha + \theta d + x'\beta + \rho V)/\sigma)]$,
averaging over the marginal distribution of $V$; its derivative in $d$,
averaged over the distribution of $(D, X)$, is the average partial effect.
Replacing $V$ by $\hat V$ and the expectation by the sample average gives
`cf_ape()`. The generalized residual for a probit first stage is
$E[V \mid D, Z, X]$ under the threshold model $D = \mathbb{1}\{z'\gamma + V > 0\}$
with $V$ standard normal, which is $\lambda(z'\gamma)$ for $D = 1$ and
$-\lambda(-z'\gamma)$ for $D = 0$, with $\lambda = \phi / \Phi$ the inverse
Mills ratio.

## Orthogonality of the IV scores {#app-orthogonal-iv}

*Partially linear IV.* $\psi = (Y - \ell - \theta(D - r))(Z - m)$. The
derivative with respect to $\ell$ in direction $h$ is $-E[h(X)(Z - m_0(X))] = 0$
because $E[Z - m_0 \mid X] = 0$; the same for $r$. The derivative with
respect to $m$ is $-E[(Y - \ell_0 - \theta_0(D - r_0)) h(X)] = -E[E[U \mid X] h(X)] = 0$
where $U$ is the structural error with $E[U \mid X] = 0$. The estimator is
the ratio of sample means of $\psi_b = (Y - \ell)(Z - m)$ and
$\psi_a = (D - r)(Z - m)$, and the influence function is
$\psi / E[\psi_a]$; the variance is the IV sandwich on residuals
(Chernozhukov et al. 2026, Theorem 13.1.1).

*LATE score.* Each bracket is the AIPW score of an average treatment effect
of $Z$ with the instrument propensity, and Section 04 showed that score is
orthogonal to its outcome regressions and its propensity. The ratio
$\theta = E[\psi_b] / E[\psi_a]$ is a smooth function of two orthogonal
functionals, so it is orthogonal, with influence function
$(\psi_b - \theta \psi_a) / E[\psi_a]$ (Chernozhukov et al. 2026, Theorem
13.2.1).

## Anderson-Rubin with cross-fitted residuals {#app-ar}

At the true $\theta_0$ the moment $\bar\psi(\theta_0) = \mathbb{E}_n[(\check Y - \theta_0 \check D)\check Z]$
built from cross-fitted residuals equals its oracle version up to a term
whose leading part is a product of nuisance errors (orthogonality) plus a
term of order $n^{-1/2}$ times the nuisance error (cross-fitting removes
the dependence). Neither involves the first stage, so
$\sqrt n \bar\psi(\theta_0) / \sqrt{\widehat{\mathrm{Var}}} \to N(0, 1)$
whatever $E[\tilde D \tilde Z]$ is. Squaring gives the $\chi^2(1)$ limit;
with $k$ instruments the quadratic form in the inverse covariance gives
$\chi^2(k)$ (Chernozhukov et al. 2026, Theorem 13.3.1). Test inversion
over $\theta_0$ yields the set; for $k = 1$ the inequality
$n(\bar m_Y - \theta \bar m_D)^2 \le c\,(S_{YY} - 2\theta S_{YD} + \theta^2 S_{DD})$
is quadratic in $\theta$, which is why the set is an interval, its
complement, or the whole line.

## Marginal treatment effects and the policy weights {#app-mte}

Under $D = \mathbb{1}\{p(Z, X) \ge V\}$ with $V \sim U(0,1) \perp (Z, X)$ and
$(Y(0), Y(1)) \perp Z \mid (X, V)$,
$E[Y \mid X, p(Z, X) = p] = E[Y(0) \mid X] + \int_0^p E[Y(1) - Y(0) \mid X, V = u]\, du$,
so the derivative in $p$ is the MTE. For any estimand of the form
$E[Y(1) - Y(0)]$ over a population defined by $V$ and $p$, write it as
$\int_0^1 \mathrm{MTE}(u)\, \omega(u)\, du$: for the ATE $\omega = 1$; for the
ATT, $E[Y(1) - Y(0) \mid D = 1] = E[\int_0^p \mathrm{MTE}(u) du] / E[p]$, so
$\omega(u) = P(p > u) / E[p]$; for the ATU, $\omega(u) = P(p \le u) / E[1 - p]$;
for the effect of a policy that moves $p$ to $p'$, the change in
$E[Y] = E[\int_0^p \mathrm{MTE}]$ divided by the change in $E[D] = E[p]$ gives
$\omega(u) = (F_p(u) - F_{p'}(u)) / (E[p'] - E[p])$ (Heckman and Vytlacil
2005, Table 1). The separable estimator regresses $Y$ on $X$, $pX$, and
a polynomial in $p$; the coefficients on $pX$ are $\beta_1 - \beta_0$ and
the derivative of the polynomial is $K'(u)$.

## Weights of 2SLS and Rotemberg weights {#app-weights}

With residualized variables, $\hat D = \sum_k \hat\pi_k Z_k$ and
$\hat\theta_{2SLS} = \mathrm{Cov}(\hat D, Y) / \mathrm{Cov}(\hat D, D)$. Expanding
the numerator, $\sum_k \hat\pi_k \mathrm{Cov}(Z_k, Y) = \sum_k \hat\pi_k \mathrm{Cov}(Z_k, D)\,\hat\theta_k$,
and dividing by the denominator $\sum_j \hat\pi_j \mathrm{Cov}(Z_j, D)$
gives the weights of Section \@ref(sec-many-weights). For a shift-share
instrument $B = \sum_k g_k s_k$, the same expansion with $\hat\pi_k = g_k$
gives the Rotemberg weights $\alpha_k$. The shock-level equivalence of
Borusyak, Hull, and Jaravel (2022) follows from writing
$\sum_i B_i D_i^\perp = \sum_k g_k \sum_i s_{ik} D_i^\perp = \sum_k s_k g_k \bar D_k$
with $s_k = \sum_i s_{ik}$ and $\bar D_k$ the exposure-weighted mean, and
likewise for the numerator, so the ratio is the shock-level IV coefficient
weighted by $s_k$ without an intercept.

## References

- Abadie, Alberto. 2003. "Semiparametric Instrumental Variable Estimation of Treatment Response Models." *Journal of Econometrics*.
- Anderson, T. W., and Herman Rubin. 1949. "Estimation of the Parameters of a Single Equation in a Complete System of Stochastic Equations." *Annals of Mathematical Statistics*.
- Andrews, Isaiah, James H. Stock, and Liyang Sun. 2019. "Weak Instruments in Instrumental Variables Regression: Theory and Practice." *Annual Review of Economics*.
- Autor, David H., David Dorn, and Gordon H. Hanson. 2013. "The China Syndrome: Local Labor Market Effects of Import Competition in the United States." *American Economic Review*.
- Balke, Alexander, and Judea Pearl. 1997. "Bounds on Treatment Effects from Studies with Imperfect Compliance." *Journal of the American Statistical Association*.
- Bartik, Timothy J. 1991. *Who Benefits from State and Local Economic Development Policies?* Upjohn Institute.
- Bekker, Paul A. 1994. "Alternative Approximations to the Distributions of Instrumental Variable Estimators." *Econometrica*.
- Belloni, Alexandre, Daniel Chen, Victor Chernozhukov, and Christian Hansen. 2012. "Sparse Models and Methods for Optimal Instruments with an Application to Eminent Domain." *Econometrica*.
- Blundell, Richard, and James L. Powell. 2004. "Endogeneity in Semiparametric Binary Response Models." *Review of Economic Studies*.
- Borusyak, Kirill, Peter Hull, and Xavier Jaravel. 2022. "Quasi-Experimental Shift-Share Research Designs." *Review of Economic Studies*.
- Bound, John, David A. Jaeger, and Regina M. Baker. 1995. "Problems with Instrumental Variables Estimation When the Correlation between the Instruments and the Endogenous Explanatory Variable Is Weak." *Journal of the American Statistical Association*.
- Chernozhukov, Victor, Christian Hansen, Nathan Kallus, Martin Spindler, and Vasilis Syrgkanis. 2026. *Applied Causal Inference Powered by ML and AI*, chapters 12 and 13.
- Chernozhukov, Victor, Carlos Cinelli, Whitney Newey, Amit Sharma, and Vasilis Syrgkanis. 2022. "Long Story Short: Omitted Variable Bias in Causal Machine Learning." NBER Working Paper 30302.
- Cinelli, Carlos, and Chad Hazlett. 2020. "Making Sense of Sensitivity: Extending Omitted Variable Bias." *Journal of the Royal Statistical Society B*.
- Conley, Timothy G., Christian B. Hansen, and Peter E. Rossi. 2012. "Plausibly Exogenous." *Review of Economics and Statistics*.
- Cornelissen, Thomas, Christian Dustmann, Anna Raute, and Uta Schönberg. 2016. "From LATE to MTE: Alternative Methods for the Evaluation of Policy Interventions." *Labour Economics*.
- Frandsen, Brigham, Lars Lefgren, and Emily Leslie. 2023. "Judging Judge Fixed Effects." *American Economic Review*.
- Frölich, Markus. 2007. "Nonparametric IV Estimation of Local Average Treatment Effects with Covariates." *Journal of Econometrics*.
- Garen, John. 1984. "The Returns to Schooling: A Selectivity Bias Approach with a Continuous Choice Variable." *Econometrica*.
- Gilchrist, Duncan Sheppard, and Emily Glassberg Sands. 2016. "Something to Talk About: Social Spillovers in Movie Consumption." *Journal of Political Economy*.
- Goldsmith-Pinkham, Paul, Isaac Sorkin, and Henry Swift. 2020. "Bartik Instruments: What, When, Why, and How." *American Economic Review*.
- Griliches, Zvi. 1977. "Estimating the Returns to Schooling: Some Econometric Problems." *Econometrica*.
- Heckman, James J. 1979. "Sample Selection Bias as a Specification Error." *Econometrica*.
- Heckman, James J., and Richard Robb. 1985. "Alternative Methods for Evaluating the Impact of Interventions." In *Longitudinal Analysis of Labor Market Data*.
- Heckman, James J., and Edward Vytlacil. 2005. "Structural Equations, Treatment Effects, and Econometric Policy Evaluation." *Econometrica*.
- Imbens, Guido W., and Joshua D. Angrist. 1994. "Identification and Estimation of Local Average Treatment Effects." *Econometrica*.
- Kitagawa, Toru. 2015. "A Test for Instrument Validity." *Econometrica*.
- Lee, David S., Justin McCrary, Marcelo J. Moreira, and Jack Porter. 2022. "Valid t-Ratio Inference for IV." *American Economic Review*.
- Manski, Charles F. 1990. "Nonparametric Bounds on Treatment Effects." *American Economic Review*.
- Mogstad, Magne, Alexander Torgovitsky, and Christopher R. Walters. 2021. "The Causal Interpretation of Two-Stage Least Squares with Multiple Instrumental Variables." *American Economic Review*.
- Montiel Olea, José Luis, and Carolin Pflueger. 2013. "A Robust Test for Weak Instruments." *Journal of Business & Economic Statistics*.
- Mourifié, Ismael, and Yuanyuan Wan. 2017. "Testing Local Average Treatment Effect Assumptions." *Review of Economics and Statistics*.
- Oreopoulos, Philip. 2006. "Estimating Average and Local Average Treatment Effects of Education When Compulsory Schooling Laws Really Matter." *American Economic Review*.
- Petrin, Amil, and Kenneth Train. 2010. "A Control Function Approach to Endogeneity in Consumer Choice Models." *Journal of Marketing Research*.
- Rivers, Douglas, and Quang H. Vuong. 1988. "Limited Information Estimators and Exogeneity Tests for Simultaneous Probit Models." *Journal of Econometrics*.
- Smith, Richard J., and Richard W. Blundell. 1986. "An Exogeneity Test for a Simultaneous Equation Tobit Model with an Application to Labor Supply." *Econometrica*.
- Staiger, Douglas, and James H. Stock. 1997. "Instrumental Variables Regression with Weak Instruments." *Econometrica*.
- Vytlacil, Edward. 2002. "Independence, Monotonicity, and Latent Index Models: An Equivalence Result." *Econometrica*.
- Wooldridge, Jeffrey M. 2015. "Control Function Methods in Applied Econometrics." *Journal of Human Resources*.
- Wright, Philip G. 1928. *The Tariff on Animal and Vegetable Oils*. Macmillan.
- Young, Alwyn. 2022. "Consistency without Inference: Instrumental Variables in Practical Application." *European Economic Review*.
