---
title: "Causal Inference in Economics and Marketing"
subtitle: "Section 09: Synthetic Control"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.2, fig.align = "center", fig.pos = "H",
  out.width = "95%", collapse = TRUE, comment = "#>"
)
# Code and output in footnotesize.
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
# This lecture lives in inst/lectures/09_sc/ of the package source. When
# knitted from the source tree it loads the development version of the
# package; when the package is installed it uses the installed helpers.
helper_rel <- file.path("vignette_helpers", "09_sc", "sc_lecture_helpers.R")
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
  helper_path <- file.path(pkg_root, "inst", helper_rel)
} else {
  library(causalmetrics)
  helper_path <- system.file(helper_rel, package = "causalmetrics")
}
library(dplyr)
library(ggplot2)
library(kableExtra)
source(helper_path)
helper_md5 <- tools::md5sum(helper_path)
# Andersson (2019) data: git-ignored raw-data folder of the package source.
data_path <- file.path(pkg_root, "data_raw", "09_Synthetic_Control", "andersson2019carbon")
has_data <- file.exists(file.path(data_path, "carbontax_data.dta"))
n_sims <- 200
fmt <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
```

# The problem {#sec-problem}

## One treated unit, many donors, no parallel trends

Section 07 compared groups of units before and after a policy. Its
assumption was parallel trends: without the policy, the treated group would
have moved like the comparison group. Many policies do not come in groups.
A country adopts a carbon tax. A state passes a tobacco law. A region
suffers a terrorist conflict. One treated unit, observed over many periods,
and a pool of untreated units that may or may not resemble it.

Sweden is the running example of this lecture (Andersson 2019). In 1990
Sweden extended its value-added tax to transport fuel and in 1991 added a
carbon tax. The question is the effect on per-capita CO2 emissions from
transport. The data cover Sweden and 14 OECD countries from 1960 to 2005:
30 pre-treatment years and 16 post-treatment years.

```{r sweden-data, eval = has_data}
carbon <- as.data.frame(haven::read_dta(file.path(data_path, "carbontax_data.dta")))
carbon$treated <- as.integer(carbon$country == "Sweden" & carbon$year >= 1990)
donor_avg <- carbon %>% filter(country != "Sweden") %>% group_by(year) %>%
  summarise(y = mean(CO2_transport_capita), .groups = "drop") %>% mutate(series = "OECD donor average")
sweden <- carbon %>% filter(country == "Sweden") %>%
  transmute(year, y = CO2_transport_capita, series = "Sweden")
```

```{r fig-sweden-raw, eval = has_data, fig.height = 3.6, fig.cap = "Per-capita CO2 emissions from transport, Sweden and the average of 14 OECD donor countries. The series diverge well before 1990."}
ggplot(bind_rows(sweden, donor_avg), aes(x = year, y = y, colour = series, linetype = series)) +
  geom_line() + geom_vline(xintercept = 1989.5, linetype = "dotted") +
  scale_colour_manual(values = c("Sweden" = "black", "OECD donor average" = "grey45")) +
  labs(x = NULL, y = "Metric tons per capita", colour = NULL, linetype = NULL) + sc_theme()
```

DiD would compare Sweden with the donor average. Figure
\@ref(fig:fig-sweden-raw) shows why that fails. Sweden's emissions rise
faster than the average through the 1960s and fall below it in the 1980s.
Nothing in the pre-treatment data supports parallel trends. With one
treated unit there is also no group of treated units to average over, and
no cluster to compute a standard error from.

Synthetic control (Abadie and Gardeazabal 2003; Abadie, Diamond, and
Hainmueller 2010) replaces the donor average by a weighted average.
The weights are nonnegative, sum to one, and are chosen so that the
weighted donors reproduce Sweden's pre-treatment path. The same weights
then give the counterfactual after 1990. Two things change relative to
DiD.

- The comparison unit is chosen by the data. Parallel trends is no longer
  assumed; it is built in the pre-treatment period and checked by eye.
- The model behind it allows unobserved common shocks with unit-specific
  loadings, not only additive unit and time effects. A donor mix that
  matches Sweden's pre-treatment path also matches its exposure to those
  shocks, if the fit is good enough.

```{r sweden-sc, eval = has_data}
sc_sweden <- synth_control(carbon, id = "country", time = "year", y = "CO2_transport_capita", d = "treated")
did_sweden <- sdid_weights(carbon, id = "country", time = "year", y = "CO2_transport_capita", d = "treated",
                           estimator = "did")
sc_sweden
c(did = did_sweden$estimate, synthetic_control = sc_sweden$estimate)
```

```{r fig-sweden-sc, eval = has_data, fig.height = 3.6, fig.cap = "Sweden and its synthetic control built from all pre-treatment outcomes. The gap after 1990 is the estimated effect."}
plot_synth(sc_sweden, type = "path")
```

The synthetic Sweden tracks the real one closely before 1990 and diverges
after. The average post-treatment gap is
`r if (has_data) fmt(sc_sweden$estimate) else "about -0.28"` tons per
capita, a reduction of roughly ten percent. DiD gives a different number,
and Figure \@ref(fig:fig-sweden-raw) says which one to trust.

Three questions follow and organize the lecture. When do such weights
recover the counterfactual, and when do they fail? How is uncertainty
assessed with one treated unit? And how is an application built and
defended in practice?

## One experiment, six estimators

The natural model for this setting is a linear factor model (Section
\@ref(sec-framework)):
$$
Y_{jt}^N = c_j + \delta_t + \lambda_t' \mu_j + \varepsilon_{jt}.
$$
Units differ in their fixed effects $c_j$ and in how they load ($\mu_j$)
on common factors $\lambda_t$. DiD assumes the loadings are equal. Synthetic
control does not. The experiment below simulates this model with
`sim_synth_panel()`: 20 donors, 20 pre-treatment periods, 5
post-treatment periods, two factors, a true effect of one. Two designs
follow Ferman and Pinto (2021, Table 1).

- **Design A, level in the tail.** The treated unit's fixed effect sits two
  standard deviations above the donors'. Its loadings are ordinary. The
  factors have mean zero after treatment, so the DiD assumption holds.
- **Design B, loading in the tail with a break.** The treated unit loads
  1.5 standard deviations above the donors on the first factor, and that
  factor shifts up by one standard deviation after treatment. Treatment
  is correlated with a time-varying unobservable.

| Estimator | Weights | Intercept | Post-treatment correction |
|---|---|---|---|
| DiD | uniform | unit and period effects | none |
| Synthetic control | simplex, matched on outcome levels | none | none |
| Demeaned synthetic control | simplex, matched on outcome deviations | yes | none |
| Augmented synthetic control | as demeaned | yes | ridge regression on the predictors |
| Synthetic DiD | simplex unit weights and period weights | unit and period effects | none |
| Unconstrained regression | least squares of the treated unit on the donors | yes | none |

```{r sc-experiment, cache = TRUE, cache.extra = helper_md5}
results <- run_sc_experiment(n_sims = n_sims, seed = 1)
tab <- summarise_sc_experiment(results, effect = 1)
pick <- function(design_prefix, estimator, column) {
  tab[[column]][startsWith(tab$design, design_prefix) & tab$estimator == estimator]
}
```

```{r tab-experiment, echo = FALSE}
tab %>%
  mutate(design = sub(":.*", "", design)) %>%
  select(design, estimator, bias, sd, rmse, pre_rmspe, l2) %>%
  knitr::kable(digits = 3, booktabs = TRUE,
               col.names = c("Design", "Estimator", "Bias", "SD", "RMSE", "Pre-RMSPE", "Concentration"),
               caption = paste0("Six estimators on the linear factor model, true effect 1, ", n_sims,
                                " simulations per design. Pre-RMSPE is the pre-treatment root mean squared prediction error; concentration is the squared norm of the donor weights."))
```

```{r fig-experiment, fig.height = 4.6, fig.cap = "Sampling distributions of the six estimators minus the true effect, by design, truncated at plus or minus 4. Most draws of the unconstrained regression fall outside that range."}
plot_sc_experiment(results, effect = 1)
```

## Reading the results

- **DiD** is unbiased in Design A (bias `r fmt(pick("A", "DiD", "bias"))`)
  and badly biased in Design B (`r fmt(pick("B", "DiD", "bias"))`). The
  donor average loads less on the shifted factor than the treated unit,
  and the shift lands on the estimate. This is the failure that motivates
  the lecture: a comparison group with the wrong exposure to common shocks.
- **Synthetic control** is biased in Design A (`r fmt(pick("A", "Synthetic control", "bias"))`).
  Weights on the simplex can only reach levels inside the donors' range;
  a treated unit above that range cannot be matched, and the level gap
  becomes a bias. Its pre-treatment fit is visibly worse
  (pre-RMSPE `r fmt(pick("A", "Synthetic control", "pre_rmspe"))`). In
  Design B it reduces the bias of DiD to
  `r fmt(pick("B", "Synthetic control", "bias"))` but does not remove it.
- **Demeaned synthetic control** matches deviations from each unit's own
  mean. It has no level problem in Design A (bias
  `r fmt(pick("A", "Demeaned synthetic control", "bias"))`), and a smaller
  standard deviation than DiD (`r fmt(pick("A", "Demeaned synthetic control", "sd"))`
  against `r fmt(pick("A", "DiD", "sd"))`). In Design B it is biased,
  like every estimator, because noise keeps the weights from fully
  matching the loadings; the bias (`r fmt(pick("B", "Demeaned synthetic control", "bias"))`)
  is a fraction of DiD's.
- **Augmented synthetic control** adds a regression correction for the
  remaining imbalance in the predictors. It trades bias for variance
  (Design B bias `r fmt(pick("B", "Augmented synthetic control", "bias"))`,
  SD `r fmt(pick("B", "Augmented synthetic control", "sd"))`). Its
  pre-treatment fit is better by construction, because the correction is
  fitted on those periods; that number is not a diagnostic.
- **Synthetic DiD** combines unit weights, period weights, and two-way
  effects (Section 07). It performs like the demeaned synthetic control
  in Design A and somewhat better in Design B.
- **Unconstrained regression** of the treated unit on all donors fits the
  pre-treatment period perfectly, because there are as many donors as
  pre-treatment periods, and then extrapolates noise. Its RMSE is the
  largest in both designs. A perfect pre-treatment fit is not evidence of
  a good counterfactual when the donors outnumber the periods.

Three lessons carry through the lecture. Matching levels and matching
trends are different things, and the intercept decides which one is
asked. A good pre-treatment fit is necessary but not sufficient: it can
come from overfitting, and it does not protect against a shock that
treated and donors feel differently. And weight concentration is a
diagnostic: among the simplex estimators, the ones that fail most spread
their weights least.

## When this is the right tool

Synthetic control is the design for a treated aggregate unit with a long
pre-treatment history and a pool of untreated units that are exposed to
the same common shocks. Its requirements, in the order they are usually
violated:

- **Unaffected donors.** No spillover from the treated unit to the donors,
  and no donor with its own policy change in the window.
- **No anticipation.** The pre-treatment path must be a path without
  treatment; date the treatment at its announcement if units respond
  early.
- **A long pre-treatment period relative to the donor pool.** With few
  periods and many donors, the fit is achieved by chance.
- **A treated unit inside the donors' range.** Weights on the simplex
  interpolate; a treated unit at the extreme of the distribution cannot
  be reproduced without extrapolation.
- **Stable factor structure.** The relation between the treated unit and
  its synthetic control, learned before treatment, must persist after it.

When these fail, the alternatives are DiD with a justified comparison
group, synthetic DiD, or a design that does not rely on the donors.

**Roadmap.** Section \@ref(sec-framework) states the factor model, shows
why the weights work under a perfect fit, and what breaks under an
imperfect fit. Section \@ref(sec-build) builds the synthetic control:
predictors, the importance weights, constraints, and augmentation.
Section \@ref(sec-inference) covers inference with one treated unit:
placebos, conformal tests, and the specification test. Section
\@ref(sec-robustness) collects the robustness checks and the checklist of
Ferman and Pinto. Section \@ref(sec-practice) turns to the applications and
Section \@ref(sec-extensions) to the recent literature. Proofs are
sketched in the text and completed in the Appendix.

# Just enough framework {#sec-framework}

## Setup and estimand {#sec-setup}

Units $j = 0, 1, \dots, J$ are observed in periods $t = 1, \dots, T$.
Unit $0$ is treated from period $T_0 + 1$ on; units $1, \dots, J$ are the
donors. Potential outcomes are $Y_{jt}^N$ without treatment and
$Y_{jt}^I = Y_{jt}^N + \alpha_{jt}$ with it. The observed outcome is
$Y_{jt} = Y_{jt}^N + D_{jt}\,\alpha_{jt}$ with $D_{jt} = 1\{j = 0,\ t > T_0\}$.

The estimand is the effect on the treated unit in each post-treatment
period, $\alpha_{0t} = Y_{0t} - Y_{0t}^N$ for $t > T_0$, and its average
over the post-treatment periods. It is not an average over units. Only
$Y_{0t}^N$ is missing.

Two assumptions come before any model.

- **No anticipation.** $Y_{0t} = Y_{0t}^N$ for $t \le T_0$. The
  pre-treatment path is a path without treatment.
- **No interference.** $Y_{jt} = Y_{jt}^N$ for every donor and every $t$.
  Donors are unaffected by the treated unit's policy.

Both are restrictions on the donor pool and the treatment date, not on
the estimator.

> **Definition (synthetic control).** For weights $w$ in the simplex
> $\Delta = \{w \in \mathbb{R}^J : w_j \ge 0,\ \sum_j w_j = 1\}$,
> $$
> \hat Y_{0t}^N(w) = \sum_{j = 1}^J w_j Y_{jt},
> \qquad
> \hat\alpha_{0t}(w) = Y_{0t} - \sum_{j = 1}^J w_j Y_{jt}.
> $$
> The synthetic control weights minimize the pre-treatment discrepancy,
> $$
> \hat w = \arg\min_{w \in \Delta} \sum_{t \le T_0} \Bigl(Y_{0t} - \sum_j w_j Y_{jt}\Bigr)^2
> $$
> when every pre-treatment outcome is a predictor, or
> $\|Z_0 - Z w\|_V^2 = (Z_0 - Zw)' V (Z_0 - Zw)$ for a vector of predictors
> $Z_0$ (outcome lags and covariates) and importance weights $V$.

## The linear factor model {#sec-factor-model}

The theory of the estimator is stated for
$$
Y_{jt}^N = c_j + \delta_t + \lambda_t' \mu_j + \varepsilon_{jt},
$$
with unit effects $c_j$, period effects $\delta_t$, $F$ common factors
$\lambda_t$ with unit-specific loadings $\mu_j$, and idiosyncratic shocks
$\varepsilon_{jt}$ with $\mathbb{E}[\varepsilon_{jt}] = 0$. Two-way fixed
effects is the case $\mu_j \equiv \mu$: the factors then act like period
effects. The factor model lets a common shock hit units with different
intensity, which is the case parallel trends rules out.

Treatment assignment is treated as fixed, together with $(c_j, \mu_j)$.
The identifying condition (Ferman and Pinto 2021, Assumption 3) is that
the idiosyncratic shocks have mean zero given the assignment. Treatment
may depend on the treated unit's level $c_0$ and loadings $\mu_0$; it may
not depend on shocks specific to the treated unit. This is what "selection
on unobservables" means here: unobserved confounders are allowed as long
as they also move the donors.

| | DiD | Synthetic control |
|---|---|---|
| Counterfactual | donor average, shifted by the pre-treatment gap between the treated unit and the average | weighted donors, weights fitted before treatment |
| Absorbs | $c_j$ and $\delta_t$ | $\delta_t$ (weights sum to one), plus $c_0$ and $\lambda_t'\mu_0$ if the weights reproduce them |
| Needs | $\mu_0 = \bar\mu$, or $\mathbb{E}[\lambda_t] = 0$ after treatment | weights $w$ with $c'w = c_0$ and $\mu' w = \mu_0$, and a fit good enough to find them |

## Why the weights work: skeleton of the argument {#sec-skeleton}

Substitute the model into the estimator. For any $w \in \Delta$ and any
$t$,
$$
\hat\alpha_{0t}(w) - \alpha_{0t}
= \underbrace{(c_0 - c'w)}_{\text{level mismatch}}
+ \underbrace{\lambda_t'(\mu_0 - \mu' w)}_{\text{factor mismatch}}
+ \underbrace{(\varepsilon_{0t} - \varepsilon_t' w)}_{\text{noise}},
$$
where $c = (c_1, \dots, c_J)'$, $\mu = (\mu_1, \dots, \mu_J)'$, and
$\varepsilon_t = (\varepsilon_{1t}, \dots, \varepsilon_{Jt})'$. The period
effect $\delta_t$ cancels because the weights sum to one. The noise term
has mean zero. Everything depends on whether the fitted weights make the
first two terms small.

**Step 1: perfect fit (Abadie, Diamond, and Hainmueller 2010).** Suppose
weights $w^*$ exist with $c'w^* = c_0$ and $\mu'w^* = \mu_0$, and suppose
the fitted weights reproduce the treated unit exactly over the
pre-treatment period. Then the pre-treatment gap is the noise term
evaluated at $\hat w$, and a mismatch in the loadings would have to be
offset by the noise in every one of $T_0$ periods. The bias of
$\hat\alpha_{0t}$ is bounded by a term that shrinks as $T_0$ grows, with a
rate governed by the noise variance and the number of donors. Appendix
\@ref(app-perfect-fit) states the bound. The lesson is practical: a long
pre-treatment period with a close fit is the case the method was
designed for.

**Step 2: imperfect fit (Ferman and Pinto 2021).** Drop the perfect-fit
condition. With every pre-treatment outcome as a predictor, the objective
divided by $T_0$ converges, as $T_0 \to \infty$ with $J$ fixed, to
$$
Q_0(w) = (c_0 - c'w)^2 + (\mu_0 - \mu'w)'\,\Omega_0\,(\mu_0 - \mu'w) + \sigma^2\,(1 + w'w),
$$
where $\Omega_0$ is the long-run second moment of the factors and
$\sigma^2$ the noise variance (Appendix \@ref(app-imperfect-fit)). The
fitted weights converge to $\bar w = \arg\min_\Delta Q_0$. The last term
is the variance of the weighted noise, and it rewards spreading the
weights. Unless $\sigma^2 = 0$, the minimizer does not set the first two
terms to zero, even when weights that would do so exist. So
$$
\hat\alpha_{0t} \;\to_p\; \alpha_{0t} + (c_0 - c'\bar w) + \lambda_t'(\mu_0 - \mu'\bar w) + (\varepsilon_{0t} - \varepsilon_t'\bar w).
$$

Four consequences, each visible in Table \@ref(tab:tab-experiment).

- **Level bias.** If $\bar w$ misses $c_0$, the estimator is biased even
  when the factors have mean zero after treatment. This is Design A, and
  it is a bias DiD does not have.
- **The demeaned fix.** Subtracting each unit's pre-treatment mean removes
  the level term from $Q_0$. The demeaned synthetic control is then
  unbiased under exactly the DiD condition, $\mathbb{E}[\lambda_t] = 0$
  for $t > T_0$, and has weakly smaller asymptotic variance than DiD when
  the factor and noise moments are stable (Ferman and Pinto, Propositions
  2 and 3). The intercept costs nothing in identification and buys
  efficiency.
- **Time-varying confounders.** If $\mathbb{E}[\lambda_t] \ne 0$ after
  treatment, both are biased: DiD by $\bar\lambda'(\mu_0 - \bar\mu)$ and the
  demeaned synthetic control by $\bar\lambda'(\mu_0 - \mu'\bar w)$. The
  second is usually smaller because $\bar w$ is pulled toward matching the
  loadings, but it is not zero. This is Design B.
- **Concentration as a warning.** The trade-off in $Q_0$ is between
  matching and spreading. Weights that are concentrated on few donors
  signal that matching was hard, and the simulations of Ferman and Pinto
  show the bias rising with $\|w\|^2$.

The following check makes Step 2 concrete. The treated unit's level and
loadings are an exact convex combination of three donors' (weights 0.5,
0.3, 0.2), so $w^*$ exists, and every unit has noise of standard deviation
one. The table reports the distance between the fitted weights and
$w^*$, the loading mismatch $\|\mu_0 - \mu'\hat w\|$, and the
concentration $\|\hat w\|^2$ as $T_0$ grows.

```{r weights-limit}
w_true <- c(0.5, 0.3, 0.2, rep(0, 5))
limit_check <- function(t_pre) {
  dat <- simulate_convex_treated(n_donors = 8, t_pre = t_pre, t_post = 5, sd_e = 1, w_true = w_true, seed = 1)
  fit <- synth_control(dat, id = "id", time = "time", y = "y", d = "d")
  mu <- attr(dat, "donor_loadings")
  c(t_pre = t_pre, distance_to_w_star = sqrt(sum((fit$weights$weight - w_true)^2)),
    loading_mismatch = sqrt(sum((crossprod(mu, fit$weights$weight - w_true))^2)),
    concentration = fit$concentration[["l2"]])
}
round(rbind(limit_check(50), limit_check(500), limit_check(5000)), 3)
```

The distance to $w^*$ does not vanish with $T_0$. The weights settle at a
more diluted vector than $w^*$ (whose concentration is
`r fmt(sum(w_true^2))`), and the loading mismatch that comes with it is
permanent. That mismatch is harmless while the factors average zero, and
becomes a bias when they do not.

## Weights plus an outcome model {#sec-weights-model}

Every estimator of Section \@ref(sec-problem) has the form
$$
\hat Y_{0t}^N = \hat m_t(Z_0) + \sum_j w_j \bigl(Y_{jt} - \hat m_t(Z_j)\bigr),
$$
an outcome model evaluated at the treated unit plus a weighted average of
the donors' residuals from that model (Ben-Michael, Feller, and Rothstein
2021; Doudchenko and Imbens 2016).

| Estimator | Outcome model $\hat m_t$ | Weights |
|---|---|---|
| DiD | unit and period effects | uniform |
| Synthetic control | none | simplex, on levels |
| Demeaned synthetic control | unit effect | simplex, on deviations |
| Augmented synthetic control | ridge regression of $Y_{jt}$ on the predictors | simplex, on deviations |
| Synthetic DiD | unit and period effects | simplex units, simplex periods |
| Unconstrained regression | none | least squares |

The error of the general form is
$$
\hat Y_{0t}^N - Y_{0t}^N
= \bigl(\hat m_t(Z_0) - m_t(Z_0)\bigr) - \sum_j w_j \bigl(\hat m_t(Z_j) - m_t(Z_j)\bigr)
+ \Bigl(\sum_j w_j\, m_t(Z_j) - m_t(Z_0)\Bigr) + \text{noise},
$$
with $m_t$ the true conditional mean. If the weights balance the
predictors exactly and $m_t$ is linear in them, the third term is zero
whatever the model. If the model is right, the first two terms are zero
whatever the weights. Otherwise the leading error is the product of the
model error and the imbalance. This is the structure of Section 04:
weights play the role of the propensity score, the outcome model plays
the role of the regression, and the remainder is a product. The augmented
synthetic control is the doubly robust member of the family, which is
why it removed most of the bias in Design B. The price is variance, and
it is paid through extrapolation: the outcome model predicts outside the
donors' range.

# Building the synthetic control {#sec-build}

Each subsection follows the same order: the choice, what it does to the
weights, the code on Sweden, and when it breaks.

## Predictors and importance weights {#sec-predictors}

**The choice.** Two conventions define the predictors $Z$. The first uses
every pre-treatment outcome, so the weights reproduce the whole path.
The second uses a short list of covariates averaged over a window, plus a
few outcome lags, each weighted by an importance weight in $V$. Andersson
(2019) uses GDP per capita, vehicles per capita, gasoline consumption per
capita, and urban population, averaged over 1980 to 1989, plus emissions
in 1989, 1980, and 1970.

**What it does.** With covariates, $V$ decides which predictors the weights
must match. `Synth` chooses $V$ by a nested problem: for each $V$, solve
the inner simplex problem for $w(V)$; then choose $V$ to minimize the
pre-treatment mean squared prediction error of the outcome path. Kaul et
al. (2022) show that when every outcome lag is a predictor, the outer
problem gives the covariates zero weight: the lags alone determine $w$.
Covariates matter only when the lags are few.

```{r sweden-predictors, eval = has_data}
xs <- c("GDP_per_capita", "vehicles_capita", "gas_cons_capita", "urban_pop")
set.seed(1)
sc_and <- synth_control(carbon, id = "country", time = "year", y = "CO2_transport_capita", d = "treated",
                        x = xs, lags = c(1989, 1980, 1970), pre_window = 1980:1989, v = "mspe",
                        v_control = list(maxit = 2000, starts = 3))
sc_and_eq <- synth_control(carbon, id = "country", time = "year", y = "CO2_transport_capita", d = "treated",
                           x = xs, lags = c(1989, 1980, 1970), pre_window = 1980:1989, v = "equal")
weights_tab <- Reduce(function(a, b) merge(a, b, by = "country", all = TRUE), list(
  setNames(sc_sweden$weights, c("country", "all_lags")),
  setNames(sc_and_eq$weights, c("country", "andersson_equal_v")),
  setNames(sc_and$weights, c("country", "andersson_mspe_v"))
))
weights_tab <- weights_tab[order(-weights_tab$andersson_mspe_v), ]
knitr::kable(weights_tab[weights_tab$andersson_mspe_v > 0.001 | weights_tab$all_lags > 0.001, ], digits = 3,
             row.names = FALSE, booktabs = TRUE,
             caption = "Donor weights for Sweden under three predictor choices. Andersson (2019, Table 2) reports Denmark 0.384, Belgium 0.195, New Zealand 0.177, Greece 0.090, United States 0.088, Switzerland 0.061.")
```

```{r sweden-balance, eval = has_data}
knitr::kable(sc_and$balance[, c("predictor", "treated", "synthetic", "donor_mean", "v")], digits = 3,
             row.names = FALSE, booktabs = TRUE,
             col.names = c("Predictor", "Sweden", "Synthetic Sweden", "Donor mean", "V"),
             caption = "Predictor balance for Andersson's specification (his Table 1). The donor mean is unweighted.")
```

Three readings. The all-lags weights and Andersson's weights agree on the
main donors (Denmark, Belgium, New Zealand, the United States) and differ
in the tail. The $V$ weights of the nested problem concentrate on two
predictors; the pre-treatment MSPE is flat in $V$ near its minimum, so
different optimizers land on different $V$ with almost the same $w$, and
`Synth`'s $V$ is not the one above. $V$ is a means to an end: report it,
but judge the fit and the weights. The balance table is the object to
read. Sweden and synthetic Sweden agree on every predictor except gasoline
consumption, exactly as in the paper.

**When it breaks.** Few lags and many covariates invite a fit driven by
the covariates rather than the path. Many lags with a short pre-period
invite overfitting (Section \@ref(sec-fit)). Predictors measured after
treatment are never allowed.

## Constraints and extrapolation {#sec-constraints}

**The choice.** Weights on the simplex interpolate: the synthetic unit
lies inside the convex hull of the donors, predictor by predictor. Adding
an intercept (`demean = TRUE`) lets the level differ while the trends
match. Dropping the sum-to-one constraint or the sign constraint
(Doudchenko and Imbens 2016; Hsiao, Ching, and Wan 2012) allows
extrapolation in every direction.

```{r sweden-constraints, eval = has_data}
fits <- list(
  simplex = sc_sweden,
  demeaned = synth_control(carbon, "country", "year", "CO2_transport_capita", "treated", demean = TRUE),
  nonnegative = synth_control(carbon, "country", "year", "CO2_transport_capita", "treated", constraints = "nonnegative"),
  unconstrained = synth_control(carbon, "country", "year", "CO2_transport_capita", "treated", constraints = "none")
)
t(sapply(fits, function(f) c(estimate = f$estimate, pre_rmspe = f$pre_rmspe, sum_w = sum(f$weights$weight),
                              positive_weights = sum(f$weights$weight > 0.001), concentration = f$concentration[["l2"]])))
```

**What it does.** The simplex gives a sparse, interpretable set of donors.
The intercept changes the estimate little here because Sweden's level is
inside the donors' range; it would matter for a country at the top or
bottom of the distribution (Design A). The unconstrained regression has
30 pre-treatment periods and 14 donors, so it does not interpolate the
data, but its weights are dense and of both signs, and its counterfactual
is an extrapolation no one can defend donor by donor.

The range check in the balance table is the first guard against
extrapolation: every predictor of the treated unit should lie within the
donors' range.

```{r sweden-range, eval = has_data}
sc_and$balance[, c("predictor", "treated", "in_range")]
```

**When it breaks.** A treated unit at the extreme of the predictor
distribution cannot be reproduced on the simplex, and the fit will show
it. The honest responses are the intercept (which extrapolates the level
only) or a different design; unconstrained weights hide the problem
rather than solve it.

## Pre-treatment fit and overfitting {#sec-fit}

**The choice.** Every synthetic control is judged first by its
pre-treatment fit. The root mean squared prediction error over the
pre-treatment periods (`pre_rmspe`) is the number; the path plot is the
picture. Ferman and Pinto (2021) recommend a second picture: subtract the
period mean of all units from every series and look at the fit again.
Common trends that all units share are removed, and what remains is the
part of the fit that the weights had to earn.

```{r fig-sweden-demeaned, eval = has_data, fig.height = 3.4, fig.cap = "Sweden and its synthetic control after subtracting the period mean of all 15 countries. The fit before 1990 is what the weights achieved beyond the common trend."}
plot_synth(sc_sweden, type = "demeaned_path")
```

**Overfitting.** A close fit is easy to obtain when the donors outnumber
the pre-treatment periods, because the weights can absorb noise. The
experiment below draws noisy panels with no treatment effect, 10
pre-treatment periods, and 5 to 80 donors. The pre-treatment fit improves
with the donor pool while the post-treatment error does not.

```{r overfit-experiment, cache = TRUE, cache.extra = helper_md5}
overfit <- run_overfit_experiment(n_sims = 100, n_donors = c(5, 10, 20, 40, 80), t_pre = 10, sd_e = 2, seed = 2)
overfit_tab <- summarise_overfit(overfit)
```

```{r tab-overfit, echo = FALSE}
knitr::kable(overfit_tab, digits = 3, booktabs = TRUE,
             col.names = c("Donors", "Pre-RMSPE", "Post-RMSPE", "Bias", "SD", "RMSE", "Concentration"),
             caption = "Demeaned synthetic control with no treatment effect, 10 pre-treatment periods, and noise standard deviation 2: fit and post-treatment error by donor-pool size, 100 simulations each.")
```

With 80 donors the pre-treatment fit is `r fmt(overfit_tab$pre_rmspe[overfit_tab$n_donors == 80] / overfit_tab$pre_rmspe[overfit_tab$n_donors == 5], 2)`
times that with 5 donors, and the post-treatment error is
`r fmt(overfit_tab$post_rmspe[overfit_tab$n_donors == 80] / overfit_tab$post_rmspe[overfit_tab$n_donors == 5], 2)`
times. The improvement in the fit is fitted noise. Read the
pre-treatment fit together with the ratio of donors to periods, and
prefer the post/pre comparison of the placebo test (Section
\@ref(sec-placebo-space)) to the pre-treatment fit alone.

**When it breaks.** Ferman and Pinto's Remark 3: a good fit with $J$
large relative to $T_0$ is not evidence that the loadings are matched.
Sweden has 14 donors and 30 pre-treatment years, which is comfortable.

## Regularization and augmentation {#sec-augment}

**The choice.** Two additions borrow from Section 07 and Section 04. A
ridge penalty on the weights (`zeta`, as in synthetic DiD) spreads them
across donors at the cost of fit. The ridge-augmented estimator
(`augment = "ridge"`) keeps the simplex weights and adds the correction
$\hat m_t(Z_0) - \sum_j w_j \hat m_t(Z_j)$ of Section \@ref(sec-weights-model),
with $\hat m_t$ a ridge regression of each period's outcome on the
predictors, fitted on donors, and its penalty chosen by leave-one-donor-out
cross-validation.

```{r sweden-augment, eval = has_data}
sc_dm <- fits$demeaned
aug_row <- function(label, ...) {
  base <- synth_control(carbon, "country", "year", "CO2_transport_capita", "treated", demean = TRUE, ...)
  aug <- synth_control(carbon, "country", "year", "CO2_transport_capita", "treated", demean = TRUE, augment = "ridge", ...)
  corr <- aug$augmentation$correction[aug$effects$post]
  data.frame(predictors = label, demeaned_sc = base$estimate, augmented = aug$estimate, lambda = aug$lambda,
             mean_correction = mean(corr), correction_share = mean(corr) / aug$estimate)
}
aug_tab <- rbind(
  aug_row("all 30 lags"),
  aug_row("Andersson's 7 predictors", x = xs, lags = c(1989, 1980, 1970), pre_window = 1980:1989),
  aug_row("all 30 lags, lambda = 1", lambda = 1)
)
knitr::kable(aug_tab, digits = 3, row.names = FALSE, booktabs = TRUE,
             col.names = c("Predictors", "Demeaned SC", "Augmented", "Lambda", "Mean correction", "Share of estimate"),
             caption = "Ridge augmentation on Sweden under three predictor sets. The correction is averaged over the post-treatment periods.")
```

**What it does.** The correction is the augmented estimator's own
diagnostic. With all 30 lags as predictors and 14 donors, the ridge
model has more parameters than observations; cross-validation keeps the
penalty near zero and the correction is
`r if (has_data) fmt(100 * aug_tab$correction_share[1], 0) else "about 70"` percent of the estimate. With Andersson's
seven predictors it is a fifth of the estimate, with the opposite sign.
Neither is small. A correction comparable to the estimate says that the
answer comes from the outcome model, not from the donors. In that case
report the synthetic control estimate as the headline and the augmented
one as a sensitivity check, or fix a larger penalty, which shrinks the
augmented estimator back to the synthetic control (last row). The
augmented estimator earns its keep when the pre-treatment fit is poor and
the predictor set is small relative to the donor pool.

**When it breaks.** The ridge model is linear in the predictors. When the
post-treatment outcomes depend on the pre-treatment path in a different
way for the treated unit than for the donors, the correction adds bias
rather than removing it. The Monte Carlo of Section \@ref(sec-problem)
showed the other cost: variance.

# Inference {#sec-inference}

## What is random {#sec-random}

There is one treated unit. The estimate is one gap per post-treatment
period, and nothing is averaged across treated units. Two things are
uncertain: the weights, which were estimated from noisy pre-treatment
data, and the idiosyncratic shocks after treatment, which the weighted
donors do not reproduce. A standard error in the usual sense needs a
population that the treated unit was sampled from, and there is none.

The literature answers with Fisher's logic. Under a sharp null of no
effect, the treated unit is exchangeable with the donors, or the
post-treatment periods are exchangeable with the pre-treatment ones, and
either exchangeability gives a reference distribution. Three tools follow.

- **In-space placebos** permute the treatment across units (Abadie,
  Diamond, and Hainmueller 2010).
- **In-time placebos and conformal inference** permute across periods
  (Abadie, Diamond, and Hainmueller 2015; Chernozhukov, Wuthrich, and Zhu
  2021).
- **The specification test** of Ferman and Pinto (2021) permutes across
  periods to compare two estimators rather than to test an effect.

## In-space placebos {#sec-placebo-space}

**Procedure.** Reassign the treatment to each donor in turn, fit its
synthetic control from the remaining donors with the same predictors, and
record its gap path. The treated unit's gap is compared with the
distribution of placebo gaps. The statistic that adjusts for fit is the
ratio of post-treatment to pre-treatment mean squared prediction error:
a large post-treatment gap counts for more when the pre-treatment fit was
tight. The p-value is the rank of the treated unit's ratio among all
units, divided by their number. Placebos that could not be fitted are
usually dropped: Andersson keeps donors whose pre-treatment MSPE is at
most 20 times Sweden's.

```{r sweden-placebo, eval = has_data}
pl_cut <- synth_placebo(sc_and, mspe_limit = 20)
pl_cut
head(pl_cut$mspe[order(pl_cut$mspe$rank), c("country", "pre_mspe", "post_mspe", "ratio", "rank")], 5)
pl_lags <- synth_placebo(sc_sweden, mspe_limit = 20)
c(rank_andersson_predictors = pl_cut$mspe$rank[1], rank_all_lags = pl_lags$mspe$rank[1],
  p_andersson_predictors = pl_cut$p_value, p_all_lags = pl_lags$p_value)
```

```{r fig-sweden-placebo, eval = has_data, fig.height = 3.6, fig.cap = "Gaps for Sweden (black) and for each donor treated as a placebo (grey). Sweden's post-1990 gap is the largest and most persistent."}
plot_synth(pl_cut)
```

```{r fig-sweden-ratio, eval = has_data, fig.height = 3.2, fig.cap = "Ratio of post-treatment to pre-treatment mean squared prediction error, Sweden against the placebos kept after the 20-times cut."}
plot_synth(pl_cut, type = "ratio")
```

**Reading it.** With Andersson's predictors, Sweden's ratio ranks first
among the `r if (has_data) sum(pl_cut$kept) else "ten"` units kept after
the cut, so the p-value is `r if (has_data) fmt(pl_cut$p_value, 2) else "0.10"`,
the paper's number. This is the smallest p-value the design can produce:
with $J$ kept donors, no result can be rarer than one in $J + 1$. Per-period
p-values (`p_value_by_period`) say in which years the gap stands out.

The predictor choice matters here. With all 30 lags, France's placebo fit
is so tight (pre-treatment MSPE
`r if (has_data) formatC(pl_lags$mspe$pre_mspe[pl_lags$mspe$country == "France"], digits = 2, format = "e") else "tiny"`)
that a modest post-1990 gap gives it a larger ratio than Sweden, which
then ranks second. The ratio statistic rewards fit as much as effect. Report
the cut on badly fitted placebos and the predictor set together with the
p-value, because both change the denominator and the ranking.

**When it breaks.** With few donors the test has little resolution. With
donors whose own outcomes are hard to fit, the ratio statistic is noisy.
And the test only ever speaks about the sharp null of no effect in any
period; it does not give a confidence interval.

## In-time placebos {#sec-placebo-time}

**Procedure.** Move the treatment date to a year before the real one,
use only the data before the real treatment, and fit the synthetic
control as if the placebo date were real. A gap opening after the placebo
date would mean the method finds effects where there are none.

```{r sweden-placebo-time, eval = has_data}
pt_1980 <- synth_placebo(sc_sweden, type = "time", placebo_time = 1980)
pt_1970 <- synth_placebo(sc_sweden, type = "time", placebo_time = 1970)
c(placebo_1980 = pt_1980$estimate, placebo_1970 = pt_1970$estimate, real_1990 = sc_sweden$estimate)
```

```{r fig-sweden-backdate, eval = has_data, fig.height = 3.4, fig.cap = "In-time placebo: treatment backdated to 1980, fitted on 1960 to 1979 and evaluated on 1980 to 1989."}
plot_synth(pt_1980, type = "path")
```

The backdated gaps are small relative to the 1990 gap, which is what the
design requires. Andersson uses fewer lags for these fits, because the
placebo pre-period is shorter; the lecture keeps all lags for simplicity,
and the replication follows the paper.

## Conformal inference {#sec-conformal}

**Procedure.** Chernozhukov, Wuthrich, and Zhu (2021) test a sharp null
that fixes the effect in every post-treatment period, $H_0: \alpha_{0t} =
\theta_t$. Subtract $\theta_t$ from the treated outcome after treatment,
refit the synthetic control on all periods with every period's outcome
as a predictor, and compute the residuals $\hat u_t$. Under the null the
post-treatment residuals are like the others, so a statistic of the
post-treatment residuals, here $(T_1^{-1}\sum_{t > T_0} |\hat u_t|^q)^{1/q}$,
is compared with its value under all cyclic shifts of the residual vector
(moving block permutations, which keep the dependence structure). The
p-value is the share of shifts with a statistic at least as large.
Inverting the test over a grid of constant effects gives a confidence set.
Appendix \@ref(app-conformal) states the validity condition.

```{r sweden-conformal, eval = has_data}
ci_sweden <- synth_conformal(sc_sweden, per_period = TRUE)
ci_sweden
ci_sweden$per_period
```

```{r fig-sweden-conformal, eval = has_data, fig.height = 3.2, fig.cap = "Per-period gaps for Sweden with 95 percent conformal intervals."}
ggplot(ci_sweden$per_period, aes(x = time, y = gap)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "grey80") +
  geom_line() + geom_point() +
  labs(x = NULL, y = "Gap with 95% conformal interval") + sc_theme()
```

**Reading it.** The joint p-value for no effect in any post-treatment
year is `r if (has_data) fmt(ci_sweden$p_value, 2) else "large"`, and the
confidence set for a constant effect is
[`r if (has_data) fmt(ci_sweden$conf.low, 2) else ""`, `r if (has_data) fmt(ci_sweden$conf.high, 2) else ""`].
The joint test has little power for Sweden: the post-treatment block is
16 of 46 periods, so every cyclic shift overlaps it and the permutation
distribution is contaminated by the effect itself. The per-period tests
use one post-treatment period at a time and reject in every year from
1991 on, at the smallest p-value that 31 shifts allow. Conformal inference
is the only tool here that produces an interval, and it needs only
stationarity and weak dependence of the residuals, not a model for them.

**When it breaks.** The method is designed for a short post-treatment
block relative to the pre-treatment period. With a long one, test period
by period, or test the average effect over a few periods at a time. A
structural break in the residuals after treatment violates the
exchangeability the test relies on; the placebo test in space does not
depend on it.

## Size and power of the three tests {#sec-inference-mc}

The experiment draws panels from the factor model with 20 donors, 20
pre-treatment periods, and 5 post-treatment periods, with no effect and
with an effect of 1.5 (about one noise standard deviation), and applies
the three tests at the ten percent level.

```{r inference-experiment, cache = TRUE, cache.extra = helper_md5}
inf_null <- run_inference_experiment(n_sims = n_sims, effect = 0, seed = 3)
inf_alt <- run_inference_experiment(n_sims = n_sims, effect = 1.5, seed = 4)
inf_tab <- summarise_inference(bind_rows(inf_null, inf_alt))
```

```{r tab-inference, echo = FALSE}
knitr::kable(inf_tab, digits = 3, booktabs = TRUE,
             col.names = c("True effect", "In-space placebo", "Conformal", "Specification test"),
             caption = paste0("Rejection rates at the ten percent level, ", n_sims, " simulations per row. The specification test is a test of the design, not of the effect, and should not reject in either row."))
```

Both effect tests hold their size under the null (rejection rates
`r fmt(inf_tab[[2]][1])` and `r fmt(inf_tab[[3]][1])`). Against an effect
of 1.5 the placebo test rejects in `r fmt(100 * inf_tab[[2]][2], 0)` percent
of the samples and the conformal test in `r fmt(100 * inf_tab[[3]][2], 0)`
percent, with five post-treatment periods and one treated unit. The
specification test does not respond to the effect, as it should not: the
factors have mean zero after treatment in this design, so the demeaned
synthetic control and DiD estimate the same thing.

## The specification test {#sec-spec-test}

**Procedure.** Under the factor model the demeaned synthetic control and
DiD are both unbiased when the factors have mean zero after treatment,
and both biased otherwise, with different biases. Ferman and Pinto (2021)
contrast the two: fit the demeaned weights on all periods, form
$$
\hat u_t = \hat w'(y_t - \bar y) - J^{-1} \iota'(y_t - \bar y),
$$
the difference between the two counterfactuals (the treated outcome
cancels), and compare the post-treatment mean of $\hat u_t$ with its
moving block permutations. A low p-value says the two estimators disagree
by more than the pre-treatment variation explains.

```{r sweden-spec, eval = has_data}
st_sweden <- synth_spec_test(sc_dm)
st_sweden
```

```{r fig-sweden-spec, eval = has_data, fig.height = 3.4, fig.cap = "Sweden, its demeaned synthetic control, and the DiD counterfactual. The specification test asks whether the two counterfactuals diverge after 1990 by more than they wander before."}
plot_synth(st_sweden)
```

**Reading it.** For Sweden the test does not reject
(p = `r if (has_data) fmt(st_sweden$p_value, 2) else "high"`): the two
counterfactuals move together, which supports the condition both need.
Ferman and Pinto's own illustration is the Basque Country (Abadie and
Gardeazabal 2003), where the test rejects with p = 0.023; the replication
data reproduce it.

```{r basque-spec}
basque_path <- file.path(pkg_root, "data_raw", "09_Synthetic_Control", "ferman2021synthetic", "base_abadie.csv")
if (file.exists(basque_path)) {
  basque <- read.csv(basque_path)
  basque <- basque[!basque$regionname %in% c("Spain (Espana)", "Syntetic Basque Country"), ]
  basque$d <- as.integer(basque$regionname == "Basque Country (Pais Vasco)" & basque$year >= 1970)
  sc_basque <- synth_control(basque, id = "regionname", time = "year", y = "gdpcap", d = "d", demean = TRUE)
  synth_spec_test(sc_basque)
}
```

**How to read a rejection.** A rejection means at least one of the two
estimators is biased. If the pre-treatment fit is close and the donors are
many relative to the periods, the synthetic control may still be valid
while DiD is not (Section \@ref(sec-skeleton), Step 1). If the fit is
imperfect, both are suspect, and the demeaned synthetic control is the
lesser evil because its weights lean toward the loadings. The test cannot
tell which case applies; the fit and the concentration can.

## Several treated units {#sec-several}

With a handful of treated units adopting at the same time, the average
mode fits one synthetic control to their average, and the placebo tools
apply unchanged. With staggered adoption, or with treated units that
differ enough that one synthetic control cannot serve all, the separate
mode fits one synthetic control per unit from the donors that are still
untreated over that unit's window, and averages the gaps in event time.
The jackknife over treated units gives a standard error for the average.

```{r staggered-demo}
stag <- sim_synth_panel(n_donors = 25, n_treated = 5, t_pre = 15, t_post = 8, effect = c(0.5, 1, 1.5, 2),
                        adoption = c(14, 15, 17, 19, 21), seed = 7)
sc_stag <- synth_control(stag, id = "id", time = "time", y = "y", d = "d",
                         treated_units = "separate", horizon = 4, demean = TRUE)
sc_stag
sc_stag$unit_table
```

```{r fig-staggered, fig.height = 3.2, fig.cap = "Average gap by event time over five treated units with staggered adoption; the true effect rises from 0.5 to 2 over the first four periods."}
plot_synth(sc_stag)
```

Each unit's donor pool shrinks as adoption spreads, and units treated
late have fewer donors. Two refinements are in the literature and not in
the package: pooled placebo p-values across treated units (Cavallo et al.
2013) and partially pooled weights that trade unit-level fit for fit of
the average (Ben-Michael, Feller, and Rothstein 2022). Synthetic DiD with
cohort-by-cohort weights (Section 07) is the other option.

# Robustness and design checks {#sec-robustness}

## Leave-one-out {#sec-loo}

Drop each donor with positive weight in turn and refit. The spread of the
refitted estimates measures how much the answer rests on any single
donor. Andersson (2019, Figure 9) runs this for the six donors with weight
above 0.001.

```{r sweden-loo, eval = has_data}
loo_sweden <- synth_loo(sc_and)
loo_sweden
```

```{r fig-sweden-loo, eval = has_data, fig.height = 3.4, fig.cap = "Synthetic Sweden with each positive-weight donor left out in turn (grey), against the full synthetic control and Sweden."}
plot_synth(loo_sweden)
```

The estimate is robust to any single donor; leaving out Denmark, the
largest donor, gives a somewhat larger reduction, as in the paper. A
leave-one-out spread comparable to the estimate would say the result is
one donor's story.

## The donor pool {#sec-donor-pool}

The donor pool is a modelling decision, and it is made before any weight
is estimated. Four rules.

- **Unaffected.** No donor should be exposed to the treatment through
  spillovers (fuel tourism across a border, trade diversion).
- **No policy of its own.** Donors with a similar policy in the window
  are removed; Andersson drops Finland, Norway, and the Netherlands,
  which introduced carbon taxes in the 1990s.
- **Comparable.** Donors should share the structure that drives the
  outcome, so that the factor loadings of the treated unit are inside
  their span.
- **Complete.** A balanced panel over the whole window.

A useful check is to restrict the pool to the units one would defend as
comparable and see whether the estimate moves.

```{r sweden-pool, eval = has_data}
europe <- c("Belgium", "Denmark", "France", "Greece", "Iceland", "Poland", "Portugal", "Spain", "Switzerland", "Sweden")
sc_europe <- synth_control(carbon[carbon$country %in% europe, ], "country", "year", "CO2_transport_capita", "treated",
                           x = xs, lags = c(1989, 1980, 1970), pre_window = 1980:1989, v = "mspe",
                           v_control = list(maxit = 2000, starts = 3))
c(all_donors = sc_and$estimate, european_donors = sc_europe$estimate,
  pre_rmspe_all = sc_and$pre_rmspe, pre_rmspe_europe = sc_europe$pre_rmspe)
```

Restricting the pool to European donors removes New Zealand and the
United States, which carry
`r if (has_data) fmt(sum(sc_and$weights$weight[sc_and$weights$country %in% c("New Zealand", "United States")])) else "0.3"`
of the weight, and the pre-treatment fit deteriorates by a factor of
`r if (has_data) fmt(sc_europe$pre_rmspe / sc_and$pre_rmspe, 1) else "five"`.
The estimate that comes with that fit changes sign. It is not an
alternative estimate; it is a synthetic control that failed, and the
right reading is that Sweden's emissions path has no counterpart among
European donors alone. A donor-pool restriction is informative only when
the fit survives it.

## Confounders that move with the treatment {#sec-confounders}

Sweden entered a deep recession in 1991 to 1993, right after the tax.
If the recession drove the emissions gap, the gap should close when the
economy recovers. Andersson's check applies the same donor weights to GDP
per capita: if synthetic Sweden tracks real Sweden's GDP through the
recession, the recession is not what separates the emission paths.
`synth_control(weights = )` reuses the weights on another outcome.

```{r sweden-gdp, eval = has_data}
# GDP per capita is missing for Poland before 1970; the check uses 1970 to 2005.
gdp_sweden <- synth_control(carbon[carbon$year >= 1970, ], "country", "year", "GDP_per_capita", "treated", weights = sc_and)
gaps <- merge(data.frame(year = sc_and$effects$time, emissions = sc_and$effects$gap),
              data.frame(year = gdp_sweden$effects$time, gdp = gdp_sweden$effects$gap), by = "year")
change <- function(v, from, to) v[gaps$year == to] - v[gaps$year == from]
recession_tab <- data.frame(
  recession = c("1975 to 1978", "1989 to 1993"),
  gdp_gap_change = c(change(gaps$gdp, 1975, 1978), change(gaps$gdp, 1989, 1993)),
  emissions_gap_change = c(change(gaps$emissions, 1975, 1978), change(gaps$emissions, 1989, 1993))
)
recession_tab
```

```{r fig-sweden-gdp, eval = has_data, fig.height = 3.6, fig.cap = "Gaps between Sweden and synthetic Sweden in CO2 emissions (left axis) and in GDP per capita (right axis, thousands of 2005 dollars), with the two recessions shaded."}
recessions <- data.frame(start = c(1976, 1991), end = c(1978, 1993))
ggplot() +
  geom_rect(data = recessions, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), fill = "grey85") +
  geom_hline(yintercept = 0, linetype = "dotted") + geom_vline(xintercept = 1989.5, linetype = "dotted") +
  geom_line(data = gaps, aes(x = year, y = emissions, colour = "CO2 emissions gap")) +
  geom_line(data = gaps, aes(x = year, y = gdp / 5000, colour = "GDP per capita gap")) +
  scale_y_continuous(sec.axis = sec_axis(~ . * 5, name = "GDP gap, thousand dollars")) +
  scale_colour_manual(values = c("CO2 emissions gap" = "black", "GDP per capita gap" = "#B22222")) +
  labs(x = NULL, y = "Emissions gap, tons per capita", colour = NULL) + sc_theme()
```

The GDP gap falls by a similar amount in both recessions. The emissions
gap rises slightly in the first and falls in the second, and it does not
close when GDP recovers after 1993. The recession is not the explanation. The same logic applies
to any confounder that can be observed: build its synthetic version with
the same weights and ask whether it moved.

## Justifying the design: the Ferman-Pinto checklist {#sec-checklist-fp}

Ferman and Pinto (2021, Section 6) turn their results into six requests to
the applied researcher. Each maps to an output of the package.

| Request | Where to look | Sweden |
|---|---|---|
| 1. Argue that shocks specific to the treated unit are unrelated to treatment; confounders that also move the donors are allowed | the design, in words | the tax followed a political process, not an emissions shock |
| 2. Show the pre-treatment fit after removing common trends, not only in levels | `plot_synth(type = "demeaned_path")` | Figure \@ref(fig:fig-sweden-demeaned): close |
| 3. Keep the donors few relative to the pre-treatment periods, or a good fit may be overfitting | `glance()`: `n_donors`, `t_pre` | 14 donors, 30 years |
| 4. Report the weight concentration $\|w\|^2$; diluted weights are the case where the estimator is asymptotically unbiased even with imperfect fit (Ferman 2021) | `concentration` | `r if (has_data) fmt(sc_and$concentration[["l2"]]) else ""`, about `r if (has_data) fmt(sc_and$concentration[["effective_donors"]], 1) else ""` effective donors |
| 5. Compare the demeaned synthetic control with DiD; a large difference says both are biased | `synth_spec_test()` | p = `r if (has_data) fmt(st_sweden$p_value, 2) else ""`, no rejection |
| 6. Use the original synthetic control only if it recovers the treated unit's level; otherwise demean or abandon | `balance`, `intercept` of the demeaned fit | intercept `r if (has_data) fmt(sc_dm$intercept, 3) else ""`: levels match |

Sweden passes each request, which is why the paper's estimate survives a
theory published two years after it. The checklist is also a map of the
ways an application fails: a fit that is good only in levels, a donor
pool larger than the pre-period, weights on one or two donors, or a
demeaned estimate far from DiD.

# Practice {#sec-practice}

## Replications and anchors {#sec-replications}

| Paper | What it anchors | Status |
|---|---|---|
| Andersson (2019), carbon taxes and CO2 emissions in Sweden | the full workflow: predictors and V, weights, balance, path and gap, in-time and in-space placebos, leave-one-out, the GDP check | `inst/replications/09_Synthetic_Control/andersson2019carbon.Rmd` |
| Ferman and Pinto (2021), imperfect pretreatment fit | the theory of Section \@ref(sec-framework), the demeaned estimator, the specification test on the Basque Country, the Monte Carlo of Section 4 | setup described below; notebook to be written |
| Abadie, Diamond, and Hainmueller (2010), California's Proposition 99 | the canonical application and the placebo test in space | data in the `synthdid` package; see lecture 07 |
| Berman and Israeli (2022), descriptive analytics and online retailers | staggered adoption in a marketing panel, cohort-by-cohort synthetic DiD | lecture 07; `treated_units = "separate"` is the synthetic control analogue |

**Ferman and Pinto (2021): setup and construction.** The replication
folder holds the authors' code and data. Two parts.

1. *Basque application.* Data: real GDP per capita for 17 Spanish regions,
   1955 to 1997, from Abadie and Gardeazabal (2003); the Basque Country is
   treated from 1970 (terrorism), so $T_0 = 15$ and $J = 16$ after dropping
   Spain and the authors' pre-built synthetic unit. Construction: (i) the
   original synthetic control with all 15 outcome lags, solved on raw
   outcomes (`synth_control()` with defaults); (ii) the demeaned version
   (`demean = TRUE`); (iii) the specification of Abadie and Gardeazabal
   with covariates, for comparison; (iv) plots in levels and after
   subtracting each period's regional mean (`plot_synth(type =
   "demeaned_path")`); (v) the specification test (`synth_spec_test()`),
   which gives p = 0.023, and the weights table (Appendix A.2 of the paper).
2. *Monte Carlo (their Table 1).* Data: monthly CPS employment rates for
   the 50 states and DC, 1982 to 2019. Construction: estimate a linear
   factor model by interactive fixed effects (Bai 2009) with the number of
   factors chosen by the Bai-Ng criterion (the authors find four), fit
   ARMA processes to each factor and each state's residuals, and simulate
   with $T_1 = 12$ and $T_0 \in \{120, 240, 480, 1200\}$. Panels A and B
   treat the state with the second largest and second smallest fixed
   effect; Panels C and D the state with the second largest and smallest
   loading on the first factor, adding a break of two standard deviations
   to that factor after treatment. Compare the original synthetic control,
   the demeaned synthetic control, and DiD on bias and standard error.
   `sim_synth_panel()` is the stylized version of this design; the
   authors' version needs an interactive fixed effects estimator (the
   `gsynth` package), which is not part of this package.

## A synthetic control checklist {#sec-checklist}

1. Name the treated unit, the treatment date, and the window. Argue no
   anticipation (or date the treatment at the announcement) and no
   interference with the donors.
2. Describe the donor pool and every exclusion, with reasons.
3. State the predictors, the window over which covariates are averaged,
   and how $V$ was chosen. Report the balance table.
4. Report the weights, the concentration $\|w\|^2$, and the effective
   number of donors.
5. Show the path plot and the demeaned path plot. Report the
   pre-treatment RMSPE and the ratio of donors to pre-treatment periods.
6. Report the average and the per-period gaps.
7. Run in-space placebos; report the ratio ranking, the cut rule, and
   the p-value. Run at least one in-time placebo.
8. Report a conformal interval, per period when the post-treatment block
   is long.
9. Report the demeaned estimate next to the original and the
   specification test against DiD.
10. Report leave-one-out, at least one donor-pool restriction, and the
    same-weights check on the confounders a referee will name.

# Extensions {#sec-extensions}

- **Synthetic DiD** (Arkhangelsky et al. 2021) adds period weights and
  two-way effects; Section 07 and `sdid_weights()`. It is the demeaned
  synthetic control with a data-driven choice of which pre-treatment
  periods matter.
- **Penalized synthetic control** (Abadie and L'Hour 2021) adds a penalty
  on the distance between the treated unit and each donor it uses. The
  weights become unique and sparse, and the method extends to many
  treated units in disaggregated data.
- **Staggered adoption with partial pooling** (Ben-Michael, Feller, and
  Rothstein 2022) fits one synthetic control per unit and one for the
  average, and trades off their fits. The separate mode of
  `synth_control()` is the unpooled corner.
- **Interactive fixed effects and generalized synthetic control** (Bai
  2009; Xu 2017) estimate the factors and loadings directly and impute
  the counterfactual from them. They need many donors and periods, and
  they extrapolate; `gsynth` implements them.
- **Matrix completion** (Athey et al. 2021) treats the untreated outcomes
  as a matrix with missing entries and fills them under a nuclear-norm
  penalty; synthetic control and regression on lags are the two
  one-sided special cases.
- **Prediction intervals** (Cattaneo, Feng, and Titiunik 2021) separate
  the uncertainty from the weights and from the post-treatment shocks and
  give intervals under explicit models for each; `scpi` implements them.
- **Bayesian structural time series** (Brodersen et al. 2015) models the
  treated unit's series with the donors as regressors and reports a
  posterior for the counterfactual; `CausalImpact` is the common choice
  in marketing.
- **Geo experiments** (GeoLift) use the augmented synthetic control to
  evaluate market-level marketing tests, where a few regions receive a
  campaign and the rest are donors. The randomization of the treated
  regions restores design-based inference.
- **Doubly robust panel identification** (Arkhangelsky and Imbens 2022)
  puts an assignment model and an outcome model on the same footing and
  shows when synthetic control, DiD, and their combinations are valid.

# Appendix {#sec-appendix}

## The bias bound under perfect fit {#app-perfect-fit}

Setting: the factor model of Section \@ref(sec-factor-model) with $F$
factors, $\Lambda = (\lambda_1, \dots, \lambda_{T_0})'$ the $T_0 \times F$
matrix of pre-treatment factors, and weights $w^*$ with
$c'w^* = c_0$ and $\mu'w^* = \mu_0$. Suppose the fitted weights $\hat w$
give an exact pre-treatment fit: $Y_{0s} = \sum_j \hat w_j Y_{js}$ for all
$s \le T_0$.

*Step 1.* By the decomposition of Section \@ref(sec-skeleton), an exact
fit means that for every $s \le T_0$,
$$
(c_0 - c'\hat w) + \lambda_s'(\mu_0 - \mu'\hat w) = \varepsilon_s'\hat w - \varepsilon_{0s}.
$$
Stack the $T_0$ equations. With $\tilde\Lambda = (\iota, \Lambda)$ of full
column rank, the level and loading mismatches are the least-squares
projection of the noise contrast $e = (\varepsilon_s'\hat w - \varepsilon_{0s})_{s \le T_0}$
on the factor span:
$$
\begin{pmatrix} c_0 - c'\hat w \\ \mu_0 - \mu'\hat w \end{pmatrix}
= (\tilde\Lambda'\tilde\Lambda)^{-1}\tilde\Lambda' e .
$$

*Step 2.* The post-treatment bias is $\lambda_t'(\mu_0 - \mu'\hat w)$ plus
the level term, a linear function of the projected noise. Its mean is
bounded by a constant times the expected absolute size of
$(\tilde\Lambda'\tilde\Lambda/T_0)^{-1}(\tilde\Lambda' e/T_0)$. When
$\tilde\Lambda'\tilde\Lambda/T_0$ converges to a positive definite matrix and
the noise has bounded moments, $\tilde\Lambda' e/T_0$ is an average of
$T_0$ mean-zero terms with variance of order $\sigma^2 (1 + \|\hat w\|^2)/T_0$.
The bound therefore shrinks in $T_0$ and grows with the noise variance
and, through $\|\hat w\|^2 \le 1$ and the sum over donors, with $J$.
Abadie, Diamond, and Hainmueller (2010, Theorem 1) give the explicit
constant. The conclusion used in the text: under an exact fit the bias
vanishes as $T_0$ grows, at a rate governed by the noise relative to the
factors.

The step that fails with an imperfect fit is the first one. When the fit
is not exact, the stacked equations hold only up to the fitting error,
and that error is chosen by the estimator to trade off against the noise
term, which is Appendix \@ref(app-imperfect-fit).

## The limit objective with imperfect fit {#app-imperfect-fit}

Setting: all pre-treatment outcomes as predictors, $J$ fixed, $T_0 \to \infty$.
Ferman and Pinto's Assumption 4: pre-treatment averages of $\lambda_s$,
$\varepsilon_s$, and their cross products converge in probability, with
$T_0^{-1}\sum_s \lambda_s \to 0$, $T_0^{-1}\sum_s \lambda_s\lambda_s' \to \Omega_0$,
$T_0^{-1}\sum_s \varepsilon_s\varepsilon_s' \to \sigma^2 I_{J+1}$, and cross
terms to zero. (Normalizing the factor mean to zero is without loss:
a nonzero mean is absorbed in $c_j$.)

*Derivation of $Q_0$.* The objective is
$$
Q_{T_0}(w) = \frac{1}{T_0}\sum_{s \le T_0}\bigl[(c_0 - c'w) + \lambda_s'(\mu_0 - \mu'w) + (\varepsilon_{0s} - \varepsilon_s'w)\bigr]^2 .
$$
Expand the square. The squared level term is constant in $s$. The
squared factor term averages to $(\mu_0 - \mu'w)'\Omega_0(\mu_0 - \mu'w)$.
The squared noise term averages to $\sigma^2(1 + w'w)$, because
$\varepsilon_{0s}$ and $\varepsilon_s'w$ are uncorrelated with variance
$\sigma^2$ and $\sigma^2 w'w$. Every cross term averages to zero. Hence
$Q_{T_0}(w) \to_p Q_0(w)$ pointwise, and uniformly on the simplex because
$Q_{T_0}$ is quadratic in $w$ with converging coefficients. Continuity
and compactness give $\hat w \to_p \bar w = \arg\min_\Delta Q_0$ when the
minimizer is unique (Proposition 1).

*Why $\bar w$ misses the loadings.* Let $\mathcal{W}^* = \{w \in \Delta : c'w = c_0,\ \mu'w = \mu_0\}$.
On $\mathcal{W}^*$ the first two terms vanish and $Q_0(w) = \sigma^2(1 + w'w)$.
The minimizer of $Q_0$ over all of $\Delta$ equals a point of $\mathcal{W}^*$
only if that point also minimizes $w'w$ over $\Delta$ up to the trade-off
with the first two terms, which happens only when $\sigma^2 = 0$ or by
coincidence. Otherwise $\bar w$ lies outside $\mathcal{W}^*$, and the
mismatch $\mu_0 - \mu'\bar w$ is a fixed nonzero vector.

*The demeaned estimator (Proposition 2).* Replace every series by its
deviation from its own pre-treatment mean. The level term drops from
the objective, so the limit is
$\tilde Q_0(w) = (\mu_0 - \mu'w)'\Omega_0(\mu_0 - \mu'w) + \sigma^2(1 + w'w)$
and the estimator converges to
$\alpha_{0t} + \lambda_t'(\mu_0 - \mu'\tilde w) + (\varepsilon_{0t} - \varepsilon_t'\tilde w)$.
It is asymptotically unbiased when $\mathbb{E}[\lambda_t] = 0$ for $t > T_0$,
whatever $\tilde w$ is.

*Efficiency against DiD (Proposition 3).* Assume the factor and noise
moments after treatment equal the pre-treatment limits. Then for any
fixed $w$ the asymptotic mean squared error of
$\hat\alpha_{0t}(w)$ is $\tilde Q_0(w)$: the squared factor mismatch plus
the noise variance. The demeaned weights minimize $\tilde Q_0$ over the
simplex, and DiD uses $w = \iota/J$, which is in the simplex. Hence
$\tilde Q_0(\tilde w) \le \tilde Q_0(\iota/J)$: the demeaned synthetic
control weakly dominates DiD in asymptotic mean squared error. The
inequality is strict unless uniform weights already minimize the
trade-off.

## One family: weights plus an outcome model {#app-family}

Write every estimator as
$\hat Y_{0t}^N = \hat m_t(Z_0) + \sum_j w_j (Y_{jt} - \hat m_t(Z_j))$.

- *Synthetic control:* $\hat m_t \equiv 0$.
- *Demeaned synthetic control:* $\hat m_t(Z_j) = \bar Y_j$, the unit's
  pre-treatment mean. Then
  $\hat Y_{0t}^N = \bar Y_0 + \sum_j w_j (Y_{jt} - \bar Y_j)$, which is the
  intercept form $\hat a + \sum_j w_j Y_{jt}$ with
  $\hat a = \bar Y_0 - \sum_j w_j \bar Y_j$. Minimizing the pre-treatment
  fit over $(a, w)$ and over $w$ with the series demeaned give the same
  $w$, because the optimal $a$ for given $w$ is exactly that expression.
- *Augmented synthetic control:* $\hat m_t(Z) = \hat a_t + \hat\beta_t' Z$
  from a ridge regression of $Y_{jt}$ on the donors' predictors.
  Substituting gives
  $\hat Y_{0t}^N = \sum_j w_j Y_{jt} + \hat\beta_t'(Z_0 - \sum_j w_j Z_j)$: the
  synthetic control plus the outcome model applied to the imbalance.
- *Synthetic DiD:* $\hat m_t(Z_j) = \hat a_j + \hat b_t$, two-way effects
  estimated with period weights $\lambda_t$ on the pre-treatment
  periods; the unit weights are then fitted to the residuals.
- *DiD:* the same outcome model with uniform unit and period weights.

The error decomposition of Section \@ref(sec-weights-model) follows by
substituting $Y_{jt} = m_t(Z_j) + e_{jt}$.

## Validity of the moving-block conformal test {#app-conformal}

Under the sharp null the adjusted treated outcome has no effect in any
period, and the refit on all $T$ periods treats every period alike. Let
$u_t$ be the population residual of that fit and $\hat u_t$ its estimate.
Chernozhukov, Wuthrich, and Zhu (2021) assume that $(u_t)$ is stationary
and weakly dependent, and that the estimator is consistent so that
$\hat u_t - u_t$ is small uniformly in $t$. For a statistic
$S(\hat u_{\pi})$ of a block of $T_1$ consecutive residuals, the set of
$T$ cyclic shifts $\pi$ treats every block of $T_1$ consecutive periods
as the post-treatment block in turn. Under stationarity each block has
the same distribution, so the rank of the observed block among the $T$
blocks is approximately uniform, and
$\mathbb{P}(\hat p \le a) \to a$ as $T \to \infty$ with $T_1$ fixed. With
independent residuals the test is exact in finite samples. The
approximation degrades when $T_1/T$ is not small, because the blocks
overlap the true post-treatment period heavily and the effect, if the
null is false, contaminates most of them; this is the low power seen for
Sweden. The specification test of Ferman and Pinto applies the same
argument to the contrast $\hat u_t$ of Section \@ref(sec-spec-test),
with the treated outcome cancelled.

## The nested problem for the importance weights {#app-nested}

With predictors $Z_0$ (a vector of length $K$) and $Z$ ($K \times J$),
standardized by their standard deviation across units, the inner problem
for given $V = \mathrm{diag}(v_1, \dots, v_K)$ is the quadratic program
$$
w(V) = \arg\min_{w \in \Delta} (Z_0 - Zw)'V(Z_0 - Zw),
$$
solved exactly by an active-set method (`quadprog`) or by Frank-Wolfe
steps on the simplex. The outer problem chooses $V$ in the simplex to
minimize the pre-treatment mean squared prediction error of the outcome,
$T_0^{-1}\sum_{s \le T_0}(Y_{0s} - Y_s'w(V))^2$. It is not convex and has
flat regions: many $V$ give nearly the same $w$. `synth_control(v =
"mspe")` runs Nelder-Mead from the equal-weight and regression-based
starts, refines by BFGS, and keeps the best. When every pre-treatment
outcome is a predictor, the outer objective is already the inner
objective restricted to the lag rows, so putting all of $V$ on the lags
attains the outer minimum and covariates receive zero weight (Kaul et
al. 2022).

## References

- Abadie, Alberto, Alexis Diamond, and Jens Hainmueller. 2010. "Synthetic Control Methods for Comparative Case Studies: Estimating the Effect of California's Tobacco Control Program." *Journal of the American Statistical Association*.
- Abadie, Alberto, Alexis Diamond, and Jens Hainmueller. 2015. "Comparative Politics and the Synthetic Control Method." *American Journal of Political Science*.
- Abadie, Alberto, and Javier Gardeazabal. 2003. "The Economic Costs of Conflict: A Case Study of the Basque Country." *American Economic Review*.
- Abadie, Alberto, and Jeremy L'Hour. 2021. "A Penalized Synthetic Control Estimator for Disaggregated Data." *Journal of the American Statistical Association*.
- Andersson, Julius J. 2019. "Carbon Taxes and CO2 Emissions: Sweden as a Case Study." *American Economic Journal: Economic Policy*.
- Arkhangelsky, Dmitry, Susan Athey, David A. Hirshberg, Guido W. Imbens, and Stefan Wager. 2021. "Synthetic Difference-in-Differences." *American Economic Review*.
- Arkhangelsky, Dmitry, and Guido W. Imbens. 2022. "Doubly Robust Identification for Causal Panel Data Models." *The Econometrics Journal*.
- Athey, Susan, Mohsen Bayati, Nikolay Doudchenko, Guido Imbens, and Khashayar Khosravi. 2021. "Matrix Completion Methods for Causal Panel Data Models." *Journal of the American Statistical Association*.
- Bai, Jushan. 2009. "Panel Data Models with Interactive Fixed Effects." *Econometrica*.
- Ben-Michael, Eli, Avi Feller, and Jesse Rothstein. 2021. "The Augmented Synthetic Control Method." *Journal of the American Statistical Association*.
- Ben-Michael, Eli, Avi Feller, and Jesse Rothstein. 2022. "Synthetic Controls with Staggered Adoption." *Journal of the Royal Statistical Society: Series B*.
- Berman, Ron, and Ayelet Israeli. 2022. "The Value of Descriptive Analytics: Evidence from Online Retailers." *Marketing Science*.
- Brodersen, Kay H., Fabian Gallusser, Jim Koehler, Nicolas Remy, and Steven L. Scott. 2015. "Inferring Causal Impact Using Bayesian Structural Time-Series Models." *Annals of Applied Statistics*.
- Cattaneo, Matias D., Yingjie Feng, and Rocio Titiunik. 2021. "Prediction Intervals for Synthetic Control Methods." *Journal of the American Statistical Association*.
- Cavallo, Eduardo, Sebastian Galiani, Ilan Noy, and Juan Pantano. 2013. "Catastrophic Natural Disasters and Economic Growth." *Review of Economics and Statistics*.
- Chernozhukov, Victor, Kaspar Wuthrich, and Yinchu Zhu. 2021. "An Exact and Robust Conformal Inference Method for Counterfactual and Synthetic Controls." *Journal of the American Statistical Association*.
- Doudchenko, Nikolay, and Guido W. Imbens. 2016. "Balancing, Regression, Difference-in-Differences and Synthetic Control Methods: A Synthesis." NBER Working Paper 22791.
- Ferman, Bruno. 2021. "On the Properties of the Synthetic Control Estimator with Many Periods and Many Controls." *Journal of the American Statistical Association*.
- Ferman, Bruno, and Cristine Pinto. 2021. "Synthetic Controls with Imperfect Pretreatment Fit." *Quantitative Economics*.
- Hsiao, Cheng, H. Steve Ching, and Shui Ki Wan. 2012. "A Panel Data Approach for Program Evaluation: Measuring the Benefits of Political and Economic Integration of Hong Kong with Mainland China." *Journal of Applied Econometrics*.
- Kaul, Ashok, Stefan Klossner, Gregor Pfeifer, and Manuel Schieler. 2022. "Standard Synthetic Control Methods: The Case of Using All Preintervention Outcomes Together with Covariates." *Journal of Business and Economic Statistics*.
- Xu, Yiqing. 2017. "Generalized Synthetic Control Method: Causal Inference with Interactive Fixed Effects Models." *Political Analysis*.
