---
title: "Replication: Leveraging Large-Scale Granular Single-Source Data for TV Advertising"
subtitle: "An Identification Strategy with a Control Function (Hsieh, Du, and Lu, Marketing Science 2026)"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{pdflscape}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/05_IV/hsieh2026leveraging/"
  n_hh: 25000
  n_days: 150
  n_boot: 50
  n_sims: 20
  n_threads: 4
  n_threads_boot: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "90%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(causalmetrics)
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(estimatr)
library(mlr3)
library(mlr3learners)
helper_path <- file.path(pkg_root, "inst", "vignette_helpers", "05_iv", "hsieh_helpers.R")
source(helper_path)
helper_md5 <- tools::md5sum(helper_path)
data_path <- params$data_path
sim_dir <- file.path(data_path, "data")
dir.create(sim_dir, showWarnings = FALSE)
n_hh <- params$n_hh; n_days <- params$n_days; n_boot <- params$n_boot; n_sims <- params$n_sims
n_threads <- params$n_threads; n_threads_boot <- params$n_threads_boot
size_key <- list(n_hh, n_days)
size_tag <- paste0("_", n_hh, "x", n_days)
sim_file <- function(name) file.path(sim_dir, paste0(name, size_tag, ".rds"))
file_cached <- function(name, expr) {
  f <- sim_file(name)
  if (file.exists(f)) return(readRDS(f))
  value <- expr
  saveRDS(value, f)
  value
}
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
fmt0 <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
stars <- function(p) ifelse(is.na(p), "", ifelse(p < 0.01, "***", ifelse(p < 0.05, "**", ifelse(p < 0.10, "*", ""))))
est_star <- function(b, se, digits = 3) ifelse(is.na(b), "---", paste0(fmt(b, digits), stars(2 * pnorm(-abs(b / se)))))
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
theme_set(theme_minimal(base_size = 10))
term_labels <- c(
  `(Intercept)` = "Intercept", A_stock = "Ad stock", CF_stock = "Control function", post = "Post conversion",
  lf = "Frequency (log)", lf2 = "Frequency (log) sq.", lr = "Recency (log)", lr2 = "Recency (log) sq.",
  `A_stock:post` = "Ad stock x post conversion", `A_stock:lf` = "Ad stock x freq (log)",
  `A_stock:lf2` = "Ad stock x freq (log) sq.", `A_stock:lr` = "Ad stock x rec (log)",
  `A_stock:lr2` = "Ad stock x rec (log) sq.", promo1m = "First month post-conversion promotion",
  comp = "Competitor ad spend", z_view = "Average TV viewing", z_sports = "Sports viewing",
  `A_stock:z_view` = "Ad stock x average TV viewing", `A_stock:z_sports` = "Ad stock x sports viewing",
  sigma1 = "Standard deviation (intercept)", sigma2 = "Standard deviation (ad stock)", rho = "Rho"
)
param_labels <- c(
  `(Intercept)` = "$\\alpha_0$", A_stock = "$\\beta_0$", CF_stock = "$\\delta$", post = "$\\alpha_{2,0}$",
  lf = "$\\alpha_{2,1}$", lf2 = "$\\alpha_{2,2}$", lr = "$\\alpha_{2,3}$", lr2 = "$\\alpha_{2,4}$",
  `A_stock:post` = "$\\beta_{2,0}$", `A_stock:lf` = "$\\beta_{2,1}$", `A_stock:lf2` = "$\\beta_{2,2}$",
  `A_stock:lr` = "$\\beta_{2,3}$", `A_stock:lr2` = "$\\beta_{2,4}$", promo1m = "$\\gamma_1$", comp = "$\\gamma_2$",
  z_view = "$\\alpha_1$", z_sports = "$\\alpha_1$", `A_stock:z_view` = "$\\beta_1$", `A_stock:z_sports` = "$\\beta_1$",
  sigma1 = "$\\sigma_1$", sigma2 = "$\\sigma_2$", rho = "$\\rho$"
)
```

# The paper in one page {#sec-paper}

**The question is whether linear TV advertising causes purchases, and how
the effect changes with a household's purchase history.** Hsieh, Du, and Lu
(2026) link second-by-second TV viewing from 1.4 million smart-TV
households to the daily orders of a food delivery platform over 133 days.
TV advertising cannot be randomized at the household level, so the answer
must come from observational exposure data.

**Observed exposure is endogenous for two reasons, and controls cannot fix
either.** Targeting bias: the brand buys shows whose audiences already
order more. Activity bias: households that watch more TV, or watch on a
given day, differ in demand; the same unexpected long workday lowers TV
time and raises the chance of ordering delivery. Both operate through
household-day unobservables, which no set of controls absorbs.

**The instrument comes from an institutional feature: networks assign
within-show ad slots quasi-randomly across advertisers.** A household that
watches only part of a targeted show sees the focal ad only if the ad
happens to air during the watched segment. Expected exposure is the
integral of the network's ad-timing density over the watched segment,
$P_{is} = \int_{View_{is}} l_s(x)\,dx$. The instrument is realized minus
expected exposure, $IV_{is} = A_{is} - A^b_s P_{is}$, summed over the day's
shows. It is nonzero only for partial viewers of targeted shows, mean
zero, positively correlated with exposure, and uncorrelated with the
confounder (Proposition 1) and with every endogenous input, the targeting
decision, the viewing probability, and their product (Proposition 2). The
last property gives falsification checks from observables.

**The purchase model is a probit with an ad stock and a control-function
stock.** Utility is $U_{it} = \alpha_{it} + \beta_{it} AS_{it} + \gamma X_{it} + \delta\, CFS_{it} + \varepsilon_{it}$
with $AS_{it} = \sum_l \lambda_A^l A_{i,t-l}$. Baseline propensity
$\alpha_{it}$ and ad responsiveness $\beta_{it}$ are quadratic in log
purchase frequency and log recency after conversion, shift with household
traits, and carry random effects. The first stage regresses daily exposure
on the instrument and the controls; its residual is accumulated into
$CFS_{it} = \sum_l \lambda_{CF}^l \hat\varepsilon_{i,t-l}$. The coefficient
$\delta$ tests exogeneity. Standard errors come from 50 household
bootstraps, and the two decay rates from an out-of-sample grid search.

**Two findings.** Ignoring endogeneity overstates ad elasticities by 55
percent (same-day 0.072 against 0.045, 30-day 0.222 against 0.143). Ad
responsiveness is an inverted U in purchase frequency and a U in recency,
so early repeat buyers and lapsed buyers respond most.

**The replication package holds no data.** The authors can share only a
simulated data-generating process (DGP) with a known truth and the
estimation code. This note reruns their scripts, matches them with the
package's control-function functions, adds the show level that their
scripts omit so that every table and figure of the paper has a counterpart
in the published layout, and then asks what the design would look like
under violations of its assumptions.

# The design, step by step {#sec-design}

## The DGP the identification rests on

**Purchases respond to exposure and to a demand shock the analyst does not
see.** For household $i$ on day $t$,
$Y_{it} = f(\alpha + \beta A_{it} + \gamma X_{it} + u_{it})$ with
$A_{it} = \sum_{s \in S_t} A_{is}$ (Equation 1). Endogeneity means
$A_{it}$ is correlated with $u_{it}$.

**Exposure is the product of three decisions made by three parties.** The
brand targets show $s$, $A^b_s = f_b(X_s, u_s)$ (Equation 2); the household
watches segments $View_{is} = f_v(X_{is}, u_{is}) \subseteq [0, 1]$
(Equation 3); the network draws the airing time $T^b_s \sim l_s(x)$
(Equation 4). Exposure is $A_{is} = A^b_s\, 1(T^b_s \in View_{is})$
(Equation 5). The first two decisions may depend on the demand shock, which
is where targeting bias and activity bias enter. The third does not, which
is the exogenous variation.

**Two assumptions make the third decision usable.** Quasi-random ordering:
$l_s$ does not depend on the brand or the household, so it can be estimated
from every brand's ads on that network. Nonstrategic viewership:
$View_{is}$ does not depend on $T^b_s$, so a household does not switch
channels because the focal ad is about to air. Each has an empirical check:
compare focal and nonfocal airing-time distributions, and compare
ad-completion rates for focal and nonfocal ads.

## The instrument

**Decompose realized exposure into an expected part and a shifter.**
$A_{is} = A^b_s P_{is} + A^b_s (1(T^b_s \in View_{is}) - P_{is})$
(Equation 7). The first term collects everything endogenous; the second is
a centered Bernoulli draw, $1 - P_{is}$ with probability $P_{is}$ and
$-P_{is}$ otherwise, whose mean is zero given the targeting and viewing
decisions. That second term is the instrument (Equation 8).

**Why the mean-zero property matters for practice.** Because $E[IV_{is} \mid A^b_s, View_{is}] = 0$,
the instrument is uncorrelated with $u_{is}$, with $A^b_s$, with $P_{is}$,
and with $A^b_s P_{is}$. The first is exogeneity and cannot be tested; the
other three can, and a rejection of any of them means expected exposure was
mis-estimated. The instrument is a residual from a known assignment
process, not from a fitted model of exposure, so its validity does not
depend on a propensity model being right.

## The response model and the control function

**Ad effects accumulate and depend on purchase history.** Equations 9 to
15: an ad stock with daily decay $\lambda_A$; $\alpha_{it}$ and $\beta_{it}$
each equal to a base level, a shift after the first purchase, a quadratic
in log frequency and log recency, a linear index in household traits
$Z_i$, and a household random effect. The log and squared-log terms let
both curves bend, which is what Figure 4 of the paper reports.

**A nonlinear second stage needs a control function, not two-stage least
squares.** Equations 16 to 19: regress $A_{it}$ on $IV_{it}$ and the
controls, keep the residual $\hat\varepsilon^f_{it}$, accumulate it with
decay $\lambda_{CF}$ into $CFS_{it}$, and add $CFS_{it}$ to the probit.
Conditioning on the residual removes the part of the demand shock that is
correlated with the ad stock; $\delta = 0$ is the exogeneity test
(Wooldridge 2015). The stock is needed because the endogenous regressor is
itself a stock: if the confounder persists, so must its control, and the
two decays need not be equal.

**Inference must account for the generated regressor.** The second stage
treats $CFS_{it}$ as data. The paper resamples households, 50 times, and
re-runs both stages.

# What the replication package allows {#sec-package}

**The data are under a nondisclosure agreement; the package simulates
them.** Table \@ref(tab:files) lists the four scripts. Their DGP is a
household-day panel with a known truth. It keeps the two biases of the
paper, an unobservable that drives targeting, viewing, and purchase, but
it has no show level: no airing-time density, no viewing segments, no other
brands' ads.

```{r files, echo = FALSE}
tibble(
  Script = c("Step1 - AdExposureSimulation.R", "Step2 - PurchaseSimulation.R", "Step3 - BootstrapEstimate.R", "Step4 - ModelEstimation.R"),
  Content = c("25,000 households x 150 days; $u_{endo}$ drives the targeting probability of three shows a day and the watched portion; exposure = Bernoulli(portion) x targeting; expected exposure; ad stock with decay 0.7",
              "probit utility with $\\alpha_0 = -2.5$, $\\beta_0 = 0.012$, state terms, a 30-day promotion after the first purchase, competitor spend, and a confounder that accumulates $u_{endo}$ with decay 0.95",
              "50 household resamples; first stage, control-function stock with decay 0.95, probit",
              "naive probit and control-function probit; a table with the true parameters"),
  `Used for` = c("benchmark panel, rerun with seed 1107", "benchmark outcomes, rerun with seed 123", "benchmark standard errors", "benchmark point estimates")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "The replication package and its role in this note.") %>%
  tab_style() %>% column_spec(1, width = "3.6cm") %>% column_spec(2, width = "8.2cm") %>% column_spec(3, width = "3.2cm")
```

**Three points where the scripts and the paper part ways.** First, the
scripts' expected exposure uses the targeting probabilities, which are
functions of the unobservable; an analyst never sees them. The paper's
expected treatment conditions on the realized targeting decision and the
watched portion. Both are mean zero given the unobservable, so both are
valid in the simulation, but only the second can be computed from data.
Second, the bootstrap keeps each sampled household once
(`filter(hh %in% boot_hh)`), so every "bootstrap" sample is a 63 percent
subsample without replacement, and its standard deviation is about
$\sqrt{1/0.632 - 1} = 0.76$ of a cluster bootstrap in expectation. Third,
the scripts have no show level, so Figures 2 and 3 and the
household-show half of Table 2 cannot be produced from them.

**What this note does with each.** Section \@ref(sec-benchmark) reruns
the scripts and matches them with `cf_residuals()` and a cluster
bootstrap that keeps duplicates, reporting both instruments and both
standard errors. Sections \@ref(sec-data) to \@ref(sec-hte) add a show
level, networks with their own airing-time densities, viewing segments,
nonfocal ads, on top of the authors' purchase model, and reproduce the
paper's tables and figures in their layouts. Section \@ref(sec-robust)
reruns the paper's robustness checks, and Section \@ref(sec-extensions)
asks what the diagnostics catch when the assumptions fail.

# Benchmark: the authors' scripts, rerun and matched {#sec-benchmark}

## The scripts, verbatim

**The helper file reproduces Steps 1 and 2 line by line, with seeds 1107
and 123.** Random streams inside `future_lapply(future.seed = TRUE)` do not
depend on the number of workers, so the panel equals the one the authors
would obtain. The vectorized purchase simulation used later in this note
is checked against the authors' household loop on the same panel.

```{r authors-data}
fs_file <- sim_file("fs_ready"); ss_file <- sim_file("ss_ready")
if (file.exists(ss_file)) {
  ss <- readRDS(ss_file)
} else {
  fs <- authors_step1(N_hh = n_hh, T_day = n_days, workers = n_threads)
  saveRDS(fs, fs_file)
  ss <- authors_step2(fs, workers = n_threads)
  saveRDS(ss, ss_file)
}
ours <- simulate_purchases(ss[order(ss$hh, ss$t), c("hh", "t", "A", "A_stock", "u_endo", "u_exo", "comp")], purchase_par(FALSE, FALSE))
c(rows = nrow(ss), purchase_rate = mean(ss$purchase), converted = mean(tapply(ss$purchase, ss$hh, max)),
  same_purchases = all(ours$purchase == ss$purchase),
  same_states = all(ours$freq == ss$freq & ours$rec == ss$rec & ours$promo1m == ss$promo1m))
```

```{r authors-desc, echo = FALSE}
conv_share <- mean(tapply(ss$purchase, ss$hh, max))
```

The authors' purchase parameters convert `r fmt(100 * conv_share, 1)` percent
of households within 150 days, against 3.8 percent in the paper; the daily
conversion rate of an unconverted household, $\Phi(-2.5) = 0.62$ percent,
is fifty times the paper's 0.012 percent. The confounder enters utility
with weight 0.005 on a stock whose standard deviation is about 3.2, so the
endogeneity is mild. We keep every value as shipped.

## Step 4 with the package

**The first stage is `cf_residuals()`, the stock is a two-line
recursion, and the second stage is `glm()`.** `cf_residuals()` returns the
data with the residual appended and the first-stage strength of
`iv_first_stage()` as an attribute. The stock $CFS_{it}$ is accumulated
within household with `adstock_by()`, defined in the helper file.

```{r adstock-def, eval = FALSE}
adstock <- function(x, lambda) as.numeric(stats::filter(x, lambda, method = "recursive"))
adstock_by <- function(x, id, lambda) as.numeric(stats::ave(x, id, FUN = function(v) adstock(v, lambda)))
```

```{r authors-match}
s4 <- authors_step4(ss)
bench <- add_states(ss)
spec_bench <- col_spec(lambda_A = 0.7, cf = TRUE, lambda_CF = 0.95, interactions = TRUE, traits = FALSE)
ours_bench <- fit_column(bench, spec_bench)
fs_strength <- attr(ours_bench$data, "first_stage")$strength
cf_authors <- rename_authors(coef(s4$full)); cf_naive <- rename_authors(coef(s4$naive))
terms_b <- names(cf_authors)
ours_cf <- coef(ours_bench$fit)[terms_b]
bench_model_se <- summary(ours_bench$fit)$coefficients[terms_b, "Std. Error"]
c(max_abs_diff = max(abs(cf_authors - ours_cf)), F_first_stage = fs_strength$F, F_robust = fs_strength$F_robust)
rm(s4, ours_bench); invisible(gc())
```

```{r authors-boot}
boot_file <- sim_file(paste0("bench_boot_", n_boot))
if (file.exists(boot_file)) {
  bb <- readRDS(boot_file)
} else {
  b_authors5 <- authors_step3(ss, B = min(5L, n_boot), seed = 1107, workers = n_threads_boot)
  bench_cols <- c("hh", "t", "purchase", "A", "IV", controls_base, "freq", "rec")
  b_unique <- cluster_bootstrap(bench[, bench_cols], spec_bench, n_boot = n_boot, seed = 1107, unique_only = TRUE,
                                workers = n_threads_boot)
  b_cluster <- cluster_bootstrap(bench[, bench_cols], spec_bench, n_boot = n_boot, seed = 1107, unique_only = FALSE,
                                 workers = n_threads_boot)
  bb <- list(authors5 = b_authors5, unique = b_unique, cluster = b_cluster)
  saveRDS(bb, boot_file)
}
auth_draws <- rename_authors(bb$authors5$draws)
n5 <- min(nrow(auth_draws), nrow(bb$unique$draws))
c(first_draws_max_diff = max(abs(auth_draws[seq_len(n5), ] - bb$unique$draws[seq_len(n5), colnames(auth_draws)])))
```

```{r bench-table, echo = FALSE}
truth_b <- authors_truth[terms_b]
model_se <- bench_model_se
bench_tab <- tibble(
  Term = term_labels[terms_b], True = ifelse(is.na(truth_b), "---", fmt(truth_b, 3)),
  `Naive` = fmt(cf_naive[terms_b], 4), `Control function` = fmt(ours_cf, 4),
  `Model SE` = fmt(model_se, 4), `Authors' bootstrap SE` = fmt(bb$unique$se[terms_b], 4), `Cluster bootstrap SE` = fmt(bb$cluster$se[terms_b], 4),
  `t (cluster)` = fmt(ours_cf / bb$cluster$se[terms_b], 2)
)
bench_tab$Naive[terms_b == "CF_stock"] <- "---"
kbl(bench_tab, booktabs = TRUE, linesep = "", escape = FALSE, align = "lrrrrrrr",
    caption = paste0("The authors' Step 4 table on their DGP (", fmt0(n_hh), " households, ", n_days, " days): true parameters, naive probit, control-function probit, and three standard errors. The authors' bootstrap keeps each sampled household once; the cluster bootstrap keeps duplicates. ", n_boot, " draws each.")) %>%
  tab_style("scale_down")
```

```{r bench-numbers, echo = FALSE}
b_naive <- cf_naive["A_stock"]; b_cf <- ours_cf["A_stock"]
ratio_se <- mean(bb$unique$se[terms_b] / bb$cluster$se[terms_b])
```

**What the benchmark shows.** The package reproduces the authors' point
estimates to `r format(signif(max(abs(cf_authors - ours_cf)), 2))`, and its
`unique_only` bootstrap reproduces their first five draws exactly, so the
50-draw column labelled "authors' bootstrap" is what their Step 3 returns.
The naive ad-stock coefficient is `r fmt(b_naive, 4)` against a truth of
0.012 and a corrected `r fmt(b_cf, 4)`. The control-function coefficient
is `r fmt(ours_cf["CF_stock"], 4)` with cluster t statistic
`r fmt(ours_cf["CF_stock"] / bb$cluster$se["CF_stock"], 2)`. The authors' bootstrap
standard errors average `r fmt(ratio_se, 2)` of the cluster bootstrap's,
close to the 0.76 that a 63 percent subsample implies, and the model-based
standard errors that treat the stock as data are smaller still for the
ad-stock terms. Table \@ref(tab:bench-instruments) compares the scripts'
oracle instrument with the paper's.

```{r bench-instruments, echo = FALSE}
iv_rows <- function(iv, label) {
  f <- fit_column(bench, modifyList(spec_bench, list(iv = iv)))
  st <- attr(f$data, "first_stage")$strength
  tibble(Instrument = label, `Mean` = fmt(mean(bench[[iv]]), 4), `Corr with $A$` = fmt(cor(bench$A, bench[[iv]]), 3),
         `Corr with $u_{endo}$` = fmt(cor(bench$u_endo, bench[[iv]]), 4), `First-stage F` = fmt0(st$F),
         `$\\beta_0$` = fmt(coef(f$fit)["A_stock"], 4), `$\\delta$` = fmt(coef(f$fit)["CF_stock"], 4))
}
two_ivs <- file_cached("bench_instruments", bind_rows(iv_rows("IV", "Scripts: $A - \\sum_k P_k\\, p^{target}_k$ (oracle)"),
                                                     iv_rows("IV_paper", "Paper: $A - \\sum_k A^b_k P_k$")))
kbl(two_ivs, booktabs = TRUE, linesep = "", escape = FALSE, caption = "Two instruments on the authors' DGP. The scripts subtract the expected exposure computed from the targeting probabilities; the paper subtracts the expected exposure given the realized targeting decisions.") %>%
  tab_style()
rm(bench, ss, ours); invisible(gc())
```

Both instruments are mean zero and orthogonal to the unobservable, and
both correct the naive estimate. The paper's version is the one an analyst
can build, and the rest of the note uses it.

# A show level for the authors' DGP {#sec-data}

## What is added, and why

**The paper's construction needs shows, densities, and segments.** The
helper simulates, for each household-day, three shows on three networks.
Each network has its own airing-time density on $[0, 1]$: trimodal like
MTV in the paper's Figure 2, bimodal, and close to uniform. The brand
targets each show with the authors' probabilities, rising in the
household-day shock. The household watches a segment whose length follows
the authors' rule (a 30 percent chance of not watching, otherwise a length
rising in the shock) plus two standardized traits, average viewing and
sports viewing, the latter acting on the second network only. The focal ad
airs at a time drawn from the network's density; nonfocal ads are drawn
from the same density. Exposure follows Equation 5. Expected exposure
integrates the empirical distribution of nonfocal ads over the segment,
and the instrument is Equation 8. A second brand with no effect on
purchases is exposed through the same viewing process for the placebo
check. The purchase model is the authors' with two additions for Table 4
columns 5 and 6 and Table 5: the two traits enter $\alpha_{it}$
($-0.05$, $0.03$) and $\beta_{it}$ ($-0.004$, $0.004$), and household random
effects have $\sigma_1 = 0.2$, $\sigma_2 = 0.01$, $\rho = -0.3$.

**Two of the authors' values are changed, and why.** The intercept is
$-3.0$ instead of $-2.5$: with $-2.5$ two thirds of households convert
within 150 days and repeat buyers reach purchase frequencies at which the
quadratic in log frequency turns the ad effect negative on average; with
$-3.0$ about a quarter convert, still far above the paper's 3.8 percent
but with an average ad effect of the paper's sign. The weight of the
confounder is 0.015 instead of 0.005: with the authors' value the naive
bias in $\beta_0$ is about 0.003, below the standard error at this sample
size, so the comparison that Table 4 is built to show would sit inside
sampling noise. Both changes are to the purchase model only; the exposure
process is the authors' with a show level added, and the benchmark of
Section \@ref(sec-benchmark) uses the authors' values throughout.

```{r show-data}
show_file <- sim_file("show_panel")
if (file.exists(show_file)) {
  sim <- readRDS(show_file)
} else {
  par_b <- purchase_par(traits = TRUE, random = TRUE, a0 = -3.0, delta = 0.015)
  sim <- simulate_exposure(n_hh = n_hh, n_days = n_days, seed = 1107, par = par_b)
  sim$panel <- add_states(simulate_purchases(sim$panel, sim$par))
  saveRDS(sim, show_file)
}
pan <- sim$panel
set.seed(1107)
sub_hh <- sort(sample(unique(pan$hh), min(5000L, n_hh)))
pan_sub <- pan[pan$hh %in% sub_hh, ]
c(rows = nrow(pan), purchase_rate = mean(pan$purchase), converted = mean(tapply(pan$purchase, pan$hh, max)),
  exposures_per_day = mean(pan$A), partial_share = sum(pan$n_partial) / sum(pan$n_target))
```

## Table 1: descriptives

**The paper's Table 1 tells the reader how rare exposures and purchases
are.** Ours reports the same five quantities for the simulated panel; show
length is one hour.

```{r table1, echo = FALSE}
conv <- pan %>% group_by(hh) %>% summarise(n = sum(purchase), first = suppressWarnings(min(t[purchase == 1])), .groups = "drop") %>% filter(n > 0)
ipt <- pan %>% filter(purchase == 1) %>% group_by(hh) %>% reframe(gap = diff(t))
t1 <- bind_rows(
  quantile_row(pan$view_hr, "TV viewing per household per day (hr)"),
  quantile_row(pan$targeted_hr, "Targeted show viewing per household per day (hr)"),
  quantile_row(pan$A, "Number of focal brand ad exposures per household per day"),
  quantile_row(conv$n, "Purchase frequency per converted household"),
  quantile_row(ipt$gap, "Interpurchase time per converted household (days)")
)
t1 %>% mutate(across(-Variable, ~ fmt(.x, 2))) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("", "Mean", "SD", "5\\%", "25\\%", "50\\%", "75\\%", "95\\%"), escape = FALSE,
      caption = paste0("Descriptive statistics (the layout of Table 1). Simulated TV viewing and purchase behavior for ", fmt0(n_hh), " households over ", n_days, " days.")) %>%
  tab_style() %>% footnote(general = paste0(fmt0(nrow(conv)), " households (", fmt(100 * nrow(conv) / n_hh, 1), " percent) converted."), general_title = "Note. ", footnote_as_chunk = TRUE)
```

## Figure 1: model-free evidence

**The paper plots purchase probability against frequency, recency, and
same-day exposures before fitting anything.** The first two panels motivate
the state-dependence terms; the third is the naive association that the
instrument is meant to shrink. The same three panels for the simulated
panel:

```{r fig1, echo = FALSE, fig.cap = "Daily purchase probability by prior purchase frequency, purchase recency, and same-day ad exposures (the layout of Figure 1). Means with 95 percent confidence intervals.", fig.width = 9, fig.height = 3}
mf <- function(df, var, xlab, maxv) {
  df %>% filter(.data[[var]] <= maxv) %>% group_by(x = .data[[var]]) %>%
    summarise(p = mean(purchase), se = sqrt(p * (1 - p) / n()), .groups = "drop") %>%
    ggplot(aes(x, p)) + geom_errorbar(aes(ymin = p - 1.96 * se, ymax = p + 1.96 * se), width = 0.2, colour = "grey50") +
    geom_point() + geom_line() + scale_y_continuous(labels = scales::percent) + labs(x = xlab, y = "Daily purchase probability")
}
(mf(pan, "freq", "Purchase frequency", 10) + ggtitle("(a) Frequency")) +
  (mf(filter(pan, post == 1), "rec", "Purchase recency (days)", 30) + ggtitle("(b) Recency")) +
  (mf(pan, "A", "Same-day ad exposures", 3) + ggtitle("(c) Same-day exposures"))
```

Converted households buy every `r fmt(t1$Mean[5], 1)` days here against 11.3
in the paper, and `r fmt(t1$Mean[4], 1)` times over the window against 2.4,
because the post-conversion shift of 1.5 in the authors' utility is large
relative to the paper's baseline. Exposures per day (`r fmt(t1$Mean[3], 2)`)
are the authors' and about four times the paper's 0.14.

Panel (a) rises with frequency because of habit formation in the DGP
($\alpha_{2,1} = 0.45$, $\alpha_{2,2} = -0.06$). Panel (b) falls with recency
($\alpha_{2,3} = -0.10$, $\alpha_{2,4} = -0.03$); the paper's weekly spikes
are absent because the DGP has no day-of-week reordering. Panel (c) is
positive, and part of it is activity bias.

# Validating the instrument {#sec-iv}

## Figure 2: is the focal brand scheduled like everyone else?

**The quasi-random ordering assumption is checked by comparing airing-time
distributions.** If the network gave the focal brand a special slot, the
focal distribution would differ from the nonfocal one, and the density
estimated from all ads would mis-state expected exposure. The paper shows
MTV; we show the first network, whose density is trimodal by construction,
and the Kolmogorov-Smirnov test for every network (Online Appendix E.2).

```{r fig2, echo = FALSE, fig.cap = "Within-show airing-time distributions of focal and nonfocal ads on the first network (the layout of Figure 2). Left: densities. Right: cumulative distributions. Airing times are normalized by show duration.", fig.width = 8, fig.height = 3.2}
at1 <- filter(sim$ad_times, network == 1)
ks1 <- ks_by_network(at1)
p_pdf <- ggplot(at1, aes(time, colour = type)) + geom_line(stat = "density", adjust = 0.8, linewidth = 0.7) +
  labs(x = "Normalized airing time", y = "Density", colour = NULL)
p_cdf <- ggplot(at1, aes(time, colour = type)) + stat_ecdf(linewidth = 0.7) +
  labs(x = "Normalized airing time", y = "Cumulative distribution", colour = NULL)
(p_pdf + p_cdf) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r ks-table, echo = FALSE}
ks_all <- ks_by_network(sim$ad_times)
ks_all %>% mutate(D = fmt(D, 3), p = fmt(p, 3)) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Network", "Focal ads", "Nonfocal ads", "KS statistic", "p-value"),
      caption = "Kolmogorov-Smirnov tests of focal against nonfocal airing-time distributions by network (the check of Online Appendix E.2).") %>%
  tab_style()
```

The focal and nonfocal distributions coincide on every network because the
DGP draws both from the same density; the KS p-values are
`r paste(fmt(ks_all$p, 2), collapse = ", ")`. The paper reports 0.545 for
MTV. Section \@ref(sec-violations) shows what the test does when the
assumption fails.

## Figure 3: expected exposure as an area under the curve

**The paper's Figure 3 turns the definition of $P_{is}$ into a picture.**
The left panel is the network's density; the right panel shades the
household's viewing window and reports the area, which is the probability
that a focal ad on that network airs while the household is watching.

```{r fig3, echo = FALSE, fig.cap = "Airing-time density of the first network and the expected exposure of a household that watched part of a targeted show (the layout of Figure 3). The shaded area is the estimated probability of exposure, the dashed line the realized airing time of the focal ad.", fig.width = 8, fig.height = 3.2}
ex <- sim$show_sample %>% filter(network == 1, A_b == 1, len > 0.3, len < 0.7) %>% slice(1)
grid_x <- seq(0, 1, by = 0.005)
dens1 <- tibble(x = grid_x, d = sim$densities[[1]]$d(grid_x))
p_left <- ggplot(dens1, aes(x, d)) + geom_area(fill = "grey80", colour = "grey30") + labs(x = "Normalized show time", y = "Density of ad airing")
p_right <- ggplot(dens1, aes(x, d)) + geom_area(fill = "grey85", colour = "grey30") +
  geom_area(data = filter(dens1, x >= ex$start, x <= ex$end), fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = ex$T_focal, linetype = "dashed") +
  labs(x = "Normalized show time", y = "Density of ad airing") +
  annotate("label", x = 0.5, y = max(dens1$d) * 0.95, size = 3,
           label = paste0("View = [", fmt(ex$start, 2), ", ", fmt(ex$end, 2), "], P = ", fmt(ex$P_hat, 2), ", A = ", ex$A, ", IV = ", fmt(ex$IV, 2)))
p_left + p_right
```

The household in the right panel watched
`r fmt(100 * ex$len, 0)` percent of a targeted show and had an expected
exposure of `r fmt(ex$P_hat, 2)`; its realized exposure was `r ex$A`, so its
instrument is `r fmt(ex$IV, 2)`. The uniform-density shortcut would have
used `r fmt(ex$len, 2)`, the viewed share, instead.

## Table 2 and the falsification checks

**Table 2 of the paper reports the moments that Propositions 1 and 2
predict.** Realized and expected exposure should have the same mean, so
the instrument is mean zero; the instrument should correlate with realized
exposure and not with the targeting decision, the exposure probability, or
their product. Household-show rows are those in which a household watched
any part of a targeted show, as in the paper; our household-show
quantiles use a `r fmt(100 * 0.02, 0)` percent sample of those rows while the means, standard
deviations, and correlations use all of them.

```{r table2, echo = FALSE}
mom <- show_moments(sim$moments)
ss_rows <- sim$show_sample %>% filter(A_b == 1, len > 0) %>% mutate(A_exp = A_b * P_hat)
t2_show <- bind_rows(quantile_row(ss_rows$A, "Realized ad exposure ($A_{is}$)"),
                     quantile_row(ss_rows$A_exp, "Expected ad exposure ($A^b_s \\hat P_{is}$)"),
                     quantile_row(ss_rows$IV, "Proposed IV ($\\widehat{IV}_{is}$)"))
t2_show$Mean <- mom$mean[c("A", "A_exp", "IV")]; t2_show$SD <- mom$sd[c("A", "A_exp", "IV")]
t2_day <- bind_rows(quantile_row(pan$A, "Realized ad exposure ($A_{it}$)"),
                    quantile_row(pan$A_exp, "Expected ad exposure ($\\sum_s A^b_s \\hat P_{is}$)"),
                    quantile_row(pan$IV, "Proposed IV ($\\widehat{IV}_{it}$)"))
bind_rows(t2_show, t2_day) %>% mutate(across(-Variable, ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("", "Mean", "SD", "5\\%", "25\\%", "50\\%", "75\\%", "95\\%"),
      caption = "Summary statistics of realized ad exposure, expected ad exposure, and the instrument (the layout of Table 2).") %>%
  pack_rows("Household-show level", 1, 3) %>% pack_rows("Household-day level", 4, 6) %>% tab_style() %>%
  footnote(general = paste0("Based on ", fmt0(mom$n), " household-show observations in which the household watched part of a targeted show, and ", fmt0(nrow(pan)), " household-day observations."), general_title = "Note. ", footnote_as_chunk = TRUE, threeparttable = TRUE)
```

```{r falsification, echo = FALSE}
tt_show <- mom$mean["IV"] / (mom$sd["IV"] / sqrt(mom$n)); tt_day <- t.test(pan$IV)$statistic
fals <- tibble(
  Check = c("Mean of the IV (t statistic)", "corr(IV, realized exposure)", "corr(IV, targeting decision $A^b_s$)",
            "corr(IV, exposure probability $\\hat P_{is}$)", "corr(IV, expected exposure $A^b_s \\hat P_{is}$)", "corr(IV, $u_{endo}$), unobservable in practice"),
  `Household-show` = c(paste0(fmt(mom$mean["IV"], 4), " (", fmt(tt_show, 2), ")"), fmt(mom$corr["IV", "A"], 3), fmt(mom$corr_iv_target_all, 4),
                       fmt(mom$corr["IV", "P_hat"], 4), fmt(mom$corr["IV", "A_exp"], 4), fmt(cor(sim$show_sample$IV, sim$show_sample$u_endo), 4)),
  `Household-day` = c(paste0(fmt(mean(pan$IV), 4), " (", fmt(tt_day, 2), ")"), fmt(cor(pan$IV, pan$A), 3), fmt(cor(pan$IV, pan$n_target), 4),
                      "---", fmt(cor(pan$IV, pan$A_exp), 4), fmt(cor(pan$IV, pan$u_endo), 4)),
  Paper = c("$-0.002$; $-0.004$", "0.42; 0.36", "$-0.001$; $-0.006$", "$-0.001$; $-0.007$", "$-0.009$; $-0.009$", "not observable")
)
kbl(fals, booktabs = TRUE, linesep = "", escape = FALSE, caption = "Relevance and falsification checks of the instrument (Propositions 1 and 2), with the paper's values.") %>% tab_style()
```

**Reading the checks.** Realized and expected exposure have the same mean
to three decimals, so the instrument is mean zero; it correlates with
realized exposure at `r fmt(cor(pan$IV, pan$A), 2)` on the day level
(0.36 in the paper) and with nothing endogenous. The last row cannot be
computed with real data and is the reason the design works: the
instrument does not move with the demand shock even though exposure does
(corr($A_{it}$, $u_{endo}$) = `r fmt(cor(pan$A, pan$u_endo), 2)`).

# First stage {#sec-first-stage}

**Table 3 shows that the instrument moves exposure one for one and that
purchase history does not, once the instrument is in the regression.** The
coefficient on the instrument is close to one because
$A_{it} = \text{expected} + IV_{it}$ and the expected part is absorbed by
the controls only in part. `cf_residuals()` runs this regression and
returns the residual and the strength statistics.

```{r table3, echo = FALSE}
spec4 <- col_spec(lambda_A = 0.7, cf = TRUE, lambda_CF = 0.95, interactions = FALSE, traits = TRUE, trait_slopes = FALSE)
aug4 <- first_stage(pan, spec4)
fs_fit <- attr(aug4, "first_stage")$fit
fs_strength <- attr(aug4, "first_stage")$strength
fs_rob <- lm_robust(formula(fs_fit), data = pan, clusters = hh, se_type = "stata")
ar_set <- iv_ar_confidence_set(pan_sub, y = "purchase", d = "A", z = "IV", x = c(controls_base, controls_traits), cluster = "hh",
                               theta_grid = seq(-0.02, 0.04, length.out = 601))
fs_terms <- c("(Intercept)", "IV", "post", "lf", "lf2", "lr", "lr2", "promo1m", "comp", "z_view", "z_sports")
fs_labels <- c("Intercept", "$\\widehat{IV}_{it}$", "Post conversion", "Frequency (log)", "Frequency (log) squared", "Recency (log)", "Recency (log) squared",
               "First month post-conversion promotion", "Competitor ad spend", "Average TV viewing", "Sports viewing")
fs_params <- c("$\\alpha^f_0$", "$\\phi^f$", "$\\alpha^f_{2,0}$", "$\\alpha^f_{2,1}$", "$\\alpha^f_{2,2}$", "$\\alpha^f_{2,3}$", "$\\alpha^f_{2,4}$", "$\\gamma^f_1$", "$\\gamma^f_2$", "$\\alpha^f_1$", "$\\alpha^f_1$")
tibble(`Model component` = fs_labels, Parameter = fs_params,
       Estimate = est_star(coef(fs_rob)[fs_terms], fs_rob$std.error[fs_terms], 4), `Standard error` = fmt(fs_rob$std.error[fs_terms], 4)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "Parameter estimates from the first-stage model (the layout of Table 3). Standard errors clustered by household.") %>%
  tab_style() %>% footnote(general = paste0("The DV is daily household focal brand ad exposures. First-stage F = ", fmt0(fs_strength$F), " (robust F = ", fmt0(fs_strength$F_robust), "). ***p < 0.01; **p < 0.05; *p < 0.1."), general_title = "Note. ", footnote_as_chunk = TRUE, escape = FALSE, threeparttable = TRUE)
```

The first-stage F is `r fmt0(fs_strength$F)` (the paper: 1,902,996), so
weak-instrument inference is not a concern here; on a 5,000-household
subsample the Anderson-Rubin 95 percent set for the linear-probability
effect of exposure,
[`r fmt(ar_set$intervals$lower, 4)`, `r fmt(ar_set$intervals$upper, 4)`],
equals the Wald interval
[`r fmt(ar_set$wald["lower"], 4)`, `r fmt(ar_set$wald["upper"], 4)`]. The
purchase-history terms do not predict exposure given the instrument, as
in the paper's Table 3, because the DGP routes all endogeneity through the
unobservable rather than through the observed states.

# Second stage {#sec-second-stage}

## Table 4: six columns, with and without the control function

**The paper builds the model up in three steps and adds the control
function at each step.** Columns 1 and 2 have no carryover; columns 3 and
4 add the two decays; columns 5 and 6 add the moderation of ad
responsiveness by purchase history and by household traits and the
household random effects. Comparing each pair isolates what the control
function does at that level of richness. Columns 5 and 6 are estimated
with a household random intercept and random slope on the ad stock through
`lme4::glmer()` (Laplace approximation, `nAGQ = 0`) instead of the
paper's simulated maximum likelihood. Bootstrap standard errors use
`r n_boot` household draws with duplicates kept; for column 6 the
coefficient rows are bootstrapped without the random effects (each
random-effects fit takes about twenty minutes at this size) and the three
random-effect rows on the 5,000-household subsample with 20 draws.

```{r table4-fits}
specs <- list(
  c1 = col_spec(0,   FALSE, 0,    FALSE, TRUE, FALSE),
  c2 = col_spec(0,   TRUE,  0,    FALSE, TRUE, FALSE),
  c3 = col_spec(0.7, FALSE, 0,    FALSE, TRUE, FALSE),
  c4 = col_spec(0.7, TRUE,  0.95, FALSE, TRUE, FALSE),
  c5 = col_spec(0.7, FALSE, 0,    TRUE,  TRUE, TRUE, random = TRUE),
  c6 = col_spec(0.7, TRUE,  0.95, TRUE,  TRUE, TRUE, random = TRUE)
)
fits_file <- sim_file("table4_fits")
if (file.exists(fits_file)) {
  t4 <- readRDS(fits_file)
} else {
  t4 <- lapply(specs, function(s) {
    f <- fit_column(pan, s)
    se <- if (inherits(f$fit, "merMod")) sqrt(diag(as.matrix(vcov(f$fit)))) else summary(f$fit)$coefficients[, 2]
    list(coef = coef_vector(f$fit), se = se)
  })
  saveRDS(t4, fits_file)
}
sapply(t4, function(x) x$coef[c("A_stock", "CF_stock")])
```

```{r table4-boot}
boot4_file <- sim_file(paste0("table4_boot_", n_boot))
if (file.exists(boot4_file)) {
  b4 <- readRDS(boot4_file)
} else {
  boot_cols <- c("hh", "t", "purchase", "A", "IV", controls_base, controls_traits, "freq", "rec")
  b4 <- list(
    c2 = cluster_bootstrap(pan[, boot_cols], specs$c2, n_boot = n_boot, seed = 1107, workers = n_threads_boot),
    c4 = cluster_bootstrap(pan[, boot_cols], specs$c4, n_boot = n_boot, seed = 1107, workers = n_threads_boot),
    c6 = cluster_bootstrap(pan[, boot_cols], modifyList(specs$c6, list(random = FALSE)), n_boot = n_boot, seed = 1107,
                           workers = n_threads_boot),
    c6_re = cluster_bootstrap(pan_sub[, boot_cols], specs$c6, n_boot = min(n_boot, 20L), seed = 1107,
                              workers = n_threads_boot)
  )
  b4$c6$se[c("sigma1", "sigma2", "rho")] <- b4$c6_re$se[c("sigma1", "sigma2", "rho")]
  saveRDS(b4, boot4_file)
}
sapply(b4[c("c2", "c4", "c6")], function(b) c(b$se[c("A_stock", "CF_stock")], failed = b$n_failed))
```

```{r table4, echo = FALSE}
rows4 <- c("(Intercept)", "A_stock", "CF_stock", "post", "lf", "lf2", "lr", "lr2", "A_stock:post", "A_stock:lf", "A_stock:lf2", "A_stock:lr", "A_stock:lr2",
           "promo1m", "comp", "z_view", "z_sports", "A_stock:z_view", "A_stock:z_sports", "sigma1", "sigma2", "rho")
truth4 <- c(-2.5, 0.012, NA, 1.5, 0.45, -0.06, -0.1, -0.03, 0, 0.02, -0.01, 0, 0.005, 0.2, 0.02, sim$par$a_view, sim$par$a_sports, sim$par$b_view, sim$par$b_sports, sim$par$sigma1, sim$par$sigma2, sim$par$rho)
col_pair <- function(j) {
  cf <- t4[[j]]$coef; se <- if (j %in% names(b4)) b4[[j]]$se else t4[[j]]$se
  b <- cf[rows4]; s <- se[rows4]
  est <- ifelse(is.na(b), "---", ifelse(rows4 %in% c("sigma1", "sigma2", "rho"), fmt(b, 3), est_star(b, s, 3)))
  se_txt <- ifelse(is.na(b) | is.na(s), "---", fmt(s, 3))
  tibble(est, se_txt)
}
t4_tab <- bind_cols(tibble(Parameter = term_labels[rows4], Symbol = param_labels[rows4], True = ifelse(is.na(truth4), "---", fmt(truth4, 3))),
                    setNames(bind_cols(lapply(names(specs), col_pair)), paste0(rep(c("est", "se"), 6), rep(1:6, each = 2))))
extra_rows <- tibble(Parameter = c("Carryover $\\lambda_A / \\lambda_{CF}$", "Household characteristics", "HH characteristics x ad stock"),
                     Symbol = c("", "$\\alpha_1$", "$\\beta_1$"), True = c("0.7 / 0.95", "Yes", "Yes"),
                     est1 = c("0 / 0", "Yes", "---"), se1 = "", est2 = c("0 / 0", "Yes", "---"), se2 = "", est3 = c("0.7 / 0", "Yes", "---"), se3 = "",
                     est4 = c("0.7 / 0.95", "Yes", "---"), se4 = "", est5 = c("0.7 / 0", "Yes", "Yes"), se5 = "", est6 = c("0.7 / 0.95", "Yes", "Yes"), se6 = "")
t4_out <- bind_rows(t4_tab[1:15, ], extra_rows, t4_tab[16:22, ])
kbl(t4_out, booktabs = TRUE, linesep = "", escape = FALSE, align = c("l", "c", "r", rep("r", 12)),
    col.names = c("", "", "True", rep(c("Est.", "SE"), 6)),
    caption = paste0("Second-stage estimation results for the ad response model (the layout of Table 4), simulated panel of ", fmt0(n_hh), " households. Columns 1 and 2 assume no carryover, no evolution of ad responsiveness, and no unobserved heterogeneity; columns 3 and 4 add carryover; columns 5 and 6 add moderation by purchase history and household traits and household random effects. Columns 2, 4, and 6 use the control function; their standard errors come from ", n_boot, " household bootstrap draws. ***p < 0.01; **p < 0.05; *p < 0.1.")) %>%
  add_header_above(c(" " = 3, "No carryover or moderation, without CF (1)" = 2, "No carryover or moderation, with CF (2)" = 2, "No moderation, without CF (3)" = 2,
                     "No moderation, with CF (4)" = 2, "Full model, without CF (5)" = 2, "Full model, with CF (6)" = 2)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 7) %>% landscape()
```

```{r table4-numbers, echo = FALSE}
b0 <- sapply(t4, function(x) unname(x$coef["A_stock"])); d0 <- sapply(t4, function(x) unname(x$coef["CF_stock"]))
```

**Reading the table against the paper.** In the paper the ad-stock
coefficient falls when the control function enters at every level (0.023
to 0.010, 0.016 to 0.007, 0.018 to 0.012), and $\delta$ is positive and
significant throughout. Here the moves are `r fmt(b0["c1"], 4)` to
`r fmt(b0["c2"], 4)`, `r fmt(b0["c3"], 4)` to `r fmt(b0["c4"], 4)`, and
`r fmt(b0["c5"], 4)` to `r fmt(b0["c6"], 4)` against a truth of 0.012 in
the full model, with $\delta$ equal to `r fmt(d0["c2"], 4)`,
`r fmt(d0["c4"], 4)`, and `r fmt(d0["c6"], 4)`. Columns 1 to 4 have no
single true value: they fit one coefficient to an effect that the DGP
makes vary with purchase history, so their $\beta_0$ is an average over
states. The random-effect rows return $\sigma_1 = `r fmt(t4$c6$coef["sigma1"], 3)`$
(DGP 0.2), $\sigma_2 = `r fmt(t4$c6$coef["sigma2"], 3)`$ (0.01) and
$\rho = `r fmt(t4$c6$coef["rho"], 2)`$ ($-0.3$); the slope variance and the
correlation have subsample bootstrap standard errors as large as the
estimates, as in the paper, where neither is significant.

## Elasticities: the headline number

**The paper converts coefficients into elasticities to compare with the
advertising literature.** We define the same-day elasticity as the percent
change in purchases when every household's same-day exposures rise by one
percent, $E[\partial P_{it}/\partial AS_{it} \cdot A_{it}] / E[P_{it}]$,
with the derivative from `cf_ape()`, and the 30-day elasticity as the
same-day value times $\sum_{l=0}^{29} \lambda_A^l$, the cumulative
carryover of one day's exposures.

```{r elasticities}
f5 <- fit_column(pan, modifyList(specs$c5, list(random = FALSE)))
f6 <- fit_column(pan, modifyList(specs$c6, list(random = FALSE)))
el <- rbind(naive = elasticities(f5$fit, f5$data), corrected = elasticities(f6$fit, f6$data))
true_ape <- mean(dnorm(pan$index) * pan$beta_it)
el
```

```{r elasticity-table, echo = FALSE}
tibble(Model = c("Naive (column 5, fixed effects)", "Control function (column 6, fixed effects)", "Truth (DGP)"),
       `Average partial effect of the ad stock` = fmt(c(el[, "ape"], true_ape), 5),
       `Same-day elasticity` = c(fmt(el[, "same_day"], 3), "---"), `30-day elasticity` = c(fmt(el[, "long_run"], 3), "---")) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Average partial effects and elasticities, naive against control function. The paper reports 0.072 and 0.222 without correction and 0.045 and 0.143 with it.") %>% tab_style()
```

The naive same-day elasticity overstates the corrected one by
`r fmt(100 * (el["naive", "same_day"] / el["corrected", "same_day"] - 1), 0)` percent (55 percent in
the paper). The corrected average partial effect,
`r fmt(el["corrected", "ape"], 5)`, is within sampling error of the DGP's
`r fmt(true_ape, 5)`.

# Ad responsiveness and purchase history {#sec-hte}

**Figure 4 is the paper's substantive result: who responds to TV ads.**
It plots $\beta_{it}$ implied by the column-6 estimates against frequency
at fixed recency and against recency at fixed frequency. The bands come
from the bootstrap draws.

```{r fig4, echo = FALSE, fig.cap = "Ad responsiveness by purchase frequency and recency (the layout of Figure 4). Solid: control function (column 6) with 95 percent bootstrap bands; dashed: naive (column 5); dotted: the DGP.", fig.width = 8, fig.height = 3.4}
truth_cf <- c(A_stock = 0.012, `A_stock:post` = 0, `A_stock:lf` = 0.02, `A_stock:lf2` = -0.01, `A_stock:lr` = 0, `A_stock:lr2` = 0.005)
curve_df <- function(x, freq, rec, label) {
  cf6 <- as.list(t4$c6$coef); cf5 <- as.list(t4$c5$coef)
  draws <- b4$c6$draws
  band <- apply(draws, 1, function(r) beta_curve(as.list(r), freq, rec))
  tibble(x = x, cf = beta_curve(cf6, freq, rec), naive = beta_curve(cf5, freq, rec), truth = beta_curve(as.list(truth_cf), freq, rec),
         lo = apply(band, 1, quantile, 0.025), hi = apply(band, 1, quantile, 0.975), panel = label)
}
cd <- bind_rows(curve_df(1:10, 1:10, 7, "(a) Purchase frequency (recency = 7 days)"), curve_df(1:30, 1, 1:30, "(b) Purchase recency (frequency = 1)"))
ggplot(cd, aes(x)) + geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = cf), colour = "steelblue4", linewidth = 0.8) + geom_line(aes(y = naive), linetype = "dashed") + geom_line(aes(y = truth), linetype = "dotted") +
  facet_wrap(~ panel, scales = "free_x") + labs(x = NULL, y = expression(beta[it]))
```

```{r targeting-table, echo = FALSE}
p_state <- function(freq, rec) pnorm(alpha_curve(as.list(t4$c6$coef), freq, rec))
states <- tibble(State = c("One prior purchase, 7 days ago", "Two prior purchases, 7 days ago", "Five prior purchases, 7 days ago", "One prior purchase, 1 day ago", "One prior purchase, 14 days ago", "One prior purchase, 28 days ago"),
                 freq = c(1, 2, 5, 1, 1, 1), rec = c(7, 7, 7, 1, 14, 28))
states <- states %>% mutate(beta = beta_curve(as.list(t4$c6$coef), freq, rec), base = p_state(freq, rec),
                            elasticity = dnorm(qnorm(base)) * beta * mean(pan$A[pan$post == 1]) / base)
states %>% transmute(State, `Baseline daily purchase probability` = paste0(fmt(100 * base, 2), "\\%"), `Ad responsiveness $\\beta_{it}$` = fmt(beta, 4), `Same-day elasticity` = fmt(elasticity, 3)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "Baseline propensity and ad responsiveness by purchase state, column 6 estimates. The paper reports same-day elasticities of 0.040, 0.056, and 0.046 after one, two, and five purchases, and 0.040, 0.027, and 0.078 one day, one week, and four weeks after the last purchase.") %>% tab_style()
```

**What the curves say.** The DGP puts an inverted U in frequency
($\beta_{2,1} = 0.02$, $\beta_{2,2} = -0.01$) and a U in recency
($\beta_{2,4} = 0.005$), the paper's two shapes. The control-function curve
tracks the DGP within the band; the naive curve sits above it where the
bias concentrates. The table is the paper's targeting argument in numbers:
lapsed households have a low baseline propensity and a high
responsiveness, so they are worth an ad but only if it comes before the
recency trap closes.

## Table 5: observed household heterogeneity

**Table 5 of the paper reports how the traits shift baseline propensity
($\alpha_1$) and ad responsiveness ($\beta_1$).** Heavy TV viewers order
less and respond less; sports viewers order more and respond more. The DGP
puts the same signs in with the same order of magnitude.

```{r table5, echo = FALSE}
t5_terms <- list(view = c("z_view", "A_stock:z_view"), sports = c("z_sports", "A_stock:z_sports"))
t5_truth <- c(z_view = sim$par$a_view, z_sports = sim$par$a_sports, `A_stock:z_view` = sim$par$b_view, `A_stock:z_sports` = sim$par$b_sports)
cf6 <- t4$c6$coef; se6 <- b4$c6$se
tibble(Trait = c("Average TV viewing", "Sports viewing"),
       `True` = fmt(t5_truth[c("z_view", "z_sports")], 3), Estimate = est_star(cf6[c("z_view", "z_sports")], se6[c("z_view", "z_sports")], 3), `Standard error` = fmt(se6[c("z_view", "z_sports")], 3),
       `True ` = fmt(t5_truth[c("A_stock:z_view", "A_stock:z_sports")], 3), `Estimate ` = est_star(cf6[c("A_stock:z_view", "A_stock:z_sports")], se6[c("A_stock:z_view", "A_stock:z_sports")], 3),
       `Standard error ` = fmt(se6[c("A_stock:z_view", "A_stock:z_sports")], 3)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = paste0("Parameter estimates for observed household heterogeneity (the layout of Table 5), column 6. Standard errors from ", n_boot, " bootstrap draws. ***p < 0.01; **p < 0.05; *p < 0.1.")) %>%
  add_header_above(c(" " = 1, "Baseline purchase propensity ($\\\\alpha_1$)" = 3, "Ad responsiveness ($\\\\beta_1$)" = 3), escape = FALSE) %>% tab_style()
```

The paper's traits are nine viewing shares calibrated on the month before
the study; the DGP has two. Their coefficients are recovered with the
signs and sizes put in, the responsiveness shifts less precisely because
they multiply a stock whose effect is small.

# The robustness checks the paper runs {#sec-robust}

## The two decay rates (footnote 17)

**The paper picks $\lambda_A$ and $\lambda_{CF}$ separately by out-of-sample
fit.** It estimates on the first 120 days, scores the log-likelihood on
the remaining 13, and searches a grid. The two decays need not agree: the
ad effect decays at the rate at which exposures are forgotten, the control
function at the rate at which the confounder persists. The DGP sets 0.7
and 0.95. We run the grid on a subsample of 5,000 households.

```{r decay-grid}
grid_res <- file_cached("decay_grid", decay_grid(pan_sub, grid = c(0, 0.3, 0.5, 0.7, 0.8, 0.9, 0.95), train_days = n_days - 30,
                                                 spec = col_spec(interactions = FALSE, traits = TRUE, trait_slopes = FALSE)))
grid_res[which.max(grid_res$loglik), ]
```

```{r fig-grid, echo = FALSE, fig.cap = "Out-of-sample log-likelihood over the grid of decay rates (the search of footnote 17); the cross marks the maximum, the DGP uses 0.7 and 0.95.", fig.width = 5, fig.height = 3.6}
best <- grid_res[which.max(grid_res$loglik), ]
ggplot(grid_res, aes(factor(lambda_A), factor(lambda_CF), fill = loglik - max(loglik))) + geom_tile() +
  geom_point(data = best, shape = 4, size = 4) + scale_fill_viridis_c(name = "Log-likelihood\nminus maximum") +
  labs(x = expression(lambda[A]), y = expression(lambda[CF]))
```

The maximum is at $\lambda_A = `r best$lambda_A`$, $\lambda_{CF} = `r best$lambda_CF`$,
against the DGP's 0.7 and 0.95, and the whole grid spans
`r fmt(max(grid_res$loglik) - min(grid_res$loglik), 1)` log-likelihood points on
`r fmt0(sum(pan_sub$t > n_days - 30))` held-out observations. The criterion
is flat near the top: a slower ad stock also proxies the slow-moving
confounder when the specification omits the moderation terms, so the
out-of-sample fit trades the two decays against each other. The paper's
search faces the same weak identification, which is why it fixes the
pair before comparing specifications rather than re-estimating it in each
column.

## Collinearity of the two stocks (footnote 18)

**A stock and its control are built from the same daily series, so the
paper reports variance inflation factors.** Theirs are 2.42 and 2.27.

```{r vif}
f4 <- fit_column(pan, specs$c4)
v <- car::vif(f4$fit)
c(cor_AS_CFS = cor(f4$data$A_stock, f4$data$CF_stock), vif_AS = unname(v["A_stock"]), vif_CFS = unname(v["CF_stock"]))
```

The two stocks correlate at `r fmt(cor(f4$data$A_stock, f4$data$CF_stock), 2)`
(0.72 in the paper) and the variance inflation factors are
`r fmt(v["A_stock"], 2)` and `r fmt(v["CF_stock"], 2)`, below the paper's and
far from the threshold of 10 that footnote 18 cites.

## Placebo: another brand's ads (footnote 19, Online Appendix E.4)

**If viewing is the confounder, any brand's exposures will predict
purchases in a naive model.** The paper replaces the focal brand's ads
with a car maker's and finds no effect after correction. The simulated
second brand reaches households through the same viewing process and has
no effect on purchases.

```{r placebo}
pan_pl <- pan; pan_pl$A <- pan_pl$A_pl; pan_pl$IV <- pan_pl$IV_pl
pl_naive <- fit_column(pan_pl, specs$c3); pl_cf <- fit_column(pan_pl, specs$c4)
pl <- rbind(naive = summary(pl_naive$fit)$coefficients["A_stock", 1:2],
            control_function = summary(pl_cf$fit)$coefficients["A_stock", 1:2])
pl
```

The naive coefficient on the placebo stock is `r fmt(pl["naive", 1], 4)`
(standard error `r fmt(pl["naive", 2], 4)`); with the control function it
is `r fmt(pl["control_function", 1], 4)` (standard error
`r fmt(pl["control_function", 2], 4)`). Activity bias alone produces an
"effect" of ads that do not work, and the control function removes
`r fmt(100 * (1 - pl["control_function", 1] / pl["naive", 1]), 0)` percent of it. What
remains is the approximation in the method: the confounder enters exposure
through two nonlinear decisions, and a linear function of the first-stage
residual captures most but not all of it.

## Functional form of the history terms (Online Appendix E.1)

**The paper compares four ways to let propensity and responsiveness bend
with frequency and recency.** Log plus squared log wins on AIC and BIC in
their data; it is also the DGP's form here.

```{r fform}
aug6 <- f6$data
base <- "purchase ~ A_stock + CF_stock + post + promo1m + comp + z_view + z_sports + A_stock:post"
forms <- list(
  linear = paste(base, "+ freq + rec + A_stock:freq + A_stock:rec"),
  quadratic = paste(base, "+ freq + I(freq^2) + rec + I(rec^2) + A_stock:freq + A_stock:I(freq^2)",
                    "+ A_stock:rec + A_stock:I(rec^2)"),
  log = paste(base, "+ lf + lr + A_stock:lf + A_stock:lr"),
  log_squared = paste(base, "+ lf + lf2 + lr + lr2 + A_stock:lf + A_stock:lf2 + A_stock:lr + A_stock:lr2")
)
ff <- file_cached("functional_forms", t(sapply(forms, function(f) {
  m <- glm(as.formula(f), data = aug6, family = binomial("probit"))
  c(loglik = as.numeric(logLik(m)), AIC = AIC(m), BIC = BIC(m), beta0 = unname(coef(m)["A_stock"]))
})))
cbind(round(ff[, 1:3], 1), beta0 = round(ff[, 4], 4))
```

Log plus squared log has the lowest AIC and BIC here as well, by
`r fmt0(ff["log", "BIC"] - ff["log_squared", "BIC"])` BIC points over the
log specification; the ad-stock coefficient is the same to three decimals
across the four forms.

# Extensions {#sec-extensions}

## The instrument is a recentered instrument {#sec-recentering}

**Realized minus expected treatment under a known assignment process is
the recentering of Borusyak and Hull (2023).** Their point is that a
shock-driven instrument is valid only after subtracting its expectation
under the assignment process, because the expectation carries the
nonrandom exposure. The paper's instrument is that construction with the
network's ad-timing density as the process. The consequence is that the
density must be right: Table \@ref(tab:three-ivs) compares the paper's
instrument, a version that assumes uniform timing (expected exposure
equals the viewed share), and the scripts' oracle.

```{r three-ivs, echo = FALSE}
iv_variants <- c(`Paper: empirical density` = "IV", `Uniform timing (viewed share)` = "IV_unif", `Oracle: targeting probabilities` = "IV_oracle")
three <- file_cached("three_ivs", bind_rows(lapply(names(iv_variants), function(nm) {
  iv <- iv_variants[[nm]]
  f <- fit_column(pan, modifyList(specs$c6, list(iv = iv, random = FALSE)))
  st <- attr(f$data, "first_stage")$strength
  tibble(Instrument = nm, `Mean (t)` = paste0(fmt(mean(pan[[iv]]), 4), " (", fmt(t.test(pan[[iv]])$statistic, 1), ")"),
         `corr with $A$` = fmt(cor(pan[[iv]], pan$A), 3), `corr with expected exposure` = fmt(cor(pan[[iv]], pan$A_exp), 4),
         `corr with $u_{endo}$` = fmt(cor(pan[[iv]], pan$u_endo), 4), `First-stage F` = fmt0(st$F),
         `$\\beta_0$` = fmt(coef(f$fit)["A_stock"], 4), `$\\delta$` = fmt(coef(f$fit)["CF_stock"], 4))
})))
kbl(three, booktabs = TRUE, linesep = "", escape = FALSE, caption = "Three instruments on the show-level panel: the paper's, one that assumes uniform within-show timing, and the scripts' oracle. Full model (column 6 without random effects). True $\\beta_0 = 0.012$.") %>% tab_style("scale_down")
```

The uniform version fails the falsification checks: its mean is not zero
(t statistic of `r three[["Mean (t)"]][2]`) and it correlates with expected
exposure at `r three[["corr with expected exposure"]][2]` and with the
unobservable at `r three[["corr with $u_{endo}$"]][2]`, because the viewed
share is not the exposure probability when the density is trimodal. Its
second stage lands at `r three[["$\\beta_0$"]][2]` against
`r three[["$\\beta_0$"]][1]` for the paper's instrument, so at this size the
leaked correlation is small enough for the control function to absorb;
the check fails before the estimate does, which is the point of having a
check. The oracle is valid but uses the targeting probabilities that no
analyst observes.

## What the diagnostics catch when the assumptions fail {#sec-violations}

**Each identifying assumption has a diagnostic, and the Monte Carlo asks
whether the diagnostic fires before the estimate is wrong.** Strategic
viewership: partial viewers with a positive demand shock skip the focal ad
with probability $\pi$. Nonrandom ordering: the network puts the focal ad
in the first pod with probability $\pi$. Small panels
(`r fmt0(2000)` households, 90 days), `r n_sims` replications each.

```{r violation-mc}
mc_grid <- expand.grid(type = c("skip", "order"), pi = c(0, 0.25, 0.5), stringsAsFactors = FALSE)
mc_grid <- mc_grid[!(mc_grid$pi == 0 & mc_grid$type == "order"), ]
mc <- file_cached(paste0("violation_mc_", n_sims), bind_rows(lapply(seq_len(nrow(mc_grid)), function(i) {
  violation_mc(n_sims = n_sims, n_hh = 2000, n_days = 90, type = if (mc_grid$pi[i] == 0) "none" else mc_grid$type[i],
               pi = mc_grid$pi[i], seed = i, workers = n_threads)
})))
mc_sum <- mc %>% mutate(type = ifelse(pi == 0, "none", type)) %>% group_by(type, pi) %>%
  summarise(n = n(), sd_cf = sd(beta_cf), beta_naive = mean(beta_naive), beta_cf = mean(beta_cf),
            ape_naive = mean(ape_naive), ape_cf = mean(ape_cf), ape_true = mean(ape_true), delta = mean(delta),
            iv_mean = mean(iv_mean), iv_t = mean(iv_t), corr_iv_exp = mean(corr_iv_exp),
            ks_reject = mean(ks_p_min < 0.05 / 3), .groups = "drop")
mc_sum
```

```{r mc-table, echo = FALSE}
mc_sum %>% mutate(Violation = c(none = "None", skip = "Strategic viewership", order = "Nonrandom ordering")[type]) %>%
  transmute(Violation, `$\\pi$` = fmt(pi, 2), `Naive $\\beta_0$` = fmt(beta_naive, 4), `CF $\\beta_0$` = fmt(beta_cf, 4),
            `SD of CF $\\beta_0$` = ifelse(is.na(sd_cf), "---", fmt(sd_cf, 4)),
            `APE naive / CF / true ($\\times 10^4$)` = paste(fmt(1e4 * ape_naive, 2), fmt(1e4 * ape_cf, 2), fmt(1e4 * ape_true, 2), sep = " / "),
            `IV mean (t)` = paste0(fmt(iv_mean, 4), " (", fmt(iv_t, 1), ")"), `corr(IV, expected)` = fmt(corr_iv_exp, 3), `KS rejects (share)` = fmt(ks_reject, 2)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "Monte Carlo under violations of the identifying assumptions: estimates of the ad-stock coefficient (full model without traits; true value 0.012), average partial effects, and the diagnostics the paper proposes. Means over replications; the KS test rejects at 5 percent with a Bonferroni correction over the three networks.") %>% tab_style("scale_down")
```

```{r mc-numbers, echo = FALSE}
mc_get <- function(ty, p, col) mc_sum[[col]][mc_sum$type == ty & mc_sum$pi == p]
```

**Reading the table.** Without a violation the corrected coefficient
averages `r fmt(mc_get("none", 0, "beta_cf"), 4)` (truth 0.012, standard
deviation across replications `r fmt(mc_get("none", 0, "sd_cf"), 4)`) and the
naive one `r fmt(mc_get("none", 0, "beta_naive"), 4)`. Under strategic
skipping, exposure falls short of its expectation for high-demand
households, so the instrument's mean turns negative with a t statistic of
`r fmt(mc_get("skip", 0.5, "iv_t"), 0)` at $\pi = 0.5$; the KS test cannot see
it because the network did nothing. Under nonrandom ordering the KS test
rejects in every replication and the mean-zero check also fails, because
the density estimated from other brands no longer describes the focal ad.
The corrected coefficient moves to `r fmt(mc_get("skip", 0.5, "beta_cf"), 4)`
under skipping and `r fmt(mc_get("order", 0.5, "beta_cf"), 4)` under
reordering at $\pi = 0.5$. This is why the paper reports both checks and
not one.

**A direct effect of the instrument is the violation that no check
detects.** If the shifter affected purchases other than through exposure,
the instrument would still be mean zero. Conley, Hansen, and Rossi (2012)
bound the estimate under such a violation of size $\gamma$;
`iv_plausibly_exogenous()` does this for the linear-probability version.

```{r plausibly}
pe <- iv_plausibly_exogenous(pan_sub, y = "purchase", d = "A", z = "IV", x = c(controls_base, controls_traits),
                             gamma_grid = c(-0.002, -0.001, 0, 0.001, 0.002), cluster = "hh")
pe
```

A direct effect of one exposure's shifter on the purchase probability of
$\pm 0.002$, the size of the linear-probability estimate itself, widens
the interval from
[`r fmt(attr(pe, "gamma_zero")[1], 4)`, `r fmt(attr(pe, "gamma_zero")[2], 4)`]
to [`r fmt(attr(pe, "union")[1], 4)`, `r fmt(attr(pe, "union")[2], 4)`]. The
mean-zero check cannot rule this violation out, only an argument about the
mechanism can.

## Estimators in probability units {#sec-estimators}

**The paper's control-function probit is one of three ways to use the
instrument.** Two-stage least squares on the linear-probability model
instruments the ad stock with the instrument stock; the partially linear
IV model of double machine learning does the same with flexible nuisance
functions of the states and traits (`est_dml(model = "pliv")`, forests on
a subsample); the control function gives an average partial effect through
`cf_ape()`. All three are in probability units and can be compared with
the DGP.

```{r estimators}
xs <- c(controls_base, controls_traits)
xs_sum <- paste(xs, collapse = " + ")
est_cache <- file_cached("estimators", {
  tsls <- iv_robust(as.formula(paste("purchase ~ A_stock +", xs_sum, "| IV_stock +", xs_sum)), data = pan,
                    clusters = hh, se_type = "stata")
  sub2 <- pan[pan$hh %in% sub_hh[seq_len(2000L)], ]
  rf <- lrn("regr.ranger", num.trees = 200, min.node.size = 50, num.threads = n_threads)
  dml <- est_dml(sub2, y = "purchase", d = "A_stock", x = xs, z = "IV_stock", model = "pliv",
                 learner_l = rf, learner_m = rf, learner_z = rf, folds = 2, seed = 1)
  list(tsls_est = unname(coef(tsls)["A_stock"]), tsls_se = unname(tsls$std.error["A_stock"]),
       dml_est = dml$estimate, dml_se = dml$std.error)
})
ape_cf <- cf_ape(f6$fit, f6$data, d = "A_stock")
se_ape <- b4$c6$se["A_stock"] * abs(as.numeric(ape_cf) / unname(t4$c6$coef["A_stock"]))
est_tab <- tibble(
  Estimator = c("2SLS, linear probability (household clusters)", "DML partially linear IV, forests (2,000 households)",
                "Control-function probit APE (column 6)", "Truth (DGP)"),
  Estimate = c(est_cache$tsls_est, est_cache$dml_est, as.numeric(ape_cf), true_ape),
  SE = c(est_cache$tsls_se, est_cache$dml_se, se_ape, NA)
)
est_tab
```

```{r estimators-table, echo = FALSE}
est_tab %>% mutate(Estimate = fmt(Estimate, 5), SE = ifelse(is.na(SE), "---", fmt(SE, 5))) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Average effect of one unit of ad stock on the daily purchase probability, three estimators against the DGP. The control-function standard error scales the bootstrap standard error of the coefficient.") %>% tab_style()
```

```{r estimator-numbers, echo = FALSE}
me_post <- tapply(dnorm(pan$index) * pan$beta_it, pan$post, mean)
share_pre <- mean(pan$post == 0)
```

**The linear estimators do not estimate the average partial effect
here.** Two-stage least squares lands at `r fmt(est_tab$Estimate[1], 5)`,
the marginal effect of a household before conversion
(`r fmt(me_post["0"], 5)` on average), not the DGP's `r fmt(true_ape, 5)`.
With one coefficient on the ad stock and linear controls, 2SLS averages
the state-specific effects with weights that put almost all mass on the
`r fmt(100 * share_pre, 0)` percent of household-days before conversion, where
the purchase probability and with it the effect are tiny, and the linear
controls cannot follow a probability that jumps by a factor of a hundred at
conversion. DML-IV on 2,000 households replaces the linear controls with
forests but is too imprecise here to separate the two. The probit
control function averages $\Phi'(\cdot)\,\beta_{it}$ over the observed
states and matches the DGP. The lesson for a rare-outcome panel with state
dependence: the estimand of a linear IV is a variance-weighted average that
need not be the average partial effect, and the nonlinear second stage is
not a cosmetic choice.

## Inference for the two-step estimator {#sec-inference}

**Three standard errors for the same coefficients.** The model-based one
treats the control-function stock as data; the authors' bootstrap keeps
unique households; the cluster bootstrap keeps duplicates. Table
\@ref(tab:se-compare) puts them side by side for the benchmark of Section
\@ref(sec-benchmark) (authors' DGP) and for column 4 of Table
\@ref(tab:table4) (show-level panel).

```{r se-compare, echo = FALSE}
se_terms <- c("A_stock", "CF_stock", "post", "promo1m")
se_cmp <- bind_rows(
  tibble(Panel = "Authors' DGP (full model)", Term = term_labels[se_terms], `Model-based` = bench_model_se[se_terms],
         `Authors' bootstrap` = bb$unique$se[se_terms], `Cluster bootstrap` = bb$cluster$se[se_terms]),
  tibble(Panel = "Show-level panel (column 4)", Term = term_labels[se_terms], `Model-based` = t4$c4$se[se_terms],
         `Authors' bootstrap` = NA_real_, `Cluster bootstrap` = b4$c4$se[se_terms])
)
se_cmp %>% mutate(across(where(is.numeric), ~ ifelse(is.na(.x), "---", fmt(.x, 4)))) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Model-based, unique-household bootstrap, and cluster bootstrap standard errors of the control-function probit.") %>%
  collapse_rows(1, latex_hline = "major", valign = "top") %>% tab_style()
```

The model-based standard error of the ad-stock coefficient is
`r fmt(se_cmp[["Model-based"]][1] / se_cmp[["Cluster bootstrap"]][1], 2)` of the cluster
bootstrap's on the authors' DGP and
`r fmt(se_cmp[["Model-based"]][5] / se_cmp[["Cluster bootstrap"]][5], 2)` on the show-level
panel; the generated regressor is the reason. The authors' bootstrap is
`r fmt(mean(se_cmp[["Authors' bootstrap"]][1:4] / se_cmp[["Cluster bootstrap"]][1:4]), 2)` of the
cluster bootstrap on average, the 0.76 of a 63 percent subsample.

# What the replication teaches about the workflow {#sec-lessons}

- **The instrument is a design, not a variable.** It is built from the
  assignment process (the network's density) and the observed viewing
  window, and it is testable in three ways that a fitted propensity score
  is not. The mean-zero and orthogonality checks of Table 2 are the
  evidence to show; corr with $u$ cannot be shown and is what the checks
  stand in for.
- **A show level is required to build it.** The authors' scripts
  simulate the household-day outcome of the design with an oracle
  expectation. Anyone applying the method needs second-by-second viewing
  and all brands' airing times, not exposures alone.
- **Control functions need stocks when regressors do.** The confounder
  persists at its own rate, so the control gets its own decay, chosen out
  of sample. `cf_residuals()` gives the residual; two lines accumulate it.
- **Bootstrap the whole pipeline, by cluster, with duplicates.** The
  model-based standard errors are too small for the generated regressor,
  and a bootstrap that keeps unique households is a subsample. The
  relabelling of duplicated households is what makes a stock accumulate
  correctly inside a cluster bootstrap.
- **The diagnostics are specific.** Strategic viewership shows in the
  instrument's mean; nonrandom ordering shows in the KS test; a direct
  effect shows in neither and needs a sensitivity bound.
- **Random coefficients are the expensive part.** Columns 5 and 6 with
  household random effects are the only fits in this note that take
  longer than a minute; the Laplace approximation of `glmer()` reproduces
  the structure of the paper's simulated maximum likelihood at a fraction
  of the cost, and the bootstrap of column 6 is the compute budget of the
  whole note.

# References

- Borusyak, K. and Hull, P. (2023). Nonrandom exposure to exogenous shocks. *Econometrica*, 91(6), 2155-2185.
- Conley, T. G., Hansen, C. B., and Rossi, P. E. (2012). Plausibly exogenous. *Review of Economics and Statistics*, 94(1), 260-272.
- Hsieh, T.-Y., Du, R. Y., and Lu, S. (2026). Leveraging large-scale granular single-source data for TV advertising: an identification strategy. *Marketing Science*, articles in advance.
- Petrin, A. and Train, K. (2010). A control function approach to endogeneity in consumer choice models. *Journal of Marketing Research*, 47(1), 3-13.
- Sethuraman, R., Tellis, G. J., and Briesch, R. A. (2011). How well does advertising work? Generalizations from a meta-analysis of brand advertising elasticity. *Journal of Marketing Research*, 48(3), 457-471.
- Wooldridge, J. M. (2015). Control function methods in applied econometrics. *Journal of Human Resources*, 50(2), 420-445.
