---
title: "Replication: Deep Learning-Based Causal Inference for Large-Scale Combinatorial Experiments"
subtitle: "Ye, Zhang, Zhang, Zhang, and Zhang (Management Science, 2025), with causalmetrics and PyTorch"
author: "Weiheng Zhang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
params:
  data_path: "/Users/waynezhang/Documents/note/R/R_package/causalmetrics/data_raw/04_Doubly_Robust_and_Double_ML/ye2025deep/"
  ridge: 0.01
  n_reps_appendix: 20
  rerun_python: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", fig.pos = "H",
  out.width = "90%", collapse = TRUE, comment = "#>"
)
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
  helper_path <- file.path(pkg_root, "inst", "vignette_helpers", "04_Doubly_Robust_and_Double_ML", "dedl_helpers.R")
  py_script <- file.path(pkg_root, "inst", "python", "dedl_nuisance.py")
} else {
  library(causalmetrics)
  helper_path <- system.file("vignette_helpers", "04_Doubly_Robust_and_Double_ML", "dedl_helpers.R", package = "causalmetrics")
  py_script <- system.file("python", "dedl_nuisance.py", package = "causalmetrics")
}
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
source(helper_path)
data_path <- params$data_path
derived <- file.path(data_path, "derived")
dir.create(derived, showWarnings = FALSE)
python <- find_python_torch()
run_python <- nzchar(python) && nzchar(py_script)
ridge_main <- params$ridge
fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f", big.mark = ",")
fmt0 <- function(x) formatC(x, digits = 0, format = "f", big.mark = ",")
tab_style <- function(k, ...) kable_styling(k, latex_options = c("hold_position", ...), font_size = 9)
tvar <- c("is_DP", "is_LP", "is_FYP")
dat <- read.csv(file.path(data_path, "empirics", "synthetic_data.csv"))
xvar <- setdiff(names(dat), c(tvar, "y", "kfold"))
key <- combo_key(dat[, tvar])
dat$.row <- seq_len(nrow(dat))
n_folds <- length(unique(dat$kfold))
# the assignment distribution of the observed combinations in the paper's design
t_dist_paper <- data.frame(combo_matrix(observed_cells), prob = 1 / length(observed_cells))
names(t_dist_paper)[1:3] <- tvar
# python runs are cached in `derived/`; delete the files (or set params$rerun_python) to retrain
py_run <- function(output, args, history = NULL, epoch_theta = NULL, input = file.path(data_path, "empirics", "synthetic_data.csv")) {
  if (!run_python && !file.exists(output)) stop("No Python with torch was found and no cached output exists: ", output)
  run_dedl_python(python, py_script, input, output, args, history = history, epoch_theta = epoch_theta,
                  rerun = params$rerun_python)
}
```

# The paper in one page {#sec-paper}

**The question is how to estimate the effect of a combination of treatments
that was never tested.** Large platforms run hundreds of A/B tests at once
under orthogonal assignment, so every user sits in some combination of
treatments, but only a few combinations have enough traffic to be measured on
their own. The common practice adds up the single-treatment effects. Ye,
Zhang, Zhang, Zhang, and Zhang (2025) show on a video platform that this
linear addition can be badly wrong (their Figure 1), and propose a method,
debiased deep learning (DeDL), that observes only $m + 2$ combinations of
$m$ treatments and recovers the average treatment effect (ATE) of all $2^m$.

The method has three parts.

1. **A structured outcome model.** $E[Y \mid X = x, T = t] = G(\theta(x), t)$
   with a known link $G$ and an unknown parameter function $\theta(x)$. Their
   choice is the generalized sigmoid,
   $G(\theta, t) = \theta_{m+1} / (1 + \exp(-(\theta_0 + \theta_1 t_1 + \dots + \theta_m t_m)))$,
   which captures diminishing or increasing returns to combining treatments
   user by user (their Theorem 1 shows it can approximate any ATE function as
   the number of distinct covariate profiles grows). Two ReLU networks output
   $\theta_0(x)$ and $(\theta_1, \dots, \theta_m)(x)$; the scale $\theta_{m+1}$
   is one trained constant.
2. **An influence function.** The plug-in estimator
   $\frac{1}{n}\sum_i [G(\hat\theta(x_i), t) - G(\hat\theta(x_i), t_0)]$ inherits
   the bias of the network. Following Farrell, Liang, and Misra (2020) they add
   the Neyman-orthogonal correction (their Proposition 2),
   $$
   \psi_i(t) = H_i - H_{\theta,i}'\,\Lambda(x_i)^{-1}\,\ell_{\theta,i}, \qquad
   H_i = G(\hat\theta_i, t) - G(\hat\theta_i, t_0),
   $$
   where $H_\theta$ is the gradient of $H$ in $\theta$,
   $\Lambda(x) = 2\,E_\nu[G_\theta G_\theta' \mid x]$ averages the outer product of
   the gradient over the *known* assignment distribution $\nu$ of the observed
   combinations, and $\ell_\theta = 2 G_\theta(\hat\theta, \check t)(G(\hat\theta, \check t) - y)$
   is the gradient of the squared loss at the realized treatment $\check t$.
   The mean of $\psi_i(t)$ over a fold whose $\hat\theta$ was trained on the
   other folds is root-$n$ normal (their Theorem 2).
3. **Four-fold cross-fitting**, the best combination as the largest estimated
   ATE, and the gap $\tau(t) = \mu(t^*) - \mu(t)$ with its own influence
   function for best-treatment identification.

Benchmarks: linear addition (LA), linear regression of $y$ on $(x, t)$ (LR), a
pure network on $(x, t)$ (PDL), and the plug-in structured network without the
correction (SDL). On the platform's data the "unobserved" combinations are in
fact observed, so every estimator is scored against the ground truth with four
metrics: the correct-direction ratio (sign and significance), MAPE, MSE, and
MAE. DeDL wins on every metric (their Table 3), identifies the best
combination (Table 4), and does so within demographic groups (Tables 5-6).
Their Section 5 lists the implementation checkpoints: stratified sampling to
balance covariates, an architecture and learning-rate search, a ridge on
$\Lambda(x)$ before inversion, and the training error as the compass (Figure 7).

**What this note does.** The platform data are proprietary. The authors ship a
synthetic data set with the same structure, a 999-subgroup file for Figure 1,
and a Keras notebook whose printed outputs are the numbers we match. We
reproduce every table and figure of the empirical section in the published
layout, using `est_dml_structured()` from `causalmetrics` for the influence
function and inference and a PyTorch script for the network, then port the
validation experiment of the paper's Appendix D. Two things are learned on the
way: the correction term is unstable without a ridge (Section
\@ref(sec-ridge)), and the authors' notebook departs from the paper's algorithm
in ways that change the ranking of DeDL and SDL (Section \@ref(sec-variants)).

# Data and ground truth {#sec-data}

## The synthetic data

The file has `r fmt0(nrow(dat))` users, the three treatment indicators, an
outcome `y` scaled to $[0, 1]$, `r length(xvar)` covariates (10 standardized
continuous usage variables and 77 dummies for operating system, activity,
gender, age, location, follower and friend counts, and phone price tier), and a
fold index. Unlike the paper's stratified sample, all eight combinations are
present in every fold.

```{r tab-cells, echo = FALSE}
tibble(combo = key, fold = dat$kfold) %>%
  count(combo, fold) %>%
  pivot_wider(names_from = fold, values_from = n, names_prefix = "Fold ") %>%
  mutate(combo = factor(combo, levels = paper_order),
         observed = ifelse(combo %in% observed_cells, "Observable", "Unobservable")) %>%
  arrange(combo) %>%
  mutate(combo = combo_label(as.character(combo))) %>%
  select(combo, observed, starts_with("Fold")) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Treatment combination", "Status in the exercise", paste("Fold", 1:4)),
      caption = "Users per treatment combination and fold in the synthetic data. The exercise treats the last three combinations as unobserved.") %>%
  tab_style()
```

## Figure 1: heterogeneous response to two experiments

The paper's opening figure uses 1,254 subgroups of users defined by
demographics and activity, each observed under all four combinations of two
experiments (D and S). The released file has 999 subgroups with the mean
outcome under (D,S), (\_,S), (D,\_), and (\_,\_). The left panel plots the true
combined effect against the linear-additive prediction; the right panel the
distribution of the percentage error of linear addition.

```{r fig1, echo = FALSE, fig.height = 3.6, fig.cap = "Heterogeneous response to two experiments (paper's Figure 1). Left: true combined effect against the linear-additive effect by subgroup, with the 45-degree line (solid) and the no-intercept linear fit (dot-dash). Right: density of the percentage error of linear addition, trimmed to its 10 to 90 percent quantiles."}
fig1_dat <- read.csv(file.path(data_path, "empirics", "Heterogeneous_Response_to_Two_Experiments.csv"))
plot_figure1(fig1_dat)
```

```{r fig1-numbers, echo = FALSE}
true_eff <- fig1_dat$app_DS - fig1_dat$app___
la_eff <- fig1_dat$app__S + fig1_dat$app_D_ - 2 * fig1_dat$app___
fig1_slope <- sum(true_eff * la_eff) / sum(true_eff^2)
fig1_share_wrong_sign <- mean(sign(true_eff) != sign(la_eff))
```

The fitted slope is `r fmt(fig1_slope, 2)` rather than one, and linear addition
gets the sign of the combined effect wrong in
`r round(100 * fig1_share_wrong_sign)` percent of the subgroups. This is the
motivation for a model of interactions.

## Table 2: ground-truth ATEs

With all eight cells observed, the ground truth is the difference in cell
means against the baseline $(0, 0, 0)$. The authors test each difference with a
pooled $t$-test when Levene's test does not reject equal variances and Welch's
test otherwise; we do the same.

```{r truth}
truth <- ground_truth_table(dat$y, key)
```

```{r tab2, echo = FALSE}
truth %>%
  mutate(combo = combo_label(combo),
         rel = ifelse(combo == "(0, 0, 0)", "0.000", paste0(fmt(100 * relative, 3), stars(p.value))),
         obs = ifelse(combo %in% combo_label(observed_cells), "Observable", "Unobservable")) %>%
  select(combo, rel, obs, n) %>%
  mutate(n = fmt0(n)) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, align = c("l", "r", "l", "r"),
      col.names = c("Treatment combination $t'$", "Relative effect size, \\% (1)", "Observed or not (2)", "Number of users (3)"),
      caption = "Ground-truth ATE and best treatment identification of eight treatment combinations (paper's Table 2, synthetic data). Relative effect size is the cell-mean difference against the baseline divided by the baseline mean. * $p<0.05$; ** $p<0.01$; *** $p<0.001$; **** $p<0.0001$.") %>%
  tab_style() %>%
  footnote(general = paste0("The optimal treatment combination is ", combo_label(truth$combo[which.max(truth$diff)]), "."),
           general_title = "Note.", footnote_as_chunk = TRUE)
```

```{r truth-numbers, echo = FALSE}
best_true <- truth$combo[which.max(truth$diff)]
```

The relative effects match the notebook's printed values (0.519, $-0.353$,
$-0.358$, 0.149, $-0.585$, 0.332 for the first six non-baseline
combinations). Every effect is significant at any conventional level, so the
correct-direction ratio will only reward sign agreement, and the best
combination is `r combo_label(best_true)`. The synthetic outcome is far less
noisy than the platform's screen time: the paper's ground truth has two
insignificant combinations and relative effects of one to two percent.

# The structured model and its influence function {#sec-model}

## Link function and gradients

Write $u = \theta_0 + \theta' t$ and $s(u) = 1 / (1 + e^{-u})$. The generalized
sigmoid is $G(\theta, t) = c\, s(u)$ with $c = \theta_{m+1}$, so
$$
G_\theta(\theta, t) = \big(c\,s(u)(1 - s(u)),\; c\,s(u)(1 - s(u))\,t',\; s(u)\big)',
$$
a vector of length $m + 2$. The plug-in contrast is
$H = G(\theta, t) - G(\theta, t_0)$ and $H_\theta = G_\theta(\theta, t) - G_\theta(\theta, t_0)$.
With the assignment distribution $\nu$ uniform over the five observed
combinations, $\Lambda(x) = \frac{2}{5}\sum_{t \in \mathcal{T}^o} G_\theta(\theta(x), t) G_\theta(\theta(x), t)'$
is a $5 \times 5$ matrix per user, a sum of five rank-one terms. It is
invertible in principle but nearly singular for users whose sigmoid is
saturated ($s(u)$ near 0 or 1), because then all but the last entry of
$G_\theta$ are close to zero for every $t$. This is where the ridge of Section
\@ref(sec-ridge) enters.

## Why the correction removes first-order bias

The plug-in $\frac{1}{n}\sum_i H_i$ has bias $E[H_\theta'(\hat\theta - \theta_0)]$
to first order. The score $\ell_\theta$ has zero conditional mean at
$\theta_0$ and derivative $E[\ell_{\theta\theta} \mid x] = \Lambda(x)$, so
$E[\Lambda(x)^{-1}\ell_\theta(\hat\theta) \mid x] \approx \hat\theta(x) - \theta_0(x)$:
the correction term estimates the nuisance error from the residuals and
subtracts its first-order effect. This is the same Neyman orthogonality as in
lecture 04, written for a parameter *function* rather than a scalar nuisance.
Cross-fitting supplies the independence between $\hat\theta$ and the scored
observations that the argument needs.

## The package function

`est_dml_structured()` takes the out-of-fold $\hat\theta$ matrix from any
engine, the realized treatments, the target combinations, the link, the
assignment distribution, and the fold index, and returns the influence
functions, the fold-by-fold and pooled estimates, the plug-in estimates (SDL)
from the same call, and the contrasts to the best combination. With the true
$\theta$ the correction has mean zero and the estimator equals the plug-in
contrast, which is the package's unit test. The first stage is
`inst/python/dedl_nuisance.py`, a PyTorch script mirroring the authors' Keras
architecture (two 3-layer, 20-node ReLU branches, a fixed sigmoid, a final
linear layer without bias, Adam with learning rate $10^{-3}$, batch 128, 50
epochs).

# First stage: the structured network {#sec-first-stage}

The paper trains on the five observed combinations of three folds and scores
the fourth. The script does this for every fold in one call and writes
$\hat\theta$ for every user out of fold; runs are cached in `derived/`.

```{r first-stage, cache = TRUE, cache.extra = params$rerun_python}
theta_paper <- py_run(
  output = file.path(derived, "theta_paper.csv"),
  history = file.path(derived, "history_paper.csv"),
  args = c("--model structured --y y --t is_DP,is_LP,is_FYP --x-all-others --fold kfold",
           "--train-cells 000,001,010,100,111 --arch two_branch --hidden 20 --depth 3",
           "--epochs 50 --batch 128 --lr 0.001 --val-split 0.1 --seed 1")
)
stopifnot(all(theta_paper$.row == dat$.row))
history_paper <- read.csv(file.path(derived, "history_paper.csv"))
theta_cols <- c("theta_0", "theta_1", "theta_2", "theta_3", "theta_c")
```

```{r tab-first-stage, echo = FALSE}
history_paper %>%
  group_by(fold) %>%
  summarise(epochs = max(epoch), train_mse = train_mse[epoch == max(epoch)], val_mse = val_mse[epoch == max(epoch)],
            best_val = min(val_mse), .groups = "drop") %>%
  left_join(theta_paper %>% group_by(fold) %>% summarise(n_scored = n(), c_hat = first(theta_c), .groups = "drop"), by = "fold") %>%
  arrange(fold) %>%
  mutate(across(c(train_mse, val_mse, best_val), ~ fmt(.x, 5)), c_hat = fmt(c_hat, 3), n_scored = fmt0(n_scored)) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Held-out fold", "Epochs", "Training MSE", "Validation MSE", "Best validation MSE", "Users scored", "$\\hat c$"),
      escape = FALSE, caption = "First stage per fold: squared loss after 50 epochs on the training rows (five observed combinations of the other three folds) and on the 10 percent validation split, and the trained scale $\\hat c$.") %>%
  tab_style()
```

```{r fig-history, echo = FALSE, fig.height = 3.2, fig.cap = "Validation mean squared error by epoch for the four cross-fitting networks."}
ggplot(history_paper, aes(epoch, val_mse, colour = factor(fold))) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Epoch", y = "Validation MSE (log scale)", colour = "Held-out fold") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

The validation loss is still falling slowly at epoch 50 in every fold. The
authors' Figure 7 and Section 5.3 make the point that the quality of this fit
is the quality of everything that follows; Section \@ref(sec-epochs) reproduces
that figure.

# Second stage: DeDL and SDL {#sec-second-stage}

## The ridge on $\Lambda(x)$ is not optional {#sec-ridge}

We first run the influence function exactly as written, with no ridge, and
then with the ridge the paper mentions in Section 5.2 (they report grid
searching values such as 0.001, 0.01, and 1; the notebook uses 0.5).

```{r ridge-grid, cache = TRUE, dependson = "first-stage"}
theta_mat <- as.matrix(theta_paper[, theta_cols])
ridge_grid <- c(0, 1e-3, 1e-2, 0.05, 0.1, 0.5, 1)
ridge_res <- lapply(ridge_grid, function(r) {
  fit <- est_dml_structured(theta_mat, dat$y, dat[, tvar], targets = paper_order[-1], t_dist = t_dist_paper,
                            fold_id = dat$kfold, ridge = r, contrast_best = FALSE)
  p <- fit$estimates[fit$estimates$fold == "pooled", ]
  ape <- 100 * abs(p$estimate - truth$diff[-1]) / abs(truth$diff[-1])
  ape_sdl <- 100 * abs(p$plugin - truth$diff[-1]) / abs(truth$diff[-1])
  corr <- fit$psi - fit$plugin
  data.frame(ridge = r, mape_all = mean(ape), mape_unobs = mean(ape[5:7]), sdl_all = mean(ape_sdl), sdl_unobs = mean(ape_sdl[5:7]),
             max_corr = max(abs(corr)), share_large = mean(abs(corr) > 1), ill = fit$diagnostics$share_ill_conditioned)
})
ridge_tab <- bind_rows(ridge_res)
```

```{r tab-ridge, echo = FALSE}
ridge_tab %>%
  mutate(across(c(mape_all, mape_unobs, sdl_all, sdl_unobs), ~ fmt(.x, 2)),
         max_corr = formatC(max_corr, digits = 2, format = "e"), share_large = fmt(100 * share_large, 2)) %>%
  select(ridge, mape_all, mape_unobs, max_corr, share_large, sdl_all, sdl_unobs) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Ridge", "MAPE all, \\%", "MAPE unobserved, \\%", "Largest $|$correction$|$", "Share $|$correction$| > 1$, \\%", "SDL MAPE all, \\%", "SDL MAPE unobserved, \\%"),
      escape = FALSE, caption = "Sensitivity of the DeDL estimates to the ridge added to $\\Lambda(x)$ before inversion. The SDL columns do not depend on the ridge.") %>%
  tab_style() %>%
  add_header_above(c(" " = 1, "DeDL" = 4, "SDL" = 2))
```

Without a ridge the estimator is useless: a handful of users with a saturated
sigmoid have a nearly singular $\Lambda(x)$, their corrections run into the
thousands, and the unobserved combinations' MAPE is in the tens of thousands
of percent, while the plug-in is within two percent. A ridge of
`r ridge_main` removes those users' leverage and leaves the correction
elsewhere intact; larger ridges shrink the correction toward zero, and DeDL
converges to SDL. The rest of the note uses `ridge = `r ridge_main``. This is
the practical content of the paper's remark that "$\lambda(x)$ may be
noninvertible for certain values of $x$".

## Table 7: fold-by-fold estimates

```{r dedl-main, cache = TRUE, dependson = "first-stage"}
fit_dedl <- est_dml_structured(theta_mat, dat$y, dat[, tvar], targets = paper_order[-1], t_dist = t_dist_paper,
                               fold_id = dat$kfold, ridge = ridge_main)
fit_dedl
# the plug-in (SDL) fold table from the same object
plugin_table <- function(fit) {
  crit <- qnorm(0.975)
  folds <- sort(unique(fit$fold_id))
  bind_rows(lapply(seq_along(fit$labels), function(k) {
    fe <- sapply(folds, function(f) mean(fit$plugin[fit$fold_id == f, k]))
    fs <- sapply(folds, function(f) sd(fit$plugin[fit$fold_id == f, k]) / sqrt(sum(fit$fold_id == f)))
    bind_rows(data.frame(target = fit$labels[k], fold = as.character(folds), estimate = fe, std.error = fs,
                         conf.low = fe - crit * fs, conf.high = fe + crit * fs),
              data.frame(target = fit$labels[k], fold = "pooled", estimate = mean(fe), std.error = sqrt(sum(fs^2)) / length(folds),
                         conf.low = mean(fe) - crit * sqrt(sum(fs^2)) / length(folds), conf.high = mean(fe) + crit * sqrt(sum(fs^2)) / length(folds)))
  }))
}
sdl_tab <- plugin_table(fit_dedl)
```

```{r tab7, echo = FALSE}
truth_lab <- truth %>% mutate(target = combo_label(combo))
t7 <- fit_dedl$estimates %>%
  left_join(truth_lab %>% select(target, truth_diff = diff, p.value), by = "target") %>%
  mutate(low = ifelse(fold == "pooled", ci_low_avg, conf.low), high = ifelse(fold == "pooled", ci_high_avg, conf.high),
         sig = p.value < 0.05,
         ape = ifelse(sig, 100 * abs(estimate - truth_diff) / abs(truth_diff), NA),
         se = ifelse(sig, (estimate - truth_diff)^2, NA), ae = ifelse(sig, abs(estimate - truth_diff), NA),
         fold = ifelse(fold == "pooled", "Mean", fold),
         truth_txt = paste0(fmt(truth_diff, 3), stars(p.value)),
         ci = paste0("[", fmt(low, 3), ", ", fmt(high, 3), "]")) %>%
  mutate(target = factor(target, levels = combo_label(paper_order[-1]))) %>%
  arrange(target, fold) %>%
  transmute(target = as.character(target), truth_txt, fold, estimate = fmt(estimate, 4), ci,
            ape = ifelse(is.na(ape), "NA", fmt(ape, 2)), se = ifelse(is.na(se), "NA", formatC(se * 1e4, digits = 3, format = "f")),
            ae = ifelse(is.na(ae), "NA", formatC(ae * 1e2, digits = 3, format = "f")))
kbl(t7, booktabs = TRUE, linesep = "", escape = FALSE,
    col.names = c("Treatment combination", "Ground-truth ATE (1)", "Fold (2)", "Estimated ATE (3)", "95\\% CI for ATE estimate (4)", "APE, \\% (5)", "SE $\\times 10^4$ (6)", "AE $\\times 10^2$ (7)"),
    caption = "Detailed results of fourfold DeDL estimators (paper's Table 7 layout). Effects are in units of the scaled synthetic outcome; the Mean row averages the fold estimates and the fold confidence bounds, as in the paper. SE and AE are scaled by constants, as in the paper.") %>%
  tab_style("scale_down") %>%
  collapse_rows(columns = 1:2, valign = "top", latex_hline = "major")
```

```{r tab7-numbers, echo = FALSE}
pooled_dedl <- fit_dedl$estimates %>% filter(fold == "pooled")
pooled_sdl <- sdl_tab %>% filter(fold == "pooled")
ape_pool <- 100 * abs(pooled_dedl$estimate - truth$diff[-1]) / abs(truth$diff[-1])
fold_spread <- fit_dedl$estimates %>% filter(fold != "pooled") %>% group_by(target) %>%
  summarise(sd = sd(estimate), mean_se = mean(std.error), .groups = "drop")
```

Fold estimates differ from each other by about as much as their standard
errors: the standard deviation across the four fold estimates reaches
`r fmt(max(fold_spread$sd), 3)` against fold standard errors of about
`r fmt(mean(fold_spread$mean_se), 3)`, because each fold's network is a
different random fit. Averaging over folds is what makes the final estimate
usable, and the paper's convention of averaging the fold confidence bounds,
rather than combining them as independent estimates, is a conservative
acknowledgement of that.

## Figure 5: DeDL against SDL

```{r fig5, echo = FALSE, fig.height = 3.8, fig.cap = "Detailed comparison between the SDL and DeDL estimators (paper's Figure 5 layout): ground-truth effect with its confidence interval, DeDL, and SDL, for the seven non-baseline combinations. Bars are 95 percent intervals."}
plot_figure5(truth, pooled_dedl %>% transmute(combo = paper_order[-1], estimate, conf.low, conf.high),
             pooled_sdl %>% transmute(combo = paper_order[-1], estimate, conf.low, conf.high))
```

On the synthetic data the two estimators are close and both track the truth;
the DeDL intervals are wider than the SDL intervals because the correction
term adds variance, which is the price of removing the network's bias. (The
ground-truth bars use the standard error of the cell-mean difference; the
notebook scales the cell standard deviation by a normal quantile of the
$p$-value, which is not an interval.) The paper's platform data show the SDL
intervals to be too narrow; here, with a nearly noiseless synthetic outcome,
the plug-in is already accurate.

# Benchmarks and metrics {#sec-benchmarks}

## LA, LR, and PDL

Linear addition uses the cell means of the full data. Linear regression is
fitted on the training rows of each fold (observed combinations only) and
evaluated on the held-out fold; the pure network is trained the same way by
the script with a 3-layer, 32-node ReLU architecture on $(x, t)$. Both report
the mean contrast of predicted outcomes, with the authors' inference: the
standard deviation of the row-wise contrast over $\sqrt{n}$ and a Welch test
between the two prediction vectors.

```{r benchmarks, cache = TRUE, dependson = "first-stage"}
la_tab <- la_estimator(dat$y, key)
lr_folds <- lapply(sort(unique(dat$kfold)), function(f) {
  train <- dat[dat$kfold != f & key %in% observed_cells, ]
  lr_estimator(train, dat[dat$kfold == f, ], "y", tvar, xvar)
})
average_folds <- function(tabs) {
  bind_rows(tabs, .id = "fold") %>%
    group_by(combo) %>%
    summarise(estimate = mean(estimate), conf.low = mean(conf.low), conf.high = mean(conf.high),
              std.error = mean(std.error), p.value = max(p.value), .groups = "drop") %>%
    mutate(combo = factor(combo, levels = paper_order)) %>% arrange(combo) %>% mutate(combo = as.character(combo))
}
lr_tab <- average_folds(lapply(lr_folds, `[[`, "table"))
pdl_paper <- py_run(
  output = file.path(derived, "pdl_paper.csv"),
  args = c("--model pure --y y --t is_DP,is_LP,is_FYP --x-all-others --fold kfold",
           "--train-cells 000,001,010,100,111 --hidden 32 --depth 3 --epochs 50 --batch 128 --lr 0.001 --seed 1")
)
pdl_folds <- lapply(sort(unique(pdl_paper$fold)), function(f) pdl_estimator(pdl_paper[pdl_paper$fold == f, ]))
pdl_tab <- average_folds(lapply(pdl_folds, `[[`, "table"))
dedl_tab <- pooled_dedl %>% transmute(combo = paper_order[-1], estimate, std.error, conf.low = ci_low_avg, conf.high = ci_high_avg) %>%
  bind_rows(data.frame(combo = "000", estimate = 0, std.error = 0, conf.low = 0, conf.high = 0), .)
sdl_tab2 <- pooled_sdl %>% transmute(combo = paper_order[-1], estimate, std.error, conf.low, conf.high) %>%
  bind_rows(data.frame(combo = "000", estimate = 0, std.error = 0, conf.low = 0, conf.high = 0), .)
estimators <- list(LA = la_tab, LR = lr_tab, PDL = pdl_tab, SDL = sdl_tab2, DeDL = dedl_tab)
tab3 <- metrics_table(estimators, truth)
```

## Table 3: comparison of the estimators of the ATE

```{r tab3, echo = FALSE}
notebook_tab3 <- data.frame(
  estimator = c("LA", "LR", "PDL", "SDL", "DeDL"),
  nb_mape_u = c(166.67, 21.99, 3.87, 1.95, 6.16), nb_mape = c(62.84, 24.89, 5.98, 2.79, 9.04))
tab3 %>%
  left_join(notebook_tab3, by = "estimator") %>%
  mutate(across(c(mape_u, mape), ~ fmt(.x, 2)), across(c(mse_u, mse), ~ formatC(.x * 1e4, digits = 3, format = "f")),
         across(c(mae_u, mae), ~ formatC(.x * 1e2, digits = 3, format = "f")), across(c(nb_mape_u, nb_mape), ~ fmt(.x, 2))) %>%
  select(estimator, cdr_u, mape_u, mse_u, mae_u, nb_mape_u, cdr, mape, mse, mae, nb_mape) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Estimator", "CDRu (1)", "MAPEu, \\% (2)", "MSEu $\\times 10^4$ (3)", "MAEu $\\times 10^2$ (4)", "Notebook MAPEu, \\%",
                    "CDR (5)", "MAPE, \\% (6)", "MSE $\\times 10^4$ (7)", "MAE $\\times 10^2$ (8)", "Notebook MAPE, \\%"),
      caption = "Comparison of different estimators of the ATE (paper's Table 3 layout) on the synthetic data, with the MAPE printed by the authors' notebook next to ours. MSE and MAE are scaled by constants, as in the paper.") %>%
  tab_style("scale_down") %>%
  add_header_above(c(" " = 1, "Unobserved treatment combinations" = 5, "All treatment combinations" = 5))
```

```{r tab3-numbers, echo = FALSE}
rank_u <- tab3$estimator[order(tab3$mape_u)]
g3 <- function(e, col) fmt(tab3[[col]][tab3$estimator == e], 1)
```

On the unobserved combinations the ranking is
`r paste(rank_u, collapse = ", ")`. Linear addition misses by
`r g3("LA", "mape_u")` percent because the synthetic interactions are strong;
linear regression by `r g3("LR", "mape_u")` percent because it cannot bend;
the pure network by `r g3("PDL", "mape_u")` percent because it never saw
those combinations and must extrapolate without structure, which is the
paper's argument for a link function. The two structured estimators are
within two percent, DeDL at `r g3("DeDL", "mape_u")` and SDL at
`r g3("SDL", "mape_u")`; they are close because the plug-in is already nearly
unbiased on this nearly noiseless outcome. Three of the notebook's columns
are not comparable: its linear addition mis-indexes the single effects (it
adds the effects of $(0,0,1)$ and $(0,1,0)$ for the combination $(1,1,0)$,
and so on), its pure network is trained on all eight cells, and its DeDL
departs from Algorithm 1 (Section \@ref(sec-variants)).

## Table 4: best-treatment identification

Each estimator names its best combination as the largest estimated ATE and
reports the gap $\tau(t)$ to every other combination. DeDL and SDL use the
difference of their influence functions (the package's `contrasts`), LA adds
variances, LR and PDL difference their predicted outcomes with a Welch test.
The ground truth is the gap in cell means, tested cell against cell.

```{r tab4, cache = TRUE, dependson = "benchmarks"}
truth_tau <- {
  yb <- dat$y[key == best_true]
  do.call(rbind, lapply(paper_order, function(cc) {
    yt <- dat$y[key == cc]
    data.frame(combo = cc, diff = mean(yb) - mean(yt), p.value = if (cc == best_true) 1 else t.test(yb, yt)$p.value)
  }))
}
contrast_from_predictions <- function(pred_list, tab) {
  best <- tab$combo[which.max(tab$estimate)]
  bind_rows(lapply(pred_list, function(P) {
    do.call(rbind, lapply(paper_order, function(cc) {
      d <- P[, best] - P[, cc]
      data.frame(combo = cc, estimate = mean(d), std.error = sd(d) / sqrt(length(d)),
                 p.value = if (cc == best) 1 else t.test(P[, best], P[, cc])$p.value)
    }))
  }), .id = "fold") %>% group_by(combo) %>%
    summarise(estimate = mean(estimate), std.error = mean(std.error), p.value = max(p.value), .groups = "drop") %>%
    mutate(conf.low = estimate - 1.96 * std.error, conf.high = estimate + 1.96 * std.error)
}
tau_lr <- contrast_from_predictions(lapply(lr_folds, `[[`, "predictions"), lr_tab)
tau_pdl <- contrast_from_predictions(lapply(pdl_folds, `[[`, "predictions"), pdl_tab)
tau_la <- {
  b <- which.max(la_tab$estimate)
  la_tab %>% transmute(combo, estimate = la_tab$estimate[b] - estimate,
                       std.error = sqrt(la_tab$std.error[b]^2 + std.error^2),
                       conf.low = estimate - 1.96 * std.error, conf.high = estimate + 1.96 * std.error)
}
tau_dedl <- fit_dedl$contrasts %>% filter(fold == "pooled") %>%
  transmute(combo = paper_order[-1], estimate, std.error, conf.low, conf.high) %>%
  bind_rows(data.frame(combo = "000", estimate = pooled_dedl$estimate[which.max(pooled_dedl$estimate)],
                       std.error = pooled_dedl$std.error[which.max(pooled_dedl$estimate)]) %>%
              mutate(conf.low = estimate - 1.96 * std.error, conf.high = estimate + 1.96 * std.error), .)
fit_sdl_contrast <- {
  kb <- which.max(pooled_sdl$estimate)
  d <- fit_dedl$plugin[, kb] - fit_dedl$plugin
  folds <- sort(unique(fit_dedl$fold_id))
  bind_rows(lapply(seq_along(fit_dedl$labels), function(k) {
    fe <- sapply(folds, function(f) mean(d[fit_dedl$fold_id == f, k]))
    fs <- sapply(folds, function(f) sd(d[fit_dedl$fold_id == f, k]) / sqrt(sum(fit_dedl$fold_id == f)))
    data.frame(combo = paper_order[-1][k], estimate = mean(fe), std.error = sqrt(sum(fs^2)) / length(folds))
  })) %>% mutate(conf.low = estimate - 1.96 * std.error, conf.high = estimate + 1.96 * std.error)
}
tau_sdl <- bind_rows(data.frame(combo = "000", estimate = pooled_sdl$estimate[which.max(pooled_sdl$estimate)],
                                std.error = pooled_sdl$std.error[which.max(pooled_sdl$estimate)]) %>%
                       mutate(conf.low = estimate - 1.96 * std.error, conf.high = estimate + 1.96 * std.error), fit_sdl_contrast)
order_tau <- function(t) t[match(paper_order, t$combo), ]
tab4 <- bind_rows(lapply(list(LA = tau_la, LR = tau_lr, PDL = tau_pdl, SDL = tau_sdl, DeDL = tau_dedl), function(t) {
  combo_metrics(order_tau(t), truth_tau)
}), .id = "estimator")
best_named <- c(LA = la_tab$combo[which.max(la_tab$estimate)], LR = lr_tab$combo[which.max(lr_tab$estimate)],
                PDL = pdl_tab$combo[which.max(pdl_tab$estimate)], SDL = paper_order[-1][which.max(pooled_sdl$estimate)],
                DeDL = paper_order[-1][which.max(pooled_dedl$estimate)])
```

```{r tab4-print, echo = FALSE}
notebook_tab4 <- c(LA = 127.79, LR = 19.75, PDL = 3.16, SDL = 1.87, DeDL = 1.69)
tab4 %>%
  mutate(best = combo_label(best_named[estimator]), nb = fmt(notebook_tab4[estimator], 2),
         mape = fmt(mape, 2), mse = formatC(mse * 1e4, digits = 3, format = "f"), mae = formatC(mae * 1e2, digits = 3, format = "f")) %>%
  select(estimator, best, cdr, mape, mse, mae, nb) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Estimator", "Identified best", "CDR (1)", "MAPE, \\% (2)", "MSE $\\times 10^4$ (3)", "MAE $\\times 10^2$ (4)", "Notebook MAPE, \\%"),
      caption = paste0("Comparison of different estimators of best treatment identification (paper's Table 4 layout). The true best combination is ",
                       combo_label(best_true), ". Metrics compare the estimated gap to the best combination with the ground-truth gap over all eight combinations.")) %>%
  tab_style()
```

# Conditional effects by demographic group {#sec-cate}

The paper splits users by residence (rural or not) and phone price tier. The
synthetic data do not carry that cross, so the notebook, and we, use the four
groups defined by `gender_M` and `feed_model_1`. The group CATE of DeDL and
SDL is the group mean of the influence functions, the LR and PDL CATEs are
group means of the predicted contrasts, and the LA CATE adds the group's
single-treatment cell effects. The ground truth is the group's cell means. As
in the notebook's Table 5 code, MAPE here averages over all non-baseline
combinations without dropping insignificant ones.

```{r cate, cache = TRUE, dependson = "tab4"}
grp <- paste0("G", 1 + 2 * dat$gender_M + dat$feed_model_1)
grp_names <- c(G1 = "female, feed model 2", G2 = "female, feed model 1", G3 = "male, feed model 2", G4 = "male, feed model 1")
lr_pred_all <- do.call(rbind, lapply(seq_along(lr_folds), function(i) {
  data.frame(.row = dat$.row[dat$kfold == sort(unique(dat$kfold))[i]], lr_folds[[i]]$predictions)
})) %>% arrange(.row)
pdl_pred_all <- do.call(rbind, lapply(seq_along(pdl_folds), function(i) {
  data.frame(.row = pdl_paper$.row[pdl_paper$fold == sort(unique(pdl_paper$fold))[i]], pdl_folds[[i]]$predictions)
})) %>% arrange(.row)
cate_group <- function(g) {
  rows <- grp == g
  tr <- ground_truth_table(dat$y[rows], key[rows])
  la <- la_estimator(dat$y[rows], key[rows])$estimate
  dedl <- c(0, colMeans(fit_dedl$psi[rows, ]))
  sdl <- c(0, colMeans(fit_dedl$plugin[rows, ]))
  lr <- colMeans(as.matrix(lr_pred_all[rows, paste0("X", paper_order)]) - lr_pred_all[rows, "X000"])
  pdl <- colMeans(as.matrix(pdl_pred_all[rows, paste0("X", paper_order)]) - pdl_pred_all[rows, "X000"])
  list(n = sum(rows), truth = tr$diff, est = list(LA = la, LR = lr, PDL = pdl, SDL = sdl, DeDL = dedl))
}
cate_res <- lapply(sort(unique(grp)), cate_group)
names(cate_res) <- sort(unique(grp))
cate_metrics <- bind_rows(lapply(names(cate_res), function(g) {
  r <- cate_res[[g]]
  bind_rows(lapply(names(r$est), function(e) {
    err <- r$est[[e]] - r$truth
    idx_all <- 2:8; idx_u <- 6:8
    data.frame(group = g, n = r$n, estimator = e,
               mape_u = 100 * mean(abs(err[idx_u] / r$truth[idx_u])), mse_u = mean(err[idx_u]^2), mae_u = mean(abs(err[idx_u])),
               mape = 100 * mean(abs(err[idx_all] / r$truth[idx_all])), mse = mean(err[idx_all]^2), mae = mean(abs(err[idx_all])))
  }))
}))
tab5 <- cate_metrics %>% group_by(estimator) %>%
  summarise(across(c(mape_u, mse_u, mae_u, mape, mse, mae), ~ weighted.mean(.x, n)), .groups = "drop") %>%
  mutate(estimator = factor(estimator, levels = c("LA", "LR", "PDL", "SDL", "DeDL"))) %>% arrange(estimator)
```

```{r tab5, echo = FALSE}
tab5 %>%
  mutate(across(c(mape_u, mape), ~ fmt(.x, 2)), across(c(mse_u, mse), ~ formatC(.x * 1e4, digits = 3, format = "f")),
         across(c(mae_u, mae), ~ formatC(.x * 1e2, digits = 3, format = "f"))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Estimator", "MAPEu, \\% (1)", "MSEu $\\times 10^4$ (2)", "MAEu $\\times 10^2$ (3)", "MAPE, \\% (4)", "MSE $\\times 10^4$ (5)", "MAE $\\times 10^2$ (6)"),
      caption = "Comparison of different estimators of the CATE (paper's Table 5 layout): group-size-weighted averages over the four demographic groups.") %>%
  tab_style() %>%
  add_header_above(c(" " = 1, "Unobserved treatment combinations" = 3, "All treatment combinations" = 3))
```

```{r tab6, echo = FALSE}
sizes <- sapply(cate_res, `[[`, "n")
t6 <- cate_metrics %>%
  mutate(estimator = factor(estimator, levels = c("LA", "LR", "PDL", "SDL", "DeDL"))) %>%
  select(group, estimator, mape_u, mape) %>%
  pivot_longer(c(mape_u, mape), names_to = "set", values_to = "mape") %>%
  mutate(set = ifelse(set == "mape_u", "Unobserved treatment combinations", "All treatment combinations"),
         set = factor(set, levels = c("Unobserved treatment combinations", "All treatment combinations"))) %>%
  pivot_wider(names_from = group, values_from = mape) %>%
  arrange(set, estimator) %>%
  mutate(across(starts_with("G"), ~ fmt(.x, 2)))
kbl(t6 %>% mutate(set = as.character(set)), booktabs = TRUE, linesep = "", escape = FALSE,
    col.names = c("", "Estimator", paste0(names(sizes), ", \\% (", fmt0(sizes), ")")),
    caption = paste0("MAPE comparison of different estimators of the CATE for four groups (paper's Table 6 layout). Groups: ",
                     paste0(names(grp_names), " = ", grp_names, collapse = "; "), ". Group sizes in parentheses.")) %>%
  tab_style() %>%
  collapse_rows(columns = 1, valign = "top", latex_hline = "major")
```

```{r cate-numbers, echo = FALSE}
best_cate <- tab5$estimator[which.min(tab5$mape)]
```

Within groups the structured estimators keep their advantage over LA and LR;
`r as.character(best_cate)` ranks first on the size-weighted MAPE. Group G2
has near-zero true effects for some combinations, so every estimator's
percentage error there is large; the notebook's Table 6 shows the same
group with errors above 100 percent. The notebook's Table 5 code reports
only the last group's metrics (its loop overwrites the earlier groups), so
its printed SDL and DeDL values, 0.88 and 0.76 percent for the unobserved
combinations, correspond to our G4 column.

# Implementation lessons {#sec-lessons}

## Figure 7: the training error as the compass {#sec-epochs}

The paper's Figure 7 retrains the fold-1 network at learning rate $10^{-4}$
and evaluates DeDL and SDL after every epoch. The script writes $\hat\theta$
for the held-out fold after each epoch; `est_dml_structured()` is then called
fifty times.

```{r epochs, cache = TRUE, dependson = "benchmarks"}
split_file <- file.path(derived, "synthetic_fold1_split.csv")
if (!file.exists(split_file)) {
  write.csv(dat %>% select(-.row) %>% mutate(split = ifelse(kfold == 1, "pred", "train")), split_file, row.names = FALSE)
}
theta_final <- py_run(
  input = split_file, output = file.path(derived, "theta_epochs_final.csv"),
  history = file.path(derived, "history_epochs.csv"), epoch_theta = file.path(derived, "theta_epochs.csv.gz"),
  args = c("--model structured --y y --t is_DP,is_LP,is_FYP --x-all-others --train-col split --train-value train",
           "--train-cells 000,001,010,100,111 --arch two_branch --hidden 20 --depth 3",
           "--epochs 50 --batch 128 --lr 0.0001 --val-split 0.1 --seed 10234")
)
theta_epochs <- read.csv(gzfile(file.path(derived, "theta_epochs.csv.gz")))
# one-split runs score every row; keep the held-out fold only
theta_epochs <- theta_epochs[theta_epochs$.row %in% which(dat$kfold == 1), ]
history_epochs <- read.csv(file.path(derived, "history_epochs.csv"))
rows1 <- which(dat$kfold == 1)
lr_mape_fold1 <- combo_metrics(lr_folds[[1]]$table[-1, ], truth[-1, ])$mape
epoch_curve <- bind_rows(lapply(sort(unique(theta_epochs$epoch)), function(e) {
  te <- theta_epochs[theta_epochs$epoch == e, ]
  stopifnot(all(te$.row == rows1))
  f <- est_dml_structured(as.matrix(te[, theta_cols]), dat$y[rows1], dat[rows1, tvar], targets = paper_order[-1],
                          t_dist = t_dist_paper, ridge = ridge_main, contrast_best = FALSE)
  p <- f$estimates[f$estimates$fold == "pooled", ]
  data.frame(epoch = e,
             mape_dedl = mean(100 * abs(p$estimate - truth$diff[-1]) / abs(truth$diff[-1])),
             mape_sdl = mean(100 * abs(p$plugin - truth$diff[-1]) / abs(truth$diff[-1])))
})) %>% left_join(history_epochs %>% select(epoch, val_mse), by = "epoch") %>% mutate(mape_lr = lr_mape_fold1)
```

```{r fig7, echo = FALSE, fig.height = 3.8, fig.cap = "MAPE comparison with the training epoch (paper's Figure 7 layout): validation MSE of the fold-1 network (left axis) and the MAPE over all combinations of DeDL, SDL, and linear regression (right axis) after each epoch."}
plot_figure7(epoch_curve)
```

```{r epochs-numbers, echo = FALSE}
first_better <- epoch_curve$epoch[which(epoch_curve$mape_dedl < epoch_curve$mape_lr)[1]]
end_curve <- epoch_curve[nrow(epoch_curve), ]
```

At the slow learning rate the network is poorly trained for the first epochs
and both structured estimators are worse than linear regression; DeDL falls
below the LR line at epoch `r first_better` and ends at
`r fmt(end_curve$mape_dedl, 2)` percent against `r fmt(end_curve$mape_sdl, 2)`
for SDL. The paper's message is that the validation loss is the only observable
guide to whether the first stage is good enough for debiasing to help, and
that when the fit is poor, debiasing does not rescue it.

## The authors' notebook against the paper's algorithm {#sec-variants}

The released notebook differs from Algorithm 1 in four places.

1. It trains on all rows of the three training folds, so the "unobserved"
   combinations are in the training data.
2. The shares in $\Lambda(x)$ are taken from the first five rows of a
   groupby over all eight cells, in sorted order, which are not the five
   observed combinations and sum to about 5/8.
3. The factor 2 in $\Lambda(x)$ is dropped and a ridge of 0.5 is added.
4. The DeDL row of Table 3 concatenates fold 4 four times.

The next table isolates these choices with the same `est_dml_structured()`
call. The notebook's $\Lambda$ is reproduced by passing its shares halved as
weights (the function's factor 2 cancels the missing one) and its ridge.

```{r variants, cache = TRUE, dependson = "dedl-main"}
theta_all <- py_run(
  output = file.path(derived, "theta_allcells.csv"), history = file.path(derived, "history_allcells.csv"),
  args = c("--model structured --y y --t is_DP,is_LP,is_FYP --x-all-others --fold kfold",
           "--arch two_branch --hidden 20 --depth 3 --epochs 50 --batch 128 --lr 0.001 --val-split 0.1 --seed 1")
)
theta_all_mat <- as.matrix(theta_all[, theta_cols])
shares_sorted <- prop.table(table(key))[c("000", "001", "010", "011", "100")]
t_dist_notebook <- data.frame(combo_matrix(c("000", "001", "100", "010", "111")), prob = as.numeric(shares_sorted) / 2)
names(t_dist_notebook)[1:3] <- tvar
variant_mape <- function(theta, t_dist, ridge) {
  f <- suppressWarnings(est_dml_structured(theta, dat$y, dat[, tvar], targets = paper_order[-1], t_dist = t_dist,
                                           fold_id = dat$kfold, ridge = ridge, contrast_best = FALSE))
  p <- f$estimates[f$estimates$fold == "pooled", ]
  ape <- 100 * abs(p$estimate - truth$diff[-1]) / abs(truth$diff[-1]); aps <- 100 * abs(p$plugin - truth$diff[-1]) / abs(truth$diff[-1])
  c(dedl_all = mean(ape), dedl_u = mean(ape[5:7]), sdl_all = mean(aps), sdl_u = mean(aps[5:7]))
}
variants <- bind_rows(
  data.frame(variant = "Paper: observed cells, uniform shares, factor 2, ridge 0.01", t(variant_mape(theta_mat, t_dist_paper, ridge_main))),
  data.frame(variant = "Paper design with the notebook's ridge 0.5", t(variant_mape(theta_mat, t_dist_paper, 0.5))),
  data.frame(variant = "Notebook: all cells in training, uniform shares, factor 2, ridge 0.01", t(variant_mape(theta_all_mat, t_dist_paper, ridge_main))),
  data.frame(variant = "Notebook: all cells, misaligned shares, no factor 2, ridge 0.5", t(variant_mape(theta_all_mat, t_dist_notebook, 0.5)))
)
pdl_all <- py_run(
  output = file.path(derived, "pdl_allcells.csv"),
  args = c("--model pure --y y --t is_DP,is_LP,is_FYP --x-all-others --fold kfold",
           "--hidden 32 --depth 3 --epochs 50 --batch 128 --lr 0.001 --seed 1")
)
pdl_all_tab <- average_folds(lapply(sort(unique(pdl_all$fold)), function(f) pdl_estimator(pdl_all[pdl_all$fold == f, ])$table))
pdl_all_metrics <- metrics_table(list(PDL = pdl_all_tab), truth)
```

```{r tab-variants, echo = FALSE}
variants %>%
  mutate(across(-variant, ~ fmt(.x, 2))) %>%
  kbl(booktabs = TRUE, linesep = "", col.names = c("Variant", "DeDL, all", "DeDL, unobserved", "SDL, all", "SDL, unobserved"),
      caption = "MAPE (percent) of DeDL and SDL under the paper's algorithm and under the notebook's departures from it.") %>%
  tab_style() %>%
  column_spec(1, width = "7.5cm") %>%
  add_header_above(c(" " = 1, "MAPE, %" = 4))
```

```{r variants-numbers, echo = FALSE}
v <- variants
```

Training on all cells improves the plug-in on the unobserved combinations
(they are no longer unobserved), which is why the notebook's SDL looks so
good, and the heavy ridge with the misaligned shares shrinks the correction,
which is why its DeDL cannot improve on SDL. The same holds for the pure
network: trained on all cells it reaches a MAPE of
`r fmt(pdl_all_metrics$mape, 2)` percent over all combinations and
`r fmt(pdl_all_metrics$mape_u, 2)` on the "unobserved" ones (the notebook
prints 5.98 and 3.87), against `r g3("PDL", "mape")` and
`r g3("PDL", "mape_u")` when it must extrapolate. Under the paper's own
algorithm, with a small ridge, DeDL has MAPE `r fmt(v$dedl_u[1], 2)` percent
on the unobserved combinations against `r fmt(v$sdl_u[1], 2)` for SDL.

# Appendix D: synthetic experiments {#sec-appendix-d}

The paper's Appendix D validates DeDL on simulated combinatorial experiments.
We port the base design: $m = 4$ experiments, $d_c = 10$ uniform covariates,
outcome $c / (1 + \exp(-(\beta(x)'\tilde t))) + \varepsilon$ with
$\beta(x) = (x'C)^3$, $\tilde t = (1, t')'$, $C$ uniform on $[-0.5, 0.5]$, $c$
uniform on $[10, 20]$, uniform noise of $\pm 0.05$, six observed combinations
(baseline, four singles, and $(1, 1, 0, 0)$) each with probability $1/6$,
2,000 training and 2,000 estimation users per replication. The network is
the appendix's single-branch one (one hidden layer of 10 ReLU units, outputs
$\theta_0, \dots, \theta_4$, sigmoid, trained scale initialized at the
maximum outcome), full-batch Adam at learning rate 0.05 for 2,000 epochs,
restarted up to five times when the validation MSE exceeds 0.3. The
estimators are LA, LR, SDL, and DeDL, scored on the estimation users against
the true ATE function.

```{r appendix-d, cache = TRUE, cache.extra = params$n_reps_appendix}
n_reps <- params$n_reps_appendix
designs <- lapply(seq_len(n_reps), function(r) appendix_d_design(seed = 100 + r))
appd_long <- bind_rows(lapply(seq_len(n_reps), function(r) {
  bind_rows(designs[[r]]$train %>% mutate(split = "train"), designs[[r]]$est %>% mutate(split = "est")) %>% mutate(rep = r)
}))
appd_in <- file.path(derived, "appd_in.csv")
if (!file.exists(appd_in) || params$rerun_python) write.csv(appd_long, appd_in, row.names = FALSE)
appd_theta <- py_run(
  input = appd_in, output = file.path(derived, "appd_theta.csv"), history = file.path(derived, "appd_history.csv"),
  args = c("--model structured --y y --t t1,t2,t3,t4 --x x1,x2,x3,x4,x5,x6,x7,x8,x9,x10 --group rep --train-col split --train-value train",
           "--arch single --hidden 10 --depth 1 --epochs 2000 --batch 0 --lr 0.05 --val-split 0.3 --restarts 5 --val-threshold 0.3 --c-init max_y --seed 1")
)
appd_results <- bind_rows(lapply(seq_len(n_reps), function(r) {
  des <- designs[[r]]
  est_rows <- which(appd_long$rep == r & appd_long$split == "est")
  th <- appd_theta[match(est_rows, appd_theta$.row), ]
  theta <- as.matrix(th[, c("theta_0", paste0("theta_", 1:4), "theta_c")])
  tv <- paste0("t", 1:4)
  keys <- des$truth_keys
  obs_keys <- combo_key(des$obs_t)
  truth_r <- data.frame(combo = keys, diff = des$truth, p.value = ifelse(keys == "0000", 1, 0))
  tdist <- data.frame(des$obs_t, prob = 1 / nrow(des$obs_t))
  fit <- est_dml_structured(theta, des$est$y, des$est[, tv], targets = des$all_t[-1, ], t_dist = tdist, ridge = ridge_main, contrast_best = FALSE)
  p <- fit$estimates[fit$estimates$fold == "pooled", ]
  crit <- qnorm(0.975)
  dedl <- data.frame(combo = keys, estimate = c(0, p$estimate), conf.low = c(0, p$conf.low), conf.high = c(0, p$conf.high))
  sdl_se <- apply(fit$plugin, 2, sd) / sqrt(nrow(fit$plugin))
  sdl <- data.frame(combo = keys, estimate = c(0, p$plugin), conf.low = c(0, p$plugin - crit * sdl_se), conf.high = c(0, p$plugin + crit * sdl_se))
  # linear addition from the estimation sample's observed cells
  ekey <- combo_key(des$est[, tv])
  cell <- function(k) { v <- des$est$y[ekey == k]; c(mean = mean(v), var = var(v) / length(v)) }
  b <- cell("0000")
  la <- do.call(rbind, lapply(keys, function(k) {
    bits <- as.numeric(strsplit(k, "")[[1]])
    if (k %in% obs_keys && k != "0000") { s <- cell(k); est <- s["mean"] - b["mean"]; vr <- s["var"] + b["var"] }
    else if (k == "0000") { est <- 0; vr <- 0 }
    else { singles <- c("1000", "0100", "0010", "0001")[bits == 1]; est <- sum(sapply(singles, function(s) cell(s)["mean"] - b["mean"])); vr <- sum(sapply(singles, function(s) cell(s)["var"] + b["var"])) }
    data.frame(combo = k, estimate = unname(est), conf.low = unname(est) - crit * sqrt(unname(vr)), conf.high = unname(est) + crit * sqrt(unname(vr)))
  }))
  lr <- lr_estimator(des$train, des$est, "y", tv, paste0("x", 1:10), combos = keys)$table
  ests <- list(LA = la, LR = lr, SDL = sdl, DeDL = dedl)
  unobs <- setdiff(keys, obs_keys)
  bind_rows(lapply(names(ests), function(e) {
    tab <- ests[[e]][match(keys, ests[[e]]$combo), ]
    mu <- combo_metrics(tab[keys %in% unobs, ], truth_r[keys %in% unobs, ])
    ma <- combo_metrics(tab[keys != "0000", ], truth_r[keys != "0000", ])
    data.frame(rep = r, estimator = e, cdr_u = mu$cdr_share, mape_u = mu$mape, mse_u = mu$mse, mae_u = mu$mae,
               cdr = ma$cdr_share, mape = ma$mape, mse = ma$mse, mae = ma$mae)
  }))
}))
appd_tab <- appd_results %>% group_by(estimator) %>%
  summarise(across(c(cdr_u, mape_u, mse_u, mae_u, cdr, mape, mse, mae), mean), .groups = "drop") %>%
  mutate(estimator = factor(estimator, levels = c("LA", "LR", "SDL", "DeDL"))) %>% arrange(estimator)
appd_hist <- read.csv(file.path(derived, "appd_history.csv"))
```

```{r tab-appd, echo = FALSE}
appd_tab %>%
  mutate(across(c(cdr_u, cdr), ~ fmt(100 * .x, 1)), across(c(mape_u, mape), ~ fmt(.x, 2)),
         across(c(mse_u, mse, mae_u, mae), ~ fmt(.x, 3))) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE,
      col.names = c("Estimator", "CDRu, \\%", "MAPEu, \\%", "MSEu", "MAEu", "CDR, \\%", "MAPE, \\%", "MSE", "MAE"),
      caption = paste0("Synthetic experiments of the paper's Appendix D (base design, m = 4, ", n_reps,
                       " replications): mean metrics over replications for the 10 unobserved and all 15 non-baseline combinations. Effects are in outcome units (scale c between 10 and 20).")) %>%
  tab_style() %>%
  add_header_above(c(" " = 1, "Unobserved treatment combinations" = 4, "All treatment combinations" = 4))
```

```{r appd-numbers, echo = FALSE}
appd_final <- appd_hist %>% group_by(group) %>% summarise(val = val_mse[epoch == max(epoch)], restarts = max(restarts), .groups = "drop")
wins <- appd_results %>% select(rep, estimator, mape_u) %>% pivot_wider(names_from = estimator, values_from = mape_u) %>%
  summarise(dedl_beats_sdl = mean(DeDL < SDL), dedl_beats_lr = mean(DeDL < LR))
```

With interactions built into the data through the sigmoid, linear addition
and linear regression miss the unobserved combinations by a wide margin,
while the structured estimators do far better, and DeDL improves on SDL by
about a third in MAPE at the cost of a slightly lower correct-direction
ratio (its intervals are wider). DeDL has a lower MAPE than SDL on the
unobserved combinations in
`r round(100 * wins$dedl_beats_sdl)` percent of the replications; the
validation MSE of the networks ends between `r fmt(min(appd_final$val), 3)`
and `r fmt(max(appd_final$val), 3)`, and `r sum(appd_final$restarts > 1)`
replications needed a restart. The paper reports, with 200 replications and a
longer tuning loop, MAPE below one percent for DeDL; our smaller run
reproduces the ordering and the size of the gap to the linear benchmarks.

# What matched, what did not, and what we learned {#sec-summary}

```{r tab-match, echo = FALSE}
match_tab <- tibble::tribble(
  ~item, ~notebook, ~ours, ~status,
  "Table 2: relative effects (001, 010, 100, 111, 110, 101)", "0.519, -0.353, -0.358, 0.149, -0.585, 0.332",
  paste(fmt(truth$relative[2:7], 3), collapse = ", "), "exact",
  "Table 3: LA MAPE all / unobserved", "62.84 / 166.67", paste(fmt(tab3$mape[tab3$estimator == "LA"], 2), "/", fmt(tab3$mape_u[tab3$estimator == "LA"], 2)), "notebook mis-indexes the single effects",
  "Table 3: LR MAPE all / unobserved", "24.89 / 21.99", paste(fmt(tab3$mape[tab3$estimator == "LR"], 2), "/", fmt(tab3$mape_u[tab3$estimator == "LR"], 2)), "close (fold averaging)",
  "Table 3: PDL MAPE all / unobserved", "5.98 / 3.87", paste(fmt(pdl_all_metrics$mape, 2), "/", fmt(pdl_all_metrics$mape_u, 2)), "same magnitude once trained on all cells, as the notebook does",
  "Table 3: SDL MAPE all / unobserved", "2.79 / 1.95", paste(fmt(tab3$mape[tab3$estimator == "SDL"], 2), "/", fmt(tab3$mape_u[tab3$estimator == "SDL"], 2)), "same magnitude",
  "Table 3: DeDL MAPE all / unobserved", "9.04 / 6.16", paste(fmt(tab3$mape[tab3$estimator == "DeDL"], 2), "/", fmt(tab3$mape_u[tab3$estimator == "DeDL"], 2)), "ours lower; notebook departs from Algorithm 1",
  "Table 4: DeDL MAPE", "1.69", fmt(tab4$mape[tab4$estimator == "DeDL"], 2), "same magnitude",
  "Table 5, unobserved: SDL / DeDL MAPE", "0.88 / 0.76 (last group only)",
  paste(fmt(cate_metrics$mape_u[cate_metrics$group == "G4" & cate_metrics$estimator == "SDL"], 2), "/",
        fmt(cate_metrics$mape_u[cate_metrics$group == "G4" & cate_metrics$estimator == "DeDL"], 2), "(G4)"), "same magnitude"
)
kbl(match_tab, booktabs = TRUE, linesep = "", col.names = c("Quantity", "Authors' notebook", "This note", "Status"),
    caption = "Match between the authors' printed outputs on the synthetic data and this note.") %>%
  tab_style() %>%
  column_spec(1, width = "4.6cm") %>% column_spec(2, width = "3.6cm") %>% column_spec(3, width = "3.6cm") %>% column_spec(4, width = "3.8cm")
```

Three lessons for the workflow of double machine learning with a structured
first stage.

- **The influence function is only as good as $\Lambda(x)^{-1}$.** The
  orthogonal correction divides by the curvature of the loss in the nuisance
  direction. Where that curvature vanishes (saturated users), the correction
  is noise with infinite variance. A ridge is a bias-variance choice that the
  theory does not make for you; report the sensitivity.
- **The first-stage fit is observable, the nuisance error is not.** Figure 7
  is the practical test: watch the validation loss and do not debias a model
  that has not converged.
- **Read the code, not only the paper.** The released notebook's departures
  (training set, assignment shares, ridge, a copy-paste in the results table,
  a mis-indexed linear-addition benchmark) change the ranking of the
  estimators. The package function
  makes each choice an explicit argument, which is what a replication should
  produce.

# Appendix: formulas used {#sec-appendix}

## Gradients of the generalized sigmoid

For $u = \theta_0 + \sum_k \theta_k t_k$, $s = 1 / (1 + e^{-u})$, $G = c\,s$:
$\partial G / \partial \theta_0 = c\,s(1 - s)$,
$\partial G / \partial \theta_k = c\,s(1 - s)\,t_k$,
$\partial G / \partial c = s$. The loss gradient is
$\ell_\theta = 2\,G_\theta(\theta, \check t)\,(G(\theta, \check t) - y)$, and
$\Lambda(x) = 2 \sum_{t \in \mathcal{T}^o} \nu(t)\, G_\theta(\theta(x), t)\, G_\theta(\theta(x), t)'$.
The package's `link = "linear"` and `"logit"` replace $G$ by $u$ and by $s(u)$
with the corresponding gradients; `link = "custom"` takes both functions from
the user.

## The authors' inference conventions

- Ground truth and best treatment: cell-mean differences; pooled $t$-test if
  Levene's test does not reject at 5 percent, Welch otherwise.
- LA: the variance of a single-treatment effect is the sum of the two cell
  variances of the mean; combinations add the single effects and their
  variances.
- LR and PDL: the contrast of predicted outcomes has standard error
  $\mathrm{sd}(\hat y_t - \hat y_0) / \sqrt{n}$ and its $p$-value is a Welch
  test between the two prediction vectors.
- DeDL and SDL: fold means and standard errors of $\psi$ (or of $H$); the
  reported interval averages the fold bounds.
- Metrics: a combination with an insignificant ground truth contributes zero
  to MAPE, MSE, and MAE and counts as correctly identified when the estimate's
  interval covers zero; otherwise the estimate must be significant with the
  right sign.

## References

- Chernozhukov, Victor, Denis Chetverikov, Mert Demirer, Esther Duflo, Christian Hansen, Whitney Newey, and James Robins. 2018. "Double/Debiased Machine Learning for Treatment and Structural Parameters." *The Econometrics Journal*.
- Farrell, Max H., Tengyuan Liang, and Sanjog Misra. 2020. "Deep Learning for Individual Heterogeneity: An Automatic Inference Framework." arXiv:2010.14694.
- Farrell, Max H., Tengyuan Liang, and Sanjog Misra. 2021. "Deep Neural Networks for Estimation and Inference." *Econometrica*.
- Ye, Zikun, Zhiqi Zhang, Dennis J. Zhang, Heng Zhang, and Renyu Zhang. 2025. "Deep Learning-Based Causal Inference for Large-Scale Combinatorial Experiments: Theory and Empirical Evidence." *Management Science*.
