---
title: "causalmetrics lecture note, Section 4"
subtitle: "Doubly Robust Estimation and Double Machine Learning"
author: "Weiheng Zhang"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
    toc_depth: 4
    keep_tex: false
    latex_engine: xelatex
header-includes:
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{float}
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "80%",
  dpi       = 150,
  message   = FALSE,
  warning   = FALSE,
  fig.pos   = "H"
)
# Code and output in footnotesize.
default_chunk_hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  paste0("\\footnotesize\n\n", x, "\n\n\\normalsize\n")
})

# This lecture lives in inst/lectures/ of the package source. Knitted from the
# source tree it loads the development version of the package; when the
# package is installed it uses the installed copy and its helpers.
pkg_root <- normalizePath(file.path("..", "..", ".."), mustWork = FALSE)
if (file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  devtools::load_all(pkg_root, quiet = TRUE)
  helper_dir <- file.path(pkg_root, "inst", "vignette_helpers")
} else {
    helper_dir <- system.file("vignette_helpers", package = "causalmetrics")
}

library(causalmetrics)
library(mlr3)
library(mlr3learners)
library(kableExtra)

# Simulation helpers for this lecture (live in inst/vignette_helpers/)
helper_path <- file.path(helper_dir, "04_Doubly_Robust_and_Double_ML", "dml_simulation_helpers.R")
source(helper_path)

# Monte Carlo size for the opening experiment. 100 replications take about
# two minutes with random forests; the chunk is cached.
n_sims <- 100
n_sims_lasso <- 60
fmt <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
```

# The problem {#sec-problem}

## Same identification, a harder estimation problem

Section 03 ended with five estimators for one estimand. All five assumed
the same thing: conditional ignorability, $(Y(1), Y(0)) \perp\!\!\!\perp D
\mid X$, plus overlap. They differed only in *how* they used two objects.
This lecture calls them the **nuisance functions**,

$$
m(X) = \mathbb{E}[D \mid X], \qquad g(d, X) = \mathbb{E}[Y \mid D = d, X],
$$

together with their residual-based cousin $\ell(X) = \mathbb{E}[Y \mid X]$.
Regression adjustment imputes with $g$. Weighting rebalances with $m$. AIPW
uses both. In Section 03 these functions were quadratic polynomials in one
covariate. That was enough to teach the geometry. It is not enough for the
data this course is about.

Two features of modern applied work break the Section 03 toolkit at the
estimation stage, not at the identification stage.

- **Many covariates.** A customer record carries purchase histories at
  many horizons, engagement counts, geography, device, and tenure. A
  firm-level panel has dozens of balance-sheet ratios. Text and image
  embeddings add hundreds of coordinates. With $p$ comparable to $n$, least
  squares on everything is noisy or impossible. Choosing controls by hand
  brings back the discretion that a pre-specified adjustment set was meant
  to remove.
- **Unknown functional form.** Even with few covariates, nothing says the
  outcome or the assignment rule is linear in them. Targeting systems,
  eligibility rules, and habit formation produce thresholds, curvature, and
  interactions. A misspecified linear adjustment is biased. No standard
  error reports that bias.

The instinct of anyone who has fitted a random forest is obvious: *let the
machine learn $g$ and $m$.* Prediction is what these methods are good at.
Both nuisance functions are conditional expectations, so they are
prediction targets. This lecture follows that instinct. It shows why the
obvious way of doing it fails, and what the repair looks like. The repair
is called **double/debiased machine learning**, DML for short.

One sentence to hold on to: **prediction is the means, not the goal.** The
target is a single number $\theta$, a causal effect, together with a
confidence interval that has its nominal coverage. Every choice below is
judged by that standard, not by out-of-sample fit.

### Notation

We observe $W = (Y, D, X)$. Two working models carry the lecture.

- The **partially linear model** (PLM):
  $$
  Y = \theta D + g(X) + \varepsilon, \quad \mathbb{E}[\varepsilon \mid D, X] = 0;
  \qquad
  D = m(X) + v, \quad \mathbb{E}[v \mid X] = 0.
  $$
  The effect is a constant $\theta$; $D$ may be binary or continuous; $g$
  and $m$ are unrestricted.
- The **interactive regression model** (IRM): $Y = g(D, X) + \varepsilon$
  with binary $D$ and no restriction on how $D$ and $X$ interact. The
  target is the ATE, $\theta = \mathbb{E}[g(1, X) - g(0, X)]$, or the ATT.

In both models $\theta$ is a **low-dimensional target** and the functions
$(g, m)$ or $(\ell, m)$ are **infinite-dimensional nuisances**. Under
conditional ignorability the PLM coefficient and the IRM contrast are
causal effects; without it they are still well-defined predictive
quantities, and DML estimates them either way. The identification argument
is the one from Section 03 and is not repeated here.

## The natural approach: plug a learner into last lecture's formula

Take the imputation view of Section 03 and replace the linear model with a
forest. Fit one random forest of $Y$ on $(D, X)$, predict every unit twice,
once with $D$ set to 1 and once with $D$ set to 0, and average the
difference:

$$
\hat\theta_{\text{plug-in}}
\;=\; \frac{1}{n} \sum_{i=1}^{n}
\bigl\{ \hat g(1, X_i) - \hat g(0, X_i) \bigr\}.
$$

This is regression adjustment with a flexible learner (an "S-learner" in
the heterogeneous-effects literature). It is tempting for three reasons.

- Forests are consistent for $g$ under mild conditions.
- No functional form is assumed.
- No propensity model is needed.

A naive analyst reports the standard deviation of the imputed differences
divided by $\sqrt{n}$, as for any sample mean.

```{r naive-one-sample}
dat <- simulate_dml_dgp(n = 500, p = 10, theta = 1, seed = 1)
x_names <- paste0("x", 1:10)

set.seed(1)
naive_plugin_forest(dat, x_names, num.trees = 100)
```

The truth is 1. The estimate is far below it, and the reported standard
error would not have warned you. Whether this is bad luck or a systematic
failure is a question for a simulation.

## One experiment, six estimators

### The design

Ten standard normal covariates; three of them matter. The propensity score
and the outcome surface are nonlinear in the *same* covariates. So the
confounding is real, and a linear adjustment is misspecified:

$$
m_0(X) = \Lambda\bigl(0.8 X_1 - 0.4 (X_2^2 - 1)\bigr),
\qquad
g_0(X) = 2\,\Lambda(2 X_1) + 0.5 X_2^2 + 0.25 X_3,
$$

$$
D \sim \text{Bernoulli}\bigl(m_0(X)\bigr),
\qquad
Y = \theta D + g_0(X) + \varepsilon,
\qquad
\varepsilon \sim N(0, 1),
\qquad
\theta = 1,
$$

where $\Lambda$ is the logistic function. The effect is constant, so the
ATE, the ATT, and the partially linear coefficient coincide and every
estimator below targets the same number. Each replication draws $n = 500$
units; there are `r n_sims` replications.

### The estimators

```{r pipe-table-1, echo = FALSE}
tibble::tibble(
  c1 = c("OLS with linear controls", "Naive plug-in forest", "Naive plug-in forest, cross-fitted", "Orthogonal score, full-sample forests", "Orthogonal score, cross-fitted", "AIPW, cross-fitted"),
  c2 = c("none", "$\\hat g(D, X)$", "$\\hat g(D, X)$", "$\\hat\\ell(X), \\hat m(X)$", "$\\hat\\ell(X), \\hat m(X)$", "$\\hat m(X), \\hat g(0, X), \\hat g(1, X)$"),
  c3 = c("full sample", "full sample", "held-out folds", "full sample", "held-out folds", "held-out folds"),
  c4 = c("Section 03's regression adjustment", "\"replace the linear model by a forest\"", "the wrong score with the right sample splitting", "the right score with the wrong sample use", "DML with the partially linear score", "Section 03's AIPW with forests: DML in the interactive model")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Estimator", "Nuisances", "Fitted on", "What it stands for")) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  column_spec(1, width = "4.3cm") %>%
  column_spec(2, width = "3.6cm") %>%
  column_spec(4, width = "5.6cm")
```

The last three are `est_dml()` calls.
The cross-fitted plug-in is there to separate two ideas that are easy to
conflate: sample splitting and the choice of score. "Orthogonal score" means the
residual-on-residual regression of Section 03's Series R: predict $Y$ and
$D$ from $X$, then regress $Y - \hat\ell(X)$ on $D - \hat m(X)$. Why that
combination deserves the name is the subject of Section
\@ref(sec-framework). "Cross-fitted" means each unit's nuisance
predictions come from forests that never saw that unit. That is the recipe
of Section 03's Appendix C and the subject of Section
\@ref(sec-crossfit). All forests use 100 trees and default tuning. The
point is not to tune well. It is to see what tuning cannot fix.

```{r dml-experiment, cache = TRUE, cache.extra = tools::md5sum(helper_path)}
results <- run_dml_experiment(
  n_sims = n_sims, n = 500, p = 10, theta = 1,
  num.trees = 100, folds = 5, seed = 1
)
tab <- summarise_dml_experiment(results)
pick <- function(estimator, column) tab[[column]][tab$estimator == estimator]
```

```{r tab-experiment, echo = FALSE}
kbl(tab, booktabs = TRUE, digits = 3, linesep = "", escape = FALSE,
    col.names = c("Estimator", "Bias", "SD", "RMSE", "Mean reported SE", "95\\% coverage"),
    caption = "Monte Carlo comparison, true effect 1, n = 500. Coverage is the share of replications whose nominal 95 percent interval contains the truth.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

```{r fig-experiment, fig.height = 8, fig.cap = "Sampling distributions of the estimators, centred at the truth. The curve in each panel is the normal approximation implied by the estimator's own average reported standard error: a histogram that is shifted relative to its curve is biased, one that is wider than its curve under-reports uncertainty."}
plot_dml_experiment(results)
```

### Reading the results

- **Linear adjustment fails as expected.** OLS with all ten covariates
  has a bias of `r fmt(pick("OLS with linear controls", "bias"))` and
  coverage of `r fmt(pick("OLS with linear controls", "coverage"))`. The
  functional form is wrong, and more covariates would not help. This is
  the problem the forest was supposed to solve.
- **The plug-in forest is worse, not better.** Its bias,
  `r fmt(pick("Naive plug-in forest", "bias"))`, is of the same size as
  OLS's, and its reported standard error,
  `r fmt(pick("Naive plug-in forest", "mean_se"), 3)`, is a fraction of
  the actual dispersion, `r fmt(pick("Naive plug-in forest", "sd"), 3)`,
  so the interval almost never covers the truth. The forest shrinks the
  treatment contrast. At every split $D$ competes against ten covariates.
  Splits on $X_1$ and $X_2$ predict $Y$ almost as well, so they absorb
  most of the effect of $D$. A learner tuned for prediction has no reason
  to protect the coefficient we care about.
- **Sample splitting alone does not help.** The cross-fitted plug-in
  keeps a bias of `r fmt(pick("Naive plug-in forest, cross-fitted", "bias"))`,
  and its reported standard error is just as misleading. The moment it
  solves is the wrong one. Cross-fitting cures a different disease.
- **The right score with the wrong sample splitting is still biased.**
  Fitting the orthogonal score's forests on the full sample gives a bias
  of `r fmt(pick("Orthogonal score, full-sample forests", "bias"))` and
  coverage of
  `r fmt(pick("Orthogonal score, full-sample forests", "coverage"))`. The
  residuals $Y - \hat\ell$ and $D - \hat m$ are too small and are
  correlated with the noise they were fitted to.
- **Cross-fitted DML recovers the target.** Both cross-fitted estimators
  are centred close to the truth (bias
  `r fmt(pick("Orthogonal score, cross-fitted", "bias"))` and
  `r fmt(pick("AIPW, cross-fitted", "bias"))`), their reported standard
  errors match the actual dispersion, and their coverage is close to
  nominal. The AIPW row is the estimator of Section 03 with forest
  nuisances: cross-fitted AIPW *is* DML for the ATE.
- **A small gap remains.** The cross-fitted rows are not exactly centred.
  That residual is nuisance estimation error leaking into $\hat\theta$ at
  second order. Section \@ref(sec-rates) shows why it shrinks with the
  accuracy of the learners, not with the size of the sample alone.

## What went wrong: two biases

The plug-in estimator fails for two separate reasons, and it is worth
keeping them apart because they have separate fixes.

### Regularization bias

Machine learning estimators of a conditional expectation are biased by
construction. Shrinkage, pruning, minimum leaf sizes, early stopping, and
penalties all trade bias for variance. The resulting error
$\hat g - g_0$ vanishes more slowly than $n^{-1/2}$. That is harmless for
prediction. It is fatal for the plug-in, because the plug-in moment
condition,

$$
M(\theta, g) = \mathbb{E}[g(1, X) - g(0, X)] - \theta,
$$

moves one-for-one with the nuisance. Perturb $g_0$ in a direction
$\Delta$. The moment moves by $\mathbb{E}[\Delta(1, X) - \Delta(0, X)]$,
which is not zero in general. The naive partially linear moment
$\mathbb{E}[(Y - \theta D - g(X))\, D] = 0$ has the same flaw; its
derivative in the direction $\Delta$ is $-\mathbb{E}[\Delta(X) D]$. So the
learner's bias in $\hat g$ passes undiminished to $\hat\theta$. That bias
shrinks slower than the standard error, so the confidence interval never
catches up. The repair is a moment whose derivative with respect to every
nuisance is zero at the truth. That property is **Neyman orthogonality**,
the subject of Section \@ref(sec-framework). The residual-on-residual
score and the AIPW score both have it.

### Overfitting bias

Orthogonality removes the *first-order* effect of nuisance error. It does
not help when the nuisance estimate is correlated with the observations on
which the score is evaluated. A forest fitted on the full sample
reproduces part of each unit's noise $\varepsilon_i$ in its own prediction
$\hat g(X_i)$. The residual $Y_i - \hat g(X_i)$ then contains too little
of $\varepsilon_i$. The residual $D_i - \hat m(X_i)$ contains too little
of $v_i$. The score averages their product, which is biased. The more
flexible the learner, the worse this "own-observation" bias.
**Cross-fitting** breaks the dependence: the nuisances are fitted on folds
that exclude the unit being scored. Section \@ref(sec-crossfit) shows why
this is enough for the estimator to behave as if the true nuisances had
been plugged in.

### The remaining gap

With an orthogonal score and cross-fitting, nuisance error enters
$\hat\theta$ only at second order. The remainders are products such as
$\|\hat\ell - \ell_0\| \cdot \|\hat m - m_0\|$ and, for the partialling-out
score, $\|\hat m - m_0\|^2$. They vanish at the $\sqrt{n}$ rate only if
each nuisance converges faster than $n^{-1/4}$. In the experiment the
forest's propensity error is close to that threshold at $n = 500$. That is
why the cross-fitted rows still show a small negative bias. Better
learners, more data, or a score that is robust to the harder nuisance all
shrink it. Section \@ref(sec-rates) quantifies this, and Section
\@ref(sec-recipe) turns it into a workflow for choosing learners.

## When this is the right tool

DML is an estimation layer. It sits on top of an identification strategy
and changes nothing about it. Reach for it when all four conditions hold.

- **The design is credible without it.** Conditional ignorability with a
  pre-treatment adjustment set (this lecture). Or another design whose
  estimator contains nuisance functions: instrumental variables (Section
  05), difference-in-differences (Section 07), policy learning (Section 08).
- **The nuisance functions are hard.** Many covariates, unknown
  functional forms, interactions, or engineered features such as
  embeddings. With three covariates and a plausible linear model, Section
  03's tools are simpler and equally valid.
- **The target is low-dimensional.** An ATE, an ATT, a partially linear
  coefficient, a group-average effect, or a small vector of such numbers.
  DML delivers $\sqrt{n}$ inference for these. It does not deliver a
  nonparametric function with the same guarantees.
- **There is enough data to learn the nuisances.** The rate condition is a
  real constraint. A few hundred observations and a forest on fifty
  covariates will not satisfy it. The residual bias above will then be
  large.

Do not expect it to do four other things. It does not choose the
adjustment set; a bad control stays bad. It does not repair a lack of
overlap; extreme propensities remain extreme, and trimming still changes
the estimand. It does not fix unobserved confounding. It does not estimate
heterogeneous effects for individual units. The typical applications in
this course fit the four conditions: targeted promotions with rich
customer histories (Ellickson, Kar, and Reeder 2023), retirement-plan
eligibility with income and demographics (Chernozhukov et al. 2018), and
price elasticities with high-dimensional product features.

## Doubly robust, double machine learning

The two names in this lecture's title are related but not synonyms.

- **Doubly robust** is a property of a *score*. The AIPW score of Section
  03 is consistent for the ATE if either the outcome model or the
  propensity model is correct. Its bias is the product of the two errors.
- **Double machine learning** is a *recipe* with three ingredients: a
  Neyman-orthogonal score, cross-fitted nuisance estimation, and learners
  that converge fast enough. The AIPW score is one orthogonal score.
  Cross-fitted AIPW with machine-learned nuisances is DML for the ATE. The
  residual-on-residual score of the partially linear model is another
  orthogonal score. It is *not* doubly robust: its bias is second order
  in the propensity error alone.

The rest of the lecture develops the recipe.

- Section \@ref(sec-framework): orthogonality, the proof skeleton, and the
  rate condition.
- Section \@ref(sec-views): three views of the score. The partially linear
  model as a generalized Frisch-Waugh-Lovell theorem, the interactive
  model with the doubly robust score, and double lasso, where the story
  started.
- Section \@ref(sec-crossfit): cross-fitting.
- Section \@ref(sec-recipe): the recipe as a workflow. Folds, pooled
  versus fold-averaged solutions, repeated splits, learner selection.
- Section \@ref(sec-diagnostics): diagnostics and failure modes.
- Section \@ref(sec-practice): the replications and a checklist.
- Section \@ref(sec-extensions): where the recipe goes next.

# Just enough framework {#sec-framework}

## Targets, nuisances, moments

Everything in this lecture has the same shape.

- A **target** $\theta_0$, a number or a short vector.
- A **nuisance** $\eta_0$, a function or a tuple of functions.
- A **score** $\psi(W; \theta, \eta)$ with moment
  $$
  M(\theta, \eta) := \mathbb{E}\bigl[\psi(W; \theta, \eta)\bigr],
  \qquad M(\theta_0, \eta_0) = 0,
  $$
  and $\theta_0$ is the unique root at the true nuisance.

The estimator replaces $\eta_0$ by a learned $\hat\eta$ and $\mathbb{E}$ by
the sample mean $\mathbb{E}_n$, then solves
$\mathbb{E}_n[\psi(W; \hat\theta, \hat\eta)] = 0$. The whole question is
how the error $\hat\eta - \eta_0$ enters $\hat\theta$.

Three scores from Sections 1 and 03, with $H(D, X) = D/m(X) - (1 - D)/(1 - m(X))$:

```{r pipe-table-2, echo = FALSE}
tibble::tibble(
  c1 = c("Plug-in (regression imputation)", "Partialling out (PLM)", "AIPW (IRM)"),
  c2 = c("$g$", "$(\\ell, m)$", "$(g, m)$"),
  c3 = c("$g(1, X) - g(0, X) - \\theta$", "$\\bigl(Y - \\ell(X) - \\theta\\,(D - m(X))\\bigr)\\,(D - m(X))$", "$g(1, X) - g(0, X) + H(D, X)\\,\\bigl(Y - g(D, X)\\bigr) - \\theta$")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Score", "Nuisance $\\eta$", "$\\psi(W; \\theta, \\eta)$")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

## Neyman orthogonality {#sec-orthogonality}

Fix a direction $\Delta = \eta - \eta_0$. The **directional (Gateaux)
derivative** of the moment at the truth is
$$
\partial_\eta M(\theta_0, \eta_0)[\Delta]
\;:=\; \frac{d}{dt}\, M(\theta_0, \eta_0 + t\Delta)\Big|_{t = 0}.
$$

> **Definition (Neyman orthogonality).** The score is Neyman orthogonal if
> $\partial_\eta M(\theta_0, \eta_0)[\Delta] = 0$ for every admissible
> direction $\Delta$.

In words: a small error in the nuisance has no first-order effect on the
moment. Only second-order terms survive. Check the three scores; one line
each, full derivations in Appendix \@ref(app-orthogonality).

- **Plug-in.** $\partial_\eta M[\Delta] = \mathbb{E}[\Delta(1, X) - \Delta(0, X)]$.
  Not zero. Not orthogonal.
- **Partialling out.** Direction $\Delta_\ell$ in $\ell$:
  $-\mathbb{E}[\Delta_\ell(X)\,(D - m_0(X))] = -\mathbb{E}\bigl[\Delta_\ell(X)\,\mathbb{E}[D - m_0(X) \mid X]\bigr] = 0$.
  Direction $\Delta_m$ in $m$: the derivative is
  $-\mathbb{E}[\Delta_m(X)\, U] + \theta_0\,\mathbb{E}[\Delta_m(X)(D - m_0(X))]$
  with $U = Y - \ell_0(X) - \theta_0 (D - m_0(X))$; both terms vanish
  because $\mathbb{E}[U \mid D, X] = 0$ and $\mathbb{E}[D - m_0 \mid X] = 0$.
  Orthogonal.
- **AIPW.** Direction $\Delta_g$ in $g$:
  $\mathbb{E}[\Delta_g(1, X) - \Delta_g(0, X)] - \mathbb{E}[H_0(D, X)\,\Delta_g(D, X)] = 0$,
  because $\mathbb{E}[H_0\,\Delta_g(D, X)] = \mathbb{E}[\Delta_g(1, X)] - \mathbb{E}[\Delta_g(0, X)]$
  (the inverse-propensity identity). Direction $\Delta_m$ in $m$: the
  derivative multiplies $Y - g_0(D, X)$, whose conditional mean is zero.
  Orthogonal.

The first bullet is Section 1's regularization bias in one line. The other
two are the reason the orthogonal estimators recovered the target.

## Why orthogonality is enough: skeleton of the proof {#sec-skeleton}

Take a score that is linear in $\theta$, $\psi = \psi^b(W; \eta) - \psi^a(W; \eta)\,\theta$.
All three scores above have this form. Then
$$
\hat\theta = \frac{\mathbb{E}_n[\psi^b(W; \hat\eta)]}{\mathbb{E}_n[\psi^a(W; \hat\eta)]},
\qquad
\sqrt{n}\,(\hat\theta - \theta_0)
= \frac{\sqrt{n}\;\mathbb{E}_n[\psi(W; \theta_0, \hat\eta)]}{\mathbb{E}_n[\psi^a(W; \hat\eta)]}.
$$
The denominator converges to $J_0 = \mathbb{E}[\psi^a(W; \eta_0)]$ once
$\hat\eta$ is consistent. Everything happens in the numerator. Add and
subtract the oracle score $\psi(W; \theta_0, \eta_0)$:
$$
\sqrt{n}\,\mathbb{E}_n[\psi(\theta_0, \hat\eta)]
= \underbrace{\sqrt{n}\,\mathbb{E}_n[\psi(\theta_0, \eta_0)]}_{(a)\ \text{oracle}}
+ \underbrace{\sqrt{n}\,(\mathbb{E}_n - \mathbb{E})\bigl[\psi(\theta_0, \hat\eta) - \psi(\theta_0, \eta_0)\bigr]}_{(b)\ \text{empirical process}}
+ \underbrace{\sqrt{n}\,\mathbb{E}\bigl[\psi(\theta_0, \hat\eta) - \psi(\theta_0, \eta_0)\bigr]}_{(c)\ \text{bias}},
$$
where $\mathbb{E}$ in $(b)$ and $(c)$ treats $\hat\eta$ as fixed.

**Step 1: the oracle term $(a)$.** An average of i.i.d. mean-zero terms.
By the central limit theorem,
$(a) \to N\bigl(0, \mathbb{E}[\psi(W; \theta_0, \eta_0)^2]\bigr)$. This is
what an estimator with known nuisances would deliver.

**Step 2: the bias term $(c)$.** Expand $M(\theta_0, \cdot)$ around
$\eta_0$:
$$
(c) = \sqrt{n}\Bigl(
\underbrace{\partial_\eta M(\theta_0, \eta_0)[\hat\eta - \eta_0]}_{=\,0\ \text{by orthogonality}}
+ \tfrac12\, \partial^2_\eta M(\theta_0, \bar\eta)[\hat\eta - \eta_0, \hat\eta - \eta_0]
\Bigr).
$$
Orthogonality removes the first-order term. The second-order term is
bounded by $C\,\|\hat\eta - \eta_0\|^2$, or by a product of the component
errors. So $(c) = o_p(1)$ if $\|\hat\eta - \eta_0\| = o_p(n^{-1/4})$. That
is the **rate condition**.

**Step 3: the empirical process term $(b)$.** This is a centred average
of a function that depends on $\hat\eta$, which was fitted on the same
data. Without sample splitting it vanishes only if the learner lives in a
"not too complex" (Donsker) class. Adaptive learners such as forests and
boosting do not. **Cross-fitting** makes $\hat\eta$ independent of the fold
on which it is evaluated. Conditional on the training folds, $(b)$ has
mean zero and variance $\mathbb{E}[(\psi(\hat\eta) - \psi(\eta_0))^2]$,
which vanishes as soon as $\hat\eta$ is consistent. Section
\@ref(sec-crossfit) gives the argument.

**Conclusion.**
$$
\sqrt{n}\,(\hat\theta - \theta_0) \;\to\; N(0, V),
\qquad
V = \frac{\mathbb{E}[\psi(W; \theta_0, \eta_0)^2]}{J_0^2},
$$
and $V$ is estimated by plugging $\hat\eta$ and $\hat\theta$ into the same
formula. Appendix \@ref(app-proof) states the conditions and fills in the
steps.

> **The three ingredients, mapped to the proof.**
> An orthogonal score kills the first-order part of $(c)$.
> The rate condition kills the second-order part of $(c)$.
> Cross-fitting kills $(b)$.
> Drop any one and the corresponding term reappears.

Two consequences matter in practice.

- **Oracle equivalence.** $\hat\theta$ behaves as if the true nuisances
  had been plugged in. Standard errors are computed from the score with
  $\hat\eta$ inside, and no correction for the first stage is needed.
- **Learner agnosticism.** Nothing in the argument names the learner.
  Any method with the rate works, and different nuisances may use different
  methods.

## The rate condition and double robustness {#sec-rates}

The second-order remainder of Step 2 has a concrete form for each score
(Appendix \@ref(app-orthogonality)).

- **Partialling out.**
  $$
  \mathbb{E}[\psi(\theta_0, \hat\eta)]
  = \mathbb{E}\bigl[(\hat\ell - \ell_0)(\hat m - m_0)\bigr]
  - \theta_0\, \mathbb{E}\bigl[(\hat m - m_0)^2\bigr].
  $$
  The bias of $\hat\theta$ is this quantity divided by $J_0 = \mathbb{E}[(D - m_0)^2]$.
  The rate condition is $\|\hat\ell - \ell_0\|\,\|\hat m - m_0\| + \|\hat m - m_0\|^2 = o(n^{-1/2})$.
  The propensity error enters *squared* and on its own. Nothing protects
  against a poor $\hat m$.
- **AIPW.**
  $$
  \mathbb{E}[\psi(\theta_0, \hat\eta)]
  = \mathbb{E}\Bigl[\tfrac{\hat m - m_0}{\hat m}\,\bigl(\hat g(1, X) - g_0(1, X)\bigr)\Bigr]
  + \mathbb{E}\Bigl[\tfrac{\hat m - m_0}{1 - \hat m}\,\bigl(\hat g(0, X) - g_0(0, X)\bigr)\Bigr].
  $$
  Every term is a *product* of a propensity error and an outcome error.
  The rate condition is $\|\hat m - m_0\|\,\|\hat g - g_0\| = o(n^{-1/2})$.
  If either factor is zero the remainder is zero: that is **double
  robustness**, now seen as a special structure of the remainder rather
  than a separate property.

Section 1's residual bias is this remainder at work. Two cross-fitted
fits on the design of Section 1, at $n = 500$ and $n = 2000$, show the
propensity error against the $n^{-1/4}$ threshold and the bias it implies
through the partialling-out formula.

```{r rate-check}
rate_check <- function(n, seed) {
  dat <- simulate_dml_dgp(n = n, p = 10, theta = 1, seed = seed)
  set.seed(seed)
  fit <- est_dml(
    dat, y = "y", d = "d", x = x_names,
    learner_l = lrn("regr.ranger", num.trees = 100),
    learner_m = lrn("classif.ranger", predict_type = "prob", num.trees = 100),
    folds = 5, seed = seed
  )
  rmse_m <- sqrt(mean((fit$nuisance$m_hat - dat$m_true)^2))
  j0 <- fit$diagnostics$identification$mean_sq_treatment_residual
  c(n = n, rmse_m = rmse_m, threshold = n^(-1/4),
    implied_bias = -1 * rmse_m^2 / j0, estimate = fit$estimate)
}
round(rbind(rate_check(500, 1), rate_check(2000, 1)), 3)
```

At $n = 500$ the forest's propensity error is of the same order as
$n^{-1/4}$, and the implied bias matches the Monte Carlo bias of
`r fmt(pick("Orthogonal score, cross-fitted", "bias"))` in Table
\@ref(tab:tab-experiment) in sign and size. At $n = 2000$ the error is
smaller, but it shrinks slowly, because a forest's rate is well below
$n^{-1/2}$. Two lessons follow. The learner for $m$ deserves as much care
as the learner for $\ell$. And when one nuisance is much harder than the
other, a doubly robust score, which only needs the product to be small,
is the safer choice.

# Three views of the score {#sec-views}

The orthogonal score appears in three settings. Each gives a different
picture of the same idea. Each subsection follows the same order:
*model, theorem or score, estimator, R, when it breaks.*

## Partially linear model: generalized Frisch-Waugh-Lovell {#sec-plm}

**Model.**
$$
Y = \theta D + g(X) + \varepsilon, \quad \mathbb{E}[\varepsilon \mid D, X] = 0;
\qquad
D = m(X) + v, \quad \mathbb{E}[v \mid X] = 0.
$$
$D$ may be binary or continuous. Only $\theta$ is restricted to be a
constant. Define $\ell(X) = \mathbb{E}[Y \mid X] = \theta\, m(X) + g(X)$
and the residuals $\tilde Y = Y - \ell(X)$, $\tilde D = D - m(X)$.
Subtracting $\ell(X)$ from both sides of the model gives
$$
\tilde Y = \theta\, \tilde D + \varepsilon .
$$
This is Section 03's Series R, now as a statement about the population.

> **Theorem (partialling out in the PLM).** If $\mathbb{E}[\tilde D^2] > 0$,
> $$
> \theta = \frac{\mathbb{E}[\tilde D \tilde Y]}{\mathbb{E}[\tilde D^2]} .
> $$
> *Proof.* Multiply $\tilde Y = \theta \tilde D + \varepsilon$ by $\tilde D$
> and take expectations. The error term drops out because
> $\mathbb{E}[\varepsilon \tilde D] = \mathbb{E}\bigl[\mathbb{E}[\varepsilon \mid D, X]\,(D - m(X))\bigr] = 0$. $\square$

With linear $\ell$ and $m$ this is the Frisch-Waugh-Lovell theorem of
Section 02. The nuisances are now arbitrary functions, and the residuals
must be learned.

**Score and estimator.** The score $\psi = (\tilde Y - \theta \tilde D)\tilde D$
is orthogonal (Section \@ref(sec-orthogonality)). With cross-fitted
residuals $\check Y_i = Y_i - \hat\ell_{[k(i)]}(X_i)$ and
$\check D_i = D_i - \hat m_{[k(i)]}(X_i)$,
$$
\hat\theta = \frac{\sum_i \check D_i \check Y_i}{\sum_i \check D_i^2},
\qquad
\widehat{\mathrm{SE}}(\hat\theta) =
\frac{\sqrt{\sum_i \check D_i^2\, \hat e_i^2 / (n - 1)}}{\sum_i \check D_i^2 / \sqrt{n}},
\qquad \hat e_i = \check Y_i - \hat\theta \check D_i .
$$
The standard error is the heteroskedasticity-robust sandwich of the
residual-on-residual regression. Nothing is added for the first stage;
that is oracle equivalence.

> **Algorithm (DML for the partially linear model).**
> 1. Split the indices into $K$ folds $I_1, \dots, I_K$.
> 2. For each $k$, fit $\hat\ell_{[k]}$ and $\hat m_{[k]}$ on the data
>    outside $I_k$; form $\check Y_i$ and $\check D_i$ for $i \in I_k$.
> 3. Regress $\check Y$ on $\check D$ by least squares; report the
>    robust standard error.

**In R.** `est_dml(model = "plr")` runs the three steps. Two checks tie
it to what you already know. With linear-regression nuisances fitted on
the full sample, it *is* least squares with the controls included
(classical FWL). The learners must be requested, because for a binary
treatment the default learner for $m$ is a logit:

```{r plm-fwl}
dat_lin <- data.frame(y = dat$y, d = dat$d, dat[, x_names])
fwl <- est_dml(dat_lin, y = "y", d = "d", x = x_names, cross_fit = FALSE,
               learner_l = lrn("regr.lm"), learner_m = lrn("regr.lm"))
c(dml_linear = fwl$estimate,
  ols        = coef(lm(y ~ ., data = dat_lin))[["d"]])
```

With cross-fitted forests it is the estimator of Section 1's fourth row:

```{r plm-forest}
set.seed(1)
fit_plm <- est_dml(
  dat, y = "y", d = "d", x = x_names,
  learner_l = lrn("regr.ranger", num.trees = 100),
  learner_m = lrn("classif.ranger", predict_type = "prob", num.trees = 100),
  folds = 5, seed = 1
)
fit_plm
```

**What $\theta$ means when the model is wrong.** Suppose the effect varies,
$Y = g(0, X) + D\,\tau(X) + \varepsilon$, so the PLM is misspecified. The
estimator still converges to the best linear predictor of $\tilde Y$ on
$\tilde D$. For binary $D$ that limit is
$$
\theta^{*} = \frac{\mathbb{E}\bigl[m(X)\,(1 - m(X))\,\tau(X)\bigr]}{\mathbb{E}\bigl[m(X)\,(1 - m(X))\bigr]},
$$
an average of $\tau(X)$ with **overlap weights** $m(X)(1 - m(X))$. Units
whose treatment is nearly determined by $X$ get almost no weight. This is
neither the ATE nor the ATT. With the true nuisances supplied, the
partially linear estimator lands on the overlap-weighted average and the
interactive estimator of the next subsection lands on the ATE:

```{r plm-heterogeneous}
hte <- simulate_hte_dgp(n = 5000, seed = 2)
w_overlap <- with(hte, m_true * (1 - m_true))
c(
  ATE               = mean(hte$tau_true),
  overlap_weighted  = with(hte, sum(w_overlap * tau_true) / sum(w_overlap)),
  plm_estimate      = est_dml(hte, "y", "d", l_hat = "l_true", m_hat = "m_true")$estimate,
  irm_estimate      = est_dml(hte, "y", "d", model = "irm",
                              p_hat = "m_true", mu0_hat = "g0_true", mu1_hat = "g1_true")$estimate
)
```

**When it breaks.**

- Effects are heterogeneous and the ATE or ATT is the question. Use the
  interactive model.
- $\mathbb{E}[\tilde D^2]$ is small: the covariates almost determine the
  treatment. The denominator is close to zero and the estimate is noisy.
  `est_dml()` reports this quantity under `diagnostics$identification`.
- $\hat m$ is poor. The bias is $-\theta_0\,\mathbb{E}[(\hat m - m_0)^2] / \mathbb{E}[\tilde D^2]$
  at second order (Section \@ref(sec-rates)); nothing protects against it.

## Interactive model: the doubly robust score {#sec-irm}

**Model.**
$$
Y = g(D, X) + \varepsilon, \quad \mathbb{E}[\varepsilon \mid D, X] = 0;
\qquad
D \in \{0, 1\}, \quad m(X) = \mathbb{P}(D = 1 \mid X).
$$
No restriction on how $D$ and $X$ interact. Targets: the ATE
$\theta = \mathbb{E}[g(1, X) - g(0, X)]$ and the ATT
$\mathbb{E}[g(1, X) - g(0, X) \mid D = 1]$.

**Three representations, one orthogonal.** Section 03 wrote the ATE three
ways. Only one of them can carry a machine learner.

```{r pipe-table-3, echo = FALSE}
tibble::tibble(
  c1 = c("Regression adjustment", "Inverse propensity weighting", "AIPW"),
  c2 = c("$\\mathbb{E}[g(1, X) - g(0, X)]$", "$\\mathbb{E}[Y\\, H(D, X)]$", "$\\mathbb{E}\\bigl[g(1, X) - g(0, X) + H(D, X)(Y - g(D, X))\\bigr]$"),
  c3 = c("$g$", "$m$", "$(g, m)$"),
  c4 = c("$\\mathbb{E}[\\Delta(1, X) - \\Delta(0, X)]$", "$-\\mathbb{E}\\bigl[Y\\,\\Delta(X)\\,\\bigl(\\tfrac{D}{m^2} + \\tfrac{1 - D}{(1 - m)^2}\\bigr)\\bigr]$", "$0$ in both directions"),
  c5 = c("no", "no", "yes")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Representation", "$\\theta = $", "Nuisance", "Derivative at the truth", "Orthogonal?")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

The AIPW representation is the other two glued together so that their
first-order errors cancel (Section \@ref(sec-orthogonality) and Appendix
\@ref(app-orthogonality)). Its remainder is the product of the two
nuisance errors (Section \@ref(sec-rates)): double robustness.

**Scores.** All three targets have scores linear in $\theta$,
$\psi = \psi^b - \psi^a \theta$, with $\bar p = \mathbb{P}(D = 1)$ and
$p_G = \mathbb{P}(G = 1)$ for a group indicator $G$:

```{r pipe-table-4, echo = FALSE}
tibble::tibble(
  c1 = c("ATE", "ATT", "GATE"),
  c2 = c("$g(1, X) - g(0, X) + H(D, X)\\,(Y - g(D, X))$", "$\\bigl(D - (1 - D)\\,\\tfrac{m(X)}{1 - m(X)}\\bigr)\\,\\bigl(Y - g(0, X)\\bigr) / \\bar p$", "$\\tfrac{G}{p_G}\\,\\psi^b_{\\text{ATE}}$"),
  c3 = c("$1$", "$D / \\bar p$", "$\\tfrac{G}{p_G}$")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Target", "$\\psi^b(W; \\eta)$", "$\\psi^a$")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

The ATT score needs only $m$ and $g(0, \cdot)$: the treated arm is never
imputed. The group score reweights the ATE score by group membership.

> **Algorithm (DML for the ATE in the interactive model).**
> 1. Split into $K$ folds.
> 2. For each $k$, fit $\hat g_{[k]}(0, \cdot)$ on untreated units outside
>    $I_k$, $\hat g_{[k]}(1, \cdot)$ on treated units outside $I_k$, and
>    $\hat m_{[k]}$ on all units outside $I_k$; clip $\hat m$ to
>    $[\epsilon, 1 - \epsilon]$.
> 3. For $i \in I_k$ compute
>    $\hat\varphi_i = \hat g(1, X_i) - \hat g(0, X_i) + \hat H_i\,(Y_i - \hat g(D_i, X_i))$.
> 4. $\hat\theta = \mathbb{E}_n[\hat\varphi]$ and
>    $\widehat{\mathrm{SE}} = \mathrm{sd}(\hat\varphi) / \sqrt{n}$.

**In R.** `est_dml(model = "irm")` runs the algorithm; `est_aipw()` is the
same estimator with the Section 03 interface, and the two agree to the
last digit on the same folds. On the Section 1 sample with forests:

```{r irm-forest}
learners_irm <- list(
  learner_p   = lrn("classif.ranger", predict_type = "prob", num.trees = 100),
  learner_mu0 = lrn("regr.ranger", num.trees = 100),
  learner_mu1 = lrn("regr.ranger", num.trees = 100)
)
set.seed(1)
fit_ate <- do.call(est_dml, c(list(dat, y = "y", d = "d", x = x_names,
                                   model = "irm", estimand = "ATE", seed = 1), learners_irm))
set.seed(1)
fit_att <- do.call(est_dml, c(list(dat, y = "y", d = "d", x = x_names,
                                   model = "irm", estimand = "ATT", seed = 1), learners_irm))
set.seed(1)
fit_aipw <- est_aipw(dat, y = "y", d = "d", x = x_names, seed = 1,
                     learner_p = learners_irm$learner_p,
                     learner_mu0 = learners_irm$learner_mu0,
                     learner_mu1 = learners_irm$learner_mu1)
rbind(
  ATE      = c(estimate = fit_ate$estimate,  std.error = fit_ate$std.error),
  ATT      = c(estimate = fit_att$estimate,  std.error = fit_att$std.error),
  est_aipw = c(estimate = fit_aipw$estimate, std.error = fit_aipw$std.error)
)
```

**When it breaks.**

- Weak overlap. $H$ divides by $\hat m$ and $1 - \hat m$. Clipping at
  $\epsilon$ caps the weights at $1/\epsilon$; trimming units outside
  $[\epsilon, 1 - \epsilon]$ changes the target population. Report both
  (Section 03).
- Both nuisances poor. The product $\|\hat m - m_0\|\,\|\hat g - g_0\|$
  must be $o(n^{-1/2})$; one accurate nuisance buys slack for the other,
  but not for both.
- Non-binary treatment. Use the partially linear model, or the
  generalizations in Section \@ref(sec-extensions).

## High-dimensional linear model: double lasso {#sec-double-lasso}

**Model.**
$$
Y = \alpha D + \beta' W + \varepsilon, \qquad D = \gamma' W + v,
$$
with $p$ controls, $p$ comparable to or larger than $n$, and
**approximate sparsity**: the sorted coefficients decay fast enough that a
few hundred of them carry the signal. This is the partially linear model
with linear nuisances and the lasso as the learner. It is where the DML
story started (Belloni, Chernozhukov, and Hansen 2014).

**The naive procedure and its failure.** *Single selection*: run a lasso
of $Y$ on $(D, W)$, keep the selected controls, refit by least squares,
report the usual standard error. The lasso keeps controls that predict
$Y$. It drops controls that predict $D$ strongly but $Y$ weakly. Those are
exactly the confounders that omitted-variable bias is made of. In moment
language the procedure solves $\mathbb{E}[(Y - \alpha D - \beta' W)\, D] = 0$
with an estimated $\beta$, and the derivative in the direction $\Delta$
is $-\mathbb{E}[W D]\,\Delta \ne 0$. Not orthogonal.

**Double lasso.** Partial out with the lasso on both sides.

1. Lasso $Y$ on $W$; residuals $\check Y$.
2. Lasso $D$ on $W$; residuals $\check D$.
3. Least squares of $\check Y$ on $\check D$; robust standard error.

Controls that matter for *either* equation survive in at least one
residual. That is the orthogonality of Section \@ref(sec-plm) again.
*Double selection* (least squares on $D$ and the union of the two selected
sets) is an equivalent variant.

**Cross-fitting or not.** With the theoretical plug-in penalty, sparsity
theory guarantees that the lasso does not overfit, and no sample
splitting is needed (Theorem 4.2.1 in the book). With a cross-validated
penalty that guarantee is gone, so cross-fit. There is a second
subtlety. The lasso *shrinks* the coefficients it keeps. Shrinkage is
nuisance error like any other, and Section \@ref(sec-rates) says it is
amplified by $1 / \mathbb{E}[\tilde D^2]$. Post-selection refitting
removes it; using the shrunken lasso predictions as $\hat\ell$ and
$\hat m$ does not. The simulation shows both effects.

**Simulation.** Example 4.3.1 of the book: $p = n = 100$,
$\beta_j = \gamma_j = 1/j^2$, $W \sim N(0, I_p)$, $v \sim N(0, 0.25^2)$,
$\alpha = 1$. Three estimators.

```{r pipe-table-5, echo = FALSE}
tibble::tibble(
  c1 = c("Single selection", "Post-double-selection", "CV lasso, cross-fitted"),
  c2 = c("lasso $Y$ on $(D, W)$; least squares on $D$ and the selected $W$", "lasso $Y$ on $W$ and $D$ on $W$; least squares on $D$ and the union of the selections", "\\texttt{est\\_dml()} with cross-validated lasso \\emph{predictions} as $\\hat\\ell$ and $\\hat m$")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Estimator", "Steps")) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  column_spec(2, width = "11.5cm")
```

```{r double-lasso-experiment, cache = TRUE, cache.extra = tools::md5sum(helper_path)}
lasso_results <- run_double_lasso_experiment(
  n_sims = n_sims_lasso, n = 100, p = 100, theta = 1, sd_v = 0.25, folds = 5, seed = 1
)
lasso_tab <- summarise_dml_experiment(lasso_results)
pick_lasso <- function(estimator, column) lasso_tab[[column]][lasso_tab$estimator == estimator]
msq_dtilde <- mean(lasso_results$msq_treatment_residual, na.rm = TRUE)
msq_m_err  <- mean(lasso_results$msq_m_error, na.rm = TRUE)
```

```{r tab-double-lasso, echo = FALSE}
kbl(lasso_tab, booktabs = TRUE, digits = 3, linesep = "", escape = FALSE,
    col.names = c("Estimator", "Bias", "SD", "RMSE", "Mean reported SE", "95\\% coverage"),
    caption = "Lasso-based estimators, true effect 1, p = n = 100, residual treatment standard deviation 0.25.") %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

```{r fig-double-lasso, fig.height = 3, fig.cap = "Single selection loses the confounders that predict the treatment. Post-double-selection keeps them and refits without shrinkage. Shrunken cross-validated lasso predictions fail here because the residual treatment variance is tiny."}
plot_dml_experiment(lasso_results, colors = lasso_colors, ncol = 3)
```

**Reading the results.**

- **Single selection** has a bias of
  `r fmt(pick_lasso("Single selection", "bias"))`. The treatment is
  unpenalized and almost collinear with the leading controls, so the
  outcome lasso keeps $D$ and drops most of those controls. Their
  effects, $\sum_j \beta_j\gamma_j$ over the dropped set, land on the
  coefficient of $D$ (Appendix \@ref(app-orthogonality)).
- **Post-double-selection** is centred (bias
  `r fmt(pick_lasso("Post-double-selection", "bias"))`) with coverage
  `r fmt(pick_lasso("Post-double-selection", "coverage"))`. The lasso for
  $D$ picks up the confounders the lasso for $Y$ misses, and the
  least-squares refit carries no shrinkage.
- **Cross-validated lasso predictions**, cross-fitted, are biased
  (`r fmt(pick_lasso("CV lasso, cross-fitted", "bias"))`)
  despite the orthogonal score and cross-fitting. The residual treatment
  variance in this design is $\mathbb{E}[\tilde D^2] \approx$
  `r fmt(msq_dtilde)`, and the lasso's squared error in $\hat m$ is
  about `r fmt(msq_m_err, 3)`. Both nuisances are shrunk toward zero in
  the same direction, so the remainder of Section \@ref(sec-rates),
  $\bigl(\mathbb{E}[(\hat\ell - \ell)(\hat m - m)] - \theta\,\mathbb{E}[(\hat m - m)^2]\bigr) / \mathbb{E}[\tilde D^2]$,
  is positive and multiplied by roughly `r round(1 / msq_dtilde)`. This
  is the weak-residual-variation failure of Section \@ref(sec-plm)
  combined with shrinkage.

Three lessons. Refit after selecting when the lasso partials out; that is
what post-lasso and the plug-in `rlasso` do. Read
$\mathbb{E}[\tilde D^2]$ before trusting any partialling-out estimate;
`est_dml()` reports it. And treat shrinkage as nuisance error: it counts
against the rate condition like any other.

**When it breaks.**

- Sparsity fails: many small coefficients that together matter. The
  lasso cannot approximate either nuisance, and no penalty choice helps.
  Move to nonlinear learners or richer dictionaries.
- Cross-validated penalties without cross-fitting: the overfitting bias
  of Section 1 returns.
- Perfect or near-perfect prediction of $D$ by $W$: the residual
  $\check D$ has little variation, and every nuisance error is amplified.

# Cross-fitting {#sec-crossfit}

## The problem it solves

Section \@ref(sec-skeleton) left one term open, the empirical process
term
$$
(b) = \sqrt{n}\,(\mathbb{E}_n - \mathbb{E})\bigl[\psi(\theta_0, \hat\eta) - \psi(\theta_0, \eta_0)\bigr].
$$
It is a centred average, but of a function that was chosen by looking
at the same data. There are two ways to make it vanish.

- **Restrict the learner.** If every function the learner can produce
  lies in a "simple" class (a Donsker class), a uniform central limit
  theorem applies and $(b) \to 0$. Parametric models and low-dimensional
  smooth nonparametrics qualify. Forests, boosting, neural networks, and
  the lasso with data-driven tuning do not.
- **Break the dependence.** Fit $\hat\eta$ on observations that are not
  used to evaluate the score. Then, conditional on the fitting data,
  $\hat\eta$ is a fixed function and $(b)$ is an ordinary centred average.

Cross-fitting takes the second route. It places no restriction on the
learner. Section 1's row "Orthogonal score, full-sample forests" shows
what happens when neither route is taken: the score is right, the
learner is adaptive, and $(b)$ does not vanish.

## The construction

> **Algorithm ($K$-fold cross-fitting).**
> 1. Split the indices $\{1, \dots, n\}$ into $K$ folds $I_1, \dots, I_K$
>    of similar size, stratified on $D$ when $D$ is binary.
> 2. For each $k$, fit the nuisances $\hat\eta_{[k]}$ on the complement
>    $I_k^c$.
> 3. For each $i \in I_k$, evaluate the score with
>    $\hat\eta_{[k(i)]}$, the fit that did not use observation $i$.
> 4. Solve $\mathbb{E}_n[\psi(W_i; \theta, \hat\eta_{[k(i)]})] = 0$ on the
>    pooled scores (or fold by fold; Section \@ref(sec-recipe)).

Every observation contributes a score computed from a model that never
saw it. Every nuisance fit uses a fraction $(K - 1)/K$ of the data. With
$K = 2$ this is the original "swap" scheme; $K = 5$ is the usual default;
larger $K$ helps in small samples (Remark 9.4.5 in the book).

## Why it works: the conditional argument

Fix a fold $k$ and condition on its training data $T_k = \{W_i : i \notin I_k\}$.
The fitted nuisance $\hat\eta_{[k]}$ is then a fixed function. For
$i \in I_k$ define
$$
Z_i = \psi(W_i; \theta_0, \hat\eta_{[k]}) - \psi(W_i; \theta_0, \eta_0).
$$
Conditional on $T_k$ the $Z_i$ are i.i.d., with mean
$\mu_k = \mathbb{E}[Z_i \mid T_k]$ and variance
$\sigma_k^2 = \mathbb{E}[Z_i^2 \mid T_k] - \mu_k^2$. The fold's
contribution to $(b)$ is $\sqrt{n_k}\,(\bar Z_k - \mu_k)$, and by
Chebyshev's inequality
$$
\mathbb{P}\bigl(\lvert \sqrt{n_k}\,(\bar Z_k - \mu_k)\rvert > t \,\big|\, T_k\bigr)
\;\le\; \frac{\sigma_k^2}{t^2}.
$$
The variance $\sigma_k^2$ is bounded by
$\mathbb{E}[(\psi(\hat\eta_{[k]}) - \psi(\eta_0))^2 \mid T_k]$, which goes
to zero whenever $\hat\eta_{[k]}$ is consistent and the score is smooth
in $\eta$. So each fold's centred average is $o_p(1)$ *after* the
$\sqrt{n}$ scaling, and summing over $K$ folds gives $(b) = o_p(1)$. The
means $\mu_k$ are exactly the bias term $(c)$ of Section
\@ref(sec-skeleton), which orthogonality handles. Appendix
\@ref(app-proof) writes this out.

Two features of the argument matter. Nothing in it depends on what the
learner is. And the only requirement on $\hat\eta$ is consistency, which
is weaker than the rate condition of Step 2; the rate is needed for
$(c)$, not for $(b)$.

## What it costs, and what it does not fix

**Costs.**

- $K$ nuisance fits instead of one, on smaller samples. With adaptive
  learners this is the price of valid inference, not a nuisance.
- The estimate depends on the random partition. Section
  \@ref(sec-recipe) removes that dependence with repeated partitions.

**What it does not fix.** Cross-fitting cures the own-observation bias.
It does nothing about the choice of moment. Table
\@ref(tab:tab-experiment) has the proof: the cross-fitted plug-in forest
keeps essentially the full bias of the in-sample plug-in. Sample
splitting with a non-orthogonal score is the "natural algorithm, draft 3"
of the Stanford lectures, and it still fails. Nor does cross-fitting
create identification, repair overlap, or shrink the rate condition.

**In R.** The residual diagnostics of `est_dml()` make the own-observation
bias visible on a single sample. In-sample forests produce residuals that
are too small, because the fitted values have absorbed part of the noise.

```{r crossfit-diagnostics}
set.seed(1)
fit_insample <- est_dml(
  dat, y = "y", d = "d", x = x_names,
  learner_l = lrn("regr.ranger", num.trees = 100),
  learner_m = lrn("classif.ranger", predict_type = "prob", num.trees = 100),
  cross_fit = FALSE
)
compare <- function(fit) {
  q <- fit$diagnostics$nuisance
  c(rmse_l = q$rmse[q$nuisance == "l_hat"],
    rmse_m = q$rmse[q$nuisance == "m_hat"],
    mean_sq_D_resid = fit$diagnostics$identification$mean_sq_treatment_residual,
    estimate = fit$estimate)
}
round(rbind(full_sample = compare(fit_insample), cross_fitted = compare(fit_plm)), 3)
```

The in-sample fit looks better on every nuisance metric. That is the
warning sign, not a success: residuals that small are impossible for a
treatment with this much unexplained variation. Cross-fitted residual
metrics are honest estimates of the out-of-sample nuisance error, and the
rate condition of Section \@ref(sec-rates) is about that error.

# The recipe in practice {#sec-recipe}

## The generic algorithm

> **Inputs.** Data $\{W_i\}_{i=1}^n$; an orthogonal score $\psi(W; \theta, \eta)$
> that identifies $\theta_0$; a learner for each component of $\eta$; the
> number of folds $K$; the number of repeated partitions $S$.
>
> **For each partition $s = 1, \dots, S$:**
> 1. Draw $K$ folds.
> 2. Fit the nuisances on each complement; predict on the held-out fold.
> 3. Form the cross-fitted scores $\psi(W_i; \theta, \hat\eta_{[k(i)]})$.
> 4. Solve $\mathbb{E}_n[\psi] = 0$ for $\hat\theta_s$; compute $\hat\sigma_s^2$.
>
> **Report** $\hat\theta_1, \hat\sigma_1$ if $S = 1$, otherwise the median
> aggregates of Section \@ref(sec-repeated).

`est_dml()` is this algorithm. The rest of the section is about the
choices inside it.

## Linear scores and the Jacobian {#sec-linear-scores}

Every score in this lecture is linear in $\theta$:
$$
\psi(W; \theta, \eta) = \psi^b(W; \eta) - \psi^a(W; \eta)\,\theta .
$$
Then step 4 has a closed form and so does the variance:
$$
\hat\theta = \frac{\sum_i \psi^b_i}{\sum_i \psi^a_i},
\qquad
\hat J = \frac{1}{n}\sum_i \psi^a_i,
\qquad
\hat\sigma^2 = \frac{1}{\hat J^2}\cdot\frac{1}{n - 1}\sum_i \bigl(\psi^b_i - \psi^a_i\,\hat\theta\bigr)^2,
\qquad
\widehat{\mathrm{SE}} = \hat\sigma / \sqrt{n}.
$$

```{r pipe-table-6, echo = FALSE}
tibble::tibble(
  c1 = c("Partialling out", "ATE (AIPW)", "ATT"),
  c2 = c("$\\tilde D^2$", "$1$", "$D / \\bar p$"),
  c3 = c("$\\tilde Y \\tilde D$", "$g(1, X) - g(0, X) + H(D, X)(Y - g(D, X))$", "$\\bigl(D - (1 - D)\\tfrac{m}{1 - m}\\bigr)(Y - g(0, X)) / \\bar p$"),
  c4 = c("$\\mathbb{E}[\\tilde D^2]$, the residual treatment variance", "$1$", "$1$")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Score", "$\\psi^a$", "$\\psi^b$", "$\\hat J$ estimates")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 9)
```

The Jacobian is where identification strength lives. For the ATE and ATT
it is one by construction. For the partially linear model it is the
residual treatment variance, and a small value inflates both the estimate
and every remainder term (Sections \@ref(sec-plm) and
\@ref(sec-double-lasso)).

## Pooled or fold-averaged solutions {#sec-solve}

Step 4 can be run in two ways.

- **Pooled** (`solve = "pooled"`, DML2 in the paper): stack the
  cross-fitted scores of all folds and solve once.
- **Fold average** (`solve = "fold_average"`, DML1): solve within each
  fold, then average the $K$ solutions.

Both use the variance formula above with the pooled scores. They agree
asymptotically. In finite samples the pooled solution is more stable,
because a fold with a small $\hat J_k$ does not get a full $1/K$ share of
the answer. Pooled is the default. The fold-level solutions are still
worth reading: their spread is a cheap diagnostic of how much the answer
depends on the partition.

```{r solve-compare}
set.seed(1)
fit_fold_avg <- est_dml(
  dat, y = "y", d = "d", x = x_names,
  learner_l = lrn("regr.ranger", num.trees = 100),
  learner_m = lrn("classif.ranger", predict_type = "prob", num.trees = 100),
  folds = 5, seed = 1, solve = "fold_average"
)
rbind(pooled = c(estimate = fit_plm$estimate, std.error = fit_plm$std.error),
      fold_average = c(estimate = fit_fold_avg$estimate, std.error = fit_fold_avg$std.error))
fit_plm$fold_estimates
```

## Repeated cross-fitting {#sec-repeated}

One partition is one random draw. Its estimate carries partition noise
on top of sampling noise. Repeating the whole procedure on $S$
independent partitions and aggregating removes the dependence on any one
of them. Chernozhukov et al. (2018, Section 3.4) recommend the median:
$$
\tilde\theta = \operatorname{median}_s\, \hat\theta_s,
\qquad
\tilde\sigma^2 = \operatorname{median}_s \Bigl(\hat\sigma_s^2 + (\hat\theta_s - \tilde\theta)^2\Bigr).
$$
The second term inside the median charges the variance for the spread
across partitions, so the interval does not pretend the partition was
fixed. The mean works too, with the same correction; the median is less
sensitive to one unlucky split.

```{r repeated}
set.seed(1)
fit_rep <- est_dml(
  dat, y = "y", d = "d", x = x_names,
  learner_l = lrn("regr.ranger", num.trees = 100),
  learner_m = lrn("classif.ranger", predict_type = "prob", num.trees = 100),
  folds = 5, n_rep = 5, seed = 1
)
fit_rep$repetitions
c(estimate = fit_rep$estimate, std.error = fit_rep$std.error)
```

Use $S \ge 5$ whenever the learners are flexible and $n$ is moderate.
With $S = 1$, two analysts with different seeds report different numbers
from the same data.

## Choosing and tuning the learners {#sec-learners}

The rate condition is about the *error* of each nuisance estimate. The
cross-fitted residuals measure that error out of sample, so they also
rank learners.

> **Selecting learners to minimise the DML bias.**
> For each candidate learner $j$ and each nuisance, compute the
> cross-fitted mean squared prediction error, for instance
> $\mathbb{E}_n[\check Y_j^2]$ and $\mathbb{E}_n[\check D_j^2]$.
> Pick the learner with the smallest error for each nuisance separately.
> Or form a linear ensemble of the candidates with the same criterion.

Three rules go with the box.

- **Few candidates.** Choosing among a handful of learners does not
  disturb the theory (Corollary 9.2.3 in the book). Searching over
  hundreds does; the validation data are then being used to fit.
- **Tune inside the training folds.** Cross-validation for
  hyperparameters must happen on the $K - 1$ training folds, never on
  the held-out fold. Data-driven tuning plus cross-fitting is fine;
  data-driven tuning without cross-fitting is Section 1's overfitting
  bias.
- **Different learners for different nuisances.** Nothing requires the
  same method for $\ell$ and $m$. A binary treatment often wants a
  calibrated classifier for $m$ and a flexible regression for $\ell$.

On the Section 1 sample, three candidates for the partially linear
nuisances: linear models, forests, and a lasso on the covariates and
their squares.

```{r learner-compare}
dat_sq <- dat
sq_names <- paste0(x_names, "_sq")
dat_sq[sq_names] <- dat[x_names]^2

candidates <- list(
  linear = list(x = x_names,
                l = lrn("regr.lm"),
                m = lrn("classif.log_reg", predict_type = "prob")),
  forest = list(x = x_names,
                l = lrn("regr.ranger", num.trees = 100),
                m = lrn("classif.ranger", predict_type = "prob", num.trees = 100)),
  lasso_squares = list(x = c(x_names, sq_names),
                       l = lrn("regr.cv_glmnet", nfolds = 5),
                       m = lrn("classif.cv_glmnet", predict_type = "prob", nfolds = 5))
)
compare_learners <- t(sapply(candidates, function(cand) {
  set.seed(1)
  fit <- est_dml(dat_sq, y = "y", d = "d", x = cand$x,
                 learner_l = cand$l, learner_m = cand$m, folds = 5, seed = 1)
  q <- fit$diagnostics$nuisance
  c(rmse_l = q$rmse[q$nuisance == "l_hat"], rmse_m = q$rmse[q$nuisance == "m_hat"],
    estimate = fit$estimate, std.error = fit$std.error)
}))
round(compare_learners, 3)
```

Read the table by column, not by row. The learner with the smallest
`rmse_l` is the right choice for $\ell$, the one with the smallest
`rmse_m` for $m$, and they need not be the same. Here the squares give
the lasso the curvature it needs, and it edges out the forest on both
nuisances. The two flexible candidates agree on the effect. The linear
candidate, with the largest errors, is the outlier.

## Folds, clipping, and trimming

- **Folds.** $K = 5$ is a sensible default. Use $K \ge 10$ when $n$ is
  a few hundred, so that each nuisance fit sees enough data. Stratify the
  folds on $D$ when $D$ is binary; `est_dml()` does this automatically.
- **Clipping.** In the interactive model, cap $\hat m$ to
  $[\epsilon, 1 - \epsilon]$ before forming $H$. The default
  $\epsilon = 0.01$ bounds the weights at 100. Clipping is a variance
  device and leaves the estimand alone.
- **Trimming.** Dropping units with $\hat m$ outside $[\epsilon, 1 - \epsilon]$
  changes the population the estimate refers to, as in Section 03.
  Report the trimmed share and, when possible, the estimate with and
  without trimming.
- **Binary treatment in the partially linear model.** Predict $D$ with a
  classifier and use its probability as $\hat m$; a regression forest on
  a 0/1 outcome works but is less well calibrated in the tails.

# Diagnostics and failure modes {#sec-diagnostics}

## What to report

A DML estimate is only as credible as the nuisances behind it. Report
the quantities below with every estimate; `est_dml()` returns all of them.

```{r pipe-table-7, echo = FALSE}
tibble::tibble(
  c1 = c("Cross-fitted RMSE and $R^2$ of each nuisance", "Residual treatment variance $\\mathbb{E}_n[\\tilde D^2]$", "Propensity range, clipped share, effective sample sizes", "Fold-level estimates", "Repetition estimates", "Estimates across two or three learners", "In-sample versus cross-fitted residuals", "Placebo outcomes, sensitivity analysis"),
  c2 = c("rate condition; learner choice", "identification strength (partially linear)", "overlap (interactive)", "dependence on one partition", "dependence across partitions", "model dependence", "overfitting of external predictions", "identification"),
  c3 = c("\\texttt{diagnostics\\$nuisance}", "\\texttt{diagnostics\\$identification}", "\\texttt{diagnostics\\$propensity}, \\texttt{diagnostics\\$weights}", "\\texttt{fold\\_estimates}", "\\texttt{repetitions}", "rerun; Section \\@ref(sec-learners)", "\\texttt{cross\\_fit = FALSE}; Section \\@ref(sec-crossfit)", "outside the estimator; Section 03")
) %>%
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, col.names = c("Diagnostic", "What it checks", "Where")) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  column_spec(c(1, 3), width = "5.2cm")
```

```{r diagnostics-print}
fit_plm$diagnostics$nuisance
fit_plm$diagnostics$identification$mean_sq_treatment_residual
fit_ate$diagnostics$propensity$clipped_summary
c(ess_treated = fit_ate$diagnostics$weights$ess_treated,
  ess_control = fit_ate$diagnostics$weights$ess_control)
```

## Failure modes

Each entry: the symptom, then the remedy.

- **Weak residual variation** (partially linear). $\mathbb{E}_n[\tilde D^2]$
  is a small fraction of $\mathrm{var}(D)$; the standard error is large;
  fold estimates scatter. Remedy: none inside the estimator. The
  covariates nearly determine the treatment. Accept the imprecision, or
  reconsider what variation identifies the effect.
- **Weak overlap** (interactive). $\hat m$ near 0 or 1 for many units;
  effective sample sizes far below $n_1$ and $n_0$. Remedy: clip, trim
  and report the trimmed share, or target the ATT if the untreated
  support is the problem.
- **Nuisances too poor.** Large cross-fitted RMSE; the residual bias of
  Section \@ref(sec-rates) is not negligible. Remedy: better learners
  chosen by cross-fitted error, more data, or a doubly robust score when
  one nuisance is much easier than the other.
- **Shrinkage with weak variation.** The lasso case of Section
  \@ref(sec-double-lasso). Remedy: refit after selection.
- **Partition dependence.** Fold estimates or repetitions disagree by
  more than a standard error. Remedy: repeated cross-fitting with median
  aggregation; larger $K$ in small samples.
- **In-sample nuisances from outside.** Residual metrics look too good;
  the estimate drifts toward the plug-in. Remedy: the contract for
  external predictions is out-of-fold, and the function cannot verify
  it. Check the fold structure of what you were handed.
- **Heterogeneous effects read as an ATE.** The partially linear
  coefficient is an overlap-weighted average (Section \@ref(sec-plm)).
  Remedy: the interactive model for a binary treatment.
- **Bad controls and hidden confounding.** No DML diagnostic detects
  them. Remedy: the placebo and sensitivity tools of Section 03, and the
  adjustment-set discipline of Section 03.
- **Too few observations per fold.** Nuisance fits fail or degrade.
  Remedy: $K \ge 10$, simpler learners, repeated splits.

The common thread: DML moves the difficulty from specifying a functional
form to learning nuisances well and saying how well they were learned.
The diagnostics are the second half of that bargain.

# Practice {#sec-practice}

## The replications

Two papers anchor this section. Their replication notebooks are
planned for `inst/replications/04_Doubly_Robust_and_Double_ML/` and are
not yet part of the package.

**Chernozhukov et al. (2018), Tables 1 and 2.**

- *Table 1, Pennsylvania reemployment bonus experiment.* The treatment is
  a randomized bonus offer; the outcome is unemployment duration. Because
  assignment is randomized, identification is not the question. The table
  teaches two things: DML with several learners agrees with the
  experimental benchmark, and the partially linear and interactive models
  give the same answer when the effect is close to constant.
- *Table 2, 401(k) eligibility and net financial assets.* Eligibility is
  observational; the design argues that income and a few demographics
  make it as good as random. The table compares lasso, trees, forests,
  boosting, and an ensemble in both models, with repeated splits. Watch
  the cross-fitted RMSE rows next to the estimates: the learners that
  predict better move together, and the estimate is far below the raw
  difference in means.

**Ellickson, Kar, and Reeder (2023), targeted promotions.** Emails are
targeted by a rule that depends on customer history, so raw comparisons
across emails are confounded. Stage one builds an orthogonal (AIPW)
score for each customer and each email-versus-baseline contrast with
forest nuisances; those are `est_dml(model = "irm")$score` plus the
estimate, one number per unit. Stage two regresses the scores on the
components that distinguish the emails. The replication teaches that DML
scores are data: once orthogonalized, they can be projected, averaged
within groups, or fed to a policy rule, and the projection inherits
valid inference.

## A reporting checklist

1. State the estimand and the model: a partially linear coefficient, an
   ATE, or an ATT.
2. Defend the adjustment set as pre-treatment and causally valid
   (Section 03).
3. Name the learner for each nuisance and how it was chosen; report the
   cross-fitted RMSE next to the estimate.
4. Report the residual treatment variance (partially linear) or the
   propensity range, clipped share, and effective sample sizes
   (interactive).
5. State $K$, the number of repeated partitions, the solution method,
   and any clipping or trimming.
6. Show the fold-level and repetition-level spread.
7. Compare two or three learners; a large gap is model dependence.
8. If external predictions were used, state that they are out-of-fold
   and how the folds were built.
9. Run a placebo outcome or a sensitivity analysis. DML does not deliver
   identification.
10. Interpret a partially linear coefficient as an overlap-weighted
    effect when heterogeneity is plausible.

# Extensions {#sec-extensions}

The recipe does not change across designs. Only the score does. Each
item below is one orthogonal score and one pair of nuisances.

- **Instrumental variables** (Section 05). Partially linear IV
  residualizes $Y$, $D$, and the instrument $Z$ on $X$, then runs
  two-stage least squares on the residuals. The interactive version
  ratios two AIPW-type scores, one for $Y$ and one for $D$, with the
  instrument's propensity in the denominators, and targets the LATE.
- **Difference-in-differences** (Section 07). The doubly robust DiD
  score applies the AIPW construction to the outcome *change*, with a
  propensity for belonging to the treated group. Conditional parallel
  trends replaces conditional ignorability.
- **Heterogeneous effects** (Section 08). The residuals of the
  partially linear model satisfy $\tilde Y \approx \tau(X)\tilde D$; the
  R-learner fits $\tau(X)$ by minimizing $(\tilde Y - \tau(X)\tilde D)^2$.
  The DR-learner regresses the AIPW score on $X$. Both reuse this
  lecture's nuisances.
- **Group and treated-population effects.** The ATT and group-ATE scores
  of Section \@ref(sec-irm); one estimator per group, same folds.
- **Automatic DML.** For any linear functional of the regression
  function there is a weight $\alpha(W)$, the Riesz representer, that
  plays the role of $H(D, X)$ in the AIPW score. Automatic DML learns
  $\alpha$ by machine learning instead of deriving it by hand, which
  extends the recipe to functionals for which no closed form exists.
- **Targeted maximum likelihood.** TMLE debiases the plug-in by
  refitting $\hat g$ along the direction $H(D, X)$ until the score has
  sample mean zero, then plugs in. It is first-order equivalent to DML
  with the same nuisances and also needs cross-fitting with adaptive
  learners.

# Appendix {#sec-appendix}

## Orthogonality calculations {#app-orthogonality}

**Notation.** $W = (Y, D, X)$. The nuisance is $\eta$, its true value
is $\eta_0$, and $\Delta = \eta - \eta_0$ is a fixed direction. The
moment is $M(\theta, \eta) = \mathbb{E}[\psi(W; \theta, \eta)]$ and
$$
\partial_\eta M(\theta_0, \eta_0)[\Delta]
= \frac{d}{dt}\, M(\theta_0, \eta_0 + t\Delta)\Big|_{t = 0} .
$$
$\|f\| = \sqrt{\mathbb{E}[f(W)^2]}$ is the $L^2$ norm. Two facts do
all the work. Under $\mathbb{E}[\varepsilon \mid D, X] = 0$, the outcome
error times any function of $(D, X)$ has mean zero. Under
$\mathbb{E}[D \mid X] = m_0(X)$, the treatment residual
$\tilde D = D - m_0(X)$ times any function of $X$ has mean zero.

### Plug-in score

Score $\psi = g(1, X) - g(0, X) - \theta$, nuisance $g$. With
$g = g_0 + t\Delta$,
$$
M(\theta_0, g_0 + t\Delta) = t\,\mathbb{E}[\Delta(1, X) - \Delta(0, X)],
\qquad
\partial_g M[\Delta] = \mathbb{E}[\Delta(1, X) - \Delta(0, X)] .
$$
The derivative is not zero. The moment is linear in $\Delta$, so there
is no second-order term to fall back on. The bias of the plug-in is
exactly $\mathbb{E}[\hat g(1, X) - g_0(1, X)] - \mathbb{E}[\hat g(0, X) - g_0(0, X)]$.
Whatever regularization does to the treatment contrast passes straight
through.

### Partialling-out score

Score $\psi = \bigl(Y - \ell(X) - \theta\,(D - m(X))\bigr)(D - m(X))$,
nuisance $\eta = (\ell, m)$. Write $U = Y - \ell_0(X) - \theta_0\tilde D$.
In the partially linear model $U = \varepsilon$, so
$\mathbb{E}[U \mid D, X] = 0$.

*Direction in $\ell$.* With $\ell = \ell_0 + t\Delta_\ell$ and $m = m_0$,
$$
\psi = \bigl(U - t\Delta_\ell(X)\bigr)\tilde D,
\qquad
\partial_\ell M[\Delta_\ell] = -\mathbb{E}[\Delta_\ell(X)\,\tilde D]
= -\mathbb{E}\bigl[\Delta_\ell(X)\,\mathbb{E}[\tilde D \mid X]\bigr] = 0 .
$$

*Direction in $m$.* With $m = m_0 + t\Delta_m$ and $\ell = \ell_0$,
$D - m = \tilde D - t\Delta_m$ and
$$
\psi = \bigl(U + t\theta_0\Delta_m\bigr)\bigl(\tilde D - t\Delta_m\bigr)
= U\tilde D - t\,U\Delta_m + t\,\theta_0\Delta_m\tilde D - t^2\theta_0\Delta_m^2 .
$$
The derivative at $t = 0$ is
$-\mathbb{E}[U\Delta_m(X)] + \theta_0\,\mathbb{E}[\Delta_m(X)\tilde D] = 0 + 0$.

*Exact remainder.* Put $\Delta_\ell = \hat\ell - \ell_0$ and
$\Delta_m = \hat m - m_0$, both fixed functions. Then
$Y - \hat\ell - \theta_0(D - \hat m) = U - \Delta_\ell + \theta_0\Delta_m$
and $D - \hat m = \tilde D - \Delta_m$, so
$$
\psi(W; \theta_0, \hat\eta)
= \bigl(U - \Delta_\ell + \theta_0\Delta_m\bigr)\bigl(\tilde D - \Delta_m\bigr) .
$$
Take expectations. The terms $U\tilde D$, $U\Delta_m$,
$\Delta_\ell\tilde D$ and $\Delta_m\tilde D$ vanish by the two facts.
What remains is
$$
\mathbb{E}[\psi(W; \theta_0, \hat\eta)]
= \mathbb{E}[\Delta_\ell\,\Delta_m] - \theta_0\,\mathbb{E}[\Delta_m^2] .
$$
The expansion is exact: the score is bilinear in the nuisance errors,
so there are no higher-order terms. Cauchy-Schwarz bounds it by
$\|\Delta_\ell\|\,\|\Delta_m\| + |\theta_0|\,\|\Delta_m\|^2$. The Jacobian
at $\hat\eta$ is $\mathbb{E}[(D - \hat m)^2] = \mathbb{E}[\tilde D^2] + \|\Delta_m\|^2$,
so with $\hat\eta$ held fixed the population solution is
$$
\theta(\hat\eta) - \theta_0
= \frac{\mathbb{E}[\Delta_\ell\Delta_m] - \theta_0\,\mathbb{E}[\Delta_m^2]}{\mathbb{E}[\tilde D^2] + \mathbb{E}[\Delta_m^2]} .
$$
Section \@ref(sec-rates) evaluates the second term of the numerator on
the fitted propensity. Section \@ref(sec-double-lasso) is this formula
with shrinkage in both errors.

### AIPW score for the ATE

Score $\psi = g(1, X) - g(0, X) + H(D, X)\,(Y - g(D, X)) - \theta$ with
$H(D, X) = D / m(X) - (1 - D) / (1 - m(X))$, nuisance $\eta = (g, m)$.
$H_0$ is $H$ at $m_0$.

*Inverse-propensity identity.* For any function $f(D, X)$,
$$
\mathbb{E}\Bigl[\frac{D}{m_0(X)}\, f(D, X)\Bigr]
= \mathbb{E}\Bigl[\frac{\mathbb{E}[D \mid X]}{m_0(X)}\, f(1, X)\Bigr]
= \mathbb{E}[f(1, X)],
\qquad
\mathbb{E}\Bigl[\frac{1 - D}{1 - m_0(X)}\, f(D, X)\Bigr] = \mathbb{E}[f(0, X)] .
$$

*Direction in $g$.* With $g = g_0 + t\Delta_g$ and $m = m_0$,
$$
\partial_g M[\Delta_g]
= \mathbb{E}[\Delta_g(1, X) - \Delta_g(0, X)] - \mathbb{E}[H_0(D, X)\,\Delta_g(D, X)] = 0
$$
by the identity with $f = \Delta_g$.

*Direction in $m$.* Only $H$ depends on $m$. With $m = m_0 + t\Delta_m$,
$$
\dot H(D, X) := \frac{d}{dt} H\Big|_{t = 0}
= -\Delta_m(X)\Bigl(\frac{D}{m_0(X)^2} + \frac{1 - D}{(1 - m_0(X))^2}\Bigr),
\qquad
\partial_m M[\Delta_m] = \mathbb{E}\bigl[\dot H(D, X)\,(Y - g_0(D, X))\bigr] = 0,
$$
because $\dot H$ is a function of $(D, X)$ and
$\mathbb{E}[Y - g_0(D, X) \mid D, X] = 0$.

*Inverse propensity weighting alone.* The score $Y H(D, X) - \theta$
has the same derivative with $Y$ in place of $Y - g_0(D, X)$:
$\mathbb{E}[\dot H\, Y] = \mathbb{E}[\dot H\, g_0(D, X)] \ne 0$. Subtracting
$g_0$ inside the weighted term is what makes the derivative vanish.
That is the middle row of the table in Section \@ref(sec-irm).

*Exact remainder.* With $\Delta_g = \hat g - g_0$, $\Delta_m = \hat m - m_0$
and $\hat H$ equal to $H$ at $\hat m$,
$$
\mathbb{E}[\psi(W; \theta_0, \hat\eta)]
= \mathbb{E}[\Delta_g(1, X) - \Delta_g(0, X)]
+ \underbrace{\mathbb{E}[\hat H\,(Y - g_0(D, X))]}_{= 0}
- \mathbb{E}[\hat H\,\Delta_g(D, X)] .
$$
Conditioning on $X$ in the last term gives
$\mathbb{E}[\tfrac{m_0}{\hat m}\Delta_g(1, X)] - \mathbb{E}[\tfrac{1 - m_0}{1 - \hat m}\Delta_g(0, X)]$.
Collecting terms,
$$
\mathbb{E}[\psi(W; \theta_0, \hat\eta)]
= \mathbb{E}\Bigl[\frac{\hat m - m_0}{\hat m}\,\Delta_g(1, X)\Bigr]
+ \mathbb{E}\Bigl[\frac{\hat m - m_0}{1 - \hat m}\,\Delta_g(0, X)\Bigr] .
$$
Again the expansion is exact. If $\hat m$ is clipped to
$[\epsilon, 1 - \epsilon]$, Cauchy-Schwarz gives
$$
\bigl|\mathbb{E}[\psi(W; \theta_0, \hat\eta)]\bigr|
\le \frac{1}{\epsilon}\,\|\hat m - m_0\|\,
\bigl(\|\hat g(1, \cdot) - g_0(1, \cdot)\| + \|\hat g(0, \cdot) - g_0(0, \cdot)\|\bigr) .
$$
The product form is double robustness. The factor $1 / \epsilon$ is the
theoretical price of weak overlap, and the reason `est_dml()` clips.

### ATT score

Score $\psi = \psi^b - \psi^a\theta$ with
$$
\psi^b = \Bigl(D - (1 - D)\frac{m(X)}{1 - m(X)}\Bigr)\frac{Y - g(0, X)}{\bar p},
\qquad
\psi^a = \frac{D}{\bar p},
$$
nuisance $\eta = (g(0, \cdot), m, \bar p)$. At the truth,
$\mathbb{E}[D\,(Y - g_0(0, X))] = \mathbb{E}[D\,(g_0(1, X) - g_0(0, X))] = \bar p\,\theta_0$
and $\mathbb{E}[(1 - D)\tfrac{m_0}{1 - m_0}(Y - g_0(0, X))] = 0$, so
$M(\theta_0, \eta_0) = 0$.

*Direction in $g(0, \cdot)$.* With $g(0, \cdot) = g_0(0, \cdot) + t\Delta$,
$$
\partial M[\Delta]
= -\frac{1}{\bar p}\,\mathbb{E}\Bigl[\Bigl(D - (1 - D)\frac{m_0}{1 - m_0}\Bigr)\Delta(X)\Bigr]
= -\frac{1}{\bar p}\bigl(\mathbb{E}[m_0\Delta] - \mathbb{E}[m_0\Delta]\bigr) = 0,
$$
using $\mathbb{E}[D \mid X] = m_0$ and $\mathbb{E}[1 - D \mid X] = 1 - m_0$.

*Direction in $m$.* Only the odds $m / (1 - m)$ depend on $m$. They
multiply $(1 - D)(Y - g_0(0, X))$, whose conditional mean given
$(D, X)$ is zero. The derivative is zero.

*Direction in $\bar p$.* The score is $\bar p^{-1}$ times a function
that does not involve $\bar p$, so the derivative is
$-M(\theta_0, \eta_0) / \bar p = 0$. In the estimator $\bar p$ cancels
between numerator and denominator, so its estimation is immaterial.

*Exact remainder.* With $\Delta_g = \hat g(0, \cdot) - g_0(0, \cdot)$ and
$\Delta_m = \hat m - m_0$,
$$
\mathbb{E}[\psi(W; \theta_0, \hat\eta)]
= \frac{1}{\bar p}\,\mathbb{E}\Bigl[\frac{\hat m - m_0}{1 - \hat m}\,\Delta_g(X)\Bigr] .
$$
Only the untreated outcome model enters, and only in product with the
propensity error. The ATT score is doubly robust in $(g(0, \cdot), m)$.

### Single and double selection in the linear model

Model $Y = \alpha D + \beta' W + \varepsilon$ and $D = \gamma' W + v$
with $\mathbb{E}[\varepsilon \mid D, W] = 0$, $\mathbb{E}[v \mid W] = 0$,
$\sigma_v^2 = \mathbb{E}[v^2]$, and $\Sigma_W = \mathbb{E}[WW']$. The
coefficient of $Y$ on $W$ alone is $\pi = \alpha\gamma + \beta$.

*Single selection is not orthogonal.* The refit after selecting on the
outcome equation solves $\mathbb{E}[(Y - \alpha D - \hat\beta' W)\, D] = 0$
in $\alpha$, with $\beta$ as the nuisance. In the direction $\Delta$,
$$
\partial_\beta M[\Delta] = -\mathbb{E}[D\,W']\,\Delta = -\gamma'\Sigma_W\Delta \ne 0
$$
unless $\gamma = 0$, that is, unless there is no confounding.

*One formula for both procedures.* Let $S$ be the set of controls kept
by a selection and refit by least squares, with $\Sigma_W = I$. Then the
partialling-out errors are $\Delta_{\ell, j} = -\pi_j$ and
$\Delta_{m, j} = -\gamma_j$ for $j \notin S$, and zero otherwise. The
exact formula of the partialling-out subsection gives
$$
\theta(\hat\eta) - \alpha
= \frac{\sum_{j \notin S}\pi_j\gamma_j - \alpha\sum_{j \notin S}\gamma_j^2}{\sigma_v^2 + \sum_{j \notin S}\gamma_j^2}
= \frac{\sum_{j \notin S}\beta_j\gamma_j}{\sigma_v^2 + \sum_{j \notin S}\gamma_j^2},
$$
which is also the limit of least squares of $Y$ on $D$ and $W_S$. It is
omitted-variable bias, summed over the dropped controls.

- *Single selection* takes $S = S_\ell$, the controls the outcome lasso
  keeps. A control with small $\beta_j$ and large $\gamma_j$ is dropped
  and contributes its full $\beta_j\gamma_j$. When $D$ is almost
  collinear with the leading controls, the lasso keeps $D$ and drops
  those controls too; the numerator then contains their large
  products.
- *Post-double-selection* takes $S = S_\ell \cup S_m$. Every dropped
  control has small $\beta_j$ and small $\gamma_j$, so each term of the
  numerator is a product of two small numbers. That is the remainder
  $\mathbb{E}[\Delta_\ell\Delta_m]$ in coordinates.
- *Shrunken predictions* keep the controls but shrink them:
  $\Delta_{m, j} = -s_j\gamma_j$ and $\Delta_{\ell, j} = -s_j'\pi_j$ with
  $s_j, s_j' \in (0, 1]$ on the kept set. The same formula applies. The
  numerator no longer vanishes on the kept controls, and the
  denominator stays close to $\sigma_v^2$. When $\sigma_v^2$ is small,
  as in Table \@ref(tab:tab-double-lasso), the ratio is large.

## Proof sketch of asymptotic normality {#app-proof}

This appendix fills in Section \@ref(sec-skeleton) for a linear score
with $K$-fold cross-fitting and the pooled solution. It follows
Chernozhukov et al. (2018, Theorems 3.1 and 3.2), simplified to a scalar
$\theta$.

### Setting

- $W_1, \dots, W_n$ are i.i.d. draws of $W$.
- The score is linear: $\psi(W; \theta, \eta) = \psi^b(W; \eta) - \psi^a(W; \eta)\,\theta$.
- The folds $I_1, \dots, I_K$ have $n_k = n / K$ observations each, and
  $K$ is fixed.
- $\hat\eta_{[k]}$ is fitted on $I_k^c$. For $i \in I_k$ write
  $\psi_i^a = \psi^a(W_i; \hat\eta_{[k]})$ and $\psi_i^b = \psi^b(W_i; \hat\eta_{[k]})$.
- The estimator is $\hat\theta = \sum_i \psi_i^b / \sum_i \psi_i^a$ with
  $\hat J = \mathbb{E}_n[\psi_i^a]$.

For a fixed function $\eta$, $M(\theta, \eta) = \mathbb{E}[\psi(W; \theta, \eta)]$
is the expectation over a fresh draw of $W$. $\mathbb{E}_{n,k}$ is the
average over $i \in I_k$. "Given $I_k^c$" means conditional on the
observations outside fold $k$, which fixes $\hat\eta_{[k]}$.

### Conditions

There are a set of functions $\mathcal{T}_n$ with
$\mathbb{P}(\hat\eta_{[k]} \in \mathcal{T}_n) \to 1$ for every $k$, and
sequences $r_n \to 0$ and $\lambda_n = o(n^{-1/2})$, such that for every
$\eta \in \mathcal{T}_n$:

- **(C1) Identification.** $M(\theta_0, \eta_0) = 0$ and
  $J_0 = \mathbb{E}[\psi^a(W; \eta_0)] \ne 0$.
- **(C2) Orthogonality.** $\partial_\eta M(\theta_0, \eta_0)[\eta - \eta_0] = 0$.
- **(C3) Second-order remainder.** With
  $f(t) = M(\theta_0, \eta_0 + t(\eta - \eta_0))$,
  $\sup_{t \in (0, 1)} |f''(t)| \le \lambda_n$.
- **(C4) Consistency.** $\|\psi(\cdot; \theta_0, \eta) - \psi(\cdot; \theta_0, \eta_0)\| \le r_n$
  and $\|\psi^a(\cdot; \eta) - \psi^a(\cdot; \eta_0)\| \le r_n$.
- **(C5) Moments.** $\sigma_0^2 = \mathbb{E}[\psi(W; \theta_0, \eta_0)^2]$
  is finite and positive, and $\mathbb{E}[\psi^a(W; \eta_0)^2] < \infty$.

Appendix \@ref(app-orthogonality) gives (C2) and (C3) for the scores of
this lecture: $\lambda_n$ is of order
$\|\hat\ell - \ell_0\|\,\|\hat m - m_0\| + \|\hat m - m_0\|^2$ for partialling
out and $\epsilon^{-1}\|\hat m - m_0\|\,\|\hat g - g_0\|$ for AIPW. Learners
whose $L^2$ error is $o(n^{-1/4})$ deliver $\lambda_n = o(n^{-1/2})$ and
$r_n \to 0$.

> **Theorem.** Under (C1) to (C5),
> $$
> \sqrt{n}\,(\hat\theta - \theta_0) \to N(0, V),
> \qquad V = \sigma_0^2 / J_0^2,
> $$
> and $\hat\sigma^2 = \hat J^{-2}\,\mathbb{E}_n\bigl[(\psi_i^b - \psi_i^a\hat\theta)^2\bigr] \to_p V$.

### Proof

**Step 0: representation.** By construction
$\sum_i (\psi_i^b - \psi_i^a\hat\theta) = 0$. Subtracting this from
$\sum_i (\psi_i^b - \psi_i^a\theta_0)$ gives
$\mathbb{E}_n[\psi(W_i; \theta_0, \hat\eta_{[k(i)]})] = (\hat\theta - \theta_0)\,\hat J$, so
$$
\sqrt{n}\,(\hat\theta - \theta_0)
= \frac{\sqrt{n}\,\mathbb{E}_n[\psi(W_i; \theta_0, \hat\eta_{[k(i)]})]}{\hat J} .
$$

**Step 1: a lemma for cross-fitted averages.** Let $h(W; \eta)$ satisfy
$\|h(\cdot; \eta) - h(\cdot; \eta_0)\| \le r_n$ for all $\eta \in \mathcal{T}_n$.
Then
$$
\mathbb{E}_{n,k}[h(W_i; \hat\eta_{[k]})] - \mathbb{E}_{n,k}[h(W_i; \eta_0)] = O_p(r_n) .
$$
*Proof.* Given $I_k^c$, the differences
$Z_i = h(W_i; \hat\eta_{[k]}) - h(W_i; \eta_0)$, $i \in I_k$, are i.i.d.
On the event $\hat\eta_{[k]} \in \mathcal{T}_n$,
$\mathbb{E}[Z_i^2 \mid I_k^c] \le r_n^2$. By Jensen,
$\bar Z_k^2 \le \mathbb{E}_{n,k}[Z_i^2]$, so
$\mathbb{E}[\bar Z_k^2 \mid I_k^c] \le r_n^2$ and Markov's inequality gives
$\mathbb{P}(|\bar Z_k| > t \mid I_k^c) \le r_n^2 / t^2$. Take expectations
and use $\mathbb{P}(\hat\eta_{[k]} \in \mathcal{T}_n) \to 1$. $\square$

The conditioning is the point. Because $\hat\eta_{[k]}$ was fitted on
$I_k^c$, it is a fixed function given $I_k^c$, and
$\mathbb{E}[Z_i^2 \mid I_k^c]$ is an ordinary population moment at that
function. Without cross-fitting $\hat\eta$ depends on $W_i$ and no such
bound is available.

**Step 2: the Jacobian.** Apply the lemma with $h = \psi^a$ and average
over the $K$ folds:
$\hat J = \mathbb{E}_n[\psi^a(W_i; \eta_0)] + O_p(r_n) = J_0 + o_p(1)$ by
the law of large numbers, (C4) and (C5).

**Step 3: decomposition of the numerator.** For each fold, with
$Z_i = \psi(W_i; \theta_0, \hat\eta_{[k]}) - \psi(W_i; \theta_0, \eta_0)$,
$$
\sqrt{n_k}\,\mathbb{E}_{n,k}[\psi(W_i; \theta_0, \hat\eta_{[k]})]
= \underbrace{\sqrt{n_k}\,\mathbb{E}_{n,k}[\psi(W_i; \theta_0, \eta_0)]}_{(a_k)}
+ \underbrace{\sqrt{n_k}\,\bigl(\mathbb{E}_{n,k}[Z_i] - \mathbb{E}[Z_i \mid I_k^c]\bigr)}_{(b_k)}
+ \underbrace{\sqrt{n_k}\,\mathbb{E}[Z_i \mid I_k^c]}_{(c_k)} .
$$
Since $n_k = n / K$,
$$
\sqrt{n}\,\mathbb{E}_n[\psi(W_i; \theta_0, \hat\eta_{[k(i)]})]
= \frac{1}{\sqrt K}\sum_{k = 1}^K \bigl[(a_k) + (b_k) + (c_k)\bigr] .
$$

**Step 4: the bias term $(c_k)$.** Let
$f(t) = M(\theta_0, \eta_0 + t(\hat\eta_{[k]} - \eta_0))$. Then
$\mathbb{E}[Z_i \mid I_k^c] = f(1) - f(0)$. By (C1), $f(0) = 0$. By (C2),
$f'(0) = 0$. Taylor's theorem with the Lagrange remainder gives
$f(1) = \tfrac12 f''(\bar t)$ for some $\bar t \in (0, 1)$, so by (C3)
$$
|(c_k)| \le \tfrac12\sqrt{n_k}\,\lambda_n = o(1) .
$$
Orthogonality removes the term of order $\sqrt{n}\,\|\hat\eta - \eta_0\|$.
The rate condition removes the term of order $\sqrt{n}\,\|\hat\eta - \eta_0\|^2$.

**Step 5: the empirical-process term $(b_k)$.** Given $I_k^c$, the $Z_i$
are i.i.d. with variance at most $\mathbb{E}[Z_i^2 \mid I_k^c] \le r_n^2$
on the event $\hat\eta_{[k]} \in \mathcal{T}_n$. So $(b_k)$ has
conditional mean zero and conditional variance at most $r_n^2$, and
Chebyshev's inequality gives
$$
\mathbb{P}\bigl(|(b_k)| > t \,\big|\, I_k^c\bigr) \le r_n^2 / t^2 \to 0 .
$$
Taking expectations, $(b_k) = o_p(1)$. This is the only step that uses
cross-fitting, and it uses only consistency, not the rate.

**Step 6: the oracle term and the limit.**
$K^{-1/2}\sum_k (a_k) = \sqrt{n}\,\mathbb{E}_n[\psi(W_i; \theta_0, \eta_0)]$
is $\sqrt{n}$ times an average of i.i.d. mean-zero variables with
variance $\sigma_0^2$. By the central limit theorem it converges to
$N(0, \sigma_0^2)$. Steps 0, 2, 4, 5 and Slutsky's lemma give
$$
\sqrt{n}\,(\hat\theta - \theta_0)
= \frac{N(0, \sigma_0^2) + o_p(1)}{J_0 + o_p(1)}
\;\to\; N\bigl(0, \sigma_0^2 / J_0^2\bigr) .
$$

**Step 7: the variance estimator.** Write
$\psi_i^b - \psi_i^a\hat\theta = \psi(W_i; \theta_0, \hat\eta_{[k(i)]}) - \psi_i^a(\hat\theta - \theta_0)$.
The second piece has mean square $O_p(n^{-1})$ by Step 6 and (C5). For
the first piece, the conditioning argument of Step 1 applied to $Z_i^2$
gives $\mathbb{E}_n[Z_i^2] = O_p(r_n^2)$, and the triangle inequality in
the empirical $L^2$ norm gives
$$
\Bigl|\sqrt{\mathbb{E}_n[\psi(W_i; \theta_0, \hat\eta_{[k(i)]})^2]}
- \sqrt{\mathbb{E}_n[\psi(W_i; \theta_0, \eta_0)^2]}\Bigr|
\le \sqrt{\mathbb{E}_n[Z_i^2]} = O_p(r_n) .
$$
So $\mathbb{E}_n[(\psi_i^b - \psi_i^a\hat\theta)^2] \to_p \sigma_0^2$ by the
law of large numbers, and $\hat\sigma^2 \to_p \sigma_0^2 / J_0^2 = V$ by
Step 2. $\square$

### Remarks

- **Where each ingredient enters.** Identification: Steps 2 and 4
  ($f(0) = 0$). Orthogonality: Step 4 ($f'(0) = 0$). The rate: Step 4
  ($\sqrt{n}\,\lambda_n \to 0$). Cross-fitting: Steps 1, 5 and 7, through
  the conditioning on $I_k^c$. Nuisance consistency: Steps 2, 5 and 7.
  Without cross-fitting, Step 5 needs a Donsker condition on the
  learner. Without orthogonality, $(c_k)$ is of order
  $\sqrt{n}\,\|\hat\eta - \eta_0\|$, which diverges for any learner slower
  than $n^{-1/2}$.
- **Fold average.** The fold-average solution replaces Step 0 by
  $\hat\theta_k = \sum_{I_k}\psi_i^b / \sum_{I_k}\psi_i^a$ and averages
  the $K$ solutions. The same steps apply fold by fold, with
  $\hat J_k \to_p J_0$, and the limit is unchanged.
- **Repeated partitions.** Steps 4 and 5 show that
  $\sqrt{n}\,(\hat\theta_s - \theta_0) = J_0^{-1}\sqrt{n}\,\mathbb{E}_n[\psi(W_i; \theta_0, \eta_0)] + o_p(1)$
  for every partition $s$, and the leading term does not depend on the
  partition. So $\sqrt{n}\,(\hat\theta_s - \hat\theta_{s'}) = o_p(1)$,
  and the mean or median over a fixed number $S$ of partitions has the
  same limit. The correction $(\hat\theta_s - \tilde\theta)^2$ in the
  median variance is $o_p(n^{-1})$ in theory; it matters in finite
  samples, where the $o_p(1)$ terms are not yet small.

## References

- Belloni, Alexandre, Victor Chernozhukov, and Christian Hansen. 2014. "Inference on Treatment Effects after Selection among High-Dimensional Controls." *Review of Economic Studies*.
- Chernozhukov, Victor, Denis Chetverikov, Mert Demirer, Esther Duflo, Christian Hansen, Whitney Newey, and James Robins. 2018. "Double/Debiased Machine Learning for Treatment and Structural Parameters." *The Econometrics Journal*.
- Chernozhukov, Victor, Christian Hansen, Nathan Kallus, Martin Spindler, and Vasilis Syrgkanis. 2026. *Applied Causal Inference Powered by ML and AI*. Online, version 0.1.2.
- Chernozhukov, Victor, Whitney K. Newey, and Rahul Singh. 2022. "Automatic Debiased Machine Learning of Causal and Structural Effects." *Econometrica*.
- Ellickson, Paul B., Wreetabrata Kar, and James C. Reeder III. 2023. "Estimating Marketing Component Effects: Double Machine Learning from Targeted Digital Promotions." *Marketing Science*.
- Kennedy, Edward H. 2023. "Towards Optimal Doubly Robust Estimation of Heterogeneous Causal Effects." *Electronic Journal of Statistics*.
- Nie, Xinkun, and Stefan Wager. 2021. "Quasi-Oracle Estimation of Heterogeneous Treatment Effects." *Biometrika*.
- Robins, James M., Andrea Rotnitzky, and Lue Ping Zhao. 1994. "Estimation of Regression Coefficients When Some Regressors Are Not Always Observed." *Journal of the American Statistical Association*.
- Robinson, Peter M. 1988. "Root-N-Consistent Semiparametric Regression." *Econometrica*.
- Sant'Anna, Pedro H. C., and Jun Zhao. 2020. "Doubly Robust Difference-in-Differences Estimators." *Journal of Econometrics*.
- Syrgkanis, Vasilis. *MS&E 228: Applied Causal Inference Powered by ML and AI.* Lecture slides, Stanford University.
- van der Laan, Mark J., and Daniel Rubin. 2006. "Targeted Maximum Likelihood Learning." *International Journal of Biostatistics*.
